### 2021-03-17 reports/manuscript.Rmd
beginning writing

---
title: "Associations of environmental and human factors with altered rodent populations in West Africa: A scoping review"
description: |
date: "`r Sys.Date()`"
author: 
  - first_name: "David"
    last_name: "Simons"
    url: https://www.dsimons.org
    orcid_id: 0000-0001-9655-1656
output: distill::distill_article
bibliography: "`r here::here('citations', 'manuscript_citations.bib')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Rodents (Rodentia) are abundant and diverse, representing around 40% of the total number of mammalian species [@american_society_of_mammologists_asm_2021]. Rodents typically demonstrate ‘fast’ life history strategies characterised by early maturation, short generation times, low juvenile and adult survival and high fecundity [@dobson_fast_2007]. There is a degree of between-species heterogeneity, with commensal species and those that are reservoirs of zoonoses having traits consistent with faster life histories (Han et al., 2015). Rodents often compete with humans for food produced in agricultural systems [@fiedler_rodent_1988], with 5-10% of rodent species classed as major pests. Commensal species thrive in human adapted landscapes nesting in houses and obtaining nutrition from human food storage and preparation areas. In sub-Saharan Africa, Mastomys spp. and Arvicanthis spp. are identified as causing a significant burden as agricultural pests [@stenseth_mice_2003]. Crop loss within Afro-Malagasy small-holder farming communities is high, with 15% of pre-harvest crop being lost to rodent pest activity [@swanepoel_systematic_2017].

Another important issue is the transmission of infectious diseases from rodents to humans. Zoonotic diseases (i.e. animal pathogens that impact human health) are transmitted via two pathways – a direct and an indirect pathway. The direct transmission pathway involves transmission of pathogens through rodents biting or scratching a human, or through contamination of food or water sources. The indirect pathway involves rodents acting as amplifying hosts for ectoparasites that themselves are vectors of zoonoses [@meerburg_rodent-borne_2009]. Several rodent-borne pathogens are global in their distribution, including bacteria such as Lyme disease (caused by *Borrelia burgdorferi*) and Leptospirosis (caused by *Leptospira spp.*) and parasites such as Toxoplasmosis (caused by *Toxoplasma gondii*) and Leishmaniasis (caused by *Leishmania spp.*). However, most rodent-borne pathogens have more limited geographic distributions, particularly viral diseases such as Hantavirus pulmonary syndrome (caused by Hantaviridae) concentrated in the Americas and Lassa fever (caused by Arenaviridae) limited to West Africa.

Given the significant economic and public health burden of rodents, monitoring rodent distributions is therefore an ongoing focus of research efforts across multiple scientific disciplines, including (but not limited to) conservation science, development studies and infectious disease epidemiology. The distribution of rodent populations can be monitored via rodent trapping efforts, with traps deployed in habitats expected to contain rodent species of interest.

There is increasing interest in the link between environmental and human factors and rodent populations to land use change and human population change. A recent review of studies included in the PREDICTS database [@hudson_predicts_2014] identified that land use change from primary (i.e. intact forests) to managed and urban systems is associated with a change in rodent assemblages towards species that are more likely to be hosts of zoonoses [@gibb_zoonotic_2020]. *Mastomys natalensis populations* in Tanzania, East Africa are sensitive to seasonal meteorological cycles, in years with below average rainfall and short wet seasons population density is lower [@makundi_reproduction_2007]. Future climate change projections include longer and wetter West African monsoon seasons which could lead to increased zoonotic disease host rodent populations [@akinsanola_projection_2019]. As West Africa contains several known rodent zoonotic reservoirs (e.g. *Mastomys natalensis* and *Arvicanthus niloticus*) and several potential reservoirs of zoonotic diseases (e.g. *Lemniscomys striatus*), this will be the geographic region of interest in this work.

To date there has been no comprehensive overview of rodent trapping studies conducted in West Africa. The aim of this scoping review is therefore to synthesise the knowledge on rodent distributions throughout West Africa obtained from trapping studies.

We specifically aim to address the following research questions :

  +	RQ1: In which West African countries or regions have rodent trapping studies been performed?
  + RQ2: What are the stated aims of these rodent trapping studies?
  + RQ3: Which rodent species are being targeted for trapping activities? Are species with high potential for hosting zoonotic pathogens targeted at higher rates?
  + RQ4: How is the type of land use and/or land use intensity classified in the literature and are classifications comparable across studies?
  +	RQ5: Is rodent trapping occurring in all habitat types or are some habitats targeted at higher rates?
  +	RQ6: Is there an association of land use change, human population change or the implementation of regulatory strategies with rodent species distributions?

## Methods

### Study design

  + Scoping review
  + Protocol pre-registered on the [Openscience Framework](...)

### Eligibility criteria

  + Studies were included if:
    + Reported rodent trapping activity (as opposed to hand captures)
    + Included trapping activity from at least one West African country: Benin, Burkina-Faso, Cape Verde, Gambia, Ghana, Guinea, Guinea-Bissau, Cote d'Ivoire, Liberia, Mali, Mauritania, Niger, Nigeria, Senegal, Sierra Leone and Togo
    + Trapping activity occurred within a stated area (e.g. national park, village, house)
    + Recorded as outcome genus or species of trapped individual
    + Were available in English
    + Published in a peer-reviewed journal, a pre-print manuscript or a report by a an identifiable organisation/consultancy

### Search strategy

The following terms were searched for in OVID Medline, Web of Science (Core collection and Zoological Record), JSTOR, BioOne, African Journals Online, Global Health and the pre-print servers BioRxiv and EcoEvoRxiv:

  1. Rodent
  2. West Africa
  3. 1. AND 2.

Other resources including the UN Official Documents System, Open Grey, AGRIS FAO and Google Scholar were searched using combinations of the following terms:

  1. Rodent
  2. Rodent Trap*
  3. West Africa

Additional articles were identified through references within included articles and reports known to the study team. Searches were completed on 2021-03-01.

### Selection of studies

One reviewer screened titles, abstracts and full texts against the inclusion criteria. A random subset of each of these (10%) were reviewed by a second  reviewer.

### Data extraction

Data were extracted by one reviewer. A second reviewer verified a subset (10%) of included studies. A data extraction form was maintained on a Google sheets document. Data extracted included i) study identifiers; ii) study aims; iii) trapping methodology; iv) geolocation data; v) method of speciation; vi) trapping locations and dates; vii) trapped species; viii) number of trap-nights and ix) microorganisms of interest.

## Results

A total of 2,200 records were identified, with 102 studies included in narrative synthesis (see Figure 1).

``` {r figure_1, echo = F, fig.cap = "Flow diagram of records returned from the search strategy. Reasons for exclusion of studies at the full text stage are given."}
knitr::include_graphics(here::here("figures", "figure_1.png"))
```

The earliest studies identified were from 1974 with the majority (62%) published since 2010. The included studies trapped in 16 countries, however, three countries were outside of the scope of this study (Cameroon, Chad and Morocco). At least a single trap site was recorded from 13 West African countries, with no rodent trapping reported from Gambia, Cape Verde or Togo.

Most studies (97, 95%) were conducted in a single country, three studies were conducted in two countries, three studies were conducted in five countries and one study was conducted across six countries. This lead to 124 activities being reported across the 13 West African countries. More trapping activity was reported from Senegal (26, 20%) than any other country. Ghana (16, 12%) and Guinea (15, 12%) were the next most frequently trapped countries, followed by Nigeria (11, 9%), Benin, Mali and Niger (all 9, 7%), Sierra Leone (8, 6%), Mauritania (5, 4%), Burkina Faso and Liberia (both 3, 2%), Guinea-Bissau had a single recorded trapping activity.

The trapping activity from the included studies was spread across 1,171 study sites. Twenty-four (24%) studies reported trapping only occurring at a single study site, 34 (33%) studies trapped at between two and five study sites the remaining 44 studies trapped at between six and 93 study sites.

### 2021-03-17 scripts/libraries.R

@@ -1,13 +1,25 @@
-library("tidyverse")
-library("magrittr")
-library("bib2df")
-library("here")
-library("googledrive")
-library("readxl")
-library("countrycode")
-library("ggimage")
-library("sf")
-library("OpenStreetMap")
-library("tmap")
-library("maptools")
-library("raster")
+if (!require("pacman")) install.packages("pacman")
+pkgs =
+  c("here",
+    "tidyverse",
+    "magrittr",
+    "bib2df",
+    "googledrive",
+    "readxl",
+    "countrycode",
+    "ggimage",
+    "sf",
+    "OpenStreetMap",
+    "tmap",
+    "maptools",
+    "raster",
+    "mapview",
+    "ggmap"
+  )
+pacman::p_load(pkgs, character.only = T)
+
+if(!exists("google_api")) {
+google_api <- rstudioapi::askForSecret("Google API Key")
+}
+
+source(here::here("scripts", "functions.R"))

### 2021-03-22 reports/manuscript.Rmd
descriptives cont

@@
   + Studies were included if:
-    + Reported rodent trapping activity (as opposed to hand captures)
+    + Trapping of small mammals including at least one species of Rodentia.
+    + Descriped trap used, length of trapping activity or location of trapping activity.
     + Included trapping activity from at least one West African country: Benin, Burkina-Faso, Cape Verde, Gambia, Ghana, Guinea, Guinea-Bissau, Cote d'Ivoire, Liberia, Mali, Mauritania, Niger, Nigeria, Senegal, Sierra Leone and Togo
-    + Trapping activity occurred within a stated area (e.g. national park, village, house)
     + Recorded as outcome genus or species of trapped individual
@@
     + Published in a peer-reviewed journal, a pre-print manuscript or a report by a an identifiable organisation/consultancy
+  + Studies were excluded if:
+    + Data were duplicated from a previously included study (i.e. secondary data analysis)
+    + No full text were available
 
@@
 
+### Study aims
+
+Where explicity stated the study aims were extracted.
+For those with no recorded aims the specific aims were inferred from the background or conclusions of manuscript or report.
+The aims of the studies were broadly categorised as pertaining to rodent ecology research or studies on risk of zoonoses.
+Within these subdivisions the aim were further subdivided into the following categories:
+
+  + Rodent ecology
+    + Cat 1
+    + Cat 2
+  + Zoonoses risk
+    + Cat 1
+    + Cat 2
+
+### Trapping methodology
+
+The brand or description of the rodent trap was obtained from each study that reported this. For those using multiple devices all types were recorded.
+
+### Geolocation of trapping activity
+
+GPS locations were extracted for the most precise location presented i.e.
+trap, trap-line, study-site or study region.
+Locations were extracted in the format reported and converted to decimal degrees.
+Where no GPS location is reported coordinates will be matched to study locations using the National Geospatial Intelligence Agency NGA GEOnet Names Server [] based on study site name/region and through maps presented in the manuscript.
+
 ## Results
 
-A total of 2,200 records were identified, with 102 studies included in narrative synthesis (see Figure 1).
+A total of 2,200 records were identified, with 101 studies included in narrative synthesis (see Figure 1).
 
@@
 The earliest studies identified were from 1974 with the majority (62%) published since 2010.
+Most of the included studies were classified as rodent ecology research (54%) with the remainder studies on the risk of zoonoses (46%).
+
+  + Aims
+  + Published/field
+  + Trap design
+  + Trap technique
+  + Trap nights/effort
+
 The included studies trapped in 16 countries, however, three countries were outside of the scope of this study (Cameroon, Chad and Morocco).
@@
 
-Most studies (97, 95%) were conducted in a single country, three studies were conducted in two countries, three studies were conducted in five countries and one study was conducted across six countries.
-This lead to 124 activities being reported across the 13 West African countries.
+Most studies (94, 95%) were conducted in a single country, three studies were conducted in two countries, three studies were conducted in five countries and one study was conducted across six countries.
+This lead to 124 trapping activities being reported across the 13 West African countries.
 More trapping activity was reported from Senegal (26, 20%) than any other country.
-Ghana (16, 12%) and Guinea (15, 12%) were the next most frequently trapped countries, followed by Nigeria (11, 9%), Benin, Mali and Niger (all 9, 7%), Sierra Leone (8, 6%), Mauritania (5, 4%), Burkina Faso and Liberia (both 3, 2%), Guinea-Bissau had a single recorded trapping activity.
+Ghana (16, 12%) and Guinea (15, 12%) were the next most frequently trapped countries, followed by Nigeria (11, 9%), Benin, Mali and Niger (all 9, 7%), Sierra Leone (8, 6%), Mauritania and the Ivory Coast (both 5, 4%), Burkina Faso and Liberia (both 3, 2%), Guinea-Bissau had a single recorded trapping activity.
 
@@
 Twenty-four (24%) studies reported trapping only occurring at a single study site, 34 (33%) studies trapped at between two and five study sites the remaining 44 studies trapped at between six and 93 study sites.
+Trap sites were mapped to level one country administrative regions and density of sites per million square kilometers was calculated (see Figure 2).
+The areas with highest trapping densities included the capital cities of Niger, Sierra Leone, Senegal, Benin, Mali, Ghana and the Ivory Coast.
+Of the areas not incorporating capital cities Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore) and Eastern Sierra Leone were most intensively studied.
+
+``` {r figure_2, echo = F, fig.cap = "Density of trap sites per thousand km^2^", layout = "l-page"}
+knitr::include_graphics(here::here("figures", "static_site_density.png"))
+```
+
+  + Habitat classification
+  + Species classification
+  + Species trapped
+  + Pathogens

### 2021-03-22 scripts/descriptive.R

@@ -2,6 +2,8 @@ source(here::here("scripts", "libraries.r"))
 studies <- read_rds(here("data_clean", "studies.rds"))
 rodent_data <- read_rds(here("data_clean", "rodent_df.rds"))
 
+
+# Publication year --------------------------------------------------------
 ggplot(studies) +
   geom_bar(aes(x = year_publication)) +
   theme_minimal() +
@@ -10,21 +12,35 @@ ggplot(studies) +
        title = "Studies reporting rodent trapping in West African countries",
        caption = paste("N =", length(unique(studies$unique_id)), sep = " "))
 
+
+# Study aim ---------------------------------------------------------------
+table(studies$aim)
+ecology_studies <- studies %>% filter(aim == "Ecology")
+zoonoses_studies <- studies %>% filter(aim == "Zoonoses risk")
+
+
+# Trap type and setup -----------------------------------------------------
+trap_type <- studies %>%
+  separate(col = trap_types, into = c("trap_1", "trap_2", "trap_3", "trap_4"), sep = ", ", remove = T) %>%
+  pivot_longer(cols = c("trap_1", "trap_2", "trap_3", "trap_4"), values_to = "trap_type") %>%
+  drop_na(trap_type)
+
+# Study location ----------------------------------------------------------
 countries <- studies %>%
   full_join(., rodent_data %>%
               distinct(unique_id, country),
             by = "unique_id") %>%
   distinct(unique_id, country)
 
-countries$iso2 <- countrycode(as.character(countries$country), "country.name", "iso2c")
+countries$iso3 <- countrycode(as.character(countries$country), "country.name", "iso3c")
 
 countries %>%
-  group_by(country, iso2) %>%
+  group_by(country, iso3) %>%
   summarise(n = n()) %>%
   drop_na(country) %>%
   ggplot(aes(x = reorder(country, n), y = n)) +
   geom_col() +
-  geom_flag(aes(image = iso2, y = -2)) +
+  #geom_flag(aes(image = iso3, y = -2)) +
   coord_flip() +
   theme_minimal() +
   labs(
@@ -34,7 +50,7 @@ countries %>%
     caption = paste("N =", length(unique(studies$unique_id)), sep = " ")
     )
 
-ggsave(plot = last_plot(),filename = here("figures", "studies_country.png"), device = "png")
+#ggsave(plot = last_plot(),filename = here("figures", "studies_country.png"), device = "png")
 
 a <- countries %>%
   filter(!country %in% c("Cameroon", "Chad", "Morocco")) %>%

### 2021-03-22 scripts/download_spatial.R

@@ -1,0 +1,55 @@
+source(here::here("scripts", "libraries.r"))
+
+countries <- rodent_spatial %>%
+  drop_na(country) %>%
+  mutate(iso3 = countrycode(as.character(country), "country.name", "iso3c"))
+countries <- unique(countries$iso3)
+# Level 0 -----------------------------------------------------------------
+#WA
+BEN_0 <- getData("GADM", country = "Benin", level = 0, path = here("data_download", "admin_spatial"))
+BFA_0 <- getData("GADM", country = "BFA", level = 0, path = here("data_download", "admin_spatial"))
+CPV_0 <- getData("GADM", country = "CPV", level = 0, path = here("data_download", "admin_spatial"))
+GMB_0 <- getData("GADM", country = "GMB", level = 0, path = here("data_download", "admin_spatial"))
+GHA_0 <- getData("GADM", country = "Ghana", level = 0, path = here("data_download", "admin_spatial"))
+GIN_0 <- getData("GADM", country = "Guinea", level = 0, path = here("data_download", "admin_spatial"))
+GNB_0 <- getData("GADM", country = "GNB", level = 0, path = here("data_download", "admin_spatial"))
+CIV_0 <- getData("GADM", country = "CIV", level = 0, path = here("data_download", "admin_spatial"))
+LBR_0 <- getData("GADM", country = "LBR", level = 0, path = here("data_download", "admin_spatial"))
+MLI_0 <- getData("GADM", country = "Mali", level = 0, path = here("data_download", "admin_spatial"))
+MRT_0 <- getData("GADM", country = "Mauritania", level = 0, path = here("data_download", "admin_spatial"))
+NER_0 <- getData("GADM", country = "Niger", level = 0, path = here("data_download", "admin_spatial"))
+NGA_0 <- getData("GADM", country = "Nigeria", level = 0, path = here("data_download", "admin_spatial"))
+SEN_0 <- getData("GADM", country = "Senegal", level = 0, path = here("data_download", "admin_spatial"))
+SLE_0 <- getData("GADM", country = "SL", level = 0, path = here("data_download", "admin_spatial"))
+TGO_0 <- getData("GADM", country = "Togo", level = 0, path = here("data_download", "admin_spatial"))
+
+#non-WA
+TCD_0 <- getData("GADM", country = "Chad", level = 0, path = here("data_download", "admin_spatial"))
[…]

### 2021-03-22 scripts/mapping_traps.R

@@ -1,5 +1,23 @@
 source(here::here("scripts", "libraries.r"))
 
+all_countries <- c("BEN", "BFA", "CIV", "CMR", "CPV", "DZA", "ESH", "GHA",
+                   "GIN", "GMB", "GNB", "LBR", "MAR", "MLI", "MRT", "NER",
+                   "NGA", "SEN", "SLE", "TCD", "TGO")
+wa_countries <- c("BEN", "BFA", "CIV", "CPV", "ESH", "GHA",
+                  "GIN", "GMB", "GNB", "LBR", "MLI", "MRT",
+                  "NER", "NGA", "SEN", "SLE", "TGO")
+no_data_countries <- c("GMB", "TGO", "CPV")
+
+level_0 <- read_rds(here("data_download", "admin_spatial", "level_0_admin.rds"))
+list2env(level_0, envir = .GlobalEnv)
+level_0 <- do.call(rbind.SpatialPolygonsDataFrame, level_0)  %>%
+  st_as_sf()
+
+level_1 <- read_rds(here("data_download", "admin_spatial", "level_1_admin.rds"))
+list2env(level_1, envir = .GlobalEnv)
+level_1_all <- do.call(rbind.SpatialPolygonsDataFrame, level_1) %>%
+  st_as_sf()
+
 rodent_spatial <- read_rds(here("data_clean", "rodent_spatial.rds"))
 bbox_rodent <- st_bbox(rodent_spatial)
 
@@ -13,55 +31,27 @@ bbox_rodent[4] <- bbox_rodent[4] + (0.1 * ycrange) # ymax - top
 bbox_rodent <- bbox_rodent %>%
   st_as_sfc()
 
-countries <- rodent_spatial %>%
-  drop_na(country)
-countries <- as.character(unique(countries$country))
-
-#WA
-benin <- getData("GADM", country = "Benin", level = 0, path = here("data_download", "admin_spatial"))
-burkina_faso <- getData("GADM", country = "BFA", level = 0, path = here("data_download", "admin_spatial"))
-cape_verde <- getData("GADM", country = "CPV", level = 0, path = here("data_download", "admin_spatial"))
-gambia <- getData("GADM", country = "GMB", level = 0, path = here("data_download", "admin_spatial"))
-ghana <- getData("GADM", country = "Ghana", level = 0, path = here("data_download", "admin_spatial"))
[…]

### 2021-03-22 reports/manuscript.Rmd
starting to describe trapping methodology

@@
 
+### Genus and species definitions
+
+Designations of species change over time as new evidence emerges. As included studies span 45 years disambiguation of historical species names was performed using Taxize...
+
 ## Results
 
-A total of 2,200 records were identified, with 101 studies included in narrative synthesis (see Figure 1).
+A total of 2,200 records were identified, with 101 studies included in narrative synthesis (see Figure \@ref(fig:prisma)).
 
-``` {r figure_1, echo = F, fig.cap = "Flow diagram of records returned from the search strategy. Reasons for exclusion of studies at the full text stage are given."}
+``` {r prisma, fig.cap = "Flow diagram of records returned from the search strategy. Reasons for exclusion of studies at the full text stage are given.", echo = F}
 knitr::include_graphics(here::here("figures", "figure_1.png"))
@@
   + Published/field
-  + Trap design
-  + Trap technique
-  + Trap nights/effort
+
+### Trap types
+
+Of the included studies 15 did not state the type of trap used with three studies reporting supplementary hand trapping of rodents alongside trap use.
+For the remaining studies three used more than four types of rodent traps, seven used three types, 31 used two types the remaining 45 studies used a single trap type.
+The most commonly used trap was the Sherman trap (67%) [] which is available in several sizes, locally made wire mesh traps were used in 29 studies (34%), less commonly used trap types included the Museum special, Tomahawk, Victor, Snaptraps and Firobind traps.
+
+### Trap setup
+
+Thirty-three studies did not describe the setup of traps.
+Twenty five studies placed transect lines of traps through study sites with thirteen studies placing traps along a line but not clearly describing a transect across a defined site.
+Sixteen studies purposefully sample houses, with four further studies purposefully placing traps in targeted locationes.
+Nine studies placed traps in a grid across single or multiple habitat types.
+Three studies described a stratified or randomised approach to placing traps.
+Trap setup was typically reported for ecology studies (80%) with transect lines the most commonly used approach.
+Trap setup was reported in 53% of zoonoses risk studies with purposeful placement of traps in houses the most common approach (20%).
+
+### Trapping effort
+
+Trapping effort was variably reported in included studies.
+Thirty studies did not report trapping effort.
+Thirty-six studies reported a measure of trapping effort at the same level of detail at which they reported rodent captures.
+Thirty-five studies reported incomplete data on trapping effort (i.e.
+number of study nights or total number of trap-nights).
+Trapping effort was more completely reported from ecology studies (53% complete information and 20% no information) compared to zoonoses risk studies (15% complete information and 41% no information).
+
+### Location of studies
 
@@
 
-Most studies (94, 95%) were conducted in a single country, three studies were conducted in two countries, three studies were conducted in five countries and one study was conducted across six countries.
+Most studies (95%) were conducted in a single country, three studies were conducted in two countries, three studies were conducted in five countries and one study was conducted across six countries.
 This lead to 124 trapping activities being reported across the 13 West African countries.
-More trapping activity was reported from Senegal (26, 20%) than any other country.
-Ghana (16, 12%) and Guinea (15, 12%) were the next most frequently trapped countries, followed by Nigeria (11, 9%), Benin, Mali and Niger (all 9, 7%), Sierra Leone (8, 6%), Mauritania and the Ivory Coast (both 5, 4%), Burkina Faso and Liberia (both 3, 2%), Guinea-Bissau had a single recorded trapping activity.
+More trapping activity was reported from Senegal (20%) than any other country.
+Ghana (12%) and Guinea (12%) were the next most frequently trapped countries, followed by Nigeria (9%), Benin, Mali and Niger (7%), Sierra Leone (6%), Mauritania and the Ivory Coast (4%), Burkina Faso and Liberia (2%), Guinea-Bissau had a single recorded trapping activity.
 
@@
 Twenty-four (24%) studies reported trapping only occurring at a single study site, 34 (33%) studies trapped at between two and five study sites the remaining 44 studies trapped at between six and 93 study sites.
-Trap sites were mapped to level one country administrative regions and density of sites per million square kilometers was calculated (see Figure 2).
+Trap sites were mapped to level one country administrative regions and density of sites per million square kilometers was calculated (see Figure \@ref(fig:density)).
 The areas with highest trapping densities included the capital cities of Niger, Sierra Leone, Senegal, Benin, Mali, Ghana and the Ivory Coast.
@@
 
-``` {r figure_2, echo = F, fig.cap = "Density of trap sites per thousand km^2^", layout = "l-page"}
+``` {r density, fig.cap = "Density of trap sites per thousand km^2^", echo = F, layout = "l-page"}
 knitr::include_graphics(here::here("figures", "static_site_density.png"))
@@
   + Habitat classification
-  + Species classification
+
+### Species identification
+
+Nine included studies did not describe the method used to identify trapped rodents.
+Most studies (52%) relied solely on morphometric identification against taxonomic keys or museum voucher specimens.
+Eight studies described molecular identification of rodents with 30% using a combination of morphological and molecular techniques.
+In more recently published studies (since 2010) 54% used molecular techniques to assist with speciation, compared with less than 16% in studies published before 2010.
+
+### Reported species and species abundance
+
   + Species trapped

### 2021-03-22 scripts/libraries.R

@@ -14,7 +14,9 @@ pkgs =
     "maptools",
     "raster",
     "mapview",
-    "ggmap"
+    "ggmap",
+    "parzer",
+    "distill"
   )
 pacman::p_load(pkgs, character.only = T)
 

### 2021-03-24 reports/manuscript.Rmd
updated manuscript

@@
 
+From the included studies in 13 West African countries 131 distinct species were trapped.
+The majority were from the order Rodentia (102) with Muridae (83) being the largest family of rodents, followed by Gliridae (6), Nesomyidae (4), Sciuridae (3), Ctenodactylidae (2) and Anomaluridae, Dipodidae, Hystricidae and Thryonomidae (all 1).
+The next most represented orders were the Soricomorpha (25), Erinaceomorpha (3) and a single Afrosoricida.
+
   + Species trapped

### 2021-03-24 reports/manuscript.Rmd
mapping rodents

--- 2021-03-22 reports/manuscript.Rmd
@@
 The earliest studies identified were from 1974 with the majority (62%) published since 2010.
-Most of the included studies were classified as rodent ecology research (54%) with the remainder studies on the risk of zoonoses (46%).
+Most of the included studies were classified as rodent ecology research (54%) with the remainder studies on zoonotic diseases (46%).
 
@@
   + Published/field
+
+### Location of studies
+
+The included studies trapped in 16 countries, however, three countries were outside of the scope of this study (Cameroon, Chad and Morocco).
+At least a single trap site was recorded from 13 West African countries, with no rodent trapping reported from Gambia, Cape Verde or Togo.
+
+Most studies (95%) were conducted in a single country, three studies were conducted in two countries, three studies were conducted in five countries and one study was conducted across six countries.
+This lead to 124 trapping activities being reported across the 13 West African countries.
+More trapping activity was reported from Senegal (20%) than any other country.
+Ghana (12%) and Guinea (12%) were the next most frequently trapped countries, followed by Nigeria (9%), Benin, Mali and Niger (7%), Sierra Leone (6%), Mauritania and the Ivory Coast (4%), Burkina Faso and Liberia (2%), Guinea-Bissau had a single recorded trapping activity.
+
+The trapping activity from the included studies was spread across 1,171 study sites.
+Twenty-four (24%) studies reported trapping only occurring at a single study site, 34 (33%) studies trapped at between two and five study sites the remaining 44 studies trapped at between six and 93 study sites.
+Trap sites were mapped to level one country administrative regions and density of sites per million square kilometers was calculated (see Figure \@ref(fig:density)).
+The areas with highest trapping densities included the capital cities of Niger, Sierra Leone, Senegal, Benin, Mali, Ghana and the Ivory Coast.
+Of the areas not incorporating capital cities Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore) and Eastern Sierra Leone were most intensively studied.
+Mapping to level two administrative regions is displayed in Supplementary figure \@ref(fig:supfig1).
+
+``` {r density, fig.cap = "Density of trap sites per thousand km^2^", echo = F, layout = "l-page"}
+knitr::include_graphics(here::here("figures", "static_site_density.png"))
+```
 
@@
 
-### Location of studies
-
-The included studies trapped in 16 countries, however, three countries were outside of the scope of this study (Cameroon, Chad and Morocco).
-At least a single trap site was recorded from 13 West African countries, with no rodent trapping reported from Gambia, Cape Verde or Togo.
-
-Most studies (95%) were conducted in a single country, three studies were conducted in two countries, three studies were conducted in five countries and one study was conducted across six countries.
-This lead to 124 trapping activities being reported across the 13 West African countries.
-More trapping activity was reported from Senegal (20%) than any other country.
-Ghana (12%) and Guinea (12%) were the next most frequently trapped countries, followed by Nigeria (9%), Benin, Mali and Niger (7%), Sierra Leone (6%), Mauritania and the Ivory Coast (4%), Burkina Faso and Liberia (2%), Guinea-Bissau had a single recorded trapping activity.
-
-The trapping activity from the included studies was spread across 1,171 study sites.
-Twenty-four (24%) studies reported trapping only occurring at a single study site, 34 (33%) studies trapped at between two and five study sites the remaining 44 studies trapped at between six and 93 study sites.
-Trap sites were mapped to level one country administrative regions and density of sites per million square kilometers was calculated (see Figure \@ref(fig:density)).
-The areas with highest trapping densities included the capital cities of Niger, Sierra Leone, Senegal, Benin, Mali, Ghana and the Ivory Coast.
-Of the areas not incorporating capital cities Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore) and Eastern Sierra Leone were most intensively studied.
-
-``` {r density, fig.cap = "Density of trap sites per thousand km^2^", echo = F, layout = "l-page"}
-knitr::include_graphics(here::here("figures", "static_site_density.png"))
-```
+  + Calculating where possible for site/study effort
 
@@
 
-  + Species trapped
+Across all West African study sites there were 59,343 trapped small mammals (592 were trapped outside of West Africa), 2,830 (4.8%) trapped individuals were identified to order level (Rodentia), 5,329 (9%) were identified to genus level, the remaining 51,184 (86%) were identified to species level.
+Of the 131 distinct identified species trapped (see Supplementary table 1) the majority were from the order Rodentia (102) with Muridae (83) being the largest family of rodents, followed by Gliridae (6), Nesomyidae (4), Sciuridae (3), Ctenodactylidae (2) and Anomaluridae, Dipodidae, Hystricidae and Thryonomidae (all 1).
+The remaining 29 species came from the orders of the Soricomorpha (25), Erinaceomorpha (3) and a single Afrosoricida.
+
+More individuals were trapped in studies conducted to investigate the risk of zoonoses (35,892, 60%) than from ecology studies (23,451 40%).
+Animals trapped as part of ecological research were more often identified to species level (93%) than those trapped in zoonoses research (82%).
+The most commonly trapped genera of rodents were *Mastomys sp.* (22,665, 39.7%), *Rattus sp.* (7,718, 13.5%), *Mus sp.* (7,317, 12.8%), *Praomys sp.* (4,887, 8.6%) and *Arvicanthis sp.* (4,408, 7.72%).
+At the species level *Mastomys natalensis* (10,419, 20.2%) and *Mastomys erythroleucus* (7,184, 13.9%) were the most commonly trapped rodents, *Rattus rattus* (6,229, 12.1%), *Mus musculus* (6,030, 11.8%), *Arvicanthis niloticus* (4,101, 8%) and *Mastomys huberti* (3,106, 6%) made up the next most commonly trapped species.
+Presence and absence maps for these six most commonly trapped species are shown in Figure \@ref(fig:dist).
+
+``` {r dist, fig.cap = "Presence and absence of the 6 most commonly trapped species in included studies, no adjustment made for trapping effort", echo = F, layout = "l-page"}
+knitr::include_graphics(here::here("figures", "top_6.png"))
+```
+
   + Pathogens
+
+## Supplementary material
+
+#### Supplementary figure 1
+
+``` {r supfig1, fig.cap = "Density of trap sites per thousand km^2^", echo = F, layout = "l-page"}
+knitr::include_graphics(here::here("figures", "static_site_density_2.png"))
+```
+
+#### Supplementary table 1
+
+The number and percentage of trapped rodents identified to species level
+
+``` {r sup_table, echo = F}
+DT::datatable(readRDS(here::here("tables", "sup_table1.rds")))
+```

### 2021-03-24 scripts/cleaning.R

@@ -43,7 +43,9 @@ rodent_data %<>%
 
 # Cleaning species names and matching to GBIF
 rodent_data %<>%
-  mutate(genus = ifelse(genus == "nannomys" & species != "-", "mus", genus), # in GBIF nonnomys are classified as mus
+  mutate(genus = ifelse(genus == "nannomys" & species != "-", "mus",
+                        ifelse(genus == "myomys", "praomys", genus)), # in GBIF nonnomys are classified as mus and myomys are a synonym of praomys
+         species = ifelse(genus == "mus" & species == "domesticus", "musculus", species), # coalescing mus musculus and mus musculus domesticus
          classification = paste(genus, ifelse(species == "-", "sp.", species), sep = " "))
 
 genus <- tibble(rodent_data %>%

### 2021-03-24 scripts/descriptive.R

@@ -106,6 +106,42 @@ speciation <- species_data %>%
   left_join(., genus_data %>%
               distinct(genus, .keep_all = T), by = "genus")
 
+species_data %>%
+  left_join(., studies %>%
+              dplyr::select(unique_id, aim),
+            by = "unique_id") %>%
+  filter(iso3c %in% wa_countries) %>%
+  group_by(aim) %>%
+  summarise(n = sum(number))
+
+species_data %>%
+  left_join(., studies %>%
+              dplyr::select(unique_id, aim),
+            by = "unique_id") %>%
+  filter(iso3c %in% wa_countries & genus == "rodentia") %>%
+  group_by(aim, unique_id) %>%
+  summarise(n = sum(number))
+
+species_data %>%
+  left_join(., studies %>%
+              dplyr::select(unique_id, aim),
+            by = "unique_id") %>%
+  filter(iso3c %in% wa_countries & species %in% c("-", "sp.1", "sp.2") & genus != "rodentia") %>%
+  group_by(aim) %>%
+  summarise(n = sum(number))
+
+species_data %>%
+  left_join(., studies %>%
+                         dplyr::select(unique_id, aim),
+                       by = "unique_id") %>%
+  filter(iso3c %in% wa_countries & !is.na(species_gbif)) %>%
+  group_by(aim) %>%
+  summarise(n = sum(number))
+
+sup_table <- species_data %>%
+  group_by(classification) %>%
+  summarise(number = sum(number))
+
 count_genus <- species_data %>%
   group_by(genus_gbif, genus) %>%
   filter(genus != "rodentia") %>%
@@ -114,5 +150,14 @@ count_genus <- species_data %>%
   arrange(-percent)
 
 count_species <- species_data %>%
-  group_by(species_gbif) %>%
-  summarise(number = sum(number))
+  group_by(species_gbif, classification) %>%
+  drop_na(species_gbif) %>%
+  summarise(number = sum(number)) %>%
+  mutate(percent = round(number/sum(.$number)*100, 2)) %>%
+  arrange(-percent) %>%
+  rename(`GBIF ID` = "species_gbif",
+         "Classification" = "classification",
+         "Number of individuals" = "number",
+         "Percent (%)" = "percent") %>%
+  mutate(Classification = snakecase::to_sentence_case(Classification))
+write_rds(count_species, here("tables", "sup_table1.rds"))

### 2021-03-24 scripts/download_spatial.R

@@ -53,3 +53,23 @@ SLE_1 <- getData(name = "GADM", country ="SLE", level = 1, path = here("data_dow
 level_one <- ls(pattern = "_1")
 countries_level_1 <- do.call("list",mget(level_one))
 write_rds(countries_level_1, here("data_download", "admin_spatial", "level_1_admin.rds"))
+
+# Level 2 -------------------------------------------------
+
+BEN_2 <- getData(name = "GADM", country ="BEN", level = 2, path = here("data_download", "admin_spatial"))
+BFA_2 <- getData(name = "GADM", country ="BFA", level = 2, path = here("data_download", "admin_spatial"))
+CIV_2 <- getData(name = "GADM", country ="CIV", level = 2, path = here("data_download", "admin_spatial"))
+GHA_2 <- getData(name = "GADM", country ="GHA", level = 2, path = here("data_download", "admin_spatial"))
+GIN_2 <- getData(name = "GADM", country ="GIN", level = 2, path = here("data_download", "admin_spatial"))
+GNB_2 <- getData(name = "GADM", country ="GNB", level = 2, path = here("data_download", "admin_spatial"))
+LBR_2 <- getData(name = "GADM", country ="LBR", level = 2, path = here("data_download", "admin_spatial"))
+MLI_2 <- getData(name = "GADM", country ="MLI", level = 2, path = here("data_download", "admin_spatial"))
+MRT_2 <- getData(name = "GADM", country ="MRT", level = 2, path = here("data_download", "admin_spatial"))
+NER_2 <- getData(name = "GADM", country ="NER", level = 2, path = here("data_download", "admin_spatial"))
+NGA_2 <- getData(name = "GADM", country ="NGA", level = 2, path = here("data_download", "admin_spatial"))
+SEN_2 <- getData(name = "GADM", country ="SEN", level = 2, path = here("data_download", "admin_spatial"))
+SLE_2 <- getData(name = "GADM", country ="SLE", level = 2, path = here("data_download", "admin_spatial"))
+
+level_two <- ls(pattern = "_2")
+countries_level_2 <- do.call("list",mget(level_two))
+write_rds(countries_level_2, here("data_download", "admin_spatial", "level_2_admin.rds"))

### 2021-03-24 scripts/mapping_presence.R

@@ -1,0 +1,45 @@
+source(here::here("scripts", "libraries.r"))
+
+wa_mainland <- c("BEN", "BFA", "CIV", "ESH", "GHA",
+                  "GIN", "GMB", "GNB", "LBR", "MLI", "MRT",
+                  "NER", "NGA", "SEN", "SLE", "TGO")
+
+level_0 <- read_rds(here("data_download", "admin_spatial", "level_0_admin.rds"))
+list2env(level_0, envir = .GlobalEnv)
+level_0 <- do.call(rbind.SpatialPolygonsDataFrame, level_0)  %>%
+  st_as_sf()
+
+level_1 <- read_rds(here("data_download", "admin_spatial", "level_1_admin.rds"))
+list2env(level_1, envir = .GlobalEnv)
+level_1_all <- do.call(rbind.SpatialPolygonsDataFrame, level_1) %>%
+  st_as_sf()
+
+level_2 <- read_rds(here("data_download", "admin_spatial", "level_2_admin.rds"))
+list2env(level_2, envir = .GlobalEnv)
+level_2_all <- do.call(rbind.SpatialPolygonsDataFrame, level_2) %>%
+  st_as_sf()
+
+rodent_spatial <- read_rds(here("data_clean", "rodent_spatial.rds")) %>%
+  filter(iso3c %in% wa_countries)
+studies <- read_rds(here("data_clean", "studies.rds"))
+
+top_6_species <- c(2438911, 2438912, 2439270, 7429082, 9733997, 4264823)
+
+top_6_spatial <- rodent_spatial %>%
+  filter(species_gbif %in% top_6_species) %>%
+  mutate(pres_abs = ifelse(number == 0, "Absent", "Present")) %>%
+  left_join(., studies %>%
+              dplyr::select(metric, unique_id, aim),
+            by = "unique_id") %>%
+  mutate(abundance = ifelse(metric == "presence", NA, number),
+         classification = snakecase::to_sentence_case(classification))
+
+afr_bbox <- st_bbox(level_0 %>%
+          filter(GID_0 %in% wa_mainland))
+
+top_6_plot <- tm_shape(level_0 %>%
+                         filter(GID_0 %in% wa_mainland), bbox = afr_bbox) +  tm_polygons(alpha = 0, lwd = 1) +
+  tm_shape(top_6_spatial) + tm_dots(col = "pres_abs", palette = "Dark2", title = "", size = .1, shape = 20) +
[…]

### 2021-03-25 reports/manuscript.Rmd
improving species and genus cleaning

@@
 
-Where explicity stated the study aims were extracted.
+Where explicitly stated the study aims were extracted.
 For those with no recorded aims the specific aims were inferred from the background or conclusions of manuscript or report.
-The aims of the studies were broadly categorised as pertaining to rodent ecology research or studies on risk of zoonoses.
-Within these subdivisions the aim were further subdivided into the following categories:
-
-  + Rodent ecology
-    + Cat 1
-    + Cat 2
-  + Zoonoses risk
-    + Cat 1
-    + Cat 2
-
-### Trapping methodology
-
-The brand or description of the rodent trap was obtained from each study that reported this. For those using multiple devices all types were recorded.
+The aims of the studies were broadly categorised as pertaining to rodent ecology research or studies on risk of zoonoses and further categorised into detailed aims.
 
@@
 
+### Trapping methodology
+
+The brand or description of the rodent trap was obtained from each study. For those using multiple devices all types were recorded.
+
 ### Genus and species definitions
 
-Designations of species change over time as new evidence emerges. As included studies span 45 years disambiguation of historical species names was performed using Taxize...
+Classification of species is variable over time as morphological and genetic information emerges.
+To handle species classification synonyms and reclassification the reported names of trapped species was compared to the Global Biodiversity Information Facility [].
+For species with updated taxonomies the synonym was converted to the accepted species name for all subsequent analyses.
+Where a subspecies is reported by study authors the species complex it belongs to will be used for subsequent analysis.
 
@@
 
-  + Aims
-  + Published/field
+### Aims of included studies
+
+The detailed aims of the included studies are broken down by the identifier of ecology or zoonosis research, the number in brackets refers to the number of studies including this as an aim:
+
+  + Ecology (n = 55)
+    + Biodiversity
+      + To investigate biodiversity of rodent populations (33)
+      + To conduct a small mammal survey (19)
+      + To characterise biodiversity of rodent parasites (4)
+      + To understand the effect of nature reserve designation on rodent biodiversity (1)
+    + Rodent population dynamics
+      + To characterise rodent population dynamics (7)
+      + To describe survival and recruitment within a rodent population (1)
+      + To investigate survival and recruitment in an island rodent population (1)
+    + Rodent species assemblages
+      + To understand the impact of land use change on rodent assemblages (5)
+      + To characterise rodent assemblages along an invasion front (4)
+      + To characterise rodent assemblages in different habitats (4)
+      + To characterise commensal rodent assemblages (1)
+      + To investigate the effect of habitat and aridity on rodent assemblages (1)
+      + To understand the effect of habitat on rodent parasite assemblages (1)
+      + To investigate the the impact of wildfires on population assemblages (1)
+      + To understand rodent assemblage structure in urban settings (1)
+      + To characterise seasonal changes in rodent assemblages (1)
+    + Rodent genetics and species classification
+      + To perform genetic based revision of species classification (2)
+      + To investigate the phylogenetics of invasive rodent species (1)
+      + To understand chromosomal variation in a specific rodent species (1)
+      + To characterise morphological variation within a rodent species (1)
+    + Rodent adaptation and behaviour
+      + To understand the effect of a water poor diet on specific rodent species (1)
+      + To investigate the feeding ecology of shrews (1)
+      + To understand the social organisation of commensal rodents (1)
+    + Other
+      + To infer historical rodent habitat range expansion (1)
+      + To characterise the impact of rodent pest control (1)
+  + Zoonoses risk (n = 46)
+    + Zoonotic disease prevalence
+      + To determine the prevalence of zoonoses in animal population (27)
+      + To investigate the prevalence of zoonotic bacteria in rodent ectoparasites (1)
+      + To identify clusters of infected rodents (1)
+      + To investigate regional distribution of zoonosis risk (1)
+    + Zoonotic disease spillover risk
+      + To investigate the potential reservoir of zoonotic pathogen (12)
+      + To assess the risk of zoonotic spillover (11)
+      + To investigate the source of a zoonotic outbreak (2)
+      + To characterise host species biodiversity (2)
+      + To investigate the effect of rodent control on zoonoses risk (1)
+      + To characterise the effect of invasive species on zoonoses risk (1)
+    + Effect of environmental factors on zoonoses
+      + To investigate the effect of environment and climate on arthropod vectors (2)
+    + Other
+      + To investigate the phylogenetics of a zoonotic pathogen (2)
+      + To investigate the evolutionary relationships of pathogen strains (1)
+      + To investigate parasite biodiversity (1)
 
@@
 Thirty studies did not report trapping effort.
-Thirty-six studies reported a measure of trapping effort at the same level of detail at which they reported rodent captures.
-Thirty-five studies reported incomplete data on trapping effort (i.e.
+Thirty-four studies reported a measure of trapping effort at the same level of detail at which they reported rodent captures (the unit of analysis).
+Thirty-seven studies reported incomplete data on trapping effort (i.e.
 number of study nights or total number of trap-nights).
-Trapping effort was more completely reported from ecology studies (53% complete information and 20% no information) compared to zoonoses risk studies (15% complete information and 41% no information).
+Trapping effort was more completely reported from ecology studies (51% complete information and 20% no information) compared to zoonoses risk studies (13% complete information and 41% no information).
 
-  + Calculating where possible for site/study effort
+Among studies that reported complete trap night information the median number of trap nights per study was 4,764 (Interquartile range (IQR) = 1,227-10,477) with a minimum of 240 trap nights and maximum of 45,274 trap nights.
+At the level of the studies unit of analysis the median number of trap nights was 360 (IQR = 246-793).
+In studies with incomplete data on trap nights at the unit of analysis there was a median of 7,893 trap nights (IQR = 1,401-10,331) conducted, for 21 studies reporting only on the number of days or nights trapping was performed activities typically lasted between 1-4 nights.
 
-  + Habitat classification
+### Habitat classification
+
+No included studies included a reference to a specific habitat classification scheme. **Describe classifications used**
 
@@
 
-  + Pathogens
+### Pathogens
 

### 2021-03-25 scripts/cleaning.R

@@ -4,136 +4,7 @@ source(here("scripts", "update_data.r"))
 # Studies -----------------------------------------------------------------
 
 studies <- read_rds(here("data_raw", "studies.rds")) %>%
-  mutate(unique_id = as_factor(unique_id))
+  mutate(unique_id = as_factor(unique_id)) %>%
+  separate(col = aim_detail, into = c("aim_detail_1", "aim_detail_2"), sep = ", ")
 
 write_rds(studies, here("data_clean", "studies.rds"))
-
-# Rodent data -------------------------------------------------------------
-
-rodent_data <- read_rds(here("data_raw", "rodent_data.rds")) %>%
-  mutate(country = as_factor(country),
-         record_id = 1:nrow(.))
-
-retain_columns <- c("unique_id", "year_trapping", "month_trapping", "country", "iso3c", "region", "town_village", "habitat", "intensity_use",
-                    "genus", "species", "classification", "genus_gbif", "species_gbif",  "number", "trap_nights", "trap_night_unit", "record_id")
-
-# Cleaning coordinates
-rodent_data %<>%
-  separate(col = longitude_DMS_W, into = c("long_degrees", "long_minutes", "long_seconds"), "_", remove = F) %>%
-  separate(col = latitude_DMS_N, into = c("lat_degrees", "lat_minutes", "lat_seconds"), "_") %>%
-  mutate(across(all_of(c("long_degrees", "long_minutes", "long_seconds","lat_degrees", "lat_minutes", "lat_seconds")), as.double),
-         long_hemi = ifelse(long_degrees<0, "E", #assign -ve numbers to E
-                            ifelse(substring(longitude_DMS_W, 1, 1) == "-", "E", "W")), #as 0 cannot be -ve we can check the sign on the text entry
-         lat_hemi = ifelse(lat_degrees<0, "S", "N"),
-         gps_dms = ifelse(is.na(lat_degrees|long_degrees), F, T),
-         long_dms = ifelse(gps_dms == T,
-                           paste(long_hemi, long_degrees, " ",
-                                 ifelse(is.na(long_minutes), 0, long_minutes), '.',
-                                 ifelse(is.na(long_seconds), 0, long_seconds),
-                                 sep = ""), NA),
-         lat_dms = ifelse(gps_dms == T,
[…]

### 2021-03-25 scripts/cleaning_species.R

@@ -1,0 +1,141 @@
+source(here::here("scripts", "libraries.r"))
+
+rodent_data <- read_rds(here("data_raw", "rodent_data.rds")) %>%
+  mutate(country = as_factor(country),
+         record_id = 1:nrow(.))
+
+retain_columns <- c("unique_id", "year_trapping", "month_trapping", "country", "iso3c", "region", "town_village", "habitat", "intensity_use",
+                    "genus", "species", "classification", "gbif_id", "genus_gbif", "species_gbif",  "number", "trap_nights", "trap_night_unit", "record_id")
+
+genus_synonym <- tibble(syn = c("myomys", "nannomys"), acc = c("praomys", "mus"))
+species_synonym <- tibble(syn = c("mus domesticus", "aethomys hypoxanthus", "crocidura megalura", "dasymys bentleyae", "erinaceus albiventris", "gerbilliscus gambianus", "tatera guineae", "gerbilliscus guineae", "gerbilliscus kempi", "gerbillus campestris", "gerbillus gambianus", "grammomys rutilans", "graphiurus hueti", "graphiurus parvus", "hemiechinus aethiopicus", "lophuromys flavipunctatus", "massouteria mzabi", "mastomys hildebrandtii", "steatomys caurianus", "tatera gambiana", "tatera guinea", "tatera guineae", "tatera kempi", "tatera robusta", "taterillus gracillis", "thamnomys rutilans"),
+                          acc = c("mus musculus", "oenomys hypoxanthus", "suncus megalura", "dasymys incomtus", "atelerix albiventris", "gerbilliscus gambiana", "gerbilliscus guinea", "gerbilliscus guinea", "gerbilliscus kempii", "dipodillus campestris", "gerbilliscus gambiana", "grammomys poensis", "graphiurus nagtglasii", "graphiurus kelleni", "paraechinus aethiopicus", "lophuromys flavopunctatus", "massoutiera mzabi", "mastomys natalensis", "steatomys caurinus", "gerbilliscus kempii", "gerbilliscus guineae", "gerbilliscus guineae", "gerbilliscus kempii", "gerbilliscus robusta", "taterillus gracilis", "grammomys poensis"))
+
+# Cleaning species names and matching to GBIF
+rodent_data$genus <- plyr::mapvalues(rodent_data$genus, from = genus_synonym$syn, to = genus_synonym$acc)
[…]

### 2021-03-25 scripts/descriptive.R

@@ -16,6 +16,9 @@ table(studies$aim)
 ecology_studies <- studies %>% filter(aim == "Ecology")
 zoonoses_studies <- studies %>% filter(aim == "Zoonoses risk")
 
+aim_detail_eco <- studies %>% filter(aim == "Ecology") %>% group_by(aim_detail_2) %>% summarise(n = n())
+aim_detail_zoo <- studies %>% filter(aim == "Zoonoses risk") %>% group_by(aim_detail_2) %>% summarise(n = n())
+
 # Trap type and setup -----------------------------------------------------
 trap_type <- studies %>%
   separate(col = trap_types, into = c("trap_1", "trap_2", "trap_3", "trap_4"), sep = ", ", remove = T) %>%
@@ -31,8 +34,36 @@ trap_technique <- studies %>%
 
 table(trap_technique$trap_method)
 
+# Trapping effort ---------------------------------------------------------
+
 table(studies$trapping_effort)
 
+t_effort <- studies %>% filter(trapping_effort == "Yes") %>% distinct(unique_id)
+
+s_effort <- as.data.frame(rodent_data) %>% filter(unique_id %in% t_effort$unique_id) %>%
+  group_by(unique_id, year_trapping, month_trapping, region, town_village, habitat) %>%
+  summarise(trap_nights = unique(trap_nights)) %>%
+  group_by(unique_id) %>%
+  summarise(trap_nights = sum(trap_nights))
+
+summary(s_effort$trap_nights)
+
+t_effort <- as.data.frame(rodent_data) %>% filter(unique_id %in% t_effort$unique_id) %>%
+  group_by(unique_id, year_trapping, month_trapping, region, town_village, habitat) %>%
+  summarise(trap_nights = unique(trap_nights))
+
+summary(t_effort$trap_nights)
+
+inc_effort <- studies %>% filter(trapping_effort == "Incomplete")
+
+inc_effort <- as.data.frame(rodent_data) %>% filter(unique_id %in% inc_effort$unique_id & !is.na(trap_nights)) %>%
+  group_by(unique_id, year_trapping, month_trapping, region, town_village, habitat) %>%
+  summarise(trap_nights = unique(trap_nights)) %>%
+  group_by(unique_id) %>%
+  summarise(trap_nights = unique(trap_nights)) %>%
+  summarise(trap_nights = sum(trap_nights))
+
+summary(inc_effort$trap_nights)
[…]

### 2021-03-26 reports/manuscript.Rmd
finishing cleaning pathogen

--- 2021-03-24 reports/manuscript.Rmd
@@
 
+### Trapping methodology
+
+The brand or description of the rodent trap was obtained from each study. For those using multiple devices all types were recorded. Type of trapping method. Effort of trapping
+
+### Habitat classification
+
 ### Genus and species definitions
 
-Designations of species change over time as new evidence emerges. As included studies span 45 years disambiguation of historical species names was performed using Taxize...
+Classification of species is variable over time as morphological and genetic information emerges.
+To handle species classification synonyms and reclassification the reported names of trapped species was compared to the Global Biodiversity Information Facility [].
+For species with updated taxonomies the synonym was converted to the accepted species name for all subsequent analyses.
+Where a subspecies is reported by study authors the species complex it belongs to will be used for subsequent analysis.
+
+### Pathogen testing
 
@@
 
-  + Aims
-  + Published/field
+### Aims of included studies
+
+The detailed aims of the included studies are shown following division into the themes of ecological or zoonosis research.
+Biodiversity of rodents was a shared aim between the two themes with 37 ecological studies and 2 zoonoses studies investigating this.
+The remaining aims were more disparate with ecological studies investigating rodent population dynamics, rodent species assemblages, the genetics of rodents and their behaviours.
+Most zoonotic disease research focussed on prevalence of pathogens in rodent populations, the risk of spillover to human populations, the identity of pathogen reservoir species with fewer studies investigating the phylogenetics of pathogen strains (see Figure \@ref(fig:aims).
+
+``` {r aims, fig.cap = "The aims of included studies following division into ecological and zoonotic infectious disease themes.
+Multiple aims were identified from each study, the number on the arrow indicates the number of distinct studies with that aim ", echo = F, layout = "l-page"}
+knitr::include_graphics(here::here("figures", "study_aims.png"))
+```
 
@@
 
-### Trap setup
+### Trap setup within the study area
 
-Thirty-three studies did not describe the setup of traps.
+Thirty-three studies did not describe the structure of the trap setup within the defined study area.
 Twenty five studies placed transect lines of traps through study sites with thirteen studies placing traps along a line but not clearly describing a transect across a defined site.
-Sixteen studies purposefully sample houses, with four further studies purposefully placing traps in targeted locationes.
-Nine studies placed traps in a grid across single or multiple habitat types.
-Three studies described a stratified or randomised approach to placing traps.
-Trap setup was typically reported for ecology studies (80%) with transect lines the most commonly used approach.
+Sixteen studies purposefully sampled houses, with four further studies purposefully placing traps in targeted locations (i.e.
+near rivers, in fields, by houses).
+Nine studies placed traps in a structured grid across a single or multiple specified habitat types.
+Three studies described a stratified or randomised approach to placing traps within a defined area.
+Trap setup was commonly reported in ecology studies (80%) with transect lines the most commonly used approach.
 Trap setup was reported in 53% of zoonoses risk studies with purposeful placement of traps in houses the most common approach (20%).
@@
 
-Trapping effort was variably reported in included studies.
-Thirty studies did not report trapping effort.
-Thirty-six studies reported a measure of trapping effort at the same level of detail at which they reported rodent captures.
-Thirty-five studies reported incomplete data on trapping effort (i.e.
-number of study nights or total number of trap-nights).
-Trapping effort was more completely reported from ecology studies (53% complete information and 20% no information) compared to zoonoses risk studies (15% complete information and 41% no information).
+Trapping effort was variably reported across included studies.
+Thirty studies did not report any measures of trapping effort.
+Thirty-four studies reported a measure of trapping effort at the same level of detail at which they reported rodent captures  (the unit of analysis) and were classified as complete reports.
+Thirty-seven studies had incomplete reports on trapping effort due to not presenting the number of trap nights at the appropriate unit of analysis (i.e.
+number of study nights only or total number of trap-nights).
+Trapping effort was more completely reported from ecology studies (51% complete information and 20% no information) compared to zoonoses risk studies (13% complete information and 41% no information).
 
-  + Calculating where possible for site/study effort
+Among studies that reported complete trap-night information the median number of trap-nights per study was 4,764 (Interquartile range (IQR) = 1,227-10,477) with a minimum of 240 trap-nights and maximum of 45,274 trap-nights.
+For studies reporting the number of trap-nights at the unit of analysis (typically study site or habitat type) the median number of trap-nights was 360 (IQR = 246-793).
+In studies with incomplete reports on trap-nights at the unit of analysis there was a median of 7,893 trap-nights (IQR = 1,401-10,331) conducted, for 21 studies reporting only on the number of days or nights trapping was performed on activities typically lasted between 1-4 nights.
 
-  + Habitat classification
+### Habitat classification
+
+No studies classified habitats with reference to a specific habitat classification scheme.
+Seven studies did not describe the local habitat where traps were placed (3 ecology, 4 zoonoses).
+The remaining studies used greater than 160 descriptions of the habitat type in which trapping was conducted, further, the resolution of the description ranged from the habitat of the location of the trap through to the study site.
+The 160 reported habitat types were combined into 32 categories (see Supplementary table 2 for the habitat dictionary) for further analysis.
+At least one habitat was recorded for 16,168 trap sites (95%), with two or more habitats for a single trap site recorded for 4,195 (26%) sites.
+As single sites could span multiple habitat types the number of habitat type trap site combinations was 20,369.
+The most commonly trapped sites were in or around buildings (31%), in areas described as the rodents "natural habitat" (26%), in agricultural areas (13%) (e.g.
+rice fields, palm plantations), forests (6%) and within or around villages (5%).
+
+Habitat type was more comprehensively reported from ecology than from zoonoses studies.
+Thirty-one habitat categories were reported from ecology studies with the most commonly investigated habitats being agriculture (23%), in or around buildings (18%) and within forests (15%).
+Among studies on zoonoses 15 habitat categories were reported, the most commonly reported classification was described as the rodents "natural habitat" (39%), within or around buildings (38%) and in agricultural settings (7%).
 
@@
 
-Across all West African study sites there were 59,343 trapped small mammals (592 were trapped outside of West Africa), 2,830 (4.8%) trapped individuals were identified to order level (Rodentia), 5,329 (9%) were identified to genus level, the remaining 51,184 (86%) were identified to species level.
-Of the 131 distinct identified species trapped (see Supplementary table 1) the majority were from the order Rodentia (102) with Muridae (83) being the largest family of rodents, followed by Gliridae (6), Nesomyidae (4), Sciuridae (3), Ctenodactylidae (2) and Anomaluridae, Dipodidae, Hystricidae and Thryonomidae (all 1).
-The remaining 29 species came from the orders of the Soricomorpha (25), Erinaceomorpha (3) and a single Afrosoricida.
+Across all West African study sites there were 59,343 trapped small mammals (592 were trapped outside of West African countries), 2,830 (4.8%) trapped individuals were identified to order level (Rodentia), 5,329 (9%) were identified to genus level, the remaining 51,184 (86%) were identified to species level.
+Of the 124 distinct identified species trapped (see Supplementary table 1) the majority were from the order Rodentia (96) with Muridae (79) being the largest family of rodents, followed by Gliridae and Nesomyidae (both 4), Sciuridae (3), Ctenodactylidae (2) and Anomaluridae, Dipodidae, Hystricidae and Thryonomidae (all 1).
+The remaining 29 species came from the orders of the Soricomorpha (25), Erinaceomorpha (2) and Afrosoricida (1).
 
@@
 Animals trapped as part of ecological research were more often identified to species level (93%) than those trapped in zoonoses research (82%).
-The most commonly trapped genera of rodents were *Mastomys sp.* (22,665, 39.7%), *Rattus sp.* (7,718, 13.5%), *Mus sp.* (7,317, 12.8%), *Praomys sp.* (4,887, 8.6%) and *Arvicanthis sp.* (4,408, 7.72%).
-At the species level *Mastomys natalensis* (10,419, 20.2%) and *Mastomys erythroleucus* (7,184, 13.9%) were the most commonly trapped rodents, *Rattus rattus* (6,229, 12.1%), *Mus musculus* (6,030, 11.8%), *Arvicanthis niloticus* (4,101, 8%) and *Mastomys huberti* (3,106, 6%) made up the next most commonly trapped species.
+The most commonly trapped genera of rodents were *Mastomys sp.* (22,665, 39.7%), *Rattus sp.* (7,718, 13.5%), *Mus sp.* (7,324, 12.8%), *Praomys sp.* (4,977, 8.7%) and *Arvicanthis sp.* (4,408, 7.72%).
+At the species level *Mastomys natalensis* (10,351, 20.2%) and *Mastomys erythroleucus* (7,052, 13.8%) were the most commonly trapped rodents, *Rattus rattus* (6,228, 12.2%), *Mus musculus* (6,030, 11.8%), *Arvicanthis niloticus* (4,086, 8%) and *Mastomys huberti* (3,106, 6.1%) made up the next most commonly trapped species.
 Presence and absence maps for these six most commonly trapped species are shown in Figure \@ref(fig:dist).

### 2021-03-26 scripts/clean_habitat.R

@@ -1,0 +1,54 @@
+source(here::here("scripts", "libraries.r"))
+
+rodent_data <- read_rds(here("data_clean", "rodent_df.rds")) %>%
+  tibble()
+habitat_split <- c("habitat_1", "habitat_2", "habitat_3", "habitat_4", "habitat_5", "habitat_6", "habitat_7")
+
+rodent_data %<>%
+  separate(habitat, into = c(habitat_split), sep = ", ", remove = F)
+
+habitat_types <- rodent_data %>%
+  dplyr::select(unique_id, country, region, town_village, all_of(habitat_split), trap_night_unit, trap_nights) %>%
+  pivot_longer(cols = all_of(habitat_split), values_to = "habitat_type") %>%
+  drop_na(habitat_type)
+
+if(!file.exists(here("data_clean", "habitat_dictionary.rds"))){
+sort(unique(habitat_types$habitat_type))
+cleaned_habitat <- as.list(c("agriculture", "agriculture", "airport", "alluvial_plain", "buildings_periphery", "nature_reserve", "botanical_garden", "buildings_periphery",
+                             "burned", "vegetation", "vegetation", "clay_soil", "buildings_periphery", "agriculture", "waterway", "waterway", "agriculture", "agriculture",
+                             "agriculture", "agriculture", "agriculture", "degraded_forest", "agriculture", "agriculture", "degraded_forest", "forest", "waterway", "waterway",
+                             "waterway", "forest", "sand_dunes", "vegetation", "fallow", "fallow", "fallow", "agriculture", "agriculture", "agriculture", "agriculture",
+                             "alluvial_plain", "forest", "forest", "degraded_forest", "degraded_forest", "gardens", "vegetation", "buildings_periphery", "vegetation", "forest",
+                             "vegetation", "vegetation", "buildings", "buildings", "agriculture", "buildings", "buildings_periphery", "buildings_periphery", "forest", "waterway",
+                             "island", "lacustrian", "waterway", "vegetation", "building_periphery", "building_periphery", "building_periphery", "marsh", "agriculture",
+                             "lateritic soil", "agriculture", "multiple", "nature_reserve", "forest_clearing", "natural_habitat", "natural_habitat", "forest", "agriculture",
+                             "forest", "natural_habitat", "agriculture", "village", "oasis", "agriculture", "agriculture", "natural_habitat", "agriculture", "agriculture",
[…]

### 2021-03-26 scripts/descriptive.R

@@ -1,7 +1,18 @@
 source(here::here("scripts", "libraries.r"))
 studies <- read_rds(here("data_clean", "studies.rds"))
-rodent_data <- read_rds(here("data_clean", "rodent_df.rds"))
+rodent_data <- read_rds(here("data_clean", "rodent_df.rds")) %>%
+  tibble()  %>%
+  dplyr::select(-all_of(c("habitat", "intensity_use", "geometry")))
 
+habitat_split <- c("habitat_1", "habitat_2", "habitat_3", "habitat_4", "habitat_5", "habitat_6", "habitat_7")
+habitat_data <- read_rds(here("data_clean", "habitat_types.rds")) %>%
+  tibble() %>%
+  dplyr::select(record_id, all_of(habitat_split), intensity_use)
+
+rodent_data %<>%
+  left_join(.,
+            habitat_data ,
+            by = "record_id")
 # Publication year --------------------------------------------------------
 ggplot(studies) +
   geom_bar(aes(x = year_publication)) +
@@ -16,54 +27,9 @@ table(studies$aim)
 ecology_studies <- studies %>% filter(aim == "Ecology")
 zoonoses_studies <- studies %>% filter(aim == "Zoonoses risk")
 
-aim_detail_eco <- studies %>% filter(aim == "Ecology") %>% group_by(aim_detail_2) %>% summarise(n = n())
-aim_detail_zoo <- studies %>% filter(aim == "Zoonoses risk") %>% group_by(aim_detail_2) %>% summarise(n = n())
+aim_detail_eco <- studies %>% filter(aim == "Ecology") %>% dplyr::select(unique_id, aim, aim_detail_1, aim_detail_2)
+aim_detail_zoo <- studies %>% filter(aim == "Zoonoses risk") %>%  dplyr::select(unique_id, aim, aim_detail_1, aim_detail_2)
 
-# Trap type and setup -----------------------------------------------------
-trap_type <- studies %>%
-  separate(col = trap_types, into = c("trap_1", "trap_2", "trap_3", "trap_4"), sep = ", ", remove = T) %>%
-  pivot_longer(cols = c("trap_1", "trap_2", "trap_3", "trap_4"), values_to = "trap_type") %>%
-  drop_na(trap_type)
-
-trap_type %>% filter(!trap_type %in% c("hand", "not_stated")) %>% group_by(unique_id) %>% summarise(n = n()) %>% count(n)
-
-trap_technique <- studies %>%
[…]

### 2021-03-28 reports/manuscript.Rmd
saving manuscript
pathogen_maps

@@
 Locations were extracted in the format reported and converted to decimal degrees.
-Where no GPS location is reported coordinates will be matched to study locations using the National Geospatial Intelligence Agency NGA GEOnet Names Server [] based on study site name/region and through maps presented in the manuscript.
+Where no GPS location is reported coordinates will be matched to study locations using the National Geospatial Intelligence Agency NGA GEOnet Names Server [ref.] based on study site name/region and through maps presented in the manuscript.
 
@@
 Classification of species is variable over time as morphological and genetic information emerges.
-To handle species classification synonyms and reclassification the reported names of trapped species was compared to the Global Biodiversity Information Facility [].
+To handle species classification synonyms and reclassification the reported names of trapped species was compared to the Global Biodiversity Information Facility [ref.].
 For species with updated taxonomies the synonym was converted to the accepted species name for all subsequent analyses.
@@
 A total of 2,200 records were identified, with 101 studies included in narrative synthesis (see Figure \@ref(fig:prisma)).
+The characteristics of the included studies are shown in Table 1.
+The earliest studies identified were from 1974 with the majority (62%) published since 2010.
+Most of the included studies were classified as rodent ecology research (54%) with the remainder of studies on zoonotic diseases (46%).
 
@@
 ```
-
-The earliest studies identified were from 1974 with the majority (62%) published since 2010.
-Most of the included studies were classified as rodent ecology research (54%) with the remainder studies on zoonotic diseases (46%).
 
@@
 For the remaining studies three used more than four types of rodent traps, seven used three types, 31 used two types the remaining 45 studies used a single trap type.
-The most commonly used trap was the Sherman trap (67%) [] which is available in several sizes, locally made wire mesh traps were used in 29 studies (34%), less commonly used trap types included the Museum special, Tomahawk, Victor, Snaptraps and Firobind traps.
+The most commonly used trap was the Sherman trap (67%) [ref.] which is available in several sizes, locally made wire mesh traps were used in 29 studies (34%), less commonly used trap types included the Museum special, Tomahawk, Victor, Snaptraps and Firobind traps.
 
@@
 At least one habitat was recorded for 16,168 trap sites (95%), with two or more habitats for a single trap site recorded for 4,195 (26%) sites.
-As single sites could span multiple habitat types the number of habitat type trap site combinations was 20,369.
+Single trap sites could span multiple habitat types resulting in 20,369 habitat or trap sites.
 The most commonly trapped sites were in or around buildings (31%), in areas described as the rodents "natural habitat" (26%), in agricultural areas (13%) (e.g.
@@
 
-Habitat type was more comprehensively reported from ecology than from zoonoses studies.
-Thirty-one habitat categories were reported from ecology studies with the most commonly investigated habitats being agriculture (23%), in or around buildings (18%) and within forests (15%).
+Habitat type was more comprehensively reported from ecology studies, 31 habitat categories were reported from ecology studies with the most commonly investigated habitats being agriculture (23%), in or around buildings (18%) and within forests (15%).
 Among studies on zoonoses 15 habitat categories were reported, the most commonly reported classification was described as the rodents "natural habitat" (39%), within or around buildings (38%) and in agricultural settings (7%).
@@
 
+Trap success rates
+
+### Rodent biodiversity
+
+Diversity indices, accumulation measures
+
 ``` {r dist, fig.cap = "Presence and absence of the 6 most commonly trapped species in included studies, no adjustment made for trapping effort", echo = F, layout = "l-page"}
@@
 ### Pathogens
+
+Fifty-one studies presented data on microorganisms that infect or are carried by small mammal species in West Africa.
+Data from a study previously excluded [@fichet-calvet_fluctuation_2007] due to duplicated data was incorporated as it provided the complementary pathogen data to the rodent trapping data.
+Six of the studies were solely investigated pathogens of rodents.
+The remaining 45 studies investigated organisms that were potentially zoonotic pathogens, pathogens of rodents or microbial organisms that were of uncertain significance to rodent or human health.
+
+Twenty-six studies used Polymerase Chain Reaction (PCR) to detect the presence of 22 different species or families of microorganisms.
+Ten studies used antibody or antigen based molecular tests to detect the presence of 9 different species or families of microorganisms.
+Seven studies conducted histological or direct visualisation assays of samples for 10 parasitic or bacterial species.
+Two studies performed direct viral culture of *Lassa mammarenavirus* to detect the presence of the virus in rodent specimens.
+
+The rodent pathogens investigated were *Hydatigera species* (previously *Taenia species*) and *Trichuris species* both of which are gastro-intestinal helminths and *Leishmania major* which is an intracellular pathogen causing disease in both rodents and humans.
+
+Forty-two of the 45 studies reported prevalence of zoonotic pathogens at a level of detail beyond a study wide summary.
+These studies typically tested for a single pathogen (35), with two studies testing for two and three pathogens and three studies testing for five or more.
+The most frequently tested for pathogens were *Lassa mammarenavirus* (11) or members of the *Arenaviridae* family (7), the spirochete bacteria *Borrelia* was investigated in six studies, *Bartonella* and *Toxoplasma gondii* were investigated in three each, the remaining 28 pathogens were reported in two or fewer studies.
+
+Ten of the studies reporting data on pathogen prevalence had incomplete data on the geolocation from which some of the animals tested were from.
+Six of these studies had no data on geolocation of the trapped rodents.
+
+``` {r paths, fig.cap = "Presence and absence of the 4 most commonly tested for species in included studies, no adjustment made for trapping effort", echo = F, layout = "l-page"}
+knitr::include_graphics(here::here("figures", "top_4.png"))
+```
 

### 2021-03-28 scripts/clean_pathogen.R

@@ -37,26 +37,29 @@ pathogen_data <- pathogen_data %>%
 pathogen_tested <- c("path_1", "path_2", "path_3", "path_4", "path_5", "path_6")
 
 if(!file.exists(here("data_clean", "pathogen_dictionary.rds"))){
-pathogen_name <- as.list(c("anaplasma", "arenaviridae_species", "arenaviridae_species", "arenaviridae_species", "arenaviridae_species", "arenaviridae_species", "arenaviridae_species",
-                           "bartonella_species", "bartonella_species", "borrelia_species", "borrelia_crocidurae", "borrelia_crocidurae", "borrelia_species",
-                           "e_coli_esbl", "ehrlichia", "flavivirus_species", "hantavirus_species", "hantavirus_species", "hydatigera_species", "k_pneumoniae_esbl",
-                           "lassa_mammaranevirus", "lassa_mammaranevirus", "lassa_mammaranevirus", "lassa_mammaranevirus", "lassa_mammaranevirus", "lassa_mammaranevirus",
-                           "lassa_mammaranevirus", "leishmania_major", "leishmania_species", "leptospirosis_species", "mammarenavirus_species", "mycobacterium_species",
-                           "mycobacterium_species", "mycoplasma_species", "orentia_species", "orthopoxvirus_species", "orthopoxvirus_species", "phlebovirus_species",
-                           "plagiorchis_species", "rickettsia_species", "rift_valley_fever_virus", "schistosoma_mansoni", "schistosoma_species", "toxoplasma_gondii", "trichuris_species",
-                           "trypanosoma_species", "usutu_virus"))
-names(pathogen_name) <- sort(c(unique(pathogen_data$path_1),
+pathogen_name <- as.list(c("anaplasma", "arenaviridae_species", "arenaviridae_species", "arenaviridae_species", "arenaviridae_species", "babesia_species", "bartonella_species",
+                           "bartonella_species", "borrelia_species", "borrelia_species", "borrelia_crocidurae", "borrelia_species", "e_coli_esbl", "ehrlichia_species", "eimeria_species",
+                           "flavivirus_species", "hantavirus_species", "hydatigera_species", "k_pneumoniae_esbl", "lassa_mammarenavirus", "lassa_mammarenavirus", "lassa_mammarenavirus",
+                           "lassa_mammarenavirus", "lassa_mammarenavirus", "lassa_mammarenavirus", "lassa_mammarenavirus", "leishmania_major", "leishmania_species",
[…]

### 2021-03-28 scripts/descriptive.R

@@ -13,6 +13,8 @@ rodent_data %<>%
   left_join(.,
             habitat_data ,
             by = "record_id")
+
+pathogen <- read_rds(here("data_clean", "pathogen.rds"))
 # Publication year --------------------------------------------------------
 ggplot(studies) +
   geom_bar(aes(x = year_publication)) +
@@ -205,5 +207,87 @@ count_genus <- species_data %>%
   arrange(-percent)
 
 # Pathogen ----------------------------------------------------------------
+pathogen_tested <- c("path_1", "path_2", "path_3", "path_4", "path_5", "path_6")
+pcr_test <- c("pcr_path_1_positive", "pcr_path_2_positive", "pcr_path_3_positive", "pcr_path_4_positive", "pcr_path_5_positive", "pcr_path_6_positive")
+ab_ag_test <- c("ab_ag_path_1_positive", "ab_ag_path_2_positive", "ab_ag_path_3_positive", "ab_ag_path_4_positive", "ab_ag_path_5_positive")
+culture_test <- c("culture_path_1_positive", "culture_path_1_positive", "culture_path_1_positive")
+direct_visualisation <- c("histo_path_1_positive", "histo_path_2_positive", "histo_path_3_positive", "histo_path_4_positive", "histo_path_5_positive", "histo_path_6_positive")
 
+group_pathogens <- as.list(c("borrelia_species", "amr_bacteria", "amr_bacteria", "arenaviridae_species", "leishmania_species", "mammarenavirus_species", "schistosoma_species"))
+names(group_pathogens) <- c("borrelia_crocidurae", "e_coli_esbl", "k_pneumoniae_esbl", "lassa_mammarenavirus", "leishmania_major", "mammarenavirus_species",
+                             "schistosoma_mansoni")
 
+pathogen_data <- tibble(pathogen) %>%
+  dplyr::select(-geometry) %>% # speed up processing by removing sf structure
+  left_join(., studies %>%
+              dplyr::select(unique_id, pathogen),
+            by = "unique_id")
+
+table(studies$pathogen)
+
+pathogen_data %>%
+  filter(pathogen == "Rodent pathogen") %>%
+  distinct(unique_id, across(all_of(pathogen_tested))) # rodent pathogens
+
+pathogen_data %>%
[…]

### 2021-03-28 scripts/update_data.R

@@ -7,8 +7,13 @@ drive_download("https://docs.google.com/spreadsheets/d/1rQYjHhk6uk1PoKZZVsgFlmuq
 studies <- read_xlsx(here("data_download", "included_studies.xlsx"), sheet = "study")
 write_rds(studies, here("data_raw", "studies.rds"))
 
-rodent_data <- read_xlsx(here("data_download", "included_studies.xlsx"), sheet = "trapping")
+rodent_data <- read_xlsx(here("data_download", "included_studies.xlsx"), sheet = "trapping", col_types = c("guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess",
+                                                                                     "guess", "guess", "guess", "numeric", "guess", "guess"))
 write_rds(rodent_data, here("data_raw", "rodent_data.rds"))
 
-pathogen_data <- read_xlsx(here("data_download", "included_studies.xlsx"), sheet = "pathogen")
+pathogen_data <- read_xlsx(here("data_download", "included_studies.xlsx"), sheet = "pathogen", col_types = c("guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess",
+                                                                                                             "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess",
+                                                                                                             "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess",
+                                                                                                             "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric",
+                                                                                                             "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric"))
 write_rds(pathogen_data, here("data_raw", "pathogen.rds"))

### 2021-03-30 reports/manuscript.Rmd
updated_search v4

@@
     orcid_id: 0000-0001-9655-1656
-output: distill::distill_article
+output: 
+  distill::distill_article
+always_allow_html: yes
 bibliography: "`r here::here('citations', 'manuscript_citations.bib')`"

### 2021-03-30 scripts/bibtex_to_df.R

@@ -16,3 +16,11 @@ included_studies <- bib2df(here("citations", "included_studies.bib"))%>%
                             ifelse(!is.na(ISSN), ISSN,
                                    ifelse(!is.na(ISBN), ISBN,
                                           ifelse(!is.na(PMID), PMID, NA))))) #load citations from the studies that were found to be suitable for inclusion in the review
+
+expanded_search <- bib2df(here("citations", "v4_search.bib"))
+
+search_4 <- expanded_search %>%
+  mutate(unique_id = ifelse(!is.na(DOI), DOI,
+                            ifelse(!is.na(ISSN), ISSN,
+                                   ifelse(!is.na(ISBN), ISBN,
+                                          ifelse(!is.na(BIBTEXKEY), BIBTEXKEY, NA)))))

### 2021-04-10 reports/manuscript.Rmd
caught up with new data

@@
 
-A total of 2,200 records were identified, with 101 studies included in narrative synthesis (see Figure \@ref(fig:prisma)).
+A total of 4,282 records were identified, with 124 studies included in narrative synthesis (see Figure \@ref(fig:prisma)).
 The characteristics of the included studies are shown in Table 1.
-The earliest studies identified were from 1974 with the majority (62%) published since 2010.
-Most of the included studies were classified as rodent ecology research (54%) with the remainder of studies on zoonotic diseases (46%).
+The earliest studies identified were from 1974 with the majority (61%) published since 2010.
+Most of the included studies were classified as rodent ecology research (56%) with the remainder of studies on zoonotic diseases (44%).
 
@@
 
-The detailed aims of the included studies are shown following division into the themes of ecological or zoonosis research.
+The detailed aims of the included studies are shown in Figure \@ref(fig:aims).
 Biodiversity of rodents was a shared aim between the two themes with 37 ecological studies and 2 zoonoses studies investigating this.
 The remaining aims were more disparate with ecological studies investigating rodent population dynamics, rodent species assemblages, the genetics of rodents and their behaviours.
-Most zoonotic disease research focussed on prevalence of pathogens in rodent populations, the risk of spillover to human populations, the identity of pathogen reservoir species with fewer studies investigating the phylogenetics of pathogen strains (see Figure \@ref(fig:aims).
+Most zoonotic disease research focussed on prevalence of pathogens in rodent populations, the risk of spillover to human populations, the identity of pathogen reservoir species with fewer studies investigating the phylogenetics of pathogen strains.
 
@@
 
-The included studies trapped in 16 countries, however, three countries were outside of the scope of this study (Cameroon, Chad and Morocco).
-At least a single trap site was recorded from 13 West African countries, with no rodent trapping reported from Gambia, Cape Verde or Togo.
+The included studies trapped in 17 countries, however, three countries were outside of the scope of this study (Cameroon, Chad and Morocco).
+At least a single trap site was recorded from 14 West African countries, with no rodent trapping reported from Gambia or Togo.
 
-Most studies (95%) were conducted in a single country, three studies were conducted in two countries, three studies were conducted in five countries and one study was conducted across six countries.
-This lead to 124 trapping activities being reported across the 13 West African countries.
-More trapping activity was reported from Senegal (20%) than any other country.
-Ghana (12%) and Guinea (12%) were the next most frequently trapped countries, followed by Nigeria (9%), Benin, Mali and Niger (7%), Sierra Leone (6%), Mauritania and the Ivory Coast (4%), Burkina Faso and Liberia (2%), Guinea-Bissau had a single recorded trapping activity.
+Most studies (93%) were conducted in a single country, five studies were conducted in two countries, three studies were conducted in five countries and one study was conducted across six countries.
+This lead to 146 trapping activities being reported across the 14 West African countries.
+More trapping activity was reported from Senegal (19%) than any other country.
+Nigeria (13%), Ghana and Guinea (both 12%) were the next most frequently trapped countries, followed by Mali and Benin (both 8%), Niger and Sierra Leone (both 6%), Mauritania and the Ivory Coast (4%), Burkina Faso (3%), Liberia (2%) with Guinea-Bissau and Cape Verde recording a single trapping activity.
 
-The trapping activity from the included studies was spread across 1,171 study sites.
-Twenty-four (24%) studies reported trapping only occurring at a single study site, 34 (33%) studies trapped at between two and five study sites the remaining 44 studies trapped at between six and 93 study sites.
-Trap sites were mapped to level one country administrative regions and density of sites per million square kilometers was calculated (see Figure \@ref(fig:density)).
+The trapping activity from the included studies was spread across 1,193 study sites.
+Thirty-one (25%) studies reported trapping only occurring at a single study site, 45 (37%) studies trapped at between two and five study sites the remaining 48 studies trapped at between six and 93 study sites.
+Trap sites were mapped to level two country administrative regions and density of sites per million square kilometers was calculated (see Figure \@ref(fig:density)).
 The areas with highest trapping densities included the capital cities of Niger, Sierra Leone, Senegal, Benin, Mali, Ghana and the Ivory Coast.
-Of the areas not incorporating capital cities Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore) and Eastern Sierra Leone were most intensively studied.
-Mapping to level two administrative regions is displayed in Supplementary figure \@ref(fig:supfig1).
+Of the areas not incorporating capital cities Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone were most intensively studied.
 
@@
 
-Of the included studies 15 did not state the type of trap used with three studies reporting supplementary hand trapping of rodents alongside trap use.
-For the remaining studies three used more than four types of rodent traps, seven used three types, 31 used two types the remaining 45 studies used a single trap type.
-The most commonly used trap was the Sherman trap (67%) [ref.] which is available in several sizes, locally made wire mesh traps were used in 29 studies (34%), less commonly used trap types included the Museum special, Tomahawk, Victor, Snaptraps and Firobind traps.
+Of the included studies 18 did not state the type of trap used with three studies reporting supplementary hand trapping of rodents alongside trap use.
+For the remaining studies three used more than four types of rodent traps, twelve used three types, 33 used two types the remaining 58 studies used a single trap type.
+The most commonly used trap was the Sherman trap (56%) which is available in several sizes, locally made wire mesh traps were used in 35 studies (28%), less commonly used trap types included the Museum special, Tomahawk, Victor, Snaptraps and Firobind traps.
 
@@
 
-Thirty-three studies did not describe the structure of the trap setup within the defined study area.
-Twenty five studies placed transect lines of traps through study sites with thirteen studies placing traps along a line but not clearly describing a transect across a defined site.
-Sixteen studies purposefully sampled houses, with four further studies purposefully placing traps in targeted locations (i.e.
+Forty studies did not describe the structure of the trap setup within the defined study area.
+Twenty seven studies placed transect lines of traps through study sites with nineteen studies placing traps along a line but not clearly describing a transect across a defined site.
+Twenty-two studies purposefully sampled houses, with nine further studies purposefully placing traps in targeted locations (i.e.
 near rivers, in fields, by houses).
-Nine studies placed traps in a structured grid across a single or multiple specified habitat types.
+Ten studies placed traps in a structured grid across a single or multiple specified habitat types.
 Three studies described a stratified or randomised approach to placing traps within a defined area.
-Trap setup was commonly reported in ecology studies (80%) with transect lines the most commonly used approach.
-Trap setup was reported in 53% of zoonoses risk studies with purposeful placement of traps in houses the most common approach (20%).
+Trap setup was commonly reported in ecology studies (81%) with transect lines the most commonly used approach.
+Trap setup was reported in 56% of zoonoses risk studies with purposeful placement of traps in houses the most common approach (24%).
 
@@
 Trapping effort was variably reported across included studies.
-Thirty studies did not report any measures of trapping effort.
-Thirty-four studies reported a measure of trapping effort at the same level of detail at which they reported rodent captures  (the unit of analysis) and were classified as complete reports.
-Thirty-seven studies had incomplete reports on trapping effort due to not presenting the number of trap nights at the appropriate unit of analysis (i.e.
+Thirty-nine studies did not report any measures of trapping effort.
+Forty-two studies reported a measure of trapping effort at the same level of detail at which they reported rodent captures  (the unit of analysis) and were classified as complete reports.
+Thirty-nine studies had incomplete reports on trapping effort due to not presenting the number of trap nights at the appropriate unit of analysis (i.e.
 number of study nights only or total number of trap-nights).
-Trapping effort was more completely reported from ecology studies (51% complete information and 20% no information) compared to zoonoses risk studies (13% complete information and 41% no information).
+Trapping effort was more completely reported from ecology studies (51% complete information and 21% no information) compared to zoonoses risk studies (11% complete information and 44% no information).
 
-Among studies that reported complete trap-night information the median number of trap-nights per study was 4,764 (Interquartile range (IQR) = 1,227-10,477) with a minimum of 240 trap-nights and maximum of 45,274 trap-nights.
-For studies reporting the number of trap-nights at the unit of analysis (typically study site or habitat type) the median number of trap-nights was 360 (IQR = 246-793).
-In studies with incomplete reports on trap-nights at the unit of analysis there was a median of 7,893 trap-nights (IQR = 1,401-10,331) conducted, for 21 studies reporting only on the number of days or nights trapping was performed on activities typically lasted between 1-4 nights.
+Among studies that reported complete trap-night information the median number of trap-nights per study was 4,418 (Interquartile range (IQR) = 1,187-10,252) with a minimum of 240 trap-nights and maximum of 45,274 trap-nights.
+For studies reporting the number of trap-nights at the unit of analysis (typically study site or habitat type) the median number of trap-nights at each site was 360 (IQR = 246-793).
+In studies with incomplete reports on trap-nights there was a median of 7,200 trap-nights (IQR = 1,184-9,424) conducted, for 26 studies reporting only on the number of days or nights trapping was performed on activities typically lasted between 1-4 nights.
 
@@
 
-No studies classified habitats with reference to a specific habitat classification scheme.
-Seven studies did not describe the local habitat where traps were placed (3 ecology, 4 zoonoses).
-The remaining studies used greater than 160 descriptions of the habitat type in which trapping was conducted, further, the resolution of the description ranged from the habitat of the location of the trap through to the study site.
-The 160 reported habitat types were combined into 32 categories (see Supplementary table 2 for the habitat dictionary) for further analysis.
-At least one habitat was recorded for 16,168 trap sites (95%), with two or more habitats for a single trap site recorded for 4,195 (26%) sites.
-Single trap sites could span multiple habitat types resulting in 20,369 habitat or trap sites.
-The most commonly trapped sites were in or around buildings (31%), in areas described as the rodents "natural habitat" (26%), in agricultural areas (13%) (e.g.
-rice fields, palm plantations), forests (6%) and within or around villages (5%).
+No studies reported trap habitats with direct reference to a habitat classification scheme.
+Thirteen studies did not describe the local habitat where traps were placed (4 ecology, 9 zoonoses).
+The remaining studies used more than 170 descriptions of the habitat type in which trapping was conducted, further, the resolution of the description varied between studies with some reporting the habitat of the location of the trap and others the study site or region.
+The reported habitat types were combined into 30 categories (see Supplementary table 2 for the habitat dictionary) for further analysis.
+At least one habitat was recorded for 17,122 trap sites (95%), with two or more habitats for a single trap site recorded for 4,403 (24%) sites.
+Single trap sites could span multiple habitat types resulting in 22,202 habitat or trap sites.
+The most commonly trapped sites were in or around buildings (29%), in areas described as the rodents "natural habitat" (24%), in agricultural areas (13%) (e.g.
+rice fields, palm plantations), forests (8%) and in the area surrounding buildings (4%).
 
-Habitat type was more comprehensively reported from ecology studies, 31 habitat categories were reported from ecology studies with the most commonly investigated habitats being agriculture (23%), in or around buildings (18%) and within forests (15%).
+Habitat type was more comprehensively reported from ecology studies, 30 habitat categories were reported from ecology studies with the most commonly investigated habitats being agriculture (23%), within forests (18%) and in or around buildings (16%).
 Among studies on zoonoses 15 habitat categories were reported, the most commonly reported classification was described as the rodents "natural habitat" (39%), within or around buildings (38%) and in agricultural settings (7%).
@@
 
-Nine included studies did not describe the method used to identify trapped rodents.
+Thirteen included studies did not describe the method used to identify trapped rodents.
 Most studies (52%) relied solely on morphometric identification against taxonomic keys or museum voucher specimens.
-Eight studies described molecular identification of rodents with 30% using a combination of morphological and molecular techniques.
-In more recently published studies (since 2010) 54% used molecular techniques to assist with speciation, compared with less than 16% in studies published before 2010.
+Nine studies described molecular identification of rodents with 30% using a combination of morphological and molecular techniques.
+In more recently published studies (since 2010) 76% used molecular techniques to assist with speciation, compared with less than 15% in studies published before 2010.
 
@@
 
-Across all West African study sites there were 59,343 trapped small mammals (592 were trapped outside of West African countries), 2,830 (4.8%) trapped individuals were identified to order level (Rodentia), 5,329 (9%) were identified to genus level, the remaining 51,184 (86%) were identified to species level.
-Of the 124 distinct identified species trapped (see Supplementary table 1) the majority were from the order Rodentia (96) with Muridae (79) being the largest family of rodents, followed by Gliridae and Nesomyidae (both 4), Sciuridae (3), Ctenodactylidae (2) and Anomaluridae, Dipodidae, Hystricidae and Thryonomidae (all 1).
-The remaining 29 species came from the orders of the Soricomorpha (25), Erinaceomorpha (2) and Afrosoricida (1).
+Across all West African study sites there were 73,164 trapped small mammals (592 were trapped outside of West African countries), 2,830 (4%) trapped individuals were identified to order level (Rodentia), 7,760 (11%) were identified to genus level, the remaining 62,574 (85%) were identified to species level.
+Of the 147 distinct identified species trapped (see Supplementary table 1) the majority were from the order Rodentia (112) with Muridae (82) being the largest family of rodents, Sciuridae (10), followed by Gliridae and Nesomyidae (both 6), Ctenodactylidae and Anomaluridae (2) and Dipodidae, Hystricidae and Thryonomidae (all 1).
+The remaining 34 species came from the orders of the Soricomorpha (31), Erinaceomorpha (2) and Afrosoricida (1).
 
-More individuals were trapped in studies conducted to investigate the risk of zoonoses (35,892, 60%) than from ecology studies (23,451 40%).
-Animals trapped as part of ecological research were more often identified to species level (93%) than those trapped in zoonoses research (82%).
-The most commonly trapped genera of rodents were *Mastomys sp.* (22,665, 39.7%), *Rattus sp.* (7,718, 13.5%), *Mus sp.* (7,324, 12.8%), *Praomys sp.* (4,977, 8.7%) and *Arvicanthis sp.* (4,408, 7.72%).
-At the species level *Mastomys natalensis* (10,351, 20.2%) and *Mastomys erythroleucus* (7,052, 13.8%) were the most commonly trapped rodents, *Rattus rattus* (6,228, 12.2%), *Mus musculus* (6,030, 11.8%), *Arvicanthis niloticus* (4,086, 8%) and *Mastomys huberti* (3,106, 6.1%) made up the next most commonly trapped species.
+A similar number of rodents were trapped for ecology studies (36,367) as for studies investigating the risk of zoonoses (36,205).
+Animals trapped as part of ecological research were more often identified to species level (90%) than those trapped in zoonoses research (82%).
+The most commonly trapped genera of rodents were *Mastomys sp.* (27,072, 38.5%), *Rattus sp.* (10,191, 14.5%), *Mus sp.* (8,624, 12.3%), *Arvicanthis sp.* (5,821, 8.3%) and *Praomys sp.* (5,409, 7.7%).
+At the species level *Mastomys natalensis* (11,221, 17.9%) and *Rattus rattus* (8,578, 13.7%) were the most commonly trapped rodents, *Mastomys erythroleucus* (8,578, 13.7%), *Mus musculus* (6,245, 8.8%), *Arvicanthis niloticus* (5,497, 8.8%) and *Mastomys huberti* (4,699, 7.5%) made up the next most commonly trapped species.
 Presence and absence maps for these six most commonly trapped species are shown in Figure \@ref(fig:dist).

### 2021-04-10 scripts/2_1_cleaning.R

@@ -1,0 +1,10 @@
+source(here::here("scripts", "libraries.r"))
+source(here("scripts", "update_data.r"))
+
+# Studies -----------------------------------------------------------------
+
+studies <- read_rds(here("data_raw", "studies.rds")) %>%
+  mutate(unique_id = as_factor(unique_id)) %>%
+  separate(col = aim_detail, into = c("aim_detail_1", "aim_detail_2"), sep = ", ")
+
+write_rds(studies, here("data_clean", "studies.rds"))

### 2021-04-10 scripts/2_2_clean_species.R

@@ -1,0 +1,145 @@
+source(here::here("scripts", "libraries.r"))
+
+rodent_data <- read_rds(here("data_raw", "rodent_data.rds")) %>%
+  mutate(country = as_factor(country),
+         record_id = 1:nrow(.))
+
+retain_columns <- c("unique_id", "year_trapping", "month_trapping", "country", "iso3c", "region", "town_village", "habitat", "intensity_use",
+                    "genus", "species", "classification", "gbif_id", "genus_gbif", "species_gbif",  "number", "trap_nights", "trap_night_unit", "record_id")
+
+genus_synonym <-  as.list(c("praomys", "mus"))
+names(genus_synonym) <- c("myomys", "nannomys")
+
+accepted_names <- as.list(c("mus musculus", "oenomys hypoxanthus", "suncus megalura", "dasymys incomtus", "atelerix albiventris", "gerbilliscus gambiana", "gerbilliscus guinea", "gerbilliscus guinea", "gerbilliscus kempii", "dipodillus campestris", "gerbilliscus gambiana", "grammomys poensis", "graphiurus nagtglasii", "graphiurus kelleni", "paraechinus aethiopicus", "lophuromys flavopunctatus", "massoutiera mzabi", "mastomys natalensis", "steatomys caurinus", "gerbilliscus kempii", "gerbilliscus guineae", "gerbilliscus guineae", "gerbilliscus kempii", "gerbilliscus robusta", "taterillus gracilis", "grammomys poensis"))
+names(accepted_names) <- c("mus domesticus", "aethomys hypoxanthus", "crocidura megalura", "dasymys bentleyae", "erinaceus albiventris", "gerbilliscus gambianus", "tatera guineae", "gerbilliscus guineae", "gerbilliscus kempi", "gerbillus campestris", "gerbillus gambianus", "grammomys rutilans", "graphiurus hueti", "graphiurus parvus", "hemiechinus aethiopicus", "lophuromys flavipunctatus", "massouteria mzabi", "mastomys hildebrandtii", "steatomys caurianus", "tatera gambiana", "tatera guinea", "tatera guineae", "tatera kempi", "tatera robusta", "taterillus gracillis", "thamnomys rutilans")
+
+write_rds(genus_synonym, here("data_clean", "genus_dictionary.rds"))
+write_rds(accepted_names, here("data_clean", "species_dictionary.rds"))
+
[…]

### 2021-04-10 scripts/bibtex_to_df.R

@@ -1,26 +1,18 @@
 source(here::here("scripts", "libraries.R"))
 
-citations <- bib2df(here("citations", "final_search.bib")) #load citations from bibtex file exported from zotero and convert to df
+citations <- bib2df(here("citations", "manuscript_citations.bib")) #load citations from bibtex file exported from zotero and convert to df
 
 missing_uid <- citations %>%
   filter(is.na(DOI) & is.na(ISSN) & is.na(ISBN) & is.na(PMID)) #identify records that do not contain a unique identifier in the form of a DOI, ISSN, ISBN or PMID
 
 citations_uid <- citations %>%
-  mutate(unique_id = ifelse(!is.na(DOI), DOI,
-                            ifelse(!is.na(ISSN), ISSN,
-                                   ifelse(!is.na(ISBN), ISBN,
-                                          ifelse(!is.na(PMID), PMID, NA)))))
+  mutate(unique_id = case_when(!is.na(DOI) ~ DOI,
+                               !is.na(ISSN) ~ ISSN,
+                               !is.na(ISBN) ~ ISBN,
+                               !is.na(PMID) ~ PMID))
 
-included_studies <- bib2df(here("citations", "included_studies.bib"))%>%
-  mutate(unique_id = ifelse(!is.na(DOI), DOI,
-                            ifelse(!is.na(ISSN), ISSN,
-                                   ifelse(!is.na(ISBN), ISBN,
-                                          ifelse(!is.na(PMID), PMID, NA))))) #load citations from the studies that were found to be suitable for inclusion in the review
-
-expanded_search <- bib2df(here("citations", "v4_search.bib"))
-
-search_4 <- expanded_search %>%
-  mutate(unique_id = ifelse(!is.na(DOI), DOI,
-                            ifelse(!is.na(ISSN), ISSN,
-                                   ifelse(!is.na(ISBN), ISBN,
-                                          ifelse(!is.na(BIBTEXKEY), BIBTEXKEY, NA)))))
+included_studies <- bib2df(here("citations", "include_final.bib"))%>%
+  mutate(unique_id = case_when(!is.na(DOI) ~ DOI,
+                               !is.na(ISSN) ~ ISSN,
+                               !is.na(ISBN) ~ ISBN,
+                               !is.na(PMID) ~ PMID)) #load citations from the studies that were found to be suitable for inclusion in the review

### 2021-04-10 scripts/clean_habitat.R

@@ -1,54 +1,0 @@
-source(here::here("scripts", "libraries.r"))
-
-rodent_data <- read_rds(here("data_clean", "rodent_df.rds")) %>%
-  tibble()
-habitat_split <- c("habitat_1", "habitat_2", "habitat_3", "habitat_4", "habitat_5", "habitat_6", "habitat_7")
-
-rodent_data %<>%
-  separate(habitat, into = c(habitat_split), sep = ", ", remove = F)
-
-habitat_types <- rodent_data %>%
-  dplyr::select(unique_id, country, region, town_village, all_of(habitat_split), trap_night_unit, trap_nights) %>%
-  pivot_longer(cols = all_of(habitat_split), values_to = "habitat_type") %>%
-  drop_na(habitat_type)
-
-if(!file.exists(here("data_clean", "habitat_dictionary.rds"))){
-sort(unique(habitat_types$habitat_type))
-cleaned_habitat <- as.list(c("agriculture", "agriculture", "airport", "alluvial_plain", "buildings_periphery", "nature_reserve", "botanical_garden", "buildings_periphery",
-                             "burned", "vegetation", "vegetation", "clay_soil", "buildings_periphery", "agriculture", "waterway", "waterway", "agriculture", "agriculture",
-                             "agriculture", "agriculture", "agriculture", "degraded_forest", "agriculture", "agriculture", "degraded_forest", "forest", "waterway", "waterway",
-                             "waterway", "forest", "sand_dunes", "vegetation", "fallow", "fallow", "fallow", "agriculture", "agriculture", "agriculture", "agriculture",
-                             "alluvial_plain", "forest", "forest", "degraded_forest", "degraded_forest", "gardens", "vegetation", "buildings_periphery", "vegetation", "forest",
-                             "vegetation", "vegetation", "buildings", "buildings", "agriculture", "buildings", "buildings_periphery", "buildings_periphery", "forest", "waterway",
-                             "island", "lacustrian", "waterway", "vegetation", "building_periphery", "building_periphery", "building_periphery", "marsh", "agriculture",
-                             "lateritic soil", "agriculture", "multiple", "nature_reserve", "forest_clearing", "natural_habitat", "natural_habitat", "forest", "agriculture",
-                             "forest", "natural_habitat", "agriculture", "village", "oasis", "agriculture", "agriculture", "natural_habitat", "agriculture", "agriculture",
[…]

### 2021-04-10 scripts/clean_pathogen.R

@@ -1,134 +1,0 @@
-source(here::here("scripts", "libraries.R"))
-
-genus_dictionary <- read_rds(here("data_clean", "genus_dictionary.rds"))
-species_dictionary <- read_rds(here("data_clean", "species_dictionary.rds"))
-
-
-# Pathogen ----------------------------------------------------------------
-
-# Cleaning species names
-pathogen_data <- read_rds(here("data_raw", "pathogen.rds")) %>%
-  mutate(country = as_factor(country),
-         record_id = 1:nrow(.),
-         unique_id = as_factor(unique_id),
-         genus = recode(genus, !!!genus_dictionary),
-         classification = paste(genus, ifelse(species == "-", "sp.", species), sep = " "),
-         classification = recode(classification, !!!species_dictionary)) %>%
-  separate(col = classification, into = c("genus", "species"), sep = " ", remove = F) %>%
-  mutate(species = ifelse(species %in% c("sp.", "sp.1", "sp.2"), "", species))
-
-# Matching to GBIF
-rodent_data <- read_rds(here("data_clean", "species_data.rds")) %>%
-  distinct(classification, gbif_id)
-
-pathogen_data <- pathogen_data %>%
-  left_join(., rodent_data,
-            by = "classification")
-
-# Cleaning habitat
-habitat_dictionary <- read_rds(here("data_clean", "habitat_dictionary.rds"))
-habitat_split <- c("habitat_1", "habitat_2", "habitat_3", "habitat_4", "habitat_5", "habitat_6", "habitat_7")
-
-pathogen_data <- pathogen_data %>%
-  separate(habitat, into = all_of(habitat_split), sep = ", ") %>%
-  mutate(across(.cols = all_of(habitat_split), ~recode(., !!!habitat_dictionary)))
-
-# Cleaning pathogen
-pathogen_tested <- c("path_1", "path_2", "path_3", "path_4", "path_5", "path_6")
-
-if(!file.exists(here("data_clean", "pathogen_dictionary.rds"))){
-pathogen_name <- as.list(c("anaplasma", "arenaviridae_species", "arenaviridae_species", "arenaviridae_species", "arenaviridae_species", "babesia_species", "bartonella_species",
-                           "bartonella_species", "borrelia_species", "borrelia_species", "borrelia_crocidurae", "borrelia_species", "e_coli_esbl", "ehrlichia_species", "eimeria_species",
[…]

### 2021-04-10 scripts/cleaning.R

@@ -1,10 +1,0 @@
-source(here::here("scripts", "libraries.r"))
-source(here("scripts", "update_data.r"))
-
-# Studies -----------------------------------------------------------------
-
-studies <- read_rds(here("data_raw", "studies.rds")) %>%
-  mutate(unique_id = as_factor(unique_id)) %>%
-  separate(col = aim_detail, into = c("aim_detail_1", "aim_detail_2"), sep = ", ")
-
-write_rds(studies, here("data_clean", "studies.rds"))

### 2021-04-10 scripts/cleaning_species.R

@@ -1,145 +1,0 @@
-source(here::here("scripts", "libraries.r"))
-
-rodent_data <- read_rds(here("data_raw", "rodent_data.rds")) %>%
-  mutate(country = as_factor(country),
-         record_id = 1:nrow(.))
-
-retain_columns <- c("unique_id", "year_trapping", "month_trapping", "country", "iso3c", "region", "town_village", "habitat", "intensity_use",
-                    "genus", "species", "classification", "gbif_id", "genus_gbif", "species_gbif",  "number", "trap_nights", "trap_night_unit", "record_id")
-
-genus_synonym <-  as.list(c("praomys", "mus"))
-names(genus_synonym) <- c("myomys", "nannomys")
-
-accepted_names <- as.list(c("mus musculus", "oenomys hypoxanthus", "suncus megalura", "dasymys incomtus", "atelerix albiventris", "gerbilliscus gambiana", "gerbilliscus guinea", "gerbilliscus guinea", "gerbilliscus kempii", "dipodillus campestris", "gerbilliscus gambiana", "grammomys poensis", "graphiurus nagtglasii", "graphiurus kelleni", "paraechinus aethiopicus", "lophuromys flavopunctatus", "massoutiera mzabi", "mastomys natalensis", "steatomys caurinus", "gerbilliscus kempii", "gerbilliscus guineae", "gerbilliscus guineae", "gerbilliscus kempii", "gerbilliscus robusta", "taterillus gracilis", "grammomys poensis"))
-names(accepted_names) <- c("mus domesticus", "aethomys hypoxanthus", "crocidura megalura", "dasymys bentleyae", "erinaceus albiventris", "gerbilliscus gambianus", "tatera guineae", "gerbilliscus guineae", "gerbilliscus kempi", "gerbillus campestris", "gerbillus gambianus", "grammomys rutilans", "graphiurus hueti", "graphiurus parvus", "hemiechinus aethiopicus", "lophuromys flavipunctatus", "massouteria mzabi", "mastomys hildebrandtii", "steatomys caurianus", "tatera gambiana", "tatera guinea", "tatera guineae", "tatera kempi", "tatera robusta", "taterillus gracillis", "thamnomys rutilans")
-
-write_rds(genus_synonym, here("data_clean", "genus_dictionary.rds"))
-write_rds(accepted_names, here("data_clean", "species_dictionary.rds"))
-
[…]

### 2021-04-10 scripts/download_spatial.R

@@ -1,75 +1,0 @@
-source(here::here("scripts", "libraries.r"))
-
-countries <- rodent_spatial %>%
-  drop_na(country) %>%
-  mutate(iso3 = countrycode(as.character(country), "country.name", "iso3c"))
-countries <- unique(countries$iso3)
-# Level 0 -----------------------------------------------------------------
-#WA
-BEN_0 <- getData("GADM", country = "Benin", level = 0, path = here("data_download", "admin_spatial"))
-BFA_0 <- getData("GADM", country = "BFA", level = 0, path = here("data_download", "admin_spatial"))
-CPV_0 <- getData("GADM", country = "CPV", level = 0, path = here("data_download", "admin_spatial"))
-GMB_0 <- getData("GADM", country = "GMB", level = 0, path = here("data_download", "admin_spatial"))
-GHA_0 <- getData("GADM", country = "Ghana", level = 0, path = here("data_download", "admin_spatial"))
-GIN_0 <- getData("GADM", country = "Guinea", level = 0, path = here("data_download", "admin_spatial"))
-GNB_0 <- getData("GADM", country = "GNB", level = 0, path = here("data_download", "admin_spatial"))
-CIV_0 <- getData("GADM", country = "CIV", level = 0, path = here("data_download", "admin_spatial"))
-LBR_0 <- getData("GADM", country = "LBR", level = 0, path = here("data_download", "admin_spatial"))
-MLI_0 <- getData("GADM", country = "Mali", level = 0, path = here("data_download", "admin_spatial"))
-MRT_0 <- getData("GADM", country = "Mauritania", level = 0, path = here("data_download", "admin_spatial"))
-NER_0 <- getData("GADM", country = "Niger", level = 0, path = here("data_download", "admin_spatial"))
-NGA_0 <- getData("GADM", country = "Nigeria", level = 0, path = here("data_download", "admin_spatial"))
-SEN_0 <- getData("GADM", country = "Senegal", level = 0, path = here("data_download", "admin_spatial"))
-SLE_0 <- getData("GADM", country = "SL", level = 0, path = here("data_download", "admin_spatial"))
-TGO_0 <- getData("GADM", country = "Togo", level = 0, path = here("data_download", "admin_spatial"))
-
-#non-WA
-TCD_0 <- getData("GADM", country = "Chad", level = 0, path = here("data_download", "admin_spatial"))
[…]

### 2021-04-10 scripts/mapping_presence.R

@@ -23,7 +23,20 @@ rodent_spatial <- read_rds(here("data_clean", "rodent_spatial.rds")) %>%
   filter(iso3c %in% wa_countries)
 studies <- read_rds(here("data_clean", "studies.rds"))
 
-top_6_species <- c(2438911, 2438912, 2439270, 7429082, 9733997, 4264823)
+count_species <- species_data %>%
+  filter(iso3c %in% wa_countries) %>%
+  group_by(species_gbif, classification) %>%
+  drop_na(species_gbif) %>%
+  summarise(number = sum(number)) %>%
+  mutate(percent = round(number/sum(.$number)*100, 2)) %>%
+  arrange(-percent) %>%
+  rename(`GBIF ID` = "species_gbif",
+         "Classification" = "classification",
+         "Number of individuals" = "number",
+         "Percent (%)" = "percent") %>%
+  mutate(Classification = snakecase::to_sentence_case(Classification)) # the number of individuals trapped identified to species level
+
+top_6_species <- head(count_species$`GBIF ID`, 6)
 
 top_6_spatial <- rodent_spatial %>%
   filter(species_gbif %in% top_6_species) %>%
@@ -43,3 +56,15 @@ top_6_plot <- tm_shape(level_0 %>%
   tm_facets(by = "classification")
 
 tmap_save(top_6_plot, here("figures", "top_6.png"))
+
+density(top_6_spatial)
+bw.diggle()
+
+library("maptools")
+p.sf.utm <- st_transform(top_6_spatial, 32619) # project from geographic to UTM
+p.sp  <- as(p.sf.utm, "Spatial")  # Create Spatial* object
+p.ppp <- as(p.sp, "ppp")
+class(p.ppp)
+
+as.ppp(test)
+

### 2021-04-10 scripts/mapping_traps.R

@@ -6,7 +6,7 @@ all_countries <- c("BEN", "BFA", "CIV", "CMR", "CPV", "DZA", "ESH", "GHA",
 wa_countries <- c("BEN", "BFA", "CIV", "CPV", "ESH", "GHA",
                   "GIN", "GMB", "GNB", "LBR", "MLI", "MRT",
                   "NER", "NGA", "SEN", "SLE", "TGO")
-no_data_countries <- c("GMB", "TGO", "CPV")
+no_data_countries <- c("GMB", "TGO")
 
 level_0 <- read_rds(here("data_download", "admin_spatial", "level_0_admin.rds"))
 list2env(level_0, envir = .GlobalEnv)
@@ -22,6 +22,8 @@ level_2 <- read_rds(here("data_download", "admin_spatial", "level_2_admin.rds"))
 list2env(level_2, envir = .GlobalEnv)
 level_2_all <- do.call(rbind.SpatialPolygonsDataFrame, level_2) %>%
   st_as_sf()
+
+studies <- read_rds(here("data_clean", "studies.rds"))
 
 rodent_spatial <- read_rds(here("data_clean", "rodent_spatial.rds"))
 bbox_rodent <- st_bbox(rodent_spatial)
@@ -75,8 +77,8 @@ afrmap <- tm_shape(afr) + tm_polygons() +
 print(afrmap, vp = grid::viewport(0.13, 0.88, width = 0.23, height = 0.23))
 }
 
-# Plot as mapview ---------------------------------------------------------
-tmap_mode("view")
+# Plot as leaflet ---------------------------------------------------------
+tmap_mode("plot")
 
 tm_shape(rodent_spatial) +
   tm_dots(col = "iso3c")
@@ -84,7 +86,32 @@ tm_shape(rodent_spatial) +
 mapview(rodent_spatial,
         zcol = "iso3c")
 
-tmap_mode("plot")
+trap_map <- rodent_spatial %>%
+  left_join(., studies %>%
+              dplyr::select(unique_id, first_author, year_publication, link),
+            by = "unique_id") %>%
+  distinct(unique_id, geometry, .keep_all = T)
+
+factcont <- colorNumeric("viridis", trap_map$year_publication, alpha = T, reverse = T)
+
+leaflet(trap_map) %>%
+  addProviderTiles(providers$CartoDB.Positron) %>%
+  addCircleMarkers(fillColor = ~factcont(year_publication),
+                   stroke = F,
+                   fillOpacity = 0.5,
[…]

### 2021-04-10 scripts/pathogen_dictionary.R

@@ -1,0 +1,72 @@
+anaplasma <- c("anaplasma")
+names(anaplasma) <- "anaplasma_species"
+arenaviridae <- c("arenaviridae_spp", "arenavirus", "arenavirus_spp", "arenaviruses")
+names(arenaviridae) <- rep("arenaviridae_species", length(arenaviridae))
+babesia <- c("babesia_spp")
+names(babesia) <- "babesia_species"
+bartonella <- c("bartonella", "bartonella_spp")
+names(bartonella) <- rep("bartonella_species", length(bartonella))
+borrelia <- c("borrelia", "borrelia_crocidurae", "borrelia_spp", "borrellia_spp")
+names(borrelia) <- rep("borrelia", length(borrelia))
+coxiella <- c("coxiella_burnetii")
+names(coxiella) <- "coxiella_burnetii"
+e_coli <- c("e_coli_esbl")
+names(e_coli) <- "e_coli_esbl"
+ehrlichia <- c("ehrlichia")
+names(ehrlichia) <- "ehrlichia_species"
+eimeria <- c("eimeria_spp")
+names(eimeria) <- "eimeria_species"
+flavivirus <- c("flavivirus")
+names(flavivirus) <- "flavivirus_species"
+hantavirus <- c("hantavirus")
+names(hantavirus) <- "hantavirus_species"
+hydatid <- "hydatigera_species"
+names(hydatid) <- "hydatigera_species"
+k_pneumoniae <- "k_pneumoniae_esbl"
+names(k_pneumoniae) <- "k_pneumoniae_esbl"
+lassa <- c("lassa", "lassa mammarenavirus", "lassa_mammarenavirus", "lassa_mammarenavirus_Ag", "lassa_mammarenavirus_antibody", "lassa_mammarenavirus_antigen",
+           "lassa_mammarenavirus_IgG_Ab")
+names(lassa) <- rep("lassa_mammarenavirus", length(lassa))
+leishmania <- c("leishmania_major", "leishmania_spp")
+names(leishmania) <- rep("leishmania_species", length(leishmania))
+leptospirosis <- "leptospirosis_spp"
+names(leptospirosis) <- "leptospirosis_species"
+mammarenavirus <- "mammarenavirus"
+names(mammarenavirus) <- "mammarenavirus_species"
+mycobacteria <- c("mycobacteria_spp", "mycobacterium_spp")
+names(mycobacteria) <- "mycobacterium_species"
[…]

### 2021-04-10 scripts/update_data.R

@@ -1,19 +1,0 @@
-source(here::here("scripts", "libraries.R"))
-
-register_google(google_api)
-
-drive_download("https://docs.google.com/spreadsheets/d/1rQYjHhk6uk1PoKZZVsgFlmuqWGicU2tTisk9ddfAwTM/edit#gid=0", path = here("data_download", "included_studies.xlsx"), overwrite = T)
-
-studies <- read_xlsx(here("data_download", "included_studies.xlsx"), sheet = "study")
-write_rds(studies, here("data_raw", "studies.rds"))
-
-rodent_data <- read_xlsx(here("data_download", "included_studies.xlsx"), sheet = "trapping", col_types = c("guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess",
-                                                                                     "guess", "guess", "guess", "numeric", "guess", "guess"))
-write_rds(rodent_data, here("data_raw", "rodent_data.rds"))
-
-pathogen_data <- read_xlsx(here("data_download", "included_studies.xlsx"), sheet = "pathogen", col_types = c("guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess",
-                                                                                                             "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess",
-                                                                                                             "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess",
-                                                                                                             "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric",
-                                                                                                             "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric"))
-write_rds(pathogen_data, here("data_raw", "pathogen.rds"))

### 2021-04-11 reports/manuscript.Rmd
share to website

@@
 
-Fifty-one studies presented data on microorganisms that infect or are carried by small mammal species in West Africa.
+X studies presented data on microorganisms that infect or are carried by small mammal species in West Africa.
 Data from a study previously excluded [@fichet-calvet_fluctuation_2007] due to duplicated data was incorporated as it provided the complementary pathogen data to the rodent trapping data.
-Six of the studies were solely investigated pathogens of rodents.
-The remaining 45 studies investigated organisms that were potentially zoonotic pathogens, pathogens of rodents or microbial organisms that were of uncertain significance to rodent or human health.
+X of the studies were solely investigated pathogens of rodents.
+The remaining X studies investigated organisms that were potentially zoonotic pathogens, pathogens of rodents or microbial organisms that were of uncertain significance to rodent or human health.
 
-Twenty-six studies used Polymerase Chain Reaction (PCR) to detect the presence of 22 different species or families of microorganisms.
-Ten studies used antibody or antigen based molecular tests to detect the presence of 9 different species or families of microorganisms.
-Seven studies conducted histological or direct visualisation assays of samples for 10 parasitic or bacterial species.
-Two studies performed direct viral culture of *Lassa mammarenavirus* to detect the presence of the virus in rodent specimens.
+X studies used Polymerase Chain Reaction (PCR) to detect the presence of X different species or families of microorganisms.
+X studies used antibody or antigen based molecular tests to detect the presence of 9 different species or families of microorganisms.
+X studies conducted histological or direct visualisation assays of samples for X parasitic or bacterial species.
+X studies performed direct viral culture of *Lassa mammarenavirus* to detect the presence of the virus in rodent specimens.
 
@@
 
-Forty-two of the 45 studies reported prevalence of zoonotic pathogens at a level of detail beyond a study wide summary.
-These studies typically tested for a single pathogen (35), with two studies testing for two and three pathogens and three studies testing for five or more.
-The most frequently tested for pathogens were *Lassa mammarenavirus* (11) or members of the *Arenaviridae* family (7), the spirochete bacteria *Borrelia* was investigated in six studies, *Bartonella* and *Toxoplasma gondii* were investigated in three each, the remaining 28 pathogens were reported in two or fewer studies.
+X of the X studies reported prevalence of zoonotic pathogens at a level of detail beyond a study wide summary.
+These studies typically tested for a single pathogen (X), with X studies testing for two and three pathogens and X studies testing for five or more.
+The most frequently tested for pathogens were *Lassa mammarenavirus* (X) or members of the *Arenaviridae* family (X), the spirochete bacteria *Borrelia* was investigated in X studies, *Bartonella* and *Toxoplasma gondii* were investigated in X each, the remaining X pathogens were reported in two or fewer studies.
 
-Ten of the studies reporting data on pathogen prevalence had incomplete data on the geolocation from which some of the animals tested were from.
-Six of these studies had no data on geolocation of the trapped rodents.
+X of the studies reporting data on pathogen prevalence had incomplete data on the geolocation from which some of the animals tested were from.
+X of these studies had no data on geolocation of the trapped rodents.
 
@@
 ## Supplementary material
-
-#### Supplementary figure 1
-
-``` {r supfig1, fig.cap = "Density of trap sites per thousand km^2^", echo = F, layout = "l-page"}
-knitr::include_graphics(here::here("figures", "static_site_density_2.png"))
-```
 

### 2021-04-11 scripts/mapping_presence.R

@@ -3,6 +3,9 @@ source(here::here("scripts", "libraries.r"))
 wa_mainland <- c("BEN", "BFA", "CIV", "ESH", "GHA",
                   "GIN", "GMB", "GNB", "LBR", "MLI", "MRT",
                   "NER", "NGA", "SEN", "SLE", "TGO")
+wa_countries <- c("BEN", "BFA", "CIV", "CPV", "ESH", "GHA",
+                  "GIN", "GMB", "GNB", "LBR", "MLI", "MRT",
+                  "NER", "NGA", "SEN", "SLE", "TGO")
 
 level_0 <- read_rds(here("data_download", "admin_spatial", "level_0_admin.rds"))
 list2env(level_0, envir = .GlobalEnv)
@@ -19,6 +22,7 @@ list2env(level_2, envir = .GlobalEnv)
 level_2_all <- do.call(rbind.SpatialPolygonsDataFrame, level_2) %>%
   st_as_sf()
 
+species_data <- read_rds(here("data_clean", "species_data.rds"))
 rodent_spatial <- read_rds(here("data_clean", "rodent_spatial.rds")) %>%
   filter(iso3c %in% wa_countries)
 studies <- read_rds(here("data_clean", "studies.rds"))
@@ -57,14 +61,25 @@ top_6_plot <- tm_shape(level_0 %>%
 
 tmap_save(top_6_plot, here("figures", "top_6.png"))
 
-density(top_6_spatial)
-bw.diggle()
-
-library("maptools")
-p.sf.utm <- st_transform(top_6_spatial, 32619) # project from geographic to UTM
-p.sp  <- as(p.sf.utm, "Spatial")  # Create Spatial* object
-p.ppp <- as(p.sp, "ppp")
-class(p.ppp)
-
-as.ppp(test)
-
+# library("maptools")
+# library("spatstat")
+# t6_sf_utm <- st_transform(top_6_spatial %>%
+#                             filter(classification == "Mastomys natalensis" & pres_abs == "Present"),
+#                           32619) # project from geographic to UTM
+# t6_sp  <- as(t6_sf_utm, "Spatial")  # Create Spatial* object
+# t6_ppp <- as(t6_sp, "ppp")
+# t6_ppp$marks <- NULL
+#
+# wa_sp <- level_0 %>%
+#   filter(GID_0 %in% wa_mainland) %>%
+#   st_union() %>%
[…]

### 2021-04-11 reports/manuscript.Rmd
knit to root

@@
   distill::distill_article
+knit: (function(inputFile, encoding) {
+    rmarkdown::render(inputFile, encoding = encoding, output_dir = here::here())
+  })
 always_allow_html: yes

### 2021-04-19 reports/manuscript.Rmd
Update manuscript.Rmd

@@
 Commensal species thrive in human adapted landscapes nesting in houses and obtaining nutrition from human food storage and preparation areas.
-In sub-Saharan Africa, Mastomys spp.
-and Arvicanthis spp.
-are identified as causing a significant burden as agricultural pests [@stenseth_mice_2003].
+In sub-Saharan Africa, *Mastomys spp.* and *Arvicanthis spp.* are identified as causing a significant burden as agricultural pests [@stenseth_mice_2003].
 Crop loss within Afro-Malagasy small-holder farming communities is high, with 15% of pre-harvest crop being lost to rodent pest activity [@swanepoel_systematic_2017].
@@
 animal pathogens that impact human health) are transmitted via two pathways – a direct and an indirect pathway.
-The direct transmission pathway involves transmission of pathogens through rodents biting or scratching a human, or through contamination of food or water sources.
+The direct transmission pathway involves transmission of pathogens through a rodent biting or scratching a human, or through contamination of food or water sources.
 The indirect pathway involves rodents acting as amplifying hosts for ectoparasites that themselves are vectors of zoonoses [@meerburg_rodent-borne_2009].
 Several rodent-borne pathogens are global in their distribution, including bacteria such as Lyme disease (caused by *Borrelia burgdorferi*) and Leptospirosis (caused by *Leptospira spp.*) and parasites such as Toxoplasmosis (caused by *Toxoplasma gondii*) and Leishmaniasis (caused by *Leishmania spp.*).
-However, most rodent-borne pathogens have more limited geographic distributions, particularly viral diseases such as Hantavirus pulmonary syndrome (caused by Hantaviridae) concentrated in the Americas and Lassa fever (caused by Arenaviridae) limited to West Africa.
+However, most rodent-borne pathogens have more limited geographic distributions, particularly viral diseases such as Hantavirus pulmonary syndrome (caused by Hantaviridae) typically concentrated in the Americas and Lassa fever (caused by Arenaviridae) limited to West Africa.
 
-Given the significant economic and public health burden of rodents, monitoring rodent distributions is therefore an ongoing focus of research efforts across multiple scientific disciplines, including (but not limited to) conservation science, development studies and infectious disease epidemiology.
-The distribution of rodent populations can be monitored via rodent trapping efforts, with traps deployed in habitats expected to contain rodent species of interest.
+The significant economic and public health burden of rodents has led to the monitoring of rodent distributions being an important focus of research efforts across multiple scientific disciplines, including (but not limited to) conservation science, development studies and infectious disease epidemiology.
+The distribution of rodent populations can be effectively monitored through rodent trapping, with traps deployed in habitats expected to contain rodent species of interest.
 
-There is increasing interest in the link between environmental and human factors and rodent populations to land use change and human population change.
+Further, there is increasing interest in the link between environmental and human factors and rodent populations to land use change and human population change.
 A recent review of studies included in the PREDICTS database [@hudson_predicts_2014] identified that land use change from primary (i.e.
 intact forests) to managed and urban systems is associated with a change in rodent assemblages towards species that are more likely to be hosts of zoonoses [@gibb_zoonotic_2020].
-*Mastomys natalensis populations* in Tanzania, East Africa are sensitive to seasonal meteorological cycles, in years with below average rainfall and short wet seasons population density is lower [@makundi_reproduction_2007].
+*Mastomys natalensis* populations in Tanzania, East Africa are sensitive to seasonal meteorological cycles, in years with below average rainfall and short wet seasons population density is lower [@makundi_reproduction_2007] with reduced pest activity as a result.
 Future climate change projections include longer and wetter West African monsoon seasons which could lead to increased zoonotic disease host rodent populations [@akinsanola_projection_2019].
@@
 *Mastomys natalensis* and *Arvicanthus niloticus*) and several potential reservoirs of zoonotic diseases (e.g.
-*Lemniscomys striatus*), this will be the geographic region of interest in this work.
+*Lemniscomys striatus*), this scoping review will be limited to considering the evidence produced from West Africa.
 
@@
 
-We specifically aim to address the following research questions :
+We specifically aim to address the following research questions:
 
@@
 ### Study design
+
+A scoping review approach was chosen to support a more descriptive and exploratory approach than using a systematic review.
 
@@
 
-The brand or description of the rodent trap was obtained from each study. For those using multiple devices all types were recorded. Type of trapping method. Effort of trapping
+The brand or description of the rodent trap was obtained from each study.
+For those using multiple devices all types were recorded.
+The method employed when setting the traps was extracted where reported.
+Studies that clearly reported setting a transect line were classified as such, for studies reporting using a line but not through a specified habitat trap line was used.
+
+The trapping effort in trap-nights was recorded.
+A trap-night is the setting of a single trap for a single night.
+If this was reported at the habitat level or the level at which the study reported rodent captures it was classified as a complete recording of trapping effort.
+For studies that reported trapping effort at different levels to the capture level it was recorded as incomplete.
+For studies reporting number of nights trapping occured for or for where the number of traps set was stated but the number of nights not recorded they were classified as incomplete.
 
 ### Habitat classification
+
+The habitat classification scheme a study used was recorded if reported.
+For studies not using standardised recording of habitat types the explicit description from the study was associated with the trap.
+For studies reporting multiple habitat types for a single trap, trap-line or trapping grid a higher order classification was recorded to encompass the types.
 

### 2021-04-20 reports/manuscript.Rmd
Update manuscript.Rmd

@@
 
-  +	RQ1: In which West African countries or regions have rodent trapping studies been performed?
-  + RQ2: What are the stated aims of these rodent trapping studies?
-  + RQ3: Which rodent species are being targeted for trapping activities? Are species with high potential for hosting zoonotic pathogens targeted at higher rates?
-  + RQ4: How is the type of land use and/or land use intensity classified in the literature and are classifications comparable across studies?
-  +	RQ5: Is rodent trapping occurring in all habitat types or are some habitats targeted at higher rates?
-  +	RQ6: Is there an association of land use change, human population change or the implementation of regulatory strategies with rodent species distributions?
+  +	1. In which West African countries or regions have rodent trapping studies been performed?
+  + 2. What are the stated aims of these rodent trapping studies?
+  + 3. Which rodent species are being targeted for trapping activities? Are species with high potential for hosting zoonotic pathogens targeted at higher rates?
+  + 4. How is the type of land use and/or land use intensity classified in the literature and are classifications comparable across studies?
+  +	5. Is rodent trapping occurring in all habitat types or are some habitats targeted at higher rates?
 
@@
 
-Trap success rates
+Trap success rates at study level
+
+The most commonly trapped species (*Mastomys natalensis*) is of particular importance in West Africa due to it being the reservoir host of Lassa fever.
+*Rattus sp.* and *Mus sp.* are also known to harbour zoonotic pathogens while *Arvicanthis sp.* and *Praomys sp.* are less commonly associated with zoonoses.
 
@@
 
+### Pathogens and their rodent hosts
+
+X rodent species were investigated for the presence of a zoonotic pathogen. X individual rodents were tested for at least one of the above pathogens.
+
+The rodents most commonly found to be hosts of zoonoses were...
+
+The species with the greatest number of detected pathogens within the studied areas were...
+
+*Lassa mammarenavirus* was detected in X individuals either through PCR (demonstrating acute infection) or antibody studies (suggesting previous infection or the acquisition of maternal antibodies).
+These individuals were from X species (X *Mastomys natalensis*, Y *Praomys* ...).
+
+*Borrelia sp.* were detected in X individuals ...
+
+X species of those tested were not found to be infected by any zoonotic pathogen investigated for.
+
+###
+
 ## Supplementary material

### 2021-04-20 reports/manuscript.Rmd
Update manuscript.Rmd

@@
 
-###
+## Discussion
+
+This scoping review aimed to consolidate and summarise research conducted on rodents and their potential zoonoses in West Africa.
+A comprehensive search identified X studies performed in Y West African countries.
+Around half of these studies were conducted to understand the ecology of rodents with the others being conducted to understand the risk of zoonotic pathogens to human populations in West Africa.
+
+Rodent trappping has been performed in most West African countries since the first recorded study from 197x, however, the distribution of effort between the countries and regions within those countries has been highly skewed.
+More than 50% of trapping activity has occurred in Senegal, Nigeria, Ghana and Guinea.
+Capital cities and their surrounding regions were more commonly studied than more sparsely populated areas.
+
+The studies conducted to assess rodent ecology typically focussed on investigating rodent biodiveristy or conducting small mammal surveys of national parks or areas of specific scientific interest.
+Other studies investigated rodnt population dynamics using mark and release methods to measure rodent recruitment (birth rates) and survival.
+
+  + 2. What are the stated aims of these rodent trapping studies?
+  + 3. Which rodent species are being targeted for trapping activities? Are species with high potential for hosting zoonotic pathogens targeted at higher rates?
+  + 4. How is the type of land use and/or land use intensity classified in the literature and are classifications comparable across studies?
+  +	5. Is rodent trapping occurring in all habitat types or are some habitats targeted at higher rates?
 

### 2021-04-23 reports/manuscript.Rmd
change to more detailed map

@@
 ``` {r density, fig.cap = "Density of trap sites per thousand km^2^", echo = F, layout = "l-page"}
-knitr::include_graphics(here::here("figures", "static_site_density.png"))
+knitr::include_graphics(here::here("figures", "static_site_density_2.png"))
 ```

### 2021-04-23 reports/manuscript.Rmd
add toc

@@
 output: 
-  distill::distill_article
+  distill::distill_article:
+  toc: true
 knit: (function(inputFile, encoding) {

### 2021-04-26 reports/manuscript.Rmd
update 26-04

@@
 
-Rodents (Rodentia) are abundant and diverse, representing around 40% of the total number of mammalian species [@american_society_of_mammologists_asm_2021].
+Rodents (Rodentia) are abundant and diverse, representing around 40% of all mammalian species [@american_society_of_mammologists_asm_2021].
 Rodents typically demonstrate ‘fast’ life history strategies characterised by early maturation, short generation times, low juvenile and adult survival and high fecundity [@dobson_fast_2007].
-There is a degree of between-species heterogeneity, with commensal species and those that are reservoirs of zoonoses having traits consistent with faster life histories (Han et al., 2015).
-Rodents often compete with humans for food produced in agricultural systems [@fiedler_rodent_1988], with 5-10% of rodent species classed as major pests.
-Commensal species thrive in human adapted landscapes nesting in houses and obtaining nutrition from human food storage and preparation areas.
-In sub-Saharan Africa, *Mastomys spp.* and *Arvicanthis spp.* are identified as causing a significant burden as agricultural pests [@stenseth_mice_2003].
+Between-species heterogeneity exists, with commensal species and those that are reservoirs of zoonoses having traits consistent with faster life histories (Han et al., 2015).
+Rodent species compete with humans for food produced in agricultural systems [@fiedler_rodent_1988], 5-10% of rodent species are classed as major crop pests.
+Commensal species thrive in human adapted landscapes nesting in houses and scavenging human food storage and preparation areas.
+In sub-Saharan Africa, *Mastomys spp.* and *Arvicanthis spp.* are two species causing significant burden as agricultural pests [@stenseth_mice_2003].
 Crop loss within Afro-Malagasy small-holder farming communities is high, with 15% of pre-harvest crop being lost to rodent pest activity [@swanepoel_systematic_2017].
 
-Another important issue is the transmission of infectious diseases from rodents to humans.
+Beyond crop loss a further threat to local communities is the transmission of infectious diseases from rodents to humans.
 Zoonotic diseases (i.e.
-animal pathogens that impact human health) are transmitted via two pathways – a direct and an indirect pathway.
+animal pathogens or microorganisms residing in animals that impact human health) are transmitted via two pathways – a direct and an indirect pathway.
 The direct transmission pathway involves transmission of pathogens through a rodent biting or scratching a human, or through contamination of food or water sources.
 The indirect pathway involves rodents acting as amplifying hosts for ectoparasites that themselves are vectors of zoonoses [@meerburg_rodent-borne_2009].
-Several rodent-borne pathogens are global in their distribution, including bacteria such as Lyme disease (caused by *Borrelia burgdorferi*) and Leptospirosis (caused by *Leptospira spp.*) and parasites such as Toxoplasmosis (caused by *Toxoplasma gondii*) and Leishmaniasis (caused by *Leishmania spp.*).
-However, most rodent-borne pathogens have more limited geographic distributions, particularly viral diseases such as Hantavirus pulmonary syndrome (caused by Hantaviridae) typically concentrated in the Americas and Lassa fever (caused by Arenaviridae) limited to West Africa.
+Rodent-borne pathogens vary in their scale from those with global distribution, such as the bacteria *Borrelia burgdorferi* (causing Lyme disease) and *Leptospira spp.* (casing Leptospirosis) and parasites such as *Toxoplasma gondii* (causing Toxoplasmosis) and *Leishmania spp.* (causing visceral and cutaneous Leishmaniasis).
+However, the majority of rodent-borne pathogens have more limited geographic distributions, viral diseases such as Hantavirus pulmonary syndrome (caused by Hantaviridae) are typically concentrated in the Americas and Lassa fever (caused by Arenaviridae) is limited to West Africa [@mills_ecologic_1998].
 
-The significant economic and public health burden of rodents has led to the monitoring of rodent distributions being an important focus of research efforts across multiple scientific disciplines, including (but not limited to) conservation science, development studies and infectious disease epidemiology.
+These economic and public health burdens have led to the monitoring of rodent distributions being performed across multiple scientific disciplines, including (but not limited to) conservation science, development studies and infectious disease epidemiology.
 The distribution of rodent populations can be effectively monitored through rodent trapping, with traps deployed in habitats expected to contain rodent species of interest.
 
-Further, there is increasing interest in the link between environmental and human factors and rodent populations to land use change and human population change.
+Further, there is increasing interest in the understanding the associations between environmental, human activity and rodent populations with land use change.
 A recent review of studies included in the PREDICTS database [@hudson_predicts_2014] identified that land use change from primary (i.e.
 intact forests) to managed and urban systems is associated with a change in rodent assemblages towards species that are more likely to be hosts of zoonoses [@gibb_zoonotic_2020].
-*Mastomys natalensis* populations in Tanzania, East Africa are sensitive to seasonal meteorological cycles, in years with below average rainfall and short wet seasons population density is lower [@makundi_reproduction_2007] with reduced pest activity as a result.
+*Mastomys natalensis* populations (an important pest and zoonotic reservoir species) in Tanzania, East Africa are sensitive to seasonal meteorological cycles, in years with below average rainfall and short rainy seasons population density is observed to be lower [@makundi_reproduction_2007] with reduced pest activity.
 Future climate change projections include longer and wetter West African monsoon seasons which could lead to increased zoonotic disease host rodent populations [@akinsanola_projection_2019].
@@
     + Trapping of small mammals including at least one species of Rodentia.
-    + Descriped trap used, length of trapping activity or location of trapping activity.
+    + Described trap used, length of trapping activity or location of trapping activity.
     + Included trapping activity from at least one West African country: Benin, Burkina-Faso, Cape Verde, Gambia, Ghana, Guinea, Guinea-Bissau, Cote d'Ivoire, Liberia, Mali, Mauritania, Niger, Nigeria, Senegal, Sierra Leone and Togo
@@
 
-### Study aims
+#### Study aims
 
 Where explicitly stated the study aims were extracted.
-For those with no recorded aims the specific aims were inferred from the background or conclusions of manuscript or report.
-The aims of the studies were broadly categorised as pertaining to rodent ecology research or studies on risk of zoonoses and further categorised into detailed aims.
+For those with no recorded aims the specific aims were inferred from the introduction or conclusions of manuscript or report.
+The aims of the studies were categorised by the research team with studies subsequently broadly categorised as pertaining to rodent ecology research or studies on risk of zoonoses.
 
-### Geolocation of trapping activity
+#### Geolocation of trapping activity
 
@@
 trap, trap-line, study-site or study region.
-Locations were extracted in the format reported and converted to decimal degrees.
-Where no GPS location is reported coordinates will be matched to study locations using the National Geospatial Intelligence Agency NGA GEOnet Names Server [ref.] based on study site name/region and through maps presented in the manuscript.
+Coordinates were extracted in the format reported and converted to decimal degrees.
+Where no GPS location is reported coordinates will be matched to study locations using the National Geospatial Intelligence Agency NGA GEOnet Names Server [@national_geospatial-intelligence_agency_nga_nodate] based on study site name/region and through maps presented in the manuscript.
 
-### Trapping methodology
+#### Trapping methodology
 
-The brand or description of the rodent trap was obtained from each study.
-For those using multiple devices all types were recorded.
+The brand, if reported, or the description of the structure of the rodent trap was obtained from each study.
+For studies using multiple devices all types were recorded.
 The method employed when setting the traps was extracted where reported.
-Studies that clearly reported setting a transect line were classified as such, for studies reporting using a line but not through a specified habitat trap line was used.
+Studies that reported setting a transect line were classified as such, for studies reporting using a line of traps but not through a specified habitat the term trap line was used.
 
@@
 If this was reported at the habitat level or the level at which the study reported rodent captures it was classified as a complete recording of trapping effort.
-For studies that reported trapping effort at different levels to the capture level it was recorded as incomplete.
-For studies reporting number of nights trapping occured for or for where the number of traps set was stated but the number of nights not recorded they were classified as incomplete.
+For studies that reported trapping effort at different levels to the capture level the recording of trap-nights was classified as incomplete.
+For studies reporting number of nights trapping occurred for or for where the number of traps set was stated but the number of nights not recorded they were classified as incomplete.
 
-### Habitat classification
+#### Habitat classification
 
-The habitat classification scheme a study used was recorded if reported.
+The habitat classification scheme a study used was recorded.
 For studies not using standardised recording of habitat types the explicit description from the study was associated with the trap.
-For studies reporting multiple habitat types for a single trap, trap-line or trapping grid a higher order classification was recorded to encompass the types.
+For studies reporting multiple habitat types for a single trap, trap-line or trapping grid a higher order classification of habitat type was recorded.
 
-### Genus and species definitions
+#### Genus and species definitions
 
-Classification of species is variable over time as morphological and genetic information emerges.
-To handle species classification synonyms and reclassification the reported names of trapped species was compared to the Global Biodiversity Information Facility [ref.].
+Species classifications can vary over time as morphological and genetic information emerges.
+To handle species classification synonyms and reclassification the reported names of trapped species was compared to the Global Biodiversity Information Facility [@global_biodiversity_information_facility_gbiforg_nodate].
 For species with updated taxonomies the synonym was converted to the accepted species name for all subsequent analyses.
@@
 
-### Pathogen testing
+#### Zoonoses testing
+
+In studies investigating rodents for potential zoonoses the pathogens tested for were extracted.
+For studies reporting multiple potential pathogens all of the organisms were extracted.
+For studies solely reporting on pathogens of rodents (i.e.
+rodent parasites) pathogens were grouped as "Rodent pathogens".
+The number of rodents tested and the number of positive or negative samples were extracted alongisde the type of assay used to detect the pathogens (e.g.
+Polymerase Chain Reaction (PCR), Enzyme Linked ImmunoSorbent Assay (ELISA) or viral culture).
+Where possible pathogens were identified to the species level, however, where an assay only allows for attribution to a family of viruses or bacteria the higher order grouping was used.
+
+### Analysis
+
+Descriptive analyis of included studies was conducted in R (v4.0.2) [@r_core_team_r_2020] within the RStudio IDE [@rstudio_team_rstudio_2020].
+The analysis code and packages used for this analysis are available on GitHub [repository](https://doi.org/10.5281/zenodo.4718375).
+
+The locations of traps, trap-sites and studies were assigned to level 2 administrative areas in West Africa using shapefiles maintained by GADM [@database_of_global_administrative_areas_gadm_nodate].
 
@@
 A total of 4,282 records were identified, with 124 studies included in narrative synthesis (see Figure \@ref(fig:prisma)).
-The characteristics of the included studies are shown in Table 1.
+A table of the included studies and their DoIs is presented in Supplementary table 1.
 The earliest studies identified were from 1974 with the majority (61%) published since 2010.
@@
 ``` {r aims, fig.cap = "The aims of included studies following division into ecological and zoonotic infectious disease themes.
-Multiple aims were identified from each study, the number on the arrow indicates the number of distinct studies with that aim ", echo = F, layout = "l-page"}
+Multiple aims were identified from each study, the number on the arrow indicates the number of distinct studies with that aim.
+Opening the image separately will allow you to zoom in.", echo = F, layout = "l-page"}
 knitr::include_graphics(here::here("figures", "study_aims.png"))
@@
 
-The included studies trapped in 17 countries, however, three countries were outside of the scope of this study (Cameroon, Chad and Morocco).
+Trapped occurred in 17 countries, however, three countries were outside of the scope of this study (Cameroon and Chad in Central Africa and Morocco in Northern Africa).
 At least a single trap site was recorded from 14 West African countries, with no rodent trapping reported from Gambia or Togo.
@@
 Most studies (93%) were conducted in a single country, five studies were conducted in two countries, three studies were conducted in five countries and one study was conducted across six countries.
-This lead to 146 trapping activities being reported across the 14 West African countries.
+In summary 146 trapping activities were reported across the 14 West African countries.
 More trapping activity was reported from Senegal (19%) than any other country.
@@
 The trapping activity from the included studies was spread across 1,193 study sites.
-Thirty-one (25%) studies reported trapping only occurring at a single study site, 45 (37%) studies trapped at between two and five study sites the remaining 48 studies trapped at between six and 93 study sites.
-Trap sites were mapped to level two country administrative regions and density of sites per million square kilometers was calculated (see Figure \@ref(fig:density)).
+Thirty-one (25%) studies reported trapping only at a single study site, 45 (37%) studies trapped at between two and five study sites, the remaining 48 studies trapped at between six and 93 study sites.
+Trap sites were mapped to level two country administrative regions and density of sites per thousand square kilometers was calculated (see Figure \@ref(fig:density)).
 The areas with highest trapping densities included the capital cities of Niger, Sierra Leone, Senegal, Benin, Mali, Ghana and the Ivory Coast.
-Of the areas not incorporating capital cities Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone were most intensively studied.
+Of the areas not incorporating capital cities Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone were the most intensively studied regions.
 
@@
 
-Of the included studies 18 did not state the type of trap used with three studies reporting supplementary hand trapping of rodents alongside trap use.
+Eighteen studies did not state the type of trap used, with three studies reporting supplementary hand trapping of rodents alongside trap use.
 For the remaining studies three used more than four types of rodent traps, twelve used three types, 33 used two types the remaining 58 studies used a single trap type.
@@
 Nine studies described molecular identification of rodents with 30% using a combination of morphological and molecular techniques.
-In more recently published studies (since 2010) 76% used molecular techniques to assist with speciation, compared with less than 15% in studies published before 2010.
+In more recently published studies since 2010, 76% used molecular techniques to assist with speciation, compared with less than 15% in studies published prior to 2010.
 
@@
 
-Across all West African study sites there were 73,164 trapped small mammals (592 were trapped outside of West African countries), 2,830 (4%) trapped individuals were identified to order level (Rodentia), 7,760 (11%) were identified to genus level, the remaining 62,574 (85%) were identified to species level.
-Of the 147 distinct identified species trapped (see Supplementary table 1) the majority were from the order Rodentia (112) with Muridae (82) being the largest family of rodents, Sciuridae (10), followed by Gliridae and Nesomyidae (both 6), Ctenodactylidae and Anomaluridae (2) and Dipodidae, Hystricidae and Thryonomidae (all 1).
+Across all West African study sites there were 73,164 trapped small mammals (592 were trapped outside of West African countries), 2,830 (4%) trapped individuals were only identified to order level (Rodentia), 7,760 (11%) were identified to genus level, the remaining 62,574 (85%) were identified to species level.
+Of the 147 distinct identified species trapped (see Supplementary table 3) the majority were from the order Rodentia (112) with Muridae (82) being the largest family of rodents, Sciuridae (10), followed by Gliridae and Nesomyidae (both 6), Ctenodactylidae and Anomaluridae (2) and Dipodidae, Hystricidae and Thryonomidae (all 1).
 The remaining 34 species came from the orders of the Soricomorpha (31), Erinaceomorpha (2) and Afrosoricida (1).
@@
 
-Trap success rates at study level
-
-The most commonly trapped species (*Mastomys natalensis*) is of particular importance in West Africa due to it being the reservoir host of Lassa fever.
-*Rattus sp.* and *Mus sp.* are also known to harbour zoonotic pathogens while *Arvicanthis sp.* and *Praomys sp.* are less commonly associated with zoonoses.
-
-### Rodent biodiversity
-
-Diversity indices, accumulation measures
-
 ``` {r dist, fig.cap = "Presence and absence of the 6 most commonly trapped species in included studies, no adjustment made for trapping effort", echo = F, layout = "l-page"}
@@
 
+### Trap success
+
+Trap success is often used as a measure of rodent abundance in the absence of mark-recapture studies.
+In the 61 studies from which trap success could be calculated 34,891 rodents were captured in 476,332 trap-nights equivalent to a trap success rate of 7.3%.
+The trap success rate was slightly higher from studies reporting the number of trap-nights at the same level (i.e.
+habitat) as rodent captures (9.4%) than from studies reporting trap-nights at a higher level (village 4%, study 4.4%).
+
+### Rodent biodiversity
+
+Twenty-six studies reported a numeric measure of rodent species diversity or richness, these included the Shannon Index, Simpson's Index and Chao Index.
+Six of these studies present a species accumulation curve to estimate the completeness of their sampling activity.
+A further 8 studies that did not calculate diversity indices produced species accumulation curves to measure completeness of sampling.
+
 ### Pathogens
 
-X studies presented data on microorganisms that infect or are carried by small mammal species in West Africa.
-Data from a study previously excluded [@fichet-calvet_fluctuation_2007] due to duplicated data was incorporated as it provided the complementary pathogen data to the rodent trapping data.
-X of the studies were solely investigated pathogens of rodents.
-The remaining X studies investigated organisms that were potentially zoonotic pathogens, pathogens of rodents or microbial organisms that were of uncertain significance to rodent or human health.
+Sixty-two studies presented data on microorganisms that infect or are carried by small mammal species in West Africa.
+Data from a study previously excluded [@fichet-calvet_fluctuation_2007] due to duplicated data was incorporated as it provided the complementary pathogen data to the rodent trapping data of the more comprehensively reported study [@fichet-calvet_spatial_2005].
+Seven studies solely investigated pathogens of rodents, the remaining fifty-five studies investigated organisms that were potentially zoonotic pathogens, pathogens of rodents or microbial organisms that were of uncertain significance to rodent or human health.
 
-X studies used Polymerase Chain Reaction (PCR) to detect the presence of X different species or families of microorganisms.
-X studies used antibody or antigen based molecular tests to detect the presence of 9 different species or families of microorganisms.
-X studies conducted histological or direct visualisation assays of samples for X parasitic or bacterial species.
-X studies performed direct viral culture of *Lassa mammarenavirus* to detect the presence of the virus in rodent specimens.
+Thirty-two studies used Polymerase Chain Reaction (PCR) to detect the presence of 22 different species or families of microorganisms.
+Eleven studies used antibody or antigen based molecular tests to detect the presence of 9 different species or families of microorganisms.
+Eight studies conducted histological or direct visualisation assays of samples for 11 parasitic or bacterial species.
+Three studies performed direct culture of *Lassa mammarenavirus* or *Leishmania species* to detect the presence of the pathogen in rodent specimens.
 
-The rodent pathogens investigated were *Hydatigera species* (previously *Taenia species*) and *Trichuris species* both of which are gastro-intestinal helminths and *Leishmania major* which is an intracellular pathogen causing disease in both rodents and humans.
+The rodent pathogens included *Hydatigera species* (previously *Taenia species*) and *Trichuris species* both of which are gastro-intestinal helminths.
 
-X of the X studies reported prevalence of zoonotic pathogens at a level of detail beyond a study wide summary.
-These studies typically tested for a single pathogen (X), with X studies testing for two and three pathogens and X studies testing for five or more.
-The most frequently tested for pathogens were *Lassa mammarenavirus* (X) or members of the *Arenaviridae* family (X), the spirochete bacteria *Borrelia* was investigated in X studies, *Bartonella* and *Toxoplasma gondii* were investigated in X each, the remaining X pathogens were reported in two or fewer studies.
-
-X of the studies reporting data on pathogen prevalence had incomplete data on the geolocation from which some of the animals tested were from.
-X of these studies had no data on geolocation of the trapped rodents.
+Most studies tested for a single pathogen (39), with two studies testing for six pathogens, one for five pathogens, two for three pathogens and 11 for two pathogens.
+The most frequently tested for pathogens were *Lassa mammarenavirus* (28%) or members of the *Arenaviridae* family (16%), the spirochete bacteria *Borrelia* was investigated in 8 studies, *Bartonella* and *Toxoplasma gondii* were investigated in 4 each, the remaining 25 pathogens were reported in three or fewer studies.
 
@@
 
-#### Supplementary table 1
+#### Supplementary table 3
 
@@
 ``` {r sup_table, echo = F}
-DT::datatable(readRDS(here::here("tables", "sup_table1.rds")))
+DT::datatable(readRDS(here::here("tables", "sup_table3.rds")))
 ```

### 2021-04-26 scripts/pathogen_dictionary.R

@@ -18,7 +18,7 @@ eimeria <- c("eimeria_spp")
 names(eimeria) <- "eimeria_species"
 flavivirus <- c("flavivirus")
 names(flavivirus) <- "flavivirus_species"
-hantavirus <- c("hantavirus")
+hantavirus <- c("hantavirus", "hantavirus_spp")
 names(hantavirus) <- "hantavirus_species"
 hydatid <- "hydatigera_species"
 names(hydatid) <- "hydatigera_species"

### 2021-04-27 reports/manuscript.Rmd
results first draft

@@
 ```
+
+## Abstract
+
+### Background
+
+### Aims
+
+### Methods
+
+### Results
+
+### Conclusions
 
@@
 Multiple aims were identified from each study, the number on the arrow indicates the number of distinct studies with that aim.
-Opening the image separately will allow you to zoom in.", echo = F, layout = "l-page"}
+Opening the image separately will allow you to zoom in.", echo = F, layout = "l-screen-inset shaded"}
 knitr::include_graphics(here::here("figures", "study_aims.png"))
@@
 
-``` {r density, fig.cap = "Density of trap sites per thousand km^2^", echo = F, layout = "l-page"}
+``` {r density, fig.cap = "Density of trap sites per thousand km^2^", echo = F, layout="l-page"}
 knitr::include_graphics(here::here("figures", "static_site_density_2.png"))
@@
 Of the 147 distinct identified species trapped (see Supplementary table 3) the majority were from the order Rodentia (112) with Muridae (82) being the largest family of rodents, Sciuridae (10), followed by Gliridae and Nesomyidae (both 6), Ctenodactylidae and Anomaluridae (2) and Dipodidae, Hystricidae and Thryonomidae (all 1).
-The remaining 34 species came from the orders of the Soricomorpha (31), Erinaceomorpha (2) and Afrosoricida (1).
+The remaining 34 species came from the orders of the Soricomorpha (30), Erinaceomorpha (2) and Afrosoricida (1).
 
@@
 The most commonly trapped genera of rodents were *Mastomys sp.* (27,072, 38.5%), *Rattus sp.* (10,191, 14.5%), *Mus sp.* (8,624, 12.3%), *Arvicanthis sp.* (5,821, 8.3%) and *Praomys sp.* (5,409, 7.7%).
-At the species level *Mastomys natalensis* (11,221, 17.9%) and *Rattus rattus* (8,578, 13.7%) were the most commonly trapped rodents, *Mastomys erythroleucus* (8,578, 13.7%), *Mus musculus* (6,245, 8.8%), *Arvicanthis niloticus* (5,497, 8.8%) and *Mastomys huberti* (4,699, 7.5%) made up the next most commonly trapped species.
+At the species level *Mastomys natalensis* (11,221, 17.9%) and *Rattus rattus* (8,578, 13.7%) were the most commonly trapped rodents, *Mastomys erythroleucus* (7,379, 11.8%), *Mus musculus* (6,245, 10%), *Arvicanthis niloticus* (5,497, 8.8%) and *Mastomys huberti* (4,699, 7.5%) made up the next most commonly trapped species.
 Presence and absence maps for these six most commonly trapped species are shown in Figure \@ref(fig:dist).
 
-``` {r dist, fig.cap = "Presence and absence of the 6 most commonly trapped species in included studies, no adjustment made for trapping effort", echo = F, layout = "l-page"}
+``` {r dist, fig.cap = "Presence and absence of the 6 most commonly trapped species in included studies, no adjustment made for trapping effort", echo = F, layout="l-page"}
 knitr::include_graphics(here::here("figures", "top_6.png"))
@@
 Three studies performed direct culture of *Lassa mammarenavirus* or *Leishmania species* to detect the presence of the pathogen in rodent specimens.
+PCR was the most commonly used diagnostic assay being performed on 21,953 rodent samples, antibody or antigen based serological assays were performed on 11,430 samples, histological or direct visualisation was used on 11,229 samples with direct culture being used for 643 samples.
+
+The most common pathogens assessed for with PCR were *Lassa mammarenavirus* or other Arenaviridae (31%) the bacteria *Borrelia sp.* (11.3%) and *Bartonella sp.* (6.5%) followed by Usutu virus (6.4%) and Hantaviridae (5.6%).
+The most common pathogens assessed for with serology were *Lassa mammarenavirus* or other Arenaviridae (78%), *Toxoplasma gondii* (10.6%), *Borrelia sp.* (6.3%) and *Leptospirosis sp.* (2.2%).
+The most common pathogens assessed for with histology or direct visualisation were *Borrelia sp.* (48.7%), *Schistosoma sp.* (20.4%) and other parasites.
+All the studies using direct culture were assessing for either *Lassa mammarenavirus* (81%) or *Leishmania sp.* (19%).
 
@@
 Most studies tested for a single pathogen (39), with two studies testing for six pathogens, one for five pathogens, two for three pathogens and 11 for two pathogens.
-The most frequently tested for pathogens were *Lassa mammarenavirus* (28%) or members of the *Arenaviridae* family (16%), the spirochete bacteria *Borrelia* was investigated in 8 studies, *Bartonella* and *Toxoplasma gondii* were investigated in 4 each, the remaining 25 pathogens were reported in three or fewer studies.
+The most frequently tested for pathogens were *Lassa mammarenavirus* (28%) or members of the *Arenaviridae* family (16%), the spirochete bacteria *Borrelia sp.* was investigated in 8 studies, *Bartonella sp.* and *Toxoplasma gondii* were investigated in 4 each, the remaining 25 pathogens were reported in three or fewer studies.
 
-``` {r paths, fig.cap = "Presence and absence of the 4 most commonly tested for species in included studies, no adjustment made for trapping effort", echo = F, layout = "l-page"}
+``` {r paths, fig.cap = "Presence and absence of the 4 most commonly tested for zoonotic pathogens, no adjustment made for trapping or diagnostic effort", echo = F, layout="l-screen-inset shaded"}
 knitr::include_graphics(here::here("figures", "top_4.png"))
@@
 
-X rodent species were investigated for the presence of a zoonotic pathogen. X individual rodents were tested for at least one of the above pathogens.
+Ninety-seven rodent species were investigated for the presence of a zoonotic pathogen, 32,014 individual rodents were tested for at least one of the above pathogens, with 42,940 assays being performed.
+The rodent species most commonly assayed for zoonotic pathogens were also the most commonly trapped, *Rattus rattus* (n = 2,977) were assessed for 24 pathogens, *Mus musculus* (n = 3,402) for 23, *Mastomys natalensis* (n = 7,189) and *Mastomys erythroleucus* (n = 3,013) for 19 and *Arvicanthis niloticus* (n = 3,840) for 18.
+All remaining species were investigated for 10 or fewer pathogens.
 
-The rodents most commonly found to be hosts of zoonoses were...
+``` {r rodents-tested, fig.cap = "The number of individuals tested from the 19 monst commonly tested species, note that the x acis differs by genus", echo = F, layout="l-page"}
+knitr::include_graphics(here::here("figures", "rodent_testing.png"))
+```
 
-The species with the greatest number of detected pathogens within the studied areas were...
+*Lassa mammarenavirus* was detected in 414 *Mastomys natalensis*, 9 *Mastomys erythroleucus*, 5 *Hylomyscus pamfi* and 3 *Mus baoulei* through PCR (demonstrating acute infection), in serological assays, evidence of prior infection or acquisition of maternal antibodies was found in 465 *Mastomys natalensis* and 361 *Mastomys sp.* individuals alongside 6 *Mus musculus* and 4 *Rattus rattus*.
+From direct viral culture 19 *Mastomys natalensis*, 2 *Rattus rattus* and 1 *Mus minutoides* were found to be infected with *Lassa mammarenavirus*.
+*Bartonella sp.* were detected in 113 *Mastomys natalensis*, 50 *Mastomys erythroleucus*, 39 *Rattus norvegicus*, 16 *Rattus rattus* and 13 individuals of 7 further species through PCR.
 
-*Lassa mammarenavirus* was detected in X individuals either through PCR (demonstrating acute infection) or antibody studies (suggesting previous infection or the acquisition of maternal antibodies).
-These individuals were from X species (X *Mastomys natalensis*, Y *Praomys* ...).
+*Borrelia sp.* were identified via PCR in 110 individuals, 37 *Mastomys erythroleucus*, 12 *Mus musculus*, 11 *Masyomus huberti* and *Praomys daltoni*, the remaining 39 positive results were from 9 further species.
+In serological assays 47 *Mastomys natalensis*, 12 *Mastomys erythroleucus* and 8 *Praomys daltoni* were found to be positive, the remaining 11 positive individuals came from three further species.
+Most *Borrelia sp.* infections were identified through histology or direct visualisation with, 243 *Arvicanthis niloticus*, 102 *Mastomys natalensis*, and 78 *Mastomys huberti* along with a further 64 individuals from 8 species infected.
 
-*Borrelia sp.* were detected in X individuals ...
+Finally *Toxoplasma gondii* was detected via PCR in 89 *Mus musculus*, 13 *Cricetomys gambianus*, 5 *Crocidura olivieri* and 3 *Rattus rattus*.
+From serological assays evidence of infection was found in 25 *Mus musculus*, 46 *Mastomys erythroleucus*, 4 *Rattus rattus*, and *Mastomys natalensis*, using histology and direct visualisation 1 *Mus musculus* and 1 *Rattus rattus* were also found to be infected.
 
-X species of those tested were not found to be infected by any zoonotic pathogen investigated for.
+``` {r rodents-positive, fig.cap = "The proportion of individuals testing positive for the assessed pathogens from the 19 monst commonly tested species, note that the x acis differs by assay", echo = F, layout="l-page"}
+knitr::include_graphics(here::here("figures", "all_assays.png"))
+```
+
+Thirty eight species of Rodentia, 3 species of Soricomorpha and 2 species of Erinaceomorpha that were tested for potential pathogens had entirely negative results.
 

### 2021-04-27 scripts/2_2_clean_species.R

@@ -10,8 +10,17 @@ retain_columns <- c("unique_id", "year_trapping", "month_trapping", "country", "
 genus_synonym <-  as.list(c("praomys", "mus"))
 names(genus_synonym) <- c("myomys", "nannomys")
 
-accepted_names <- as.list(c("mus musculus", "oenomys hypoxanthus", "suncus megalura", "dasymys incomtus", "atelerix albiventris", "gerbilliscus gambiana", "gerbilliscus guinea", "gerbilliscus guinea", "gerbilliscus kempii", "dipodillus campestris", "gerbilliscus gambiana", "grammomys poensis", "graphiurus nagtglasii", "graphiurus kelleni", "paraechinus aethiopicus", "lophuromys flavopunctatus", "massoutiera mzabi", "mastomys natalensis", "steatomys caurinus", "gerbilliscus kempii", "gerbilliscus guineae", "gerbilliscus guineae", "gerbilliscus kempii", "gerbilliscus robusta", "taterillus gracilis", "grammomys poensis"))
-names(accepted_names) <- c("mus domesticus", "aethomys hypoxanthus", "crocidura megalura", "dasymys bentleyae", "erinaceus albiventris", "gerbilliscus gambianus", "tatera guineae", "gerbilliscus guineae", "gerbilliscus kempi", "gerbillus campestris", "gerbillus gambianus", "grammomys rutilans", "graphiurus hueti", "graphiurus parvus", "hemiechinus aethiopicus", "lophuromys flavipunctatus", "massouteria mzabi", "mastomys hildebrandtii", "steatomys caurianus", "tatera gambiana", "tatera guinea", "tatera guineae", "tatera kempi", "tatera robusta", "taterillus gracillis", "thamnomys rutilans")
+accepted_names <- as.list(c("mus musculus", "oenomys hypoxanthus", "suncus megalura", "dasymys incomtus", "atelerix albiventris", "gerbilliscus gambiana",
+                            "gerbilliscus guinea", "gerbilliscus guinea", "gerbilliscus kempii", "dipodillus campestris", "gerbilliscus gambiana",
+                            "grammomys poensis", "graphiurus nagtglasii", "graphiurus kelleni", "paraechinus aethiopicus", "lophuromys flavopunctatus",
+                            "massoutiera mzabi", "mastomys natalensis", "steatomys caurinus", "gerbilliscus kempii", "gerbilliscus guineae",
[…]

### 2021-04-27 scripts/2_4_clean_pathogen.R

@@ -85,6 +85,7 @@ dd_path <- pathogen_data %>%
 
 utm_path <- pathogen_data %>%
   drop_na(UTM_coordinates) %>%
+  mutate(UTM_coordinates = str_replace_all(UTM_coordinates, " ", "_")) %>%
   separate(col = UTM_coordinates, into = c("zone", "easting", "northing"), "_")
 
 utm_q_path <- utm_path %>%
@@ -115,4 +116,46 @@ path_all <- bind_rows(pathogen_gps, path_no_gps) %>%
                 culture_path_1_positive, culture_path_2_positive, culture_path_3_positive, histo_path_1_positive, histo_path_2_positive,
                 histo_path_3_positive, histo_path_4_positive, histo_path_5_positive, histo_path_6_positive, record_id)
 
+pathogen_tested <- c("path_1", "path_2", "path_3", "path_4", "path_5", "path_6")
+pcr_test <- c("pcr_path_1_positive", "pcr_path_2_positive", "pcr_path_3_positive", "pcr_path_4_positive", "pcr_path_5_positive", "pcr_path_6_positive")
+ab_ag_test <- c("ab_ag_path_1_positive", "ab_ag_path_2_positive", "ab_ag_path_3_positive", "ab_ag_path_4_positive", "ab_ag_path_5_positive")
+culture_test <- c("culture_path_1_positive", "culture_path_1_positive", "culture_path_1_positive")
+direct_visualisation <- c("histo_path_1_positive", "histo_path_2_positive", "histo_path_3_positive", "histo_path_4_positive", "histo_path_5_positive", "histo_path_6_positive")
+
+wide_pathogen <- path_all %>%
+  tibble() %>%
+  pivot_longer(cols = all_of(pathogen_tested), values_to = "pathogen") %>%
+  mutate(name = pathogen) %>%
+  drop_na(pathogen) %>%
+  mutate(row = row_number()) %>%
+  pivot_wider(values_from = c(path_1_tested, path_2_tested, path_3_tested, path_4_tested, path_5_tested, path_6_tested,
+                              pcr_path_1_positive, pcr_path_2_positive, pcr_path_3_positive, pcr_path_4_positive, pcr_path_5_positive, pcr_path_6_positive,
+                              ab_ag_path_1_positive, ab_ag_path_2_positive, ab_ag_path_3_positive, ab_ag_path_4_positive, ab_ag_path_5_positive,
[…]

### 2021-04-27 scripts/mapping_pathogens.R

@@ -23,6 +23,15 @@ level_2_all <- do.call(rbind.SpatialPolygonsDataFrame, level_2) %>%
 
 pathogen <- read_rds(here("data_clean", "pathogen.rds")) %>%
   filter(iso3c %in% wa_mainland)
+wide_pathogen <- read_rds(here("data_clean", "wide_pathogen.rds")) %>%
+  filter(iso3c %in% wa_mainland)
+
+species_data <- read_rds(here("data_clean", "species_data.rds")) %>%
+  distinct(genus, species, classification, gbif_id, genus_gbif, species_gbif)
+
+pathogen <- pathogen %>%
+  left_join(., species_data,
+            by = "gbif_id")
 
 pathogen_tested <- c("path_1", "path_2", "path_3", "path_4", "path_5", "path_6")
 pcr_test <- c("pcr_path_1_positive", "pcr_path_2_positive", "pcr_path_3_positive", "pcr_path_4_positive", "pcr_path_5_positive", "pcr_path_6_positive")
@@ -31,113 +40,217 @@ culture_test <- c("culture_path_1_positive", "culture_path_1_positive", "culture
 direct_visualisation <- c("histo_path_1_positive", "histo_path_2_positive", "histo_path_3_positive", "histo_path_4_positive", "histo_path_5_positive", "histo_path_6_positive")
 
 arenaviridae <- c("arenaviridae_species", "lassa_mammarenavirus", "mammarenavirus_species")
-
-arenavirus_map <- pathogen %>%
-  filter(path_1_tested != 0) %>%
-  filter(path_1 %in% arenaviridae | path_2 %in% arenaviridae | path_3 %in% arenaviridae | path_5 %in% arenaviridae) %>%
-  janitor::remove_empty("cols")  %>%
-  pivot_longer(cols = c("path_1", "path_2", "path_3"), values_to = "pathogen") %>%
-  drop_na(pathogen) %>%
-  filter(pathogen %in% arenaviridae) %>%
-  mutate(p_pos = if_else(name == "path_1",
-                         if_else(!is.na(pcr_path_1_positive), pcr_path_1_positive/path_1_tested*100,
-                                 ifelse(!is.na(ab_ag_path_1_positive), ab_ag_path_1_positive/path_1_tested*100,
-                                        ifelse(!is.na(culture_path_1_positive), culture_path_1_positive/path_1_tested*100,
[…]

### 2021-04-27 scripts/pathogen_dictionary.R

@@ -7,7 +7,7 @@ names(babesia) <- "babesia_species"
 bartonella <- c("bartonella", "bartonella_spp")
 names(bartonella) <- rep("bartonella_species", length(bartonella))
 borrelia <- c("borrelia", "borrelia_crocidurae", "borrelia_spp", "borrellia_spp")
-names(borrelia) <- rep("borrelia", length(borrelia))
+names(borrelia) <- rep("borrelia_species", length(borrelia))
 coxiella <- c("coxiella_burnetii")
 names(coxiella) <- "coxiella_burnetii"
 e_coli <- c("e_coli_esbl")
@@ -19,7 +19,7 @@ names(eimeria) <- "eimeria_species"
 flavivirus <- c("flavivirus")
 names(flavivirus) <- "flavivirus_species"
 hantavirus <- c("hantavirus", "hantavirus_spp")
-names(hantavirus) <- "hantavirus_species"
+names(hantavirus) <- rep("hantavirus_species", length(hantavirus))
 hydatid <- "hydatigera_species"
 names(hydatid) <- "hydatigera_species"
 k_pneumoniae <- "k_pneumoniae_esbl"
@@ -34,13 +34,13 @@ names(leptospirosis) <- "leptospirosis_species"
 mammarenavirus <- "mammarenavirus"
 names(mammarenavirus) <- "mammarenavirus_species"
 mycobacteria <- c("mycobacteria_spp", "mycobacterium_spp")
-names(mycobacteria) <- "mycobacterium_species"
+names(mycobacteria) <- rep("mycobacterium_species", length(mycobacteria))
 mycoplasma <- "mycoplasma_spp"
 names(mycoplasma) <- "mycoplasma_species"
 orentia <- "orentia"
 names(orentia) <- "orentia_species"
 orthopoxvirus <- c("orthopoxvirus", "orthopoxvirus_spp")
-names(orthopoxvirus) <- "orthopoxvirus_species"
+names(orthopoxvirus) <- rep("orthopoxvirus_species", length(orthopoxvirus))
 phlebovirus <- "phleboviruses"
 names(phlebovirus) <- "phlebovirus_species"
 plagiorchis <- "plagiorchis_species"

### 2021-04-28 reports/manuscript.Rmd
add draft abstract

@@
 
+Rodents represent around 40% of all mammalian species, they are important components of small mammal assemblages globally.
+In sub-Saharan Africa particularly they are significant contributors to agricultural crop loss in addition to being hosts for a number of zoonotic pathogens of importance to local human populations.
+Trapping of individual rodents, identification to species and assays for the presence of potential zoonotic pathogens provides a source of data for understanding the potential ecological and human health impact of changing small mammal populations.
+
 ### Aims
+
+This scoping review aimed to synthesise the research conducted in West Africa specifically to; a) identify the aims of rodent research and summarise methodological approaches b) identify the locations, habitats and rodent species that have been studied, c) identify what potential pathogens are tested for and which host species they are found in.
 
@@
 
+This was conducted as a scoping review of all rodent trapping studies identified through a systematic search of online databases for studies conducted within the United Nations West Africa subregion.
+
 ### Results
 
+4,282 records were identified from the initial search with 124 studies published between 1974 and 2021 included in narrative synthesis.
+The majority of studies investigated rodent ecology (56%) with the remainder investigating potential zoonotic diseases.
+Study methodology was comprehensively reported from 28 (23%) studies.
+Studies were conducted in 14 of 16 West African countries.
+Specific trapping habitats were not reported using standardised approaches.
+The included studies reported a combined 73,164 trapped small mammals, predominantly from the order Rodentia.
+These individuals were identified to 147 species with *Mastomys natalensis*, *Rattus rattus* and *Mastomys erythroleucus* the most commonly trapped species.
+55 studies investigated potential zoonotic pathogens, 7 further studies investigated rodent pathogens.
+Thirty-two microorganisms were tested for including viruses, bacteria and parasites.
+*Mastomys natalensis* was found to be most commonly infected by *Lassa mammarenavirus* and *Bartonella sp*, *Arvicanthis niloticus* was most commonly infected with *Borrelia sp*, *Mus musculus* was most commonly infected with *Toxoplasma gondii*.
+
 ### Conclusions
+
+This scoping review identified 124 studies, conducted in 14 countries at 1,193 study sites trapping 73,164 small rodents.
+Current research efforts to describe the rodent populations of West Africa has provided some information about the distribution of species across this heterogenous region and the distribution of potential pathogens within individual species.
+We make some recommendations for this area of research to support data sharing and re-use to better understand dynamic changes within these complex systems as a result of alterations to local habitats or climate.
 
@@
 Seven studies solely investigated pathogens of rodents, the remaining fifty-five studies investigated organisms that were potentially zoonotic pathogens, pathogens of rodents or microbial organisms that were of uncertain significance to rodent or human health.
+Thirty-two microorganisms were tested for, 8 of these at the species level, with the remaining 24 at levels higher than species.
 
@@
 
-``` {r rodents-tested, fig.cap = "The number of individuals tested from the 19 monst commonly tested species, note that the x acis differs by genus", echo = F, layout="l-page"}
+``` {r rodents-tested, fig.cap = "The number of individuals tested from the 19 monst commonly tested species, note that the x axis differs by genus", echo = F, layout="l-page"}
 knitr::include_graphics(here::here("figures", "rodent_testing.png"))
@@
 
-*Borrelia sp.* were identified via PCR in 110 individuals, 37 *Mastomys erythroleucus*, 12 *Mus musculus*, 11 *Masyomus huberti* and *Praomys daltoni*, the remaining 39 positive results were from 9 further species.
+*Borrelia sp.* were identified via PCR in 110 individuals, 37 *Mastomys erythroleucus*, 12 *Mus musculus*, 11 *Mastomys huberti* and *Praomys daltoni*, the remaining 39 positive results were from 9 further species.
 In serological assays 47 *Mastomys natalensis*, 12 *Mastomys erythroleucus* and 8 *Praomys daltoni* were found to be positive, the remaining 11 positive individuals came from three further species.
@@
 
-``` {r rodents-positive, fig.cap = "The proportion of individuals testing positive for the assessed pathogens from the 19 monst commonly tested species, note that the x acis differs by assay", echo = F, layout="l-page"}
+``` {r rodents-positive, fig.cap = "The proportion of individuals testing positive for the assessed pathogens from the 19 most commonly tested species, note that the x axis differs by assay", echo = F, layout="l-page"}
 knitr::include_graphics(here::here("figures", "all_assays.png"))

### 2021-04-28 reports/manuscript.Rmd
add toc

@@
     toc: true
+    toc_depth: 3
+    toc_float: true
 knit: (function(inputFile, encoding) {

### 2021-04-28 scripts/descriptive.R

@@ -86,6 +86,15 @@ rodent_data %>%
   summarise(n = n()) %$%
   table(n)
 
+
+# Study methodology -------------------------------------------------------
+
+studies %>%
+  filter(trap_types != "not_stated" & geolocation_level %in% c("study_site", "trap_site") & trapping_effort == "Yes") %>%
+  nrow()
+
+table(studies$pathogen)
+
 # Trap type and setup -----------------------------------------------------
 trap_type <- studies %>%
   separate(col = trap_types, into = c("trap_1", "trap_2", "trap_3", "trap_4"), sep = ", ", remove = T) %>%
@@ -398,6 +407,13 @@ long_pathogen %>%
   assign(x = "tested", value = ., pos = 1) %>%
   summarise(n_pathogen_tested = n()) %T>%
   arrange(-n_pathogen_tested) # The number of distinct pathogens tested for by rodent species
+
+long_pathogen %>%
+  filter(str_detect(assay, regex("tested"))) %>%
+  group_by(pathogen_tested) %>%
+  summarise(n_pathogen_tested = sum(number)) %>%
+  arrange(-n_pathogen_tested) %>% # The pathogens tested for
+  print(n = 32)
 
 long_pathogen %>%
   filter(str_detect(assay, regex("path_1_tested"))) %>%

### 2021-05-11 reports/manuscript.Rmd
draft manuscript

@@
 knitr::opts_chunk$set(echo = TRUE)
+source(here::here("scripts", "manuscript_library.r"))
 ```
@@
 
-Rodents represent around 40% of all mammalian species, they are important components of small mammal assemblages globally.
-In sub-Saharan Africa particularly they are significant contributors to agricultural crop loss in addition to being hosts for a number of zoonotic pathogens of importance to local human populations.
-Trapping of individual rodents, identification to species and assays for the presence of potential zoonotic pathogens provides a source of data for understanding the potential ecological and human health impact of changing small mammal populations.
+Rodents represent an estimated 40% of mammalian species and are important components of small mammal assemblages globally.
+In sub-Saharan Africa they are significant contributors to agricultural crop loss alongside being hosts for a number of zoonotic pathogens of importance to local human populations.
+Trapping of individual rodents, identification to species and assays for the presence of potential zoonotic pathogens produces important data for understanding the potential ecological and human health impact of changing small mammal populations.
 
@@
 
-This was conducted as a scoping review of all rodent trapping studies identified through a systematic search of online databases for studies conducted within the United Nations West Africa subregion.
+A scoping review of all rodent trapping studies identified through a systematic search of online databases for studies conducted within the United Nations West Africa sub-region was performed.
 
@@
 4,282 records were identified from the initial search with 124 studies published between 1974 and 2021 included in narrative synthesis.
+Studies were conducted in 14 West African countries.
 The majority of studies investigated rodent ecology (56%) with the remainder investigating potential zoonotic diseases.
 Study methodology was comprehensively reported from 28 (23%) studies.
-Studies were conducted in 14 of 16 West African countries.
 Specific trapping habitats were not reported using standardised approaches.
-The included studies reported a combined 73,164 trapped small mammals, predominantly from the order Rodentia.
-These individuals were identified to 147 species with *Mastomys natalensis*, *Rattus rattus* and *Mastomys erythroleucus* the most commonly trapped species.
+The included studies reported 73,164 trapped small mammals, predominantly from the order Rodentia.
+These individuals were identified to 147 species with *Mastomys natalensis*, *Rattus rattus* and *Mastomys erythroleucus* the most common.
 55 studies investigated potential zoonotic pathogens, 7 further studies investigated rodent pathogens.
 Thirty-two microorganisms were tested for including viruses, bacteria and parasites.
-*Mastomys natalensis* was found to be most commonly infected by *Lassa mammarenavirus* and *Bartonella sp*, *Arvicanthis niloticus* was most commonly infected with *Borrelia sp*, *Mus musculus* was most commonly infected with *Toxoplasma gondii*.
+*Mastomys natalensis* was found to be most commonly infected by *Lassa mammarenavirus* and *Bartonella sp*, *Arvicanthis niloticus* was most commonly infected with *Borrelia sp* and *Mus musculus* was most commonly infected with *Toxoplasma gondii*.
 
@@
 This scoping review identified 124 studies, conducted in 14 countries at 1,193 study sites trapping 73,164 small rodents.
-Current research efforts to describe the rodent populations of West Africa has provided some information about the distribution of species across this heterogenous region and the distribution of potential pathogens within individual species.
-We make some recommendations for this area of research to support data sharing and re-use to better understand dynamic changes within these complex systems as a result of alterations to local habitats or climate.
+Current research to describe the rodent populations of West Africa have provided some information about the distribution of species across this heterogenous region and the distribution of potential pathogens within individual species.
+We make recommendations for subsequent research to enable wider data sharing and re-use to to support meta-analysis of obtained data to address questions on rodent ecology and zoonotic disease risk.
 
@@
 
-Beyond crop loss a further threat to local communities is the transmission of infectious diseases from rodents to humans.
+Alongside crop loss a further risk to local communities is the transmission of infectious diseases from rodents to humans.
 Zoonotic diseases (i.e.
@@
 To date there has been no comprehensive overview of rodent trapping studies conducted in West Africa.
-The aim of this scoping review is therefore to synthesise the knowledge on rodent distributions throughout West Africa obtained from trapping studies.
+The aim of this scoping review is therefore to systematically map the research conducted in this area, to identify currently used methods, gaps in current knowledge and to synthesise the knowledge on rodent distributions throughout West Africa obtained from included trapping studies.
 
@@
 
-  +	1. In which West African countries or regions have rodent trapping studies been performed?
-  + 2. What are the stated aims of these rodent trapping studies?
-  + 3. Which rodent species are being targeted for trapping activities? Are species with high potential for hosting zoonotic pathogens targeted at higher rates?
-  + 4. How is the type of land use and/or land use intensity classified in the literature and are classifications comparable across studies?
-  +	5. Is rodent trapping occurring in all habitat types or are some habitats targeted at higher rates?
+  1. In which West African countries or regions have rodent trapping studies been performed?
+  2. What are the stated aims of these rodent trapping studies?
+  3. Which rodent species are being targeted for trapping activities? Are species with high potential for hosting zoonotic pathogens targeted at higher rates?
+  4. How is the type of land use and/or land use intensity classified in the literature and are classifications comparable across studies?
+  5. Is rodent trapping occurring in all habitat types or are some habitats targeted at higher rates?
 
@@
 
-A scoping review approach was chosen to support a more descriptive and exploratory approach than using a systematic review.
-
-  + Scoping review
-  + Protocol pre-registered on the [Openscience Framework](...)
+A scoping review approach was used to identify and map the available information on rodent assemblages, abundance and diversity across West Africa.
+This approach is recommended when examining how research is conducted within a topic area [@munn_systematic_2018].
+The review followed the recommended appreach from the PRISMA extension for Scoping Review guidance [@tricco_prisma_2018].
+The protocol for this review was not pre-registered.
 
@@
     + Recorded as outcome genus or species of trapped individual
-    + Were available in English
     + Published in a peer-reviewed journal, a pre-print manuscript or a report by a an identifiable organisation/consultancy
@@
     + No full text were available
+    + Not available in English
 
@@
 
-The following terms were searched for in OVID Medline, Web of Science (Core collection and Zoological Record), JSTOR, BioOne, African Journals Online, Global Health and the pre-print servers BioRxiv and EcoEvoRxiv:
+The following terms as keywords were searched for in OVID Medline, Web of Science (Core collection and Zoological Record), JSTOR, BioOne, African Journals Online, Global Health and the pre-print servers BioRxiv and EcoEvoRxiv:
 
@@
 
-Additional articles were identified through references within included articles and reports known to the study team. Searches were completed on 2021-03-01.
+Additional articles were identified through references within included articles and reports known to the study team.
+No time constraints were placed on searches.
+Searches were completed on 2021-03-01.
+The exploded search terms and links to the respective web portals are provided in supplementary material.
 
@@
 
-One reviewer screened titles, abstracts and full texts against the inclusion criteria. A random subset of each of these (10%) were reviewed by a second reviewer.
+One reviewer screened titles, abstracts and full texts against the inclusion and exclusion criteria. A random subset of each of these (10%) were reviewed by a second reviewer.
 
@@
 
-Data were extracted by one reviewer.
-A second reviewer verified a subset (10%) of included studies.
+Data from eligible studies were extracted using a standardised data abstraction tool designed for this study on a subset of eligible studies and subsequently refined.
+Data were extracted by a single reviewer.
+A second reviewer verified a random subset (10%) of included studies.
 A data extraction form was maintained on a Google sheets document.
 Data extracted included i) study identifiers; ii) study aims; iii) trapping methodology; iv) geolocation data; v) method of speciation; vi) trapping locations and dates; vii) trapped species; viii) number of trap-nights and ix) microorganisms of interest.
+The data extraction forms and data dictionaries used during the analysis are reproduced in the supplementary material.
 
@@
 
-Where explicitly stated the study aims were extracted.
-For those with no recorded aims the specific aims were inferred from the introduction or conclusions of manuscript or report.
-The aims of the studies were categorised by the research team with studies subsequently broadly categorised as pertaining to rodent ecology research or studies on risk of zoonoses.
+When explicitly stated the study aims were extracted, for those with no specified aims the aims were inferred from the introduction or conclusions of the manuscript or report.
+The aims of studies were categorised with a subsequent higher order allocation to rodent ecology research or studies on risk of zoonoses.
 
@@
 Coordinates were extracted in the format reported and converted to decimal degrees.
-Where no GPS location is reported coordinates will be matched to study locations using the National Geospatial Intelligence Agency NGA GEOnet Names Server [@national_geospatial-intelligence_agency_nga_nodate] based on study site name/region and through maps presented in the manuscript.
+Where no GPS location was reported coordinates were matched to study locations using the National Geospatial Intelligence Agency NGA GEOnet Names Server [@national_geospatial-intelligence_agency_nga_nodate] based on study site name/region and comparison to maps presented in the manuscript (when available).
 
@@
 For studies using multiple devices all types were recorded.
-The method employed when setting the traps was extracted where reported.
-Studies that reported setting a transect line were classified as such, for studies reporting using a line of traps but not through a specified habitat the term trap line was used.
+The method employed when setting the traps was also extracted, studies that reported setting a transect line were classified as such, for studies reporting using a line of traps but not through a specified habitat the term trap line was used.
 
-The trapping effort in trap-nights was recorded.
-A trap-night is the setting of a single trap for a single night.
+The trapping effort in trap-nights (the setting of a single trap for a single night) was recorded.
 If this was reported at the habitat level or the level at which the study reported rodent captures it was classified as a complete recording of trapping effort.
@@
 
+#### Genus/species' trapped
+
+For studies reporting on all trapped species (i.e.
+not those only reporting on specific species of interest) that do not report absence at study sites all species trapped within the study will be recorded at all study sites with explicitly zero observations.
+
 #### Zoonoses testing
 
-In studies investigating rodents for potential zoonoses the pathogens tested for were extracted.
-For studies reporting multiple potential pathogens all of the organisms were extracted.
-For studies solely reporting on pathogens of rodents (i.e.
-rodent parasites) pathogens were grouped as "Rodent pathogens".
-The number of rodents tested and the number of positive or negative samples were extracted alongisde the type of assay used to detect the pathogens (e.g.
+In studies investigating rodents for potential zoonoses all pathogens tested for were extracted.
+The number of rodents tested and the number of positive or negative samples were extracted alongside the type of assay used to detect the pathogens (e.g.
 Polymerase Chain Reaction (PCR), Enzyme Linked ImmunoSorbent Assay (ELISA) or viral culture).
-Where possible pathogens were identified to the species level, however, where an assay only allows for attribution to a family of viruses or bacteria the higher order grouping was used.
+Where possible pathogens were identified to the species level, however, where an assay only allows for attribution to a family of viruses or bacteria the higher order grouping was used (i.e.
+PCR using a non-specific arenavirus primer).
+
+#### Risk of bias
+
+No formal risk of bias assessment tool was used.
 
@@
 A total of 4,282 records were identified, with 124 studies included in narrative synthesis (see Figure \@ref(fig:prisma)).
-A table of the included studies and their DoIs is presented in Supplementary table 1.
+A summary table of the included studies and the citations of the study or report are presented in supplementary table 1.
 The earliest studies identified were from 1974 with the majority (61%) published since 2010.
@@
 
+Table 1 summarises the included studies by publication year, and methodology with grouping by study aim.
+
+``` {r table-1, echo=F}
+studies <- read_rds(here("data_clean", "studies.rds"))
+
+studies %>%
+  mutate(decade_publication = case_when(year_publication < 1980 ~ "1970-79",
+                                        year_publication < 1990 ~ "1980-89",
+                                        year_publication < 2000 ~ "1990-99",
+                                        year_publication < 2010 ~ "2000-09",
+                                        year_publication < 2020 ~ "2010-19",
+                                        year_publication < 2030 ~ "2020-",
+                                        TRUE ~ "other"),
+         geolocation_level = case_when(geolocation_level == "island" ~ "Study site",
+                                       geolocation_level == "none" ~ "No geolocation",
+                                       geolocation_level == "study" ~ "Study",
+                                       geolocation_level == "study_region" ~ "Study",
+                                       geolocation_level == "study_site" ~ "Study site",
+                                       geolocation_level == "trap_line" ~ "Specific trap site",
+                                       geolocation_level == "trap_site" ~ "Specific trap site",
+                                       geolocation_level == "trap_site, study_site" ~ "Specific trap site",
+                                       TRUE ~ "other"),
+         speciation = case_when(speciation == "molecular" ~ "Molecular",
+                                speciation == "morphological" ~ "Morphological",
+                                speciation == "morphological, molecular" ~ "Morphological and molecular",
+                                speciation == "not_stated" ~ "Not stated",
+                                TRUE ~ "other"),
+         pathogen = case_when(pathogen == "Yes, second paper" ~ "Yes",
+                               TRUE ~ pathogen),
+         country = sub('(^[^_]+_[^_]+)_(.*)$', '\\2', unique_id),
+         country = case_when(country == "multiple" ~ "Multiple",
+                             TRUE ~ countryname(sub("_2", "", country), destination = "un.name.en"))) %>%
+  dplyr::select(country, decade_publication, aim, repeated_visit, geolocation_level,
+                speciation, species_accumulation, diversity_measurement, trapping_effort, pathogen) %>%
+  tbl_summary(by = aim,
+              label = c(country ~ "Country",
+                        decade_publication ~ "Publication year",
+                        repeated_visit ~ "Repeat study visits - Yes",
+                        geolocation_level ~ "Level of geolocation",
+                        speciation ~ "Method of speciation",
+                        species_accumulation ~ "Use of a species accumulation curve - Yes",
+                        diversity_measurement ~ "Use of a measure of species diversity - Yes",
+                        trapping_effort ~ "Reporting of trapping effort",
+                        pathogen ~ "Reporting of potential pathogens")) %>%
+  add_overall() %>%
+  as_gt() %>%
+  tab_header(title = "Table 1: Characteristics of included studies summarised by study aim") %>%
+  tab_options(table.font.size = px(12),
+              data_row.padding = px(2),
+              container.height = 600) %>%
+  fmt_missing(columns = everything(), missing_text = " ")
+```
+
 ### Aims of included studies
@@
 Biodiversity of rodents was a shared aim between the two themes with 37 ecological studies and 2 zoonoses studies investigating this.
-The remaining aims were more disparate with ecological studies investigating rodent population dynamics, rodent species assemblages, the genetics of rodents and their behaviours.
-Most zoonotic disease research focussed on prevalence of pathogens in rodent populations, the risk of spillover to human populations, the identity of pathogen reservoir species with fewer studies investigating the phylogenetics of pathogen strains.
 
 ``` {r aims, fig.cap = "The aims of included studies following division into ecological and zoonotic infectious disease themes.
-Multiple aims were identified from each study, the number on the arrow indicates the number of distinct studies with that aim.
-Opening the image separately will allow you to zoom in.", echo = F, layout = "l-screen-inset shaded"}
+Multiple aims were identified from each study, the number on the arrow indicates the number of distinct studies with that aim.", echo = F, layout = "l-screen-inset shaded"}
 knitr::include_graphics(here::here("figures", "study_aims.png"))
@@
 
-Trapped occurred in 17 countries, however, three countries were outside of the scope of this study (Cameroon and Chad in Central Africa and Morocco in Northern Africa).
-At least a single trap site was recorded from 14 West African countries, with no rodent trapping reported from Gambia or Togo.
+Rodent trapping occurred in 17 countries, three countries were outside of the scope of this study (Cameroon and Chad in Central Africa and Morocco in Northern Africa).
+At least a single trap site was recorded from 14 West African countries.
+No rodent trapping was identified from Gambia or Togo.
 
-Most studies (93%) were conducted in a single country, five studies were conducted in two countries, three studies were conducted in five countries and one study was conducted across six countries.
+Most studies (92%) were conducted in a single country, six studies were conducted in two countries, three studies were conducted in five countries and one study was conducted across six countries.
 In summary 146 trapping activities were reported across the 14 West African countries.
-More trapping activity was reported from Senegal (19%) than any other country.
-Nigeria (13%), Ghana and Guinea (both 12%) were the next most frequently trapped countries, followed by Mali and Benin (both 8%), Niger and Sierra Leone (both 6%), Mauritania and the Ivory Coast (4%), Burkina Faso (3%), Liberia (2%) with Guinea-Bissau and Cape Verde recording a single trapping activity.
 
-The trapping activity from the included studies was spread across 1,193 study sites.
+The trapping activity from the included studies was performed at 1,193 study sites.
 Thirty-one (25%) studies reported trapping only at a single study site, 45 (37%) studies trapped at between two and five study sites, the remaining 48 studies trapped at between six and 93 study sites.
 Trap sites were mapped to level two country administrative regions and density of sites per thousand square kilometers was calculated (see Figure \@ref(fig:density)).
-The areas with highest trapping densities included the capital cities of Niger, Sierra Leone, Senegal, Benin, Mali, Ghana and the Ivory Coast.
-Of the areas not incorporating capital cities Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone were the most intensively studied regions.
+The areas with highest trapping densities included the capital cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako) and Ghana (Accra) and the largest cities of Ivory Coast (Abidjan) and Benin (Cotonou).
+Of the areas not incorporating these cities; Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone were the most intensively studied regions.
 
@@
 
-Forty studies did not describe the structure of the trap setup within the defined study area.
-Twenty seven studies placed transect lines of traps through study sites with nineteen studies placing traps along a line but not clearly describing a transect across a defined site.
-Twenty-two studies purposefully sampled houses, with nine further studies purposefully placing traps in targeted locations (i.e.
+Forty studies did not describe the trap setup within a defined study area.
+Twenty seven studies placed transect lines of traps through study sites with nineteen studies placing traps along a line but not clearly describing a transect across a defined habitat.
+Twenty-two studies purposefully sampled domestic structures (i.e.
+homes), with nine further studies purposefully placing traps in targeted locations (i.e.
 near rivers, in fields, by houses).
@@
 Trap setup was commonly reported in ecology studies (81%) with transect lines the most commonly used approach.
-Trap setup was reported in 56% of zoonoses risk studies with purposeful placement of traps in houses the most common approach (24%).
+Trap setup was reported in 56% of zoonoses risk studies with purposeful placement of traps in houses the most common method (24%).
 
@@
 
-Trapping effort was variably reported across included studies.
 Thirty-nine studies did not report any measures of trapping effort.
-Forty-two studies reported a measure of trapping effort at the same level of detail at which they reported rodent captures  (the unit of analysis) and were classified as complete reports.
-Thirty-nine studies had incomplete reports on trapping effort due to not presenting the number of trap nights at the appropriate unit of analysis (i.e.
+Forty-two studies reported a measure of trapping effort at the same level of detail at which they reported rodent captures  (the unit of analysis) and were classed as complete reports.
+Thirty-nine studies were classed as incomplete reports on trapping effort due to not presenting the number of trap nights at the same scale as the unit of analysis (i.e.
 number of study nights only or total number of trap-nights).
@@
 For studies reporting the number of trap-nights at the unit of analysis (typically study site or habitat type) the median number of trap-nights at each site was 360 (IQR = 246-793).
-In studies with incomplete reports on trap-nights there was a median of 7,200 trap-nights (IQR = 1,184-9,424) conducted, for 26 studies reporting only on the number of days or nights trapping was performed on activities typically lasted between 1-4 nights.
+In studies with incomplete reports on trap-nights there was a median of 7,200 trap-nights (IQR = 1,184-9,424) conducted.
+For 26 studies reporting only on the number of days or nights trapping was performed the trapping activity typically lasted between 1-4 nights.
 
@@
 No studies reported trap habitats with direct reference to a habitat classification scheme.
-Thirteen studies did not describe the local habitat where traps were placed (4 ecology, 9 zoonoses).
-The remaining studies used more than 170 descriptions of the habitat type in which trapping was conducted, further, the resolution of the description varied between studies with some reporting the habitat of the location of the trap and others the study site or region.
-The reported habitat types were combined into 30 categories (see Supplementary table 2 for the habitat dictionary) for further analysis.
+Thirteen studies did not describe the local habitat in which traps were placed (4 ecology, 9 zoonoses).
+The remaining studies used more than 170 descriptions of the habitat type in which trapping was conducted, further, the scale of habitat of the description varied.
+The extracted habitat types were grouped into 30 categories (see Supplementary table 2 for the habitat dictionary) for further analysis.
 At least one habitat was recorded for 17,122 trap sites (95%), with two or more habitats for a single trap site recorded for 4,403 (24%) sites.
@@
 
-Habitat type was more comprehensively reported from ecology studies, 30 habitat categories were reported from ecology studies with the most commonly investigated habitats being agriculture (23%), within forests (18%) and in or around buildings (16%).
-Among studies on zoonoses 15 habitat categories were reported, the most commonly reported classification was described as the rodents "natural habitat" (39%), within or around buildings (38%) and in agricultural settings (7%).
+Habitat type was more comprehensively reported from ecology studie and were most commonly agricultural (23%), within forests (18%) and in or around buildings (16%).
+The most commonly reported habitats from zoonoses studies were the rodents "natural habitat" (39%), within or around buildings (38%) and in agricultural settings (7%).
 
@@
 Nine studies described molecular identification of rodents with 30% using a combination of morphological and molecular techniques.
-In more recently published studies since 2010, 76% used molecular techniques to assist with speciation, compared with less than 15% in studies published prior to 2010.
+In more recently published studies since 2010, 76% used molecular techniques to aid speciation, compared with less than 15% in studies published prior to 2010.
 
@@
 
-Twenty-six studies reported a numeric measure of rodent species diversity or richness, these included the Shannon Index, Simpson's Index and Chao Index.
+Twenty-six studies reported a numeric measure of rodent species diversity or richness, the indices used included the Shannon Index, Simpson's Index and Chao Index.
 Six of these studies present a species accumulation curve to estimate the completeness of their sampling activity.
-A further 8 studies that did not calculate diversity indices produced species accumulation curves to measure completeness of sampling.
+A further 8 studies that did not present diversity indices produced species accumulation curves to measure completeness of sampling.
 
@@
 Thirty-two microorganisms were tested for, 8 of these at the species level, with the remaining 24 at levels higher than species.
+
+``` {r pathogen-table, echo=F}
+readRDS(here::here("tables", "pathogen_table.rds")) %>%
+  ungroup() %>%
+  gt::gt(rowname_col = "Pathogen tested") %>%
+  gt::tab_header(title = "Table 2: The number of studies testing for each microorganism or group of microorganisms by assay type") %>%
+  tab_options(table.font.size = px(12),
+              data_row.padding = px(2)) %>%
+  gt::fmt_missing(columns = everything(), missing_text = " ")
+```
 
@@
 The most common pathogens assessed for with histology or direct visualisation were *Borrelia sp.* (48.7%), *Schistosoma sp.* (20.4%) and other parasites.
-All the studies using direct culture were assessing for either *Lassa mammarenavirus* (81%) or *Leishmania sp.* (19%).
+All  studies using direct culture were investigating either *Lassa mammarenavirus* (81%) or *Leishmania sp.* (19%).
 
@@
 
-Most studies tested for a single pathogen (39), with two studies testing for six pathogens, one for five pathogens, two for three pathogens and 11 for two pathogens.
+Most studies tested for a single pathogen (39), with 16 studies testing for two or more pathogens.
 The most frequently tested for pathogens were *Lassa mammarenavirus* (28%) or members of the *Arenaviridae* family (16%), the spirochete bacteria *Borrelia sp.* was investigated in 8 studies, *Bartonella sp.* and *Toxoplasma gondii* were investigated in 4 each, the remaining 25 pathogens were reported in three or fewer studies.
@@
 
-*Lassa mammarenavirus* was detected in 414 *Mastomys natalensis*, 9 *Mastomys erythroleucus*, 5 *Hylomyscus pamfi* and 3 *Mus baoulei* through PCR (demonstrating acute infection), in serological assays, evidence of prior infection or acquisition of maternal antibodies was found in 465 *Mastomys natalensis* and 361 *Mastomys sp.* individuals alongside 6 *Mus musculus* and 4 *Rattus rattus*.
+*Lassa mammarenavirus* was detected in 414 *Mastomys natalensis*, 9 *Mastomys erythroleucus*, 5 *Hylomyscus pamfi* and 3 *Mus baoulei* through PCR (demonstrating acute infection).
+In serological assays, evidence of prior infection or acquisition of maternal antibodies to *Lassa mammarenavirus* was found in; 465 *Mastomys natalensis*, 361 *Mastomys sp.* and 6 *Mus musculus* and 4 *Rattus rattus*.
 From direct viral culture 19 *Mastomys natalensis*, 2 *Rattus rattus* and 1 *Mus minutoides* were found to be infected with *Lassa mammarenavirus*.
-*Bartonella sp.* were detected in 113 *Mastomys natalensis*, 50 *Mastomys erythroleucus*, 39 *Rattus norvegicus*, 16 *Rattus rattus* and 13 individuals of 7 further species through PCR.
+*Bartonella sp.* were detected in; 113 *Mastomys natalensis*, 50 *Mastomys erythroleucus*, 39 *Rattus norvegicus*, 16 *Rattus rattus* and 13 individuals of 7 further species through PCR.
 
-*Borrelia sp.* were identified via PCR in 110 individuals, 37 *Mastomys erythroleucus*, 12 *Mus musculus*, 11 *Mastomys huberti* and *Praomys daltoni*, the remaining 39 positive results were from 9 further species.
-In serological assays 47 *Mastomys natalensis*, 12 *Mastomys erythroleucus* and 8 *Praomys daltoni* were found to be positive, the remaining 11 positive individuals came from three further species.
+*Borrelia sp.* infections were identified through PCR in 110 individuals; 37 *Mastomys erythroleucus*, 12 *Mus musculus*, 11 *Mastomys huberti* and *Praomys daltoni*, the remaining 39 positive results were from 9 further species.
+In serological assays; 47 *Mastomys natalensis*, 12 *Mastomys erythroleucus* and 8 *Praomys daltoni* were found to be positive, the remaining 11 positive individuals came from three further species.
 Most *Borrelia sp.* infections were identified through histology or direct visualisation with, 243 *Arvicanthis niloticus*, 102 *Mastomys natalensis*, and 78 *Mastomys huberti* along with a further 64 individuals from 8 species infected.
 
-Finally *Toxoplasma gondii* was detected via PCR in 89 *Mus musculus*, 13 *Cricetomys gambianus*, 5 *Crocidura olivieri* and 3 *Rattus rattus*.
+Finally *Toxoplasma gondii* was detected via PCR in; 89 *Mus musculus*, 13 *Cricetomys gambianus*, 5 *Crocidura olivieri* and 3 *Rattus rattus*.
 From serological assays evidence of infection was found in 25 *Mus musculus*, 46 *Mastomys erythroleucus*, 4 *Rattus rattus*, and *Mastomys natalensis*, using histology and direct visualisation 1 *Mus musculus* and 1 *Rattus rattus* were also found to be infected.
@@
 This scoping review aimed to consolidate and summarise research conducted on rodents and their potential zoonoses in West Africa.
-A comprehensive search identified X studies performed in Y West African countries.
+A comprehensive search identified 124 studies performed in 14 West African countries.
 Around half of these studies were conducted to understand the ecology of rodents with the others being conducted to understand the risk of zoonotic pathogens to human populations in West Africa.
 
-Rodent trappping has been performed in most West African countries since the first recorded study from 197x, however, the distribution of effort between the countries and regions within those countries has been highly skewed.
+Rodent trappping has been performed in most West African countries since the first recorded study from 1974, however, the distribution of effort between the countries and regions within those countries has been highly skewed.
 More than 50% of trapping activity has occurred in Senegal, Nigeria, Ghana and Guinea.
-Capital cities and their surrounding regions were more commonly studied than more sparsely populated areas.
+Large cities and their surrounding regions were more commonly studied than less densely populated areas.
 
-The studies conducted to assess rodent ecology typically focussed on investigating rodent biodiveristy or conducting small mammal surveys of national parks or areas of specific scientific interest.
-Other studies investigated rodnt population dynamics using mark and release methods to measure rodent recruitment (birth rates) and survival.
+The studies conducted to assess rodent ecology typically focussed on rodent biodiversity or conducting small mammal surveys of national parks or areas of specific interest such as those about to be exploited for mining dam building.
+Other studies investigated rodent population dynamics using mark and release methods to measure rodent recruitment (birth rates) and survival.
+The majority of zoonoses studies were conducted to determine the prevalence of zonoses in the rodent population and to identify the potential reservoirs of pathogens.
 
-  + 2. What are the stated aims of these rodent trapping studies?
-  + 3. Which rodent species are being targeted for trapping activities? Are species with high potential for hosting zoonotic pathogens targeted at higher rates?
-  + 4. How is the type of land use and/or land use intensity classified in the literature and are classifications comparable across studies?
-  +	5. Is rodent trapping occurring in all habitat types or are some habitats targeted at higher rates?
+The most commonly trapped species are likely those of most direct importance to human populations in West Africa.
+Of the 62,574 individuals identified to species level the *Mastomys* species' and *Arvicanthis niloticus* are commensal 'native' rodents with *Mus musculus* and *Rattus sp.* commensal 'introduced' species that are important pest species of human agriculture in addition to being potential vectors of zoonotic pathogens.
+A further concern is that these species that thrive in human dominated landscapes will continue to expand in their habitable range and so continue to exert pressure on the distribution of other rodent species.
+The *Mastomys* species alongside *Mus musculus* and *Rattus rattus* had some evidence of infection with the four most studied pathogens (Arenaviruses, Bartonellae, Borrelia and Toxoplasma) this compares with *Arvicanthis niloticus* which despite being as widely tested was only found to harbour Borrelia in Mali.
+
+From the data obtained for this scoping review it would appear that rodents that have the potential to carry zoonotic pathogens or species that are significant agricultural pests are studied with greater intensity.
+
+Land use has not been consistently classified either within ecological studies or zoonoses studies which limits the ability to synthesise data across multiple studies.
+Areas of human modified habitats were the most commonly used for these rodent trapping studies with less information obtained about rodent assemblages and their potential zoonoses in less human dominated landscapes.
+
+### Limitations
+
+While the search terms favoured sensitivity over specificity it is possible that studies conducted for other purposes (i.e.
+agricultural studies) may have trapped rodents without it being a key component of the study and therefore not detected through the predominantly keyword based search.
+
+While entirely duplicated studies were excluded at the inclusion and exclusion stage of the search, however, it appears that several studies have incorporated data that had been published previously in other studies.
+One notable example was a synthesis of trapping studies undertaken by a research group based in Dakar, Senegal over 30 years, it was not possible to identify which data had not been published in prior publications and thus introduces the risk that some data were double counted in this review.
+
+Some studies did not present detailed information about the location of rodent trapping.
+Coordinates for the study were derived from information within the manuscript, this may have introduced errors as locations within much of West Africa are not uniquely named.
+
+Absence of individuals of a species was often not explicitly recorded by the study teams, the absence of a species was imputed when it's presence was reported elsewhere in the study, this assumption may not always have been correct and may introduce inappropriate absences.
+
+### Recommendations to support future meta-analyses
+
+Meta-analyses to synthesise data obtained from multiple observational studies are increasingly prevalent in both ecology and infectious disease research.
+This methodological approach could be helpful to understand both rodent ecology and the risk of zoonotic disease emergence.
+This method would currently face several limitations, we highlight these below and suggest adaptations to practice that could support meta-analysis and future re-use of study data.
+
+First, study reporting remains heterogenous with insufficient detail presented on study methodology to assess potential risks of bias, for example when choosing suitable trap locations or selecting the time period to perform the trapping study.
+This could be improved by pre-registering study protocols prior to analysis to make clear the rationale underlying these decisions.
+
+Second, the selection of trap type (i.e the size of the trap) can influence the species of rodents that are caught.
+If specific species are targeted through choice of trap size this should be explicitly stated in the methods section.
+
+Third, trapping effort was often incompletely recorded.
+Trapping effort equates to sample size and without sufficient information on the structure or quantity of this parameter incorporation of studies into meta-analyses are challenging.
+At a minimum study level trapping effort should be reported, ideally the number of trap-nights at each defined trap site should be reported.
+
+Fourth, habitat classification when recorded did not use a standardised approach limiting direct comparison between studies.
+The use of and referencing a standardised scale such as the IUCN Habitat Classification Scheme (Version 3.1) would support synthesis of results across studies.
+
+Finally, collection of data on rodents and their pathogens is often costly and is associated with risk to researchers and local communities, particularly when investigating zoonotic pathogens.
+Increasing the availability of primary data for further analysis should be promoted and is increasingly being mandated by governmental and non-governmental research funders.
+Further, the often lengthy delay between collection of data and it's subsequent publication can also lead to unnecessary duplication of research by different teams.
+This can be minimised by pre-registering studies and releasing study-protocols.
+These approaches are key components of the growing Open Science movement within ecology and beyond [@powers_open_2019].
+
+## Conclusions
+
+This scoping review identified 124 studies, conducted in 14 countries at 1,193 study sites trapping 73,164 small rodents.
+Current research to describe the rodent populations of West Africa have provided some information about the distribution of species across this heterogenous region and the distribution of potential pathogens within individual species.
+We make recommendations for subsequent research to enable wider data sharing and re-use to to support meta-analysis of obtained data to address questions on rodent ecology and zoonotic disease risk.
 

### 2021-05-11 scripts/double_checking.R

@@ -1,0 +1,28 @@
+source(here::here("scripts", "libraries.R"))
+
+# double_checking <- bib2df(here("citations", "v4_search.bib")) %>%
+#   sample_n(size = nrow(.)/10, replace = F)
+#
+# double_checking %>%
+#   dplyr::select(BIBTEXKEY, JOURNAL, TITLE, ABSTRACT, ISSN, ISBN, DOI) %>%
+#   write_csv(file = here("citations", "double_checks", "title_abstract.csv"))
+#
+# data_extraction <-  bib2df(here("citations", "include_final.bib")) %>%
+#   sample_n(size = nrow(.)/10, replace = F)
+#
+# data_extraction %>%
+#   dplyr::select(BIBTEXKEY, JOURNAL, TITLE, ABSTRACT, ISSN, ISBN, DOI) %>%
+#   write_csv(file = here("citations", "double_checks", "data_extraction.csv"))
+#
+# title_abstract <- read_csv(here("citations", "double_checks", "title_abstract.csv"))
+#
+# double_checking <- bib2df(here("citations", "v4_search.bib"))
+#
+# double_checking %>%
+#   filter(BIBTEXKEY %in% title_abstract$BIBTEXKEY) %>%
+#   df2bib(file = here("citations", "double_checks", "title_abstract.bib"))
+#
+# double_checking %>%
+#   filter(BIBTEXKEY %in% data_extraction$BIBTEXKEY) %>%
+#   df2bib(file = here("citations", "double_checks", "data_extraction.bib"))
+

### 2021-05-11 scripts/mapping_pathogens.R

@@ -254,3 +254,29 @@ all_path_fig <- tm_shape(level_0 %>%
 
 tmap_save(all_path_fig, here("figures", "top_4.png"))
 
+species <- all_path_map %>%
+  tibble() %>%
+  group_by(classification.x) %>%
+  summarise(n = n()) %>%
+  arrange(-n) %>%
+  head(n = 10)
+
+species_path <- all_path_map %>%
+  filter(classification.x %in% c("mastomys natalensis", "mastomys erythroleucus", "crocidura sp.",
+                                 "rattus rattus", "mus musculus", "praomys daltoni",
+                                 "arvicanthis niloticus", "mus minutoides", "taterillus sp.",
+                                 "mastomys huberti")) %>%
+  filter(pos_neg == "Positive")
+
+
+species_path_fig <- tm_shape(level_0 %>%
+           filter(GID_0 %in% wa_mainland)) +
+  tm_polygons(alpha = 0.5) +
+  tm_shape(species_path) +
+  tm_dots(col = "pos_neg", size = 0.05, palette = "black",
+          jitter = 0.08, alpha = 1, legend.show = F) +
+  tm_facets(by = c("pathogen", "classification.x"), free.coords = F) +
+  tm_layout() +
+  tm_add_legend(title = "Positive")
+
+tmap_save(species_path_fig, here("figures", "top_10_species.png"))

### 2021-05-18 reports/manuscript.Rmd
update May 18

@@
 ---
-title: "Associations of environmental and human factors with altered rodent populations in West Africa: A scoping review"
+title: "Rodent trapping to explore rodent ecology and zoonotic disease risk: A scoping review"
 description: |
@@
 
-A scoping review approach was used to identify and map the available information on rodent assemblages, abundance and diversity across West Africa.
-This approach is recommended when examining how research is conducted within a topic area [@munn_systematic_2018].
-The review followed the recommended appreach from the PRISMA extension for Scoping Review guidance [@tricco_prisma_2018].
+A scoping review method was adopted to identify and map the available information on rodent assemblages, abundance and diversity across West Africa.
+This is a recommended approach when examining how research is conducted within a topic area [@munn_systematic_2018].
+The review followed the PRISMA extension for Scoping Review guidance [@tricco_prisma_2018].
 The protocol for this review was not pre-registered.
@@
     + Described trap used, length of trapping activity or location of trapping activity.
-    + Included trapping activity from at least one West African country: Benin, Burkina-Faso, Cape Verde, Gambia, Ghana, Guinea, Guinea-Bissau, Cote d'Ivoire, Liberia, Mali, Mauritania, Niger, Nigeria, Senegal, Sierra Leone and Togo
+    + Included trapping activity from at least one West African country: Benin, Burkina-Faso, Cape Verde, Gambia, Ghana, Guinea, Guinea-Bissau, C&#244;te d'Ivoire, Liberia, Mali, Mauritania, Niger, Nigeria, Senegal, Sierra Leone and Togo
     + Recorded as outcome genus or species of trapped individual
@@
 
-When explicitly stated the study aims were extracted, for those with no specified aims the aims were inferred from the introduction or conclusions of the manuscript or report.
-The aims of studies were categorised with a subsequent higher order allocation to rodent ecology research or studies on risk of zoonoses.
+Explicitly stated study aims were extracted, for those with no specified aims the aims were inferred from the introduction or conclusions of the manuscript or report.
+The aims of studies were categorised to a higher order grouping, namely, rodent ecology research or studies on the risk of zoonoses.
 
@@
 
-Species classifications can vary over time as morphological and genetic information emerges.
-To handle species classification synonyms and reclassification the reported names of trapped species was compared to the Global Biodiversity Information Facility [@global_biodiversity_information_facility_gbiforg_nodate].
-For species with updated taxonomies the synonym was converted to the accepted species name for all subsequent analyses.
+Species classifications may vary over time as morphological and genetic information emerges, to handle potential species classification synonyms and reclassification the reported names of trapped species was mapped to the Global Biodiversity Information Facility [@global_biodiversity_information_facility_gbiforg_nodate] identification code.
+For species with updated taxonomies the reported species identity was converted to the accepted species name for all subsequent analyses.
 Where a subspecies is reported by study authors the species complex it belongs to will be used for subsequent analysis.
@@
 
-For studies reporting on all trapped species (i.e.
-not those only reporting on specific species of interest) that do not report absence at study sites all species trapped within the study will be recorded at all study sites with explicitly zero observations.
+The presence and absence of trapped individuals and genus/species will be extracted.
+For studies reporting on all trapped individuals (i.e.
+not those only reporting on the presence of specific species of interest) the absence of a reported capture of a species recorded elsewhere in the study will be explicitly recorded as an absence at the study location.
 
@@
 
-Descriptive analyis of included studies was conducted in R (v4.0.2) [@r_core_team_r_2020] within the RStudio IDE [@rstudio_team_rstudio_2020].
+Descriptive analysis of included studies was conducted in R (v4.0.2) [@r_core_team_r_2020] within the RStudio IDE [@rstudio_team_rstudio_2020].
 The analysis code and packages used for this analysis are available on GitHub [repository](https://doi.org/10.5281/zenodo.4718375).
@@
   as_gt() %>%
-  tab_header(title = "Table 1: Characteristics of included studies summarised by study aim") %>%
   tab_options(table.font.size = px(12),
@@
 Thirty-one (25%) studies reported trapping only at a single study site, 45 (37%) studies trapped at between two and five study sites, the remaining 48 studies trapped at between six and 93 study sites.
-Trap sites were mapped to level two country administrative regions and density of sites per thousand square kilometers was calculated (see Figure \@ref(fig:density)).
+Trap sites were mapped to level two country administrative regions and density of sites per thousand square kilometres was calculated (see Figure \@ref(fig:density)).
 The areas with highest trapping densities included the capital cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako) and Ghana (Accra) and the largest cities of Ivory Coast (Abidjan) and Benin (Cotonou).
@@
 Three studies performed direct culture of *Lassa mammarenavirus* or *Leishmania species* to detect the presence of the pathogen in rodent specimens.
-PCR was the most commonly used diagnostic assay being performed on 21,953 rodent samples, antibody or antigen based serological assays were performed on 11,430 samples, histological or direct visualisation was used on 11,229 samples with direct culture being used for 643 samples.
+PCR was performed on 21,953 rodent samples, antibody or antigen based serological assays were performed on 11,430 samples, histological or direct visualisation was used on 11,229 samples with direct culture being used for 643 samples.
 
@@
 
-While the search terms favoured sensitivity over specificity it is possible that studies conducted for other purposes (i.e.
-agricultural studies) may have trapped rodents without it being a key component of the study and therefore not detected through the predominantly keyword based search.
+Search terms favoured sensitivity over specificity, however, it remains possible that studies conducted for other purposes (i.e.
+agricultural studies) may have trapped rodents without it being a key component of the study and thus not detected through the predominantly keyword based searches.
 
-While entirely duplicated studies were excluded at the inclusion and exclusion stage of the search, however, it appears that several studies have incorporated data that had been published previously in other studies.
-One notable example was a synthesis of trapping studies undertaken by a research group based in Dakar, Senegal over 30 years, it was not possible to identify which data had not been published in prior publications and thus introduces the risk that some data were double counted in this review.
+Entirely duplicated studies were removed at the inclusion and exclusion stage of the search, however, several studies have incorporated data that had been published previously within other studies.
+One notable example was a synthesis of trapping studies undertaken by a research group based in Senegal over 30 years.
+In some cases it was not possible to identify the novel data based on presented variables, thus, introducing the risk that some data were double counted in this review.
 
-Some studies did not present detailed information about the location of rodent trapping.
-Coordinates for the study were derived from information within the manuscript, this may have introduced errors as locations within much of West Africa are not uniquely named.
+Several studies did not present detailed information about the location of rodent trapping.
+For these studies coordinates were derived from information within the manuscript.
+This may introduce errors with the geolocation of these studies as locations within much of West Africa are not uniquely named.
 
-Absence of individuals of a species was often not explicitly recorded by the study teams, the absence of a species was imputed when it's presence was reported elsewhere in the study, this assumption may not always have been correct and may introduce inappropriate absences.
+Absence of individuals of a species was often not explicitly recorded in the included studies.
+The absence of a species was imputed when it's presence was reported elsewhere in the study, this assumption may not hold and where authors had variable data collection methods and may introduce inappropriate absences.
 
-### Recommendations to support future meta-analyses
+### Recommendations
 
 Meta-analyses to synthesise data obtained from multiple observational studies are increasingly prevalent in both ecology and infectious disease research.
-This methodological approach could be helpful to understand both rodent ecology and the risk of zoonotic disease emergence.
-This method would currently face several limitations, we highlight these below and suggest adaptations to practice that could support meta-analysis and future re-use of study data.
+This approach is helpful to understand both rodent ecology and the risk of zoonotic disease emergence.
+This method would currently face several limitations, we discuss some of these below and suggest adaptations to practice that could support meta-analysis and future re-use of study data.
 
@@
 
+## Data availability
+
+All data required to reproduce this analysis are available [here](https://docs.google.com/spreadsheets/d/1rQYjHhk6uk1PoKZZVsgFlmuqWGicU2tTisk9ddfAwTM/edit?usp=sharing) (this will be converted to an archived link at manuscript submission)
+
+## Code availability
+
+All code required to reproduce this analysis, including this manuscript are available [here](https://github.com/DidDrog11/scoping_review) (this repository will be versioned at manuscript submission)
+
 ## Supplementary material

### 2021-08-26 reports/manuscript.Rmd
Adding imputed trapnights

@@
 output: 
-  distill::distill_article:
-    toc: true
-    toc_depth: 3
-    toc_float: true
+  # distill::distill_article:
+  #   toc: true
+  #   toc_depth: 3
+  #   toc_float: true
+  word_document
 knit: (function(inputFile, encoding) {

### 2021-08-26 reports/manuscript_v2.Rmd
Adding imputed trapnights

---
title: "Rodent trapping studies to understand the prevalence of zoonotic disease in West Africa: A scoping review."
description: |
date: "`r Sys.Date()`"
author: 
  - first_name: "David"
    last_name: "Simons"
    url: https://www.dsimons.org
    orcid_id: 0000-0001-9655-1656
output: 
  distill::distill_article:
    toc: true
    toc_depth: 3
    toc_float: true
  # word_document
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding, output_dir = here::here())
  })
always_allow_html: yes
bibliography: "`r here::here('citations', 'manuscript_citations.bib')`"
---

# Introduction

  * The role of zoonotic infectious diseases in human health and society have been dramatically highlighted through the ongoing SARS-CoV-2 pandemic and recent Ebola virus outbreaks in the Democratic Republic of Congo and Guinea. 
  * The number and intensity of zoonotic spillover events is projected to increase under anthropogenic pressure (e.g. increased human populations and increasing urbanisation) and global climate change [@morse_prediction_2012].
  * Rodents are abundant and diverse [@burgin_how_2018], their widespread occurrence and proximity to human activity make the understanding of the effect of environmental change on their host-pathogen interactions vitally important. 
  * Rodents are important hosts of zoonotic disease. Of 2,220 extant rodent species, 244 (10.7%) are described as being reservoirs of 85 zoonotic pathogens [@han_global_2016]. This is an example of an ecosystem disservice provided by rodents.
  * Rodents, however, also provide important ecosystem services within these habitats, including, pest regulation and seed dispersal [@fischer_ecosystem_2018].
  * Rodents typically demonstrate “fast” life histories [@dobson_fast_2007] and within rodents, these traits are associated with a species being a reservoir of zoonotic diseases [@han_rodent_2015], but see [@albery_fast-lived_2021]. Further, these traits are prevalent in species that thrive in human dominated landscapes, displacing species that are less likely to be reservoirs of zoonotic disease [@gibb_zoonotic_2020].
  * West Africa has previously been identified to be at relatively high risk of emerging zoonotic infectious disease events from wildlife populations [@jones_global_2008]. Ongoing changes in land use and population demographics may lead to an increased burden from zoonotic diseases. 
  * Rodents are implicated in the transmission pathways of several diseases in the region, including, Lassa fever, Schistosomiasis and Leptospirosis.
  * Studies have been conducted in West Africa both to identify novel potential zoonotic pathogens within rodents alongside studies investigating the prevalence and burden of known pathogens within rodent hosts (e.g. for Lassa fever [@fichet-calvet_diversity_2009] and Schistosomiasis [@catalano_multihost_2020]).
  * Our understanding of the associations between rodents and their pathogens on broader geographic scales are typically based on global datasets (e.g. PREDICTS, GIDEON and GBIF) that consolidate information obtained from multiple sources of information [@gibb_zoonotic_2020; @smith_global_2014; @pigott_mapping_2014]. These datasets may lack the contextually rich information on the presence, absence and abundance of individual species and the habitats in which they were trapped that is potentially provided by individual rodent trapping studies [@bovendorp_optimising_2017].
  * Alongside disease specific studies these datasets have been used to identify potential geographic hotspots where virus and host species diversity may be expected to be at its greatest and therefore quantify the risk of a zoonotic disease spillover event occurring at a defined location [@han_global_2016]. However, there remains the potential for important confounding introduced to these associations through bias generated by study design or through selection of sampling sites.
  * For example, systematically increased sampling (e.g. more intensive studies over longer time periods) or over-representation of certain habitats (e.g. periurban landscapes) could lead to an apparent association between locations and risk that is driven by these factors rather than an underlying host and virus association [@wille_how_2021; @gibb_mammal_2021].
  * Conversely some regions may not be sampled adequately and therefore under-represented in these datasets due to sparse human populations or inaccessible habitats which may lead to erroneously identifying regions as low risk of zoonotic disease spillover events.
  * Here, we identify rodent trapping studies performed across West Africa and identify the location and habitat types in which they have been conducted, the pathogens assessed and the host-pathogen associations that have been reported in order to quantify the potential bias and identify areas requiring further focussed investigation.

# Methods

## Literature search

  * We conducted a search in Ovid MEDLINE, Web of Science (Core collection and Zoological Record), JSTOR, BioOne, African Journals Online, Global Health and the pre-print servers, BioRxiv and EcoEvoRxiv for the following terms as keywords, no date constraints were used:

    1. Rodent OR Rodent trap*
    2. West Africa (or the individual countries)
    3. 1. AND 2.

  * We searched other resources including the UN Official Documents System, Open Grey, AGRIS FAO and Google Scholar using combinations of the above terms.
  * We ran the search on 2021-03-01.
  * We included studies if they met all of the following inclusion criteria; i) reported findings from trapping studies where the target was a small mammal, ii) described the type of trap used or the length of trapping activity or the location of the trapping activity, iii) included trapping activity from at least one West African country, iv) recorded the genus or species of trapped individuals, v) were published in a peer-reviewed journal or as a pre-print on a digital platform or as a report by a credible organisation.
  * We excluded studies if they met any of the following exclusion criteria: i) reported data that were duplicated from a previously included study, ii) no full text available, iii) not available in English.
  * One reviewer screened titles, abstracts and full texts against the inclusion and exclusion criteria. At each stage, a random subset (10%) was reviewed by a second reviewer.

## Data extraction

  * We extracted data from eligible studies to input into a Google sheets document. Extracted variables included i) study identifiers; ii) study aims; iii) trapping methodology; iv) geolocation data; v) method of speciation; vi) trapping locations and dates; vii) trapped species; viii) number of trap-nights and ix) microorganisms/pathogens of interest. The data extraction tool is archived and available in Supplementary Material 1.

### Location of rodent trapping studies and habitats studied
  * We extracted GPS locations for the most precise location presented (i.e. trap, trap-line, study site or study region). We extracted coordinates in the format reported, subsequently converting them to decimal degrees. 
  * We recorded the habitat classification scheme a study used (e.g.  IUCN Habitat Classification Scheme (Version 3.1)). For studies not using standardised recording, the explicit description from of the habitat in which the trap was placed was extracted. For studies reporting multiple habitat types (e.g. rice field, corn field and vegetable garden) for a single trap, trap-line or trapping grid, a higher order classification of habitat type was recorded (e.g.  agricultural land).

### Rodent presence, absence, abundance
  * We mapped genus and species names to the species names used in the GLobal Biodiversity Information Facility (GBIF) taxonomy  [@gbif_the_global_biodiversity_information_facility_global_2021]
  * We extacted information on the presence, absence and number of trapped individuals. For studies reporting on all trapped individuals (i.e. not those only reporting on the presence of a specific species of interest), the pseudo-absence of a species reported as present elsewhere in the study was explicitly recorded as an absence at that trap location.

### Pathogen presence and absence
  * We extracted data on all pathogens assayed for in studies investigating rodents for potential zoonoses. The number of rodents tested and the number of positive or negative samples were recorded alongside the type of assay used (e.g. Polymerase Chain Reaction (PCR), Enzyme Linked ImmunoSorbent Assay (ELISA) or viral culture). If studies reported indeterminate results this was noted. Where possible, pathogens were identified to species level. However, where an assay only allowed for attribution to a family of viruses or bacteria, the higher order grouping was used (i.e. PCR using a non-specific arenavirus primer).

## Analysis

### Location and habitats of rodent trapping to investigate potential biases

  * We summarised the number of studies, the year in which trapping occurred and the country in which they were conducted. In studies not reporting the year of trapping we used the year of publication as a proxy.
  * We used the GPS coordinates of single trap, trapping grid/line or study site and the number of trap nights to calculate trapping effort (trap night density) within level 2 administrative areas in West Africa. 
  * For studies not reporting the number of trap nights we imputed the number based on the median trap success rate from studies which reported the number of trap nights (matched to building or non-building based study sites).
  * We summarised the habitat types of trap sites based on information reported in the study.
  * For the subset of studies investigating rodent zoonoses we compared the location of trapping sites was to SEDAC Global Population Density estimates (1970-2000 (10 year) and 2000-2020 (5 year)) and to land cover classifications (1975, 2000 and 2013) to explore the association of human population and rodent trapping effort.
  * We mapped the presence and absence of rodent species and compared this to the presence and absence reported by both GBIF and IUCN to give a measure of the extent of each species range in which they have been sampled. 
  * **Which statistical test to compare?** Ranked correlation to compare presence of habitat type against number of trap nights in habitat? Same analysis but weighted for human population?

### Rodent pathogen associations

  * We summarised the presence and absence of pathogens, the assays used, their host species and the locations from where the samples were obtained.
  * We investigated the association between rodent species and the detection of potential pathogens. We report proportion of positive and negative tests for each species and pathogen pair.

# Results

## Included studies

  * 4,282 citations were identified, with 126 rodent trapping studies included in narrative synthesis.
  * The earliest included studies were from 1974 with increasing numbers of studies being performed annually since 2005.
  * The median length of rodent trapping activity is one year (IQR 0-2 years). The median time from completion of rodent trapping to publication is 3 years (IQR 2-5 years) (see Figure 1 - Panel A).

## Location and habitats of rodent trapping studies to investigate potential biases

  * Rodent trapping took place across 1,307 study sites with at least one trap site recorded from 14 West African countries. No rodent trapping studies were identified from Gambia or Togo.
  * Thirty-one (25%) studies reported trapping at a single study site, 46 (37%) studies trapped at between two and five study sites, the remaining 49 studies trapped at between six and 93 study sites. The areas with highest trapping activity included the capital cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako) and Ghana (Accra) and the largest cities of Ivory Coast (Abidjan) and Benin (Cotonou). Outside of these cities, Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone were the most intensively studied regions (see Figure 1 - Panel B).
  * No studies reported trap habitats with reference to a standardised habitat classification scheme. 
  * Extracted habitat types were grouped into 30 categories (see Supplementary Material 2 for the habitat dictionary). At least one habitat was recorded for 17,122 trap sites (95%), with two or more habitats for a single trap site recorded for 4,403 (24%) sites. Single trap sites could span multiple habitat types resulting in 22,202 distinct habitat and trap sites. 
  * The most commonly trapped sites were in or around buildings (29%), in areas described as the rodents “natural habitat” (24%), in agricultural areas (13%) (e.g. rice fields, palm plantations), forests (8%) and in the area surrounding buildings (4%).
  * Explore whether trap sites are more likely to have been conducted within high population regions or whether there is an association with trap location and habitat as obtained from remote sensing data (Need to do this. The remote sensing data will be matched to the trap dates there will be several years between the trapping and data this is compared to it seems impractical to try and get annual population estimates or land classification so this will be chosen pragmatically)

Figure 1 - Panel A - location of traps, Panel B - The distribution of population density at trap sites, Panel C - The distribution of habitat types at trap sites compared to all habitat types in West Africa

## Rodent presence, absence, abundance
  * 73,164 small mammals were trapped (592 were trapped outside of West African countries), 2,830 (4%) trapped individuals were identified to order level (Rodentia), 7,760 (11%) were identified to genus level, the remaining 62,574 (85%) were identified to species level. 
  * In studies reporting the number of trap nights the median trap success rate of traps placed in or around buildings was 13% (IQR 6-23%), this compares to a median trap success rate of 3% (IQR 1-9%) in other habitats.
  * Of the 147 distinct identified species trapped (see Supplementary Material 3) the majority were from the order Rodentia (112). Muridae (82) represented the largest family of rodents, followed by, Sciuridae (10), Gliridae and Nesomyidae (both 6), Ctenodactylidae and Anomaluridae (2) and Dipodidae, Hystricidae and Thryonomidae (all 1). The remaining 34 species came from the orders of the Soricomorpha (30), Erinaceomorpha (2) and Afrosoricida (1).
  * The most commonly trapped genera of rodents were Mastomys sp. (27,072, 38.5%), Rattus sp. (10,191, 14.5%), Mus sp. (8,624, 12.3%), Arvicanthis sp. (5,821, 8.3%) and Praomys sp. (5,409, 7.7%). At species level Mastomys natalensis (11,221, 17.9%), Rattus rattus (8,578, 13.7%), Mastomys erythroleucus (7,379, 11.8%), Mus musculus (6,245, 10%), Arvicanthis niloticus (5,497, 8.8%) and Mastomys huberti (4,699, 7.5%) were the most commonly trapped. 
  * Presence and absence maps for the six most commonly trapped species are compared to GBIF and IUCN data. (Need to do this)
  * The number of individuals trapped provides a measure of abundance, the interpretation of this is limited by incomplete reporting on the trapping effort (i.e. number of trap nights) within a study. The most abundant species at presence locations were … (Need to do this. I will explore the number of individuals trapped by species, I imagine it will show that if a species is detected at a site the most abundant ones will be the synanthropic species such as mus, rattus etc. while the more specialised species will be less abundant in areas where they are detected)

Figure 2 - Three plots side by side with 6 rows, one each for the most commonly trapped species L - presence/absence from this review, C - presence/absence from GBIF, R - Species range from IUCN

## Pathogen presence and absence
  * Sixty-two studies presented data on microorganisms that infect or are carried by small mammal species in West Africa.
  * Seven studies solely investigated pathogens of rodents, including Hydatigera species (previously Taenia species) and Trichuris species.
  * Fifty-five studies investigated organisms that were potentially zoonotic pathogens. Thirty-two microorganisms were tested for, 8 of these were defined at species level, with the remaining 24 at levels of higher taxonomic classification.
  * Thirty-two studies used Polymerase Chain Reaction (PCR) to detect the presence of 22 different species or families of microorganisms in 21,953 rodent samples. Eleven studies used antibody or antigen based molecular tests to detect the presence of 9 different species or families of microorganisms in 11,430 samples. Eight studies conducted histological or direct visualisation assays of samples for 11 parasitic or bacterial species in 11,229 samples. Three studies performed direct culture of Lassa mammarenavirus or Leishmania species to detect the presence of the pathogen in 643 samples.
  * The most common pathogens assessed for using PCR were Lassa mammarenavirus or other Arenaviridae (31%) the bacteria Borrelia sp. (11.3%) and Bartonella sp. (6.5%) followed by Usutu virus (6.4%) and Hantaviridae (5.6%). The most common pathogens assessed for with serology were Lassa mammarenavirus or other Arenaviridae (78%), Toxoplasma gondii (10.6%), Borrelia sp. (6.3%) and Leptospirosis sp. (2.2%). The most common pathogens assessed for with histology or direct visualisation were Borrelia sp. (48.7%), Schistosoma sp. (20.4%) and other parasites. All studies using direct culture were investigating either Lassa mammarenavirus (81%) or Leishmania sp. (19%). 
  * Most studies tested for a single pathogen (39), with 16 studies testing for two or more pathogens. The most frequently tested for pathogens were Lassa mammarenavirus (28%) or members of the Arenaviridae family (16%), the spirochete bacteria Borrelia sp. was investigated in 8 studies, Bartonella sp. and Toxoplasma gondii were investigated in 4 each, the remaining 25 pathogens were reported in three or fewer studies.
Define presence and absence of pathogens by country (Need to do)

Figure 3 - Presence/absence plots for the following pathogens Arenaviruses with LASV highlighted, Bartonella, Borrelia and Toxoplasma.

## Host-pathogen associations
  * Ninety-seven rodent species were investigated for the presence of a zoonotic pathogen. 42,940 assays were performed on 32,014 individual rodents. 
  * The rodent species most commonly assayed for zoonotic pathogens were, Rattus rattus (n = 2,977) assessed for 24 pathogens, Mus musculus (n = 3,402) for 23, Mastomys natalensis (n = 7,189) and Mastomys erythroleucus (n = 3,013) both 19 and Arvicanthis niloticus (n = 3,840) for 18.    * All remaining species were investigated for 10 or fewer pathogens.
  * Lassa mammarenavirus was detected in 414 Mastomys natalensis, 9 Mastomys erythroleucus, 5 Hylomyscus pamfi and 3 Mus baoulei through PCR. In serological assays, evidence of prior infection or acquisition of maternal antibodies to Lassa mammarenavirus was found in 465 Mastomys natalensis, 361 Mastomys sp., 6 Mus musculus and 4 Rattus rattus. From direct viral culture 19 Mastomys natalensis, 2 Rattus rattus and 1 Mus minutoides were found to be infected with Lassa mammarenavirus. Bartonella sp. were detected in; 113 Mastomys natalensis, 50 Mastomys erythroleucus, 39 Rattus norvegicus, 16 Rattus rattus and 13 individuals of 7 further species through PCR.
  * Borrelia sp. infections were identified through PCR in 110 individuals; 37 Mastomys erythroleucus, 12 Mus musculus, 11 Mastomys huberti and Praomys daltoni., tThe remaining 39 positive results were from 9 further species. In serological assays, 47 Mastomys natalensis, 12 Mastomys erythroleucus and 8 Praomys daltoni were found to be positive, the remaining 11 positive individuals came from three further species. Most Borrelia sp. infections were identified through histology or direct visualisation with, 243 Arvicanthis niloticus, 102 Mastomys natalensis, and 78 Mastomys huberti along with a further 64 individuals from 8 species infected.
  * Finally, Toxoplasma gondii was detected via PCR in 89 Mus musculus, 13 Cricetomys gambianus, 5 Crocidura olivieri and 3 Rattus rattus. From serological assays evidence of infection was found in 25 Mus musculus, 46 Mastomys erythroleucus, 4 Rattus rattus, and Mastomys natalensis, using histology and direct visualisation 1 Mus musculus and 1 Rattus rattus were also found to be infected.
  * Thirty-eight species of Rodentia, 3 species of Soricomorpha and 2 species of Erinaceomorpha that were tested for potential pathogens had entirely negative results.

Figure 4 - Matrix heat plot Y - rodent species/genera, X - pathogen species/genera. Colour relates to proportion positive (Perhaps use bivariate colour to also highlight the number of test performed in that species).

## Discussion

  * 124 trapping studies were identified across 14 West African countries. 54 studies provided additional information about potential zoonotic pathogens of trapped rodents.

### 2021-08-26 scripts/1_download_spatial.R

@@ -1,9 +1,12 @@
 source(here::here("scripts", "libraries.r"))
+rodent_spatial <- read_rds(here("data_clean", "rodent_spatial.rds"))
+
 
 countries <- rodent_spatial %>%
   drop_na(country) %>%
   mutate(iso3 = countrycode(as.character(country), "country.name", "iso3c"))
 countries <- unique(countries$iso3)
+
 # Level 0 -----------------------------------------------------------------
 #WA
 BEN_0 <- getData("GADM", country = "Benin", level = 0, path = here("data_download", "admin_spatial"))

### 2021-08-26 scripts/1_update_data.R

@@ -1,14 +1,12 @@
 source(here::here("scripts", "libraries.R"))
 
-register_google(google_api)
-
 drive_download("https://docs.google.com/spreadsheets/d/1rQYjHhk6uk1PoKZZVsgFlmuqWGicU2tTisk9ddfAwTM/edit#gid=0", path = here("data_download", "included_studies.xlsx"), overwrite = T)
 
 studies <- read_xlsx(here("data_download", "included_studies.xlsx"), sheet = "study")
 write_rds(studies, here("data_raw", "studies.rds"))
 
 rodent_data <- read_xlsx(here("data_download", "included_studies.xlsx"), sheet = "trapping", col_types = c("guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess",
-                                                                                     "guess", "guess", "guess", "numeric", "guess", "guess"))
+                                                                                     "guess", "guess", "text", "text", "guess", "guess"))
 write_rds(rodent_data, here("data_raw", "rodent_data.rds"))
 
 pathogen_data <- read_xlsx(here("data_download", "included_studies.xlsx"), sheet = "pathogen", col_types = c("guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess",

### 2021-08-26 scripts/2_1_cleaning.R

@@ -1,5 +1,5 @@
-source(here::here("scripts", "libraries.r"))
-source(here("scripts", "update_data.r"))
+source(here::here("scripts", "libraries.R"))
+source(here("scripts", "update_data.R"))
 
 # Studies -----------------------------------------------------------------
 

### 2021-08-26 scripts/analysis.R

@@ -1,0 +1,1009 @@
+source(here::here("scripts", "libraries.r"))
+
+# Data --------------------------------------------------------------------
+studies <- read_rds(here("data_clean", "studies.rds"))
+rodent_data <- read_rds(here("data_clean", "rodent_df.rds")) %>%
+  tibble()  %>%
+  dplyr::select(-all_of(c("intensity_use", "geometry")))
+
+habitat_split <- c("habitat_1", "habitat_2", "habitat_3", "habitat_4", "habitat_5", "habitat_6", "habitat_7")
+habitat_data <- read_rds(here("data_clean", "habitat_types.rds")) %>%
+  tibble() %>%
+  dplyr::select(record_id, all_of(habitat_split), intensity_use)
+
+rodent_data %<>%
+  left_join(.,
+            habitat_data ,
+            by = "record_id")
+
+imputed_tn <- read_rds(here("data_clean", "imputed_trap_nights.rds"))
+
+rodent_data <- full_join(rodent_data, imputed_tn,
+                  by = c("unique_id", "year_trapping", "month_trapping",
+                         "town_village", "habitat")) %>%
+  mutate(trap_nights.x = case_when(trap_nights.x == "50-200" ~ as.numeric(125),
+                                   TRUE ~ as.numeric(trap_nights.x)),
+         trap_nights = case_when(is.na(trap_nights.y) ~ trap_nights.x,
+                                 TRUE ~ trap_nights.y),
+         trap_night_data = case_when(is.na(trap_night_data) ~ "Actual",
+                                     TRUE ~ trap_night_data)) %>%
+  select(-c(trap_nights.x, trap_nights.y))
+
+pathogen <- read_rds(here("data_clean", "pathogen.rds"))
+wide_pathogen <- read_rds(here("data_clean", "wide_pathogen.rds"))
+species_gbif <- read_rds(here("data_clean", "species_data.rds")) %>%
+  distinct(genus, species, gbif_id, genus_gbif, species_gbif)
+
+long_pathogen <- read_rds(here("data_clean", "long_pathogen.rds"))
+
+
+# Summary table -----------------------------------------------------------
+
+studies %>%
+  mutate(decade_publication = case_when(year_publication < 1980 ~ "1970's",
+                                        year_publication < 1990 ~ "1980's",
+                                        year_publication < 2000 ~ "1990's",
+                                        year_publication < 2010 ~ "2000's",
+                                        year_publication < 2020 ~ "2010's",
+                                        year_publication < 2030 ~ "2020's",
[…]

### 2021-08-26 scripts/descriptive.R

@@ -1,943 +1,0 @@
-source(here::here("scripts", "libraries.r"))
-
-
-# Data --------------------------------------------------------------------
-studies <- read_rds(here("data_clean", "studies.rds"))
-rodent_data <- read_rds(here("data_clean", "rodent_df.rds")) %>%
-  tibble()  %>%
-  dplyr::select(-all_of(c("habitat", "intensity_use", "geometry")))
-
-habitat_split <- c("habitat_1", "habitat_2", "habitat_3", "habitat_4", "habitat_5", "habitat_6", "habitat_7")
-habitat_data <- read_rds(here("data_clean", "habitat_types.rds")) %>%
-  tibble() %>%
-  dplyr::select(record_id, all_of(habitat_split), intensity_use)
-
-rodent_data %<>%
-  left_join(.,
-            habitat_data ,
-            by = "record_id")
-
-pathogen <- read_rds(here("data_clean", "pathogen.rds"))
-wide_pathogen <- read_rds(here("data_clean", "wide_pathogen.rds"))
-species_gbif <- read_rds(here("data_clean", "species_data.rds")) %>%
-  distinct(genus, species, gbif_id, genus_gbif, species_gbif)
-
-long_pathogen <- read_rds(here("data_clean", "long_pathogen.rds"))
-
-
-# Summary table -----------------------------------------------------------
-
-studies %>%
-  mutate(decade_publication = case_when(year_publication < 1980 ~ "1970's",
-                                        year_publication < 1990 ~ "1980's",
-                                        year_publication < 2000 ~ "1990's",
-                                        year_publication < 2010 ~ "2000's",
-                                        year_publication < 2020 ~ "2010's",
-                                        year_publication < 2030 ~ "2020's",
-                                        TRUE ~ "other"),
-         geolocation_level = case_when(geolocation_level == "island" ~ "Study site",
-                                       geolocation_level == "none" ~ "No geolocation",
-                                       geolocation_level == "study" ~ "Study",
-                                       geolocation_level == "study_region" ~ "Study",
-                                       geolocation_level == "study_site" ~ "Study site",
-                                       geolocation_level == "trap_line" ~ "Specific trap site",
[…]

### 2021-08-26 scripts/imputing_trapnights.R

@@ -1,0 +1,124 @@
+source(here::here("scripts", "libraries.R"))
+
+# Data --------------------------------------------------------------------
+rodent_data <- read_rds(here("data_clean", "rodent_df.rds")) %>%
+  tibble()
+habitat_split <- c("habitat_1", "habitat_2", "habitat_3", "habitat_4", "habitat_5", "habitat_6", "habitat_7")
+
+rodent_data %<>%
+  separate(habitat, into = c(habitat_split), sep = ", ", remove = F)
+
+source(here("scripts", "habitat_dictionary.R"))
+cleaned_habitat <- read_rds(here("data_clean", "habitat_dictionary.rds"))
+
+rodent_data %<>%
+  mutate(habitat_1 = recode(habitat_1, !!!cleaned_habitat),
+         habitat_2 = recode(habitat_2, !!!cleaned_habitat),
+         habitat_3 = recode(habitat_3, !!!cleaned_habitat),
+         habitat_4 = recode(habitat_4, !!!cleaned_habitat),
+         habitat_5 = recode(habitat_5, !!!cleaned_habitat),
+         habitat_6 = recode(habitat_6, !!!cleaned_habitat),
+         habitat_7 = recode(habitat_7, !!!cleaned_habitat)) # the mapping is then applied to all the reported habitats
+# Imputing traping effort -------------------------------------------------
+
+study_site <- tibble(rodent_data) %>%
+  mutate(trap_night_unit = case_when(trap_night_unit %in% c("trap_site", "habitat", "study_site", "village", "study_visit", "study_habitat", "site", "trap_session") ~ "study_site",
+                                     TRUE ~ trap_night_unit)) %>%
+  filter(trap_night_unit == "study_site" & is.na(study_nights)) %>%
+  distinct(unique_id, year_trapping, month_trapping, iso3c, region, town_village, habitat,
+           habitat_1, habitat_2, habitat_3, habitat_4, habitat_5, habitat_6, habitat_7,
+           classification, number, trap_nights, capture_rate) %>%
+  mutate(trap_nights = case_when(trap_nights == "50-200" ~ as.numeric(125),
+                                 TRUE ~ as.numeric(trap_nights))) %>%
+  group_by(unique_id, year_trapping, month_trapping, iso3c, region, town_village, habitat, habitat_1,
+           habitat_2, habitat_3, habitat_4, habitat_5, habitat_6, habitat_7, trap_nights, capture_rate) %>%
+  summarise(total_captures = sum(number))
+
[…]

### 2021-08-27 reports/manuscript_v2.Rmd
updating map/analysis

@@
   * We mapped genus and species names to the species names used in the GLobal Biodiversity Information Facility (GBIF) taxonomy  [@gbif_the_global_biodiversity_information_facility_global_2021]
-  * We extacted information on the presence, absence and number of trapped individuals.
+  * We extracted information on the presence, absence and number of trapped individuals.
 For studies reporting on all trapped individuals (i.e.
@@
   * For studies not reporting the number of trap nights we imputed the number based on the median trap success rate from studies which reported the number of trap nights (matched to building or non-building based study sites).
+The median trap success rate for all rodents at a defined trap site was calculated for both trap sites including built environments and non-built environments.
+The number of rodents trapped at a trap site was then used in combination with these values of trap success to impute the number of trap nights for trap sites with no reported trap nights.
   * We summarised the habitat types of trap sites based on information reported in the study.
@@
 
-  * Rodent trapping took place across 1,307 study sites with at least one trap site recorded from 14 West African countries. No rodent trapping studies were identified from Gambia or Togo.
+  * Rodent trapping took place across 1,331 trap sites with at least one trap site recorded from 14 West African countries. No rodent trapping studies were identified from Gambia or Togo.
   * Thirty-one (25%) studies reported trapping at a single study site, 46 (37%) studies trapped at between two and five study sites, the remaining 49 studies trapped at between six and 93 study sites.
@@
 
-Figure 1 - Panel A - location of traps, Panel B - The distribution of population density at trap sites, Panel C - The distribution of habitat types at trap sites compared to all habitat types in West Africa
+Figure 1 - Panel A - plot of studies, Panel B - location of traps, Panel C - The distribution of population density at trap sites, Panel D- The distribution of habitat types at trap sites compared to all habitat types in West Africa
+
+Figure 1: **Panel A:** Each row represents one of the 126 included studies, green points designate the first year of data collection, blue points designate the end of data collection.
+For studies completed within one year the blue point overlies the green.
+Studies with a transparent grey point did not report the year in which trapping was conducted.
+The year of publication is shown by a red point.
+**Panel B:** Map of West Africa, countries where rodent trapping has occurred are mapped to level 2 administrative areas (where available).
+These regions are coloured by the total number of trap nights performed at trap sites within their boundaries.
+**Panel C:** Histogram of human population density at 5km^2^ resolution across West Africa and for areas where trapping has occurred.
+**Panel D:** Histogram of habitat types at 1km^2^ resolution across West Africa and for areas where trapping has occurred.
 
@@
   * 73,164 small mammals were trapped (592 were trapped outside of West African countries), 2,830 (4%) trapped individuals were identified to order level (Rodentia), 7,760 (11%) were identified to genus level, the remaining 62,574 (85%) were identified to species level.
-  * In studies reporting the number of trap nights the median trap success rate of traps placed in or around buildings was 13% (IQR 6-23%), this compares to a median trap success rate of 3% (IQR 1-9%) in other habitats.
+  * In studies reporting the number of trap nights the median trap success rate of traps placed in or around buildings was 13% (IQR 6-24%), this compares to a median trap success rate of 3% (IQR 1-9%) in other habitats.
   * Of the 147 distinct identified species trapped (see Supplementary Material 3) the majority were from the order Rodentia (112).

### 2021-08-27 scripts/1_download_spatial.R

@@ -1,7 +1,6 @@
 source(here::here("scripts", "libraries.r"))
 rodent_spatial <- read_rds(here("data_clean", "rodent_spatial.rds"))
 
-
 countries <- rodent_spatial %>%
   drop_na(country) %>%
   mutate(iso3 = countrycode(as.character(country), "country.name", "iso3c"))
@@ -9,72 +8,122 @@ countries <- unique(countries$iso3)
 
 # Level 0 -----------------------------------------------------------------
 #WA
-BEN_0 <- getData("GADM", country = "Benin", level = 0, path = here("data_download", "admin_spatial"))
-BFA_0 <- getData("GADM", country = "BFA", level = 0, path = here("data_download", "admin_spatial"))
-CPV_0 <- getData("GADM", country = "CPV", level = 0, path = here("data_download", "admin_spatial"))
-GMB_0 <- getData("GADM", country = "GMB", level = 0, path = here("data_download", "admin_spatial"))
-GHA_0 <- getData("GADM", country = "Ghana", level = 0, path = here("data_download", "admin_spatial"))
-GIN_0 <- getData("GADM", country = "Guinea", level = 0, path = here("data_download", "admin_spatial"))
-GNB_0 <- getData("GADM", country = "GNB", level = 0, path = here("data_download", "admin_spatial"))
-CIV_0 <- getData("GADM", country = "CIV", level = 0, path = here("data_download", "admin_spatial"))
-LBR_0 <- getData("GADM", country = "LBR", level = 0, path = here("data_download", "admin_spatial"))
-MLI_0 <- getData("GADM", country = "Mali", level = 0, path = here("data_download", "admin_spatial"))
-MRT_0 <- getData("GADM", country = "Mauritania", level = 0, path = here("data_download", "admin_spatial"))
-NER_0 <- getData("GADM", country = "Niger", level = 0, path = here("data_download", "admin_spatial"))
-NGA_0 <- getData("GADM", country = "Nigeria", level = 0, path = here("data_download", "admin_spatial"))
-SEN_0 <- getData("GADM", country = "Senegal", level = 0, path = here("data_download", "admin_spatial"))
-SLE_0 <- getData("GADM", country = "SL", level = 0, path = here("data_download", "admin_spatial"))
[…]

### 2021-08-27 scripts/2_5_imputing_trapnights.R

@@ -19,7 +19,7 @@ rodent_data %<>%
          habitat_5 = recode(habitat_5, !!!cleaned_habitat),
          habitat_6 = recode(habitat_6, !!!cleaned_habitat),
          habitat_7 = recode(habitat_7, !!!cleaned_habitat)) # the mapping is then applied to all the reported habitats
-# Imputing traping effort -------------------------------------------------
+# Imputing trapping effort -------------------------------------------------
 
 study_site <- tibble(rodent_data) %>%
   mutate(trap_night_unit = case_when(trap_night_unit %in% c("trap_site", "habitat", "study_site", "village", "study_visit", "study_habitat", "site", "trap_session") ~ "study_site",
@@ -60,6 +60,18 @@ non_building_study_site <- study_site %>%
 summary(non_building_study_site$trap_success)
 general_ts <- median(non_building_study_site$trap_success, na.rm = T)
 
+complete_tn <- tibble(rodent_data) %>%
+  mutate(trap_night_unit = case_when(trap_night_unit %in% c("trap_site", "habitat", "study_site", "village", "study_visit", "study_habitat", "site", "trap_session") ~ "study_site",
+                                     TRUE ~ trap_night_unit)) %>%
+  filter(trap_night_unit == "study_site" & is.na(study_nights)) %>%
+  mutate(trap_nights = case_when(trap_nights == "50-200" ~ as.numeric(125),
+                                 TRUE ~ as.numeric(trap_nights))) %>%
+  group_by(unique_id, year_trapping, month_trapping, region, town_village, habitat, trap_nights, capture_rate) %>%
+  summarise(total_captures = sum(number)) %>%
+  mutate(trap_nights = round(case_when(is.na(trap_nights) ~ total_captures/as.numeric(capture_rate),
+                                       TRUE ~ trap_nights), 0),
+         trap_night_data = "Actual")
+
 missing_tn <- tibble(rodent_data) %>%
   mutate(trap_night_unit = case_when(trap_night_unit %in% c("trap_site", "habitat", "study_site", "village", "study_visit", "study_habitat", "site", "trap_session") ~ "study_site",
                                      TRUE ~ trap_night_unit)) %>%
@@ -69,21 +81,21 @@ missing_tn <- tibble(rodent_data) %>%
            classification, number, trap_nights, capture_rate, trap_night_unit) %>%
[…]

### 2021-08-27 scripts/mapping_traps.R

@@ -9,24 +9,26 @@ wa_countries <- c("BEN", "BFA", "CIV", "CPV", "ESH", "GHA",
 no_data_countries <- c("GMB", "TGO")
 
 level_0 <- read_rds(here("data_download", "admin_spatial", "level_0_admin.rds"))
-list2env(level_0, envir = .GlobalEnv)
-level_0 <- do.call(rbind.SpatialPolygonsDataFrame, level_0)  %>%
-  st_as_sf()
 
 level_1 <- read_rds(here("data_download", "admin_spatial", "level_1_admin.rds"))
-list2env(level_1, envir = .GlobalEnv)
-level_1_all <- do.call(rbind.SpatialPolygonsDataFrame, level_1) %>%
-  st_as_sf()
 
 level_2 <- read_rds(here("data_download", "admin_spatial", "level_2_admin.rds"))
-list2env(level_2, envir = .GlobalEnv)
-level_2_all <- do.call(rbind.SpatialPolygonsDataFrame, level_2) %>%
-  st_as_sf()
 
 studies <- read_rds(here("data_clean", "studies.rds"))
 
-rodent_spatial <- read_rds(here("data_clean", "rodent_spatial.rds"))
+rodent_spatial <- read_rds(here("data_clean", "rodent_spatial.rds")) %>%
+  select(-trap_nights)
+
+imputed_tn <- read_rds(here("data_clean", "imputed_trap_nights.rds"))
+
+rodent_spatial <- rodent_spatial %>%
+  left_join(., imputed_tn,
+            by = c("unique_id", "year_trapping", "month_trapping",
+                   "region", "town_village", "habitat"))
+
 bbox_rodent <- st_bbox(rodent_spatial)
+
+tmap_options(check.and.fix = TRUE)
 
 #expand the bounding box
 xcrange <- bbox_rodent$xmax - bbox_rodent$xmin # range of x values
@@ -39,42 +41,40 @@ bbox_rodent <- bbox_rodent %>%
   st_as_sfc()
 
 if(!file.exists(here("figures", "static_site_map.png"))){
-trapping_map <- tm_shape(BEN_0, bbox = bbox_rodent) + tm_polygons(col = "#f2f0f0") + tm_layout(frame = F) +
-  tm_shape(BFA_0) + tm_polygons(col = "#f2f0f0") +
-  tm_shape(GMB_0) + tm_polygons(col = "white") +
[…]

### 2021-08-30 reports/manuscript_v2.Rmd
looking at pop

@@
   * Thirty-one (25%) studies reported trapping at a single study site, 46 (37%) studies trapped at between two and five study sites, the remaining 49 studies trapped at between six and 93 study sites.
+Trap sites were situated in 273 (19.3%) of 1414 level 2 administrative regions from the 14 West African nations.
 The areas with highest trapping activity included the capital cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako) and Ghana (Accra) and the largest cities of Ivory Coast (Abidjan) and Benin (Cotonou).
@@
 
-Figure 2 - Three plots side by side with 6 rows, one each for the most commonly trapped species L - presence/absence from this review, C - presence/absence from GBIF, R - Species range from IUCN
+Figure 2 - Three plots side by side with 6 rows, one each for the most commonly trapped species Panel A - presence/absence from this review, Panel B - presence/absence from GBIF, Panel C - Species range from IUCN
+
+Figure 2 - **Panel A** shows the presence and absence of the 6 most commonly trapped rodent species from the current study.
+The perimeter polygon (blue) is the convex hull of the observation points produced from presence only data.
+**Panel B**
 

### 2021-08-30 scripts/mapping_presence.R

@@ -1,85 +1,0 @@
-source(here::here("scripts", "libraries.r"))
-
-wa_mainland <- c("BEN", "BFA", "CIV", "ESH", "GHA",
-                  "GIN", "GMB", "GNB", "LBR", "MLI", "MRT",
-                  "NER", "NGA", "SEN", "SLE", "TGO")
-wa_countries <- c("BEN", "BFA", "CIV", "CPV", "ESH", "GHA",
-                  "GIN", "GMB", "GNB", "LBR", "MLI", "MRT",
-                  "NER", "NGA", "SEN", "SLE", "TGO")
-
-level_0 <- read_rds(here("data_download", "admin_spatial", "level_0_admin.rds"))
-list2env(level_0, envir = .GlobalEnv)
-level_0 <- do.call(rbind.SpatialPolygonsDataFrame, level_0)  %>%
-  st_as_sf()
-
-level_1 <- read_rds(here("data_download", "admin_spatial", "level_1_admin.rds"))
-list2env(level_1, envir = .GlobalEnv)
-level_1_all <- do.call(rbind.SpatialPolygonsDataFrame, level_1) %>%
-  st_as_sf()
-
-level_2 <- read_rds(here("data_download", "admin_spatial", "level_2_admin.rds"))
-list2env(level_2, envir = .GlobalEnv)
-level_2_all <- do.call(rbind.SpatialPolygonsDataFrame, level_2) %>%
-  st_as_sf()
-
-species_data <- read_rds(here("data_clean", "species_data.rds"))
-rodent_spatial <- read_rds(here("data_clean", "rodent_spatial.rds")) %>%
-  filter(iso3c %in% wa_countries)
-studies <- read_rds(here("data_clean", "studies.rds"))
-
-count_species <- species_data %>%
-  filter(iso3c %in% wa_countries) %>%
-  group_by(species_gbif, classification) %>%
-  drop_na(species_gbif) %>%
-  summarise(number = sum(number)) %>%
-  mutate(percent = round(number/sum(.$number)*100, 2)) %>%
-  arrange(-percent) %>%
-  rename(`GBIF ID` = "species_gbif",
-         "Classification" = "classification",
-         "Number of individuals" = "number",
-         "Percent (%)" = "percent") %>%
-  mutate(Classification = snakecase::to_sentence_case(Classification)) # the number of individuals trapped identified to species level
-
-top_6_species <- head(count_species$`GBIF ID`, 6)
[…]

### 2021-08-30 reports/manuscript_v2.Rmd

@@
   * We summarised the habitat types of trap sites based on information reported in the study.
-  * For the subset of studies investigating rodent zoonoses we compared the location of trapping sites was to SEDAC Global Population Density estimates (1970-2000 (10 year) and 2000-2020 (5 year)) and to land cover classifications (1975, 2000 and 2013) to explore the association of human population and rodent trapping effort.
+  * For the subset of studies investigating rodent zoonoses we compared the location of trapping sites with 2005 SEDAC Global Population Density estimates, the median year studies were commenced, using a Generalised Additive Model (GAM) incorporating a spatial interaction term to investigate the association of number of trap nights and human population density.
+The model structure was specified as below:
+    * *trap night density* ~ *Tweedie*(*log*(Population density (2005)) + (Longitude * Latitude))   
+  * We repeated this approach for land cover classifications obtained from the European Space Agency Copernicus dataset (2005) using a GAM incorporating a spatial interaction term to investigate the association of trap nights and land cover.
   * We mapped the presence and absence of rodent species and compared this to the presence and absence reported by both GBIF and IUCN to give a measure of the extent of each species range in which they have been sampled.
-  * **Which statistical test to compare?** Ranked correlation to compare presence of habitat type against number of trap nights in habitat? Same analysis but weighted for human population?
 
@@
 Outside of these cities, Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone were the most intensively studied regions (see Figure 1 - Panel B).
+  * Greater numbers of trap nights were conducted in areas with higher population densities (*R~2~* = 0.167).
+In particular Mauritania, Northern Senegal, Eastern Sierra Leone, Eastern Guinea and South West Nigeria were trapped at increased rates than would be expected based on their population densities.
+South West Nigeria, Northern Nigeria, Liberia, the Ivory Coast, Ghana, Niger and Burkina Faso were trapped at lower rates.
   * No studies reported trap habitats with reference to a standardised habitat classification scheme. 
@@
 The remote sensing data will be matched to the trap dates there will be several years between the trapping and data this is compared to it seems impractical to try and get annual population estimates or land classification so this will be chosen pragmatically)
+
+```{r, figure_1}
+
+cowplot::plot_grid(plotlist = )
+
+```
 

### 2021-08-31 reports/manuscript_v2.Rmd
finishing fig 2

@@
 ---
+```{r, echo=FALSE}
+source(here::here("scripts", "manuscript_library.R"))
+```
 
@@
     * *trap night density* ~ *Tweedie*(*log*(Population density (2005)) + (Longitude * Latitude))   
-  * We repeated this approach for land cover classifications obtained from the European Space Agency Copernicus dataset (2005) using a GAM incorporating a spatial interaction term to investigate the association of trap nights and land cover.
+  * We obtained land cover classifications from the European Space Agency Copernicus dataset at 300m^2^ spatial resolution (2005) we extracted the proportion of land cover classes within all regions of West Africa and all regions in which rodent trapping for zoonotic diseases.
+We compared the proportion of possible land cover classes with those of regions where trapping occurred.
   * We mapped the presence and absence of rodent species and compared this to the presence and absence reported by both GBIF and IUCN to give a measure of the extent of each species range in which they have been sampled.
@@
   * The earliest included studies were from 1974 with increasing numbers of studies being performed annually since 2005.
-  * The median length of rodent trapping activity is one year (IQR 0-2 years). The median time from completion of rodent trapping to publication is 3 years (IQR 2-5 years) (see Figure 1 - Panel A).
+  * The median length of rodent trapping activity is one year (IQR 0-2 years). The median time from completion of rodent trapping to publication is 3 years (IQR 2-5 years) (see Figure 1).
+
+```{r, figure_1, echo=FALSE}
+study_timings <- read_rds(here("plots", "study_timings.rds"))
+
+plot_grid(study_timings)
+
+```
+
+*Figure 1*: Each row represents one of the 126 included studies, green points designate the first year of data collection, blue points designate the end of data collection.
+For studies completed within one year the blue point completely overlies the green.
+Studies with a transparent grey point did not report the year in which trapping was conducted.
+The year of publication is shown by a red point.
 
@@
 The areas with highest trapping activity included the capital cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako) and Ghana (Accra) and the largest cities of Ivory Coast (Abidjan) and Benin (Cotonou).
-Outside of these cities, Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone were the most intensively studied regions (see Figure 1 - Panel B).
+Outside of these cities, Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone were the most intensively studied regions (see Figure 2 - Panel A).
   * Greater numbers of trap nights were conducted in areas with higher population densities (*R~2~* = 0.167).
 In particular Mauritania, Northern Senegal, Eastern Sierra Leone, Eastern Guinea and South West Nigeria were trapped at increased rates than would be expected based on their population densities.
-South West Nigeria, Northern Nigeria, Liberia, the Ivory Coast, Ghana, Niger and Burkina Faso were trapped at lower rates.
+South West Nigeria, Northern Nigeria, Liberia, the Ivory Coast, Ghana, Niger and Burkina Faso were trapped at lower rates (see Figure 2 - Panel B) .
   * No studies reported trap habitats with reference to a standardised habitat classification scheme. 
@@
 
-```{r, figure_1}
+Figure 1 - Panel A - plot of studies, Panel B - location of traps, Panel C - The distribution of population density at trap sites, Panel D- The distribution of habitat types at trap sites compared to all habitat types in West Africa.
 
-cowplot::plot_grid(plotlist = )
-
+```{r, figure_2, layout = "l_page", echo=FALSE}
+knitr::include_graphics(here("figures", "Figure_2.png"))
 ```
 
-Figure 1 - Panel A - plot of studies, Panel B - location of traps, Panel C - The distribution of population density at trap sites, Panel D- The distribution of habitat types at trap sites compared to all habitat types in West Africa
-
-Figure 1: **Panel A:** Each row represents one of the 126 included studies, green points designate the first year of data collection, blue points designate the end of data collection.
-For studies completed within one year the blue point overlies the green.
-Studies with a transparent grey point did not report the year in which trapping was conducted.
-The year of publication is shown by a red point.
-**Panel B:** Map of West Africa, countries where rodent trapping has occurred are mapped to level 2 administrative areas (where available).
+*Figure 2*: **Panel A:** Map of West Africa, countries where rodent trapping has occurred are mapped to level 2 administrative areas (where available).
 These regions are coloured by the total number of trap nights performed at trap sites within their boundaries.
-**Panel C:** Histogram of human population density at 5km^2^ resolution across West Africa and for areas where trapping has occurred.
-**Panel D:** Histogram of habitat types at 1km^2^ resolution across West Africa and for areas where trapping has occurred.
+**Panel B:** (n.b.
+will not include map in final manuscript) The number of trap nights conducted is associated with a regions population density.
+The population density for all regions in West Africa is shown in the yellow box plot, the population density and trap night density for each region is show on the purple scatterplot.
+The line of best fit is a GAM model not incorporating spatial interactions.
+The map panel on the right is the product of the GAM model incorporating spatial interactions.
+**Panel C:** For each of the 10 land cover classes from the ESA dataset we measure the proportion of 300m^2^ pixels within a level 2 administrative area.
+Zoonotic trapping studies occurred in regions over-representative for Cropland, Mosaic landscapes and Shrubland while being under-representative for Bare and Sparse vegetation land cover classes.
 
@@
 
-Figure 2 - Three plots side by side with 6 rows, one each for the most commonly trapped species Panel A - presence/absence from this review, Panel B - presence/absence from GBIF, Panel C - Species range from IUCN
+Figure 3 - Three plots side by side with 6 rows, one each for the most commonly trapped species Panel A - presence/absence from this review, Panel B - presence/absence from GBIF, Panel C - Species range from IUCN
 
-Figure 2 - **Panel A** shows the presence and absence of the 6 most commonly trapped rodent species from the current study.
+Figure 3 - **Panel A** shows the presence and absence of the 6 most commonly trapped rodent species from the current study.
 The perimeter polygon (blue) is the convex hull of the observation points produced from presence only data.
@@
 
-Figure 3 - Presence/absence plots for the following pathogens Arenaviruses with LASV highlighted, Bartonella, Borrelia and Toxoplasma.
+Figure 4 - Presence/absence plots for the following pathogens Arenaviruses with LASV highlighted, Bartonella, Borrelia and Toxoplasma.
 
@@
 
-Figure 4 - Matrix heat plot Y - rodent species/genera, X - pathogen species/genera.
+Figure 5 - Matrix heat plot Y - rodent species/genera, X - pathogen species/genera.
 Colour relates to proportion positive (Perhaps use bivariate colour to also highlight the number of test performed in that species).

### 2021-08-31 scripts/analysis.R

@@ -313,22 +313,22 @@ if(!file.exists(here("data_clean", "pop_tn_analysis.rds"))) {
   write_rds(sites_2, here("data_clean", "traps_level_2_zoonoses.rds"))
   human_pop <- rast(here("data_download", "pop_2005","pop_2005.tif"))
 
-vect_sites <- vect(sites_2)
+  vect_sites <- vect(sites_2)
 
-crop_pop <- crop(human_pop, vect_sites)
-wa_pop <- writeRaster(crop_pop, here("data_download", "pop_2005","wa_pop_2005.tif"))
-region_pop <- terra::extract(crop_pop, vect_sites, fun = "median", method = "simple", na.rm = TRUE, touches = TRUE)
-region_pop_sf <- cbind(vect_sites, region_pop) %>%
-  st_as_sf() %>%
-  st_centroid() %>%
-  mutate(x = st_coordinates(.$geometry)[,1],
-         y = st_coordinates(.$geometry)[,2],
-         tn_density = case_when(is.na(tn_density) ~ 0,
-         TRUE ~ tn_density)) %>%
-  tibble() %>%
-  select(-geometry)
+  crop_pop <- crop(human_pop, vect_sites)
+  wa_pop <- writeRaster(crop_pop, here("data_download", "pop_2005","wa_pop_2005.tif"))
+  region_pop <- terra::extract(crop_pop, vect_sites, fun = "median", method = "simple", na.rm = TRUE, touches = TRUE)
+  region_pop_sf <- cbind(vect_sites, region_pop) %>%
+    st_as_sf() %>%
+    st_centroid() %>%
+    mutate(X = st_coordinates(.$geometry)[,1],
+           Y = st_coordinates(.$geometry)[,2],
+           tn_density = case_when(is.na(tn_density) ~ 0,
+                                  TRUE ~ tn_density)) %>%
+    tibble() %>%
+    select(-geometry)
 
-write_rds(region_pop_sf, here("data_clean", "pop_tn_analysis.rds"))
+  write_rds(region_pop_sf, here("data_clean", "pop_tn_analysis.rds"))
 } else {
   region_pop_sf <- read_rds(here("data_clean", "pop_tn_analysis.rds"))
   human_pop <- rast(here("data_download", "pop_2005","wa_pop_2005.tif"))
@@ -336,7 +336,7 @@ write_rds(region_pop_sf, here("data_clean", "pop_tn_analysis.rds"))
 
[…]

### 2021-08-31 scripts/libraries.R

@@ -12,8 +12,10 @@ pkgs =
     "sf",
     "OpenStreetMap",
     "tmap",
+    "cowplot",
     "maptools",
     "raster",
+    "parzer",
     "terra",
     "mapview",
     "ggmap",
@@ -24,7 +26,9 @@ pkgs =
     "lubridate",
     "gt",
     "gtsummary",
-    "lubridate"
+    "lubridate",
+    "see",
+    "scales"
   )
 pacman::p_load(pkgs, character.only = T)
 

### 2021-08-31 scripts/mapping_traps.R

@@ -194,24 +194,36 @@ population_map <- tm_shape(region_pop_sf) +
 write_rds(population_map, here("plots", "pop_map_2005.rds"))
 tmap_save(population_map, filename = here("figures", "pop_density_2.png"))
 
+non_trapped_regions <- tibble(region_pop_sf) %>%
+  select(-geometry) %>%
+  mutate(tn_density = case_when(is.na(tn_density) ~ 0,
+                                TRUE ~ tn_density)) %>%
+  filter(tn_density == 0)
+
 log_pop_tn <- ggplot(tibble(region_pop_sf) %>%
-         select(-geometry) %>%
-         mutate(tn_density = case_when(is.na(tn_density) ~ 0,
-                                       TRUE ~ tn_density))) +
+                       select(-geometry) %>%
+                       mutate(tn_density = case_when(is.na(tn_density) ~ 0,
+                                                     TRUE ~ tn_density))) +
   geom_smooth(aes(x = tn_density, y = log(pop_2005)),
-              formula = y ~ s(x, bs = "cs")) +
+              formula = y ~ s(x, bs = "cs"),
+              method = "gam") +
   geom_point(data = . %>%
                filter(tn_density > 0),
              aes(x = tn_density, y = log(pop_2005)),
              colour = "#440154") +
-  geom_point(data = . %>%
-               filter(tn_density == 0),
-             aes(x = tn_density, y = log(pop_2005)),
-             colour = "#fde725",
-             alpha = 0.2) +
+  geom_boxplot(non_trapped_regions,
+               mapping = aes(x = 0.01, y = log(pop_2005)),
+               colour = "#fde725",
+               fill = "#fde725",
+               alpha = 0.6,
+               inherit.aes = FALSE,
+               position = position_nudge(x = - 2)) +
   theme_minimal() +
-  labs(x = parse(text = paste("Trap~night~density~per~1000~km^2")),
-       y = parse(text = paste("log~Population~density~per~1~km^2~(2005~level)")))
+  scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
+                labels = trans_format("log10", math_format(10^.x))) +
+  annotation_logticks(sides = "b") +
+  labs(x = parse(text = paste("log~Trap~night~density~per~1000~km^2")),
[…]

### 2021-08-31 reports/manuscript_v2.Rmd
updating report

@@
   * The most commonly trapped genera of rodents were Mastomys sp.
-(27,072, 38.5%), Rattus sp.
-(10,191, 14.5%), Mus sp.
-(8,624, 12.3%), Arvicanthis sp.
-(5,821, 8.3%) and Praomys sp.
-(5,409, 7.7%).
-At species level Mastomys natalensis (11,221, 17.9%), Rattus rattus (8,578, 13.7%), Mastomys erythroleucus (7,379, 11.8%), Mus musculus (6,245, 10%), Arvicanthis niloticus (5,497, 8.8%) and Mastomys huberti (4,699, 7.5%) were the most commonly trapped.
-  * Presence and absence maps for the six most commonly trapped species are compared to GBIF and IUCN data. (Need to do this)
+(27,079, 37.7%), Rattus sp.
+(11,472, 16.1%), Mus sp.
+(8,624, 12.0%), Arvicanthis sp.
+(5,821, 8.1%) and Praomys sp.
+(5,409, 7.5%).
+At species level Mastomys natalensis (11,116, 17.4%), Rattus rattus (9,959, 15.6%), Mastomys erythroleucus (7,386, 11.6%), Mus musculus (6,245, 9.8%), Arvicanthis niloticus (5,497, 8.6%), Mastomys huberti (4,699, 7.4%) and Praomys daltoni (1,854, 2.9%) were the most commonly trapped.
+  * Presence and absence maps for the seven most commonly trapped species are compared to GBIF and IUCN data in Figure 3 Panel A-G. 
   * The number of individuals trapped provides a measure of abundance, the interpretation of this is limited by incomplete reporting on the trapping effort (i.e.
@@
 
-Figure 3 - Three plots side by side with 6 rows, one each for the most commonly trapped species Panel A - presence/absence from this review, Panel B - presence/absence from GBIF, Panel C - Species range from IUCN
+```{r, figure_3, layout = "l_page", echo=FALSE}
+knitr::include_graphics(here("figures", "Figure_3.png"))
+```
 
-Figure 3 - **Panel A** shows the presence and absence of the 6 most commonly trapped rodent species from the current study.
-The perimeter polygon (blue) is the convex hull of the observation points produced from presence only data.
-**Panel B**
+Figure 3 - Three plots side by side with 7 rows, one each for the most commonly trapped species Panel A - presence/absence from this review, Panel B - presence/absence from GBIF, Panel C - Species range from IUCN
+
+Figure 3 - **Panel A-G** shows the presence and absence of the 7 most commonly trapped rodent species from the current study.
 

### 2021-08-31 scripts/manuscript_library.R

@@ -4,6 +4,7 @@ pkgs =
     "tidyverse",
     "gt",
     "gtsummary",
-    "countrycode"
+    "countrycode",
+    "cowplot"
   )
 pacman::p_load(pkgs, character.only = T)

### 2021-09-02 reports/manuscript_v2.Rmd
mapping pathogens

@@
 At species level Mastomys natalensis (11,116, 17.4%), Rattus rattus (9,959, 15.6%), Mastomys erythroleucus (7,386, 11.6%), Mus musculus (6,245, 9.8%), Arvicanthis niloticus (5,497, 8.6%), Mastomys huberti (4,699, 7.4%) and Praomys daltoni (1,854, 2.9%) were the most commonly trapped.
-  * Presence and absence maps for the seven most commonly trapped species are compared to GBIF and IUCN data in Figure 3 Panel A-G. 
-  * The number of individuals trapped provides a measure of abundance, the interpretation of this is limited by incomplete reporting on the trapping effort (i.e.
-number of trap nights) within a study.
-The most abundant species at presence locations were … (Need to do this.
-I will explore the number of individuals trapped by species, I imagine it will show that if a species is detected at a site the most abundant ones will be the synanthropic species such as mus, rattus etc.
-while the more specialised species will be less abundant in areas where they are detected)
+  * Locations of points for the presence or absence of the seven most commonly trapped species are displayed in Figure 3 (left column).
+These are able to identify locations of absence within an individual species documented range.
+Whether an absence represents a "true absence" (i.e.
+not present at this location) or a "pseudo absence" (i.e.
+not detected at this location) should be interpreted with consideration to the trapping effort accompanying the report.
+This is compared to reported presence obtained from GBIF (centre column) and IUCN species range data (left column) in Figure 3.
 
@@
 
-Figure 3 - Three plots side by side with 7 rows, one each for the most commonly trapped species Panel A - presence/absence from this review, Panel B - presence/absence from GBIF, Panel C - Species range from IUCN
-
-Figure 3 - **Panel A-G** shows the presence and absence of the 7 most commonly trapped rodent species from the current study.
+*Figure 3*: Each row corresponds to a single rodent species.
+The column on the left shows the presence and absence of a rodent species from the individual studies included in this review.
+The centre column shows the presence of a rodent species obtained from GBIF (September 2021) for records where longitude and latitude have been provided.
+The right column shows the range of rodent species as proposed by the IUCN (2021) (red shaded area), overlaid are the presence points from both this review and GBIF records.
 

### 2021-09-02 scripts/mapping_pathogens.R

@@ -1,4 +1,4 @@
-source(here::here("scripts", "libraries.r"))
+source(here::here("scripts", "libraries.R"))
 
 studies <- read_rds(here("data_clean", "studies.rds"))
 
@@ -7,19 +7,10 @@ wa_mainland <- c("BEN", "BFA", "CIV", "ESH", "GHA",
                  "NER", "NGA", "SEN", "SLE", "TGO")
 
 level_0 <- read_rds(here("data_download", "admin_spatial", "level_0_admin.rds"))
-list2env(level_0, envir = .GlobalEnv)
-level_0 <- do.call(rbind.SpatialPolygonsDataFrame, level_0)  %>%
-  st_as_sf()
 
 level_1 <- read_rds(here("data_download", "admin_spatial", "level_1_admin.rds"))
-list2env(level_1, envir = .GlobalEnv)
-level_1_all <- do.call(rbind.SpatialPolygonsDataFrame, level_1) %>%
-  st_as_sf()
 
 level_2 <- read_rds(here("data_download", "admin_spatial", "level_2_admin.rds"))
-list2env(level_2, envir = .GlobalEnv)
-level_2_all <- do.call(rbind.SpatialPolygonsDataFrame, level_2) %>%
-  st_as_sf()
 
 pathogen <- read_rds(here("data_clean", "pathogen.rds")) %>%
   filter(iso3c %in% wa_mainland)
@@ -50,7 +41,74 @@ four_paths_wide <- wide_pathogen %>%
   tibble() %>%
   dplyr::select(1:16, matches(four_paths)) %>%
   left_join(., species_data,
-            by = "gbif_id")
+            by = c("gbif_id", "classification")) %>%
+  distinct(record_id, .keep_all = T)
+
+pathogen_map <- function(pathogen_genus) {
+
+  included_countries = level_0 %>%
+    filter(GID_0 %in% wa_mainland)
+
+  pathogen_groups = list(arenaviridae = c("arenaviridae_species", "lassa_mammarenavirus", "mammarenavirus_species"),
+                         borrelia = c("borrelia_species", "borrelia"),
+                         bartonella = c("bartonella_species"),
+                         toxo = c("toxoplasma_gondii"))
+
+  pathogen_data = list(pathogen = four_paths_wide %>%
+                         dplyr::select(1:16, matches(c(pathogen_groups[[pathogen_genus]])),
+                                       genus, species, genus_gbif, species_gbif) %>%
+                         janitor::remove_empty("cols")  %>%
[…]

### 2021-09-02 scripts/mapping_species.R

@@ -17,7 +17,8 @@ studies <- read_rds(here("data_clean", "studies.rds"))
 rodent_spatial <- read_rds(here("data_clean", "rodent_spatial.rds")) %>%
   select(-trap_nights)
 species_data <- read_rds(here("data_clean", "species_data.rds"))
-rodent_iucn <- st_read(here("data_download", "iucn_data", "data_0.shp"))
+rodent_iucn <- st_read(here("data_download", "iucn_data", "data_0.shp")) %>%
+  bind_rows(st_read(here("data_download", "iucn_data", "data_1.shp")))
 
 count_species <- species_data %>%
   filter(iso3c %in% wa_countries) %>%
@@ -53,14 +54,149 @@ top_7_plot <- tm_shape(level_0 %>%
 
 # Figure 3 ----------------------------------------------------------------
 
-m_nat_trap <- rodent_spatial %>%
-  filter(classification == "mastomys natalensis" &
-           iso3c %in% wa_countries) %>%
-  mutate(pres_abs = ifelse(number == 0, "Absent", "Present"),
-         source = "This review")
 
-tm_shape(level_0 %>%
-           filter(GID_0 %in% wa_countries), bbox = afr_bbox) +
-  tm_polygons(alpha = 0, lwd = 1) +
-  tm_shape(m_nat_trap) +
-  tm_dots(col = "pres_abs", palette = "Dark2", title = "", size = .1, shape = 20)
+# M natalensis ------------------------------------------------------------
+
+included_countries <- level_0 %>%
+  filter(GID_0 %in% wa_countries)
+
+plot_species <- function(species_name) {
+
+  species_name = tolower(species_name)
+
+  data <- list(
+    review = rodent_spatial %>%
+      filter(classification == species_name &
+               iso3c %in% wa_countries) %>%
+      mutate(pres_abs = ifelse(number == 0, "Absent", "Present"),
+             source = "This review",
+             x = st_coordinates(.)[,1],
+             y = st_coordinates(.)[,2]),
+    iucn = rodent_iucn %>%
+      filter(BINOMIAL == snakecase::to_sentence_case(species_name)),
+    gbif = read_tsv(here("data_download", "gbif_species", paste0(gsub(" ", "_", species_name), "_gbif.csv"))) %>%
+      select(gbifID, species, countryCode, decimalLatitude, decimalLongitude))
+
+  data$gbif <- data$gbif  %>%
[…]

### 2021-09-03 reports/manuscript_v2.Rmd
v2 for comments

@@
     toc_float: true
-  # word_document
+  word_document:
 knit: (function(inputFile, encoding) {
@@
   * The role of zoonotic infectious diseases in human health and society have been dramatically highlighted through the ongoing SARS-CoV-2 pandemic and recent Ebola virus outbreaks in the Democratic Republic of Congo and Guinea.
-  * The number and intensity of zoonotic spillover events is projected to increase under anthropogenic pressure (e.g.
+  * The number and intensity of zoonotic spillover events are projected to increase under anthropogenic pressure (e.g.
 increased human populations and increasing urbanisation) and global climate change [@morse_prediction_2012].
@@
 This is an example of an ecosystem disservice provided by rodents.
-  * Rodents, however, also provide important ecosystem services within these habitats, including, pest regulation and seed dispersal [@fischer_ecosystem_2018].
-  * Rodents typically demonstrate “fast” life histories [@dobson_fast_2007] and within rodents, these traits are associated with a species being a reservoir of zoonotic diseases [@han_rodent_2015], but see [@albery_fast-lived_2021].
-Further, these traits are prevalent in species that thrive in human dominated landscapes, displacing species that are less likely to be reservoirs of zoonotic disease [@gibb_zoonotic_2020].
+  * Rodents, also provide important ecosystem services within their habitats, including, pest regulation and seed dispersal [@fischer_ecosystem_2018].
+  * Rodents typically demonstrate “fast” life histories [@dobson_fast_2007], these traits are also associated with a species being a reservoir of zoonotic diseases [@han_rodent_2015], but see [@albery_fast-lived_2021].
+Further, these traits are prevalent in species that thrive in human dominated landscapes where they displace species that are less likely to be reservoirs of zoonotic disease [@gibb_zoonotic_2020].
   * West Africa has previously been identified to be at relatively high risk of emerging zoonotic infectious disease events from wildlife populations [@jones_global_2008].
-Ongoing changes in land use and population demographics may lead to an increased burden from zoonotic diseases.
+Ongoing changes in land use and population demographics may lead to an increased burden from zoonotic diseases on human populations in West Africa.
   * Rodents are implicated in the transmission pathways of several diseases in the region, including, Lassa fever, Schistosomiasis and Leptospirosis.
-  * Studies have been conducted in West Africa both to identify novel potential zoonotic pathogens within rodents alongside studies investigating the prevalence and burden of known pathogens within rodent hosts (e.g.
+  * Rodent trapping studies have been conducted in West Africa in order to identify novel potential zoonotic pathogens within rodents and to investigate the prevalence and burden of known pathogens within known rodent hosts (e.g.
 for Lassa fever [@fichet-calvet_diversity_2009] and Schistosomiasis [@catalano_multihost_2020]).
-  * Our understanding of the associations between rodents and their pathogens on broader geographic scales are typically based on global datasets (e.g.
+  * Our understanding of the associations between rodents and their pathogens at broader geographic scales are typically based on global datasets (e.g.
 PREDICTS, GIDEON and GBIF) that consolidate information obtained from multiple sources of information [@gibb_zoonotic_2020; @smith_global_2014; @pigott_mapping_2014].
-These datasets may lack the contextually rich information on the presence, absence and abundance of individual species and the habitats in which they were trapped that is potentially provided by individual rodent trapping studies [@bovendorp_optimising_2017].
-  * Alongside disease specific studies these datasets have been used to identify potential geographic hotspots where virus and host species diversity may be expected to be at its greatest and therefore quantify the risk of a zoonotic disease spillover event occurring at a defined location [@han_global_2016].
-However, there remains the potential for important confounding introduced to these associations through bias generated by study design or through selection of sampling sites.
+These datasets lack the contextually rich information on the presence, absence, abundance and habitats of individual rodent species that is provided by individual rodent trapping studies [@bovendorp_optimising_2017].
+  * These datasets have been used to identify potential geographic hotspots where virus and host species diversity may be expected to be at its greatest and therefore quantify the risk of a zoonotic disease spillover event occurring within a defined region [@han_global_2016].
+However, there remains the potential for important confounding in these spatial distributions through bias generated by study design and selection of sampling sites.
   * For example, systematically increased sampling (e.g.
@@
 periurban landscapes) could lead to an apparent association between locations and risk that is driven by these factors rather than an underlying host and virus association [@wille_how_2021; @gibb_mammal_2021].
-  * Conversely some regions may not be sampled adequately and therefore under-represented in these datasets due to sparse human populations or inaccessible habitats which may lead to erroneously identifying regions as low risk of zoonotic disease spillover events.
-  * Here, we identify rodent trapping studies performed across West Africa and identify the location and habitat types in which they have been conducted, the pathogens assessed and the host-pathogen associations that have been reported in order to quantify the potential bias and identify areas requiring further focussed investigation.
+  * Conversely some regions may not be sampled adequately and therefore under-represented in these datasets due to sparse human populations or inaccessible habitats which may lead to proposing that these regions are at low risk of zoonotic disease spillover events.
+  * Here, we identify rodent trapping studies performed across West Africa and identify the location and habitat types in which they have been conducted, the pathogens assessed and the host-pathogen associations that have been reported in order to quantify the potential bias and to identify regions requiring further focussed investigation.
 
@@
 
-  * We conducted a search in Ovid MEDLINE, Web of Science (Core collection and Zoological Record), JSTOR, BioOne, African Journals Online, Global Health and the pre-print servers, BioRxiv and EcoEvoRxiv for the following terms as keywords, no date constraints were used:
+  * We conducted a search in Ovid MEDLINE, Web of Science (Core collection and Zoological Record), JSTOR, BioOne, African Journals Online, Global Health and the pre-print servers, BioRxiv and EcoEvoRxiv for the following terms as keywords, no date limits were set:
 
@@
 
-  * We extracted data from eligible studies to input into a Google sheets document.
+  * We extracted data from eligible studies into a Google sheets document.
 Extracted variables included i) study identifiers; ii) study aims; iii) trapping methodology; iv) geolocation data; v) method of speciation; vi) trapping locations and dates; vii) trapped species; viii) number of trap-nights and ix) microorganisms/pathogens of interest.
@@
 trap, trap-line, study site or study region).
-We extracted coordinates in the format reported, subsequently converting them to decimal degrees.
+We extracted coordinates in the format reported and converted them to decimal degrees.
   * We recorded the habitat classification scheme a study used (e.g.
@@
 ### Rodent presence, absence, abundance
-  * We mapped genus and species names to the species names used in the GLobal Biodiversity Information Facility (GBIF) taxonomy  [@gbif_the_global_biodiversity_information_facility_global_2021]
+  * We mapped genus and species names to the species names used in the Global Biodiversity Information Facility (GBIF) taxonomy  [@gbif_the_global_biodiversity_information_facility_global_2021]
   * We extracted information on the presence, absence and number of trapped individuals.
@@
 ### Pathogen presence and absence
-  * We extracted data on all pathogens assayed for in studies investigating rodents for potential zoonoses.
+  * We extracted data on all pathogens assayed in studies investigating rodents for potential zoonoses.
 The number of rodents tested and the number of positive or negative samples were recorded alongside the type of assay used (e.g.
@@
 Where possible, pathogens were identified to species level.
-However, where an assay only allowed for attribution to a family of viruses or bacteria, the higher order grouping was used (i.e.
-PCR using a non-specific arenavirus primer).
+However, where an assay only allowed for attribution to a family of viruses or bacteria, the higher order grouping was used (e.g.
+Arenaviridae for a PCR using a non-specific arenavirus primer).
 
@@
   * We summarised the number of studies, the year in which trapping occurred and the country in which they were conducted.
-In studies not reporting the year of trapping we used the year of publication as a proxy.
   * We used the GPS coordinates of single trap, trapping grid/line or study site and the number of trap nights to calculate trapping effort (trap night density) within level 2 administrative areas in West Africa.
   * For studies not reporting the number of trap nights we imputed the number based on the median trap success rate from studies which reported the number of trap nights (matched to building or non-building based study sites).
-The median trap success rate for all rodents at a defined trap site was calculated for both trap sites including built environments and non-built environments.
-The number of rodents trapped at a trap site was then used in combination with these values of trap success to impute the number of trap nights for trap sites with no reported trap nights.
+The median trap success rate for all rodents at a defined trap site was calculated separately for trap sites which included built environments and non-built environments.
+The number of rodents trapped at a trap site was then used in combination with these two values of trap success to impute the number of trap nights for trap sites with no reported trap nights.
   * We summarised the habitat types of trap sites based on information reported in the study.
-  * For the subset of studies investigating rodent zoonoses we compared the location of trapping sites with 2005 SEDAC Global Population Density estimates, the median year studies were commenced, using a Generalised Additive Model (GAM) incorporating a spatial interaction term to investigate the association of number of trap nights and human population density.
+  * For the subset of studies investigating rodent zoonoses we compared the location of trapping sites with SEDAC Global Population Density estimates for 2005, the median year studies were commenced.
+We used a Generalised Additive Model (GAM) incorporating a spatial interaction term to investigate the association of number of trap nights and human population density.
 The model structure was specified as below:
-    * *trap night density* ~ *Tweedie*(*log*(Population density (2005)) + (Longitude * Latitude))   
-  * We obtained land cover classifications from the European Space Agency Copernicus dataset at 300m^2^ spatial resolution (2005) we extracted the proportion of land cover classes within all regions of West Africa and all regions in which rodent trapping for zoonotic diseases.
+    * *trap night density* ~ *Tweedie*(log(Population density (2005)) + (Longitude * Latitude))   
+  * We obtained land cover classifications from the European Space Agency Copernicus dataset at 300m^2^ spatial resolution (2005) we extracted the proportion of land cover classes within all regions of West Africa and all regions in which rodent trapping occurred for investigation of zoonotic diseases.
 We compared the proportion of possible land cover classes with those of regions where trapping occurred.
@@
 
-  * We summarised the presence and absence of pathogens, the assays used, their host species and the locations from where the samples were obtained.
+  * We summarised the presence and absence of microorganisms, the assays used, their host species and the locations from where the samples were obtained.
   * We investigated the association between rodent species and the detection of potential pathogens. We report proportion of positive and negative tests for each species and pathogen pair.
@@
 
-  * 4,282 citations were identified, with 126 rodent trapping studies included in narrative synthesis.
-  * The earliest included studies were from 1974 with increasing numbers of studies being performed annually since 2005.
-  * The median length of rodent trapping activity is one year (IQR 0-2 years). The median time from completion of rodent trapping to publication is 3 years (IQR 2-5 years) (see Figure 1).
+  * 4,282 relevant citations were identified, with 126 rodent trapping studies included in narrative synthesis.
+  * The earliest included studies were published in 1974 with increasing numbers of studies being performed annually since 2005.
+  * The median length of rodent trapping activity was one year (IQR 0-2 years). The median time from completion of rodent trapping to publication is 3 years (IQR 2-5 years) (see Figure 1).
 
-```{r, figure_1, echo=FALSE}
+```{r, figure_1, layout = "l_page", echo=FALSE}
 study_timings <- read_rds(here("plots", "study_timings.rds"))
@@
 
-  * Rodent trapping took place across 1,331 trap sites with at least one trap site recorded from 14 West African countries. No rodent trapping studies were identified from Gambia or Togo.
+  * Rodent trapping took place at 1,331 trap sites with at least one trap site recorded from 14 West African countries. No rodent trapping studies were identified from Gambia or Togo.
   * Thirty-one (25%) studies reported trapping at a single study site, 46 (37%) studies trapped at between two and five study sites, the remaining 49 studies trapped at between six and 93 study sites.
 Trap sites were situated in 273 (19.3%) of 1414 level 2 administrative regions from the 14 West African nations.
-The areas with highest trapping activity included the capital cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako) and Ghana (Accra) and the largest cities of Ivory Coast (Abidjan) and Benin (Cotonou).
+The areas with highest trapping activity included the capital cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako) and Ghana (Accra) and the largest cities of Côte d'Ivoire (Abidjan) and Benin (Cotonou).
 Outside of these cities, Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone were the most intensively studied regions (see Figure 2 - Panel A).
@@
 In particular Mauritania, Northern Senegal, Eastern Sierra Leone, Eastern Guinea and South West Nigeria were trapped at increased rates than would be expected based on their population densities.
-South West Nigeria, Northern Nigeria, Liberia, the Ivory Coast, Ghana, Niger and Burkina Faso were trapped at lower rates (see Figure 2 - Panel B) .
+South West Nigeria, Northern Nigeria, Liberia, Côte d'Ivoire, Ghana, Niger and Burkina Faso were trapped at lower rates (see Figure 2 - Panel B) .
   * No studies reported trap habitats with reference to a standardised habitat classification scheme. 
@@
 The remote sensing data will be matched to the trap dates there will be several years between the trapping and data this is compared to it seems impractical to try and get annual population estimates or land classification so this will be chosen pragmatically)
-
-Figure 1 - Panel A - plot of studies, Panel B - location of traps, Panel C - The distribution of population density at trap sites, Panel D- The distribution of habitat types at trap sites compared to all habitat types in West Africa.
 
@@
 The remaining 34 species came from the orders of the Soricomorpha (30), Erinaceomorpha (2) and Afrosoricida (1).
-  * The most commonly trapped genera of rodents were Mastomys sp.
-(27,079, 37.7%), Rattus sp.
-(11,472, 16.1%), Mus sp.
-(8,624, 12.0%), Arvicanthis sp.
-(5,821, 8.1%) and Praomys sp.
-(5,409, 7.5%).
-At species level Mastomys natalensis (11,116, 17.4%), Rattus rattus (9,959, 15.6%), Mastomys erythroleucus (7,386, 11.6%), Mus musculus (6,245, 9.8%), Arvicanthis niloticus (5,497, 8.6%), Mastomys huberti (4,699, 7.4%) and Praomys daltoni (1,854, 2.9%) were the most commonly trapped.
+  * The most commonly trapped genera of rodents were *Mastomys sp.* (27,079, 37.7%), *Rattus sp.* (11,472, 16.1%), *Mus sp.* (8,624, 12.0%), *Arvicanthis sp.* (5,821, 8.1%) and *Praomys sp.* (5,409, 7.5%).
+At species level *Mastomys natalensis* (11,116, 17.4%), *Rattus rattus* (9,959, 15.6%), *Mastomys erythroleucus* (7,386, 11.6%), *Mus musculus* (6,245, 9.8%), *Arvicanthis niloticus* (5,497, 8.6%), *Mastomys huberti* (4,699, 7.4%) and Praomys daltoni (1,854, 2.9%) were the most commonly trapped.
   * Locations of points for the presence or absence of the seven most commonly trapped species are displayed in Figure 3 (left column).
@@
 not present at this location) or a "pseudo absence" (i.e.
-not detected at this location) should be interpreted with consideration to the trapping effort accompanying the report.
+not detected at this location) should be interpreted with consideration to the trapping effort at the site.
 This is compared to reported presence obtained from GBIF (centre column) and IUCN species range data (left column) in Figure 3.
@@
   * Seven studies solely investigated pathogens of rodents, including Hydatigera species (previously Taenia species) and Trichuris species.
-  * Fifty-five studies investigated organisms that were potentially zoonotic pathogens.
+  * Fifty-five studies investigated microorganisms that were potentially zoonotic pathogens.
 Thirty-two microorganisms were tested for, 8 of these were defined at species level, with the remaining 24 at levels of higher taxonomic classification.
@@
 Three studies performed direct culture of Lassa mammarenavirus or Leishmania species to detect the presence of the pathogen in 643 samples.
-  * The most common pathogens assessed for using PCR were Lassa mammarenavirus or other Arenaviridae (31%) the bacteria Borrelia sp.
-(11.3%) and Bartonella sp.
-(6.5%) followed by Usutu virus (6.4%) and Hantaviridae (5.6%).
-The most common pathogens assessed for with serology were Lassa mammarenavirus or other Arenaviridae (78%), Toxoplasma gondii (10.6%), Borrelia sp.
-(6.3%) and Leptospirosis sp.
-(2.2%).
-The most common pathogens assessed for with histology or direct visualisation were Borrelia sp.
-(48.7%), Schistosoma sp.
+  * The most common microorganisms assessed for using PCR were Lassa mammarenavirus or other Arenaviridae (31%) the bacteria *Borrelia sp.* (11.3%) and *Bartonella sp.* (6.5%), followed by Usutu virus (6.4%) and Hantaviridae (5.6%).
+The most common pathogens assessed for with serology were *Lassa mammarenavirus* or other Arenaviridae (78%), *Toxoplasma gondii* (10.6%), *Borrelia sp.* (6.3%) and *Leptospirosis sp.* (2.2%).
+The most common pathogens assessed for with histology or direct visualisation were *Borrelia sp.* (48.7%), *Schistosoma sp*.
 (20.4%) and other parasites.
-All studies using direct culture were investigating either Lassa mammarenavirus (81%) or Leishmania sp.
-(19%).
-  * Most studies tested for a single pathogen (39), with 16 studies testing for two or more pathogens.
-The most frequently tested for pathogens were Lassa mammarenavirus (28%) or members of the Arenaviridae family (16%), the spirochete bacteria Borrelia sp.
-was investigated in 8 studies, Bartonella sp.
-and Toxoplasma gondii were investigated in 4 each, the remaining 25 pathogens were reported in three or fewer studies.
-Define presence and absence of pathogens by country (Need to do)
+All studies using direct culture were investigating either *Lassa mammarenavirus* (81%) or *Leishmania sp.* (19%).
+  * Most studies tested for a single microorganisms (39), with 16 studies testing for two or more microorganisms.
+The most frequently tested for microorganisms were *Lassa mammarenavirus* (28%) or members of the Arenaviridae family (16%), the spirochete bacteria *Borrelia sp.* was investigated in 8 studies, *Bartonella sp.* and *Toxoplasma gondii* were investigated in 4 each, the remaining 25 microorganisms were reported in three or fewer studies.
+  * There is spatial heterogeneity in the location of trapped rodents that were tested for the four most commonly assayed microorganisms. 
+  * The presence and absence of Arenaviridae, predominantly *Lassa mammarenavirus*, in rodents was assessed in Guinea, Sierra Leone, Mali, Côte d'Ivoire, Benin and Nigeria.
+The known reservoir species of this zoonotic pathogen *Mastomys natalensis* was additionally detected in Senegal, Guinea-Bissau, Burkina-Faso and Niger with likely presence in Liberia and Togo.
+Arenaviridae were detected in rodents from 17 of 54 species or genera assayed from Guinea, Sierra Leone, Mali, Côte d'Ivoire, Ghana and Nigeria, no positive rodents were detected in Benin (Figure 4 top row).
+  * The presence and absence of *Borrelia sp.* was investigated in Senegal, Mauritania, Mali, Niger and Benin.
+Microorganisms of this genera were detected from samples in Mali with negative samples obtained from Senegal, Mauritania, Niger and Benin (Figure 4 second row).
+*Borrelia sp.* were detected in 8 of 44 species or genera of rodents assayed.
+  * The presence and absence of *Bartonella sp.* was investigated in Senegal, Mali and Nigeria with positive samples obtained at most sites from all three countries (Figure 4 third row).
+*Bartonella sp.* were detected in 11 of 11 specis or genera of rodents assayed.
+  * The presence of *Toxoplasma gondii* was only investigated in Senegal with relatively high proportions of positive samples at all sites (Figure 4 fourth row).
+*Toxoplasma gondii* was detected in 6 of 8 rodent species assayed.
 
-Figure 4 - Presence/absence plots for the following pathogens Arenaviruses with LASV highlighted, Bartonella, Borrelia and Toxoplasma.
+```{r, figure_4, layout = "l-screen-inset", echo=FALSE}
+
+knitr::include_graphics(here("figures", "Figure_4a.png"))
+knitr::include_graphics(here("figures", "Figure_4b.png"))
+knitr::include_graphics(here("figures", "Figure_4_legend.png"))
+
+```
+
+*Figure 4*: Presence/absence plots at each unique trapping site for the four most commonly assayed microorganisms Arenaviridae (top), *Bartonella sp.* (second row), *Borrelia sp.* (third row) and *Toxoplasma gondii* (fourth row).
+The tables to the right of each map highlight the 10 (or number applicable) most commonly positive and tested rodent species and genera assayed.
 
@@
   * Ninety-seven rodent species were investigated for the presence of a zoonotic pathogen. 42,940 assays were performed on 32,014 individual rodents. 
-  * The rodent species most commonly assayed for zoonotic pathogens were, Rattus rattus (n = 2,977) assessed for 24 pathogens, Mus musculus (n = 3,402) for 23, Mastomys natalensis (n = 7,189) and Mastomys erythroleucus (n = 3,013) both 19 and Arvicanthis niloticus (n = 3,840) for 18.
+  * The rodent species most commonly assayed for zoonotic pathogens were, *Rattus rattus* (n = 2,977) assessed for 24 pathogens, *Mus musculus* (n = 3,402) for 23, *Mastomys natalensis* (n = 7,189) and *Mastomys erythroleucus* (n = 3,013) both 19 and *Arvicanthis niloticus* (n = 3,840) for 18.
   * All remaining species were investigated for 10 or fewer pathogens.
-  * Lassa mammarenavirus was detected in 414 Mastomys natalensis, 9 Mastomys erythroleucus, 5 Hylomyscus pamfi and 3 Mus baoulei through PCR.
-In serological assays, evidence of prior infection or acquisition of maternal antibodies to Lassa mammarenavirus was found in 465 Mastomys natalensis, 361 Mastomys sp., 6 Mus musculus and 4 Rattus rattus.
-From direct viral culture 19 Mastomys natalensis, 2 Rattus rattus and 1 Mus minutoides were found to be infected with Lassa mammarenavirus.
-Bartonella sp.
-were detected in; 113 Mastomys natalensis, 50 Mastomys erythroleucus, 39 Rattus norvegicus, 16 Rattus rattus and 13 individuals of 7 further species through PCR.
-  * Borrelia sp.
-infections were identified through PCR in 110 individuals; 37 Mastomys erythroleucus, 12 Mus musculus, 11 Mastomys huberti and Praomys daltoni., tThe remaining 39 positive results were from 9 further species.
-In serological assays, 47 Mastomys natalensis, 12 Mastomys erythroleucus and 8 Praomys daltoni were found to be positive, the remaining 11 positive individuals came from three further species.
-Most Borrelia sp.
-infections were identified through histology or direct visualisation with, 243 Arvicanthis niloticus, 102 Mastomys natalensis, and 78 Mastomys huberti along with a further 64 individuals from 8 species infected.
-  * Finally, Toxoplasma gondii was detected via PCR in 89 Mus musculus, 13 Cricetomys gambianus, 5 Crocidura olivieri and 3 Rattus rattus.
-From serological assays evidence of infection was found in 25 Mus musculus, 46 Mastomys erythroleucus, 4 Rattus rattus, and Mastomys natalensis, using histology and direct visualisation 1 Mus musculus and 1 Rattus rattus were also found to be infected.
+  * *Lassa mammarenavirus* was detected in 414 *Mastomys natalensis*, 9 *Mastomys erythroleucus*, 5 *Hylomyscus pamfi* and 3 *Mus baoulei* through PCR.
+In serological assays, evidence of prior infection or acquisition of maternal antibodies to *Lassa mammarenavirus* was detected in 465 *Mastomys natalensis*, 361 *Mastomys sp.*, 6 *Mus musculus* and 4 *Rattus rattus*.
+From direct viral culture 19 *Mastomys natalensis*, 2 *Rattus rattus* and 1 *Mus minutoides* were found to be infected with *Lassa mammarenavirus*.
+*Bartonella sp.* were detected in; 113 *Mastomys natalensis*, 50 *Mastomys erythroleucus*, 39 *Rattus norvegicus*, 16 *Rattus rattus* and 13 individuals of 7 further species through PCR.
+  * *Borrelia sp.* infections were identified through PCR in 110 individuals; 37 *Mastomys erythroleucus*, 12 *Mus musculus*, 11 *Mastomys huberti* and *Praomys daltoni*.
+The remaining 39 positive results were from 9 further species.
+In serological assays, 47 *Mastomys natalensis*, 12 *Mastomys erythroleucus* and 8 *Praomys daltoni* were positive, with the remaining 11 positive individuals from three further species.
+Most *Borrelia sp.* infections were identified through histology or direct visualisation with, 243 *Arvicanthis niloticus*, 102 *Mastomys natalensis*, and 78 *Mastomys huberti* along with a further 64 individuals from 8 species infected.
+  * Finally, *Toxoplasma gondii* was detected through PCR in 89 Mus musculus, 13 *Cricetomys gambianus*, 5 *Crocidura olivieri* and 3 *Rattus rattus*.
+From serological assays evidence of prior infection was found in 25 *Mus musculus*, 46 *Mastomys erythroleucus*, 4 *Rattus rattus* and *Mastomys natalensis*.
+A single *Mus musculus* and *Rattus rattus* were found to be infected using histology and direct visualisation.
   * Thirty-eight species of Rodentia, 3 species of Soricomorpha and 2 species of Erinaceomorpha that were tested for potential pathogens had entirely negative results.
+
+```{r figure_5, layout = "l_page", echo=FALSE}
+
+knitr::include_graphics(here("figures", "Figure_5.png"))
+
+```
 

### 2021-09-03 scripts/analysis.R

@@ -757,6 +757,142 @@ names(pathogen_groups) <- tested %>%
   distinct(pathogen_tested) %>%
   pull()
 
+
+# Top four pathogens, rodents tested --------------------------------------
+four_paths <- vctrs::vec_c(arenaviridae, borrelia, bartonella, toxoplasma)
+
+four_paths_wide <- wide_pathogen %>%
+  tibble() %>%
+  dplyr::select(1:16, matches(four_paths)) %>%
+  left_join(., species_data,
+            by = c("gbif_id", "classification")) %>%
+  distinct(record_id, .keep_all = T)
+
+pathogen_table <- function(pathogen_genus) {
+  pathogen_groups = list(arenaviridae = c("arenaviridae_species", "lassa_mammarenavirus", "mammarenavirus_species"),
+                         borrelia = c("borrelia_species", "borrelia"),
+                         bartonella = c("bartonella_species"),
+                         toxoplasma = c("toxoplasma_gondii"))
+
+  four_paths_wide %>%
+    dplyr::select(1:16, matches(c(pathogen_groups[[pathogen_genus]])),
+                  genus, species, genus_gbif, species_gbif) %>%
+    janitor::remove_empty("cols")  %>%
+    mutate(number_tested = rowSums(.[grep("tested", names(.))], na.rm = T)) %>%
+    filter(number_tested != 0) %>%
+    mutate(pcr_positive = rowSums(.[grep("pcr", names(.))], na.rm = T),
+           ab_ag_positive = rowSums(.[grep("ab_ag", names(.))], na.rm = T),
+           culture_positive = rowSums(.[grep("culture", names(.))], na.rm = T),
+           pos_neg = case_when(pcr_positive + ab_ag_positive + culture_positive > 0 ~ "Positive",
+                               TRUE ~ "Negative")) %>%
+    dplyr::select(1:15, all_of(c("number_tested", "pcr_positive", "ab_ag_positive", "culture_positive", "pos_neg")),
+                  genus, species, genus_gbif, species_gbif) %>%
+    tibble() %>%
+    mutate(genus = snakecase::to_sentence_case(genus),
+           `Species` = snakecase::to_sentence_case(classification),
+           pos_neg = case_when(pos_neg == "Positive" ~ 1,
+                               TRUE ~ 0)) %>%
+    group_by(genus, `Species`) %>%
[…]

### 2021-09-03 scripts/mapping_traps.R

@@ -138,19 +138,51 @@ level_2_sites <- level_2 %>%
          tn_density = region_trap_nights/(as.numeric(area_m2)/1000000),
          tn_density = ifelse(is.na(tn_density), NA, tn_density))
 
-trap_night_density_level2 <- tm_shape(level_2_sites) +
-  tm_polygons(col = "tn_density", style = "fixed",
-              breaks = c(0, 0.001, 0.01, 0.1, 1, 10, 100, 200),
-              legend.format = list(fun = function(x) paste0(formatC(x, drop0trailing = TRUE))),
-              palette = "-viridis", colorNA = NULL, border.alpha = 1, border.col = "grey", lwd = 0.1,
-              title = parse(text = paste("Density~of~trap~nights~per~1000~km^2"))) +
-  tm_shape(level_0 %>%
-             filter(GID_0 %in% wa_countries)) +  tm_polygons(alpha = 0, lwd = 1) +
-  tm_compass(type = "arrow", size = 1, position = c("left", "bottom")) +
-  tm_scale_bar(position = c("left", "bottom")) +
-  tm_graticules(labels.show = TRUE,
-                lwd = 0.5,
-                alpha = 0.5)
+ggplot() +
+  geom_sf(data = included_countries, fill = "#808080", alpha = 0.1, lwd = 0.1) +
+  geom_point(data = data$review %>%
+               filter(pres_abs == "Present"), mapping = aes(colour = pres_abs, x = x, y = y), size = .5,
+             crs = 4326) +
+  geom_spatial_point(data = anti_join(tibble(data$review), tibble(data$review) %>%
+                                        filter(pres_abs == "Present"),
+                                      by = c("x", "y")),
+                     mapping = aes(colour = pres_abs, x = x, y = y), size = .5,
+                     crs = 4326) +
+  scale_colour_manual(values = c("#ff8c00", "#440154")) +
+  labs(colour = "",
+       title = paste0(snakecase::to_sentence_case(species_name), " - this review"),
+       x = element_blank(),
+       y = element_blank()) +
+  annotation_north_arrow(height = unit(1, "cm"),
+                         style = north_arrow_minimal(text_size = 8)) +
+  annotation_scale(height = unit(0.1, "cm"),
+                   location = "tr") +
+  theme_minimal()
+
+level_2_sites$breaks <- cut(level_2_sites$tn_density,
[…]

### 2021-09-05 reports/manuscript_v2.Rmd
improving fig. 5

@@
     toc_float: true
-  word_document:
+  # word_document:
 knit: (function(inputFile, encoding) {
@@
 ```{r, figure_1, layout = "l_page", echo=FALSE}
-study_timings <- read_rds(here("plots", "study_timings.rds"))
 
-plot_grid(study_timings)
+knitr::include_graphics(here("figures", "Fig_1_Panel_A.png"))
 
@@
 
-```{r, figure_2, layout = "l_page", echo=FALSE}
+```{r, figure_2, layout = "l-screen-inset", echo=FALSE}
 knitr::include_graphics(here("figures", "Figure_2.png"))
@@
 
-```{r, figure_3, layout = "l_page", echo=FALSE}
+```{r, figure_3, layout = "l-screen-inset", echo=FALSE}
 knitr::include_graphics(here("figures", "Figure_3.png"))
@@
 
-```{r figure_5, layout = "l_page", echo=FALSE}
+```{r figure_5, layout ="l-screen-inset", echo=FALSE}
 
@@
 
-Figure 5 - Matrix heat plot Y - rodent species/genera, X - pathogen species/genera.
-Colour relates to proportion positive (Perhaps use bivariate colour to also highlight the number of test performed in that species).
+*Figure 5*: The percentage of assays from each included species and microorganism tested for is represented by colour.
+The size of a point relates to the squareroot of the number of individuals assayed.
+The plot on the left is for species/genera assayed for four or more microorganisms, the plot on rights are for species/genera assayed for 3 or less microorganisms.
 

### 2021-09-27 reports/manuscript_v2.Rmd
update manuscript

--- 2021-09-03 reports/manuscript_v2.Rmd
@@
 ---
-title: "Rodent trapping studies to understand the prevalence of zoonotic disease in West Africa: A scoping review."
+title: " A scoping review of rodent trapping studies to understand the prevalence of zoonotic disease in West Africa."
 description: |
@@
 
+# Abstract
+
+Rodents are important reservoirs of zoonotic infectious diseases; 255 species are known hosts of 85 zoonotic pathogens.
+These reservoir species are globally distributed with West Africa containing multiple known and several further predicted reservoir species.
+Typically International Union for Conservation of Nature (IUCN) species distribution maps and Global Biodiversity Information Facility (GBIF) presence maps are used to determine regions at risk for zoonotic spillover events, however, these datasets are biased by incomplete sampling.
+To investigate the impact of rodent species sampling heterogeneity on zoonotic spillover risk we have systematically reviewed rodent trapping studies to produce a contextually rich dataset that can be used to explore this risk at finer spatial resolution.
+Here we show that sampling of rodents and their pathogens in the West African region are spatially biased by country and trapping habitat.
+We found that rodent trapping effort was associated with regional population density and was increased in habitats modified by human activity.
+We produce updated rodent species ranges compared to available IUCN maps and enrich GBIF data by including locations of rodent species absence and trapping effort.
+Furthermore, we report the spatial bias in the investigation of four important rodent zoonoses, Arenaviridae, *Borellia sp.*, *Bartonella sp.* and *Toxoplasma sp.*.
+Our results highlight that incorporating sampling bias is important when assessing the risk for zoonotic spillover events from rodents.
+The synthesis of contextually rich rodent trapping data contributes important information that is lacking in IUCN distribution maps and GBIF species presence data.
+We anticipate this dataset can support the production of more complete spatial risk assessments of zoonotic spillover events.
+For example, the inclusion of absence data and trapping effort can help identify regions where data sparsity may produce inappropriately predicted risk.
+Furthermore, this data specifies regions in need of comprehensive study for four targeted rodent zoonoses, Arenaviridae, *Borellia sp.*, *Bartonella sp.* and *Toxoplasma sp.*.
+
 # Introduction
 
-  * The role of zoonotic infectious diseases in human health and society have been dramatically highlighted through the ongoing SARS-CoV-2 pandemic and recent Ebola virus outbreaks in the Democratic Republic of Congo and Guinea.
-  * The number and intensity of zoonotic spillover events are projected to increase under anthropogenic pressure (e.g.
-increased human populations and increasing urbanisation) and global climate change [@morse_prediction_2012].
-  * Rodents are abundant and diverse [@burgin_how_2018], their widespread occurrence and proximity to human activity make the understanding of the effect of environmental change on their host-pathogen interactions vitally important.
-  * Rodents are important hosts of zoonotic disease.
-Of 2,220 extant rodent species, 244 (10.7%) are described as being reservoirs of 85 zoonotic pathogens [@han_global_2016].
-This is an example of an ecosystem disservice provided by rodents.
-  * Rodents, also provide important ecosystem services within their habitats, including, pest regulation and seed dispersal [@fischer_ecosystem_2018].
-  * Rodents typically demonstrate “fast” life histories [@dobson_fast_2007], these traits are also associated with a species being a reservoir of zoonotic diseases [@han_rodent_2015], but see [@albery_fast-lived_2021].
+The potential effect of zoonotic infectious diseases outbreaks on human health and societies has been dramatically highlighted through the ongoing SARS-CoV-2 pandemic and recent Ebola virus outbreaks.
+The number of zoonotic spillover events are projected to increase under intensifying anthropogenic pressure such as, increased human populations [@Allen], increasing urbanisation [@Hassell], and global climate change [@morse_prediction_2012].
+In addition, two taxa are proposed to contribute to the highest burden of potential zoonoses globally, rodents (Rodentia) and bats (Chiroptera) [@han].
+Of 2,220 extant rodent species, 244 (10.7%) are described as being reservoirs of 85 zoonotic pathogens [@han].
+Specifically, West Africa has been identified as region at increased hazard for rodent-borne zoonotic spillover events under different projected scenarios [@garcia-pena_land-use_2021].
+Rodents are implicated in the transmission pathways of several diseases in this region, including, Lassa fever, Schistosomiasis and Leptospirosis.
+
+Rodent species form diverse assemblages providing important ecosystem services such as pest regulation and seed dispersal [@fischer_ecosystem_2018].
+However, the role a minority of species play in zoonotic infectious disease spillover can be considered an ecosystem disservice.
+Rodents typically demonstrate “fast” life histories [@dobson_fast_2007], these traits are also associated with a species being a reservoir of zoonotic pathogens [@han_rodent_2015; @albery_fast-lived_2021].
 Further, these traits are prevalent in species that thrive in human dominated landscapes where they displace species that are less likely to be reservoirs of zoonotic disease [@gibb_zoonotic_2020].
-  * West Africa has previously been identified to be at relatively high risk of emerging zoonotic infectious disease events from wildlife populations [@jones_global_2008].
-Ongoing changes in land use and population demographics may lead to an increased burden from zoonotic diseases on human populations in West Africa.
-  * Rodents are implicated in the transmission pathways of several diseases in the region, including, Lassa fever, Schistosomiasis and Leptospirosis.
-  * Rodent trapping studies have been conducted in West Africa in order to identify novel potential zoonotic pathogens within rodents and to investigate the prevalence and burden of known pathogens within known rodent hosts (e.g.
+The widespread occurrence of reservoir species and their proximity to human activity make the description of rodent species assemblages and host-pathogen interactions vitally important.
+
+Rodent trapping studies provide a useful method to describe rodent population structures and survey for potential zoonoses.
+Studies have been conducted in West Africa to both identify novel potential zoonotic pathogens within rodents [@predicts] and to investigate the prevalence and burden of known pathogens within known rodent hosts (e.g.
 for Lassa fever [@fichet-calvet_diversity_2009] and Schistosomiasis [@catalano_multihost_2020]).
-  * Our understanding of the associations between rodents and their pathogens at broader geographic scales are typically based on global datasets (e.g.
-PREDICTS, GIDEON and GBIF) that consolidate information obtained from multiple sources of information [@gibb_zoonotic_2020; @smith_global_2014; @pigott_mapping_2014].
-These datasets lack the contextually rich information on the presence, absence, abundance and habitats of individual rodent species that is provided by individual rodent trapping studies [@bovendorp_optimising_2017].
-  * These datasets have been used to identify potential geographic hotspots where virus and host species diversity may be expected to be at its greatest and therefore quantify the risk of a zoonotic disease spillover event occurring within a defined region [@han_global_2016].
+These studies provide contextually rich information on when, where and under what conditions rodents were trapped that are typically missing from the global datasets [@bovendorp_optimising_2017].
+Despite this, these global datasets (e.g.
+IUCN, GBIF and GIDEON) are typically used to inform species distribution maps and zoonotic disease spillover predictions [@han_global_2016; @smith_global_2014; @pigott_mapping_2014].
+.
+These studies have been used to identify potential geographic hotspots where virus and host species diversity may be expected to be at its greatest to predict regional zoonotic disease spillover risks.
 However, there remains the potential for important confounding in these spatial distributions through bias generated by study design and selection of sampling sites.
-  * For example, systematically increased sampling (e.g.
+For example, systematically increased sampling (e.g.
 more intensive studies over longer time periods) or over-representation of certain habitats (e.g.
 periurban landscapes) could lead to an apparent association between locations and risk that is driven by these factors rather than an underlying host and virus association [@wille_how_2021; @gibb_mammal_2021].
-  * Conversely some regions may not be sampled adequately and therefore under-represented in these datasets due to sparse human populations or inaccessible habitats which may lead to proposing that these regions are at low risk of zoonotic disease spillover events.
-  * Here, we identify rodent trapping studies performed across West Africa and identify the location and habitat types in which they have been conducted, the pathogens assessed and the host-pathogen associations that have been reported in order to quantify the potential bias and to identify regions requiring further focussed investigation.
+Conversely some regions may not be sampled adequately and therefore under-represented in these datasets due to sparse human populations or inaccessible habitats which may lead to proposing that these regions are at low risk of zoonotic disease spillover events.
+
+Here, we identify rodent trapping studies performed across West Africa and identify the location and habitat types in which they have been conducted, the pathogens assessed and the host-pathogen associations that have been reported in order to quantify the potential bias and to identify regions requiring further focussed investigation.
 
@@
 
-  * We conducted a search in Ovid MEDLINE, Web of Science (Core collection and Zoological Record), JSTOR, BioOne, African Journals Online, Global Health and the pre-print servers, BioRxiv and EcoEvoRxiv for the following terms as keywords, no date limits were set:
+We conducted a search in Ovid MEDLINE, Web of Science (Core collection and Zoological Record), JSTOR, BioOne, African Journals Online, Global Health and the pre-print servers, BioRxiv and EcoEvoRxiv for the following terms as keywords, no date limits were set:
 
@@
 
-  * We searched other resources including the UN Official Documents System, Open Grey, AGRIS FAO and Google Scholar using combinations of the above terms.
-  * We ran the search on 2021-03-01.
-  * We included studies if they met all of the following inclusion criteria; i) reported findings from trapping studies where the target was a small mammal, ii) described the type of trap used or the length of trapping activity or the location of the trapping activity, iii) included trapping activity from at least one West African country, iv) recorded the genus or species of trapped individuals, v) were published in a peer-reviewed journal or as a pre-print on a digital platform or as a report by a credible organisation.
-  * We excluded studies if they met any of the following exclusion criteria: i) reported data that were duplicated from a previously included study, ii) no full text available, iii) not available in English.
-  * One reviewer screened titles, abstracts and full texts against the inclusion and exclusion criteria. At each stage, a random subset (10%) was reviewed by a second reviewer.
+We searched other resources including the UN Official Documents System, Open Grey, AGRIS FAO and Google Scholar using combinations of the above terms. Searches were run on 2021-03-01.
+
+We included studies if they met all of the following inclusion criteria; i) reported findings from trapping studies where the target was a small mammal, ii) described the type of trap used or the length of trapping activity or the location of the trapping activity, iii) included trapping activity from at least one West African country, iv) recorded the genus or species of trapped individuals, v) were published in a peer-reviewed journal or as a pre-print on a digital platform or as a report by a credible organisation.
+We excluded studies if they met any of the following exclusion criteria: i) reported data that were duplicated from a previously included study, ii) no full text available, iii) not available in English.
+One reviewer screened titles, abstracts and full texts against the inclusion and exclusion criteria.
+At each stage, a random subset (10%) was reviewed by a second reviewer.
 
@@
 
-  * We extracted data from eligible studies into a Google sheets document.
+We extracted data from eligible studies into a Google sheets document.
 Extracted variables included i) study identifiers; ii) study aims; iii) trapping methodology; iv) geolocation data; v) method of speciation; vi) trapping locations and dates; vii) trapped species; viii) number of trap-nights and ix) microorganisms/pathogens of interest.
@@
 ### Location of rodent trapping studies and habitats studied
-  * We extracted GPS locations for the most precise location presented (i.e.
+
+We extracted GPS locations for the most precise location presented (i.e.
 trap, trap-line, study site or study region).
 We extracted coordinates in the format reported and converted them to decimal degrees.
-  * We recorded the habitat classification scheme a study used (e.g.
+We recorded the habitat classification scheme a study used (e.g.
 IUCN Habitat Classification Scheme (Version 3.1)).
@@
 ### Rodent presence, absence, abundance
-  * We mapped genus and species names to the species names used in the Global Biodiversity Information Facility (GBIF) taxonomy  [@gbif_the_global_biodiversity_information_facility_global_2021]
-  * We extracted information on the presence, absence and number of trapped individuals.
+
+We mapped genus and species names to the species names used in the Global Biodiversity Information Facility (GBIF) taxonomy  [@gbif_the_global_biodiversity_information_facility_global_2021].
+We extracted information on the presence, absence and number of trapped individuals.
 For studies reporting on all trapped individuals (i.e.
@@
 ### Pathogen presence and absence
-  * We extracted data on all pathogens assayed in studies investigating rodents for potential zoonoses.
+
+We extracted data on all pathogens assayed in studies investigating rodents for potential zoonoses.
 The number of rodents tested and the number of positive or negative samples were recorded alongside the type of assay used (e.g.
@@
 
-  * We summarised the number of studies, the year in which trapping occurred and the country in which they were conducted.
-  * We used the GPS coordinates of single trap, trapping grid/line or study site and the number of trap nights to calculate trapping effort (trap night density) within level 2 administrative areas in West Africa.
-  * For studies not reporting the number of trap nights we imputed the number based on the median trap success rate from studies which reported the number of trap nights (matched to building or non-building based study sites).
+We summarised the number of studies, the year in which trapping occurred and the country in which they were conducted.
+We used the GPS coordinates of single trap, trapping grid/line or study site and the number of trap nights to calculate trapping effort (trap night density) within level 2 administrative areas in West Africa.
+For studies not reporting the number of trap nights we imputed the number based on the median trap success rate from studies which reported the number of trap nights (matched to building or non-building based study sites).
 The median trap success rate for all rodents at a defined trap site was calculated separately for trap sites which included built environments and non-built environments.
 The number of rodents trapped at a trap site was then used in combination with these two values of trap success to impute the number of trap nights for trap sites with no reported trap nights.
-  * We summarised the habitat types of trap sites based on information reported in the study.
-  * For the subset of studies investigating rodent zoonoses we compared the location of trapping sites with SEDAC Global Population Density estimates for 2005, the median year studies were commenced.
+We summarised the habitat types of trap sites based on information reported in the study.
+
+For the subset of studies investigating rodent zoonoses we compared the location of trapping sites with SEDAC Global Population Density estimates for 2005, the median year studies were commenced.
 We used a Generalised Additive Model (GAM) incorporating a spatial interaction term to investigate the association of number of trap nights and human population density.
-The model structure was specified as below:
-    * *trap night density* ~ *Tweedie*(log(Population density (2005)) + (Longitude * Latitude))   
-  * We obtained land cover classifications from the European Space Agency Copernicus dataset at 300m^2^ spatial resolution (2005) we extracted the proportion of land cover classes within all regions of West Africa and all regions in which rodent trapping occurred for investigation of zoonotic diseases.
+The model structure was specified as:
+
+  + *Trap night density* ~ *Tweedie*(log(Population density (2005)) + (Longitude * Latitude))
+
+We obtained land cover classifications from the European Space Agency Copernicus dataset at 300m^2^ spatial resolution (2005) we extracted the proportion of land cover classes within all regions of West Africa and all regions in which rodent trapping occurred for investigation of zoonotic diseases.
 We compared the proportion of possible land cover classes with those of regions where trapping occurred.
-  * We mapped the presence and absence of rodent species and compared this to the presence and absence reported by both GBIF and IUCN to give a measure of the extent of each species range in which they have been sampled.
+
+Finally, we mapped the presence and absence of rodent species and compared this to the presence and absence reported by both GBIF and IUCN to give a measure of the extent of each species range in which they have been sampled.
 
@@
 
-  * We summarised the presence and absence of microorganisms, the assays used, their host species and the locations from where the samples were obtained.
-  * We investigated the association between rodent species and the detection of potential pathogens. We report proportion of positive and negative tests for each species and pathogen pair.
+We summarised the presence and absence of microorganisms, the assays used, their host species and the locations from where the samples were obtained.
+We investigated the association between rodent species and the detection of potential pathogens and report the proportion of positive and negative tests for each species and pathogen pair.
 
@@
 
-  * 4,282 relevant citations were identified, with 126 rodent trapping studies included in narrative synthesis.
-  * The earliest included studies were published in 1974 with increasing numbers of studies being performed annually since 2005.
-  * The median length of rodent trapping activity was one year (IQR 0-2 years). The median time from completion of rodent trapping to publication is 3 years (IQR 2-5 years) (see Figure 1).
+4,282 relevant citations were identified, with 126 rodent trapping studies included in narrative synthesis.
+The earliest included studies were published in 1974 with increasing numbers of studies being performed annually since 2005.
+The median length of rodent trapping activity was one year (IQR 0-2 years).
+The median time from completion of rodent trapping to publication is 3 years (IQR 2-5 years) (Figure 1.).
 
@@
 
-  * Rodent trapping took place at 1,331 trap sites with at least one trap site recorded from 14 West African countries. No rodent trapping studies were identified from Gambia or Togo.
-  * Thirty-one (25%) studies reported trapping at a single study site, 46 (37%) studies trapped at between two and five study sites, the remaining 49 studies trapped at between six and 93 study sites.
+Rodent trapping took place at 1,331 trap sites with at least one trap site recorded from 14 West African countries.
+No rodent trapping studies were identified from Gambia or Togo.
+Thirty-one (25%) studies reported trapping at a single study site, 46 (37%) studies trapped at between two and five study sites, the remaining 49 studies trapped at between six and 93 study sites.
 Trap sites were situated in 273 (19.3%) of 1414 level 2 administrative regions from the 14 West African nations.
 The areas with highest trapping activity included the capital cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako) and Ghana (Accra) and the largest cities of Côte d'Ivoire (Abidjan) and Benin (Cotonou).
-Outside of these cities, Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone were the most intensively studied regions (see Figure 2 - Panel A).
-  * Greater numbers of trap nights were conducted in areas with higher population densities (*R~2~* = 0.167).
+Outside of these cities, Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone were the most intensively studied regions (Figure 2.
+A).
+
+Greater numbers of trap nights were conducted in areas with higher population densities (*R~2~* = 0.167).
 In particular Mauritania, Northern Senegal, Eastern Sierra Leone, Eastern Guinea and South West Nigeria were trapped at increased rates than would be expected based on their population densities.
-South West Nigeria, Northern Nigeria, Liberia, Côte d'Ivoire, Ghana, Niger and Burkina Faso were trapped at lower rates (see Figure 2 - Panel B) .
-  * No studies reported trap habitats with reference to a standardised habitat classification scheme. 
-  * Extracted habitat types were grouped into 30 categories (see Supplementary Material 2 for the habitat dictionary).
+South West Nigeria, Northern Nigeria, Liberia, Côte d'Ivoire, Ghana, Niger and Burkina Faso were trapped at lower rates (Figure 2.
+B) .
+
+No studies reported trap habitats with reference to a standardised habitat classification scheme.
+Extracted habitat types were grouped into 30 categories (see Supplementary Material 2 for the habitat dictionary).
 At least one habitat was recorded for 17,122 trap sites (95%), with two or more habitats for a single trap site recorded for 4,403 (24%) sites.
 Single trap sites could span multiple habitat types resulting in 22,202 distinct habitat and trap sites.
-  * The most commonly trapped sites were in or around buildings (29%), in areas described as the rodents “natural habitat” (24%), in agricultural areas (13%) (e.g.
+The most commonly trapped sites were in or around buildings (29%), in areas described as the rodents “natural habitat” (24%), in agricultural areas (13%) (e.g.
 rice fields, palm plantations), forests (8%) and in the area surrounding buildings (4%).
-  * Explore whether trap sites are more likely to have been conducted within high population regions or whether there is an association with trap location and habitat as obtained from remote sensing data (Need to do this.
-The remote sensing data will be matched to the trap dates there will be several years between the trapping and data this is compared to it seems impractical to try and get annual population estimates or land classification so this will be chosen pragmatically)
 
@@
 ## Rodent presence, absence, abundance
-  * 73,164 small mammals were trapped (592 were trapped outside of West African countries), 2,830 (4%) trapped individuals were identified to order level (Rodentia), 7,760 (11%) were identified to genus level, the remaining 62,574 (85%) were identified to species level.
-  * In studies reporting the number of trap nights the median trap success rate of traps placed in or around buildings was 13% (IQR 6-24%), this compares to a median trap success rate of 3% (IQR 1-9%) in other habitats.
-  * Of the 147 distinct identified species trapped (see Supplementary Material 3) the majority were from the order Rodentia (112).
+
+73,164 small mammals were trapped (592 were trapped outside of West African countries), 2,830 (4%) trapped individuals were identified to order level (Rodentia), 7,760 (11%) were identified to genus level, the remaining 62,574 (85%) were identified to species level.
+In studies reporting the number of trap nights the median trap success rate of traps placed in or around buildings was 13% (IQR 6-24%), this compares to a median trap success rate of 3% (IQR 1-9%) in other habitats.
+Of the 147 distinct species trapped (see Supplementary Material 3) the majority were from the order Rodentia (112).
 Muridae (82) represented the largest family of rodents, followed by, Sciuridae (10), Gliridae and Nesomyidae (both 6), Ctenodactylidae and Anomaluridae (2) and Dipodidae, Hystricidae and Thryonomidae (all 1).
 The remaining 34 species came from the orders of the Soricomorpha (30), Erinaceomorpha (2) and Afrosoricida (1).
-  * The most commonly trapped genera of rodents were *Mastomys sp.* (27,079, 37.7%), *Rattus sp.* (11,472, 16.1%), *Mus sp.* (8,624, 12.0%), *Arvicanthis sp.* (5,821, 8.1%) and *Praomys sp.* (5,409, 7.5%).
+
+The most commonly trapped genera of rodents were *Mastomys sp.* (27,079, 37.7%), *Rattus sp.* (11,472, 16.1%), *Mus sp.* (8,624, 12.0%), *Arvicanthis sp.* (5,821, 8.1%) and *Praomys sp.* (5,409, 7.5%).
 At species level *Mastomys natalensis* (11,116, 17.4%), *Rattus rattus* (9,959, 15.6%), *Mastomys erythroleucus* (7,386, 11.6%), *Mus musculus* (6,245, 9.8%), *Arvicanthis niloticus* (5,497, 8.6%), *Mastomys huberti* (4,699, 7.4%) and Praomys daltoni (1,854, 2.9%) were the most commonly trapped.
-  * Locations of points for the presence or absence of the seven most commonly trapped species are displayed in Figure 3 (left column).
+
+Locations of points for the presence or absence of the seven most commonly trapped species are shown in Figure 3 (left column).
 These are able to identify locations of absence within an individual species documented range.
@@
 ## Pathogen presence and absence
-  * Sixty-two studies presented data on microorganisms that infect or are carried by small mammal species in West Africa.
-  * Seven studies solely investigated pathogens of rodents, including Hydatigera species (previously Taenia species) and Trichuris species.
-  * Fifty-five studies investigated microorganisms that were potentially zoonotic pathogens.
+
+Sixty-two studies presented data on microorganisms that infect or are carried by small mammal species in West Africa.
+Seven studies solely investigated pathogens of rodents, including Hydatigera species (previously Taenia species) and Trichuris species.
+Fifty-five studies investigated microorganisms that were potentially zoonotic pathogens.
 Thirty-two microorganisms were tested for, 8 of these were defined at species level, with the remaining 24 at levels of higher taxonomic classification.
-  * Thirty-two studies used Polymerase Chain Reaction (PCR) to detect the presence of 22 different species or families of microorganisms in 21,953 rodent samples.
+
+Thirty-two studies used Polymerase Chain Reaction (PCR) to detect the presence of 22 different species or families of microorganisms in 21,953 rodent samples.
 Eleven studies used antibody or antigen based molecular tests to detect the presence of 9 different species or families of microorganisms in 11,430 samples.
@@
 Three studies performed direct culture of Lassa mammarenavirus or Leishmania species to detect the presence of the pathogen in 643 samples.
-  * The most common microorganisms assessed for using PCR were Lassa mammarenavirus or other Arenaviridae (31%) the bacteria *Borrelia sp.* (11.3%) and *Bartonella sp.* (6.5%), followed by Usutu virus (6.4%) and Hantaviridae (5.6%).
+The most common microorganisms assessed for using PCR were Lassa mammarenavirus or other Arenaviridae (31%) the bacteria *Borrelia sp.* (11.3%) and *Bartonella sp.* (6.5%), followed by Usutu virus (6.4%) and Hantaviridae (5.6%).
 The most common pathogens assessed for with serology were *Lassa mammarenavirus* or other Arenaviridae (78%), *Toxoplasma gondii* (10.6%), *Borrelia sp.* (6.3%) and *Leptospirosis sp.* (2.2%).
@@
 All studies using direct culture were investigating either *Lassa mammarenavirus* (81%) or *Leishmania sp.* (19%).
-  * Most studies tested for a single microorganisms (39), with 16 studies testing for two or more microorganisms.
+
+Most studies tested for a single microorganisms (39), with 16 studies testing for two or more microorganisms.
 The most frequently tested for microorganisms were *Lassa mammarenavirus* (28%) or members of the Arenaviridae family (16%), the spirochete bacteria *Borrelia sp.* was investigated in 8 studies, *Bartonella sp.* and *Toxoplasma gondii* were investigated in 4 each, the remaining 25 microorganisms were reported in three or fewer studies.
-  * There is spatial heterogeneity in the location of trapped rodents that were tested for the four most commonly assayed microorganisms. 
-  * The presence and absence of Arenaviridae, predominantly *Lassa mammarenavirus*, in rodents was assessed in Guinea, Sierra Leone, Mali, Côte d'Ivoire, Benin and Nigeria.
+
+There is spatial heterogeneity in the location of trapped rodents that were tested for the four most commonly assayed microorganisms.
+The presence and absence of Arenaviridae, predominantly *Lassa mammarenavirus*, in rodents was assessed in Guinea, Sierra Leone, Mali, Côte d'Ivoire, Benin and Nigeria.
 The known reservoir species of this zoonotic pathogen *Mastomys natalensis* was additionally detected in Senegal, Guinea-Bissau, Burkina-Faso and Niger with likely presence in Liberia and Togo.
 Arenaviridae were detected in rodents from 17 of 54 species or genera assayed from Guinea, Sierra Leone, Mali, Côte d'Ivoire, Ghana and Nigeria, no positive rodents were detected in Benin (Figure 4 top row).
-  * The presence and absence of *Borrelia sp.* was investigated in Senegal, Mauritania, Mali, Niger and Benin.
+The presence and absence of *Borrelia sp.* was investigated in Senegal, Mauritania, Mali, Niger and Benin.
 Microorganisms of this genera were detected from samples in Mali with negative samples obtained from Senegal, Mauritania, Niger and Benin (Figure 4 second row).
 *Borrelia sp.* were detected in 8 of 44 species or genera of rodents assayed.
-  * The presence and absence of *Bartonella sp.* was investigated in Senegal, Mali and Nigeria with positive samples obtained at most sites from all three countries (Figure 4 third row).
+The presence and absence of *Bartonella sp.* was investigated in Senegal, Mali and Nigeria with positive samples obtained at most sites from all three countries (Figure 4 third row).
 *Bartonella sp.* were detected in 11 of 11 specis or genera of rodents assayed.
-  * The presence of *Toxoplasma gondii* was only investigated in Senegal with relatively high proportions of positive samples at all sites (Figure 4 fourth row).
+The presence of *Toxoplasma gondii* was only investigated in Senegal with relatively high proportions of positive samples at all sites (Figure 4 fourth row).
 *Toxoplasma gondii* was detected in 6 of 8 rodent species assayed.
@@
 ## Host-pathogen associations
-  * Ninety-seven rodent species were investigated for the presence of a zoonotic pathogen. 42,940 assays were performed on 32,014 individual rodents. 
-  * The rodent species most commonly assayed for zoonotic pathogens were, *Rattus rattus* (n = 2,977) assessed for 24 pathogens, *Mus musculus* (n = 3,402) for 23, *Mastomys natalensis* (n = 7,189) and *Mastomys erythroleucus* (n = 3,013) both 19 and *Arvicanthis niloticus* (n = 3,840) for 18.
-  * All remaining species were investigated for 10 or fewer pathogens.
-  * *Lassa mammarenavirus* was detected in 414 *Mastomys natalensis*, 9 *Mastomys erythroleucus*, 5 *Hylomyscus pamfi* and 3 *Mus baoulei* through PCR.
+
+Ninety-seven rodent species were investigated for the presence of a zoonotic pathogen.
+42,940 assays were performed on 32,014 individual rodents.
+The rodent species most commonly assayed for zoonotic pathogens were, *Rattus rattus* (n = 2,977) assessed for 24 pathogens, *Mus musculus* (n = 3,402) for 23, *Mastomys natalensis* (n = 7,189) and *Mastomys erythroleucus* (n = 3,013) both 19 and *Arvicanthis niloticus* (n = 3,840) for 18.
+All remaining species were investigated for 10 or fewer pathogens.
+
+*Lassa mammarenavirus* was detected in 414 *Mastomys natalensis*, 9 *Mastomys erythroleucus*, 5 *Hylomyscus pamfi* and 3 *Mus baoulei* through PCR.
 In serological assays, evidence of prior infection or acquisition of maternal antibodies to *Lassa mammarenavirus* was detected in 465 *Mastomys natalensis*, 361 *Mastomys sp.*, 6 *Mus musculus* and 4 *Rattus rattus*.
@@
 *Bartonella sp.* were detected in; 113 *Mastomys natalensis*, 50 *Mastomys erythroleucus*, 39 *Rattus norvegicus*, 16 *Rattus rattus* and 13 individuals of 7 further species through PCR.
-  * *Borrelia sp.* infections were identified through PCR in 110 individuals; 37 *Mastomys erythroleucus*, 12 *Mus musculus*, 11 *Mastomys huberti* and *Praomys daltoni*.
+
+*Borrelia sp.* infections were identified through PCR in 110 individuals; 37 *Mastomys erythroleucus*, 12 *Mus musculus*, 11 *Mastomys huberti* and *Praomys daltoni*.
 The remaining 39 positive results were from 9 further species.
@@
 Most *Borrelia sp.* infections were identified through histology or direct visualisation with, 243 *Arvicanthis niloticus*, 102 *Mastomys natalensis*, and 78 *Mastomys huberti* along with a further 64 individuals from 8 species infected.
-  * Finally, *Toxoplasma gondii* was detected through PCR in 89 Mus musculus, 13 *Cricetomys gambianus*, 5 *Crocidura olivieri* and 3 *Rattus rattus*.
+
+*Toxoplasma gondii* was detected through PCR in 89 Mus musculus, 13 *Cricetomys gambianus*, 5 *Crocidura olivieri* and 3 *Rattus rattus*.
 From serological assays evidence of prior infection was found in 25 *Mus musculus*, 46 *Mastomys erythroleucus*, 4 *Rattus rattus* and *Mastomys natalensis*.
 A single *Mus musculus* and *Rattus rattus* were found to be infected using histology and direct visualisation.
-  * Thirty-eight species of Rodentia, 3 species of Soricomorpha and 2 species of Erinaceomorpha that were tested for potential pathogens had entirely negative results.
+Finally, 38 species of Rodentia, 3 species of Soricomorpha and 2 species of Erinaceomorpha that were tested for potential pathogens had entirely negative results.
 
@@
 
-## Discussion
+# Discussion
 

### 2021-10-04 reports/manuscript_v2.Rmd
from mac

@@
 We extracted coordinates in the format reported and converted them to decimal degrees.
-We recorded the habitat classification scheme a study used (e.g.
+We recorded the habitat classification scheme used by a study (e.g.
 IUCN Habitat Classification Scheme (Version 3.1)).
-For studies not using standardised recording, the explicit description from of the habitat in which the trap was placed was extracted.
+For studies not using standardised habitat recording, the explicit description of the habitat in which the trap was placed were extracted.
 For studies reporting multiple habitat types (e.g.

### 2021-10-04 scripts/2_4_clean_pathogen.R
clean repository

@@ -42,7 +42,7 @@ pathogen_data <- pathogen_data %>%
   mutate(across(
     all_of(pathogen_tested),
     ~recode(., !!!pathogen_name)
-    ))
+  ))
 
 # Cleaning location data
 pathogen_data %<>%

### 2021-10-04 scripts/analysis.R

@@ -129,7 +129,11 @@ study_timings <- ggplot(study_start) +
        x = element_blank(),
        alpha = "Study dates reported")
 
-ggsave(plot = study_timings, filename = here("figures", "Fig_1_Panel_A.png"), dpi = 300)
+<<<<<<< HEAD
+ggsave(plot = study_timings, filename = here("figures", "Fig_1_Panel_A.png"), dpi = 300, height = 8, width = 10)
+=======
+ggsave(plot = study_timings, filename = here("figures", "Fig_1_Panel_A.png"), dpi = 300, height = 8)
+>>>>>>> 55927033cfb2829238efa72cd4a7eeb6ca51203d
 write_rds(study_timings, here("plots", "study_timings.rds"))
 
 described_studies <- study_start %>%
@@ -272,7 +276,7 @@ summary(non_buildings_tn$trap_success)
 # Trap night and population -----------------------------------------------
 if(!file.exists(here("data_clean", "pop_tn_analysis.rds"))) {
   rodent_spatial <- read_rds(here("data_clean", "rodent_spatial.rds")) %>%
-    select(-trap_nights)
+    dplyr::select(-trap_nights)
 
   imputed_tn <- read_rds(here("data_clean", "imputed_trap_nights.rds"))
 
@@ -316,17 +320,17 @@ if(!file.exists(here("data_clean", "pop_tn_analysis.rds"))) {
   vect_sites <- vect(sites_2)
 
   crop_pop <- crop(human_pop, vect_sites)
-  wa_pop <- writeRaster(crop_pop, here("data_download", "pop_2005","wa_pop_2005.tif"))
+  wa_pop <- writeRaster(crop_pop, here("data_download", "pop_2005","wa_pop_2005.tif"), overwrite = TRUE)
   region_pop <- terra::extract(crop_pop, vect_sites, fun = "median", method = "simple", na.rm = TRUE, touches = TRUE)
   region_pop_sf <- cbind(vect_sites, region_pop) %>%
     st_as_sf() %>%
     st_centroid() %>%
-    mutate(X = st_coordinates(.$geometry)[,1],
-           Y = st_coordinates(.$geometry)[,2],
+    mutate(x = st_coordinates(.$geometry)[,1],
+           y = st_coordinates(.$geometry)[,2],
            tn_density = case_when(is.na(tn_density) ~ 0,
                                   TRUE ~ tn_density)) %>%
     tibble() %>%
[…]

### 2021-10-04 scripts/libraries.R

@@ -3,6 +3,7 @@ pkgs =
   c("here",
     "tidyverse",
     "mgcv",
+    "mgcViz",
     "magrittr",
     "bib2df",
     "googledrive",
@@ -17,6 +18,7 @@ pkgs =
     "cowplot",
     "maptools",
     "raster",
+    "ncdf4",
     "parzer",
     "terra",
     "mapview",
@@ -28,6 +30,7 @@ pkgs =
     "lubridate",
     "gt",
     "gtsummary",
+    "gridExtra",
     "lubridate",
     "see",
     "scales"

### 2021-10-04 scripts/mapping_pathogens.R

@@ -70,11 +70,17 @@ pathogen_map <- function(pathogen_genus) {
                          st_as_sf() %>%
                          mutate(x = round(st_coordinates(.)[,1], 3),
                                 y = round(st_coordinates(.)[,2], 3),
-                                pos_neg_count = case_when(pos_neg == "Positive" ~ 1,
-                                                          TRUE ~ 0)) %>%
+                                genus = snakecase::to_sentence_case(genus),
+                                `Species` = snakecase::to_sentence_case(classification),
+                                pos_neg = case_when(pos_neg == "Positive" ~ 1,
+                                                    TRUE ~ 0),
+                                unique_positive = case_when(pcr_positive == 0 ~ ab_ag_positive + culture_positive,
+                                                            pcr_positive != 0 & ab_ag_positive != 0 ~ ab_ag_positive + culture_positive,
+                                                            pcr_positive != 0 & ab_ag_positive == 0 ~ pcr_positive + culture_positive,
+                                                            TRUE ~ ab_ag_positive + culture_positive)) %>%
                          group_by(x, y) %>%
                          summarise(number_tested = sum(number_tested),
-                                   number_positive = sum(pos_neg_count),
+                                   number_positive = sum(unique_positive),
                                    number_negative = number_tested-number_positive,
                                    prop_positive = number_positive/number_tested) %>%
                          mutate(pos_neg = case_when(number_positive > 0 ~ "Positive",
@@ -99,9 +105,13 @@ pathogen_map <- function(pathogen_genus) {
                           `Species` = snakecase::to_sentence_case(classification),
                           pos_neg = case_when(pos_neg == "Positive" ~ 1,
                                               TRUE ~ 0)) %>%
+                   mutate(unique_positive = case_when(pcr_positive == 0 ~ ab_ag_positive + culture_positive,
[…]

### 2021-10-04 scripts/mapping_traps.R

@@ -16,10 +16,13 @@ level_2 <- read_rds(here("data_download", "admin_spatial", "level_2_admin.rds"))
 
 non_trapped <- read_rds(here("data_download", "admin_spatial", "level_2_TGOGMB.rds"))
 
+included_countries <- level_0 %>%
+  filter(GID_0 %in% wa_countries)
+
 studies <- read_rds(here("data_clean", "studies.rds"))
 
 rodent_spatial <- read_rds(here("data_clean", "rodent_spatial.rds")) %>%
-  select(-trap_nights)
+  dplyr::select(-trap_nights)
 
 imputed_tn <- read_rds(here("data_clean", "imputed_trap_nights.rds"))
 
@@ -160,9 +163,9 @@ ggplot() +
   theme_minimal()
 
 level_2_sites$breaks <- cut(level_2_sites$tn_density,
-                   breaks = c(0, 0.001, 0.01, 0.1, 1, 10, 100, 200),
-                   labels = c("0 - 0.001", "0.001 - 0.01", "0.01 - 0.1",
-                            "0.1 - 1", "1 - 10", "10 - 100", "100 - 200"))
+                            breaks = c(0, 0.001, 0.01, 0.1, 1, 10, 100, 200),
+                            labels = c("0 - 0.001", "0.001 - 0.01", "0.01 - 0.1",
+                                       "0.1 - 1", "1 - 10", "10 - 100", "100 - 200"))
 
 trap_night_density_level2 <- ggplot() +
   geom_sf(data = level_2_sites,
@@ -307,7 +310,27 @@ tn_to_pop <- tm_shape(pop_predict) +
 
 write_rds(tn_to_pop, here("plots", "tn_to_pop.rds"))
 
-panel_b <- plot_grid(log_pop_tn, tmap_grob(tn_to_pop))
+tn_pop_map <- read_rds(here("data_clean", "tn_pop_model.rds"))
+
+map_1 <- getViz(tn_pop_model)
+(
+  m1 <- plot(sm(map_1, 1), n = 150, too.far = 0.02) +
+    l_fitRaster(pTrans = zto1(0.05, 2, 0.1)) +
+    geom_sf(data = included_countries, alpha = 0.1, lwd = 0.1, inherit.aes = FALSE) +
+    scale_fill_viridis_c(na.value = "white") +
+    annotation_north_arrow(height = unit(1, "cm"),
+                           style = north_arrow_minimal(text_size = 8)) +
[…]

### 2021-10-04 reports/manuscript_v2.Rmd
clean merge

@@
 ---
-title: " A scoping review of rodent trapping studies to understand the prevalence of zoonotic disease in West Africa."
+title: "A scoping review of rodent trapping studies to understand the prevalence of zoonotic disease in West Africa."
 description: |
@@
   # word_document:
+  # pdf_document:
 knit: (function(inputFile, encoding) {
@@
 For example, the inclusion of absence data and trapping effort can help identify regions where data sparsity may produce inappropriately predicted risk.
-Furthermore, this data specifies regions in need of comprehensive study for four targeted rodent zoonoses, Arenaviridae, *Borellia sp.*, *Bartonella sp.* and *Toxoplasma sp.*.
+Furthermore, these data identify regions in need of comprehensive study for our four targeted rodent zoonoses, Arenaviridae, *Borellia sp.*, *Bartonella sp.* and *Toxoplasma sp.*.
 
@@
 The potential effect of zoonotic infectious diseases outbreaks on human health and societies has been dramatically highlighted through the ongoing SARS-CoV-2 pandemic and recent Ebola virus outbreaks.
-The number of zoonotic spillover events are projected to increase under intensifying anthropogenic pressure such as, increased human populations [@Allen], increasing urbanisation [@Hassell], and global climate change [@morse_prediction_2012].
-In addition, two taxa are proposed to contribute to the highest burden of potential zoonoses globally, rodents (Rodentia) and bats (Chiroptera) [@han].
-Of 2,220 extant rodent species, 244 (10.7%) are described as being reservoirs of 85 zoonotic pathogens [@han].
-Specifically, West Africa has been identified as region at increased hazard for rodent-borne zoonotic spillover events under different projected scenarios [@garcia-pena_land-use_2021].
-Rodents are implicated in the transmission pathways of several diseases in this region, including, Lassa fever, Schistosomiasis and Leptospirosis.
+The number of zoonotic spillover events are projected to increase under intensifying anthropogenic pressure such as, increased human populations [@allen_global_2017], increasing urbanisation [@hassell_urbanization_2017], and global climate change [@morse_prediction_2012].
+In addition, two taxa are proposed to contribute to the greatest number of potential zoonotuc pathogens globally, rodents (Rodentia) and bats (Chiroptera) [@han_rodent_2015].
+Of 2,220 extant rodent species, 244 (10.7%) are described as being reservoirs of 85 zoonoses [@han_global_2016].
+Specifically, West Africa has previously been identified as region at increased hazard for rodent-borne zoonotic spillover events under different projected scenarios [@garcia-pena_land-use_2021].
+Rodents are implicated in the transmission pathways of several human diseases in this region, including, Lassa fever, Schistosomiasis and Leptospirosis.
 
-Rodent species form diverse assemblages providing important ecosystem services such as pest regulation and seed dispersal [@fischer_ecosystem_2018].
-However, the role a minority of species play in zoonotic infectious disease spillover can be considered an ecosystem disservice.
+Rodent species form diverse assemblages, which provide important and beneficial ecosystem services including pest regulation and seed dispersal [@fischer_ecosystem_2018].
+The role of rodent species' in zoonotic infectious disease spillover or novel pathogen emergence can be viewed as an ecosystem disservice.
 Rodents typically demonstrate “fast” life histories [@dobson_fast_2007], these traits are also associated with a species being a reservoir of zoonotic pathogens [@han_rodent_2015; @albery_fast-lived_2021].
-Further, these traits are prevalent in species that thrive in human dominated landscapes where they displace species that are less likely to be reservoirs of zoonotic disease [@gibb_zoonotic_2020].
-The widespread occurrence of reservoir species and their proximity to human activity make the description of rodent species assemblages and host-pathogen interactions vitally important.
+Further, these traits are prevalent in species that thrive in human dominated landscapes where they displace species that are less likely to be reservoirs of zoonotic pathogen [@gibb_zoonotic_2020].
+The widespread occurrence of reservoir species and their proximity to human activity make the description of rodent species assemblages and host-pathogen interactions vitally important to understanding the risk of zoonotic spillover and novel pathogen emergence.
 
 Rodent trapping studies provide a useful method to describe rodent population structures and survey for potential zoonoses.
-Studies have been conducted in West Africa to both identify novel potential zoonotic pathogens within rodents [@predicts] and to investigate the prevalence and burden of known pathogens within known rodent hosts (e.g.
+Studies have been conducted in West Africa to both identify novel potential zoonotic pathogens within rodents [@usaid_one_2021] and to investigate the prevalence and burden of known pathogens within known rodent hosts (e.g.
 for Lassa fever [@fichet-calvet_diversity_2009] and Schistosomiasis [@catalano_multihost_2020]).
@@
 IUCN, GBIF and GIDEON) are typically used to inform species distribution maps and zoonotic disease spillover predictions [@han_global_2016; @smith_global_2014; @pigott_mapping_2014].
-.
+
 These studies have been used to identify potential geographic hotspots where virus and host species diversity may be expected to be at its greatest to predict regional zoonotic disease spillover risks.
@@
 more intensive studies over longer time periods) or over-representation of certain habitats (e.g.
-periurban landscapes) could lead to an apparent association between locations and risk that is driven by these factors rather than an underlying host and virus association [@wille_how_2021; @gibb_mammal_2021].
-Conversely some regions may not be sampled adequately and therefore under-represented in these datasets due to sparse human populations or inaccessible habitats which may lead to proposing that these regions are at low risk of zoonotic disease spillover events.
+peri-urban landscapes) could lead to an apparent association between locations and risk that is driven by these factors rather than an underlying host and virus association [@wille_how_2021; @gibb_mammal_2021].
+Conversely some regions may not be sampled adequately and therefore under-represented in these datasets due to sparse human populations or inaccessible habitats leading to these regions being reported to be at low risk of zoonotic disease spillover events.
 
@@
 
-For the subset of studies investigating rodent zoonoses we compared the location of trapping sites with SEDAC Global Population Density estimates for 2005, the median year studies were commenced.
+For studies investigating rodent zoonoses we compared the location of trapping sites with SEDAC Global Population Density estimates for 2005 - the median year studies were commenced.
 We used a Generalised Additive Model (GAM) incorporating a spatial interaction term to investigate the association of number of trap nights and human population density.
@@
 
-4,282 relevant citations were identified, with 126 rodent trapping studies included in narrative synthesis.
+We identified 4,282 relevant citations, with 126 rodent trapping studies included.
 The earliest included studies were published in 1974 with increasing numbers of studies being performed annually since 2005.
@@
 
+if(knitr::is_html_output()) {
+  knitr::include_graphics(here("figures", "Fig_1_Panel_A.png"), dpi = 320)
+} else {
   knitr::include_graphics(here("figures", "Fig_1_Panel_A.png"))
+}
 
@@
 No rodent trapping studies were identified from Gambia or Togo.
+The most frequently studied countries were Senegal, Ghana, Guinea and Nigeria (more than 10 studies each).
+Fewer than 5 studies were conducted in Liberia, Burkina Faso and Guinea-Bissau.
 Thirty-one (25%) studies reported trapping at a single study site, 46 (37%) studies trapped at between two and five study sites, the remaining 49 studies trapped at between six and 93 study sites.
-Trap sites were situated in 273 (19.3%) of 1414 level 2 administrative regions from the 14 West African nations.
-The areas with highest trapping activity included the capital cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako) and Ghana (Accra) and the largest cities of Côte d'Ivoire (Abidjan) and Benin (Cotonou).
-Outside of these cities, Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone were the most intensively studied regions (Figure 2.
+Trap sites were situated in 273 (19.3%) of 1414 level 2 administrative regions in the 14 West African nations.
+The areas with highest trap night density included the capital cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako) and Ghana (Accra) and the largest cities of Côte d'Ivoire (Abidjan) and Benin (Cotonou).
+Outside of these cities, Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone had the highest density of trap nights (Figure 2.
 A).
 
-Greater numbers of trap nights were conducted in areas with higher population densities (*R~2~* = 0.167).
-In particular Mauritania, Northern Senegal, Eastern Sierra Leone, Eastern Guinea and South West Nigeria were trapped at increased rates than would be expected based on their population densities.
-South West Nigeria, Northern Nigeria, Liberia, Côte d'Ivoire, Ghana, Niger and Burkina Faso were trapped at lower rates (Figure 2.
-B) .
+Greater numbers of trap nights were conducted in areas with higher population densities (*R~2~* = 0.18) adjusting for spatial autocorrelation.
+In particular Mauritania, Northern Senegal, Eastern Sierra Leone, Eastern Guinea and South West Nigeria were trapped at higher rates than would be expected based on their population densities.
+South West Nigeria, Northern Nigeria, Liberia, Côte d'Ivoire, Ghana, Niger and Burkina Faso were trapped at lower rates than would be expected.
 
@@
 ```{r, figure_2, layout = "l-screen-inset", echo=FALSE}
+
 knitr::include_graphics(here("figures", "Figure_2.png"))
+
 ```
@@
 In studies reporting the number of trap nights the median trap success rate of traps placed in or around buildings was 13% (IQR 6-24%), this compares to a median trap success rate of 3% (IQR 1-9%) in other habitats.
-Of the 147 distinct species trapped (see Supplementary Material 3) the majority were from the order Rodentia (112).
-Muridae (82) represented the largest family of rodents, followed by, Sciuridae (10), Gliridae and Nesomyidae (both 6), Ctenodactylidae and Anomaluridae (2) and Dipodidae, Hystricidae and Thryonomidae (all 1).
-The remaining 34 species came from the orders of the Soricomorpha (30), Erinaceomorpha (2) and Afrosoricida (1).
+The majority of the 147 species trapped were Rodentia (112) (Supplementary Material 3), of these Muridae (82) were the largest family of rodents, followed by, Sciuridae (10), Gliridae and Nesomyidae (both 6), Ctenodactylidae, Anomaluridae, Dipodidae, Hystricidae and Thryonomidae (2 or fewer).
+The remaining 34 species were from the orders of Soricomorpha (30), Erinaceomorpha (2) and Afrosoricida (1).
 
-The most commonly trapped genera of rodents were *Mastomys sp.* (27,079, 37.7%), *Rattus sp.* (11,472, 16.1%), *Mus sp.* (8,624, 12.0%), *Arvicanthis sp.* (5,821, 8.1%) and *Praomys sp.* (5,409, 7.5%).
-At species level *Mastomys natalensis* (11,116, 17.4%), *Rattus rattus* (9,959, 15.6%), *Mastomys erythroleucus* (7,386, 11.6%), *Mus musculus* (6,245, 9.8%), *Arvicanthis niloticus* (5,497, 8.6%), *Mastomys huberti* (4,699, 7.4%) and Praomys daltoni (1,854, 2.9%) were the most commonly trapped.
+*Mastomys sp.* were the most commonly trapped genus (27,079, 37.7%), followed by, *Rattus sp.* (11,472, 16.1%), *Mus sp.* (8,624, 12.0%), *Arvicanthis sp.* (5,821, 8.1%) and *Praomys sp.* (5,409, 7.5%).
+*Mastomys natalensis* were the most commonly trapped species (11,116, 17.4%), followed by, *Rattus rattus* (9,959, 15.6%), *Mastomys erythroleucus* (7,386, 11.6%), *Mus musculus* (6,245, 9.8%), *Arvicanthis niloticus* (5,497, 8.6%), *Mastomys huberti* (4,699, 7.4%) and *Praomys daltoni* (1,854, 2.9%).
 
 Locations of points for the presence or absence of the seven most commonly trapped species are shown in Figure 3 (left column).
-These are able to identify locations of absence within an individual species documented range.
-Whether an absence represents a "true absence" (i.e.
-not present at this location) or a "pseudo absence" (i.e.
-not detected at this location) should be interpreted with consideration to the trapping effort at the site.
-This is compared to reported presence obtained from GBIF (centre column) and IUCN species range data (left column) in Figure 3.
+Absence include true absences and pseudo-absences dependent on the trapping effort at the site.
+This are shown compared to species presence obtained from GBIF (centre column) and IUCN species range data (left column) (Figure 3).
 
 ```{r, figure_3, layout = "l-screen-inset", echo=FALSE}
+
 knitr::include_graphics(here("figures", "Figure_3.png"))
+
 ```
@@
 Seven studies solely investigated pathogens of rodents, including Hydatigera species (previously Taenia species) and Trichuris species.
-Fifty-five studies investigated microorganisms that were potentially zoonotic pathogens.
-Thirty-two microorganisms were tested for, 8 of these were defined at species level, with the remaining 24 at levels of higher taxonomic classification.
+Zoonotic pathogens and potentially zoonotic micro-organisms were investigated in 55 studies.
+Thirty-two microorganisms were tested for, 8 of these were specified to species level, the remaining 24 were specified at higher taxonomic classifications.
 
-Thirty-two studies used Polymerase Chain Reaction (PCR) to detect the presence of 22 different species or families of microorganisms in 21,953 rodent samples.
-Eleven studies used antibody or antigen based molecular tests to detect the presence of 9 different species or families of microorganisms in 11,430 samples.
-Eight studies conducted histological or direct visualisation assays of samples for 11 parasitic or bacterial species in 11,229 samples.
-Three studies performed direct culture of Lassa mammarenavirus or Leishmania species to detect the presence of the pathogen in 643 samples.
-The most common microorganisms assessed for using PCR were Lassa mammarenavirus or other Arenaviridae (31%) the bacteria *Borrelia sp.* (11.3%) and *Bartonella sp.* (6.5%), followed by Usutu virus (6.4%) and Hantaviridae (5.6%).
+Polymerase Chain Reaction (PCR) was used in 32 studies to detect the presence of 22 different species or families of microorganisms from 21,953 rodent samples.
+The most common microorganisms assayed for using PCR were *Lassa mammarenavirus* or other Arenaviridae (31%) the bacteria *Borrelia sp.* (11.3%) and *Bartonella sp.* (6.5%), followed by Usutu virus (6.4%) and Hantaviridae (5.6%).
+Molecular tests (antibody or antigen based) were used in 11 studies to detect the presence of 9 different species or families of microorganisms from 11,430 rodent samples.
 The most common pathogens assessed for with serology were *Lassa mammarenavirus* or other Arenaviridae (78%), *Toxoplasma gondii* (10.6%), *Borrelia sp.* (6.3%) and *Leptospirosis sp.* (2.2%).
+Histological or direct visualisation assays were used in 8 studies investigating 11 parasitic or bacterial species from 11,229 rodent samples.
 The most common pathogens assessed for with histology or direct visualisation were *Borrelia sp.* (48.7%), *Schistosoma sp*.
 (20.4%) and other parasites.
-All studies using direct culture were investigating either *Lassa mammarenavirus* (81%) or *Leishmania sp.* (19%).
+Direct culture of *Lassa mammarenavirus* or Leishmania species were reported from 3 studies to detect the presence of these pathogen in 643 rodent samples.
 
-Most studies tested for a single microorganisms (39), with 16 studies testing for two or more microorganisms.
-The most frequently tested for microorganisms were *Lassa mammarenavirus* (28%) or members of the Arenaviridae family (16%), the spirochete bacteria *Borrelia sp.* was investigated in 8 studies, *Bartonella sp.* and *Toxoplasma gondii* were investigated in 4 each, the remaining 25 microorganisms were reported in three or fewer studies.
+Most studies tested for a single microorganisms (39), 16 studies tested for two or more microorganisms.
+The most frequently investigated microorganisms were *Lassa mammarenavirus* (17 studies) or members of the Arenaviridae family (10), the spirochete bacteria *Borrelia sp.* (8), *Bartonella sp.* and *Toxoplasma gondii* (4), the remaining 25 microorganisms were reported in three or fewer studies.
 
-There is spatial heterogeneity in the location of trapped rodents that were tested for the four most commonly assayed microorganisms.
-The presence and absence of Arenaviridae, predominantly *Lassa mammarenavirus*, in rodents was assessed in Guinea, Sierra Leone, Mali, Côte d'Ivoire, Benin and Nigeria.
-The known reservoir species of this zoonotic pathogen *Mastomys natalensis* was additionally detected in Senegal, Guinea-Bissau, Burkina-Faso and Niger with likely presence in Liberia and Togo.
-Arenaviridae were detected in rodents from 17 of 54 species or genera assayed from Guinea, Sierra Leone, Mali, Côte d'Ivoire, Ghana and Nigeria, no positive rodents were detected in Benin (Figure 4 top row).
-The presence and absence of *Borrelia sp.* was investigated in Senegal, Mauritania, Mali, Niger and Benin.
-Microorganisms of this genera were detected from samples in Mali with negative samples obtained from Senegal, Mauritania, Niger and Benin (Figure 4 second row).
-*Borrelia sp.* were detected in 8 of 44 species or genera of rodents assayed.
-The presence and absence of *Bartonella sp.* was investigated in Senegal, Mali and Nigeria with positive samples obtained at most sites from all three countries (Figure 4 third row).
-*Bartonella sp.* were detected in 11 of 11 specis or genera of rodents assayed.
-The presence of *Toxoplasma gondii* was only investigated in Senegal with relatively high proportions of positive samples at all sites (Figure 4 fourth row).
-*Toxoplasma gondii* was detected in 6 of 8 rodent species assayed.
+The location of trapped rodents that were tested for the four most commonly assayed microorganisms did not encompass the entire range of these hosts.
+The presence and absence of Arenaviridae, predominantly *Lassa mammarenavirus*, in rodents was investigated in Benin, Côte d'Ivoire, Guinea, Mali, Nigeria and Sierra Leone.
+No studies investigated the prevalence of these pathogens in hosts from Burkina-Faso, Guinea-Bissau, Liberia, Niger, Senegal and Togo.
+Arenaviridae were detected in rodents from 17 of 54 species or genera assayed in all the countries it was investigated except Benin (Figure 4 top row).
+The presence and absence of *Bartonella sp.* was investigated in Mali and Nigeria and Senegal with positive samples detected in 11 of 11 species or genera of rodents obtained at most sites from these countries (Figure 4 second row).
+The presence and absence of *Borrelia sp.* was investigated in Benin, Mali, Mauritania, Niger and Senegal.
+*Borrelia sp.* were detected in 8 of 44 species or genera of rodents assayed in Mali with negative samples obtained from Benin, Mauritania, Niger and Senegal (Figure 4 third row).
+The presence of *Toxoplasma gondii* was only investigated in Senegal and was detected in 6 of 8 rodent species assayed (Figure 4 fourth row).
 
-```{r, figure_4, layout = "l-screen-inset", echo=FALSE}
+```{r, figure_4, echo=FALSE}
 
@@
 
-Ninety-seven rodent species were investigated for the presence of a zoonotic pathogen.
-42,940 assays were performed on 32,014 individual rodents.
+Ninety-seven rodent species were investigated for the presence of a zoonotic pathogen, 42,940 assays were performed on 32,014 individual rodent samples.
 The rodent species most commonly assayed for zoonotic pathogens were, *Rattus rattus* (n = 2,977) assessed for 24 pathogens, *Mus musculus* (n = 3,402) for 23, *Mastomys natalensis* (n = 7,189) and *Mastomys erythroleucus* (n = 3,013) both 19 and *Arvicanthis niloticus* (n = 3,840) for 18.
@@
 
-*Lassa mammarenavirus* was detected in 414 *Mastomys natalensis*, 9 *Mastomys erythroleucus*, 5 *Hylomyscus pamfi* and 3 *Mus baoulei* through PCR.
-In serological assays, evidence of prior infection or acquisition of maternal antibodies to *Lassa mammarenavirus* was detected in 465 *Mastomys natalensis*, 361 *Mastomys sp.*, 6 *Mus musculus* and 4 *Rattus rattus*.
-From direct viral culture 19 *Mastomys natalensis*, 2 *Rattus rattus* and 1 *Mus minutoides* were found to be infected with *Lassa mammarenavirus*.
-*Bartonella sp.* were detected in; 113 *Mastomys natalensis*, 50 *Mastomys erythroleucus*, 39 *Rattus norvegicus*, 16 *Rattus rattus* and 13 individuals of 7 further species through PCR.
+*Lassa mammarenavirus* was detected in 14 species of rodents and Crocidura genera.
+*Mastomys natalensis* was the most common species to test positive (Figure 4.).
+Fifty-four species and genera were tested for arenaviruses, with 37 testing negative.
+*Bartonella sp.* infections were identified in all 11 species of rodents and shrews that were tested.
+*Mastomys erythroleucus* was the most common species to test positive.
+*Borrelia sp.* infections were identified in 13 species of rodents and shrews.
+*Mastomys natalensis* was the most common species to test positive.
+Forty-four species and genera were tested for *Borrelia sp.* with 36 testing negative.
+*Toxoplasma gondii* was detected in 6 species of rodents and shrews, with all individuals from two species *Arvicanthis niloticus* and *Praomys daltoni* testing negative.
+*Mus musculus* was the most common species to test positive.
+Finally, 38 species of Rodentia, 3 species of Soricomorpha and 2 species of Erinaceomorpha that were tested for pathogens had entirely negative results.
 
-*Borrelia sp.* infections were identified through PCR in 110 individuals; 37 *Mastomys erythroleucus*, 12 *Mus musculus*, 11 *Mastomys huberti* and *Praomys daltoni*.
-The remaining 39 positive results were from 9 further species.
-In serological assays, 47 *Mastomys natalensis*, 12 *Mastomys erythroleucus* and 8 *Praomys daltoni* were positive, with the remaining 11 positive individuals from three further species.
-Most *Borrelia sp.* infections were identified through histology or direct visualisation with, 243 *Arvicanthis niloticus*, 102 *Mastomys natalensis*, and 78 *Mastomys huberti* along with a further 64 individuals from 8 species infected.
-
-*Toxoplasma gondii* was detected through PCR in 89 Mus musculus, 13 *Cricetomys gambianus*, 5 *Crocidura olivieri* and 3 *Rattus rattus*.
-From serological assays evidence of prior infection was found in 25 *Mus musculus*, 46 *Mastomys erythroleucus*, 4 *Rattus rattus* and *Mastomys natalensis*.
-A single *Mus musculus* and *Rattus rattus* were found to be infected using histology and direct visualisation.
-Finally, 38 species of Rodentia, 3 species of Soricomorpha and 2 species of Erinaceomorpha that were tested for potential pathogens had entirely negative results.
-
-```{r figure_5, layout ="l-screen-inset", echo=FALSE}
+```{r figure_5, echo=FALSE}
 
@@
 # Discussion    
+We identified 124 rodent trapping studies conducted in the West African region since 1965.
+These studies were conducted in 14 countries with 54 studies providing information about zoonotic pathogens in trapped rodents.
+Studies were conducted in all West African countries except Gambia and Togo with between- and within-country variability in the number of trap nights conducted.
+Over 70,000 rodents identified to genus or species level were trapped from 112 species of Rodentia.
+Thirty-two potential pathogens were investigated across all studies, 5 rodent species were assayed for at least 18 potential pathogens.
+We identify several important limitations to available data, potentially biasing previous zoonotic risk studies, first, in the timing of studies, and the locations in which they've been conducted, second, the locations of species presence and third, the selection of pathogens to assay.
 
-  * 124 trapping studies were identified across 14 West African countries. 54 studies provided additional information about potential zoonotic pathogens of trapped rodents.
+First, timing and location of trapping studies can introduce bias to inference from these data when investigating the risk of zoonotic disease emergence.
+Most studies have been conducted within the last twenty years.
+Human population, intensification of agriculture and urbanisation have proceeded rapidly over this time across West Africa.
+These changes modify the available habitats for rodent species increasing the prevalence of commensal species that are more likely to be reservoirs of zoonoses.
+Not accounting for the time period in which data were collected will introduce bias into inferences made for zoonotic emergence based on reservoir species distributions.
+Further, several regions, namely North West Senegal, Eastern Sierra Leone, Southern Ghana, Southern Benin, South West Nigeria  Central Nigeria have been sampled at much higher intensity than elsewhere in West Africa.
+Studies modelling risk of zoonoses without incorporating this important sampling heterogeneity will introduce bias when inferring current or projected risk.
+
+Second, rodent species were found to have wider distributions than the ranges reported by IUCN data.
+Trapping studies were conducted from more regions across West Africa than were available from GBIF records and where therefore able to report presence and absence of species at greater spatial resolution.
+Further, the additional data produced from species' absence and pseudoabsence highlighted the variability of rodent presence within their ranges across the region.
+Failing to incorporate this heterogeneity in rodent reservoir species' distribution can importantly bias estimations of the risk of zoonotic disease emergence.
+For example, the apparent absence of *Mastomys natalensis* at several trapping sites in Nigeria may explain some of the observed spatial variation in Lassa fever spillover that would not be expected based on it's IUCN proposed range or presence only data from GBIF.
+
+Third, while up to 32 potential zoonoses were investigated only four were investigated in more than three studies.
+Of these four studies conducting assays for *Lassa mammarenavirus* or other Arenaviridae where the most widely conducted in regards to number of countries and number of study sites within countries.
+For example, *Toxoplasma sp.* were found to be prevalent in rodent hosts within Senegal and despite a significant burden of human disease no studies investigated the prevalence of this pathogen in other countries of West Africa [https://www.nature.com/articles/s41598-020-69078-9].
+The limited geographic scope of sampling for these pathogens hampers efforts to understand the risk of zoonotic spillover across this heterogeneous region.
+
+Finally, we found that only five of the 112 rodent species trapped had been tested for more than half of the potential pathogens.
+There is increasing emphasis on the role of co-infection and multi-species pathogen transmission systems on zoonotic disease spillover risk which depends on comprehensive testing of both known and un-known reservoir species.
+More comprehensive testing of potential reservoir species across West Africa is required to better understand the role of individual species in disease systems.
+This will also reduce the risk of bias in predictive risk modelling by better characterising host-pathogen associations in species rich environments.
+
+## Implications
+
+## Limitations
+
+## Future research
+
+# Conclusion
+
+# Notes
+
+First, we found that study locations were biased towards regions with higher human population densities and were more likely to be conducted within habitats associated with anthropogenic land use change.
+Second, we were able to obtain contextually rich information on the presence and absence of 147 small-mammal species to compare to IUCN and GBIF distributions and investigate spatial bias in species distributions in these datasets.
+Third, we synthesised data on the presence and absence of zoonotic pathogens detected within small-mammals from these studies and identified biases in the spatial and species sampling across West Africa.
+Finally, we synthesise host-pathogen associations from the included studies.
+
+## Spatial and habitat biases in rodent sampling
+
+Between countries there were large differences in the number of studies performed in each country and the intensity of trapping.
+Important heterogeneity was observed within countries with most countries sparsely sampled, Senegal and Sierra Leone were most comprehensively sampled with studies conducted in all but 7 level-2 administrative regions.
+
+We found a non-linear positive relationship between human population density and the trap-night density of studied regions, suggesting that studies were preferentially conducted in regions of higher human habitation.
+Further, we found that regions consisting of habitat types associated with anthropogenic disturbance (e.g.
+mosaic vegetation, cropland) were more commonly sampled.
+Together, these results suggest that rodent sampling in West Africa is limited by biases introduced by the selection of study sites that impacts the inference and generalisability of results.
+This can importantly limit the applicability of out-of-sample predictions generated from models dependent on these data.
+
+## Rodent presence and absence
+
+Included studies trapped over 70,000 rodents identified to genus or species level, 112 species of Rodentia were reported.
+The most commonly trapped species were the commensal rodents *Mastomys natalensis*, *Rattus rattus*, *Mastomys erythroleucus* and *Mus musculus*.
+Rodents were more frequently trapped in or around buildings (13% trap success) compared with other habitats (3%).
+This difference potentially explains some of the observed bias in habitat selection for studies as more rodents are trapped for the same trapping effort.
+However, this may also suggest that rodents residing in human dominated landscapes are less "trap-shy" or that there is an increased abundance of rodents in these settings.
+Together these factors complicate comparisons made between urban and non-urban studies.
+
+The current study provided further information than is available from GBIF and IUCN.
+Absence and pseudo-absence of species was produced from included studies and provides important additional information compared to presence only data.
+For the three most commonly trapped species, *Mastomys natalensis*, *Rattus rattus* and *Mastomys erythroleucus* we observed absences interspersed with presences suggesting that populations are less spatially heterogeneous than previously assumed.
+Acknowledging these heterogeneities in spatial distribution is important for modelling of zoonotic risk based on host species distributions [ref].
+These heterogeneities are not clearly captured in GBIF and IUCN data.
+GBIF data across West Africa demonstrates spatial bias, for example Togo
+
+## Potential journals
+
+Philosophical transactions of the royal society B
+Ecology letters
+Emerging Microbes and Infection
+PLoS one/PLoS NTD
+The American journal of tropical medicine and hygiene
+Scientific reports 
+Vector borne and zoonotic diseases

### 2021-10-05 reports/_notes.Rmd
from mac

--- 2021-10-04 reports/manuscript_v2.Rmd
@@
 ---
-title: "A scoping review of rodent trapping studies to understand the prevalence of zoonotic disease in West Africa."
-description: |
-date: "`r Sys.Date()`"
-author: 
-  - first_name: "David"
-    last_name: "Simons"
-    url: https://www.dsimons.org
-    orcid_id: 0000-0001-9655-1656
-output: 
-  distill::distill_article:
-    toc: true
-    toc_depth: 3
-    toc_float: true
-  # word_document:
-  # pdf_document:
-knit: (function(inputFile, encoding) {
-    rmarkdown::render(inputFile, encoding = encoding, output_dir = here::here())
-  })
-always_allow_html: yes
-bibliography: "`r here::here('citations', 'manuscript_citations.bib')`"
+title: "Notes"
+author: "David Simons"
+date: "04/10/2021"
+output: pdf_document
 ---
-```{r, echo=FALSE}
-source(here::here("scripts", "manuscript_library.R"))
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
 ```
-
-# Abstract
-
-Rodents are important reservoirs of zoonotic infectious diseases; 255 species are known hosts of 85 zoonotic pathogens.
-These reservoir species are globally distributed with West Africa containing multiple known and several further predicted reservoir species.
-Typically International Union for Conservation of Nature (IUCN) species distribution maps and Global Biodiversity Information Facility (GBIF) presence maps are used to determine regions at risk for zoonotic spillover events, however, these datasets are biased by incomplete sampling.
-To investigate the impact of rodent species sampling heterogeneity on zoonotic spillover risk we have systematically reviewed rodent trapping studies to produce a contextually rich dataset that can be used to explore this risk at finer spatial resolution.
-Here we show that sampling of rodents and their pathogens in the West African region are spatially biased by country and trapping habitat.
-We found that rodent trapping effort was associated with regional population density and was increased in habitats modified by human activity.
-We produce updated rodent species ranges compared to available IUCN maps and enrich GBIF data by including locations of rodent species absence and trapping effort.
-Furthermore, we report the spatial bias in the investigation of four important rodent zoonoses, Arenaviridae, *Borellia sp.*, *Bartonella sp.* and *Toxoplasma sp.*.
-Our results highlight that incorporating sampling bias is important when assessing the risk for zoonotic spillover events from rodents.
-The synthesis of contextually rich rodent trapping data contributes important information that is lacking in IUCN distribution maps and GBIF species presence data.
-We anticipate this dataset can support the production of more complete spatial risk assessments of zoonotic spillover events.
-For example, the inclusion of absence data and trapping effort can help identify regions where data sparsity may produce inappropriately predicted risk.
-Furthermore, these data identify regions in need of comprehensive study for our four targeted rodent zoonoses, Arenaviridae, *Borellia sp.*, *Bartonella sp.* and *Toxoplasma sp.*.
-
-# Introduction
-
-The potential effect of zoonotic infectious diseases outbreaks on human health and societies has been dramatically highlighted through the ongoing SARS-CoV-2 pandemic and recent Ebola virus outbreaks.
-The number of zoonotic spillover events are projected to increase under intensifying anthropogenic pressure such as, increased human populations [@allen_global_2017], increasing urbanisation [@hassell_urbanization_2017], and global climate change [@morse_prediction_2012].
-In addition, two taxa are proposed to contribute to the greatest number of potential zoonotuc pathogens globally, rodents (Rodentia) and bats (Chiroptera) [@han_rodent_2015].
-Of 2,220 extant rodent species, 244 (10.7%) are described as being reservoirs of 85 zoonoses [@han_global_2016].
-Specifically, West Africa has previously been identified as region at increased hazard for rodent-borne zoonotic spillover events under different projected scenarios [@garcia-pena_land-use_2021].
-Rodents are implicated in the transmission pathways of several human diseases in this region, including, Lassa fever, Schistosomiasis and Leptospirosis.
-
-Rodent species form diverse assemblages, which provide important and beneficial ecosystem services including pest regulation and seed dispersal [@fischer_ecosystem_2018].
-The role of rodent species' in zoonotic infectious disease spillover or novel pathogen emergence can be viewed as an ecosystem disservice.
-Rodents typically demonstrate “fast” life histories [@dobson_fast_2007], these traits are also associated with a species being a reservoir of zoonotic pathogens [@han_rodent_2015; @albery_fast-lived_2021].
-Further, these traits are prevalent in species that thrive in human dominated landscapes where they displace species that are less likely to be reservoirs of zoonotic pathogen [@gibb_zoonotic_2020].
-The widespread occurrence of reservoir species and their proximity to human activity make the description of rodent species assemblages and host-pathogen interactions vitally important to understanding the risk of zoonotic spillover and novel pathogen emergence.
-
-Rodent trapping studies provide a useful method to describe rodent population structures and survey for potential zoonoses.
-Studies have been conducted in West Africa to both identify novel potential zoonotic pathogens within rodents [@usaid_one_2021] and to investigate the prevalence and burden of known pathogens within known rodent hosts (e.g.
-for Lassa fever [@fichet-calvet_diversity_2009] and Schistosomiasis [@catalano_multihost_2020]).
-These studies provide contextually rich information on when, where and under what conditions rodents were trapped that are typically missing from the global datasets [@bovendorp_optimising_2017].
-Despite this, these global datasets (e.g.
-IUCN, GBIF and GIDEON) are typically used to inform species distribution maps and zoonotic disease spillover predictions [@han_global_2016; @smith_global_2014; @pigott_mapping_2014].
-
-These studies have been used to identify potential geographic hotspots where virus and host species diversity may be expected to be at its greatest to predict regional zoonotic disease spillover risks.
-However, there remains the potential for important confounding in these spatial distributions through bias generated by study design and selection of sampling sites.
-For example, systematically increased sampling (e.g.
-more intensive studies over longer time periods) or over-representation of certain habitats (e.g.
-peri-urban landscapes) could lead to an apparent association between locations and risk that is driven by these factors rather than an underlying host and virus association [@wille_how_2021; @gibb_mammal_2021].
-Conversely some regions may not be sampled adequately and therefore under-represented in these datasets due to sparse human populations or inaccessible habitats leading to these regions being reported to be at low risk of zoonotic disease spillover events.
-
-Here, we identify rodent trapping studies performed across West Africa and identify the location and habitat types in which they have been conducted, the pathogens assessed and the host-pathogen associations that have been reported in order to quantify the potential bias and to identify regions requiring further focussed investigation.
-
-# Methods
-
-## Literature search
-
-We conducted a search in Ovid MEDLINE, Web of Science (Core collection and Zoological Record), JSTOR, BioOne, African Journals Online, Global Health and the pre-print servers, BioRxiv and EcoEvoRxiv for the following terms as keywords, no date limits were set:
-
-  1. Rodent OR Rodent trap*
-  2. West Africa (or the individual countries)
-  3. 1. AND 2.
-
-We searched other resources including the UN Official Documents System, Open Grey, AGRIS FAO and Google Scholar using combinations of the above terms. Searches were run on 2021-03-01.
-
-We included studies if they met all of the following inclusion criteria; i) reported findings from trapping studies where the target was a small mammal, ii) described the type of trap used or the length of trapping activity or the location of the trapping activity, iii) included trapping activity from at least one West African country, iv) recorded the genus or species of trapped individuals, v) were published in a peer-reviewed journal or as a pre-print on a digital platform or as a report by a credible organisation.
-We excluded studies if they met any of the following exclusion criteria: i) reported data that were duplicated from a previously included study, ii) no full text available, iii) not available in English.
-One reviewer screened titles, abstracts and full texts against the inclusion and exclusion criteria.
-At each stage, a random subset (10%) was reviewed by a second reviewer.
-
-## Data extraction
-
-We extracted data from eligible studies into a Google sheets document.
-Extracted variables included i) study identifiers; ii) study aims; iii) trapping methodology; iv) geolocation data; v) method of speciation; vi) trapping locations and dates; vii) trapped species; viii) number of trap-nights and ix) microorganisms/pathogens of interest.
-The data extraction tool is archived and available in Supplementary Material 1.
-
-### Location of rodent trapping studies and habitats studied
-
-We extracted GPS locations for the most precise location presented (i.e.
-trap, trap-line, study site or study region).
-We extracted coordinates in the format reported and converted them to decimal degrees.
-We recorded the habitat classification scheme used by a study (e.g.
-IUCN Habitat Classification Scheme (Version 3.1)).
-For studies not using standardised habitat recording, the explicit description of the habitat in which the trap was placed were extracted.
-For studies reporting multiple habitat types (e.g.
-rice field, corn field and vegetable garden) for a single trap, trap-line or trapping grid, a higher order classification of habitat type was recorded (e.g.
-agricultural land).
-
-### Rodent presence, absence, abundance
-
-We mapped genus and species names to the species names used in the Global Biodiversity Information Facility (GBIF) taxonomy  [@gbif_the_global_biodiversity_information_facility_global_2021].
-We extracted information on the presence, absence and number of trapped individuals.
-For studies reporting on all trapped individuals (i.e.
-not those only reporting on the presence of a specific species of interest), the pseudo-absence of a species reported as present elsewhere in the study was explicitly recorded as an absence at that trap location.
-
-### Pathogen presence and absence
-
-We extracted data on all pathogens assayed in studies investigating rodents for potential zoonoses.
-The number of rodents tested and the number of positive or negative samples were recorded alongside the type of assay used (e.g.
-Polymerase Chain Reaction (PCR), Enzyme Linked ImmunoSorbent Assay (ELISA) or viral culture).
-If studies reported indeterminate results this was noted.
-Where possible, pathogens were identified to species level.
-However, where an assay only allowed for attribution to a family of viruses or bacteria, the higher order grouping was used (e.g.
-Arenaviridae for a PCR using a non-specific arenavirus primer).
-
-## Analysis
-
-### Location and habitats of rodent trapping to investigate potential biases
-
-We summarised the number of studies, the year in which trapping occurred and the country in which they were conducted.
-We used the GPS coordinates of single trap, trapping grid/line or study site and the number of trap nights to calculate trapping effort (trap night density) within level 2 administrative areas in West Africa.
-For studies not reporting the number of trap nights we imputed the number based on the median trap success rate from studies which reported the number of trap nights (matched to building or non-building based study sites).
-The median trap success rate for all rodents at a defined trap site was calculated separately for trap sites which included built environments and non-built environments.
-The number of rodents trapped at a trap site was then used in combination with these two values of trap success to impute the number of trap nights for trap sites with no reported trap nights.
-We summarised the habitat types of trap sites based on information reported in the study.
-
-For studies investigating rodent zoonoses we compared the location of trapping sites with SEDAC Global Population Density estimates for 2005 - the median year studies were commenced.
-We used a Generalised Additive Model (GAM) incorporating a spatial interaction term to investigate the association of number of trap nights and human population density.
-The model structure was specified as:
-
-  + *Trap night density* ~ *Tweedie*(log(Population density (2005)) + (Longitude * Latitude))
-
-We obtained land cover classifications from the European Space Agency Copernicus dataset at 300m^2^ spatial resolution (2005) we extracted the proportion of land cover classes within all regions of West Africa and all regions in which rodent trapping occurred for investigation of zoonotic diseases.
-We compared the proportion of possible land cover classes with those of regions where trapping occurred.
-
-Finally, we mapped the presence and absence of rodent species and compared this to the presence and absence reported by both GBIF and IUCN to give a measure of the extent of each species range in which they have been sampled.
-
-### Rodent pathogen associations
-
-We summarised the presence and absence of microorganisms, the assays used, their host species and the locations from where the samples were obtained.
-We investigated the association between rodent species and the detection of potential pathogens and report the proportion of positive and negative tests for each species and pathogen pair.
-
-# Results
-
-## Included studies
-
-We identified 4,282 relevant citations, with 126 rodent trapping studies included.
-The earliest included studies were published in 1974 with increasing numbers of studies being performed annually since 2005.
-The median length of rodent trapping activity was one year (IQR 0-2 years).
-The median time from completion of rodent trapping to publication is 3 years (IQR 2-5 years) (Figure 1.).
-
-```{r, figure_1, layout = "l_page", echo=FALSE}
-
-if(knitr::is_html_output()) {
-  knitr::include_graphics(here("figures", "Fig_1_Panel_A.png"), dpi = 320)
-} else {
-  knitr::include_graphics(here("figures", "Fig_1_Panel_A.png"))
-}
-
-```
-
-*Figure 1*: Each row represents one of the 126 included studies, green points designate the first year of data collection, blue points designate the end of data collection.
-For studies completed within one year the blue point completely overlies the green.
-Studies with a transparent grey point did not report the year in which trapping was conducted.
-The year of publication is shown by a red point.
-
-## Location and habitats of rodent trapping studies to investigate potential biases
-
-Rodent trapping took place at 1,331 trap sites with at least one trap site recorded from 14 West African countries.
-No rodent trapping studies were identified from Gambia or Togo.
-The most frequently studied countries were Senegal, Ghana, Guinea and Nigeria (more than 10 studies each).
-Fewer than 5 studies were conducted in Liberia, Burkina Faso and Guinea-Bissau.
-Thirty-one (25%) studies reported trapping at a single study site, 46 (37%) studies trapped at between two and five study sites, the remaining 49 studies trapped at between six and 93 study sites.
-Trap sites were situated in 273 (19.3%) of 1414 level 2 administrative regions in the 14 West African nations.
-The areas with highest trap night density included the capital cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako) and Ghana (Accra) and the largest cities of Côte d'Ivoire (Abidjan) and Benin (Cotonou).
-Outside of these cities, Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone had the highest density of trap nights (Figure 2.
-A).
-
-Greater numbers of trap nights were conducted in areas with higher population densities (*R~2~* = 0.18) adjusting for spatial autocorrelation.
-In particular Mauritania, Northern Senegal, Eastern Sierra Leone, Eastern Guinea and South West Nigeria were trapped at higher rates than would be expected based on their population densities.
-South West Nigeria, Northern Nigeria, Liberia, Côte d'Ivoire, Ghana, Niger and Burkina Faso were trapped at lower rates than would be expected.
-
-No studies reported trap habitats with reference to a standardised habitat classification scheme.
-Extracted habitat types were grouped into 30 categories (see Supplementary Material 2 for the habitat dictionary).
-At least one habitat was recorded for 17,122 trap sites (95%), with two or more habitats for a single trap site recorded for 4,403 (24%) sites.
-Single trap sites could span multiple habitat types resulting in 22,202 distinct habitat and trap sites.
-The most commonly trapped sites were in or around buildings (29%), in areas described as the rodents “natural habitat” (24%), in agricultural areas (13%) (e.g.
-rice fields, palm plantations), forests (8%) and in the area surrounding buildings (4%).
-
-```{r, figure_2, layout = "l-screen-inset", echo=FALSE}
-
-knitr::include_graphics(here("figures", "Figure_2.png"))
-
-```
-
-*Figure 2*: **Panel A:** Map of West Africa, countries where rodent trapping has occurred are mapped to level 2 administrative areas (where available).
-These regions are coloured by the total number of trap nights performed at trap sites within their boundaries.
-**Panel B:** (n.b.
-will not include map in final manuscript) The number of trap nights conducted is associated with a regions population density.
-The population density for all regions in West Africa is shown in the yellow box plot, the population density and trap night density for each region is show on the purple scatterplot.
-The line of best fit is a GAM model not incorporating spatial interactions.
-The map panel on the right is the product of the GAM model incorporating spatial interactions.
-**Panel C:** For each of the 10 land cover classes from the ESA dataset we measure the proportion of 300m^2^ pixels within a level 2 administrative area.
-Zoonotic trapping studies occurred in regions over-representative for Cropland, Mosaic landscapes and Shrubland while being under-representative for Bare and Sparse vegetation land cover classes.
-
-## Rodent presence, absence, abundance
-
-73,164 small mammals were trapped (592 were trapped outside of West African countries), 2,830 (4%) trapped individuals were identified to order level (Rodentia), 7,760 (11%) were identified to genus level, the remaining 62,574 (85%) were identified to species level.
-In studies reporting the number of trap nights the median trap success rate of traps placed in or around buildings was 13% (IQR 6-24%), this compares to a median trap success rate of 3% (IQR 1-9%) in other habitats.
-The majority of the 147 species trapped were Rodentia (112) (Supplementary Material 3), of these Muridae (82) were the largest family of rodents, followed by, Sciuridae (10), Gliridae and Nesomyidae (both 6), Ctenodactylidae, Anomaluridae, Dipodidae, Hystricidae and Thryonomidae (2 or fewer).
-The remaining 34 species were from the orders of Soricomorpha (30), Erinaceomorpha (2) and Afrosoricida (1).
-
-*Mastomys sp.* were the most commonly trapped genus (27,079, 37.7%), followed by, *Rattus sp.* (11,472, 16.1%), *Mus sp.* (8,624, 12.0%), *Arvicanthis sp.* (5,821, 8.1%) and *Praomys sp.* (5,409, 7.5%).
-*Mastomys natalensis* were the most commonly trapped species (11,116, 17.4%), followed by, *Rattus rattus* (9,959, 15.6%), *Mastomys erythroleucus* (7,386, 11.6%), *Mus musculus* (6,245, 9.8%), *Arvicanthis niloticus* (5,497, 8.6%), *Mastomys huberti* (4,699, 7.4%) and *Praomys daltoni* (1,854, 2.9%).
-
-Locations of points for the presence or absence of the seven most commonly trapped species are shown in Figure 3 (left column).
-Absence include true absences and pseudo-absences dependent on the trapping effort at the site.
-This are shown compared to species presence obtained from GBIF (centre column) and IUCN species range data (left column) (Figure 3).
-
-```{r, figure_3, layout = "l-screen-inset", echo=FALSE}
-
-knitr::include_graphics(here("figures", "Figure_3.png"))
-
-```
-
-*Figure 3*: Each row corresponds to a single rodent species.
-The column on the left shows the presence and absence of a rodent species from the individual studies included in this review.
-The centre column shows the presence of a rodent species obtained from GBIF (September 2021) for records where longitude and latitude have been provided.
-The right column shows the range of rodent species as proposed by the IUCN (2021) (red shaded area), overlaid are the presence points from both this review and GBIF records.
-
-## Pathogen presence and absence
-
-Sixty-two studies presented data on microorganisms that infect or are carried by small mammal species in West Africa.
-Seven studies solely investigated pathogens of rodents, including Hydatigera species (previously Taenia species) and Trichuris species.
-Zoonotic pathogens and potentially zoonotic micro-organisms were investigated in 55 studies.
-Thirty-two microorganisms were tested for, 8 of these were specified to species level, the remaining 24 were specified at higher taxonomic classifications.
-
-Polymerase Chain Reaction (PCR) was used in 32 studies to detect the presence of 22 different species or families of microorganisms from 21,953 rodent samples.
-The most common microorganisms assayed for using PCR were *Lassa mammarenavirus* or other Arenaviridae (31%) the bacteria *Borrelia sp.* (11.3%) and *Bartonella sp.* (6.5%), followed by Usutu virus (6.4%) and Hantaviridae (5.6%).
-Molecular tests (antibody or antigen based) were used in 11 studies to detect the presence of 9 different species or families of microorganisms from 11,430 rodent samples.
-The most common pathogens assessed for with serology were *Lassa mammarenavirus* or other Arenaviridae (78%), *Toxoplasma gondii* (10.6%), *Borrelia sp.* (6.3%) and *Leptospirosis sp.* (2.2%).
-Histological or direct visualisation assays were used in 8 studies investigating 11 parasitic or bacterial species from 11,229 rodent samples.
-The most common pathogens assessed for with histology or direct visualisation were *Borrelia sp.* (48.7%), *Schistosoma sp*.
-(20.4%) and other parasites.
-Direct culture of *Lassa mammarenavirus* or Leishmania species were reported from 3 studies to detect the presence of these pathogen in 643 rodent samples.
-
-Most studies tested for a single microorganisms (39), 16 studies tested for two or more microorganisms.
-The most frequently investigated microorganisms were *Lassa mammarenavirus* (17 studies) or members of the Arenaviridae family (10), the spirochete bacteria *Borrelia sp.* (8), *Bartonella sp.* and *Toxoplasma gondii* (4), the remaining 25 microorganisms were reported in three or fewer studies.
-
-The location of trapped rodents that were tested for the four most commonly assayed microorganisms did not encompass the entire range of these hosts.
-The presence and absence of Arenaviridae, predominantly *Lassa mammarenavirus*, in rodents was investigated in Benin, Côte d'Ivoire, Guinea, Mali, Nigeria and Sierra Leone.
-No studies investigated the prevalence of these pathogens in hosts from Burkina-Faso, Guinea-Bissau, Liberia, Niger, Senegal and Togo.
-Arenaviridae were detected in rodents from 17 of 54 species or genera assayed in all the countries it was investigated except Benin (Figure 4 top row).
-The presence and absence of *Bartonella sp.* was investigated in Mali and Nigeria and Senegal with positive samples detected in 11 of 11 species or genera of rodents obtained at most sites from these countries (Figure 4 second row).
-The presence and absence of *Borrelia sp.* was investigated in Benin, Mali, Mauritania, Niger and Senegal.
-*Borrelia sp.* were detected in 8 of 44 species or genera of rodents assayed in Mali with negative samples obtained from Benin, Mauritania, Niger and Senegal (Figure 4 third row).
-The presence of *Toxoplasma gondii* was only investigated in Senegal and was detected in 6 of 8 rodent species assayed (Figure 4 fourth row).
-
-```{r, figure_4, echo=FALSE}
-
-knitr::include_graphics(here("figures", "Figure_4a.png"))
-knitr::include_graphics(here("figures", "Figure_4b.png"))
-knitr::include_graphics(here("figures", "Figure_4_legend.png"))
-
-```
-
-*Figure 4*: Presence/absence plots at each unique trapping site for the four most commonly assayed microorganisms Arenaviridae (top), *Bartonella sp.* (second row), *Borrelia sp.* (third row) and *Toxoplasma gondii* (fourth row).
-The tables to the right of each map highlight the 10 (or number applicable) most commonly positive and tested rodent species and genera assayed.
-
-## Host-pathogen associations
-
-Ninety-seven rodent species were investigated for the presence of a zoonotic pathogen, 42,940 assays were performed on 32,014 individual rodent samples.
-The rodent species most commonly assayed for zoonotic pathogens were, *Rattus rattus* (n = 2,977) assessed for 24 pathogens, *Mus musculus* (n = 3,402) for 23, *Mastomys natalensis* (n = 7,189) and *Mastomys erythroleucus* (n = 3,013) both 19 and *Arvicanthis niloticus* (n = 3,840) for 18.
-All remaining species were investigated for 10 or fewer pathogens.
-
-*Lassa mammarenavirus* was detected in 14 species of rodents and Crocidura genera.
-*Mastomys natalensis* was the most common species to test positive (Figure 4.).
-Fifty-four species and genera were tested for arenaviruses, with 37 testing negative.
-*Bartonella sp.* infections were identified in all 11 species of rodents and shrews that were tested.
-*Mastomys erythroleucus* was the most common species to test positive.
-*Borrelia sp.* infections were identified in 13 species of rodents and shrews.
-*Mastomys natalensis* was the most common species to test positive.
-Forty-four species and genera were tested for *Borrelia sp.* with 36 testing negative.
-*Toxoplasma gondii* was detected in 6 species of rodents and shrews, with all individuals from two species *Arvicanthis niloticus* and *Praomys daltoni* testing negative.
-*Mus musculus* was the most common species to test positive.
-Finally, 38 species of Rodentia, 3 species of Soricomorpha and 2 species of Erinaceomorpha that were tested for pathogens had entirely negative results.
-
-```{r figure_5, echo=FALSE}
-
-knitr::include_graphics(here("figures", "Figure_5.png"))
-
-```
-
-*Figure 5*: The percentage of assays from each included species and microorganism tested for is represented by colour.
-The size of a point relates to the squareroot of the number of individuals assayed.
-The plot on the left is for species/genera assayed for four or more microorganisms, the plot on rights are for species/genera assayed for 3 or less microorganisms.
-
-# Discussion    
-We identified 124 rodent trapping studies conducted in the West African region since 1965.
-These studies were conducted in 14 countries with 54 studies providing information about zoonotic pathogens in trapped rodents.
-Studies were conducted in all West African countries except Gambia and Togo with between- and within-country variability in the number of trap nights conducted.
-Over 70,000 rodents identified to genus or species level were trapped from 112 species of Rodentia.
-Thirty-two potential pathogens were investigated across all studies, 5 rodent species were assayed for at least 18 potential pathogens.
-We identify several important limitations to available data, potentially biasing previous zoonotic risk studies, first, in the timing of studies, and the locations in which they've been conducted, second, the locations of species presence and third, the selection of pathogens to assay.
-
-First, timing and location of trapping studies can introduce bias to inference from these data when investigating the risk of zoonotic disease emergence.
-Most studies have been conducted within the last twenty years.
-Human population, intensification of agriculture and urbanisation have proceeded rapidly over this time across West Africa.
-These changes modify the available habitats for rodent species increasing the prevalence of commensal species that are more likely to be reservoirs of zoonoses.
-Not accounting for the time period in which data were collected will introduce bias into inferences made for zoonotic emergence based on reservoir species distributions.
-Further, several regions, namely North West Senegal, Eastern Sierra Leone, Southern Ghana, Southern Benin, South West Nigeria  Central Nigeria have been sampled at much higher intensity than elsewhere in West Africa.
-Studies modelling risk of zoonoses without incorporating this important sampling heterogeneity will introduce bias when inferring current or projected risk.
-
-Second, rodent species were found to have wider distributions than the ranges reported by IUCN data.
-Trapping studies were conducted from more regions across West Africa than were available from GBIF records and where therefore able to report presence and absence of species at greater spatial resolution.
-Further, the additional data produced from species' absence and pseudoabsence highlighted the variability of rodent presence within their ranges across the region.
-Failing to incorporate this heterogeneity in rodent reservoir species' distribution can importantly bias estimations of the risk of zoonotic disease emergence.
-For example, the apparent absence of *Mastomys natalensis* at several trapping sites in Nigeria may explain some of the observed spatial variation in Lassa fever spillover that would not be expected based on it's IUCN proposed range or presence only data from GBIF.
-
-Third, while up to 32 potential zoonoses were investigated only four were investigated in more than three studies.
-Of these four studies conducting assays for *Lassa mammarenavirus* or other Arenaviridae where the most widely conducted in regards to number of countries and number of study sites within countries.
-For example, *Toxoplasma sp.* were found to be prevalent in rodent hosts within Senegal and despite a significant burden of human disease no studies investigated the prevalence of this pathogen in other countries of West Africa [https://www.nature.com/articles/s41598-020-69078-9].
-The limited geographic scope of sampling for these pathogens hampers efforts to understand the risk of zoonotic spillover across this heterogeneous region.
-
-Finally, we found that only five of the 112 rodent species trapped had been tested for more than half of the potential pathogens.
-There is increasing emphasis on the role of co-infection and multi-species pathogen transmission systems on zoonotic disease spillover risk which depends on comprehensive testing of both known and un-known reservoir species.
-More comprehensive testing of potential reservoir species across West Africa is required to better understand the role of individual species in disease systems.
-This will also reduce the risk of bias in predictive risk modelling by better characterising host-pathogen associations in species rich environments.
-
-## Implications
-
-## Limitations
-
-## Future research
-
-# Conclusion
 

### 2021-10-05 reports/manuscript_v2.Rmd
from mac

@@
 output: 
-  distill::distill_article:
+  # distill::distill_article:
+  #   toc: true
+  #   toc_depth: 3
+  #   toc_float: true
+  # word_document:
+  pdf_document:
     toc: true
     toc_depth: 3
-    toc_float: true
-  # word_document:
-  # pdf_document:
 knit: (function(inputFile, encoding) {
@@
 ---
-```{r, echo=FALSE}
+```{r, echo=FALSE, include=FALSE}
 source(here::here("scripts", "manuscript_library.R"))
@@
 
-```{r, figure_1, layout = "l_page", echo=FALSE}
+```{r, figure_1, echo=FALSE, out.width="100%"}
 
@@
 
-```{r, figure_2, layout = "l-screen-inset", echo=FALSE}
+```{r, figure_2, layout = "l-screen-inset", echo=FALSE, out.width="100%"}
 
@@
 
-```{r, figure_3, layout = "l-screen-inset", echo=FALSE}
+```{r, figure_3, layout = "l-screen-inset", echo=FALSE, out.width="100%"}
 
@@
 
-```{r, figure_4, echo=FALSE}
+```{r, figure_4, echo=FALSE, out.width="100%"}
 
@@
 
-```{r figure_5, echo=FALSE}
+```{r figure_5, echo=FALSE, out.width="100%"}
 
@@
 
+## Limitations
+
+We identified several limitations to our approach.
+We identified studies from a diverse range of sources, however, it remains possible that rodent surveys have been performed for additional purposes that would not be captured in a survey of the scientific and "grey" literature, for example, risk assessments or ecological assessments for development work may not have been identified through our search.
+We were limited by the reporting from included studies, particularly for measuring trapping effort and identifying habitat types trapped.
+We imputed trapping effort for studies that did not report any measure of effort, removing these studies had no meaningful effect on the association of human population density and trapping effort in sensitivity analysis.
+Habitat type and human population density were obtained from 2005, the median year in which studies were conducted.
+Changes in both human populations and land use in the studied period (1969-2020) have been large across West Africa, this introduces the possibility that these values were not appropriate for the region at the time when the rodent trapping study was conducted.
+
 ## Implications
 
-## Limitations
-
-## Future research
+We have identified important biases in the locations of studies, the trapping effort for different rodent species and the selection of zoonotic pathogens to assay across West Africa.
+The use of these data to infer risk of zoonotic spillover or novel disease emergence is likely inappropriate unless these systematic biases are addressed.
+We hope that the current synthesis of extant data will encourage further comprehensive ecological studies on rodent species assemblages and their pathogens.
+One priority would be to better understand the true distribution of known rodent reservoirs (particularly *Mastomys natalensis*) across the region with particular emphasis made to the habitat in which they reside and the structure of rodent assemblages in the region.
 
@@
 
-# Notes
+In conclusion we present a comprehensive synthesis of rodent trapping studies across West Africa, we identify important biases in the sampling of rodents and their pathogens that limits inference able to be drawn from current evidence.
+We identify evidence gaps in the current literature and highlight geographic regions, habitat types and host-pathogen associations that would benefit from targeted research effort.
 
-First, we found that study locations were biased towards regions with higher human population densities and were more likely to be conducted within habitats associated with anthropogenic land use change.
-Second, we were able to obtain contextually rich information on the presence and absence of 147 small-mammal species to compare to IUCN and GBIF distributions and investigate spatial bias in species distributions in these datasets.
-Third, we synthesised data on the presence and absence of zoonotic pathogens detected within small-mammals from these studies and identified biases in the spatial and species sampling across West Africa.
-Finally, we synthesise host-pathogen associations from the included studies.
-
-## Spatial and habitat biases in rodent sampling
-
-Between countries there were large differences in the number of studies performed in each country and the intensity of trapping.
-Important heterogeneity was observed within countries with most countries sparsely sampled, Senegal and Sierra Leone were most comprehensively sampled with studies conducted in all but 7 level-2 administrative regions.
-
-We found a non-linear positive relationship between human population density and the trap-night density of studied regions, suggesting that studies were preferentially conducted in regions of higher human habitation.
-Further, we found that regions consisting of habitat types associated with anthropogenic disturbance (e.g.
-mosaic vegetation, cropland) were more commonly sampled.
-Together, these results suggest that rodent sampling in West Africa is limited by biases introduced by the selection of study sites that impacts the inference and generalisability of results.
-This can importantly limit the applicability of out-of-sample predictions generated from models dependent on these data.
-
-## Rodent presence and absence
-
-Included studies trapped over 70,000 rodents identified to genus or species level, 112 species of Rodentia were reported.
-The most commonly trapped species were the commensal rodents *Mastomys natalensis*, *Rattus rattus*, *Mastomys erythroleucus* and *Mus musculus*.
-Rodents were more frequently trapped in or around buildings (13% trap success) compared with other habitats (3%).
-This difference potentially explains some of the observed bias in habitat selection for studies as more rodents are trapped for the same trapping effort.
-However, this may also suggest that rodents residing in human dominated landscapes are less "trap-shy" or that there is an increased abundance of rodents in these settings.
-Together these factors complicate comparisons made between urban and non-urban studies.
-
-The current study provided further information than is available from GBIF and IUCN.
-Absence and pseudo-absence of species was produced from included studies and provides important additional information compared to presence only data.
-For the three most commonly trapped species, *Mastomys natalensis*, *Rattus rattus* and *Mastomys erythroleucus* we observed absences interspersed with presences suggesting that populations are less spatially heterogeneous than previously assumed.
-Acknowledging these heterogeneities in spatial distribution is important for modelling of zoonotic risk based on host species distributions [ref].
-These heterogeneities are not clearly captured in GBIF and IUCN data.
-GBIF data across West Africa demonstrates spatial bias, for example Togo
-
-## Potential journals
-
-Philosophical transactions of the royal society B
-Ecology letters
-Emerging Microbes and Infection
-PLoS one/PLoS NTD
-The American journal of tropical medicine and hygiene
-Scientific reports 
-Vector borne and zoonotic diseases
+# References

### 2021-10-05 reports/manuscript_v2.Rmd
need to sort figures

@@
 title: "A scoping review of rodent trapping studies to understand the prevalence of zoonotic disease in West Africa."
-description: |
 date: "`r Sys.Date()`"
 author: 
-  - first_name: "David"
-    last_name: "Simons"
-    url: https://www.dsimons.org
-    orcid_id: 0000-0001-9655-1656
+  - "David Simons"
+    # url: https://www.dsimons.org
+    # orcid_id: 0000-0001-9655-1656
 output: 
@@
 The number of zoonotic spillover events are projected to increase under intensifying anthropogenic pressure such as, increased human populations [@allen_global_2017], increasing urbanisation [@hassell_urbanization_2017], and global climate change [@morse_prediction_2012].
-In addition, two taxa are proposed to contribute to the greatest number of potential zoonotuc pathogens globally, rodents (Rodentia) and bats (Chiroptera) [@han_rodent_2015].
+In addition, two taxa - rodents (Rodentia) and bats (Chiroptera) - are proposed to contribute to the greatest number of zoonotic pathogens and the greatest risk of novel pathogen emergence [@han_rodent_2015].
 Of 2,220 extant rodent species, 244 (10.7%) are described as being reservoirs of 85 zoonoses [@han_global_2016].
 Specifically, West Africa has previously been identified as region at increased hazard for rodent-borne zoonotic spillover events under different projected scenarios [@garcia-pena_land-use_2021].
-Rodents are implicated in the transmission pathways of several human diseases in this region, including, Lassa fever, Schistosomiasis and Leptospirosis.
+Rodents are implicated in the transmission pathways of several human diseases in this region, including, Lassa fever, Schistosomiasis and Leptospirosis [@meerburg_rodent-borne_2009].
 
@@
 Further, these traits are prevalent in species that thrive in human dominated landscapes where they displace species that are less likely to be reservoirs of zoonotic pathogen [@gibb_zoonotic_2020].
-The widespread occurrence of reservoir species and their proximity to human activity make the description of rodent species assemblages and host-pathogen interactions vitally important to understanding the risk of zoonotic spillover and novel pathogen emergence.
+The widespread occurrence of reservoir species and their proximity to human activity make the description of rodent species assemblages and host-pathogen interactions vitally important to understanding the risk of zoonotic spillover and novel pathogen emergence [@han_global_2016].
 
@@
 These studies have been used to identify potential geographic hotspots where virus and host species diversity may be expected to be at its greatest to predict regional zoonotic disease spillover risks.
-However, there remains the potential for important confounding in these spatial distributions through bias generated by study design and selection of sampling sites.
+However, there remains the potential for important confounding in these spatial distributions through bias generated by study design and selection of sampling sites [@plowright_sampling_2019].
 For example, systematically increased sampling (e.g.
@@
 peri-urban landscapes) could lead to an apparent association between locations and risk that is driven by these factors rather than an underlying host and virus association [@wille_how_2021; @gibb_mammal_2021].
-Conversely some regions may not be sampled adequately and therefore under-represented in these datasets due to sparse human populations or inaccessible habitats leading to these regions being reported to be at low risk of zoonotic disease spillover events.
+Conversely some regions may not be sampled adequately and therefore under-represented in these datasets due to sparse human populations or inaccessible habitats this may result in these regions being reported inappropriately as at low risk of novel pathogen emergence or zoonotic disease spillover events.
 
@@
 Extracted variables included i) study identifiers; ii) study aims; iii) trapping methodology; iv) geolocation data; v) method of speciation; vi) trapping locations and dates; vii) trapped species; viii) number of trap-nights and ix) microorganisms/pathogens of interest.
-The data extraction tool is archived and available in Supplementary Material 1.
+The data extraction tool is archived and available in Supplementary 1 or at this [link](https://docs.google.com/spreadsheets/d/1rQYjHhk6uk1PoKZZVsgFlmuqWGicU2tTisk9ddfAwTM/edit?usp=sharing).
 
@@
 We used the GPS coordinates of single trap, trapping grid/line or study site and the number of trap nights to calculate trapping effort (trap night density) within level 2 administrative areas in West Africa.
-For studies not reporting the number of trap nights we imputed the number based on the median trap success rate from studies which reported the number of trap nights (matched to building or non-building based study sites).
-The median trap success rate for all rodents at a defined trap site was calculated separately for trap sites which included built environments and non-built environments.
-The number of rodents trapped at a trap site was then used in combination with these two values of trap success to impute the number of trap nights for trap sites with no reported trap nights.
-We summarised the habitat types of trap sites based on information reported in the study.
+The median trap success rate for all rodents at a defined trap site was calculated separately for trap sites which included built environments and non-built environments due to important differences in trap success between these different environments.
+The number of rodents trapped at a trap site was then used in combination with these values of trap success to impute the number of trap nights for trap sites where there was no record of the number of trap nights.
+We summarised the habitat types of trap sites based on information reported in the study, the dictionary used to match habitat types is available in Supplementary 2.
 
-For studies investigating rodent zoonoses we compared the location of trapping sites with SEDAC Global Population Density estimates for 2005 - the median year studies were commenced.
-We used a Generalised Additive Model (GAM) incorporating a spatial interaction term to investigate the association of number of trap nights and human population density.
+For studies investigating rodent zoonoses we compared the location of trapping sites with SEDAC Global Population Density estimates for 2005 - the median year studies were started [@socioeconomic_data_and_applications_center_gridded_2021].
+We used a Generalised Additive Model (GAM) incorporating a spatial interaction term to investigate the association of number of trap nights and human population density [@pedersen_hierarchical_2019].
 The model structure was specified as:
 
-  + *Trap night density* ~ *Tweedie*(log(Population density (2005)) + (Longitude * Latitude))
+$$
+\text{Trap night density} \sim Tweedie(log(Population density_{2005}) + (Longitude * Latitude))   
+$$
 
-We obtained land cover classifications from the European Space Agency Copernicus dataset at 300m^2^ spatial resolution (2005) we extracted the proportion of land cover classes within all regions of West Africa and all regions in which rodent trapping occurred for investigation of zoonotic diseases.
+We obtained land cover classifications from the European Space Agency Copernicus dataset at 300m^2^ spatial resolution generated using Copernicus Climate Change Service information (2005) we extracted the proportion of land cover classes within all regions of West Africa and all regions in which rodent trapping occurred for investigation of zoonotic diseases.
 We compared the proportion of possible land cover classes with those of regions where trapping occurred.
 
-Finally, we mapped the presence and absence of rodent species and compared this to the presence and absence reported by both GBIF and IUCN to give a measure of the extent of each species range in which they have been sampled.
+Finally, we mapped the presence and absence of rodent species and compared this to the presence and absence produced from GBIF [@gbif_the_global_biodiversity_information_facility_occurrence_2021] and IUCN [@iucn_iucn_2021] datasets to give a measure of the extent of each species range in which they have been sampled.
 
@@
 We identified 4,282 relevant citations, with 126 rodent trapping studies included.
+The list of included studies in available in Supplementary 3 or through the interactive web based application produced to explore this dataset, available [here](https://diddrog11.shinyapps.io/scoping_review_app/).
 The earliest included studies were published in 1974 with increasing numbers of studies being performed annually since 2005.
@@
 
-```{r, figure_1, echo=FALSE, out.width="100%"}
-
-if(knitr::is_html_output()) {
-  knitr::include_graphics(here("figures", "Fig_1_Panel_A.png"), dpi = 320)
-} else {
-  knitr::include_graphics(here("figures", "Fig_1_Panel_A.png"))
-}
-
-```
-
-*Figure 1*: Each row represents one of the 126 included studies, green points designate the first year of data collection, blue points designate the end of data collection.
+![Each row represents one of the 126 included studies, green points designate the first year of data collection, blue points designate the end of data collection.
 For studies completed within one year the blue point completely overlies the green.
 Studies with a transparent grey point did not report the year in which trapping was conducted.
-The year of publication is shown by a red point.
+The year of publication is shown by a red point.](../figures/Fig_1_Panel_A.png)
 
@@
 
+![**A:** Map of West Africa, countries where rodent trapping has occurred are mapped to level 2 administrative areas.
+The colour of the region corresponds to the trap night density per 1000 km^2.
+Colourless regions reported no trapping.
+**B:** This shows the linear predictor from obtained from the GAM.
+Predictions are limited to coordinates are trap sites (coloured areas), uncertainty in the linear predictor is represented by th transparency of the colour.
+Yellow regions represent areas with higher than expected trapping effort for the regional population and purple regions represent areas trapped at lower than expected rates based on their populations](../figures/Figure_2.png)
+
 No studies reported trap habitats with reference to a standardised habitat classification scheme.
@@
 
-```{r, figure_2, layout = "l-screen-inset", echo=FALSE, out.width="100%"}
-
-knitr::include_graphics(here("figures", "Figure_2.png"))
-
-```
-
-*Figure 2*: **Panel A:** Map of West Africa, countries where rodent trapping has occurred are mapped to level 2 administrative areas (where available).
-These regions are coloured by the total number of trap nights performed at trap sites within their boundaries.
-**Panel B:** (n.b.
-will not include map in final manuscript) The number of trap nights conducted is associated with a regions population density.
-The population density for all regions in West Africa is shown in the yellow box plot, the population density and trap night density for each region is show on the purple scatterplot.
-The line of best fit is a GAM model not incorporating spatial interactions.
-The map panel on the right is the product of the GAM model incorporating spatial interactions.
-**Panel C:** For each of the 10 land cover classes from the ESA dataset we measure the proportion of 300m^2^ pixels within a level 2 administrative area.
-Zoonotic trapping studies occurred in regions over-representative for Cropland, Mosaic landscapes and Shrubland while being under-representative for Bare and Sparse vegetation land cover classes.
+![The proportion of the 10 land cover classes reported by the ESA Copernicus dataset at 300m^2 resolution are shown for all regions of West Africa (yellow) and for regions in which trapping occurred (purple).
+Rodent trapping occurred in regions over-representative for cropland, mosaic landscapes and shrubland wile occurring in regions under-representative for bare- and sparse vegetation land cover classes.](../figures/Figure_3.png)
 
@@
 
-```{r, figure_3, layout = "l-screen-inset", echo=FALSE, out.width="100%"}
-
-knitr::include_graphics(here("figures", "Figure_3.png"))
-
-```
-
-*Figure 3*: Each row corresponds to a single rodent species.
+![Each row corresponds to a single rodent species.
 The column on the left shows the presence and absence of a rodent species from the individual studies included in this review.
 The centre column shows the presence of a rodent species obtained from GBIF (September 2021) for records where longitude and latitude have been provided.
-The right column shows the range of rodent species as proposed by the IUCN (2021) (red shaded area), overlaid are the presence points from both this review and GBIF records.
+The right column shows the range of rodent species as proposed by the IUCN (2021) (red shaded area), overlaid are the presence points from both this review and GBIF records.](../figures/Figure_4.png)
 
@@
 
-```{r, figure_4, echo=FALSE, out.width="100%"}
+```{r, figure_5, echo=FALSE, out.width="100%", fig.cap="Presence/absence plots at each unique trapping site for the four most commonly assayed microorganisms Arenaviridae (top), *Bartonella sp.* (second row), *Borrelia sp.* (third row) and *Toxoplasma gondii* (fourth row).
+The tables to the right of each map highlight the 10 (or number applicable) most commonly positive and tested rodent species and genera assayed."}
 
-knitr::include_graphics(here("figures", "Figure_4a.png"))
-knitr::include_graphics(here("figures", "Figure_4b.png"))
-knitr::include_graphics(here("figures", "Figure_4_legend.png"))
+knitr::include_graphics(here("figures", "Figure_5a.png"),
+                        here("figures", "Figure_5b.png"),
+                        here("figures", "Figure_5_legend.png"))
 
 ```
-
-*Figure 4*: Presence/absence plots at each unique trapping site for the four most commonly assayed microorganisms Arenaviridae (top), *Bartonella sp.* (second row), *Borrelia sp.* (third row) and *Toxoplasma gondii* (fourth row).
-The tables to the right of each map highlight the 10 (or number applicable) most commonly positive and tested rodent species and genera assayed.
 
@@
 
-```{r figure_5, echo=FALSE, out.width="100%"}
-
-knitr::include_graphics(here("figures", "Figure_5.png"))
-
-```
-
-*Figure 5*: The percentage of assays from each included species and microorganism tested for is represented by colour.
-The size of a point relates to the squareroot of the number of individuals assayed.
-The plot on the left is for species/genera assayed for four or more microorganisms, the plot on rights are for species/genera assayed for 3 or less microorganisms.
+![For the 27 most commonly assayed rodent species and genera, the percentage of positive assays against each of the 32 microorganisms tested for is shown.
+The size of a point is proportional to the the number of individuals assayed for that microorganism.](../figures/Figure_6.png)
 
 # Discussion
+
 We identified 124 rodent trapping studies conducted in the West African region since 1965.
@@
 Thirty-two potential pathogens were investigated across all studies, 5 rodent species were assayed for at least 18 potential pathogens.
-We identify several important limitations to available data, potentially biasing previous zoonotic risk studies, first, in the timing of studies, and the locations in which they've been conducted, second, the locations of species presence and third, the selection of pathogens to assay.
+We identify several important limitations to available data, potentially biasing previous zoonotic risk studies, first, in the timing of studies and the locations in which they've been conducted, second, the locations where a species was recored as present and third, the limited selection of pathogens assayed at these trapping sites.
 
-First, timing and location of trapping studies can introduce bias to inference from these data when investigating the risk of zoonotic disease emergence.
-Most studies have been conducted within the last twenty years.
-Human population, intensification of agriculture and urbanisation have proceeded rapidly over this time across West Africa.
-These changes modify the available habitats for rodent species increasing the prevalence of commensal species that are more likely to be reservoirs of zoonoses.
-Not accounting for the time period in which data were collected will introduce bias into inferences made for zoonotic emergence based on reservoir species distributions.
-Further, several regions, namely North West Senegal, Eastern Sierra Leone, Southern Ghana, Southern Benin, South West Nigeria  Central Nigeria have been sampled at much higher intensity than elsewhere in West Africa.
-Studies modelling risk of zoonoses without incorporating this important sampling heterogeneity will introduce bias when inferring current or projected risk.
+First, the time period and location of trapping studies, when not systematically conducted, will introduce bias that must be accounted for when attempting to infer the risk of zoonotic disease spillover events or novel pathogen emergence.
+Most studies were conducted in a relatively short period since 2000.
+During this period human populations, agricultural farming and livestock intensification and urbanisation have progressed dramatically across West Africa [@acheampong_deforestation_2019; @connolly_extended_2021].
+These changes alter the availability of habitats for rodent species, potentially increasing the prevalence of commensal species which are more likely to be reservoirs of zoonoses.
+Thus, incorporating the time period in which data were collected is important when modelling risk of zoonotic spillover or novel pathogen emergence based on known reservoir species distributions.
+In addition, several regions, namely North West Senegal, Eastern Sierra Leone, Southern Ghana, Southern Benin, South West and Central Nigeria have been sampled at much higher intensity than elsewhere in West Africa.
+This sampling heterogeneity will again introduce bias when inferring current or projected risk of zoonotic spillover or novel pathogen emergence.
 
-Second, rodent species were found to have wider distributions than the ranges reported by IUCN data.
-Trapping studies were conducted from more regions across West Africa than were available from GBIF records and where therefore able to report presence and absence of species at greater spatial resolution.
-Further, the additional data produced from species' absence and pseudoabsence highlighted the variability of rodent presence within their ranges across the region.
-Failing to incorporate this heterogeneity in rodent reservoir species' distribution can importantly bias estimations of the risk of zoonotic disease emergence.
-For example, the apparent absence of *Mastomys natalensis* at several trapping sites in Nigeria may explain some of the observed spatial variation in Lassa fever spillover that would not be expected based on it's IUCN proposed range or presence only data from GBIF.
+Second, rodent species were found to have greater species ranges than those reported by IUCN data.
+Trapping studies were conducted from more regions across West Africa than were available from GBIF records.
+This increased spatial resolution and the inclusion of absence data alongside presence data presents a much more complete representation of the spatial heterogeneity of rodent species' locations within their ranges.
+Failing to incorporate this heterogeneity in favour of using ranges or presence only data to produce rodent reservoir species' distributions can importantly bias estimations of risk of zoonotic spillover or novel pathogen emergence.
+For example, the apparent absence of *Mastomys natalensis* at several trapping sites in Nigeria may explain some of the observed spatial variation in human Lassa fever cases [@redding_geographical_2021] that would not be expected based on this species IUCN range or presence only data from GBIF.
 
-Third, while up to 32 potential zoonoses were investigated only four were investigated in more than three studies.
-Of these four studies conducting assays for *Lassa mammarenavirus* or other Arenaviridae where the most widely conducted in regards to number of countries and number of study sites within countries.
-For example, *Toxoplasma sp.* were found to be prevalent in rodent hosts within Senegal and despite a significant burden of human disease no studies investigated the prevalence of this pathogen in other countries of West Africa [https://www.nature.com/articles/s41598-020-69078-9].
-The limited geographic scope of sampling for these pathogens hampers efforts to understand the risk of zoonotic spillover across this heterogeneous region.
+Third, while 32 potential zoonoses were investigated in at least one study, only four pathogen species or groups were investigated in more than three.
+Of these four pathogens, studies for *Lassa mammarenavirus* or other Arenaviridae where the most widely performed, both in absolute numbers but also by number of countries and number of study sites within countries.
+In contrast, *Toxoplasma sp.* were found to be prevalent in rodent hosts within Senegal - and despite a significant, known, burden of human disease - no studies investigated the prevalence of this pathogen in other West African countries [@bigna_global_2020].
+The limited geographic scope of sampling for these pathogens hampers efforts to understand the risk and human health burden of zoonotic spillover across this diverse region.
 
 Finally, we found that only five of the 112 rodent species trapped had been tested for more than half of the potential pathogens.
-There is increasing emphasis on the role of co-infection and multi-species pathogen transmission systems on zoonotic disease spillover risk which depends on comprehensive testing of both known and un-known reservoir species.
-More comprehensive testing of potential reservoir species across West Africa is required to better understand the role of individual species in disease systems.
+There is increasing emphasis on the role of co-infection and multi-species pathogen transmission systems on zoonotic disease spillover risk which depends on comprehensive testing of both known and un-known reservoir species [@sweeny_synzootics_2021].
+More comprehensive testing of potential reservoir species across West Africa is required to better understand the role of individual species in disease systems [@douglass_how_2016].
 This will also reduce the risk of bias in predictive risk modelling by better characterising host-pathogen associations in species rich environments.
@@
 
-We have identified important biases in the locations of studies, the trapping effort for different rodent species and the selection of zoonotic pathogens to assay across West Africa.
+Here we have identified important biases in the locations of studies, the trapping effort for different rodent species and the selection of zoonotic pathogens to assay across the West African region.
 The use of these data to infer risk of zoonotic spillover or novel disease emergence is likely inappropriate unless these systematic biases are addressed.
-We hope that the current synthesis of extant data will encourage further comprehensive ecological studies on rodent species assemblages and their pathogens.
-One priority would be to better understand the true distribution of known rodent reservoirs (particularly *Mastomys natalensis*) across the region with particular emphasis made to the habitat in which they reside and the structure of rodent assemblages in the region.
+We hope that this current synthesis of data will encourage further comprehensive ecological studies on rodent species assemblages and their pathogens.
+One priority would be to better understand the true distribution of known rodent reservoirs (particularly *Mastomys natalensis*) across the region with particular emphasis made to the habitat in which they reside and the structure of rodent assemblages.
 
@@
 
+# Acknowledgments
+The European Commission nor ECMWF are not responsible for any use that may be made of the Copernicus information or data it contains.
+
+# Supplementary 1
+
+**Data extraction tool**
+
+# Supplementary 2
+
+**Habitat dictionary**
+
+# Supplementary 3
+
+**Included studies table**
+
 # References

### 2021-10-05 scripts/mapping_pathogens.R

@@ -172,17 +172,37 @@ save_plot(plot_grid(plotlist = list(av_row,
                     ncol = 1,
                     greedy = FALSE,
                     rel_heights = c(1, 1)),
-          filename = here("figures", "Figure_4a.png"), base_height = 10, base_width = 16)
+          filename = here("figures", "Figure_5a.png"), base_height = 10, base_width = 16)
 
 save_plot(plot_grid(plotlist = list(bo_row,
                                     to_row),
                     ncol = 1,
                     greedy = FALSE,
                     rel_heights = c(1, 1)),
-          filename = here("figures", "Figure_4b.png"), base_height = 10, base_width = 16)
+          filename = here("figures", "Figure_5b.png"), base_height = 10, base_width = 16)
 
 save_plot(plot_grid(plot = legend,
                     ncol = 1,
                     greedy = FALSE,
                     rel_heights = c(1)),
-          filename = here("figures", "Figure_4_legend.png"), base_height = 2, base_width = 10)
+          filename = here("figures", "Figure_5_legend.png"), base_height = 2, base_width = 10)
+
+save_plot(plot_grid(plot_grid(plotlist = list(av_row,
+                                    ba_row),
+                    ncol = 1,
+                    greedy = FALSE,
+                    rel_heights = c(1, 1)),
+          plot_grid(plotlist = list(bo_row,
+                                    to_row),
+                    ncol = 1,
+                    greedy = FALSE,
+                    rel_heights = c(1, 1)),
+          plot_grid(plot = legend,
+                    ncol = 1,
+                    greedy = FALSE,
+                    rel_heights = c(1)),
+          rel_heights = c(1, 1, 0.2),
+          rel_widths = c(1, 1, 0.6)),
+          filename = here("figures", "Figure_5_combined.png"),
+          base_height = 20,
+          base_width = 32)

### 2021-10-05 scripts/mapping_traps.R

@@ -12,7 +12,9 @@ level_0 <- read_rds(here("data_download", "admin_spatial", "level_0_admin.rds"))
 
 level_1 <- read_rds(here("data_download", "admin_spatial", "level_1_admin.rds"))
 
-level_2 <- read_rds(here("data_download", "admin_spatial", "level_2_admin.rds"))
+level_2 <- read_rds(here("data_download", "admin_spatial", "level_2_admin.rds")) %>%
+  bind_rows(level_1 %>%
+              filter(GID_0 == "CPV")) #Cape Verde doesn't have level_2 administrative regions
 
 non_trapped <- read_rds(here("data_download", "admin_spatial", "level_2_TGOGMB.rds"))
 
@@ -129,38 +131,17 @@ if(!file.exists(here("data_clean", "traps_level_2.rds"))) {
 }
 
 trap_nights_region <- tibble(sites_2) %>%
-  group_by(NAME_2, GID_2, unique_id, year_trapping, month_trapping, region, town_village, habitat) %>%
+  group_by(GID_1, NAME_2, GID_2, unique_id, year_trapping, month_trapping, region, town_village, habitat) %>%
   summarise(total_trap_nights = sum(trap_nights)) %>%
-  group_by(GID_2, NAME_2) %>%
+  group_by(GID_1, GID_2, NAME_2) %>%
   summarise(region_trap_nights = sum(total_trap_nights))
 
 level_2_sites <- level_2 %>%
   left_join(., trap_nights_region,
-            by = c("GID_2", "NAME_2")) %>%
+            by = c("GID_1", "GID_2", "NAME_2")) %>%
   mutate(area_m2 = st_area(.),
          tn_density = region_trap_nights/(as.numeric(area_m2)/1000000),
          tn_density = ifelse(is.na(tn_density), NA, tn_density))
-
-ggplot() +
-  geom_sf(data = included_countries, fill = "#808080", alpha = 0.1, lwd = 0.1) +
-  geom_point(data = data$review %>%
-               filter(pres_abs == "Present"), mapping = aes(colour = pres_abs, x = x, y = y), size = .5,
-             crs = 4326) +
-  geom_spatial_point(data = anti_join(tibble(data$review), tibble(data$review) %>%
-                                        filter(pres_abs == "Present"),
-                                      by = c("x", "y")),
-                     mapping = aes(colour = pres_abs, x = x, y = y), size = .5,
-                     crs = 4326) +
[…]

### 2021-10-05 reports/manuscript_v2.Rmd
update for appraisal

@@
 
-```{r, figure_5, echo=FALSE, out.width="100%", fig.cap="Presence/absence plots at each unique trapping site for the four most commonly assayed microorganisms Arenaviridae (top), *Bartonella sp.* (second row), *Borrelia sp.* (third row) and *Toxoplasma gondii* (fourth row).
-The tables to the right of each map highlight the 10 (or number applicable) most commonly positive and tested rodent species and genera assayed."}
-
-knitr::include_graphics(here("figures", "Figure_5a.png"),
-                        here("figures", "Figure_5b.png"),
-                        here("figures", "Figure_5_legend.png"))
-
-```
+![Presence/absence plots at each unique trapping site for the four most commonly assayed microorganisms Arenaviridae (top), *Bartonella sp.* (second row), *Borrelia sp.* (third row) and *Toxoplasma gondii* (fourth row).
+The tables to the right of each map highlight the 10 (or number applicable) most commonly positive and tested rodent species and genera assayed.](../figures/Figure_5_combined.png)
 

### 2021-10-05 scripts/analysis.R

@@ -129,7 +129,7 @@ study_timings <- ggplot(study_start) +
        x = element_blank(),
        alpha = "Study dates reported")
 
-ggsave(plot = study_timings, filename = here("figures", "Fig_1_Panel_A.png"), dpi = 300, height = 8)
+ggsave(plot = study_timings, filename = here("figures", "Fig_1_Panel_A.png"), dpi = 300, height = 8, width = 10)
 write_rds(study_timings, here("plots", "study_timings.rds"))
 
 described_studies <- study_start %>%

### 2021-10-05 scripts/mapping_pathogens.R

@@ -79,11 +79,11 @@ pathogen_map <- function(pathogen_genus) {
                                                             pcr_positive != 0 & ab_ag_positive == 0 ~ pcr_positive + culture_positive,
                                                             TRUE ~ ab_ag_positive + culture_positive)) %>%
                          group_by(x, y) %>%
-                         summarise(number_tested = sum(number_tested),
-                                   number_positive = sum(unique_positive),
-                                   number_negative = number_tested-number_positive,
-                                   prop_positive = number_positive/number_tested) %>%
-                         mutate(pos_neg = case_when(number_positive > 0 ~ "Positive",
+                         mutate(number_tested = sum(number_tested),
+                                number_positive = sum(unique_positive),
+                                number_negative = number_tested-number_positive,
+                                prop_positive = number_positive/number_tested,
+                                pos_neg = case_when(number_positive > 0 ~ "Positive",
                                                     TRUE ~ "Negative"))
   )
 
@@ -165,44 +165,17 @@ bartonella_plots <- pathogen_map("bartonella")
 ba_row <- plot_grid(plotlist = bartonella_plots[c(1:2)], nrow = 1)
 toxplasma_plots <- pathogen_map("toxoplasma")
 to_row <- plot_grid(plotlist = toxplasma_plots[c(1:2)], nrow = 1)
-legend <- plot_grid(as_grob(arenaviridae_plots[[3]][[1]]), rel_heights = 0.5)
+legend <- plot_grid(as_grob(arenaviridae_plots[[3]][[1]]))
+all_plots <- plot_grid(av_row,
+                       bo_row,
+                       ba_row,
+                       to_row,
+                       ncol = 2)
 
-save_plot(plot_grid(plotlist = list(av_row,
-                                    ba_row),
-                    ncol = 1,
+save_plot(plot_grid(all_plots, legend,
+                    nrow = 2,
                     greedy = FALSE,
-                    rel_heights = c(1, 1)),
-          filename = here("figures", "Figure_5a.png"), base_height = 10, base_width = 16)
-
-save_plot(plot_grid(plotlist = list(bo_row,
-                                    to_row),
-                    ncol = 1,
-                    greedy = FALSE,
-                    rel_heights = c(1, 1)),
-          filename = here("figures", "Figure_5b.png"), base_height = 10, base_width = 16)
-
-save_plot(plot_grid(plot = legend,
-                    ncol = 1,
-                    greedy = FALSE,
[…]

### 2021-10-05 scripts/mapping_traps.R

@@ -292,29 +292,32 @@ write_rds(tn_to_pop, here("plots", "tn_to_pop.rds"))
 
 tn_pop_map <- read_rds(here("data_clean", "tn_pop_model.rds"))
 
-map_1 <- getViz(tn_pop_model)
-(
-  m1 <- plot(sm(map_1, 1), n = 150, too.far = 0.02) +
-    l_fitRaster(pTrans = zto1(0.05, 2, 0.1)) +
-    geom_sf(data = included_countries, alpha = 0.1, lwd = 0.1, inherit.aes = FALSE) +
-    scale_fill_viridis_c(na.value = "white") +
-    annotation_north_arrow(height = unit(1, "cm"),
-                           style = north_arrow_minimal(text_size = 8)) +
-    annotation_scale(height = unit(0.1, "cm"),
-                     location = "tr") +
-    theme_minimal() +
-    labs(title = element_blank(),
-         x = element_blank(),
-         y = element_blank(),
-         fill = "Linear predictor")
-)
+map_1 <- getViz(tn_pop_map)
+
+m1 <- plot(sm(map_1, 1), n = 150, too.far = 0.02) +
+  l_fitRaster(pTrans = zto1(0.05, 2, 0.1)) +
+  geom_sf(data = included_countries, alpha = 0.1, lwd = 0.1, inherit.aes = FALSE) +
+  scale_fill_viridis_c(na.value = "#ffffff00") +
+  theme_minimal() +
+  labs(title = element_blank(),
+       x = element_blank(),
+       y = element_blank(),
+       fill = "Linear predictor") +
+  annotation_north_arrow(height = unit(1, "cm"),
+                         style = north_arrow_minimal(text_size = 8)) +
+  annotation_scale(height = unit(0.1, "cm"),
+                   location = "tr")
 
 
-panel_b <- plot_grid(as.grob(m1$ggObj))
+B <- as.grob(m1$ggObj)
 
-fig_2 <- plot_grid(trap_night_density_level2,
-                   panel_b,
+A <- trap_night_density_level2
+
+fig_2 <- plot_grid(A,
+                   B,
                    nrow = 2,
+                   align = "v",
+                   axis = "l",
                    labels = "AUTO")
 
-save_plot(here("figures", "Figure_2.png"), fig_2, base_height = 10, base_width = 17)
+save_plot(here("figures", "Figure_2.png"), fig_2, base_height = 10, base_width = 12)

### 2021-10-21 reports/manuscript_v2.Rmd
final draft

@@
 author: 
-  - "David Simons"
+  - "David Simons, Lauren Attfield, Kate Jones, Deborah Watson-Jones and Richard Kock"
     # url: https://www.dsimons.org
@@
   #   toc_float: true
-  # word_document:
-  pdf_document:
+  word_document:
     toc: true
     toc_depth: 3
+  # pdf_document:
+  #   toc: true
+  #   toc_depth: 3
 knit: (function(inputFile, encoding) {
@@
 always_allow_html: yes
-bibliography: "`r here::here('citations', 'manuscript_citations.bib')`"
+bibliography: citations/manuscript_citations.bib
 ---
@@
 These reservoir species are globally distributed with West Africa containing multiple known and several further predicted reservoir species.
-Typically International Union for Conservation of Nature (IUCN) species distribution maps and Global Biodiversity Information Facility (GBIF) presence maps are used to determine regions at risk for zoonotic spillover events, however, these datasets are biased by incomplete sampling.
+Typically, International Union for Conservation of Nature (IUCN) species distribution maps and Global Biodiversity Information Facility (GBIF) presence maps are used to determine regions at risk for zoonotic spillover events, however, these datasets are biased by incomplete sampling.
 To investigate the impact of rodent species sampling heterogeneity on zoonotic spillover risk we have systematically reviewed rodent trapping studies to produce a contextually rich dataset that can be used to explore this risk at finer spatial resolution.
@@
 We found that rodent trapping effort was associated with regional population density and was increased in habitats modified by human activity.
-We produce updated rodent species ranges compared to available IUCN maps and enrich GBIF data by including locations of rodent species absence and trapping effort.
+We compare rodent species ranges produced by the IUCN with rodent presence data from GBIF and with our enriched data including rodent species absence and trapping effort.
 Furthermore, we report the spatial bias in the investigation of four important rodent zoonoses, Arenaviridae, *Borellia sp.*, *Bartonella sp.* and *Toxoplasma sp.*.
@@
 We anticipate this dataset can support the production of more complete spatial risk assessments of zoonotic spillover events.
-For example, the inclusion of absence data and trapping effort can help identify regions where data sparsity may produce inappropriately predicted risk.
-Furthermore, these data identify regions in need of comprehensive study for our four targeted rodent zoonoses, Arenaviridae, *Borellia sp.*, *Bartonella sp.* and *Toxoplasma sp.*.
+For example, the inclusion of absence data and trapping effort can help identify regions where data sparsity may hinder unbiased risk prediction.
+Furthermore, these data identify regions in need of comprehensive study for our four targeted rodent zoonoses.
 
@@
 
-The potential effect of zoonotic infectious diseases outbreaks on human health and societies has been dramatically highlighted through the ongoing SARS-CoV-2 pandemic and recent Ebola virus outbreaks.
-The number of zoonotic spillover events are projected to increase under intensifying anthropogenic pressure such as, increased human populations [@allen_global_2017], increasing urbanisation [@hassell_urbanization_2017], and global climate change [@morse_prediction_2012].
+The potential effect of zoonotic infectious disease outbreaks on human health and societies has been dramatically highlighted through the ongoing SARS-CoV-2 pandemic and recent Ebola virus outbreaks.
+The number of zoonotic spillover events is projected to increase under intensifying anthropogenic pressure such as; increased human populations [@allen_global_2017], increasing urbanisation [@hassell_urbanization_2017], and global climate change [@morse_prediction_2012].
 In addition, two taxa - rodents (Rodentia) and bats (Chiroptera) - are proposed to contribute to the greatest number of zoonotic pathogens and the greatest risk of novel pathogen emergence [@han_rodent_2015].
@@
 Rodent species form diverse assemblages, which provide important and beneficial ecosystem services including pest regulation and seed dispersal [@fischer_ecosystem_2018].
-The role of rodent species' in zoonotic infectious disease spillover or novel pathogen emergence can be viewed as an ecosystem disservice.
-Rodents typically demonstrate “fast” life histories [@dobson_fast_2007], these traits are also associated with a species being a reservoir of zoonotic pathogens [@han_rodent_2015; @albery_fast-lived_2021].
+The role of rodent species in zoonotic infectious disease spillover or novel pathogen emergence can be viewed as an ecosystem disservice.
+Rodents typically demonstrate “fast” life histories [@dobson_fast_2007] due to traits such as early maturation; traits which are also associated with zoonotic reservoirs [@han_rodent_2015; @albery_fast-lived_2021].
 Further, these traits are prevalent in species that thrive in human dominated landscapes where they displace species that are less likely to be reservoirs of zoonotic pathogen [@gibb_zoonotic_2020].
@@
 
-Rodent trapping studies provide a useful method to describe rodent population structures and survey for potential zoonoses.
-Studies have been conducted in West Africa to both identify novel potential zoonotic pathogens within rodents [@usaid_one_2021] and to investigate the prevalence and burden of known pathogens within known rodent hosts (e.g.
-for Lassa fever [@fichet-calvet_diversity_2009] and Schistosomiasis [@catalano_multihost_2020]).
+Rodent trapping studies provide a useful method to describe rodent population assemblages and survey for potential zoonoses.
+Studies have been conducted in West Africa to both identify novel potential zoonotic pathogens within rodents [@usaid_one_2021] and investigate the prevalence of known pathogens within known rodent hosts  to understand the distribution and burden of human disease (e.g., for Lassa fever [@fichet-calvet_diversity_2009] and Schistosomiasis [@catalano_multihost_2020]).
 These studies provide contextually rich information on when, where and under what conditions rodents were trapped that are typically missing from the global datasets [@bovendorp_optimising_2017].
-Despite this, these global datasets (e.g.
-IUCN, GBIF and GIDEON) are typically used to inform species distribution maps and zoonotic disease spillover predictions [@han_global_2016; @smith_global_2014; @pigott_mapping_2014].
+Despite this, these global datasets are typically used to inform species distribution maps (e.g., IUCN, GBIF) and zoonotic spillover and novel pathogen emergence predictions (e.g., EID2 and GIDEON) [@han_global_2016; @smith_global_2014; @pigott_mapping_2014].
 
-These studies have been used to identify potential geographic hotspots where virus and host species diversity may be expected to be at its greatest to predict regional zoonotic disease spillover risks.
-However, there remains the potential for important confounding in these spatial distributions through bias generated by study design and selection of sampling sites [@plowright_sampling_2019].
+Multiple studies have identified potential geographic hotspots where virus and host species diversity may be expected to be at its greatest, this has been used to predict regional zoonosis spillover and novel pathogen emergence risks [@grace_mapping_2012; @han_global_2016; @allen_global_2017].
+However, there remains the potential for confounding factors having important negative effect on these spatial distributions through bias generated by study design and selection of sampling sites [@plowright_sampling_2019].
 For example, systematically increased sampling (e.g.
-more intensive studies over longer time periods) or over-representation of certain habitats (e.g.
-peri-urban landscapes) could lead to an apparent association between locations and risk that is driven by these factors rather than an underlying host and virus association [@wille_how_2021; @gibb_mammal_2021].
-Conversely some regions may not be sampled adequately and therefore under-represented in these datasets due to sparse human populations or inaccessible habitats this may result in these regions being reported inappropriately as at low risk of novel pathogen emergence or zoonotic disease spillover events.
+more intensive studies over longer time periods) or over-representation of certain habitats (e.g., peri-urban landscapes) could lead to an apparent association between locations and risk that is driven by these factors rather than an underlying host and virus association [@wille_how_2021; @gibb_mammal_2021].
+Conversely some regions may not be sampled adequately and therefore under-represented in these datasets due to sparse human populations or inaccessible habitats, which may result in these regions being inappropriately reported as at low risk of novel pathogen emergence or zoonotic spillover events.
 
-Here, we identify rodent trapping studies performed across West Africa and identify the location and habitat types in which they have been conducted, the pathogens assessed and the host-pathogen associations that have been reported in order to quantify the potential bias and to identify regions requiring further focussed investigation.
+Here, we synthesis rodent trapping studies performed across West Africa and identify the location and habitat types in which they have been conducted, the microorganisms assessed and the host-pathogen associations that have been reported in order to quantify the potential bias and to identify regions requiring further focussed investigation.
 
@@
 
-We included studies if they met all of the following inclusion criteria; i) reported findings from trapping studies where the target was a small mammal, ii) described the type of trap used or the length of trapping activity or the location of the trapping activity, iii) included trapping activity from at least one West African country, iv) recorded the genus or species of trapped individuals, v) were published in a peer-reviewed journal or as a pre-print on a digital platform or as a report by a credible organisation.
+We included studies if they met all of the following inclusion criteria; i) reported findings from trapping studies where the target was a small mammal, ii) described the type of trap used or the length of trapping activity or the location of the trapping activity, iii) included trapping activity from at least one West African country (Benin, Burkina Faso, Cape Verde, Côte d'Ivoire, Eastern Sahara, Gambia, Ghana, Guinea, Guinea-Bissau, Liberia, Mali, Mauritania, Niger, Nigeria, Saint Helena, Ascension and Tristan da Cunha, Senegal, Sierra Leone and Togo), iv) recorded the genus or species of trapped individuals, v) were published in a peer-reviewed journal or as a pre-print on a digital platform or as a report by a credible organisation.
 We excluded studies if they met any of the following exclusion criteria: i) reported data that were duplicated from a previously included study, ii) no full text available, iii) not available in English.
@@
 
-We extracted data from eligible studies into a Google sheets document.
+We extracted data from eligible studies.
 Extracted variables included i) study identifiers; ii) study aims; iii) trapping methodology; iv) geolocation data; v) method of speciation; vi) trapping locations and dates; vii) trapped species; viii) number of trap-nights and ix) microorganisms/pathogens of interest.
-The data extraction tool is archived and available in Supplementary 1 or at this [link](https://docs.google.com/spreadsheets/d/1rQYjHhk6uk1PoKZZVsgFlmuqWGicU2tTisk9ddfAwTM/edit?usp=sharing).
+The data extraction tool and the resulting dataset is archived and available in [Supplementary 1](https://docs.google.com/spreadsheets/d/1rQYjHhk6uk1PoKZZVsgFlmuqWGicU2tTisk9ddfAwTM/edit?usp=sharing).
 
@@
 
-We extracted GPS locations for the most precise location presented (i.e.
-trap, trap-line, study site or study region).
-We extracted coordinates in the format reported and converted them to decimal degrees.
-We recorded the habitat classification scheme used by a study (e.g.
-IUCN Habitat Classification Scheme (Version 3.1)).
-For studies not using standardised habitat recording, the explicit description of the habitat in which the trap was placed were extracted.
-For studies reporting multiple habitat types (e.g.
-rice field, corn field and vegetable garden) for a single trap, trap-line or trapping grid, a higher order classification of habitat type was recorded (e.g.
-agricultural land).
+We extracted GPS locations for the most precise location presented (i.e., trap, trap-line, study site or study region) by extracting coordinates in the format reported and converting them to decimal degrees.
+We recorded the habitat classification scheme used by a study (e.g., IUCN Habitat Classification Scheme (Version 3.1)).
+For studies not using standardised habitat recording, the explicit description of the habitat in which the trap was placed was extracted.
+For studies reporting multiple habitat types (e.g., rice field, corn field and vegetable garden) for a single trap, trap-line or trapping grid, a higher order classification of habitat type was recorded (e.g., agricultural land).
 
@@
 We extracted information on the presence, absence and number of trapped individuals.
-For studies reporting on all trapped individuals (i.e.
-not those only reporting on the presence of a specific species of interest), the pseudo-absence of a species reported as present elsewhere in the study was explicitly recorded as an absence at that trap location.
+For studies reporting on all trapped individuals (i.e., not those only reporting on the presence of a specific species of interest), the pseudo-absence of a species reported as present elsewhere in the study was explicitly recorded as an absence at that trap location.
 
@@
 We extracted data on all pathogens assayed in studies investigating rodents for potential zoonoses.
-The number of rodents tested and the number of positive or negative samples were recorded alongside the type of assay used (e.g.
-Polymerase Chain Reaction (PCR), Enzyme Linked ImmunoSorbent Assay (ELISA) or viral culture).
+The number of rodents tested and the number of positive or negative samples were recorded alongside the type of assay used (e.g., Polymerase Chain Reaction (PCR), Enzyme Linked ImmunoSorbent Assay (ELISA) or viral culture).
 If studies reported indeterminate results this was noted.
 Where possible, pathogens were identified to species level.
-However, where an assay only allowed for attribution to a family of viruses or bacteria, the higher order grouping was used (e.g.
-Arenaviridae for a PCR using a non-specific arenavirus primer).
+However, where an assay only allowed for attribution to a family of viruses or bacteria, the higher order grouping was used (e.g., Arenaviridae for a PCR using a non-specific arenavirus primer).
 
@@
 The number of rodents trapped at a trap site was then used in combination with these values of trap success to impute the number of trap nights for trap sites where there was no record of the number of trap nights.
-We summarised the habitat types of trap sites based on information reported in the study, the dictionary used to match habitat types is available in Supplementary 2.
+We summarised the habitat types of trap sites based on information reported in the study; the dictionary used to match habitat types is available in Supplementary 2.
 
-For studies investigating rodent zoonoses we compared the location of trapping sites with SEDAC Global Population Density estimates for 2005 - the median year studies were started [@socioeconomic_data_and_applications_center_gridded_2021].
-We used a Generalised Additive Model (GAM) incorporating a spatial interaction term to investigate the association of number of trap nights and human population density [@pedersen_hierarchical_2019].
+For studies investigating rodent zoonoses we compared the location of trapping sites with SEDAC Global Population Density estimates for 2005, the median year studies were started [@socioeconomic_data_and_applications_center_gridded_2021].
+We used a Generalised Additive Model (GAM) incorporating a spatial interaction term to investigate the association between number of trap nights and human population density [@pedersen_hierarchical_2019].
 The model structure was specified as:
@@
 $$
-\text{Trap night density} \sim Tweedie(log(Population density_{2005}) + (Longitude * Latitude))   
+\begin{equation}
+\tag{1}
+\text{Trap night density} \sim \text{Tweedie(log(Population density_{2005}) + (Longitude * Latitude))}
+\end{equation}
 $$
 
-We obtained land cover classifications from the European Space Agency Copernicus dataset at 300m^2^ spatial resolution generated using Copernicus Climate Change Service information (2005) we extracted the proportion of land cover classes within all regions of West Africa and all regions in which rodent trapping occurred for investigation of zoonotic diseases.
+The linear predictor output of this model is the trapping density adjusted for human population density in 2005. We performed sensitivity analysis removing sites with imputed trapping effort.
+
+We obtained land cover classifications from the European Space Agency Copernicus dataset at 300m^2^ spatial resolution generated using Copernicus Climate Change Service information (2005), then extracted the proportion of land cover classes within all regions of West Africa and all regions in which rodent trapping occurred for investigation of zoonotic diseases.
 We compared the proportion of possible land cover classes with those of regions where trapping occurred.
 
-Finally, we mapped the presence and absence of rodent species and compared this to the presence and absence produced from GBIF [@gbif_the_global_biodiversity_information_facility_occurrence_2021] and IUCN [@iucn_iucn_2021] datasets to give a measure of the extent of each species range in which they have been sampled.
+Finally, we mapped the presence and absence of rodent species and compared this graphically by overlaying trapping locations to the presence produced from GBIF [@gbif_the_global_biodiversity_information_facility_occurrence_2021] and IUCN [@iucn_iucn_2021] datasets to give a measure of the extent to which each species' range has been sampled.
 
@@
 
-We summarised the presence and absence of microorganisms, the assays used, their host species and the locations from where the samples were obtained.
+We summarised the assays used for microorganisms, the presence and absence of these microorganisms, their host species and the locations from where the samples were obtained.
+We describe the different assays used for each microorganism across included studies.
+The presence of a microorganism was determined by a report of a positive assay (e.g., PCR, serology, viral culture) in a rodent sample at a defined location, absence was determined by negative results in all rodent samples at a defined location.
+We describe the number of microorganisms tested for in each rodent species and explore the geographic distributions these samples were obtained from.
 We investigated the association between rodent species and the detection of potential pathogens and report the proportion of positive and negative tests for each species and pathogen pair.
@@
 We identified 4,282 relevant citations, with 126 rodent trapping studies included.
-The list of included studies in available in Supplementary 3 or through the interactive web based application produced to explore this dataset, available [here](https://diddrog11.shinyapps.io/scoping_review_app/).
+The list of included studies in available in Supplementary 3 or through the interactive web-based application produced to explore this dataset, available [here](https://diddrog11.shinyapps.io/scoping_review_app/).
 The earliest included studies were published in 1974 with increasing numbers of studies being performed annually since 2005.
-The median length of rodent trapping activity was one year (IQR 0-2 years).
+The mean length of rodent trapping activity was 1.8 years (range 0-23 years).
 The median time from completion of rodent trapping to publication is 3 years (IQR 2-5 years) (Figure 1.).
 
-![Each row represents one of the 126 included studies, green points designate the first year of data collection, blue points designate the end of data collection.
-For studies completed within one year the blue point completely overlies the green.
+![**Year trapping activity begun, ended and the year of publication for included studies.** Each row represents one of the 126 included studies, green points indicate when the trapping was initiated, blue points indicated the year trapping was concluded.
+For studies completed within one year the blue point overlies the green.
 Studies with a transparent grey point did not report the year in which trapping was conducted.
-The year of publication is shown by a red point.](../figures/Fig_1_Panel_A.png)
+The year of publication is indicated by a red point.](../figures/Fig_1_Panel_A.png)
 
@@
 Rodent trapping took place at 1,331 trap sites with at least one trap site recorded from 14 West African countries.
-No rodent trapping studies were identified from Gambia or Togo.
-The most frequently studied countries were Senegal, Ghana, Guinea and Nigeria (more than 10 studies each).
-Fewer than 5 studies were conducted in Liberia, Burkina Faso and Guinea-Bissau.
-Thirty-one (25%) studies reported trapping at a single study site, 46 (37%) studies trapped at between two and five study sites, the remaining 49 studies trapped at between six and 93 study sites.
+No rodent trapping studies were identified from Gambia, Saint Helena, Ascension and Tristan da Cunha or Togo.
+The most frequently studied countries were Senegal, Ghana, Guinea, and Nigeria (more than 10 studies each).
+Fewer than 5 studies were conducted in Liberia, Burkina Faso, and Guinea-Bissau.
+Thirty-one (25%) studies reported trapping at a single study site, 46 (37%) studies trapped at between two and five study sites, abd the remaining 49 studies trapped at between six and 93 study sites.
 Trap sites were situated in 273 (19.3%) of 1414 level 2 administrative regions in the 14 West African nations.
-The areas with highest trap night density included the capital cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako) and Ghana (Accra) and the largest cities of Côte d'Ivoire (Abidjan) and Benin (Cotonou).
+The areas with highest trap night density included the capital cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako), and Ghana (Accra) and the largest cities of Côte d'Ivoire (Abidjan), and Benin (Cotonou).
 Outside of these cities, Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone had the highest density of trap nights (Figure 2.
@@
 In particular Mauritania, Northern Senegal, Eastern Sierra Leone, Eastern Guinea and South West Nigeria were trapped at higher rates than would be expected based on their population densities.
-South West Nigeria, Northern Nigeria, Liberia, Côte d'Ivoire, Ghana, Niger and Burkina Faso were trapped at lower rates than would be expected.
+South West Nigeria, Northern Nigeria, Liberia, Côte d'Ivoire, Ghana, Niger, and Burkina Faso were trapped at lower rates than would be expected.
+Excluding sites with imputed trap nights had no meaningful effect on the association of human population density and trapping effort in sensitivity analysis.
 
-![**A:** Map of West Africa, countries where rodent trapping has occurred are mapped to level 2 administrative areas.
-The colour of the region corresponds to the trap night density per 1000 km^2.
-Colourless regions reported no trapping.
-**B:** This shows the linear predictor from obtained from the GAM.
-Predictions are limited to coordinates are trap sites (coloured areas), uncertainty in the linear predictor is represented by th transparency of the colour.
-Yellow regions represent areas with higher than expected trapping effort for the regional population and purple regions represent areas trapped at lower than expected rates based on their populations](../figures/Figure_2.png)
+![**Regional trap night density.****A** West Africa mapped to level 2 administrative areas.
+The colour of the region corresponds to trap night density per 1000 km^2.
+Purple regions correspond to low trap night density.
+Colourless regions are level 2 administrative areas with no reported trapping.
+**B** Map of West Africa coloured by trapping density adjusted for human population density (2005) obtained from our GAM (equation 1).
+Yellow regions represent areas with higher-than-expected trapping effort for the regional population density and purple regions represent areas trapped at lower-than-expected rates based on their population density.
+Increased transparency of shading indicates uncertainty in the linear predictor due to limited precision of reported trap site location.](../figures/Figure_2.png)
 
@@
 Single trap sites could span multiple habitat types resulting in 22,202 distinct habitat and trap sites.
-The most commonly trapped sites were in or around buildings (29%), in areas described as the rodents “natural habitat” (24%), in agricultural areas (13%) (e.g.
-rice fields, palm plantations), forests (8%) and in the area surrounding buildings (4%).
+The most commonly trapped sites were in or around buildings (29%), in areas described as the rodents “natural habitat” (24%), in agricultural areas (13%) (e.g., rice fields, palm plantations), forests (8%) and in the area surrounding buildings (4%) (Figure 3.).
 
-![The proportion of the 10 land cover classes reported by the ESA Copernicus dataset at 300m^2 resolution are shown for all regions of West Africa (yellow) and for regions in which trapping occurred (purple).
+![**Regional land cover of West Africa comparing all regions and trapped regions.** The proportion of the 10 land cover classes reported by the ESA Copernicus dataset at 300m^2^ resolution are shown for all regions of West Africa (yellow) and for regions in which trapping occurred (purple).
 Rodent trapping occurred in regions over-representative for cropland, mosaic landscapes and shrubland wile occurring in regions under-representative for bare- and sparse vegetation land cover classes.](../figures/Figure_3.png)
@@
 
-73,164 small mammals were trapped (592 were trapped outside of West African countries), 2,830 (4%) trapped individuals were identified to order level (Rodentia), 7,760 (11%) were identified to genus level, the remaining 62,574 (85%) were identified to species level.
-In studies reporting the number of trap nights the median trap success rate of traps placed in or around buildings was 13% (IQR 6-24%), this compares to a median trap success rate of 3% (IQR 1-9%) in other habitats.
-The majority of the 147 species trapped were Rodentia (112) (Supplementary Material 3), of these Muridae (82) were the largest family of rodents, followed by, Sciuridae (10), Gliridae and Nesomyidae (both 6), Ctenodactylidae, Anomaluridae, Dipodidae, Hystricidae and Thryonomidae (2 or fewer).
-The remaining 34 species were from the orders of Soricomorpha (30), Erinaceomorpha (2) and Afrosoricida (1).
+74,017 small mammals were trapped (excluding 592 were trapped outside of West African countries).
+2,830 (4%) trapped individuals were identified to order level (Rodentia), 7,283 (10%) were identified to genus level, and the remaining 63,904 (86%) were identified to species level.
+In studies reporting the number of trap nights the median trap success rate of traps placed in or around buildings was 13% (IQR 6-24%), compared with a median trap success rate of 3% (IQR 1-9%) in other habitats.
+The majority of the 145 species trapped were Rodentia (112) (Supplementary Material 3), and of these Muridae (82) were the largest family of rodents, followed by; Sciuridae (10), Gliridae and Nesomyidae (both 6), Ctenodactylidae, Anomaluridae, Dipodidae, Hystricidae and Thryonomidae (2 or fewer).
+The remaining 33 species were from the orders of Soricomorpha (30), Erinaceomorpha (2) and Afrosoricida (1).
 
-*Mastomys sp.* were the most commonly trapped genus (27,079, 37.7%), followed by, *Rattus sp.* (11,472, 16.1%), *Mus sp.* (8,624, 12.0%), *Arvicanthis sp.* (5,821, 8.1%) and *Praomys sp.* (5,409, 7.5%).
+*Mastomys sp.* were the most commonly trapped genus (27,079, 37.7%), followed by; *Rattus sp.* (11,472, 16.1%), *Mus sp.* (8,624, 12.0%), *Arvicanthis sp.* (5,821, 8.1%) and *Praomys sp.* (5,409, 7.5%).
 *Mastomys natalensis* were the most commonly trapped species (11,116, 17.4%), followed by, *Rattus rattus* (9,959, 15.6%), *Mastomys erythroleucus* (7,386, 11.6%), *Mus musculus* (6,245, 9.8%), *Arvicanthis niloticus* (5,497, 8.6%), *Mastomys huberti* (4,699, 7.4%) and *Praomys daltoni* (1,854, 2.9%).
 
-Locations of points for the presence or absence of the seven most commonly trapped species are shown in Figure 3 (left column).
-Absence include true absences and pseudo-absences dependent on the trapping effort at the site.
-This are shown compared to species presence obtained from GBIF (centre column) and IUCN species range data (left column) (Figure 3).
+We identified locations of the presence or absence points of the seven most commonly trapped species, including true absences and pseudo-absences dependent on the trapping effort at the site.
+When comparing these with species presence obtained from GBIF and IUCN species range data we found that our data were obtained from more geographic locations within a species predicted range than within the GBIF data which were overrepresented from some West African countries, specifically Benin.
+We observed that both our included studies and GBIF data suggested that species have greater ranges than were projected from IUCN range maps.
+This was particularly evident for the invasive species, *Mus musculus* and *Rattus rattus*, which were projected to have more limited distributions within West Africa.
+Further, we identified species absences were interspersed within presence locations obtained from both trapping studies and GBIF presence locations.
+Together this suggests greater heterogeneity in species distributions than would be apparent from either GBIF or IUCN data (Figure 4.).
 
-![Each row corresponds to a single rodent species.
-The column on the left shows the presence and absence of a rodent species from the individual studies included in this review.
-The centre column shows the presence of a rodent species obtained from GBIF (September 2021) for records where longitude and latitude have been provided.
-The right column shows the range of rodent species as proposed by the IUCN (2021) (red shaded area), overlaid are the presence points from both this review and GBIF records.](../figures/Figure_4.png)
+![**Rodent presence, absence and species ranges from this study, GBIF and IUCN.** Each row corresponds to a single rodent species.
+The left column shows the presence (purple points) and absence (yellow points) of a rodent species from the individual studies included in this review.
+The centre column shows the presence (green points) of a rodent species obtained from GBIF (September 2021) for records where longitude and latitude have been provided.
+The right column shows the range of rodent species as proposed by the IUCN (2021) (red shaded area), overlaid with the presence points from both this review and GBIF records.](../figures/Figure_4.png)
 
@@
 Seven studies solely investigated pathogens of rodents, including Hydatigera species (previously Taenia species) and Trichuris species.
-Zoonotic pathogens and potentially zoonotic micro-organisms were investigated in 55 studies.
-Thirty-two microorganisms were tested for, 8 of these were specified to species level, the remaining 24 were specified at higher taxonomic classifications.
+Zoonotic pathogens and potentially zoonotic microorganisms were investigated in 55 studies.
+Thirty-two microorganisms were tested for, 8 of which were specified to species level, with the remaining 24 specified at higher taxonomic levels.
 
 Polymerase Chain Reaction (PCR) was used in 32 studies to detect the presence of 22 different species or families of microorganisms from 21,953 rodent samples.
-The most common microorganisms assayed for using PCR were *Lassa mammarenavirus* or other Arenaviridae (31%) the bacteria *Borrelia sp.* (11.3%) and *Bartonella sp.* (6.5%), followed by Usutu virus (6.4%) and Hantaviridae (5.6%).
+The most common microorganisms assayed for using PCR were *Lassa mammarenavirus* or other Arenaviridae (31%), the bacteria *Borrelia sp.* (11.3%) and *Bartonella sp.* (6.5%), followed by Usutu virus (6.4%) and Hantaviridae (5.6%).
 Molecular tests (antibody or antigen based) were used in 11 studies to detect the presence of 9 different species or families of microorganisms from 11,430 rodent samples.
@@
 
-Most studies tested for a single microorganisms (39), 16 studies tested for two or more microorganisms.
+Most studies tested for a single microorganism (39), 16 studies tested for two or more microorganisms.
 The most frequently investigated microorganisms were *Lassa mammarenavirus* (17 studies) or members of the Arenaviridae family (10), the spirochete bacteria *Borrelia sp.* (8), *Bartonella sp.* and *Toxoplasma gondii* (4), the remaining 25 microorganisms were reported in three or fewer studies.
@@
 No studies investigated the prevalence of these pathogens in hosts from Burkina-Faso, Guinea-Bissau, Liberia, Niger, Senegal and Togo.
-Arenaviridae were detected in rodents from 17 of 54 species or genera assayed in all the countries it was investigated except Benin (Figure 4 top row).
-The presence and absence of *Bartonella sp.* was investigated in Mali and Nigeria and Senegal with positive samples detected in 11 of 11 species or genera of rodents obtained at most sites from these countries (Figure 4 second row).
+Arenaviridae were detected in rodents all the countries in which it was investigated except Benin (Figure 5A.).
 The presence and absence of *Borrelia sp.* was investigated in Benin, Mali, Mauritania, Niger and Senegal.
-*Borrelia sp.* were detected in 8 of 44 species or genera of rodents assayed in Mali with negative samples obtained from Benin, Mauritania, Niger and Senegal (Figure 4 third row).
-The presence of *Toxoplasma gondii* was only investigated in Senegal and was detected in 6 of 8 rodent species assayed (Figure 4 fourth row).
+*Borrelia sp.* were detected in 8 of 44 species or genera of rodents and shrews assayed in Mali with negative samples obtained from Benin, Mauritania, Niger and Senegal (Figure 5B.).
+The presence and absence of *Bartonella sp.* was investigated in Mali and Nigeria and Senegal with positive samples detected at all but two study sites in Senegal (Figure 5C.).
+The presence of *Toxoplasma gondii* was only investigated in Senegal and was detected in 6 of 8 rodent and shrew species assayed (Figure 5D.).
 
-![Presence/absence plots at each unique trapping site for the four most commonly assayed microorganisms Arenaviridae (top), *Bartonella sp.* (second row), *Borrelia sp.* (third row) and *Toxoplasma gondii* (fourth row).
-The tables to the right of each map highlight the 10 (or number applicable) most commonly positive and tested rodent species and genera assayed.](../figures/Figure_5_combined.png)
+![**Presence and absence of the four most commonly assayed microorganisms.** Arenaviridae (**A**), *Borrelia sp.* (**B**)  *Bartonella sp.* (**C**) and *Toxoplasma gondii* (**D**).
+The tables to the right of each map highlight the most commonly positive and most commonly assayed rodent species and genera.](../figures/Figure_5_combined.png)
 
@@
 
-Ninety-seven rodent species were investigated for the presence of a zoonotic pathogen, 42,940 assays were performed on 32,014 individual rodent samples.
-The rodent species most commonly assayed for zoonotic pathogens were, *Rattus rattus* (n = 2,977) assessed for 24 pathogens, *Mus musculus* (n = 3,402) for 23, *Mastomys natalensis* (n = 7,189) and *Mastomys erythroleucus* (n = 3,013) both 19 and *Arvicanthis niloticus* (n = 3,840) for 18.
-All remaining species were investigated for 10 or fewer pathogens.
+Ninety-seven rodent species were investigated for the presence of a microorganism of interest, and 42,940 assays were performed on 32,014 individual rodent samples.
+The rodent species most commonly assayed for these microorganisms were; *Rattus rattus* (n = 2,977) assessed for 24 microorganisms, *Mus musculus* (n = 3,402) for 23, *Mastomys natalensis* (n = 7,189) for 19, and *Mastomys erythroleucus* (n = 3,013) for 19 and *Arvicanthis niloticus* (n = 3,840) for 18.
+All remaining species were investigated for 10 or fewer microorganisms.
 
 *Lassa mammarenavirus* was detected in 14 species of rodents and Crocidura genera.
-*Mastomys natalensis* was the most common species to test positive (Figure 4.).
+*Mastomys natalensis* was the most common species to test positive (Figure 6.).
 Fifty-four species and genera were tested for arenaviruses, with 37 testing negative.
@@
 *Mus musculus* was the most common species to test positive.
-Finally, 38 species of Rodentia, 3 species of Soricomorpha and 2 species of Erinaceomorpha that were tested for pathogens had entirely negative results.
+Finally, 38 species of Rodentia, 3 species of Soricomorpha and 2 species of Erinaceomorpha that were tested for microorganisms had entirely negative results.
 
-![For the 27 most commonly assayed rodent species and genera, the percentage of positive assays against each of the 32 microorganisms tested for is shown.
-The size of a point is proportional to the the number of individuals assayed for that microorganism.](../figures/Figure_6.png)
+![**Number of individuals tested for each microorganism and proportion positive.** For the 27 most commonly assayed rodent species and genera, the percentage of positive assays against each of the 32 microorganisms tested for is shown.
+The size of a point is proportional to the number of individuals assayed for that microorganism.](../figures/Figure_6.png)
 
@@
 These studies were conducted in 14 countries with 54 studies providing information about zoonotic pathogens in trapped rodents.
-Studies were conducted in all West African countries except Gambia and Togo with between- and within-country variability in the number of trap nights conducted.
-Over 70,000 rodents identified to genus or species level were trapped from 112 species of Rodentia.
-Thirty-two potential pathogens were investigated across all studies, 5 rodent species were assayed for at least 18 potential pathogens.
-We identify several important limitations to available data, potentially biasing previous zoonotic risk studies, first, in the timing of studies and the locations in which they've been conducted, second, the locations where a species was recored as present and third, the limited selection of pathogens assayed at these trapping sites.
+Studies were conducted in all West African countries except Gambia, Saint Helena, Ascension and Tristan da Cunha and Togo with between- and within-country variability observed in the number of trap nights conducted.
+Greater than 70,000 rodents identified to genus or species level were trapped from 112 species of Rodentia.
+Thirty-two microorganisms were investigated across all studies, and five rodent species were assayed for at least 18 microorganisms.
+We identify several important limitations to available data, potentially biasing previous zoonotic risk studies; first, in the timing of studies and the locations in which they've been conducted, second, in the locations where a species was recorded as present, and third, in the limited selection of pathogens assayed at these trapping sites.
 
-First, the time period and location of trapping studies, when not systematically conducted, will introduce bias that must be accounted for when attempting to infer the risk of zoonotic disease spillover events or novel pathogen emergence.
-Most studies were conducted in a relatively short period since 2000.
-During this period human populations, agricultural farming and livestock intensification and urbanisation have progressed dramatically across West Africa [@acheampong_deforestation_2019; @connolly_extended_2021].
+First, bias is introduced through the absence of a systematic approach to rodent trapping across time and location.
+The variability in sampling effort made by year and location must be accounted for when attempting to infer the risk of zoonotic spillover events or novel pathogen emergence [@lloyd-smith_nine_2015].
+The majority of studies have been conducted since 2000, during this period human population numbers, urbanisation, agricultural- and livestock intensification have progressed dramatically across West Africa [@acheampong_deforestation_2019; @connolly_extended_2021].
 These changes alter the availability of habitats for rodent species, potentially increasing the prevalence of commensal species which are more likely to be reservoirs of zoonoses.
-Thus, incorporating the time period in which data were collected is important when modelling risk of zoonotic spillover or novel pathogen emergence based on known reservoir species distributions.
+Thus, accounting for the year in which data were collected and the length of time of the study is important when modelling risk of zoonotic spillover or novel pathogen emergence based on known reservoir species distributions.
 In addition, several regions, namely North West Senegal, Eastern Sierra Leone, Southern Ghana, Southern Benin, South West and Central Nigeria have been sampled at much higher intensity than elsewhere in West Africa.
-This sampling heterogeneity will again introduce bias when inferring current or projected risk of zoonotic spillover or novel pathogen emergence.
+This sampling heterogeneity will again introduce bias when inferring current or projected risk of zoonotic spillover or novel pathogen emergence [@plowright_sampling_2019].
 
@@
 Trapping studies were conducted from more regions across West Africa than were available from GBIF records.
-This increased spatial resolution and the inclusion of absence data alongside presence data presents a much more complete representation of the spatial heterogeneity of rodent species' locations within their ranges.
-Failing to incorporate this heterogeneity in favour of using ranges or presence only data to produce rodent reservoir species' distributions can importantly bias estimations of risk of zoonotic spillover or novel pathogen emergence.
-For example, the apparent absence of *Mastomys natalensis* at several trapping sites in Nigeria may explain some of the observed spatial variation in human Lassa fever cases [@redding_geographical_2021] that would not be expected based on this species IUCN range or presence only data from GBIF.
+This increased spatial resolution and the inclusion of presence-absence data presents a more complete representation of the spatial heterogeneity of rodent species' locations within their ranges.
+Failing to incorporate this heterogeneity in favour of using ranges or presence only data to produce rodent reservoir species' distributions can bias estimations of risk of zoonotic spillover or novel pathogen emergence through neglecting these important confounders.
+For example, the apparent absence of *Mastomys natalensis* at several trapping sites in Nigeria may explain some of the observed spatial variation in human Lassa fever cases that would not be expected based on this species IUCN range or presence only data from GBIF [@redding_geographical_2021].
 
 Third, while 32 potential zoonoses were investigated in at least one study, only four pathogen species or groups were investigated in more than three.
-Of these four pathogens, studies for *Lassa mammarenavirus* or other Arenaviridae where the most widely performed, both in absolute numbers but also by number of countries and number of study sites within countries.
-In contrast, *Toxoplasma sp.* were found to be prevalent in rodent hosts within Senegal - and despite a significant, known, burden of human disease - no studies investigated the prevalence of this pathogen in other West African countries [@bigna_global_2020].
+Of these four pathogens, studies for *Lassa mammarenavirus* or other Arenaviridae were the most widely performed, both in absolute numbers but also by number of countries and number of study sites within countries.
+In contrast, *Toxoplasma sp.* were found to be prevalent in rodent hosts within Senegal and - despite a substantial, known, burden of human disease [@bigna_global_2020]- no studies investigated the prevalence of this pathogen in other West African countries.
 The limited geographic scope of sampling for these pathogens hampers efforts to understand the risk and human health burden of zoonotic spillover across this diverse region.
@@
 Finally, we found that only five of the 112 rodent species trapped had been tested for more than half of the potential pathogens.
-There is increasing emphasis on the role of co-infection and multi-species pathogen transmission systems on zoonotic disease spillover risk which depends on comprehensive testing of both known and un-known reservoir species [@sweeny_synzootics_2021].
+There is increasing emphasis on the role of co-infection and multi-species pathogen transmission systems on zoonotic disease spillover risk which depends on comprehensive testing of both known and unknown reservoir species [@sweeny_synzootics_2021].
 More comprehensive testing of potential reservoir species across West Africa is required to better understand the role of individual species in disease systems [@douglass_how_2016].
@@
 We identified several limitations to our approach.
-We identified studies from a diverse range of sources, however, it remains possible that rodent surveys have been performed for additional purposes that would not be captured in a survey of the scientific and "grey" literature, for example, risk assessments or ecological assessments for development work may not have been identified through our search.
-We were limited by the reporting from included studies, particularly for measuring trapping effort and identifying habitat types trapped.
-We imputed trapping effort for studies that did not report any measure of effort, removing these studies had no meaningful effect on the association of human population density and trapping effort in sensitivity analysis.
+We collated studies from a diverse range of sources, however, it remains possible that rodent surveys have been performed for additional purposes that would not be captured in our survey of the scientific and "grey" literature, such as risk assessments or ecological assessments for development work and resource extraction.
+Collated data limited by the reporting from included studies, particularly for measuring trapping effort and identifying habitat types trapped.
+However, we found that this did not bias our investigation into the association between human population density and trapping effort.
 Habitat type and human population density were obtained from 2005, the median year in which studies were conducted.
-Changes in both human populations and land use in the studied period (1969-2020) have been large across West Africa, this introduces the possibility that these values were not appropriate for the region at the time when the rodent trapping study was conducted.
+Since changes in both human populations and land use in the studied period (1969-2020) have been large across West Africa, this introduces the possibility that these values were not appropriate for the region at the time when the rodent trapping study was conducted.
 
@@
 
-Here we have identified important biases in the locations of studies, the trapping effort for different rodent species and the selection of zoonotic pathogens to assay across the West African region.
+Here we have identified important biases in the locations of studies, the trapping effort for different rodent species and the selection of zoonotic pathogens assayed across the West African region.
 The use of these data to infer risk of zoonotic spillover or novel disease emergence is likely inappropriate unless these systematic biases are addressed.
-We hope that this current synthesis of data will encourage further comprehensive ecological studies on rodent species assemblages and their pathogens.
+We hope that this synthesis of current data will encourage further comprehensive ecological studies on rodent species assemblages and their pathogens.
+We would encourage studies to report uncertainty introduced by these potential biases in model outputs.
 One priority would be to better understand the true distribution of known rodent reservoirs (particularly *Mastomys natalensis*) across the region with particular emphasis made to the habitat in which they reside and the structure of rodent assemblages.
@@
 
-In conclusion we present a comprehensive synthesis of rodent trapping studies across West Africa, we identify important biases in the sampling of rodents and their pathogens that limits inference able to be drawn from current evidence.
+In conclusion we present a comprehensive synthesis of rodent trapping studies across West Africa, identifying important biases in the sampling of rodents and their pathogens that limit inference able to be drawn from current evidence.
 We identify evidence gaps in the current literature and highlight geographic regions, habitat types and host-pathogen associations that would benefit from targeted research effort.
@@
 
-**Included studies table**
+```{r supplementary-3, include=TRUE, echo=FALSE}
+supplementary_table <- readRDS(here::here("./data_clean/supplementary_3.rds")) %>%
+  dplyr::select(-Author)
+
+set_flextable_defaults(
+  font.family = "Calibri",
+  font.size = 8,
+  font.color = "black",
+  theme_fun = "theme_booktabs",
+  fonts_ignore = TRUE
+)
+
+flextable(supplementary_table) %>%
+  colformat_md(j = c("Reference", "Link")) %>%
+  colformat_date(j = "Year publication", fmt_date = "%Y") %>%
+  autofit()
+```
 

### 2021-10-21 scripts/1_update_data.R

@@ -2,7 +2,7 @@ source(here::here("scripts", "libraries.R"))
 
 drive_download("https://docs.google.com/spreadsheets/d/1rQYjHhk6uk1PoKZZVsgFlmuqWGicU2tTisk9ddfAwTM/edit#gid=0", path = here("data_download", "included_studies.xlsx"), overwrite = T)
 
-studies <- read_xlsx(here("data_download", "included_studies.xlsx"), sheet = "study")
+studies <- read_xlsx(here("data_download", "included_studies.xlsx"), sheet = "study", col_types = "text")
 write_rds(studies, here("data_raw", "studies.rds"))
 
 rodent_data <- read_xlsx(here("data_download", "included_studies.xlsx"), sheet = "trapping", col_types = c("guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess", "guess",

### 2021-10-21 scripts/analysis.R

@@ -444,10 +444,11 @@ plot_habitats <- bind_rows(plot_wa_habitats,
        y = "Proportion of region",
        fill = "Region") +
   scale_fill_manual(values = c("#fde725", "#440154")) +
-  theme_minimal()
+  theme_minimal() +
+  theme(text = element_text(size = 16))
 
 write_rds(plot_habitats, here("plots", "trap_habitats.rds"))
-save_plot(here("figures", "Figure_3.png"), plot_grid(trap_habitats), base_height = 10, base_width = 17)
+save_plot(here("figures", "Figure_3.png"), plot_grid(plot_habitats), base_height = 8, base_width = 12)
 
 habitat_types <- rodent_data %>%
   dplyr::select(unique_id, country, region, town_village, all_of(habitat_split), trap_night_unit, trap_nights) %>%
@@ -895,7 +896,10 @@ matrix_hp <- long_pathogen %>%
 greater_4 <- matrix_hp %>%
   group_by(classification) %>%
   add_count() %>%
-  filter(n >= 4)
+  filter(n >= 4) %>%
+  mutate(pathogen_tested = case_when(pathogen_tested == "K pneumoniae esbl" ~ "K. pneumonia ESBL",
+                                     pathogen_tested == "E coli esbl" ~ "E. coli ESBL",
+                                     TRUE ~ pathogen_tested))
 
 less_4 <- matrix_hp %>%
   group_by(classification) %>%
@@ -924,7 +928,8 @@ hp_g4 <- ggplot(greater_4,
   labs(x = element_blank(),
        y = element_blank(),
        colour = "Proportion positive",
-       size = "Number tested")
+       size = "Number tested") +
+  theme(text = element_text(size = 18))
 
 hp_l4 <- ggplot(less_4,
                 aes(x = pathogen_tested, y = fct_rev(classification), colour = Percent_positive, size = Tested)) +

### 2021-10-21 scripts/mapping_traps.R

@@ -154,7 +154,7 @@ trap_night_density_level2 <- ggplot() +
           lwd = 0.05,
           colour = "gray") +
   scale_fill_viridis_d(aesthetics = "fill",
-                       direction = -1,
+                       direction = 1,
                        labels = c("0 - 0.001", "0.001 - 0.01", "0.01 - 0.1",
                                   "0.1 - 1", "1 - 10", "10 - 100", "100 - 200",
                                   "No trapping")) +
@@ -302,7 +302,7 @@ m1 <- plot(sm(map_1, 1), n = 150, too.far = 0.02) +
   labs(title = element_blank(),
        x = element_blank(),
        y = element_blank(),
-       fill = "Linear predictor") +
+       fill = "Relative trapping density") +
   annotation_north_arrow(height = unit(1, "cm"),
                          style = north_arrow_minimal(text_size = 8)) +
   annotation_scale(height = unit(0.1, "cm"),

### 2022-01-29 reports/manuscript_v2.Rmd
updated

@@
 author: 
-  - "David Simons, Lauren Attfield, Kate Jones, Deborah Watson-Jones and Richard Kock"
+  - "David Simons, Lauren A. Attfield, Kate E. Jones, Deborah Watson-Jones and Richard Kock"
     # url: https://www.dsimons.org
@@
 These reservoir species are globally distributed with West Africa containing multiple known and several further predicted reservoir species.
-Typically, International Union for Conservation of Nature (IUCN) species distribution maps and Global Biodiversity Information Facility (GBIF) presence maps are used to determine regions at risk for zoonotic spillover events, however, these datasets are biased by incomplete sampling.
+Typically, International Union for Conservation of Nature (IUCN) species distribution maps and Global Biodiversity Information Facility (GBIF) presence maps are used to determine regions at risk for zoonotic spillover events.
+However, these datasets are biased by incomplete sampling.
 To investigate the impact of rodent species sampling heterogeneity on zoonotic spillover risk we have systematically reviewed rodent trapping studies to produce a contextually rich dataset that can be used to explore this risk at finer spatial resolution.
@@
 
-The potential effect of zoonotic infectious disease outbreaks on human health and societies has been dramatically highlighted through the ongoing SARS-CoV-2 pandemic and recent Ebola virus outbreaks.
-The number of zoonotic spillover events is projected to increase under intensifying anthropogenic pressure such as; increased human populations [@allen_global_2017], increasing urbanisation [@hassell_urbanization_2017], and global climate change [@morse_prediction_2012].
+The potential effect of zoonotic infectious disease outbreaks on human health and societies has been dramatically highlighted by the ongoing SARS-CoV-2 pandemic and recent Ebola virus outbreaks.
+The number of zoonotic spillover events is projected to increase under intensifying anthropogenic pressure such as; increased human populations [@allen_global_2017], increasing urbanisation [@hassell_urbanization_2017], land use change [@@garcia-pena_land-use_2021] and global climate change [@morse_prediction_2012].
 In addition, two taxa - rodents (Rodentia) and bats (Chiroptera) - are proposed to contribute to the greatest number of zoonotic pathogens and the greatest risk of novel pathogen emergence [@han_rodent_2015].
 Of 2,220 extant rodent species, 244 (10.7%) are described as being reservoirs of 85 zoonoses [@han_global_2016].
-Specifically, West Africa has previously been identified as region at increased hazard for rodent-borne zoonotic spillover events under different projected scenarios [@garcia-pena_land-use_2021].
+Specifically, West Africa has previously been identified as a region at increased hazard for rodent-borne zoonotic spillover events under different projected scenarios [@garcia-pena_land-use_2021].
 Rodents are implicated in the transmission pathways of several human diseases in this region, including, Lassa fever, Schistosomiasis and Leptospirosis [@meerburg_rodent-borne_2009].
@@
 The role of rodent species in zoonotic infectious disease spillover or novel pathogen emergence can be viewed as an ecosystem disservice.
-Rodents typically demonstrate “fast” life histories [@dobson_fast_2007] due to traits such as early maturation; traits which are also associated with zoonotic reservoirs [@han_rodent_2015; @albery_fast-lived_2021].
+Rodents typically demonstrate “fast” life histories [@dobson_fast_2007] due to traits such as early maturation and short gestation, which are also associated with zoonotic reservoirs [@han_rodent_2015; @albery_fast-lived_2021].
 Further, these traits are prevalent in species that thrive in human dominated landscapes where they displace species that are less likely to be reservoirs of zoonotic pathogen [@gibb_zoonotic_2020].
@@
 Rodent trapping studies provide a useful method to describe rodent population assemblages and survey for potential zoonoses.
-Studies have been conducted in West Africa to both identify novel potential zoonotic pathogens within rodents [@usaid_one_2021] and investigate the prevalence of known pathogens within known rodent hosts  to understand the distribution and burden of human disease (e.g., for Lassa fever [@fichet-calvet_diversity_2009] and Schistosomiasis [@catalano_multihost_2020]).
+Studies have been conducted in West Africa to identify novel potential zoonotic pathogens within rodents [@usaid_one_2021] and to investigate the prevalence of known pathogens within known rodent hosts  to understand the distribution and burden of human disease (e.g., for Lassa fever [@fichet-calvet_diversity_2009] and Schistosomiasis [@catalano_multihost_2020]).
 These studies provide contextually rich information on when, where and under what conditions rodents were trapped that are typically missing from the global datasets [@bovendorp_optimising_2017].
@@
 
-Multiple studies have identified potential geographic hotspots where virus and host species diversity may be expected to be at its greatest, this has been used to predict regional zoonosis spillover and novel pathogen emergence risks [@grace_mapping_2012; @han_global_2016; @allen_global_2017].
-However, there remains the potential for confounding factors having important negative effect on these spatial distributions through bias generated by study design and selection of sampling sites [@plowright_sampling_2019].
+Multiple studies have identified potential geographic hotspots where virus and host species diversity may be expected to be at its greatest, this has been used to predict regional zoonotic spillover and novel pathogen emergence risk [@grace_mapping_2012; @han_global_2016; @allen_global_2017].
+However, these spatial distributions could be confounded by bias generated by study design and selection of sampling sites [@plowright_sampling_2019].
 For example, systematically increased sampling (e.g.
 more intensive studies over longer time periods) or over-representation of certain habitats (e.g., peri-urban landscapes) could lead to an apparent association between locations and risk that is driven by these factors rather than an underlying host and virus association [@wille_how_2021; @gibb_mammal_2021].
-Conversely some regions may not be sampled adequately and therefore under-represented in these datasets due to sparse human populations or inaccessible habitats, which may result in these regions being inappropriately reported as at low risk of novel pathogen emergence or zoonotic spillover events.
+Conversely some regions may not be sampled adequately and therefore be under-represented in these datasets due to sparse human populations or inaccessible habitats, which may result in these regions being inappropriately reported as at low risk of novel pathogen emergence or zoonotic spillover events.
 
-Here, we synthesis rodent trapping studies performed across West Africa and identify the location and habitat types in which they have been conducted, the microorganisms assessed and the host-pathogen associations that have been reported in order to quantify the potential bias and to identify regions requiring further focussed investigation.
+Here, we synthesise rodent trapping studies performed across West Africa and identify the location and habitat types in which they have been conducted, the microorganisms assessed and the host-pathogen associations that have been reported in order to quantify the potential bias and to identify regions requiring further focussed investigation.
 
@@
 
-We included studies if they met all of the following inclusion criteria; i) reported findings from trapping studies where the target was a small mammal, ii) described the type of trap used or the length of trapping activity or the location of the trapping activity, iii) included trapping activity from at least one West African country (Benin, Burkina Faso, Cape Verde, Côte d'Ivoire, Eastern Sahara, Gambia, Ghana, Guinea, Guinea-Bissau, Liberia, Mali, Mauritania, Niger, Nigeria, Saint Helena, Ascension and Tristan da Cunha, Senegal, Sierra Leone and Togo), iv) recorded the genus or species of trapped individuals, v) were published in a peer-reviewed journal or as a pre-print on a digital platform or as a report by a credible organisation.
+We included studies if they met all of the following inclusion criteria; i) reported findings from trapping studies where the target was a small mammal, ii) described the type of trap used or the length of trapping activity or the location of the trapping activity, iii) included trapping activity from at least one West African country (Benin, Burkina Faso, Cape Verde, Côte d'Ivoire, Eastern Sahara, Gambia, Ghana, Guinea, Guinea-Bissau, Liberia, Mali, Mauritania, Niger, Nigeria, Senegal, Sierra Leone and Togo), iv) recorded the genus or species of trapped individuals, v) were published in a peer-reviewed journal or as a pre-print on a digital platform or as a report by a credible organisation.
 We excluded studies if they met any of the following exclusion criteria: i) reported data that were duplicated from a previously included study, ii) no full text available, iii) not available in English.
@@
 Extracted variables included i) study identifiers; ii) study aims; iii) trapping methodology; iv) geolocation data; v) method of speciation; vi) trapping locations and dates; vii) trapped species; viii) number of trap-nights and ix) microorganisms/pathogens of interest.
-The data extraction tool and the resulting dataset is archived and available in [Supplementary 1](https://docs.google.com/spreadsheets/d/1rQYjHhk6uk1PoKZZVsgFlmuqWGicU2tTisk9ddfAwTM/edit?usp=sharing).
+The data extraction tool and the resulting dataset is archived and available in [Supplementary table 1](https://docs.google.com/spreadsheets/d/1rQYjHhk6uk1PoKZZVsgFlmuqWGicU2tTisk9ddfAwTM/edit?usp=sharing).
 
@@
 
-The linear predictor output of this model is the trapping density adjusted for human population density in 2005. We performed sensitivity analysis removing sites with imputed trapping effort.
+The linear predictor output of this model is the trapping density adjusted for human population density in 2005. We performed sensitivity analysis by removing sites with imputed trapping effort.
 
@@
 
-Finally, we mapped the presence and absence of rodent species and compared this graphically by overlaying trapping locations to the presence produced from GBIF [@gbif_the_global_biodiversity_information_facility_occurrence_2021] and IUCN [@iucn_iucn_2021] datasets to give a measure of the extent to which each species' range has been sampled.
+Finally, we mapped the presence and absence of rodent species and compared this graphically by overlaying trapping locations to the presence produced from GBIF [@gbif_the_global_biodiversity_information_facility_occurrence_2021] and IUCN [@iucn_iucn_2021] datasets to give a measure of the extent to which each species' range has been sampled, understand the discrepencies between these data sources and to understand the potential impact of biases within these data on subsequently produced models.
 
@@
 We summarised the assays used for microorganisms, the presence and absence of these microorganisms, their host species and the locations from where the samples were obtained.
-We describe the different assays used for each microorganism across included studies.
-The presence of a microorganism was determined by a report of a positive assay (e.g., PCR, serology, viral culture) in a rodent sample at a defined location, absence was determined by negative results in all rodent samples at a defined location.
-We describe the number of microorganisms tested for in each rodent species and explore the geographic distributions these samples were obtained from.
-We investigated the association between rodent species and the detection of potential pathogens and report the proportion of positive and negative tests for each species and pathogen pair.
+We described the different assays used for each microorganism across included studies.
+The presence of a microorganism was determined by a report of a positive assay (e.g., PCR, serology, viral culture) in a rodent sample at a defined location and absence of a microorganism was determined by negative results in all rodent samples at a defined location.
+We described the number of microorganisms tested for in each rodent species and explored the geographic distributions these samples were obtained from.
+We investigated the association between rodent species and the detection of potential pathogens and reported the proportion of positive and negative tests for each species and pathogen pair.
 
@@
 
-![**Year trapping activity begun, ended and the year of publication for included studies.** Each row represents one of the 126 included studies, green points indicate when the trapping was initiated, blue points indicated the year trapping was concluded.
+![**Year trapping activity begun, ended and the year of publication for included studies.** Each row represents one of the 126 included studies.
+Green points indicate when the trapping was initiated and blue points indicated the year trapping was concluded.
 For studies completed within one year the blue point overlies the green.
@@
 Fewer than 5 studies were conducted in Liberia, Burkina Faso, and Guinea-Bissau.
-Thirty-one (25%) studies reported trapping at a single study site, 46 (37%) studies trapped at between two and five study sites, abd the remaining 49 studies trapped at between six and 93 study sites.
+Thirty-one (25%) studies reported trapping at a single study site, 46 (37%) studies trapped at between two and five study sites, and the remaining 49 studies trapped at between six and 93 study sites.
 Trap sites were situated in 273 (19.3%) of 1414 level 2 administrative regions in the 14 West African nations.
 The areas with highest trap night density included the capital cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako), and Ghana (Accra) and the largest cities of Côte d'Ivoire (Abidjan), and Benin (Cotonou).
-Outside of these cities, Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone had the highest density of trap nights (Figure 2.
+Outside of these cities, Northern Senegal (Fatique, Thies, Saint-Louis and Kédougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone had the highest density of trap nights (Figure 2.
 A).
 
-Greater numbers of trap nights were conducted in areas with higher population densities (*R~2~* = 0.18) adjusting for spatial autocorrelation.
+Greater numbers of trap nights were conducted in areas with higher population densities (*R^2^* = 0.26), adjusting for spatial autocorrelation (Figure 2.
+B).
 In particular Mauritania, Northern Senegal, Eastern Sierra Leone, Eastern Guinea and South West Nigeria were trapped at higher rates than would be expected based on their population densities.
 South West Nigeria, Northern Nigeria, Liberia, Côte d'Ivoire, Ghana, Niger, and Burkina Faso were trapped at lower rates than would be expected.
-Excluding sites with imputed trap nights had no meaningful effect on the association of human population density and trapping effort in sensitivity analysis.
+Excluding sites with imputed trap nights reduced the number of trapping sites contributing data.
+Despite this limitation, Mauritania, Northern Senegal and Sierra Leone remained as regions trapped at higher than expected rates with Nigeria as being trapped at lower than expected rates (Supplementary Figure 1.).
 
@@
 The colour of the region corresponds to trap night density per 1000 km^2.
-Purple regions correspond to low trap night density.
+Purple regions correspond to low trap night density and yellow regions correspond to high trap night density.
 Colourless regions are level 2 administrative areas with no reported trapping.
@@
 Yellow regions represent areas with higher-than-expected trapping effort for the regional population density and purple regions represent areas trapped at lower-than-expected rates based on their population density.
-Increased transparency of shading indicates uncertainty in the linear predictor due to limited precision of reported trap site location.](../figures/Figure_2.png)
+Increased transparency of shading indicates uncertainty in the linear predictor introduced by the smoothing terms on geographical coordinates and population density.](../figures/Figure_2.png)
 
@@
 
-![**Regional land cover of West Africa comparing all regions and trapped regions.** The proportion of the 10 land cover classes reported by the ESA Copernicus dataset at 300m^2^ resolution are shown for all regions of West Africa (yellow) and for regions in which trapping occurred (purple).
+![**Regional land cover of West Africa comparing all regions and trapped regions.** The proportion of the 10 land cover classes reported by the ESA Copernicus dataset (2005) at 300m^2^ resolution are shown for all regions of West Africa (yellow) and for regions in which trapping occurred (purple).
 Rodent trapping occurred in regions over-representative for cropland, mosaic landscapes and shrubland wile occurring in regions under-representative for bare- and sparse vegetation land cover classes.](../figures/Figure_3.png)
@@
 
-74,017 small mammals were trapped (excluding 592 were trapped outside of West African countries).
-2,830 (4%) trapped individuals were identified to order level (Rodentia), 7,283 (10%) were identified to genus level, and the remaining 63,904 (86%) were identified to species level.
+Overall, 4,017 small mammals were trapped (excluding 592 which were trapped outside West African countries).
+Of these trapped individuals, 2,830 (4%) were identified to order level (Rodentia), 7,283 (10%) were identified to genus level, and the remaining 63,904 (86%) were identified to species level.
 In studies reporting the number of trap nights the median trap success rate of traps placed in or around buildings was 13% (IQR 6-24%), compared with a median trap success rate of 3% (IQR 1-9%) in other habitats.
@@
 
-*Mastomys sp.* were the most commonly trapped genus (27,079, 37.7%), followed by; *Rattus sp.* (11,472, 16.1%), *Mus sp.* (8,624, 12.0%), *Arvicanthis sp.* (5,821, 8.1%) and *Praomys sp.* (5,409, 7.5%).
+*Mastomys sp.* were the most commonly trapped genus (27,079, 37.7%), followed by *Rattus sp.* (11,472, 16.1%), *Mus sp.* (8,624, 12.0%), *Arvicanthis sp.* (5,821, 8.1%) and *Praomys sp.* (5,409, 7.5%).
 *Mastomys natalensis* were the most commonly trapped species (11,116, 17.4%), followed by, *Rattus rattus* (9,959, 15.6%), *Mastomys erythroleucus* (7,386, 11.6%), *Mus musculus* (6,245, 9.8%), *Arvicanthis niloticus* (5,497, 8.6%), *Mastomys huberti* (4,699, 7.4%) and *Praomys daltoni* (1,854, 2.9%).
@@
 We identified locations of the presence or absence points of the seven most commonly trapped species, including true absences and pseudo-absences dependent on the trapping effort at the site.
-When comparing these with species presence obtained from GBIF and IUCN species range data we found that our data were obtained from more geographic locations within a species predicted range than within the GBIF data which were overrepresented from some West African countries, specifically Benin.
-We observed that both our included studies and GBIF data suggested that species have greater ranges than were projected from IUCN range maps.
+When comparing these with species presence obtained from GBIF and IUCN species range data (Figure 4.) we found that our data were obtained from more geographic locations within a species predicted range than within the GBIF data which were overrepresented from some West African countries, specifically Benin.
+We observed that our included studies combined with GBIF data suggested that species have greater ranges than were projected from IUCN range maps.
 This was particularly evident for the invasive species, *Mus musculus* and *Rattus rattus*, which were projected to have more limited distributions within West Africa.
 Further, we identified species absences were interspersed within presence locations obtained from both trapping studies and GBIF presence locations.
-Together this suggests greater heterogeneity in species distributions than would be apparent from either GBIF or IUCN data (Figure 4.).
 
@@
 
-Most studies tested for a single microorganism (39), 16 studies tested for two or more microorganisms.
+Most (39) studies tested for a single microorganism while 16 studies tested for two or more microorganisms.
 The most frequently investigated microorganisms were *Lassa mammarenavirus* (17 studies) or members of the Arenaviridae family (10), the spirochete bacteria *Borrelia sp.* (8), *Bartonella sp.* and *Toxoplasma gondii* (4), the remaining 25 microorganisms were reported in three or fewer studies.
@@
 The location of trapped rodents that were tested for the four most commonly assayed microorganisms did not encompass the entire range of these hosts.
-The presence and absence of Arenaviridae, predominantly *Lassa mammarenavirus*, in rodents was investigated in Benin, Côte d'Ivoire, Guinea, Mali, Nigeria and Sierra Leone.
+The presence and absence of Arenaviridae, predominantly *Lassa mammarenavirus*, in rodents was investigated in Benin, Côte d'Ivoire, Guinea, Mali, Mauritania, Nigeria, Senegal and Sierra Leone.
 No studies investigated the prevalence of these pathogens in hosts from Burkina-Faso, Guinea-Bissau, Liberia, Niger, Senegal and Togo.
-Arenaviridae were detected in rodents all the countries in which it was investigated except Benin (Figure 5A.).
+Arenaviridae were detected in rodents all the countries in which it was investigated except Benin and Mauritania (Figure 5A.).
 The presence and absence of *Borrelia sp.* was investigated in Benin, Mali, Mauritania, Niger and Senegal.
-*Borrelia sp.* were detected in 8 of 44 species or genera of rodents and shrews assayed in Mali with negative samples obtained from Benin, Mauritania, Niger and Senegal (Figure 5B.).
+*Borrelia sp.* were detected in 8 of 44 species or genera of rodents and shrews assayed in Mali and Senegal (Figure 5B.).
 The presence and absence of *Bartonella sp.* was investigated in Mali and Nigeria and Senegal with positive samples detected at all but two study sites in Senegal (Figure 5C.).
-The presence of *Toxoplasma gondii* was only investigated in Senegal and was detected in 6 of 8 rodent and shrew species assayed (Figure 5D.).
+The presence of *Toxoplasma gondii* was investigated in Cape Verde (not shown), Nigeria and Senegal and was detected in 6 of 8 rodent and shrew species assayed (Figure 5D.).
 
@@
 
-Ninety-seven rodent species were investigated for the presence of a microorganism of interest, and 42,940 assays were performed on 32,014 individual rodent samples.
+Ninety-seven rodent species were investigated for the presence of a microorganism of interest, from 54 studies, in which 42,940 assays were performed on 32,014 individual rodent samples.
 The rodent species most commonly assayed for these microorganisms were; *Rattus rattus* (n = 2,977) assessed for 24 microorganisms, *Mus musculus* (n = 3,402) for 23, *Mastomys natalensis* (n = 7,189) for 19, and *Mastomys erythroleucus* (n = 3,013) for 19 and *Arvicanthis niloticus* (n = 3,840) for 18.
@@
 *Lassa mammarenavirus* was detected in 14 species of rodents and Crocidura genera.
-*Mastomys natalensis* was the most common species to test positive (Figure 6.).
-Fifty-four species and genera were tested for arenaviruses, with 37 testing negative.
+*Mastomys natalensis* was the most commonly tested species with high rates of positive samples (14%) (Figure 6.).
+Fifty-four species and genera were tested for arenaviruses, with 37 entirely testing negative.
 *Bartonella sp.* infections were identified in all 11 species of rodents and shrews that were tested.
-*Mastomys erythroleucus* was the most common species to test positive.
 *Borrelia sp.* infections were identified in 13 species of rodents and shrews.
-*Mastomys natalensis* was the most common species to test positive.
-Forty-four species and genera were tested for *Borrelia sp.* with 36 testing negative.
+*Mastomys natalensis* were found to have the highest seroprevalence (11%).
 *Toxoplasma gondii* was detected in 6 species of rodents and shrews, with all individuals from two species *Arvicanthis niloticus* and *Praomys daltoni* testing negative.
-*Mus musculus* was the most common species to test positive.
 Finally, 38 species of Rodentia, 3 species of Soricomorpha and 2 species of Erinaceomorpha that were tested for microorganisms had entirely negative results.
@@
 ![**Number of individuals tested for each microorganism and proportion positive.** For the 27 most commonly assayed rodent species and genera, the percentage of positive assays against each of the 32 microorganisms tested for is shown.
-The size of a point is proportional to the number of individuals assayed for that microorganism.](../figures/Figure_6.png)
+The size of a point is proportional to the number of individual animals assayed for that microorganism.](../figures/Figure_6.png)
 
@@
 We identified 124 rodent trapping studies conducted in the West African region since 1965.
-These studies were conducted in 14 countries with 54 studies providing information about zoonotic pathogens in trapped rodents.
-Studies were conducted in all West African countries except Gambia, Saint Helena, Ascension and Tristan da Cunha and Togo with between- and within-country variability observed in the number of trap nights conducted.
+These studies were conducted in 14 countries, with 54 studies providing information about zoonotic pathogens in trapped rodents.
+Studies were conducted in all West African countries except Gambia and Togo with between- and within-country variability observed in the number of trap nights conducted.
 Greater than 70,000 rodents identified to genus or species level were trapped from 112 species of Rodentia.
@@
 The variability in sampling effort made by year and location must be accounted for when attempting to infer the risk of zoonotic spillover events or novel pathogen emergence [@lloyd-smith_nine_2015].
-The majority of studies have been conducted since 2000, during this period human population numbers, urbanisation, agricultural- and livestock intensification have progressed dramatically across West Africa [@acheampong_deforestation_2019; @connolly_extended_2021].
+The majority of studies have been conducted since 2000 during which time human population numbers, urbanisation, agricultural- and livestock intensification have progressed dramatically across West Africa [@acheampong_deforestation_2019; @connolly_extended_2021].
 These changes alter the availability of habitats for rodent species, potentially increasing the prevalence of commensal species which are more likely to be reservoirs of zoonoses.
@@
 This increased spatial resolution and the inclusion of presence-absence data presents a more complete representation of the spatial heterogeneity of rodent species' locations within their ranges.
-Failing to incorporate this heterogeneity in favour of using ranges or presence only data to produce rodent reservoir species' distributions can bias estimations of risk of zoonotic spillover or novel pathogen emergence through neglecting these important confounders.
+Failing to incorporate this heterogeneity in favour of using ranges or presence only data to produce rodent reservoir species' distributions can bias estimations of risk of zoonotic spillover or novel pathogen emergence through neglecting these important confounding factors.
 For example, the apparent absence of *Mastomys natalensis* at several trapping sites in Nigeria may explain some of the observed spatial variation in human Lassa fever cases that would not be expected based on this species IUCN range or presence only data from GBIF [@redding_geographical_2021].
@@
 Of these four pathogens, studies for *Lassa mammarenavirus* or other Arenaviridae were the most widely performed, both in absolute numbers but also by number of countries and number of study sites within countries.
-In contrast, *Toxoplasma sp.* were found to be prevalent in rodent hosts within Senegal and - despite a substantial, known, burden of human disease [@bigna_global_2020]- no studies investigated the prevalence of this pathogen in other West African countries.
+In contrast, *Toxoplasma sp.* were found to be prevalent in rodent hosts within Nigeria and Senegal - despite a substantial known, burden of human disease [@bigna_global_2020] - no studies investigated the prevalence of this pathogen in other West African countries.
 The limited geographic scope of sampling for these pathogens hampers efforts to understand the risk and human health burden of zoonotic spillover across this diverse region.
 
-Finally, we found that only five of the 112 rodent species trapped had been tested for more than half of the potential pathogens.
+Finally, we found that only five of the 112 rodent species trapped had been tested for more than half of the potential pathogens (Figure 6.).
 There is increasing emphasis on the role of co-infection and multi-species pathogen transmission systems on zoonotic disease spillover risk which depends on comprehensive testing of both known and unknown reservoir species [@sweeny_synzootics_2021].
@@
 We identified several limitations to our approach.
-We collated studies from a diverse range of sources, however, it remains possible that rodent surveys have been performed for additional purposes that would not be captured in our survey of the scientific and "grey" literature, such as risk assessments or ecological assessments for development work and resource extraction.
-Collated data limited by the reporting from included studies, particularly for measuring trapping effort and identifying habitat types trapped.
+We collated studies from a diverse range of sources but it is possible that rodent surveys that were performed for additional purposes would not be captured in our survey of the scientific and "grey" literature, such as for risk assessments or for  environmental impact assessments for development work or resource extraction.
+Collated data were limited by data availability from included studies, particularly for measuring trapping effort and identifying habitat types trapped.
 However, we found that this did not bias our investigation into the association between human population density and trapping effort.
@@
 
-# Supplementary 1
+# Supplementary table 1
 
@@
 
-# Supplementary 2
+# Supplementary table 2
 
@@
 
-# Supplementary 3
+# Supplementary table 3
 
@@
 
+# Supplementary figure 1
+![**Regional trap night density.
+Study sites where trap nights were imputed based on the number of rodents caught were removed from the model.
+This map of West Africa coloured by trapping density adjusted for human population density (2005) were obtained from our GAM (equation 1) in sensitivity analysis.
+Yellow regions represent areas with higher-than-expected trapping effort for the regional population density and purple regions represent areas trapped at lower-than-expected rates based on their population density.
+Increased transparency of shading indicates uncertainty in the linear predictor introduced by the smoothing terms on geographical coordinates and population density.](../figures/Figure_2_sensitivity.png)
+
 # References

### 2022-01-29 scripts/analysis.R

@@ -292,11 +292,24 @@ if(!file.exists(here("data_clean", "pop_tn_analysis.rds"))) {
 
   trapping_effort <- rodent_spatial %>%
     filter(unique_id %in% zoonotic_studies) %>%
-    distinct(unique_id, year_trapping, month_trapping, country, region, town_village, habitat, trap_nights, geometry)
+    distinct(unique_id, year_trapping, month_trapping, country, region, town_village, habitat, trap_nights, trap_night_data, geometry)
+
+  trapping_effort_sensitivity <- rodent_spatial %>%
+    filter(unique_id %in% zoonotic_studies) %>%
+    filter(trap_night_data != "Imputed") %>%
+    distinct(unique_id, year_trapping, month_trapping, country, region, town_village, habitat, trap_nights, trap_night_data, geometry)
 
   sites_2 <- st_intersection(x = level_2, y = trapping_effort)
 
+  sites_2_sensitivity <- st_intersection(x = level_2, y = trapping_effort_sensitivity)
+
   trap_nights_region <- tibble(sites_2) %>%
+    group_by(NAME_2, GID_2, unique_id, year_trapping, month_trapping, region, town_village, habitat) %>%
+    summarise(total_trap_nights = sum(trap_nights)) %>%
+    group_by(GID_2, NAME_2) %>%
+    summarise(region_trap_nights = sum(total_trap_nights))
+
+  trap_nights_region_sensitivity <- tibble(sites_2_sensitivity) %>%
     group_by(NAME_2, GID_2, unique_id, year_trapping, month_trapping, region, town_village, habitat) %>%
     summarise(total_trap_nights = sum(trap_nights)) %>%
     group_by(GID_2, NAME_2) %>%
@@ -310,13 +323,24 @@ if(!file.exists(here("data_clean", "pop_tn_analysis.rds"))) {
            tn_density = region_trap_nights/(as.numeric(area_m2)/1000000),
            tn_density = ifelse(is.na(tn_density), NA, tn_density))
 
+  sites_2_sensitivity <- level_2 %>%
+    left_join(., trap_nights_region_sensitivity, by = c("GID_2", "NAME_2")) %>%
+    mutate(region_trap_nights = case_when(is.na(region_trap_nights) ~ 0,
[…]

### 2022-01-29 scripts/mapping_pathogens.R

@@ -52,9 +52,16 @@ pathogen_map <- function(pathogen_genus) {
   pathogen_groups = list("A) Arenaviridae" = c("arenaviridae_species", "lassa_mammarenavirus", "mammarenavirus_species"),
                          "B) Borrelia" = c("borrelia_species", "borrelia"),
                          "C) Bartonella" = c("bartonella_species"),
-                         "D) Toxoplasma" = c("toxoplasma_gondii"))
+                         "D) Toxoplasma" = c("toxoplasma_gondii"),
+                         "Leptospirosis" = c("leptospirosis_species"),
+                         "Usutu" = c("usutu_virus"))
 
-  pathogen_data = list(pathogen = four_paths_wide %>%
+  pathogen_data = list(pathogen = wide_pathogen %>%
+                         tibble() %>%
+                         dplyr::select(1:16, matches(c(pathogen_groups[[pathogen_genus]]))) %>%
+                         left_join(., species_data,
+                                   by = c("gbif_id", "classification")) %>%
+                         distinct(record_id, name, .keep_all = T) %>%
                          dplyr::select(1:16, matches(c(pathogen_groups[[pathogen_genus]])),
                                        genus, species, genus_gbif, species_gbif) %>%
                          janitor::remove_empty("cols")  %>%
@@ -87,7 +94,12 @@ pathogen_map <- function(pathogen_genus) {
                                                     TRUE ~ "Negative"))
   )
 
-  species = list(rodents = four_paths_wide %>%
+  species = list(rodents = wide_pathogen %>%
+                   tibble() %>%
+                   dplyr::select(1:16, matches(c(pathogen_groups[[pathogen_genus]]))) %>%
+                   left_join(., species_data,
+                             by = c("gbif_id", "classification")) %>%
+                   distinct(record_id, name, .keep_all = T) %>%
                    dplyr::select(1:16, matches(c(pathogen_groups[[pathogen_genus]])),
                                  genus, species, genus_gbif, species_gbif) %>%
                    janitor::remove_empty("cols")  %>%
@@ -113,7 +125,7 @@ pathogen_map <- function(pathogen_genus) {
                    summarise(`Tested` = sum(number_tested),
                              `Positive` = sum(unique_positive),
                              `Negative` = `Tested`-`Positive`) %>%
-                   mutate(`Prop. positive` = round(`Positive`/`Tested`, 3)) %>%
[…]

### 2022-01-29 scripts/mapping_traps.R

@@ -308,6 +308,51 @@ m1 <- plot(sm(map_1, 1), n = 150, too.far = 0.02) +
   annotation_scale(height = unit(0.1, "cm"),
                    location = "tr")
 
+tn_pop_map_sens <- read_rds(here("data_clean", "tn_pop_model_sens.rds"))
+
+m1_ss <- plot(sm(map_1, 1), n = 150, too.far = 0.02) +
+  l_fitRaster(pTrans = function(.p) .p<0.05) +
+  geom_sf(data = included_countries, alpha = 0.1, lwd = 0.1, inherit.aes = FALSE) +
+  scale_fill_viridis_c(na.value = "#ffffff00") +
+  theme_minimal() +
+  labs(title = element_blank(),
+       x = element_blank(),
+       y = element_blank(),
+       fill = "Relative trapping density") +
+  annotation_north_arrow(height = unit(1, "cm"),
+                         style = north_arrow_minimal(text_size = 8)) +
+  annotation_scale(height = unit(0.1, "cm"),
+                   location = "tr")
+
+map_1_s <- getViz(tn_pop_map_sens)
+
+m1_s <- plot(sm(map_1_s, 1), n = 150, too.far = 0.02) +
+  l_fitRaster(pTrans = zto1(0.05, 2, 0.1)) +
+  geom_sf(data = included_countries, alpha = 0.1, lwd = 0.1, inherit.aes = FALSE) +
+  scale_fill_viridis_c(na.value = "#ffffff00") +
+  theme_minimal() +
+  labs(title = element_blank(),
+       x = element_blank(),
+       y = element_blank(),
+       fill = "Relative trapping density") +
+  annotation_north_arrow(height = unit(1, "cm"),
+                         style = north_arrow_minimal(text_size = 8)) +
+  annotation_scale(height = unit(0.1, "cm"),
+                   location = "tr")
+
+m1_s_ss <- plot(sm(map_1_s, 1), n = 150, too.far = 0.02) +
+  l_fitRaster(pTrans = function(.p) .p<0.05) +
+  geom_sf(data = included_countries, alpha = 0.1, lwd = 0.1, inherit.aes = FALSE) +
+  scale_fill_viridis_c(na.value = "#ffffff00") +
+  theme_minimal() +
+  labs(title = element_blank(),
+       x = element_blank(),
+       y = element_blank(),
+       fill = "Relative trapping density") +
+  annotation_north_arrow(height = unit(1, "cm"),
[…]

### 2022-03-09 reports/manuscript_v2.Rmd
updating analysis

@@
 ---
-title: "A scoping review of rodent trapping studies to understand the prevalence of zoonotic disease in West Africa."
+title: "Synthesising rodent trapping studies in West Africa to understand the hazard of zoonotic infectious disease outbreaks."
 date: "`r Sys.Date()`"
 author: 
-  - "David Simons, Lauren A. Attfield, Kate E. Jones, Deborah Watson-Jones and Richard Kock"
+  - "David Simons^1, 2, 3^, Lauren A. Attfield, Kate E. Jones, Deborah Watson-Jones and Richard Kock"
     # url: https://www.dsimons.org
@@
 The potential effect of zoonotic infectious disease outbreaks on human health and societies has been dramatically highlighted by the ongoing SARS-CoV-2 pandemic and recent Ebola virus outbreaks.
-The number of zoonotic spillover events is projected to increase under intensifying anthropogenic pressure such as; increased human populations [@allen_global_2017], increasing urbanisation [@hassell_urbanization_2017], land use change [@@garcia-pena_land-use_2021] and global climate change [@morse_prediction_2012].
+The number of zoonotic spillover events is projected to increase under intensifying anthropogenic pressure such as; increased human populations [@allen_global_2017], increasing urbanisation [@hassell_urbanization_2017], land use change [@garcia-pena_land-use_2021] and global climate change [@morse_prediction_2012].
 In addition, two taxa - rodents (Rodentia) and bats (Chiroptera) - are proposed to contribute to the greatest number of zoonotic pathogens and the greatest risk of novel pathogen emergence [@han_rodent_2015].

### 2022-03-09 reports/manuscript_v3.Rmd
updating analysis

---
title: "Synthesising rodent trapping studies in West Africa to understand the hazard of zoonotic infectious disease outbreaks."
date: "`r Sys.Date()`"
author: 
  - "David Simons"
  - "Lauren A. Attfield"
  - "Kate E. Jones"
  - "Deborah Watson-Jones"
  - "Richard Kock"
output: 
  word_document
always_allow_html: yes
bibliography: citations/manuscript_citations.bib
csl: elsevier_vancouver.csl
---

```{r echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
source(here::here("scripts", "manuscript_library.R"))
```

# Abstract

**To complete after written**

# Introduction

Zoonotic infectious disease spillover events into human populations  (animal-to-human transmission) can have important health and economic consequences at both the global and local level (e.g. SARS-CoV-2 and Ebola virus disease) (reference). The number of zoonotic disease spillover events and the emergence of novel zoonotic pathogens are projected to increase under intensifying anthropenic pressure driven by increased human populations, urbanisation, climate change and wildlife defaunation [@allen_global_2017; @hassell_urbanization_2017; @garcia-pena_land-use_2021; @mcmahon_ecosystem_2018]. West Africa has been identified as a nexus for these processes and a hotspot for both the risk of large epidemics of known zoonotic diseases and novel zoonotic pathogen emergence and [@grace_mapping_2012; @han_global_2016; @allen_global_2017]. Further, two taxa - rodents (Rodentia) and bats (Chiroptera) - contribute the greatest number of predicted novel zoonotic pathogens and known zoonoses [@han_rodent_2015; @gibb_mammal_2021]. Of 2,220 extant rodent species, 244 (10.7%) are described as reservoirs of 85 zoonotic pathogens [@han_global_2016]. Specifically, West Africa is identified as a region at increased hazard for rodent-borne zoonotic disease spillover events under different projected future land-use change scenarios [@garcia-pena_land-use_2021]. In West Africa rodents are involved in the transmission of multiple zoonotic diseases, including, Lassa fever, Schistosomiasis, Leptospirosis and Toxoplasmosis [@meerburg_rodent-borne_2009; @galeh_global_2020].

Rodent species form diverse assemblages, which provide important and beneficial ecosystem services including pest regulation and seed dispersal [@fischer_ecosystem_2018]. The role of rodent species in zoonotic disease spillover or novel zoonotic pathogens emergence are examples of ecosystem disservices. Rodents typically demonstrate “fast” life histories [@dobson_fast_2007] with traits such as early maturation and short gestation further associated with being zoonotic reservoirs [@han_rodent_2015; @albery_fast-lived_2021]. Traits associated with "fast" life histories are prevalent in species that thrive in human dominated landscapes, displacing species less likely to be reservoirs of zoonotic pathogens [@gibb_zoonotic_2020]. The widespread occurrence of reservoir species and their proximity to human activity make the description of rodent species assemblages and host-pathogen interactions vitally important to understanding the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence [@han_global_2016].

Despite the importance of understanding these complex systems in the hazard of zoonotic spillover, current evidence on host-pathogen associations is affected by taxonomic and geographical sampling biases [@gibb_ecosystem_2020; @gibb_mammal_2021]. Studies identifying potential geographic hotspots of zoonotic disease spillover and novel pathogen emergence are based on global datasets of host species distributions and host-pathogen associations (e.g., IUCN, GBIF, GIDEON) that suffer from several biases [@plowright_sampling_2019; @carlson_future_2021]. For example, systematically increased sampling, over-representation of certain habitats and clustering around areas of high human population could lead to an apparent association between locations and hazard that is driven by these factors rather than underlying host-pathogen associations [@wille_how_2021; @redding_evaluating_2017; @gibb_mammal_2021]. While other regions remain systematically undersampled (e.g., areas of sparse human populations), resulting in a reduced hazard of zoonotic spillover events or novel zoonotic pathogen emergence being attributed [@beck_spatial_2014]. Predictions of zoonotic disease spillover and novel zoonotic pathogen emergence must account for these biases to understand the future hazard of zoonotic diseases [@carlson_future_2021].

Rodent trapping studies provide contextually rich information on when, where and under what conditions rodents were trapped, potentially enriching global datasets and counteracting inherent biases [@bovendorp_optimising_2017]. Rodent trapping provides a useful method to describe rodent population assemblages, their geographic distribution and host-pathogen associations beyond what is currently available in global datasets. Studies have been conducted in West Africa to investigate the distribution of rodent speciesm, their species assemblages, the prevalence of known zoonoses within rodent hosts (e.g., Lassa fever, Schistosomiasis) and to identify novel zoonotic pathogens [@fichet-calvet_diversity_2009; @catalano_multihost_2020; @usaid_one_2021]. However, data from these studies have not been synthesised for inclusion in assessments of zoonotic disease spillover and novel zoonotic pathogen emergence.

Here, we synthesise the rodent trapping studies conducted across West African from a search of literature between 1964-2019. First, we investigate the geographic biases in the rodent trapping dataset in relation to human population density and land use classification. Second, we compare this to global host datasets (IUCN and GBIF) to understand differences in reported host geographic distributions. Third, we compare identified host-pathogen associations with global datasets (CLOVER) to understand discrepencies in rodent host-pathogen associations and report the proportion positive for each pathogen. Finally, within our dataset we investigate host-pathogen spatial extent to identify areas of undersampling of pathogens within host ranges. We expect that rodent trapping studies will provide an important additional source of high resolution data that can be used to enrich global datasets to better understand the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence across West Africa. We find that rodent trapping studies identify greater geographic ranges of important rodent hosts of zoonotic diseases than are reported from global datasets while also identifying locations and habitats in which further sampling is required. We also identify host-pathogen associations that are not reported in global host-pathogen association data, however, there remains systematic undersampling for these pathogens across the host ranges.

# Methods

## Data sources

### Host and pathogen trapping data

We conducted a search in Ovid MEDLINE, Web of Science (Core collection and Zoological Record), JSTOR, BioOne, African Journals Online, Global Health and the pre-print servers, BioRxiv and EcoEvoRxiv for the following terms as keywords, no date limits were set: (1) Rodent OR Rodent trap* AND (2) West Africa (or the individual countries).

We searched other resources including the UN Official Documents System, Open Grey, AGRIS FAO and Google Scholar using combinations of the above terms. Searches were run on 2021-03-01, returning studies conducted between 1964-2019.

We included studies if they met all of the following inclusion criteria; i) reported findings from trapping studies where the target was a small mammal, ii) described the type of trap used or the length of trapping activity or the location of the trapping activity, iii) included trapping activity from at least one West African country, iv) recorded the genus or species of trapped individuals, v) were published in a peer-reviewed journal or as a pre-print on a digital platform or as a report by a credible organisation. We excluded studies if they met any of the following exclusion criteria: i) reported data that were duplicated from a previously included study, ii) no full text available, iii) not available in English. One reviewer screened titles, abstracts and full texts against the inclusion and exclusion criteria. At each stage, a random subset (10%) was reviewed by a second reviewer.

We extracted data from eligible studies using a standardised tool that was piloted on 5 studies (Supplementary 1.). Data was abstracted into a Google Sheets document, which was archived on completion of data extraction (Supplementary 2.). Extracted variables included i) study identifiers; ii) study aims; iii) trapping methodology; iv) geolocation data; v) method of speciation; vi) trapping locations and dates; vii) trapped species; viii) number of trap-nights and ix) microorganisms/pathogens of interest. We summarised the number of studies, the year in which trapping occurred and the country in which they were conducted.

We extracted GPS locations for the most precise location presented, converting to decimal degrees as required. For studies not using standardised habitat recording (e.g., IUCN Habitat Classification Scheme (Version 3.1)), the explicit description of the habitat in which the trap was placed was extracted. For studies reporting multiple habitat types for a single trap, trap-line or trapping grid, a higher order classification of habitat type was recorded.

We mapped genus and species names to those in the Global Biodiversity Information Facility (GBIF) taxonomy  [@gbif_the_global_biodiversity_information_facility_global_2021]. We extracted information on the detection, non-detection and number of trapped individuals.

We extracted data on all pathogens assayed. The number of rodents tested and the number of positive or negative samples were recorded alongside the type of assay used. If studies reported indeterminate results this was noted. Pathogens were identified to species level, however, where an assay only allowed for attribution to a family of viruses or bacteria, the higher order grouping was used.

### Environmental covariates

### Rodent species ranges

### Rodent presence

### Host pathogen associations

CLOVER a synthesis of four host-pathogen datasets (GMPD2, EID2, HP3 and Shaw 2020) was released in 2021 and represents the most comprehensive global repository of host-pathogen associations for Bacteria, Viruses, Helminth, Protozoa and Fungi [clover_paper]. Data were obtained from an archived version of the dataset [clover_dataset].

## Included studies

We identified 4,282 relevant citations, with 126 rodent trapping studies included. The earliest trapping studies were conducted in 1964, with increasing numbers of studies being performed annually. The median year of first trapping activity was 2005, with the median length of trapping activity being 1 year (IQR 0-2 years) (Supplementary figure 1.). Studies were conducted in 14 West African countries with no studies reported from Gambia or Togo, at 1,331 sites (Figure 1A.). A minority (31, 25%) of studies trapped at a single study site, with 46 (37%) trapping at between two and five sites, the remaining 49 studies (38%) trapped at between six and 93 study sites. There were 581,426 reported trap nights, with 384,983 trap nights imputed for studies with no recording of trapping effort (Figure 1B.).

```{r trap-locations-trapnights, out.width = "120%", fig.cap = "Figure 1. A) The location of trapping sites in West Africa. No sites were recorded from Togo or Gambia. There is important heterogeneity observed in the coverage of each country by trap night (colour) and location of sites. For example Senegal, Mali and Sierra Leone have generally good coverage compared to Guinea and Burkina Faso. B) The distribution of trap nights performed at each study site on a log scale with a median of 244 trap nights (IQR 115-494) performed at each site"}

knitr::include_graphics(here("figures", "Figure_1_updated.png"))

```

Within the included studies trap success varied importantly between traps placed in or around buildings (13%, IQR 6-24%) compared to other habitats (3%, IQR 1-9%). In total 73,164 small mammals were trapped with 62,574 (85%) identified to species level and 7,760 (11%) identified to genus. The majority of the 147 identified species were Rodentia (112, 76%) of which Muridae (82, 73%) were the most common family.

62 studies tested for 32 microorganisms, defined to species or genus leve that are known or potential pathogens. The majority of studies tested for a single microorganism (39, 62%). The most frequently assayed microorganisms were *Lassa mammarenavirus* or Arenaviridae (27, 43%), *Borrelia sp.* (8, 13%), *Bartonella sp.* (4, 6%) and *Toxoplasma gondii* (4, 6%). The majority of studies used Polymerase Chain Reaction to detect microorganisms (32, 52%), with fewer studies using molecular based tests (11, 18%) or histological or direct visualisation assays (8, 13%).

From 32,014 individual rodent samples we produced 339 host-pathogen pairs. With *Rattus rattus*, *Mus musculus*, *Mastomys erythroleucus*, *Mastomys natalensis* and *A. niloticus* being assayed for at least 18 microorganisms.

## Analysis

### What is the extent of spatial bias in the rodent trapping data?

We used the GPS coordinates of single trap, trapping grid/line or study site and the number of trap nights to calculate trapping effort (trap night density) within level 2 administrative regions in West Africa. The median trap success rate for all rodents at a defined trap site was calculated separately for trap sites which included built environments and non-built environments due to important differences in trap success between these different environments. The number of rodents trapped at a trap site was then used in combination with these values of trap success to impute the number of trap nights for trap sites where there was no record of the number of trap nights. We summarised the habitat types of trap sites based on information reported in the study; the dictionary used to match habitat types is available in Supplementary 3.

To better understand the potential bias associated with studies for zoonotic disease spillover or novel zoonotic pathogen emergence we limited further analyses to zoonotic disease studies only. We used a Generalised Additive Model (GAM) incorporating a spatial interaction term to investigate the association between number of trap nights, proportion of urban, cropland and forest land classification and human population density [@pedersen_hierarchical_2019]. Covariates were obtained for 2005, the median year studies were conducted, proportions of habitat types were obtained from Copernicus 2005 land cover classification, global population density estimates were obtained from SEDAC [@socioeconomic_data_and_applications_center_gridded_2021]. We performed sensitivity analysis by removing sites with imputed trapping effort. The model structure was specified as:

$$
\begin{equation}
\tag{1}
\text{TN density} \sim \text{Tweedie(log(Population density) + (Urban) + (Cropland) + (Shrubland) + (Forest) + (Longitude * Latitude))}
\end{equation}
$$

The model was used to identify locations of high and low trapping effort relative to population density and land use classification. Predictions were limited to regions around observed trap sites.

### What is the difference between global datasets rodent host distributions and distributions from trapping studies?

We assessed the taxonomic coverage of the trapped taxa and for the seven species (*M. natalensis*, *R. rattus*, *M. erythroleucus*, *M. musculus*, *Arvicanthis niloticus*, *Praomys daltoni* and *Cricetomys gambianus*) with the most trap locations we mapped the detection and non-detection of rodent species and compared this to detection data obtained from GBIF and species ranges from IUCN [@gbif_the_global_biodiversity_information_facility_occurrence_2021; @iucn_iucn_2021]. We identified rodent locations from GBIF and the rodent trapping studies within and outside the IUCN expected range. We located these to level 2 administrative regions to calculate the discrepencies between observed detections and non-detections with IUCN distributions.

### Are rodent trapping derived host-pathogen associations present in global zoonoses datasets?

For studies investigating rodent zoonoses we produced host-pathogen pairs. For each host-pathogen pair we compared our records with those available in CLOVER reporting concordance and discordance. We identify host-pathogen pairs that are not present in CLOVER and contain assay results consistent with acute or prior infection.

### What is the spatial extent of pathogen testing within a hosts range?

For the four pathogens sampled at the most locations (Borreliaceae, Leptospiraceae, *Lassa mammarenavirus* and *Toxoplasma gondii*) we identify their host-pathogen associations. We identified the five species with most spatially complete sampling and associated these sampling locations with level 2 administrative regions. We compared the area of these regions to their IUCN range and detection range to calculate sampling coverage.

# Results

Further descriptive information from the included studies including geolocated trapping of species, their detection and non-detection alongside microorganism data has been made available online [@simons_rodents_2022].

## What is the extent of spatial bias in the rodent trapping data?

Trap sites were situated in 273 (19.3%) of 1414 level 2 administrative regions in 14 West African nations. The areas with highest trap night density included the capital and large cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako), Ghana (Accra),  Côte d'Ivoire (Abidjan), and Benin (Cotonou). Outside of these cities, regions in, Northern Senegal, Southern Guinea, Edo and Osun States in Nigeria and Eastern Sierra Leone had the highest density of trap nights (Figure 1A.).

Trapping sites were biased towards human modified landscapes (i.e., cropland, grassland and mosaic habitats), with under-representation of forest and bare habitats. However, urban habitats were undersampled.

Trapping effort, measured through trap nights, was biased towards areas with higher population density and predominantly urban and cropland land classifications (Figure 2.). In particular Northern Senegal, Eastern Sierra Leone, Eastern Guinea and South West Nigeria had relatively high trapping efforts while South East Nigeria, Northern Nigeria, Liberia, Côte d'Ivoire, Ghana, Niger, and Burkina Faso were trapped at lower trapping effort than would be expected. Excluding sites with imputed trap nights reduced the number of trapping sites contributing data. Despite this limitation, Mauritania, Northern Senegal and Sierra Leone remained as regions trapped at higher than expected rates with Nigeria as being trapped at lower than expected rates (Supplementary Figure 2.).

```{r gam-trapping-effort, out.width = "120%", fig.cap = "Figure 2. Relative trapping effort bias across West Africa from included studies adjusted for human population density, proportion cropland cover, proportion forest cover and proportion urban habitat. Predictions are limited to areas around trap sites (coloured areas), uncertainty in the estimate of bias is represented by colour transparency/pallor. Purple regions represent areas with higher than expected trapping effort, yellow regions represent areas lower than expected trapping effort"}

knitr::include_graphics(here("figures", "Figure_2_updated.png"))

```

## What is the difference between global datasets rodent host distributions and distributions from trapping studies?

For our species of interest, except *C. gambianus*, trapping studies provided more locations of detection and non-detection than were available from GBIF. For the West African endemic species IUCN ranges showed generally good concordance to both trapping studies and GBIF, although the range of *A. niloticus* and *P. daltoni* are greater than expected. The invasive species *R. rattus* and *M. musculus* are present in a much larger range than is expected (Figure 3A. and 3B.).

```{r comparing-trap-locations-A, out.width = "120%", fig.cap = "Figure 3A. Each row corresponds to a single rodent species. L) Presence recorded in GBIF overlaid on IUCN species range. R) Detection and non-detection from rodent trapping studies overlaid on IUCN species ranges. No IUCN range for *M. musculus* is available for the study region."}

knitr::include_graphics(here("figures", "Figure_3_test_a.png"))

```

```{r comparing-trap-locations-B, out.width = "120%", fig.cap = "Figure 3B. Each row corresponds to a single rodent species. L) Presence recorded in GBIF overlaid on IUCN species range. R) Detection and non-detection from rodent trapping studies overlaid on IUCN species ranges."}

knitr::include_graphics(here("figures", "Figure_3_test_b.png"))

```

We identified the proportion of a rodents expected range in which detections and non-detections occurred (Table 1.). GBIF sampling occurred over > 15% of a rodents expected range with detections occuring outside of IUCN ranges for all species, most noticeably for *A. niloticus* and *P. daltoni*. Trapping studies occurred over > 13% for all species except *C. gambianus* with detections occurring outside of IUCN ranges most noticeably for *A. niloticus* and *R. rattus*. Combining GBIF and trapping data increased coverage by a mean of 9% of the species range suggesting limited overlap between the regions studied. Non-detections of rodents occurred across large areas of their ranges with *A. niloticus* and *M. erythroleucus* not being detected in >15% of their expected range.

```{r table-1}

table_1 <- read_rds(here("tables", "Table_1_updated.rds"))

table_1

```

## Are rodent trapping derived host-pathogen associations present in global zoonoses datasets?

When limited to confirmed pathogens and classification of both pathogen and host to species level we identified 21 host-pathogen pairs among 11 rodent and 6 pathogen species (Figure 4.). We identified negative associations for 41 host-pathogen pairs among 31 rodent and 7 pathogen species. CLOVER contained 8 (38%) of our identified host-pathogen associations and positive associations for 6 (15%) of the negative associations.

```{r comparing-host-pathogen-interactions, out.width = "120%", fig.cap = "Figure 4. L) Identified species level host-pathogen associations through detection of acute infection (i.e. PCR, culture). Percentages and colour relate to the proportion of all assays that were positive. Associations with a black border are present in the CLOVER dataset. R) Identified species level host-pathogen associations through serological assays (i.e. ELISA). Percentages and colour relate to the proportion of all assays that were positive. Associations with a black border are present in the CLOVER dataset."}

knitr::include_graphics(here("figures", "Figure_4_updated.png"))

```

CLOVER included 399 host-pathogen associations we do not observe in rodent trapping studies among our included rodent species. The majority of these 329 (82%) pairs are from species with global distributions (*M. musculus*, *R. rattus* and *R. norvegicus*), or from wide ranging distributions in sub-Saharan Africa (31, 9%) (i.e., *A. niloticus*, *M. natalensis* and *A. albiventris*).

For pathogens not identified to species level (i.e. family or lower taxa) we identified 128 host-pathogen pairs among 32 rodent species and 25 pathogen families (Supplementary figure 3.), with CLOVER containing 52 (41%) of these associations.

Rodent trapping studies identified rodent host species for six pathogens, *Toxoplasma gondii* (4), *Lassa mammarenavirus* (3), Usutu virus (2), *Coxiella burnetii* (2), *Escherichia coli* and *Klebsiella pneumoniae* (both 1), that were not present in this global host-pathogen association dataset.

## What is the spatial extent of pathogen testing within a hosts range?

The four most widely sampled pathogen species/families were were Borreliaceae, *Lassa mammarenavirus*, Leptospiraceae and *Toxoplasma gondii* (Table 2.). Infection with species of Borreliaceae was assessed in 42 species, with evidence of infection in 17 rodent species. Among species with more than 500 samples the highest rates of infection were among *A. niloticus* (16%), *M. huberti* (11%) and *M. erythroleucus* (8%). The five most widely assayed species were sampled from a relative small area of their species ranges (7.3-16%), however, there was good representation when compared to where rodent trapping studies have occurred (24-85%). Infection with *Lassa mammarenavirus* was assessed in 28 species, with evidence of infection in 7 rodent species. Among species with more than 100 samples the highest rates of infection were among *M. natalensis* (22%), *M. erythroleucus* (6%) and *R. rattus* (4%). Species sampling occurred from limited areas within IUCN ranges (2-3%), however, sampling was more representative for rodent trapping studies (3-24%). Infection with species of Leptospiraceae was assessed in 8 species, with evidence of infection in 5 rodent species. Among species with more than 100 samples *M. natalensis* (19%), *R. rattus* (11%) and *Crocidura olivieri* (10%) were most frequently infected. Species sampling occurred across less than 1% of these species IUCN ranges and for all except *R. norvegicus* for less than 1% of their detection range in included studies. *Toxoplasma gondii* infection was assessed in 8 species, with evidence of infection in 6 rodent species. Among species with more than 100 samples *M. musculus* (7%), *M. erythroleucus* (4%) and *M. natalensis* (2%) were most commonly infected. Sampling occurred across less than 6% of their IUCN ranges, however, for *M. musculus*, *R. rattus* and *M. erythroleucus* sampling occurred across greater than 12% of their detection range.

```{r table-2}

table_2 <- read_rds(here("tables", "Table_2_updated.rds"))

table_2

```

# Discussion

We synthesised data from 126 rodent trapping studies containing information on more than 72,000 trapped rodents across 1,331 trapping sites producing an estimated 966,000 trap-nights. Trapping occurred across 14 West African countries, studies investigating known or potential zoonotic pathogens occurred within 19% of level 2 administrative regions across West Africa with important bias towards areas with high population density and human dominated landscapes. For several important rodent species we identified discordance between IUCN species ranges, GBIF presence data and detections and non-detections from rodent trapping studies, particularly among non-native species (*R. rattus* and *M. musculus*). We compared host-pathogen associations from rodent trapping studies with a global dataset and identified missing associations for several important zoonotic infectious diseases including, Lassa fever, Toxoplasmosis and Q fever in commonly occurring rodent species. Finally, we assessed the spatial coverage of testing for important zoonotic pathogens and found low representativeness across their host species range.

Rodent trapping studies provide important information on the distribution of rodent species and their assemblages, we have identified important bias in the locations in which studies to data have been performed with large regions of countries including Nigeria, Guinea and Cote d'Ivoire being understudied. Trapping sites are clustered in areas with relatively high human population density and anthropogenic land use which can importantly bias our understanding of the distribution of rodent species and their pathogens across West Africa. Further effort is required to minimise this bias by sampling currently undersampled regions.

Global datasets of rodent host distribution (e.g., IUCN) and presence (e.g., GBIF) provide useful resources to model the hazard of zoonotic disease spillover and novel pathogen emergence. We show here that non-detection data from rodent trapping studies can provide additional information to describe the heterogeneity of rodent assemblages across diverse habitats in West Africa. Combining presence data from GBIF and detection/non-detection data from rodent trapping studies can substantially increase coverage across a rodents range, allowing improved estimates of rodent host prevalence for zoonotic infectious disease studies.

Host-pathogen association datasets are becoming increasing comprehensive alongside increased access to molecular discovery techniques for potential pathogens. We show that currently available data from rodent trapping studies can supplement these resources to better understand the potential hosts and reservoirs of known zoonotic pathogens. Rodent trapping data can support studies using these global datasets to predict novel pathogen emergence and to quantify the risk of known zoonotic disease spillover based on host distributions.

We have identified important limitations in the sampling of rodent zoonoses for known diseases across West Africa. For example, studies of *Lassa mammarenavirus* have taken place across 3% of *M. natalensis*'s range in West Africa. Lassa fever is known to be endemic in eight West African countries, although it is potentially more widely spread. Further studies are required to investigate this pathogen across a wider geographic range. Other important zoonotic diseases including Leptospirosis and Toxoplasmosis have been assessed over very limited regions of their hosts distribution.

We propose that rodent trapping studies can provide important data to supplement global datasets to quantify the hazard of zoonotic disease spillover and novel pathogen emergence. They can reduce some of the bias in global datasets however, important limitations remain. Further effort is required to survey rodent populations in non-human dominated landscapes and to assay pathogens across more of a hosts range.

# Supplementary

## Supplementary figure 3

```{r comparing-host-pathogen-interactions-supplementary, out.width = "120%", fig.cap = "Figure 3. L) Identified host-pathogen associations at pathogen family level through detection of acute infection (i.e. PCR, culture). Percentages and colour relate to the proportion of all assays that were positive. Associations with a black border are present in the CLOVER dataset. R) Identified host-pathogen associations at pathogen family level through serological assays (i.e. ELISA). Percentages and colour relate to the proportion of all assays that were positive. Associations with a black border are present in the CLOVER dataset."}

knitr::include_graphics(here("figures", "Supplementary_figure_3_updated.png"))

```

### 2022-03-09 scripts/analysis.R

@@ -269,149 +269,11 @@ non_buildings_tn <- trap_nights_complete %>%
 
 summary(non_buildings_tn$trap_success)
 
-# Trap night and population -----------------------------------------------
-if(!file.exists(here("data_clean", "pop_tn_analysis.rds"))) {
-  rodent_spatial <- read_rds(here("data_clean", "rodent_spatial.rds")) %>%
-    dplyr::select(-trap_nights)
 
-  imputed_tn <- read_rds(here("data_clean", "imputed_trap_nights.rds"))
+# Trap night bias ---------------------------------------------------------
 
-  rodent_spatial <- rodent_spatial %>%
-    left_join(., imputed_tn,
-              by = c("unique_id", "year_trapping", "month_trapping",
-                     "region", "town_village", "habitat"))
+source(here("scripts", "tn_gam.R"))
 
-  level_2 <- read_rds(here("data_download", "admin_spatial", "level_2_admin.rds"))
-  non_trapped <- read_rds(here("data_download", "admin_spatial", "level_2_TGOGMB.rds"))
-
-  level_2 <- bind_rows(level_2, non_trapped)
-
-  zoonotic_studies <- studies %>%
-    filter(str_detect(aim, "Zoonoses")) %>%
-    pull(unique_id)
-
-  trapping_effort <- rodent_spatial %>%
-    filter(unique_id %in% zoonotic_studies) %>%
-    distinct(unique_id, year_trapping, month_trapping, country, region, town_village, habitat, trap_nights, trap_night_data, geometry)
-
-  trapping_effort_sensitivity <- rodent_spatial %>%
-    filter(unique_id %in% zoonotic_studies) %>%
-    filter(trap_night_data != "Imputed") %>%
-    distinct(unique_id, year_trapping, month_trapping, country, region, town_village, habitat, trap_nights, trap_night_data, geometry)
-
-  sites_2 <- st_intersection(x = level_2, y = trapping_effort)
-
-  sites_2_sensitivity <- st_intersection(x = level_2, y = trapping_effort_sensitivity)
-
-  trap_nights_region <- tibble(sites_2) %>%
-    group_by(NAME_2, GID_2, unique_id, year_trapping, month_trapping, region, town_village, habitat) %>%
-    summarise(total_trap_nights = sum(trap_nights)) %>%
[…]

### 2022-03-09 scripts/libraries.R

@@ -33,7 +33,9 @@ pkgs =
     "gridExtra",
     "lubridate",
     "see",
-    "scales"
+    "scales",
+    "units",
+    "flextable"
   )
 pacman::p_load(pkgs, character.only = T)
 

### 2022-03-09 scripts/pathogen_spatial.R

@@ -1,0 +1,267 @@
+# Load data ---------------------------------------------------------------
+sf_use_s2(use_s2 = FALSE)
+
+rodent_spatial <- read_rds(here("data_clean", "rodent_spatial.rds")) %>%
+  dplyr::select(-trap_nights)
+
+pathogen <- read_rds(here("data_clean", "long_pathogen.rds")) %>%
+  select(record_id, classification, assay, number, pathogen_tested, geometry) %>%
+  mutate(assay = case_when(str_detect(assay, "tested") ~ "n_tested",
+                           str_detect(assay, "pcr") ~ "n_positive_pcr",
+                           str_detect(assay, "ab_ag") ~ "n_positive_ab_ag",
+                           str_detect(assay, "histo") ~ "n_positive_histo",
+                           str_detect(assay, "culture") ~ "n_positive_culture")) %>%
+  group_by(pathogen_tested) %>%
+  group_split()
+
+all_pathogens <- lapply(pathogen, function(x) { x %>%
+    pivot_wider(names_from = assay, values_from = number) })
+
+names(all_pathogens) <- lapply(pathogen, function(x) {
+
+  pathogen_names <- x %>%
+    pull(pathogen_tested) %>%
+    unique(.) }) %>%
+  unlist()
+
+all_pathogens[["lassa_mammarenavirus"]] <- all_pathogens[["lassa_mammarenavirus"]] %>%
+  unnest_wider(n_tested, names_sep = "_") %>%
+  unnest_wider(n_positive_pcr, names_sep = "_") %>%
+  unnest_wider(n_positive_ab_ag, names_sep = "_") %>%
+  unnest_wider(n_positive_culture, names_sep = "_") %>%
+  rowwise(record_id) %>%
+  mutate(n_tested = max(n_tested_1, n_tested_2, na.rm = TRUE),
+         n_positive_ab_ag = case_when(!is.na(n_positive_ab_ag_1) & !is.na(n_positive_ab_ag_2) ~ max(n_positive_ab_ag_1, n_positive_ab_ag_2, na.rm = TRUE),
+                                      TRUE ~ as.numeric(NA))) %>%
+  rename("n_positive_pcr" = n_positive_pcr_1,
+         "n_positive_culture" = n_positive_culture_1) %>%
+  select(record_id, classification, pathogen_tested, geometry, n_tested, n_positive_pcr, n_positive_ab_ag, n_positive_culture)
+
[…]

### 2022-05-11 reports/manuscript_v4.Rmd
updating scripts

--- 2022-03-09 reports/manuscript_v3.Rmd
@@
 ---
-title: "Synthesising rodent trapping studies in West Africa to understand the hazard of zoonotic infectious disease outbreaks."
+title: "Rodent trapping studies can improve our understanding of the hazards of zoonotic pathogen spillover and novel zoonotic pathogen emergence."
 date: "`r Sys.Date()`"
@@
 output: 
-  word_document
+  officedown::rdocx_document:
+    tables:
+      caption:
+       style: false
+    plots:
+      caption:
+        style: false
 always_allow_html: yes
@@
                       message = FALSE,
-                      warning = FALSE)
+                      warning = FALSE,
+                      crop = knitr::hook_pdfcrop)
 source(here::here("scripts", "manuscript_library.R"))
@@
 
-**To complete after written**
+Rodents are important reservoirs of known zoonotic infectious diseases and novel zoonotic pathogens.
+Rodent reservoirs are globally distributed with West Africa containing multiple host species of zoonotic infectious diseases.
+Ongoing anthropogenic land use change alters the composition of rodent species assemblages which can alter the hazard of zoonotic disease spillover and emergence.
+Current global datasets to understand these hazards suffer from biases.
+Here, we synthesise data from 127 rodent trapping studies including over 72,000 individual rodent detections to understand the range and presence of important rodent reservoir species to quantify the bias in global datasets.
+We identify the rodent trapping studies are biased towards human dominated landscapes with important sampling biases across West Africa.
+We find that these rodent trapping studies can complement current global datasets on rodent population distributions which can minimise some of these observed biases.
+For four regionally important zoonotic infectious diseases we identify host-pathogen associations not included in a comprehensive global dataset.
+Rodent trapping studies can further enrich these resources through geolocating these associations and producing measures of pathogen prevalence within hosts.
+Finally, we find that for most of these diseases the proportion of a rodent hosts range that has been sampled remains small.
+Further rodent trapping studies are required to understand the true hazard of zoonotic spillover of these diseases across West Africa.
+Our results highlight that understanding rodent sampling bias is important when assessing the risk for zoonotic spillover events from rodents.
+The synthesis of contextually rich rodent trapping data contributes important information that is lacking in IUCN distribution maps and GBIF species presence data.
+We anticipate that this dataset can support the production of more complete understanding of the hazard of zoonotic spillover events.
 
@@
 
-Zoonotic infectious disease spillover events into human populations  (animal-to-human transmission) can have important health and economic consequences at both the global and local level (e.g.
-SARS-CoV-2 and Ebola virus disease) (reference).
-The number of zoonotic disease spillover events and the emergence of novel zoonotic pathogens are projected to increase under intensifying anthropenic pressure driven by increased human populations, urbanisation, climate change and wildlife defaunation [@allen_global_2017; @hassell_urbanization_2017; @garcia-pena_land-use_2021; @mcmahon_ecosystem_2018].
-West Africa has been identified as a nexus for these processes and a hotspot for both the risk of large epidemics of known zoonotic diseases and novel zoonotic pathogen emergence and [@grace_mapping_2012; @han_global_2016; @allen_global_2017].
-Further, two taxa - rodents (Rodentia) and bats (Chiroptera) - contribute the greatest number of predicted novel zoonotic pathogens and known zoonoses [@han_rodent_2015; @gibb_mammal_2021].
+Zoonotic infectious disease spillover events into human populations (animal-to-human transmission) can have important health and economic consequences at both the global and regional level (e.g.
+SARS-CoV-2 and Ebola virus disease) [@judson_zoonoses_2021].
+The number of zoonotic disease spillover events and the frequency of the emergence of novel zoonotic pathogens are predicted to increase under intensifying anthropenic pressure driven by increased human populations, urbanisation, climate change and wildlife defaunation [@allen_global_2017; @hassell_urbanization_2017; @garcia-pena_land-use_2021; @mcmahon_ecosystem_2018].
+West Africa is a nexus for these processes and a hotspot for both the risk of large epidemics of known zoonotic diseases and novel zoonotic pathogen emergence [@grace_mapping_2012; @han_global_2016; @allen_global_2017].
+Two taxa - rodents (Rodentia) and bats (Chiroptera) - contribute the greatest number of predicted novel zoonotic pathogens and known zoonoses [@han_rodent_2015; @gibb_mammal_2021].
 Of 2,220 extant rodent species, 244 (10.7%) are described as reservoirs of 85 zoonotic pathogens [@han_global_2016].
-Specifically, West Africa is identified as a region at increased hazard for rodent-borne zoonotic disease spillover events under different projected future land-use change scenarios [@garcia-pena_land-use_2021].
-In West Africa rodents are involved in the transmission of multiple zoonotic diseases, including, Lassa fever, Schistosomiasis, Leptospirosis and Toxoplasmosis [@meerburg_rodent-borne_2009; @galeh_global_2020].
+Specifically, West Africa has been identified as region at increased hazard for rodent-borne zoonotic disease spillover events under different projected future land-use change scenarios [@garcia-pena_land-use_2021].
+Currently within West Africa rodents are involved in the transmission of multiple zoonotic diseases with large burdens on human health, including, Lassa fever, Schistosomiasis, Leptospirosis and Toxoplasmosis [@meerburg_rodent-borne_2009; @galeh_global_2020].
 
@@
 The role of rodent species in zoonotic disease spillover or novel zoonotic pathogens emergence are examples of ecosystem disservices.
-Rodents typically demonstrate “fast” life histories [@dobson_fast_2007] with traits such as early maturation and short gestation further associated with being zoonotic reservoirs [@han_rodent_2015; @albery_fast-lived_2021].
-Traits associated with "fast" life histories are prevalent in species that thrive in human dominated landscapes, displacing species less likely to be reservoirs of zoonotic pathogens [@gibb_zoonotic_2020].
-The widespread occurrence of reservoir species and their proximity to human activity make the description of rodent species assemblages and host-pathogen interactions vitally important to understanding the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence [@han_global_2016].
+Rodents typically demonstrate “fast” life histories [@dobson_fast_2007] with traits such as early maturation and short gestation times further associated with being zoonotic reservoirs [@han_rodent_2015; @albery_fast-lived_2021].
+Rodent species with "fast" life histories thrive in human dominated landscapes, displacing species less likely to be reservoirs of zoonotic pathogens [@gibb_zoonotic_2020].
+The widespread occurrence of reservoir species and their proximity to human activity make the description of rodent species assemblages and host-pathogen associations vitally important to understanding the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence [@han_global_2016].
 
-Despite the importance of understanding these complex systems in the hazard of zoonotic spillover, current evidence on host-pathogen associations is affected by taxonomic and geographical sampling biases [@gibb_ecosystem_2020; @gibb_mammal_2021].
-Studies identifying potential geographic hotspots of zoonotic disease spillover and novel pathogen emergence are based on global datasets of host species distributions and host-pathogen associations (e.g., IUCN, GBIF, GIDEON) that suffer from several biases [@plowright_sampling_2019; @carlson_future_2021].
+Despite the importance of understanding these complex systems, current evidence on host-pathogen associations is considerably affected by taxonomic and geographical sampling biases [@gibb_ecosystem_2020; @gibb_mammal_2021].
+Studies identifying potential geographic hotspots of zoonotic disease spillover and novel pathogen emergence are based on global datasets of host species distributions and host-pathogen associations (e.g., IUCN, GBIF, GIDEON) which can result in biased hazard estimates [@plowright_sampling_2019; @carlson_future_2021].
 For example, systematically increased sampling, over-representation of certain habitats and clustering around areas of high human population could lead to an apparent association between locations and hazard that is driven by these factors rather than underlying host-pathogen associations [@wille_how_2021; @redding_evaluating_2017; @gibb_mammal_2021].
@@
 
-Rodent trapping studies provide contextually rich information on when, where and under what conditions rodents were trapped, potentially enriching global datasets and counteracting inherent biases [@bovendorp_optimising_2017].
+Rodent trapping studies provide contextually rich information on when, where and under what conditions rodents were trapped, potentially enriching global datasets and mitigating against their inherent biases [@bovendorp_optimising_2017].
 Rodent trapping provides a useful method to describe rodent population assemblages, their geographic distribution and host-pathogen associations beyond what is currently available in global datasets.
-Studies have been conducted in West Africa to investigate the distribution of rodent speciesm, their species assemblages, the prevalence of known zoonoses within rodent hosts (e.g., Lassa fever, Schistosomiasis) and to identify novel zoonotic pathogens [@fichet-calvet_diversity_2009; @catalano_multihost_2020; @usaid_one_2021].
-However, data from these studies have not been synthesised for inclusion in assessments of zoonotic disease spillover and novel zoonotic pathogen emergence.
+Studies have been conducted in West Africa to investigate the distribution of rodent species, their species assemblages, the prevalence of known zoonoses within rodent hosts (e.g., Lassa fever, Schistosomiasis) and to identify novel zoonotic pathogens [@fichet-calvet_diversity_2009; @catalano_multihost_2020; @usaid_one_2021].
+However, data from these studies have not previously been synthesised for inclusion in assessments of zoonotic disease spillover and novel zoonotic pathogen emergence.
 
-Here, we synthesise the rodent trapping studies conducted across West African from a search of literature between 1964-2019.
+Here, we synthesise rodent trapping studies conducted across West African from a search of literature between 1964-2019.
 First, we investigate the geographic biases in the rodent trapping dataset in relation to human population density and land use classification.
 Second, we compare this to global host datasets (IUCN and GBIF) to understand differences in reported host geographic distributions.
-Third, we compare identified host-pathogen associations with global datasets (CLOVER) to understand discrepencies in rodent host-pathogen associations and report the proportion positive for each pathogen.
-Finally, within our dataset we investigate host-pathogen spatial extent to identify areas of undersampling of pathogens within host ranges.
-We expect that rodent trapping studies will provide an important additional source of high resolution data that can be used to enrich global datasets to better understand the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence across West Africa.
+Third, we compare identified host-pathogen associations with global datasets (CLOVER) to understand discrepencies in rodent host-pathogen associations and report the proportion positive for pathogens of interest.
+Finally, within our dataset we investigate the spatial extent of current host-pathogen sampling to identify areas of undersampling of pathogens within their host ranges.
+We expect that rodent trapping studies will provide an important additional source of high-resolution data that can be used to enrich global datasets to better understand the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence across West Africa.
 We find that rodent trapping studies identify greater geographic ranges of important rodent hosts of zoonotic diseases than are reported from global datasets while also identifying locations and habitats in which further sampling is required.
-We also identify host-pathogen associations that are not reported in global host-pathogen association data, however, there remains systematic undersampling for these pathogens across the host ranges.
+We also identify host-pathogen associations that are not reported in global host-pathogen association data, while identifying systematic undersampling for these pathogens across their host ranges.
 
@@
 
-We extracted data from eligible studies using a standardised tool that was piloted on 5 studies (Supplementary 1.).
-Data was abstracted into a Google Sheets document, which was archived on completion of data extraction (Supplementary 2.).
+We extracted data from eligible studies using a standardised tool that was piloted on 5 studies (Supplementary Table 1.).
+Data was abstracted into a Google Sheets document, which was archived on completion of data extraction [ref to zenodo].
 Extracted variables included i) study identifiers; ii) study aims; iii) trapping methodology; iv) geolocation data; v) method of speciation; vi) trapping locations and dates; vii) trapped species; viii) number of trap-nights and ix) microorganisms/pathogens of interest.
@@
 
-### Environmental covariates
+### Spatial covariates
 
-### Rodent species ranges
+Boundaries of countries and level-2 administrative regions were obtained from GADM 4.0.4 for the study region [@database_of_global_administrative_areas_gadm_2022].
+Land cover classification in 2005, the median year in which trapping activity occurred, was obtained from Copernicus climate change service at ~300m resolution [@european_space_agency_climate_change_initiative_land_2022].
+Proportion of land cover type within a level-2 administrative region was calculated.
+Human population density for 2005 was obtained from SEDAC at ~ 1km resolution, with median population density within each region extracted [@socioeconomic_data_and_applications_center_gridded_2021].
 
-### Rodent presence
+### Rodent species ranges and presence
+
+Rodent species distributions were obtained as shapefiles from the IUCN redlist, shapefiles were cropped to the study region [@iucn_iucn_2021].
+Rodent presence locations were obtained from GBIF as point data limited to the study region [@gbif_the_global_biodiversity_information_facility_occurrence_2021].
 
@@
 
-CLOVER a synthesis of four host-pathogen datasets (GMPD2, EID2, HP3 and Shaw 2020) was released in 2021 and represents the most comprehensive global repository of host-pathogen associations for Bacteria, Viruses, Helminth, Protozoa and Fungi [clover_paper].
-Data were obtained from an archived version of the dataset [clover_dataset].
+CLOVER a synthesis of four host-pathogen datasets (GMPD2, EID2, HP3 and Shaw 2020) was released in 2021 and represents a comprehensive global repository of host-pathogen associations for Bacteria, Viruses, Helminth, Protozoa and Fungi [@gibb_data_2021].
+Data were obtained from an archived version of the dataset [@gibb_viralemergenceclover_2021].
+The dataset was limited to species trapped within our included studies, no spatial subsetting was possible.
 
@@
 
-We identified 4,282 relevant citations, with 126 rodent trapping studies included.
+We identified 4,282 relevant citations, with 126 rodent trapping studies included (Supplementary Table 2.).
 The earliest trapping studies were conducted in 1964, with increasing numbers of studies being performed annually.
-The median year of first trapping activity was 2005, with the median length of trapping activity being 1 year (IQR 0-2 years) (Supplementary figure 1.).
+The median year of first trapping activity was 2005, with the median length of trapping activity being 1 year (IQR 0-2 years) (Supplementary Figure 1.).
 Studies were conducted in 14 West African countries with no studies reported from Gambia or Togo, at 1,331 sites (Figure 1A.).
@@
 
-```{r trap-locations-trapnights, out.width = "120%", fig.cap = "Figure 1.
-A) The location of trapping sites in West Africa.
+![Figure 1: A) The location of trapping sites in West Africa.
 No sites were recorded from Togo or Gambia.
 There is important heterogeneity observed in the coverage of each country by trap night (colour) and location of sites.
-For example Senegal, Mali and Sierra Leone have generally good coverage compared to Guinea and Burkina Faso.
-B) The distribution of trap nights performed at each study site on a log scale with a median of 244 trap nights (IQR 115-494) performed at each site"}
-
-knitr::include_graphics(here("figures", "Figure_1_updated.png"))
-
-```
+For example, Senegal, Mali and Sierra Leone have generally good coverage compared to Guinea and Burkina Faso.
+B) The distribution of trap nights performed at each study site on a log scale with a median of 244 trap nights (IQR 115-494) performed at each site](`r here("figures", "Figure_1_updated.png")`)
 
@@
 In total 73,164 small mammals were trapped with 62,574 (85%) identified to species level and 7,760 (11%) identified to genus.
-The majority of the 147 identified species were Rodentia (112, 76%) of which Muridae (82, 73%) were the most common family.
+The majority of the 140 identified species were Rodentia (110, 76%) and Soricomorpha (30, 21%), of which Muridae (82, 73%) were the most common family.
 
-62 studies tested for 32 microorganisms, defined to species or genus leve that are known or potential pathogens.
-The majority of studies tested for a single microorganism (39, 62%).
+62 studies tested for 32 microorganisms, defined to species or genus level that are known or potential pathogens.
+Most studies tested for a single microorganism (39, 62%).
 The most frequently assayed microorganisms were *Lassa mammarenavirus* or Arenaviridae (27, 43%), *Borrelia sp.* (8, 13%), *Bartonella sp.* (4, 6%) and *Toxoplasma gondii* (4, 6%).
-The majority of studies used Polymerase Chain Reaction to detect microorganisms (32, 52%), with fewer studies using molecular based tests (11, 18%) or histological or direct visualisation assays (8, 13%).
+Most studies used Polymerase Chain Reaction to detect microorganisms (32, 52%), with fewer studies using molecular based tests (11, 18%) or histological or direct visualisation assays (8, 13%).
+From 32,014 individual rodent samples we produced 339 host-pathogen pairs.
+With *Rattus rattus*, *Mus musculus*, *Mastomys erythroleucus*, *Mastomys natalensis* and *Arvicanthis niloticus* being assayed for at least 18 microorganisms.
 
-From 32,014 individual rodent samples we produced 339 host-pathogen pairs.
-With *Rattus rattus*, *Mus musculus*, *Mastomys erythroleucus*, *Mastomys natalensis* and *A.
-niloticus* being assayed for at least 18 microorganisms.
+Further descriptive information from the included studies including geolocated trapping of species, their detection and non-detection alongside microorganism data has been made available online [@simons_rodents_2022].
 
@@
 
-We used the GPS coordinates of single trap, trapping grid/line or study site and the number of trap nights to calculate trapping effort (trap night density) within level 2 administrative regions in West Africa.
-The median trap success rate for all rodents at a defined trap site was calculated separately for trap sites which included built environments and non-built environments due to important differences in trap success between these different environments.
-The number of rodents trapped at a trap site was then used in combination with these values of trap success to impute the number of trap nights for trap sites where there was no record of the number of trap nights.
-We summarised the habitat types of trap sites based on information reported in the study; the dictionary used to match habitat types is available in Supplementary 3.
+We used the GPS coordinates of trapping locations and the number of trap nights to calculate trapping effort (trap night density) within level-2 administrative regions in West Africa.
+We imputed trap nights for studies with missing records using the number of trapped rodents and median trap success rate from studies reporting trapping effort stratified by habitat type.
+We calculated the proportion of land cover classes within each of these level-2 administrative, comparing surveyed regions to non-surveyed regions.
 
-To better understand the potential bias associated with studies for zoonotic disease spillover or novel zoonotic pathogen emergence we limited further analyses to zoonotic disease studies only.
+To better understand the potential bias associated with studies for zoonotic disease spillover or novel zoonotic pathogen emergence we limited further analyses to zoonotic pathogen studies only.
 We used a Generalised Additive Model (GAM) incorporating a spatial interaction term to investigate the association between number of trap nights, proportion of urban, cropland and forest land classification and human population density [@pedersen_hierarchical_2019].
-Covariates were obtained for 2005, the median year studies were conducted, proportions of habitat types were obtained from Copernicus 2005 land cover classification, global population density estimates were obtained from SEDAC [@socioeconomic_data_and_applications_center_gridded_2021].
+The model was constructed in *mgcv* using the R statistical programme [@wood_generalized_2017; @r_core_team_r_2020].
+Selection of the most parsimonious model was based on the adjusted-R^2^ and Deviance explained for each model.
 We performed sensitivity analysis by removing sites with imputed trapping effort.
-The model structure was specified as:
+The final selected model structure was specified as:
 
@@
 \tag{1}
-\text{TN density} \sim \text{Tweedie(log(Population density) + (Urban) + (Cropland) + (Shrubland) + (Forest) + (Longitude * Latitude))}
+\text{TN density} \sim \text{Tweedie((Urban) + (Forest) + (Longitude * Latitude))}
 \end{equation}
@@
 
-The model was used to identify locations of high and low trapping effort relative to population density and land use classification. Predictions were limited to regions around observed trap sites.
+The model identified locations of high and low trapping effort relative to land use classification, with proportion urban land very highly correlated with human population density (*r* = 0.835, 95% C.I.
+0.82-0.85, *p* < 0.001).
+Model inference was limited to regions around observed trap sites.
 
@@
 
-We assessed the taxonomic coverage of the trapped taxa and for the seven species (*M.
+We assessed the taxonomic coverage of the trapped taxa and for the seven species with the most trap locations (*M.
 natalensis*, *R.
@@
 erythroleucus*, *M.
-musculus*, *Arvicanthis niloticus*, *Praomys daltoni* and *Cricetomys gambianus*) with the most trap locations we mapped the detection and non-detection of rodent species and compared this to detection data obtained from GBIF and species ranges from IUCN [@gbif_the_global_biodiversity_information_facility_occurrence_2021; @iucn_iucn_2021].
+musculus*, *A.
+niloticus*, *Praomys daltoni* and *Cricetomys gambianus*) we mapped the detection and non-detection of rodent species and compared this to detection data obtained from GBIF and species ranges from IUCN [@gbif_the_global_biodiversity_information_facility_occurrence_2021; @iucn_iucn_2021].
 We identified rodent locations from GBIF and the rodent trapping studies within and outside the IUCN expected range.
-We located these to level 2 administrative regions to calculate the discrepencies between observed detections and non-detections with IUCN distributions.
+We located these to level-2 administrative regions to calculate the discrepencies between observed detections and non-detections with IUCN distributions.
 
@@
 For studies investigating rodent zoonoses we produced host-pathogen pairs.
-For each host-pathogen pair we compared our records with those available in CLOVER reporting concordance and discordance.
-We identify host-pathogen pairs that are not present in CLOVER and contain assay results consistent with acute or prior infection.
+For each host-pathogen pair we compared our records with those available in CLOVER, reporting concordance and discordance.
+For host-pathogen pairs with assay results consistent with acute or prior infection we calculate the proportion positive and identify those absent from CLOVER.
+We expand the analysis to host-pathogen pairs with pathogens identified to genus level in Supplementary Figure 4.
 
@@
 
-For the four pathogens sampled at the most locations (Borreliaceae, Leptospiraceae, *Lassa mammarenavirus* and *Toxoplasma gondii*) we identify their host-pathogen associations.
-We identified the five species with most spatially complete sampling and associated these sampling locations with level 2 administrative regions.
-We compared the area of these regions to their IUCN range and detection range to calculate sampling coverage.
+For the four pathogens sampled at the most locations (Borreliaceae, Leptospiraceae, *Lassa mammarenavirus* and *Toxoplasma gondii*) we identify their potential host species.
+For the five species with most spatially complete sampling we associate their sampling locations with level-2 administrative regions and compare this to their IUCN range and rodent trapping detection range to calculate a measure sampling coverage.
 
@@
 
-Further descriptive information from the included studies including geolocated trapping of species, their detection and non-detection alongside microorganism data has been made available online [@simons_rodents_2022].
-
 ## What is the extent of spatial bias in the rodent trapping data?
 
-Trap sites were situated in 273 (19.3%) of 1414 level 2 administrative regions in 14 West African nations.
+Trap sites were situated in 273 of 1,414 (19.3%) level-2 administrative regions in 14 West African nations.
 The areas with highest trap night density included the capital and large cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako), Ghana (Accra),  Côte d'Ivoire (Abidjan), and Benin (Cotonou).
@@
 Trapping sites were biased towards human modified landscapes (i.e., cropland, grassland and mosaic habitats), with under-representation of forest and bare habitats.
-However, urban habitats were undersampled.
+However, urban habitats were comparitively undersampled (Supplementary Figure 2.).
 
-Trapping effort, measured through trap nights, was biased towards areas with higher population density and predominantly urban and cropland land classifications (Figure 2.).
-In particular Northern Senegal, Eastern Sierra Leone, Eastern Guinea and South West Nigeria had relatively high trapping efforts while South East Nigeria, Northern Nigeria, Liberia, Côte d'Ivoire, Ghana, Niger, and Burkina Faso were trapped at lower trapping effort than would be expected.
-Excluding sites with imputed trap nights reduced the number of trapping sites contributing data.
-Despite this limitation, Mauritania, Northern Senegal and Sierra Leone remained as regions trapped at higher than expected rates with Nigeria as being trapped at lower than expected rates (Supplementary Figure 2.).
+Trapping effort, measured through trap nights, was biased with increased effort in areas with higher proportion of urban land use and no-significant association with proportion of forest cover.
+The spatial term shows trapping bias towards increased effort in Northern Senegal, Eastern Sierra Leone, Eastern Guinea and coastal regions of Nigeria, Benin and Ghana, in contrast South-East Nigeria, Northern Nigeria and Liberia had relatively lower trapping effort (Figure 2).
+The most parsimonious GAM model (adjusted R^2^ = 0.523, Deviance explained = 61%) included terms for coordinates, proportion urban coverage and proportion tree coverage, model selection is shown in Supplementary Tables 3.1-3.4.
+In sensitivity analysis excluding sites with imputed trap nights reduced contributing data, however, Mauritania, Northern Senegal and Sierra Leone remained as regions trapped at higher rates with Nigeria being trapped at lower than expected rates (Supplementary Figure 3.).
 
-```{r gam-trapping-effort, out.width = "120%", fig.cap = "Figure 2.
-Relative trapping effort bias across West Africa from included studies adjusted for human population density, proportion cropland cover, proportion forest cover and proportion urban habitat.
+![Figure 2.
+Relative trapping effort bias across West Africa from included studies adjusted for proportion urban land cover and proportion tree cover.
 Predictions are limited to areas around trap sites (coloured areas), uncertainty in the estimate of bias is represented by colour transparency/pallor.
-Purple regions represent areas with higher than expected trapping effort, yellow regions represent areas lower than expected trapping effort"}
-
-knitr::include_graphics(here("figures", "Figure_2_updated.png"))
-
-```
+Purple regions represent areas with higher than expected trapping effort, yellow regions represent areas lower than expected trapping effort](`r here("figures", "Figure_2_updated.png")`)
 
@@
 For our species of interest, except *C.
-gambianus*, trapping studies provided more locations of detection and non-detection than were available from GBIF.
+gambianus*, trapping studies provided more distinct locations of detection and non-detection than were available from GBIF.
 For the West African endemic species IUCN ranges showed generally good concordance to both trapping studies and GBIF, although the range of *A.
@@
 
-```{r comparing-trap-locations-A, out.width = "120%", fig.cap = "Figure 3A.
+![Figure 3A.
 Each row corresponds to a single rodent species.
-L) Presence recorded in GBIF overlaid on IUCN species range.
+L) Presence recorded in GBIF overlaid on IUCN species range (red-shaded area).
 R) Detection and non-detection from rodent trapping studies overlaid on IUCN species ranges.
-No IUCN range for *M.
-musculus* is available for the study region."}
+M.
+musculus has no range in West African in IUCN data.](`r here("figures", "Figure_3_test_a.png")`)
 
-knitr::include_graphics(here("figures", "Figure_3_test_a.png"))
+![Figure 3B.
+Each row corresponds to a single rodent species.
+L) Presence recorded in GBIF overlaid on IUCN species range (red-shaded area).
+R) Detection and non-detection from rodent trapping studies overlaid on IUCN species ranges.](`r here("figures", "Figure_3_test_b.png")`)
 
-```
-
-```{r comparing-trap-locations-B, out.width = "120%", fig.cap = "Figure 3B.
-Each row corresponds to a single rodent species.
-L) Presence recorded in GBIF overlaid on IUCN species range.
-R) Detection and non-detection from rodent trapping studies overlaid on IUCN species ranges."}
-
-knitr::include_graphics(here("figures", "Figure_3_test_b.png"))
-
-```
-
-We identified the proportion of a rodents expected range in which detections and non-detections occurred (Table 1.).
-GBIF sampling occurred over > 15% of a rodents expected range with detections occuring outside of IUCN ranges for all species, most noticeably for *A.
+We identified the proportion of a rodent's expected range in which detections and non-detections occurred (Table 1.).
+GBIF sampling occurred over > 15% of a rodent's expected range with detections occurring outside of IUCN ranges for all species, most noticeably for *A.
 niloticus* and *P.
@@
 niloticus* and *M.
-erythroleucus* not being detected in >15% of their expected range.
+erythroleucus* not being detected in >15% of their expected range, suggestive of substantial heterogeneity fo presence across their range.
 
@@
 
-table_1 <- read_rds(here("tables", "Table_1_updated.rds"))
+table_1 <- read_rds(here("tables", "Table_1_updated.rds")) %>%
+  set_caption(caption = "Table 1: Comparison of IUCN, GBIF and rodent trapping ranges for the 7 species trapped at the most sites.") %>%
+  autofit()
 
@@
 When limited to confirmed pathogens and classification of both pathogen and host to species level we identified 21 host-pathogen pairs among 11 rodent and 6 pathogen species (Figure 4.).
-We identified negative associations for 41 host-pathogen pairs among 31 rodent and 7 pathogen species.
+We identified negative associations, through entirely negative assays, for 41 host-pathogen pairs among 31 rodent and 7 pathogen species.
 CLOVER contained 8 (38%) of our identified host-pathogen associations and positive associations for 6 (15%) of the negative associations.
 
-```{r comparing-host-pathogen-interactions, out.width = "120%", fig.cap = "Figure 4.
+![Figure 4.
 L) Identified species level host-pathogen associations through detection of acute infection (i.e.
@@
 Percentages and colour relate to the proportion of all assays that were positive.
-Associations with a black border are present in the CLOVER dataset."}
-
-knitr::include_graphics(here("figures", "Figure_4_updated.png"))
-
-```
+Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Figure_4_updated.png")`)
 
@@
 rattus* and *R.
-norvegicus*), or from wide ranging distributions in sub-Saharan Africa (31, 9%) (i.e., *A.
+norvegicus*), or from those with wide ranging distributions in sub-Saharan Africa (31, 9%) (i.e., *A.
 niloticus*, *M.
-natalensis* and *A.
-albiventris*).
+natalensis* and *Atelerix albiventris*).
 
 For pathogens not identified to species level (i.e.
-family or lower taxa) we identified 128 host-pathogen pairs among 32 rodent species and 25 pathogen families (Supplementary figure 3.), with CLOVER containing 52 (41%) of these associations.
+family or higher taxa only) we identified 128 host-pathogen pairs among 32 rodent species and 25 pathogen families (Supplementary Figure 4.), with CLOVER containing 52 (41%) of these associations.
 
@@
 
-The four most widely sampled pathogen species/families were were Borreliaceae, *Lassa mammarenavirus*, Leptospiraceae and *Toxoplasma gondii* (Table 2.).
+The four most widely sampled pathogen species/families were Borreliaceae, *Lassa mammarenavirus*, Leptospiraceae and *Toxoplasma gondii* (Table 2.).
 Infection with species of Borreliaceae was assessed in 42 species, with evidence of infection in 17 rodent species.
@@
 erythroleucus* (8%).
-The five most widely assayed species were sampled from a relative small area of their species ranges (7.3-16%), however, there was good representation when compared to where rodent trapping studies have occurred (24-85%).
+The five most widely assayed species were sampled from a relatively small area of their species ranges (7.3-16%), however, there was good representation when compared to where rodent trapping studies have occurred (24-85%).
+
 Infection with *Lassa mammarenavirus* was assessed in 28 species, with evidence of infection in 7 rodent species.
@@
 Species sampling occurred from limited areas within IUCN ranges (2-3%), however, sampling was more representative for rodent trapping studies (3-24%).
+
 Infection with species of Leptospiraceae was assessed in 8 species, with evidence of infection in 5 rodent species.
@@
 Species sampling occurred across less than 1% of these species IUCN ranges and for all except *R.
-norvegicus* for less than 1% of their detection range in included studies.
+norvegicus* for less than 1% of their detection range in rodnet trapping studies.
+
 *Toxoplasma gondii* infection was assessed in 8 species, with evidence of infection in 6 rodent species.
@@
 
+<!---BLOCK_LANDSCAPE_START--->
+
 ```{r table-2}
 
-table_2 <- read_rds(here("tables", "Table_2_updated.rds"))
+table_2 <- read_rds(here("tables", "Table_2_updated.rds"))%>%
+  set_caption(caption = "Table 2: Comparison of pathogen sampling ranges for the 4 most widely sampled pathogens and the 5 most sampled potential rodent host species.
+* no IUCN range in West African") %>%
+  autofit()
 
@@
 
+<!---BLOCK_LANDSCAPE_STOP--->
+
 # Discussion
@@
 We synthesised data from 126 rodent trapping studies containing information on more than 72,000 trapped rodents across 1,331 trapping sites producing an estimated 966,000 trap-nights.
-Trapping occurred across 14 West African countries, studies investigating known or potential zoonotic pathogens occurred within 19% of level 2 administrative regions across West Africa with important bias towards areas with high population density and human dominated landscapes.
-For several important rodent species we identified discordance between IUCN species ranges, GBIF presence data and detections and non-detections from rodent trapping studies, particularly among non-native species (*R.
+Trapping occurred across 14 West African countries, studies investigating known or potential zoonotic pathogens occurred within 19% of level-2 administrative regions across West Africa with important bias towards areas with high population density and human dominated landscapes.
+For several important rodent species, we identified discordance between IUCN species ranges, GBIF presence data and detections and non-detections from rodent trapping studies, particularly among non-native species (*R.
 rattus* and *M.
@@
 Trapping sites are clustered in areas with relatively high human population density and anthropogenic land use which can importantly bias our understanding of the distribution of rodent species and their pathogens across West Africa.
-Further effort is required to minimise this bias by sampling currently undersampled regions.
+Further effort is required to counteract this bias by sampling currently undersampled regions.
 
@@
 We show here that non-detection data from rodent trapping studies can provide additional information to describe the heterogeneity of rodent assemblages across diverse habitats in West Africa.
-Combining presence data from GBIF and detection/non-detection data from rodent trapping studies can substantially increase coverage across a rodents range, allowing improved estimates of rodent host prevalence for zoonotic infectious disease studies.
+Combining presence data from GBIF and detection/non-detection data from rodent trapping studies can substantially increase coverage across a rodent's range, allowing improved estimates of rodent host prevalence for zoonotic infectious disease studies.
 
 Host-pathogen association datasets are becoming increasing comprehensive alongside increased access to molecular discovery techniques for potential pathogens.
-We show that currently available data from rodent trapping studies can supplement these resources to better understand the potential hosts and reservoirs of known zoonotic pathogens.
-Rodent trapping data can support studies using these global datasets to predict novel pathogen emergence and to quantify the risk of known zoonotic disease spillover based on host distributions.
+We show that currently available data from rodent trapping studies can enrich these resources, particularly with prevalence data and geolocation to better understand the potential hosts and reservoirs of known zoonotic pathogens.
+Most of the host-pathogen associations from CLOVER not identified from rodent trapping studies occurred in species with ranges that go beyond our study area, suggesting that these pathogens do not exist in West Africa.
+Rodent trapping data can support analyses using these global datasets to predict novel pathogen emergence and to quantify the risk of known zoonotic disease spillover based on host distributions.
 
@@
 Lassa fever is known to be endemic in eight West African countries, although it is potentially more widely spread.
-Further studies are required to investigate this pathogen across a wider geographic range.
+Further rodent trapping studies are required to investigate this pathogen across a wider geographic range.
 Other important zoonotic diseases including Leptospirosis and Toxoplasmosis have been assessed over very limited regions of their hosts distribution.
+
+Due to data sparsity, we were unable to account for temporal change over the six decades of rodent trapping studies.
+Land use change and population density have changed dramatically over this period in West Africa.
+We attempted to mitigate this by using the median year of trapping to understand the spatial biases in trapping activity, however, it is possible that land use and population density at trapping sites varied importantly between when the study was conducted and the situation in 2005.
+Despite this limitation, the finding that trapping is biased towards high density, human dominated landscapes is unlikely to importantly change.
 
@@
 They can reduce some of the bias in global datasets however, important limitations remain.
-Further effort is required to survey rodent populations in non-human dominated landscapes and to assay pathogens across more of a hosts range.
+Further effort is required to survey rodent populations in non-human dominated landscapes and to assay pathogens across more of a host's range.
+
+# References
+
+<div id="refs"></div>
 
 # Supplementary
+<!---BLOCK_LANDSCAPE_START--->
+## Supplementary Table 1
 
-## Supplementary figure 3
+```{r supplementary-table-1}
 
-```{r comparing-host-pathogen-interactions-supplementary, out.width = "120%", fig.cap = "Figure 3.
+read_rds(here("tables", "supplementary_table_1.rds")) %>%
+  set_caption(caption = "Supplementary Table 1: Data extraction tool for studies meeting inclusion criteria")
+
+```
+
+## Supplementary Table 2
+
+```{r supplementary-table-2}
+
+read_rds(here("tables", "supplementary_table_2.rds"))
+
+```
+<!---BLOCK_LANDSCAPE_STOP--->
+## Supplementary Table 3
+
+```{r supplementary-table-3-1}
+
+read_rds(here("tables", "supplementary_table_3_1.rds"))
+
+```
+
+```{r supplementary-table-3-2}
+
+read_rds(here("tables", "supplementary_table_3_2.rds"))
+
+```
+
+```{r supplementary-table-3-3}
+
+read_rds(here("tables", "supplementary_table_3_3.rds"))
+
+```
+
+```{r supplementary-table-3-4}
+
+read_rds(here("tables", "supplementary_table_3_4.rds"))
+
+```
+
+## Supplementary Figure 1
+
+![Supplementary Figure 1.
+Green points represent the start date of rodent trapping studies, blue points representing the final trapping activity.
+Red points indicate the publication of studies.
+Increasing numbers of studies have been published since 2000 with more studies being conducted over repeated visits.](`r here("figures", "Supplementary_figure_1.png")`)
+
+## Supplementary Figure 2
+
+![Supplementary Figure 2.
+Comparison between proportion of land use between regions trapped and all land use in West Africa.
+Cropland, Shrubland and Grassland were trapped at than representative rates while Bare habitats, Forested regions and Urban landscapes were trapped at lower than representative rates.](`r here("figures", "Supplementary_figure_2.png")`)
+
+## Supplementary Figure 3
+
+![Supplementary Figure 3.
+Relative trapping effort bias across West Africa from the subset of included studies reporting trapping effort adjusted for proportion urban land cover and proportion tree cover.
+Predictions are limited to areas around trap sites (coloured areas), uncertainty in the estimate of bias is represented by colour transparency/pallor.
+Purple regions represent areas with higher than expected trapping effort, yellow regions represent areas lower than expected trapping effort](`r here("figures", "Supplementary_figure_3.png")`)
+
+## Supplementary Figure 4
+
+![Supplementary Figure 4.
 L) Identified host-pathogen associations at pathogen family level through detection of acute infection (i.e.
@@
 Percentages and colour relate to the proportion of all assays that were positive.
-Associations with a black border are present in the CLOVER dataset."}
-
-knitr::include_graphics(here("figures", "Supplementary_figure_3_updated.png"))
-
-```
+Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Supplementary_figure_4_updated.png")`)

### 2022-05-11 scripts/analysis.R

@@ -237,6 +237,13 @@ rodent_data %>%
   group_by(unique_id, trap_night_data) %>%
   summarise(total_trap_nights = sum(trap_nights))
 
+# The total number of trap nights
+
+rodent_data %>%
+  distinct(unique_id, year_trapping, month_trapping, town_village, habitat, trap_nights, .keep_all = T) %>%
+  group_by(trap_night_data) %>%
+  summarise(tn = sum(trap_nights, na.rm = TRUE))
+
 ggplot(sense_check %>%
          arrange(-total_trap_nights)) +
   geom_histogram(aes(x = total_trap_nights, fill = trap_night_data)) +
@@ -418,18 +425,29 @@ wa_countries <- c("BEN", "BFA", "CIV", "CPV", "ESH", "GHA",
 genus_data <- read_rds(here("data_clean", "genus_hierarchy.rds"))
 species_data <- read_rds(here("data_clean", "species_data.rds"))
 
+total_number <- sum(rodent_data$number)
+
 count_species <- species_data %>%
   filter(iso3c %in% wa_countries) %>%
-  group_by(species_gbif, classification) %>%
+  group_by(species_gbif) %>%
   drop_na(species_gbif) %>%
   summarise(number = sum(number)) %>%
   mutate(percent = round(number/sum(.$number)*100, 2)) %>%
   arrange(-percent) %>%
+  left_join(., species %>%
+              drop_na(gbif_id) %>%
+              filter(!str_detect(classification, "/")),
+            by = c("species_gbif" = "gbif_id")) %>%
   rename(`GBIF ID` = "species_gbif",
          "Classification" = "classification",
          "Number of individuals" = "number",
          "Percent (%)" = "percent") %>%
+  filter(`Number of individuals` > 0) %>%
+  distinct(`GBIF ID`, `Number of individuals`, .keep_all = T) %>%
   mutate(Classification = snakecase::to_sentence_case(Classification)) # the number of individuals trapped identified to species level
+
+total_species_level <- sum(count_species$`Number of individuals`)
+total_genus_level <- sum(species_data$number[is.na(species_data$species_gbif) & !str_detect(species_data$classification, "rodent*")])
 
 speciation <- count_species %>%
   mutate(`GBIF ID` = as.character(`GBIF ID`)) %>%
@@ -445,7 +463,8 @@ speciation <- count_species %>%
 speciation %>%
   group_by(order, family) %>%
   count() %>%
[…]

### 2022-05-11 scripts/rodent_distributions_pixel.R

@@ -149,20 +149,20 @@ analysis_proportion_pixel <- function(species_name, trap_data = rodent_spatial, 
     area_gbif_outside_iucn <- mask(gbif_r, iucn_v, inverse = TRUE) %>%
       expanse(unit = "km")
 
-    prop_gbif_coverage <- (mask(gbif_r, iucn_v) %>%
-                             expanse(unit = "km")/expanse(iucn_r, unit = "km")) * 100
+    prop_gbif_coverage <- round((mask(gbif_r, iucn_v) %>%
+                             expanse(unit = "km")/expanse(iucn_r, unit = "km")) * 100, 2)
 
     area_trapping_detection_in_iucn <- mask(trap_r_detection, iucn_v) %>%
       expanse(unit = "km")
 
-    prop_detection_coverage <- (mask(trap_r_detection, iucn_v) %>%
-                                  expanse(unit = "km")/expanse(iucn_r, unit = "km")) * 100
+    prop_detection_coverage <- round((mask(trap_r_detection, iucn_v) %>%
+                                  expanse(unit = "km")/expanse(iucn_r, unit = "km")) * 100, 2)
 
     area_trapping_non_detection_in_iucn <- mask(trap_r_non_detection, iucn_v) %>%
       expanse(unit = "km")
 
-    prop_non_detection_coverage <- (mask(trap_r_non_detection, iucn_v) %>%
-                                      expanse(unit = "km")/expanse(iucn_r, unit = "km")) * 100
+    prop_non_detection_coverage <- round((mask(trap_r_non_detection, iucn_v) %>%
+                                      expanse(unit = "km")/expanse(iucn_r, unit = "km")) * 100, 2)
 
     area_detection_outside_iucn <- mask(trap_r_detection, iucn_v, inverse = TRUE) %>%
       expanse(unit = "km")
@@ -170,14 +170,16 @@ analysis_proportion_pixel <- function(species_name, trap_data = rodent_spatial, 
     area_combined_in_iucn <- mask(combined_r, iucn_v) %>%
       expanse(unit = "km")
 
-    prop_combined_coverage <- (mask(combined_r, iucn_v) %>%
-                                 expanse(unit = "km")/expanse(iucn_r, unit = "km")) * 100
+    prop_combined_coverage <- round((mask(combined_r, iucn_v) %>%
+                                 expanse(unit = "km")/expanse(iucn_r, unit = "km")) * 100, 2)
 
[…]

### 2022-05-11 scripts/tn_gam.R

@@ -32,11 +32,11 @@ if(!file.exists(here("data_clean", "pop_tn_analysis.rds"))) {
     filter(trap_night_data != "Imputed") %>%
     distinct(unique_id, year_trapping, month_trapping, country, region, town_village, habitat, trap_nights, trap_night_data, geometry)
 
-  sites_2 <- st_intersection(x = level_2, y = trapping_effort)
+  sites_in_2 <-  st_join(x = trapping_effort, y = level_2, join = st_within)
 
-  sites_2_sensitivity <- st_intersection(x = level_2, y = trapping_effort_sensitivity)
+  sites_in_2_sensitivity <- st_join(x = trapping_effort_sensitivity, y = level_2, join = st_within)
 
-  trap_nights_region <- tibble(sites_2) %>%
+  trap_nights_region <- tibble(sites_in_2) %>%
     group_by(NAME_2, GID_2, unique_id, year_trapping, month_trapping, region, town_village, habitat) %>%
     summarise(total_trap_nights = sum(trap_nights)) %>%
     group_by(GID_2, NAME_2) %>%
@@ -52,16 +52,16 @@ if(!file.exists(here("data_clean", "pop_tn_analysis.rds"))) {
     left_join(., trap_nights_region, by = c("GID_2", "NAME_2")) %>%
     mutate(region_trap_nights = case_when(is.na(region_trap_nights) ~ 0,
                                           TRUE ~ region_trap_nights))  %>%
-    mutate(area_m2 = st_area(.),
-           tn_density = region_trap_nights/(as.numeric(area_m2)/1000000),
+    mutate(area_km2 = set_units(st_area(.), "km^2"),
+           tn_density = region_trap_nights/as.numeric(area_km2),
            tn_density = ifelse(is.na(tn_density), NA, tn_density))
 
   sites_2_sensitivity <- level_2 %>%
     left_join(., trap_nights_region_sensitivity, by = c("GID_2", "NAME_2")) %>%
     mutate(region_trap_nights = case_when(is.na(region_trap_nights) ~ 0,
                                           TRUE ~ region_trap_nights))  %>%
-    mutate(area_m2 = st_area(.),
-           tn_density = region_trap_nights/(as.numeric(area_m2)/1000000),
+    mutate(area_km2 = set_units(st_area(.), "km^2"),
[…]

### 2022-05-11 scripts/tn_gam_pixel.R

@@ -43,7 +43,7 @@ if(!file.exists(here("data_clean", "pop_tn_analysis.rds"))) {
   trapping_effort <- rodent_spatial %>%
     filter(unique_id %in% zoonotic_studies) %>%
     distinct(unique_id, year_trapping, month_trapping, trap_nights, geometry) %>%
-    drop_na(geometry) %>%
+    drop_na(geometry, trap_nights) %>%
     mutate(x = st_coordinates(geometry)[,1],
            y = st_coordinates(geometry)[,2]) %>%
     group_by(x, y) %>%
@@ -56,7 +56,7 @@ if(!file.exists(here("data_clean", "pop_tn_analysis.rds"))) {
     crop(contiguous_boundary) %>%
     mask(contiguous_boundary)
 
-  wa_pop <- log10(wa_pop$pop_2005)
+  wa_pop <- wa_pop$pop_2005
 
   source(here("scripts", "land_use_classification.R"))
 
@@ -84,11 +84,12 @@ if(!file.exists(here("data_clean", "pop_tn_analysis.rds"))) {
                            hab == 8 ~ "bare",
                            hab == 9 ~ "water")),
            tn_density = trap_nights/area,
+           tn = trap_nights,
            x = st_coordinates(geometry)[,1],
            y = st_coordinates(geometry)[,2]) %>%
-    select(tn_density, pop, hab, x, y)
+    select(tn, tn_density, pop, hab, x, y)
 
-  pixel_non_trapped <- tibble(spatSample(resample_hab, size = nrow(pixel_trapping), method = "random", na.rm = TRUE, cells = TRUE, xy = TRUE)) %>%
+  pixel_non_trapped <- tibble(spatSample(resample_hab, size = nrow(pixel_trapping)*5, method = "stratified", na.rm = TRUE, cells = TRUE, xy = TRUE)) %>%
     mutate(hab = factor(case_when(lccs_class == 1 ~ "agriculture",
                                   lccs_class == 2 ~ "forest",
                                   lccs_class == 3 ~ "grassland",
@@ -98,27 +99,36 @@ if(!file.exists(here("data_clean", "pop_tn_analysis.rds"))) {
                                   lccs_class == 7 ~ "urban",
                                   lccs_class == 8 ~ "bare",
                                   lccs_class == 9 ~ "water")),
-           tn_density = 0) %>%
-    select(tn_density, hab, x, y)
[…]

### 2022-05-19 reports/manuscript_v4.Rmd
For moving

@@
 
-table_1 <- read_rds(here("tables", "Table_1_updated.rds")) %>%
+table_1 <- read_rds(here("tables", "Table_1_updated_pixel.rds")) %>%
   set_caption(caption = "Table 1: Comparison of IUCN, GBIF and rodent trapping ranges for the 7 species trapped at the most sites.") %>%

### 2022-05-19 scripts/3_0_descriptive.R

@@ -1,0 +1,402 @@
+source(here::here("scripts", "libraries.R"))
+
+# Data --------------------------------------------------------------------
+# Load study details
+studies <- read_rds(here("data_clean", "studies.rds")) %>%
+  mutate(year_publication = as.numeric(year_publication))
+
+# Load in the cleaned habitat types of each study
+habitat_split <- c("habitat_1", "habitat_2", "habitat_3", "habitat_4", "habitat_5", "habitat_6", "habitat_7")
+habitat_data <- read_rds(here("data_clean", "habitat_types.rds")) %>%
+  tibble() %>%
+  dplyr::select(record_id, all_of(habitat_split), intensity_use)
+
+# Load the number of trap nights for studies not presenting this data
+# These values have been previously imputed
+imputed_tn <- read_rds(here("data_clean", "imputed_trap_nights.rds"))
+
+# Load the cleaned rodent data, each line represents a species within a trap site within a study
+# Join this data to the cleaned habitats and imputed trap night data
+rodent_data <- read_rds(here("data_clean", "rodent_df.rds")) %>%
+  tibble()  %>%
+  dplyr::select(-all_of(c("intensity_use", "geometry", "trap_nights"))) %>%
+  left_join(.,
+            habitat_data ,
+            by = "record_id") %>%
+  left_join(., imputed_tn,
+            by = c("unique_id", "year_trapping", "month_trapping",
+                   "region", "town_village", "habitat")) %>%
+  mutate(unique_id = as_factor(unique_id))
+
+# Load in the cleaned pathogen data in both wide and long formats
+pathogen <- read_rds(here("data_clean", "pathogen.rds"))
+wide_pathogen <- read_rds(here("data_clean", "wide_pathogen.rds"))
+long_pathogen <- read_rds(here("data_clean", "long_pathogen.rds"))
+
+# Load in the cleaned species and gbif associations
+# This includes missnamed species and those with uncertainty
+species_gbif <- read_rds(here("data_clean", "species_data.rds")) %>%
+  distinct(genus, species, gbif_id, genus_gbif, species_gbif)
+
+# This is the list of ISO3 codes for the countries of West Africa
+wa_countries <- c("BEN", "BFA", "CIV", "CPV", "ESH", "GHA",
+                  "GIN", "GMB", "GNB", "LBR", "MLI", "MRT",
+                  "NER", "NGA", "SEN", "SLE", "TGO")
+
+# Summary table of included studies ---------------------------------------
+# This is not presented in the manuscript but is included here for interest
[…]

### 2022-05-19 scripts/3_1_descriptive.R

@@ -1,402 +1,0 @@
-source(here::here("scripts", "libraries.R"))
-
-# Data --------------------------------------------------------------------
-# Load study details
-studies <- read_rds(here("data_clean", "studies.rds")) %>%
-  mutate(year_publication = as.numeric(year_publication))
-
-# Load in the cleaned habitat types of each study
-habitat_split <- c("habitat_1", "habitat_2", "habitat_3", "habitat_4", "habitat_5", "habitat_6", "habitat_7")
-habitat_data <- read_rds(here("data_clean", "habitat_types.rds")) %>%
-  tibble() %>%
-  dplyr::select(record_id, all_of(habitat_split), intensity_use)
-
-# Load the number of trap nights for studies not presenting this data
-# These values have been previously imputed
-imputed_tn <- read_rds(here("data_clean", "imputed_trap_nights.rds"))
-
-# Load the cleaned rodent data, each line represents a species within a trap site within a study
-# Join this data to the cleaned habitats and imputed trap night data
-rodent_data <- read_rds(here("data_clean", "rodent_df.rds")) %>%
-  tibble()  %>%
-  dplyr::select(-all_of(c("intensity_use", "geometry", "trap_nights"))) %>%
-  left_join(.,
-            habitat_data ,
-            by = "record_id") %>%
-  left_join(., imputed_tn,
-            by = c("unique_id", "year_trapping", "month_trapping",
-                   "region", "town_village", "habitat")) %>%
-  mutate(unique_id = as_factor(unique_id))
-
-# Load in the cleaned pathogen data in both wide and long formats
-pathogen <- read_rds(here("data_clean", "pathogen.rds"))
-wide_pathogen <- read_rds(here("data_clean", "wide_pathogen.rds"))
-long_pathogen <- read_rds(here("data_clean", "long_pathogen.rds"))
-
-# Load in the cleaned species and gbif associations
-# This includes missnamed species and those with uncertainty
-species_gbif <- read_rds(here("data_clean", "species_data.rds")) %>%
-  distinct(genus, species, gbif_id, genus_gbif, species_gbif)
-
-# This is the list of ISO3 codes for the countries of West Africa
-wa_countries <- c("BEN", "BFA", "CIV", "CPV", "ESH", "GHA",
-                  "GIN", "GMB", "GNB", "LBR", "MLI", "MRT",
-                  "NER", "NGA", "SEN", "SLE", "TGO")
-
-# Summary table of included studies ---------------------------------------
-# This is not presented in the manuscript but is included here for interest
[…]

### 2022-05-19 scripts/3_1_spatial_bias_analysis.R

@@ -1,0 +1,401 @@
+source(here::here("scripts", "libraries.R"))
+
+if(!file.exists(here("models", "all_models.rds"))) {
+
+  habitat_split <- c("habitat_1", "habitat_2", "habitat_3", "habitat_4", "habitat_5", "habitat_6", "habitat_7")
+  habitat_data <- read_rds(here("data_clean", "habitat_types.rds")) %>%
+    tibble() %>%
+    dplyr::select(record_id, all_of(habitat_split), intensity_use)
+
+  imputed_tn <- read_rds(here("data_clean", "imputed_trap_nights.rds"))
+
+  level_1 <- read_rds(here("data_download", "admin_spatial", "level_1_admin.rds"))
+  level_2 <- read_rds(here("data_download", "admin_spatial", "level_2_admin.rds"))
+  non_trapped <- read_rds(here("data_download", "admin_spatial", "level_2_TGOGMB.rds"))
+
+  level_2 <- bind_rows(level_2, non_trapped) %>%
+    filter(!str_detect(GID_0, "CPV"))
+
+  # Trap night and population -----------------------------------------------
+
+  # Load the cleaned rodent data, each line represents a species within a trap site within a study
+  # Join this data to the cleaned habitats and imputed trap night data
+  rodent_data <- read_rds(here("data_clean", "rodent_df.rds")) %>%
+    tibble()  %>%
+    dplyr::select(-all_of(c("intensity_use", "geometry", "trap_nights"))) %>%
+    left_join(.,
+              habitat_data ,
+              by = "record_id") %>%
+    left_join(., imputed_tn,
+              by = c("unique_id", "year_trapping", "month_trapping",
+                     "region", "town_village", "habitat")) %>%
+    mutate(unique_id = as_factor(unique_id)) %>%
+    drop_na(longitude, latitude) %>%
+    st_as_sf(coords = c("longitude", "latitude"), crs = st_crs(level_2))
+
+  trapping_effort <- rodent_data %>%
+    distinct(unique_id, year_trapping, month_trapping, country, region, town_village, habitat, trap_nights, trap_night_data, geometry)
+
+  trapping_effort_sensitivity <- rodent_data %>%
+    filter(trap_night_data != "Imputed") %>%
+    distinct(unique_id, year_trapping, month_trapping, country, region, town_village, habitat, trap_nights, trap_night_data, geometry)
+
+  # Associated trap night density with administrative regions
[…]

### 2022-05-19 scripts/3_2_rodent_host_distributions.R

@@ -1,0 +1,377 @@
+# Data --------------------------------------------------------------------
+
+all_countries <- c("BEN", "BFA", "CIV", "CMR", "CPV", "DZA", "ESH", "GHA",
+                   "GIN", "GMB", "GNB", "LBR", "MAR", "MLI", "MRT", "NER",
+                   "NGA", "SEN", "SLE", "TCD", "TGO")
+wa_countries <- c("BEN", "BFA", "CIV", "CPV", "ESH", "GHA",
+                  "GIN", "GMB", "GNB", "LBR", "MLI", "MRT",
+                  "NER", "NGA", "SEN", "SLE", "TGO")
+continental_countries <- c("BEN", "BFA", "CIV", "ESH", "GHA",
+                           "GIN", "GMB", "GNB", "LBR", "MLI", "MRT",
+                           "NER", "NGA", "SEN", "SLE", "TGO")
+no_data_countries <- c("GMB", "TGO")
+
+level_0 <- read_rds(here("data_download", "admin_spatial", "level_0_admin.rds"))
+
+level_1 <- read_rds(here("data_download", "admin_spatial", "level_1_admin.rds"))
+
+level_2 <- read_rds(here("data_download", "admin_spatial", "level_2_admin.rds"))
+
+non_trapped <- read_rds(here("data_download", "admin_spatial", "level_2_TGOGMB.rds"))
+
+included_countries <- level_0 %>%
+  filter(GID_0 %in% continental_countries)
+
+contiguous_boundary <- included_countries %>%
+  filter(!GID_0 == "CPV") %>%
+  summarise()
+
+write_rds(contiguous_boundary, here("data_clean", "WA_continental_boundary.rds"))
+
+studies <- read_rds(here("data_clean", "studies.rds"))
+
+rodent_spatial <- read_rds(here("data_clean", "rodent_spatial.rds")) %>%
+  dplyr::select(-trap_nights)
+
+imputed_tn <- read_rds(here("data_clean", "imputed_trap_nights.rds"))
+
+rodent_spatial <- rodent_spatial %>%
+  left_join(., imputed_tn,
+            by = c("unique_id", "year_trapping", "month_trapping",
+                   "region", "town_village", "habitat"))
+
+bbox_rodent <- st_bbox(rodent_spatial)
+
+# IUCN data
+# As some of these ranges cross the dateline we use S1 methods
+sf::sf_use_s2(FALSE)
+
+
[…]

### 2022-05-19 scripts/3_2_spatial_bias_analysis.R

@@ -1,401 +1,0 @@
-source(here::here("scripts", "libraries.R"))
-
-if(!file.exists(here("models", "all_models.rds"))) {
-
-  habitat_split <- c("habitat_1", "habitat_2", "habitat_3", "habitat_4", "habitat_5", "habitat_6", "habitat_7")
-  habitat_data <- read_rds(here("data_clean", "habitat_types.rds")) %>%
-    tibble() %>%
-    dplyr::select(record_id, all_of(habitat_split), intensity_use)
-
-  imputed_tn <- read_rds(here("data_clean", "imputed_trap_nights.rds"))
-
-  level_1 <- read_rds(here("data_download", "admin_spatial", "level_1_admin.rds"))
-  level_2 <- read_rds(here("data_download", "admin_spatial", "level_2_admin.rds"))
-  non_trapped <- read_rds(here("data_download", "admin_spatial", "level_2_TGOGMB.rds"))
-
-  level_2 <- bind_rows(level_2, non_trapped) %>%
-    filter(!str_detect(GID_0, "CPV"))
-
-  # Trap night and population -----------------------------------------------
-
-  # Load the cleaned rodent data, each line represents a species within a trap site within a study
-  # Join this data to the cleaned habitats and imputed trap night data
-  rodent_data <- read_rds(here("data_clean", "rodent_df.rds")) %>%
-    tibble()  %>%
-    dplyr::select(-all_of(c("intensity_use", "geometry", "trap_nights"))) %>%
-    left_join(.,
-              habitat_data ,
-              by = "record_id") %>%
-    left_join(., imputed_tn,
-              by = c("unique_id", "year_trapping", "month_trapping",
-                     "region", "town_village", "habitat")) %>%
-    mutate(unique_id = as_factor(unique_id)) %>%
-    drop_na(longitude, latitude) %>%
-    st_as_sf(coords = c("longitude", "latitude"), crs = st_crs(level_2))
-
-  trapping_effort <- rodent_data %>%
-    distinct(unique_id, year_trapping, month_trapping, country, region, town_village, habitat, trap_nights, trap_night_data, geometry)
-
-  trapping_effort_sensitivity <- rodent_data %>%
-    filter(trap_night_data != "Imputed") %>%
-    distinct(unique_id, year_trapping, month_trapping, country, region, town_village, habitat, trap_nights, trap_night_data, geometry)
-
-  # Associated trap night density with administrative regions
[…]

### 2022-05-19 scripts/3_3_host_pathogen_associations.R

@@ -1,0 +1,332 @@
+# Load data ---------------------------------------------------------------
+
+rodent_spatial <- read_rds(here("data_clean", "rodent_spatial.rds")) %>%
+  dplyr::select(-trap_nights)
+
+species_data <- read_rds(here("data_clean", "species_data.rds")) %>%
+  filter(!is.na(species_gbif)) %>%
+  distinct(classification, species_gbif)
+
+tested_pcr <- tibble(read_rds(here("data_clean", "long_pathogen.rds"))) %>%
+  select(-any_of(contains("habitat"))) %>%
+  select(-geometry) %>%
+  pivot_wider(names_from = assay, values_from = number) %>%
+  filter(!is.na(pcr_path_1_positive) | !is.na(pcr_path_2_positive) | !is.na(pcr_path_3_positive) |
+           !is.na(pcr_path_4_positive) | !is.na(pcr_path_5_positive) | !is.na(pcr_path_6_positive)) %>%
+  select(-any_of(c(contains("ab_ag"), contains("culture"), contains("histo")))) %>%
+  mutate(record_id = row_number(),
+         assay = "PCR") %>%
+  group_by(record_id) %>%
+  mutate(tested = case_when(!is.na(pcr_path_1_positive) ~ path_1_tested,
+                                !is.na(pcr_path_2_positive) ~ path_2_tested,
+                                !is.na(pcr_path_3_positive) ~ path_3_tested,
+                                !is.na(pcr_path_4_positive) ~ path_4_tested,
+                                !is.na(pcr_path_5_positive) ~ path_5_tested,
+                                !is.na(pcr_path_6_positive) ~ path_6_tested),
+         positive = case_when(!is.na(path_1_tested) ~ pcr_path_1_positive,
+                                  !is.na(path_2_tested) ~ pcr_path_2_positive,
+                                  !is.na(path_3_tested) ~ pcr_path_3_positive,
+                                  !is.na(path_4_tested) ~ pcr_path_4_positive,
+                                  !is.na(path_5_tested) ~ pcr_path_5_positive,
+                                  !is.na(path_6_tested) ~ pcr_path_6_positive)) %>%
+  select(-any_of(contains("path_")))
+
+tested_serology <- tibble(read_rds(here("data_clean", "long_pathogen.rds"))) %>%
+  select(-any_of(contains("habitat"))) %>%
+  select(-geometry) %>%
[…]

### 2022-05-19 scripts/3_4_host_pathogen_associations.R

@@ -1,332 +1,0 @@
-# Load data ---------------------------------------------------------------
-
-rodent_spatial <- read_rds(here("data_clean", "rodent_spatial.rds")) %>%
-  dplyr::select(-trap_nights)
-
-species_data <- read_rds(here("data_clean", "species_data.rds")) %>%
-  filter(!is.na(species_gbif)) %>%
-  distinct(classification, species_gbif)
-
-tested_pcr <- tibble(read_rds(here("data_clean", "long_pathogen.rds"))) %>%
-  select(-any_of(contains("habitat"))) %>%
-  select(-geometry) %>%
-  pivot_wider(names_from = assay, values_from = number) %>%
-  filter(!is.na(pcr_path_1_positive) | !is.na(pcr_path_2_positive) | !is.na(pcr_path_3_positive) |
-           !is.na(pcr_path_4_positive) | !is.na(pcr_path_5_positive) | !is.na(pcr_path_6_positive)) %>%
-  select(-any_of(c(contains("ab_ag"), contains("culture"), contains("histo")))) %>%
-  mutate(record_id = row_number(),
-         assay = "PCR") %>%
-  group_by(record_id) %>%
-  mutate(tested = case_when(!is.na(pcr_path_1_positive) ~ path_1_tested,
-                                !is.na(pcr_path_2_positive) ~ path_2_tested,
-                                !is.na(pcr_path_3_positive) ~ path_3_tested,
-                                !is.na(pcr_path_4_positive) ~ path_4_tested,
-                                !is.na(pcr_path_5_positive) ~ path_5_tested,
-                                !is.na(pcr_path_6_positive) ~ path_6_tested),
-         positive = case_when(!is.na(path_1_tested) ~ pcr_path_1_positive,
-                                  !is.na(path_2_tested) ~ pcr_path_2_positive,
-                                  !is.na(path_3_tested) ~ pcr_path_3_positive,
-                                  !is.na(path_4_tested) ~ pcr_path_4_positive,
-                                  !is.na(path_5_tested) ~ pcr_path_5_positive,
-                                  !is.na(path_6_tested) ~ pcr_path_6_positive)) %>%
-  select(-any_of(contains("path_")))
-
-tested_serology <- tibble(read_rds(here("data_clean", "long_pathogen.rds"))) %>%
-  select(-any_of(contains("habitat"))) %>%
-  select(-geometry) %>%
[…]

### 2022-05-19 scripts/3_4_host_pathogen_distributions.R

@@ -1,0 +1,297 @@
+# Load data ---------------------------------------------------------------
+
+contiguous_boundary <- read_rds(here("data_clean", "WA_continental_boundary.rds"))
+
+rodent_spatial <- read_rds(here("data_clean", "rodent_spatial.rds")) %>%
+  dplyr::select(-trap_nights)
+
+tested_pcr <- tibble(read_rds(here("data_clean", "long_pathogen.rds"))) %>%
+  select(-any_of(contains("habitat"))) %>%
+  drop_na(geometry) %>%
+  filter(!geometry %in% "c(NA, NA)") %>%
+  pivot_wider(names_from = assay, values_from = number) %>%
+  filter(!is.na(pcr_path_1_positive) | !is.na(pcr_path_2_positive) | !is.na(pcr_path_3_positive) |
+           !is.na(pcr_path_4_positive) | !is.na(pcr_path_5_positive) | !is.na(pcr_path_6_positive)) %>%
+  select(-any_of(c(contains("ab_ag"), contains("culture"), contains("histo")))) %>%
+  mutate(record_id = row_number(),
+         assay = "PCR") %>%
+  group_by(record_id) %>% # use the following to coalesce row-wise
+  mutate(tested = case_when(!is.na(pcr_path_1_positive) ~ path_1_tested,
+                            !is.na(pcr_path_2_positive) ~ path_2_tested,
+                            !is.na(pcr_path_3_positive) ~ path_3_tested,
+                            !is.na(pcr_path_4_positive) ~ path_4_tested,
+                            !is.na(pcr_path_5_positive) ~ path_5_tested,
+                            !is.na(pcr_path_6_positive) ~ path_6_tested),
+         positive = case_when(!is.na(path_1_tested) ~ pcr_path_1_positive,
+                              !is.na(path_2_tested) ~ pcr_path_2_positive,
+                              !is.na(path_3_tested) ~ pcr_path_3_positive,
+                              !is.na(path_4_tested) ~ pcr_path_4_positive,
+                              !is.na(path_5_tested) ~ pcr_path_5_positive,
+                              !is.na(path_6_tested) ~ pcr_path_6_positive)) %>%
+  select(-any_of(contains("path_"))) %>%
+  filter(tested > 0) %>%
+  st_as_sf(crs = crs(rodent_spatial))
+
[…]

### 2022-05-24 reports/manuscript_v4.Rmd
update for v4

@@
 
-Rodents are important reservoirs of known zoonotic infectious diseases and novel zoonotic pathogens.
-Rodent reservoirs are globally distributed with West Africa containing multiple host species of zoonotic infectious diseases.
-Ongoing anthropogenic land use change alters the composition of rodent species assemblages which can alter the hazard of zoonotic disease spillover and emergence.
-Current global datasets to understand these hazards suffer from biases.
-Here, we synthesise data from 127 rodent trapping studies including over 72,000 individual rodent detections to understand the range and presence of important rodent reservoir species to quantify the bias in global datasets.
-We identify the rodent trapping studies are biased towards human dominated landscapes with important sampling biases across West Africa.
-We find that these rodent trapping studies can complement current global datasets on rodent population distributions which can minimise some of these observed biases.
-For four regionally important zoonotic infectious diseases we identify host-pathogen associations not included in a comprehensive global dataset.
-Rodent trapping studies can further enrich these resources through geolocating these associations and producing measures of pathogen prevalence within hosts.
-Finally, we find that for most of these diseases the proportion of a rodent hosts range that has been sampled remains small.
-Further rodent trapping studies are required to understand the true hazard of zoonotic spillover of these diseases across West Africa.
-Our results highlight that understanding rodent sampling bias is important when assessing the risk for zoonotic spillover events from rodents.
-The synthesis of contextually rich rodent trapping data contributes important information that is lacking in IUCN distribution maps and GBIF species presence data.
-We anticipate that this dataset can support the production of more complete understanding of the hazard of zoonotic spillover events.
+NEEDS UPDATING
 
@@
 
-Zoonotic infectious disease spillover events into human populations (animal-to-human transmission) can have important health and economic consequences at both the global and regional level (e.g.
-SARS-CoV-2 and Ebola virus disease) [@judson_zoonoses_2021].
-The number of zoonotic disease spillover events and the frequency of the emergence of novel zoonotic pathogens are predicted to increase under intensifying anthropenic pressure driven by increased human populations, urbanisation, climate change and wildlife defaunation [@allen_global_2017; @hassell_urbanization_2017; @garcia-pena_land-use_2021; @mcmahon_ecosystem_2018].
-West Africa is a nexus for these processes and a hotspot for both the risk of large epidemics of known zoonotic diseases and novel zoonotic pathogen emergence [@grace_mapping_2012; @han_global_2016; @allen_global_2017].
-Two taxa - rodents (Rodentia) and bats (Chiroptera) - contribute the greatest number of predicted novel zoonotic pathogens and known zoonoses [@han_rodent_2015; @gibb_mammal_2021].
-Of 2,220 extant rodent species, 244 (10.7%) are described as reservoirs of 85 zoonotic pathogens [@han_global_2016].
-Specifically, West Africa has been identified as region at increased hazard for rodent-borne zoonotic disease spillover events under different projected future land-use change scenarios [@garcia-pena_land-use_2021].
-Currently within West Africa rodents are involved in the transmission of multiple zoonotic diseases with large burdens on human health, including, Lassa fever, Schistosomiasis, Leptospirosis and Toxoplasmosis [@meerburg_rodent-borne_2009; @galeh_global_2020].
+There is increasing awareness of the global health and economic impacts of novel zoonotic pathogen spillover, evidenced by the ongoing SARS-CoV-2 pandemic and previous HIV/AIDs and Spanish Influenza pandemics [@Bernstein].
+The number of zoonotic disease spillover events and the frequency of the emergence of novel zoonotic pathogens are predicted to increase under intensifying anthropogenic pressure driven by increased human populations, urbanisation, intensification of agriculture, climate change and wildlife defaunation [@allen_global_2017; @hassell_urbanization_2017; @garcia-pena_land-use_2021; @mcmahon_ecosystem_2018].
+The impact of endemic zoonoses meanwhile remains underestimated [@Maudlin].
+Endemic zoonoses disproportionally affect those in the poorest sections of society, those living in close contact with their animals and those with limited access to healthcare [@Halliday; @Molyneux; @judson_zoonoses_2021].
+West Africa is a nexus for these processes and is identified as a hotspot for increased risk of large epidemics of known zoonotic diseases and novel zoonotic pathogen emergence [@allen_global_2017; @grace_mapping_2012; @han_global_2016].
+Two taxonomic groups - rodents (Rodentia) and bats (Chiroptera) - contribute the greatest number of predicted novel zoonotic pathogens and known zoonoses [@han_rodent_2015; @gibb_mammal_2021].
+Of 2,220 extant rodent species, 244 (10.7%) are described as reservoirs of 85 zoonotic pathogens [@han_rodent_2015].
+Specifically, West Africa has been identified as a region at increased hazard for rodent-borne zoonotic disease spillover events under different projected future land-use change scenarios [@garcia-pena_land-use_2021].
+Currently within West Africa, rodents are involved in the transmission of multiple endemic zoonoses with large burdens on human health, these pathogens include, Lassa fever, Schistosomiasis, Leptospirosis and Toxoplasmosis [@meerburg_rodent-borne_2009; @galeh_global_2020].
 
@@
 The role of rodent species in zoonotic disease spillover or novel zoonotic pathogens emergence are examples of ecosystem disservices.
-Rodents typically demonstrate “fast” life histories [@dobson_fast_2007] with traits such as early maturation and short gestation times further associated with being zoonotic reservoirs [@han_rodent_2015; @albery_fast-lived_2021].
-Rodent species with "fast" life histories thrive in human dominated landscapes, displacing species less likely to be reservoirs of zoonotic pathogens [@gibb_zoonotic_2020].
+Rodents typically demonstrate “fast” life histories [@dobson_fast_2007] with traits such as early maturation and short gestation times (<4 days) further associated with being zoonotic reservoirs [@han_rodent_2015; @albery_fast-lived_2021].
+Rodent species with “fast” life histories thrive in human dominated landscapes, displacing species less likely to be reservoirs of zoonotic pathogens [@gibb_zoonotic_2020].
 The widespread occurrence of reservoir species and their proximity to human activity make the description of rodent species assemblages and host-pathogen associations vitally important to understanding the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence [@han_global_2016].
@@
 Despite the importance of understanding these complex systems, current evidence on host-pathogen associations is considerably affected by taxonomic and geographical sampling biases [@gibb_ecosystem_2020; @gibb_mammal_2021].
-Studies identifying potential geographic hotspots of zoonotic disease spillover and novel pathogen emergence are based on global datasets of host species distributions and host-pathogen associations (e.g., IUCN, GBIF, GIDEON) which can result in biased hazard estimates [@plowright_sampling_2019; @carlson_future_2021].
+Curated biodiversity datasets such as the Global Biodiversity Information Facility (GBIF) and resources produced by the International Union for Conservation of Nature suffer from well described sampling biases [@Bowler; @Boakes].
+These sampling biases can gravely distort produced species distribution models [@Beck].
+Datasets on host-pathogen associations also suffer from biases introduced from literature selection criteria and taxanomic discrepancies a recently produced consolidated dataset (CLOVER) has been produced to reduce this [@Gibb_clover].
+Studies identifying potential geographic hotspots of zoonotic disease spillover and novel pathogen emergence are typically based on these datasources for modelling host species distributions and host-pathogen associations which can result in biased hazard estimates [@plowright_sampling_2019; @carlson_future_2021].
 For example, systematically increased sampling, over-representation of certain habitats and clustering around areas of high human population could lead to an apparent association between locations and hazard that is driven by these factors rather than underlying host-pathogen associations [@wille_how_2021; @redding_evaluating_2017; @gibb_mammal_2021].
-While other regions remain systematically undersampled (e.g., areas of sparse human populations), resulting in a reduced hazard of zoonotic spillover events or novel zoonotic pathogen emergence being attributed [@beck_spatial_2014].
+While other regions remain systematically undersampled (e.g., areas of sparse human populations, endemic zoonoses), potentially resulting in a reduced hazard of zoonotic spillover events or novel zoonotic pathogen emergence being attributed [@beck_spatial_2014].
 Predictions of zoonotic disease spillover and novel zoonotic pathogen emergence must account for these biases to understand the future hazard of zoonotic diseases [@carlson_future_2021].
 
-Rodent trapping studies provide contextually rich information on when, where and under what conditions rodents were trapped, potentially enriching global datasets and mitigating against their inherent biases [@bovendorp_optimising_2017].
-Rodent trapping provides a useful method to describe rodent population assemblages, their geographic distribution and host-pathogen associations beyond what is currently available in global datasets.
-Studies have been conducted in West Africa to investigate the distribution of rodent species, their species assemblages, the prevalence of known zoonoses within rodent hosts (e.g., Lassa fever, Schistosomiasis) and to identify novel zoonotic pathogens [@fichet-calvet_diversity_2009; @catalano_multihost_2020; @usaid_one_2021].
-However, data from these studies have not previously been synthesised for inclusion in assessments of zoonotic disease spillover and novel zoonotic pathogen emergence.
+Rodent trapping studies provide contextually rich information on when, where and under what conditions rodents were trapped, potentially enriching consolidated datasets and providing additional data to minimise inherent biases [@bovendorp_optimising_2017].
+Rodent trapping studies describe rodent population assemblages, their geographic distribution and host-pathogen associations beyond what is currently available in consolidated datasets.
+Studies have been conducted in West Africa to investigate the distribution of rodent species, their species assemblages, the prevalence of endemic zoonoses within rodent hosts (e.g., Lassa fever, Schistosomiasis) and to identify emerging and novel zoonotic pathogens [@fichet-calvet_diversity_2009; @catalano_multihost_2020; @usaid_one_2021].
+However, individual level data from these studies have not previously been synthesised for inclusion in assessments of zoonotic disease spillover and novel zoonotic pathogen emergence.
 
-Here, we synthesise rodent trapping studies conducted across West African from a search of literature between 1964-2019.
-First, we investigate the geographic biases in the rodent trapping dataset in relation to human population density and land use classification.
-Second, we compare this to global host datasets (IUCN and GBIF) to understand differences in reported host geographic distributions.
-Third, we compare identified host-pathogen associations with global datasets (CLOVER) to understand discrepencies in rodent host-pathogen associations and report the proportion positive for pathogens of interest.
-Finally, within our dataset we investigate the spatial extent of current host-pathogen sampling to identify areas of undersampling of pathogens within their host ranges.
-We expect that rodent trapping studies will provide an important additional source of high-resolution data that can be used to enrich global datasets to better understand the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence across West Africa.
-We find that rodent trapping studies identify greater geographic ranges of important rodent hosts of zoonotic diseases than are reported from global datasets while also identifying locations and habitats in which further sampling is required.
-We also identify host-pathogen associations that are not reported in global host-pathogen association data, while identifying systematic undersampling for these pathogens across their host ranges.
+Here, we synthesise rodent trapping studies conducted across West Africa from a search of literature between 1964-2022.
+First, we use this dataset to investigate the geographic sampling biases in relation to human population density and land use classification.
+Second, we compare this to curated host datasets (IUCN and GBIF) to understand differences in reported host geographic distributions.
+Third, we compare identified host-pathogen associations with a consolidated dataset (CLOVER) to explore discrepancies in rodent host-pathogen associations and report the proportion of positive assays for pathogens of interest.
+Finally, within our dataset we investigate the spatial extent of current host-pathogen sampling to identify areas of sparse sampling of pathogens within their host ranges.
+We expect that rodent trapping studies provide an important additional source of high-resolution data that can be used to enrich available consolidated datasets to better understand the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence across West Africa.
 
@@
 
-We conducted a search in Ovid MEDLINE, Web of Science (Core collection and Zoological Record), JSTOR, BioOne, African Journals Online, Global Health and the pre-print servers, BioRxiv and EcoEvoRxiv for the following terms as keywords, no date limits were set: (1) Rodent OR Rodent trap* AND (2) West Africa (or the individual countries).
+To identify relevant literature, we conducted a search in Ovid MEDLINE, Web of Science (Core collection and Zoological Record), JSTOR, BioOne, African Journals Online, Global Health and the pre-print servers, BioRxiv and EcoEvoRxiv for the following terms as exploded keywords: (1) Rodent OR Rodent trap* AND (2) West Africa, no date limits were set.
+We also searched other resources including the UN Official Documents System, Open Grey, AGRIS FAO and Google Scholar using combinations of the above terms.
+Searches were run on 2022-05-01, and returned studies conducted between 1964-2021.
 
-We searched other resources including the UN Official Documents System, Open Grey, AGRIS FAO and Google Scholar using combinations of the above terms.
-Searches were run on 2021-03-01, returning studies conducted between 1964-2019.
-
-We included studies if they met all of the following inclusion criteria; i) reported findings from trapping studies where the target was a small mammal, ii) described the type of trap used or the length of trapping activity or the location of the trapping activity, iii) included trapping activity from at least one West African country, iv) recorded the genus or species of trapped individuals, v) were published in a peer-reviewed journal or as a pre-print on a digital platform or as a report by a credible organisation.
+For further analysis, we included studies if they met all of the following inclusion criteria; i) reported findings from trapping studies where the target was a small mammal, ii) described the type of trap used or the length of trapping activity or the location of the trapping activity, iii) included trapping activity from at least one West African country, iv) recorded the genus or species of trapped individuals, and v) were published in a peer-reviewed journal or as a pre-print on a digital platform or as a report by a credible organisation.
 We excluded studies if they met any of the following exclusion criteria: i) reported data that were duplicated from a previously included study, ii) no full text available, iii) not available in English.
-One reviewer screened titles, abstracts and full texts against the inclusion and exclusion criteria.
-At each stage, a random subset (10%) was reviewed by a second reviewer.
+One author screened titles, abstracts and full texts against the inclusion and exclusion criteria.
+At each stage; title screening, abstract screening and full text review, a random subset (10%) was reviewed by a second author.
 
 We extracted data from eligible studies using a standardised tool that was piloted on 5 studies (Supplementary Table 1.).
-Data was abstracted into a Google Sheets document, which was archived on completion of data extraction [ref to zenodo].
-Extracted variables included i) study identifiers; ii) study aims; iii) trapping methodology; iv) geolocation data; v) method of speciation; vi) trapping locations and dates; vii) trapped species; viii) number of trap-nights and ix) microorganisms/pathogens of interest.
-We summarised the number of studies, the year in which trapping occurred and the country in which they were conducted.
+Data was abstracted into a Google Sheets document, which was archived on completion of data extraction [ref to zenodo on finalised].
+We identified the aims of included studies, for example, whether it was conducted as a survey of small mammal species or specifically to assess the risk of zoonotic disease spillover.
+To understand potential limitations in the representativeness of rodent trapping we extracted data on study methodology, such as, the number of trap nights, the type of traps used and whether the study attempted to estimate abundance.
+For studies not reporting number of trap nights we used imputation based on the number of trapped individuals, stratified by the habitat type from which they were obtained.
+This was performed by multiplying the total number of trapped individuals within that study site by the median trap success for study sites with the same reported habitat type.
+Stratification was used as trap success varied importantly between traps placed in or around buildings (13%, IQR 6-24%) compared with other habitats (3%, IQR 1-9%)
 
-We extracted GPS locations for the most precise location presented, converting to decimal degrees as required.
-For studies not using standardised habitat recording (e.g., IUCN Habitat Classification Scheme (Version 3.1)), the explicit description of the habitat in which the trap was placed was extracted.
+The method of taxonomic identification applied within a study was recorded, reported species identification was assumed to be accurate.
+The number of individuals of these species or genera was extracted with species and genus names mapped to GBIF taxonomy [@gbif_the_global_biodiversity_information_facility_global_2021] for internal consistency.
+We expanded species detection and non-detection records by explicitly specifying non-detection at a trap site if a species was recorded as detected at other trapping locations within the same study.
+
+Geographic locations of trapping studies was extracted using GPS locations for the most precise location presented.
+GPS coordinates for locations not reported were produced using the National Geospatial-Intelligence Agency GEOnet Names Server [@GeoNames] based on place-names and maps presented in the study.
+All locations were converted to decimal degrees for internal consistency.
+The year of rodent trapping was extracted alongside the length of the trapping activity to understand seasonal representativeness of trapping activity.
+The habitats of trapping sites were mapped to the IUCN Habitat Classification Scheme (Version 3.1).
 For studies reporting multiple habitat types for a single trap, trap-line or trapping grid, a higher order classification of habitat type was recorded.
 
-We mapped genus and species names to those in the Global Biodiversity Information Facility (GBIF) taxonomy  [@gbif_the_global_biodiversity_information_facility_global_2021].
-We extracted information on the detection, non-detection and number of trapped individuals.
+For relevant studies we extracted data on all microorganisms and zoonotic pathogens tested and the method used (e.g., molecular or serological diagnosis).
+Where assays were able to identify the microorganism to species level this was recorded, non-specific assays higher order attribution was used (e.g.
+to family level).
+We recorded the species of rodent host tested, the number of individuals tested and the number of positive and negative results.
+For studies reporting summary results all testing data were extracted, this may introduce double counting of individual rodents, for example, if a single rodent was tested using both molecular and serological assays.
+Where studies reported indeterminate results, these were also recorded.
 
-We extracted data on all pathogens assayed.
-The number of rodents tested and the number of positive or negative samples were recorded alongside the type of assay used.
-If studies reported indeterminate results this was noted.
-Pathogens were identified to species level, however, where an assay only allowed for attribution to a family of viruses or bacteria, the higher order grouping was used.
+We summarised the number of studies, the year in which trapping occurred and the country in which they were conducted.
+We identified 4,692 relevant citations, with 127 rodent trapping studies included (Supplementary Table 2.).
+The earliest trapping studies were conducted in 1964, with a trend of increasing numbers of studies being performed annually since 2000.
+The median year of first trapping activity was 2007, with the median length of trapping activity being 1 year (IQR 0-2 years) (Supplementary Figure 1.).
+Studies were conducted in 14 West African countries, with no studies reported from The Gambia or Togo, at 1,611 trap sites (Figure 1A.).
 
-### Spatial covariates
+Included studies explicitly reported on 601,184 trap nights, a further 341,445 trap nights were imputed from studies with no recording of trapping effort based on trap success.
+A total of 942,629 trap nights were estimated (Figure 1B.).
+A minority of studies trapped at a single study site (30, 24%), with 46 (36%) trapping at between two and five sites, the remaining 51 studies (40%) trapped at between six and 93 study sites.
 
-Boundaries of countries and level-2 administrative regions were obtained from GADM 4.0.4 for the study region [@database_of_global_administrative_areas_gadm_2022].
-Land cover classification in 2005, the median year in which trapping activity occurred, was obtained from Copernicus climate change service at ~300m resolution [@european_space_agency_climate_change_initiative_land_2022].
-Proportion of land cover type within a level-2 administrative region was calculated.
-Human population density for 2005 was obtained from SEDAC at ~ 1km resolution, with median population density within each region extracted [@socioeconomic_data_and_applications_center_gridded_2021].
-
-### Rodent species ranges and presence
-
-Rodent species distributions were obtained as shapefiles from the IUCN redlist, shapefiles were cropped to the study region [@iucn_iucn_2021].
-Rodent presence locations were obtained from GBIF as point data limited to the study region [@gbif_the_global_biodiversity_information_facility_occurrence_2021].
-
-### Host pathogen associations
-
-CLOVER a synthesis of four host-pathogen datasets (GMPD2, EID2, HP3 and Shaw 2020) was released in 2021 and represents a comprehensive global repository of host-pathogen associations for Bacteria, Viruses, Helminth, Protozoa and Fungi [@gibb_data_2021].
-Data were obtained from an archived version of the dataset [@gibb_viralemergenceclover_2021].
-The dataset was limited to species trapped within our included studies, no spatial subsetting was possible.
-
-## Included studies
-
-We identified 4,282 relevant citations, with 126 rodent trapping studies included (Supplementary Table 2.).
-The earliest trapping studies were conducted in 1964, with increasing numbers of studies being performed annually.
-The median year of first trapping activity was 2005, with the median length of trapping activity being 1 year (IQR 0-2 years) (Supplementary Figure 1.).
-Studies were conducted in 14 West African countries with no studies reported from Gambia or Togo, at 1,331 sites (Figure 1A.).
-A minority (31, 25%) of studies trapped at a single study site, with 46 (37%) trapping at between two and five sites, the remaining 49 studies (38%) trapped at between six and 93 study sites.
-There were 581,426 reported trap nights, with 384,983 trap nights imputed for studies with no recording of trapping effort (Figure 1B.).
+In total 76,275 small mammals were trapped with 65,628 (90%) identified to species level and 7,439 (10%)  identified to genus, with the remaining classified to higher taxonomic level.
+The majority of the 132 identified species were Rodentia (102, 78%), of which Muridae (73, 72%) were the most common family.
+Soricomorpha were the second most identified order of small mammals (28, 21%).
+57 studies tested for 32 microorganisms, defined to species or genus level that are known or potential pathogens.
+Most studies tested for a single microorganism (48, 84%).
+The most frequently assayed microorganisms were *Lassa mammarenavirus* or Arenaviridae (21, 37%), *Borrelia sp.* (9, 16%), *Bartonella sp.* (4, 7%) and *Toxoplasma gondii* (4, 7%).
+Most studies used Polymerase Chain Reaction (PCR) to detect microorganisms (37, 65%), with fewer studies using serology-based tests (11, 19%) or histological or direct visualisation assays (11, 21%).
+From 32,920 individual rodent samples we produced 351 host-pathogen pairs.
+With *Rattus rattus*, *Mus musculus*, *Mastomys erythroleucus*, *Mastomys natalensis* and *Arvicanthis niloticus* being assayed for at least 18 microorganisms.
+Further descriptive information from the included studies including geolocated trapping of species, their detection and non-detection alongside microorganism data is available as supplementary material [@simons_rodents_2022].
 
 ![Figure 1: A) The location of trapping sites in West Africa.
-No sites were recorded from Togo or Gambia.
-There is important heterogeneity observed in the coverage of each country by trap night (colour) and location of sites.
+No sites were recorded from Togo or The Gambia.
+Heterogeneity is observed in the coverage of each country by trap night (colour) and location of sites.
 For example, Senegal, Mali and Sierra Leone have generally good coverage compared to Guinea and Burkina Faso.
-B) The distribution of trap nights performed at each study site on a log scale with a median of 244 trap nights (IQR 115-494) performed at each site](`r here("figures", "Figure_1_updated.png")`)
-
-Within the included studies trap success varied importantly between traps placed in or around buildings (13%, IQR 6-24%) compared to other habitats (3%, IQR 1-9%).
-In total 73,164 small mammals were trapped with 62,574 (85%) identified to species level and 7,760 (11%) identified to genus.
-The majority of the 140 identified species were Rodentia (110, 76%) and Soricomorpha (30, 21%), of which Muridae (82, 73%) were the most common family.
-
-62 studies tested for 32 microorganisms, defined to species or genus level that are known or potential pathogens.
-Most studies tested for a single microorganism (39, 62%).
-The most frequently assayed microorganisms were *Lassa mammarenavirus* or Arenaviridae (27, 43%), *Borrelia sp.* (8, 13%), *Bartonella sp.* (4, 6%) and *Toxoplasma gondii* (4, 6%).
-Most studies used Polymerase Chain Reaction to detect microorganisms (32, 52%), with fewer studies using molecular based tests (11, 18%) or histological or direct visualisation assays (8, 13%).
-From 32,014 individual rodent samples we produced 339 host-pathogen pairs.
-With *Rattus rattus*, *Mus musculus*, *Mastomys erythroleucus*, *Mastomys natalensis* and *Arvicanthis niloticus* being assayed for at least 18 microorganisms.
-
-Further descriptive information from the included studies including geolocated trapping of species, their detection and non-detection alongside microorganism data has been made available online [@simons_rodents_2022].
+B) Histogram of trap nights performed at each study site, a median of 248 trap nights (IQR 116-500) was performed at each site](`r here("figures", "Figure_1_updated.png")`)
 
@@
 
-We used the GPS coordinates of trapping locations and the number of trap nights to calculate trapping effort (trap night density) within level-2 administrative regions in West Africa.
-We imputed trap nights for studies with missing records using the number of trapped rodents and median trap success rate from studies reporting trapping effort stratified by habitat type.
-We calculated the proportion of land cover classes within each of these level-2 administrative, comparing surveyed regions to non-surveyed regions.
+To investigate the risk to human health from endemic zoonoses and novel pathogen emergence using synthesised data from rodent trapping data understanding of sampling biases is vital.
+Sampling of rodents has not been performed randomly across space in West Africa (Figure 1.) and at varying intensity, we aimed quantify this bias as the relationship between trapping effort (measured as trap night density (TN density)), human population density and land use type.
+We define this as relative trapping effort bias.
 
-To better understand the potential bias associated with studies for zoonotic disease spillover or novel zoonotic pathogen emergence we limited further analyses to zoonotic pathogen studies only.
-We used a Generalised Additive Model (GAM) incorporating a spatial interaction term to investigate the association between number of trap nights, proportion of urban, cropland and forest land classification and human population density [@pedersen_hierarchical_2019].
-The model was constructed in *mgcv* using the R statistical programme [@wood_generalized_2017; @r_core_team_r_2020].
-Selection of the most parsimonious model was based on the adjusted-R^2^ and Deviance explained for each model.
-We performed sensitivity analysis by removing sites with imputed trapping effort.
-The final selected model structure was specified as:
+To address this question, we calculated TN density within each West African level-2 administrative region, administrative boundaries were obtained from GADM 4.0.4 [@Database of Global Administrative Areas].
+We assigned trap nights to specific regions using GPS coordinates obtained from trapping studies as described above.
+The sf package in the R statistical language was used to handle shapefiles and geographic data [@Pebesma].
+Density was calculated by dividing the number of trap nights by the area of a level-2 administrative area.
+For studies not reporting trap nights imputation was used as previously described.
+Human population density was obtained for the closest year (2005) to the median year of trapping (2007) from Socioeconomic Data and Applications Center (SEDAC) gridded population of the world v4 at ~ 1km resolution [@SEDAC].
+Median population density was then calculated for each level-2 administrative region.
+Land cover classification was obtained from the Copernicus climate change service at ~300m resolution [@COPERNICUS].
+The proportion of cropland, shrubland, tree cover and urban land cover within a level-2 administrative region in 2005 was calculated.
+
+We investigated the role of these covariates on the spatial bias of TN density by comparing to a spatial model (equation 1), as follows:
 
@@
 \tag{1}
-\text{TN density} \sim \text{Tweedie((Urban) + (Forest) + (Longitude * Latitude))}
+\text{TN density} \sim \text{Tweedie((Longitude * Latitude))}
+\end{equation}
+
+\begin{equation}
+\tag{2}
+\text{TN density} \sim \text{Tweedie((Population density) + (Longitude * Latitude))}
+\end{equation}
+
+\begin{equation}
+\tag{3}
+\text{TN density} \sim \text{Tweedie((Population density) + (Region area) + (Longitude * Latitude))}
+\end{equation}
+
+\begin{equation}
+\tag{4}
+\text{TN density} \sim \text{Tweedie((Proportion cropland) + (Proportion tree cover) + (Proportion urban) + (Longitude * Latitude))}
+\end{equation}
+
+\begin{equation}
+\tag{5}
+\text{TN density} \sim \text{Tweedie((Population density) + (Region area) + (Proportion urban) + (Longitude * Latitude))}
 \end{equation}
@@
 
-The model identified locations of high and low trapping effort relative to land use classification, with proportion urban land very highly correlated with human population density (*r* = 0.835, 95% C.I.
-0.82-0.85, *p* < 0.001).
-Model inference was limited to regions around observed trap sites.
+We formulated these models as a Generalised Additive Models (GAM) incorporating a spatial interaction term (longitude by latitude) [@pedersen_hierarchical_2019].
+These models investigated the association between TN density, proportion of urban, cropland, tree cover and human population density.
+The models were constructed in the *mgcv* package using the R programming language [@wood_generalized_2017; @r_core_team_r_2020].
+Selection of the most parsimonious model was based on Deviance explained and the Akaike information criterion for each model.
+We performed sensitivity analysis by removing sites with imputed trapping effort and performing a pixel-based analysis.
 
-### What is the difference between global datasets rodent host distributions and distributions from trapping studies?
+### What is the difference in rodent host distributions between curated datasets and rodent trapping studies?
+
+Curated datasets (i.e., IUCN and GBIF) are commonly used in assessments of biodiversity in response to land use and climate change in addition to being used to explore zoonotic pathogen host distributions.
+To understand any biases introduced from these sources into these analyses we aimed to assess the concordance of curated rodent host distributions from IUCN and GBIF with observed rodent presence and absence from rodent trapping studies.
+We obtained rodent species distribution maps as shapefiles from the IUCN red list and translated these to a ~20km resolution raster [@iucn_iucn_2021].
+Distributions were cropped to the study region for globally distributed rodent species.
+We obtained rodent presence locations from GBIF as point data limited to the study region [@gbif_the_global_biodiversity_information_facility_occurrence_2021].
+Presence locations were associated to cells of raster with a ~20km resolution produced for the study region.
 
@@
 musculus*, *A.
-niloticus*, *Praomys daltoni* and *Cricetomys gambianus*) we mapped the detection and non-detection of rodent species and compared this to detection data obtained from GBIF and species ranges from IUCN [@gbif_the_global_biodiversity_information_facility_occurrence_2021; @iucn_iucn_2021].
-We identified rodent locations from GBIF and the rodent trapping studies within and outside the IUCN expected range.
-We located these to level-2 administrative regions to calculate the discrepencies between observed detections and non-detections with IUCN distributions.
+niloticus*, *Praomys daltoni* and *Cricetomys gambianus*) we mapped detection and non-detection, associating each location with a ~20km resolution raster.
+For these seven species we calculated the area of the IUCN expected range.
+We calculated the percentage of this range covered by presence detections in GBIF, this process was repeated for presence detections from rodent trapping studies.
+For both GBIF data and trapping studies we calculated the area of detections outside of the IUCN expected range.
+For rodent trapping studies only, we were able to calculate the area of non-detections within the IUCN expected area.
+Finally, we calculated the combined area of detection when combining GBIF and rodent trapping study data.
 
-### Are rodent trapping derived host-pathogen associations present in global zoonoses datasets?
+### Are rodent trapping derived host-pathogen associations present in a consolidated zoonoses dataset?
 
-For studies investigating rodent zoonoses we produced host-pathogen pairs.
-For each host-pathogen pair we compared our records with those available in CLOVER, reporting concordance and discordance.
+Studies investigating the risk of endemic zoonoses and novel pathogen emergence generally rely on consolidated datasets of host-pathogen associations.
+To examine the usefulness of rodent trapping studies as an additional source of data we compared identified host-pathogen asociations with a consolidated zoonoses dataset (CLOVER) [@Gibb_data_proliferation].
+CLOVER is a synthesis of four host-pathogen datasets (GMPD2, EID2, HP3 and Shaw, 2020) and was released in 2021, it contains more than 25,000 host-pathogen associations for Bacteria, Viruses, Helminth, Protozoa and Fungi.
+An archived version of the CLOVER dataset was obtained and subset for host species trapped within included studies, spatial filtering of CLOVER for host-pathogen associations limited to the study region was not possible [@Gibb_clover_data].
+
+We produced host-pathogen pairs from rodent trapping studies investigating zoonoses.
+For each host and pathogen, with both identified to species level, we compared our host-pathogen pair network with one produced using CLOVER, reporting concordance and discordance.
+We describe these networks and report differences in network composition.
 For host-pathogen pairs with assay results consistent with acute or prior infection we calculate the proportion positive and identify those absent from CLOVER.
@@
 
-For the four pathogens sampled at the most locations (Borreliaceae, Leptospiraceae, *Lassa mammarenavirus* and *Toxoplasma gondii*) we identify their potential host species.
-For the five species with most spatially complete sampling we associate their sampling locations with level-2 administrative regions and compare this to their IUCN range and rodent trapping detection range to calculate a measure sampling coverage.
+Sampling of zoonotic pathogens may also be biased within their hosts ranges, which may hinder inference across the entire range of the species.
+To quantify this, we focused on the five pathogen groups and species sampled at the most locations (Arenaviridae, Borreliaceae, Leptospiraceae, *Lassa mammarenavirus* and *Toxoplasma gondii*).
+We geolocated individuals of host species tested for these pathogens and calculated the area sampled.
+We adopt a similar approach to above by mapping sampled hosts to ~20km pixels and calculating the proportion of their IUCN range in which sampling has occurred.
+We also present the area covered of all areas in which the host has been detected in this study to produce a measure of relative completeness of sampling within the included rodent trapping studies.
+
+Data and code implemented in the R statistical language has been made available alongside this manuscript to reproduce these analyses [ref_zenodo].
 
@@
 
-Trap sites were situated in 273 of 1,414 (19.3%) level-2 administrative regions in 14 West African nations.
-The areas with highest trap night density included the capital and large cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako), Ghana (Accra),  Côte d'Ivoire (Abidjan), and Benin (Cotonou).
-Outside of these cities, regions in, Northern Senegal, Southern Guinea, Edo and Osun States in Nigeria and Eastern Sierra Leone had the highest density of trap nights (Figure 1A.).
+We found a non-random spatial distribution of rodent trapping locations across the study region, suggestive of underlying bias in the sampling or rodents across West Africa.
+Trap sites were situated in 256 of 1,450 (17.6%) level-2 administrative regions in 14 West African nations.
+The regions with highest TN density included the capitals and large cities of Niger (Niamey), Nigeria (Ibadan), Ghana (Accra), Senegal (Dakar), Ghana (Accra) and Benin (Cotonou).
+Outside of these cities, regions in, Northern Senegal, Southern Guinea, Edo and Ogun States in Nigeria and Eastern Sierra Leone had the greatest TN density (Figure 1A.).
 
 Trapping sites were biased towards human modified landscapes (i.e., cropland, grassland and mosaic habitats), with under-representation of forest and bare habitats.
-However, urban habitats were comparitively undersampled (Supplementary Figure 2.).
+However, urban habitats were comparatively undersampled (Supplementary Figure 2.).
 
-Trapping effort, measured through trap nights, was biased with increased effort in areas with higher proportion of urban land use and no-significant association with proportion of forest cover.
-The spatial term shows trapping bias towards increased effort in Northern Senegal, Eastern Sierra Leone, Eastern Guinea and coastal regions of Nigeria, Benin and Ghana, in contrast South-East Nigeria, Northern Nigeria and Liberia had relatively lower trapping effort (Figure 2).
-The most parsimonious GAM model (adjusted R^2^ = 0.523, Deviance explained = 61%) included terms for coordinates, proportion urban coverage and proportion tree coverage, model selection is shown in Supplementary Tables 3.1-3.4.
-In sensitivity analysis excluding sites with imputed trap nights reduced contributing data, however, Mauritania, Northern Senegal and Sierra Leone remained as regions trapped at higher rates with Nigeria being trapped at lower than expected rates (Supplementary Figure 3.).
+The most parsimonious GAM model (adjusted R2 = 0.3, Deviance explained = 48.7%) included smooth terms for geographic coordinates, human population density, proportion of urban landscape, and the area of the region.
+All smooth terms demonstrate significant non-linear associations (*p* < 0.002) with TN density (Supplementary Table 3.5).
+The spatial smooth term adjusted for population density, proportion of urban landscape and region area is shown in Figure 2.
+The final model identified relative trapping effort bias, with a bias towards increased trapping effort in North West Senegal, North and East Sierra Leone, Central Guinea and coastal regions of Nigeria, Benin and Ghana; in contrast South East Nigeria, Northern Nigeria and Burkina Faso had an observed bias towards a reduced trapping effort (Figure 2).
+Model summaries informing model selection are shown in Supplementary Tables 3.1-3.5.
+In sensitivity analysis, excluding sites with imputed trap nights reduced contributing data; however, Mauritania, Northern Senegal and Sierra Leone remained as regions trapped at higher rates, with Nigeria being trapped at lower than expected rates (Supplementary Figure 3a.).
+In pixel based sensitivity analysis spatial coverage was reduced with similar patterns of bias observed (Supplementary Figure 3b.).
 
 ![Figure 2.
-Relative trapping effort bias across West Africa from included studies adjusted for proportion urban land cover and proportion tree cover.
-Predictions are limited to areas around trap sites (coloured areas), uncertainty in the estimate of bias is represented by colour transparency/pallor.
-Purple regions represent areas with higher than expected trapping effort, yellow regions represent areas lower than expected trapping effort](`r here("figures", "Figure_2_updated.png")`)
+Relative trapping effort bias across West Africa from included studies adjusted for human population density, proportion urban land cover and area of the administrative region.
+Uncertainty in the point estimate of relative trapping effort bias is represented by colour transparency/pallor (e.g., Eastern Liberia and Central Mali).
+Yellow regions represent areas with a bias towards increased trapping effort (e.g., North West Senegal), purple regions represent areas with a bias towards reduced trapping effort (e.g.,Northern Nigeria).](`r here("figures", "Figure_2_updated.png")`)
 
-## What is the difference between global datasets rodent host distributions and distributions from trapping studies?
+## What is the difference in rodent host distributions between curated datasets and rodent trapping studies?
 
-For our species of interest, except *C.
+We found that rodent trapping studies could provide a useful supplement to curated data describing rodent host distributions.
+For the seven rodent species of interest (*M.
+natalensis*, *R.
+rattus*, *M.
+erythroleucus*, *M.
+musculus*, *A.
+niloticus*, *P.
+daltoni* and *C.
+gambianus*), except *C.
 gambianus*, trapping studies provided more distinct locations of detection and non-detection than were available from GBIF.
-For the West African endemic species IUCN ranges showed generally good concordance to both trapping studies and GBIF, although the range of *A.
+For the endemic rodent species (*M.
+natalensis*, *M.
+erythroleucus*, *A.
+niloticus*, *P.
+daltoni* and *C.
+gambianus*) IUCN ranges had good concordance to both trapping studies and GBIF, however, individuals of *A.
 niloticus* and *P.
-daltoni* are greater than expected.
-The invasive species *R.
+daltoni* were detected outside of IUCN ranges.
+In contrast, the non-native species *R.
 rattus* and *M.
-musculus* are present in a much larger range than is expected (Figure 3A.
+musculus* were detected across much greater ranges than were expected from IUCN distributions (Figure 3A.
 and 3B.).
@@
 M.
-musculus has no range in West African in IUCN data.](`r here("figures", "Figure_3_test_a.png")`)
+musculus has no IUCN West African range.](`r here("figures", "Figure_3_test_a.png")`)
 
@@
 L) Presence recorded in GBIF overlaid on IUCN species range (red-shaded area).
-R) Detection and non-detection from rodent trapping studies overlaid on IUCN species ranges.](`r here("figures", "Figure_3_test_b.png")`)
+R) Detection (purple) and non-detection (orange) from rodent trapping studies overlaid on IUCN species ranges.](`r here("figures", "Figure_3_test_b.png")`)
 
-We identified the proportion of a rodent's expected range in which detections and non-detections occurred (Table 1.).
-GBIF sampling occurred over > 15% of a rodent's expected range with detections occurring outside of IUCN ranges for all species, most noticeably for *A.
-niloticus* and *P.
-daltoni*.
-Trapping studies occurred over > 13% for all species except *C.
-gambianus* with detections occurring outside of IUCN ranges most noticeably for *A.
+To further understand the sampling coverage of curated datasets and included rodent trapping studies we identified the proportion of a species IUCN range in which detections and non-detections occurred for the seven species most detected in the included studies (Table 1.).
+Apart from *M.
+musculus* these species had IUCN ranges greater than 1 million km2, with the Mastomys species having the largest geographic ranges.
+Sampling locations of these species within GBIF covered between 0.09-0.26% of expected ranges with detections occurring outside IUCN ranges for all species, most noticeably for *A.
 niloticus* and *R.
 rattus*.
-Combining GBIF and trapping data increased coverage by a mean of 9% of the species range suggesting limited overlap between the regions studied.
-Non-detections of rodents occurred across large areas of their ranges with *A.
-niloticus* and *M.
-erythroleucus* not being detected in >15% of their expected range, suggestive of substantial heterogeneity fo presence across their range.
+Sampling locations within rodent trapping studies covered 0.03-0.24% of expected ranges with detections occurring outside IUCN ranges most noticeably for *A.
+niloticus* and *R.
+rattus*.
+Combining GBIF and rodent trapping data increased the sampled area by a mean of 65% (SD 36.5%) , suggesting limited overlap between the locations sampled.
+Non-detection of species occurred across species ranges (mean = 0.11%, SD = 0.03%), suggestive of heterogeneity of presence across IUCN ranges.
 
@@
 
-## Are rodent trapping derived host-pathogen associations present in global zoonoses datasets?
+## Are rodent trapping derived host-pathogen associations present in a consolidated zoonoses dataset?
 
-When limited to confirmed pathogens and classification of both pathogen and host to species level we identified 21 host-pathogen pairs among 11 rodent and 6 pathogen species (Figure 4.).
-We identified negative associations, through entirely negative assays, for 41 host-pathogen pairs among 31 rodent and 7 pathogen species.
-CLOVER contained 8 (38%) of our identified host-pathogen associations and positive associations for 6 (15%) of the negative associations.
+We found potentially important differences between the host-pathogen networks produced from included rodent trapping studies and the consolidated CLOVER dataset.
+When limited to taxonomic classification of both pathogen and host to species level we identified 25 host-pathogen pairs among 14 rodent and 6 pathogen species (Figure 4.).
+We identified negative associations (non-detection through specific assays) for 45 host-pathogen pairs among 35 rodent and 7 pathogen species.
+CLOVER contained 10 (40%) of our identified host-pathogen associations, the remaining 15 (60%) were not found to be present in CLOVER, additionally CLOVER recorded positive associations for 4 (9%) of the negative associations produced from the rodent trapping studies.
+
+![](`r here("figures", "Figure_4a_updated.png")`)
 
 ![Figure 4.
-L) Identified species level host-pathogen associations through detection of acute infection (i.e.
+A) Identified species level host-pathogen associations through detection of acute infection (i.e.
 PCR, culture).
@@
 Associations with a black border are present in the CLOVER dataset.
-R) Identified species level host-pathogen associations through serological assays (i.e.
+B) Identified species level host-pathogen associations through serological assays (i.e.
 ELISA).
 Percentages and colour relate to the proportion of all assays that were positive.
-Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Figure_4_updated.png")`)
+Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Figure_4b_updated.png")`)
 
-CLOVER included 399 host-pathogen associations we do not observe in rodent trapping studies among our included rodent species.
-The majority of these 329 (82%) pairs are from species with global distributions (*M.
+CLOVER included an additional 492 host-pathogen associations we do not observe in rodent trapping studies.
+The majority of these 392 (80%) pairs are from species with global distributions (*M.
 musculus*, *R.
 rattus* and *R.
-norvegicus*), or from those with wide ranging distributions in sub-Saharan Africa (31, 9%) (i.e., *A.
+norvegicus*), or from those with wide ranging distributions in sub-Saharan Africa (38, 8%) (i.e., *A.
 niloticus*, *M.
@@
 For pathogens not identified to species level (i.e.
-family or higher taxa only) we identified 128 host-pathogen pairs among 32 rodent species and 25 pathogen families (Supplementary Figure 4.), with CLOVER containing 52 (41%) of these associations.
+family or higher taxa only), we identified 148 host-pathogen pairs among 32 rodent species and 25 pathogen families (Supplementary Figure 4.), with CLOVER containing 66 (45%) of these associations.
 
-Rodent trapping studies identified rodent host species for six pathogens, *Toxoplasma gondii* (4), *Lassa mammarenavirus* (3), Usutu virus (2), *Coxiella burnetii* (2), *Escherichia coli* and *Klebsiella pneumoniae* (both 1), that were not present in this global host-pathogen association dataset.
+Rodent trapping studies identified additional rodent host species for six pathogens; *Lassa mammarenavirus* (5), *Toxoplasma gondii* (4), Usutu virus (2), *Coxiella burnetii* (2), *Escherichia coli* and *Klebsiella pneumoniae* (both 1), that were not present in this consolidated host-pathogen association dataset.
 
-## What is the spatial extent of pathogen testing within a hosts range?
+## What is the spatial extent of pathogen testing within a host’s range?
 
-The four most widely sampled pathogen species/families were Borreliaceae, *Lassa mammarenavirus*, Leptospiraceae and *Toxoplasma gondii* (Table 2.).
+The five most widely sampled pathogen species/families in included studies were Arenaviridae, Borreliaceae, *Lassa mammarenavirus*, Leptospiraceae and *Toxoplasma gondii* (Table 2.).
+Assays to identify Arenaviridae infection were performed in 44 rodent species with evidence of viral infection in 15 species.
+*Lassa mammarenavirus* was specifically tested for in 43 species with 10 showing evidence of viral infection.
+The most commonly infected species for both Arenaviridae, generally, and *Lassa mammarenavirus* specifically, were *M.
+natalensis* and *M.
+erythroleucus*.
+Assays were conducted on individuals trapped within less than a third of the locations in which the species were trapped (Table 2.).
+
 Infection with species of Borreliaceae was assessed in 42 species, with evidence of infection in 17 rodent species.
 Among species with more than 500 samples the highest rates of infection were among *A.
-niloticus* (16%), *M.
-huberti* (11%) and *M.
-erythroleucus* (8%).
-The five most widely assayed species were sampled from a relatively small area of their species ranges (7.3-16%), however, there was good representation when compared to where rodent trapping studies have occurred (24-85%).
-
-Infection with *Lassa mammarenavirus* was assessed in 28 species, with evidence of infection in 7 rodent species.
-Among species with more than 100 samples the highest rates of infection were among *M.
-natalensis* (22%), *M.
-erythroleucus* (6%) and *R.
-rattus* (4%).
-Species sampling occurred from limited areas within IUCN ranges (2-3%), however, sampling was more representative for rodent trapping studies (3-24%).
-
-Infection with species of Leptospiraceae was assessed in 8 species, with evidence of infection in 5 rodent species.
-Among species with more than 100 samples *M.
-natalensis* (19%), *R.
-rattus* (11%) and *Crocidura olivieri* (10%) were most frequently infected.
-Species sampling occurred across less than 1% of these species IUCN ranges and for all except *R.
-norvegicus* for less than 1% of their detection range in rodnet trapping studies.
-
-*Toxoplasma gondii* infection was assessed in 8 species, with evidence of infection in 6 rodent species.
-Among species with more than 100 samples *M.
-musculus* (7%), *M.
-erythroleucus* (4%) and *M.
-natalensis* (2%) were most commonly infected.
-Sampling occurred across less than 6% of their IUCN ranges, however, for *M.
-musculus*, *R.
-rattus* and *M.
-erythroleucus* sampling occurred across greater than 12% of their detection range.
-
-<!---BLOCK_LANDSCAPE_START--->
+niloticus* (16%), *Mastomys huberti* (11%) and *M.
+erythroleucus* (9%).
+The coverage of sampling within the trapping locations was slightly greater than that for Arenaviridae but remains a small proportion (<1%) of the host species IUCN range.
+Infection with species of Leptospiraceae and *Toxoplasma gondii* was assessed in 8 species, with evidence of infection in 5 and 6 rodent species respectively.
+Testing for these pathogens was more limited with few individuals tested and reduced coverage of host ranges.
 
@@
 
-table_2 <- read_rds(here("tables", "Table_2_updated.rds"))%>%
-  set_caption(caption = "Table 2: Comparison of pathogen sampling ranges for the 4 most widely sampled pathogens and the 5 most sampled potential rodent host species.
+table_2 <- read_rds(here("tables", "Table_2_updated_pixel.rds"))%>%
+  set_caption(caption = "Table 2: Comparison of pathogen sampling ranges for the 5 most widely sampled pathogens and the 5 most sampled potential rodent host species.
 * no IUCN range in West African") %>%
@@
 
-<!---BLOCK_LANDSCAPE_STOP--->
-
 # Discussion
 
-We synthesised data from 126 rodent trapping studies containing information on more than 72,000 trapped rodents across 1,331 trapping sites producing an estimated 966,000 trap-nights.
-Trapping occurred across 14 West African countries, studies investigating known or potential zoonotic pathogens occurred within 19% of level-2 administrative regions across West Africa with important bias towards areas with high population density and human dominated landscapes.
-For several important rodent species, we identified discordance between IUCN species ranges, GBIF presence data and detections and non-detections from rodent trapping studies, particularly among non-native species (*R.
-rattus* and *M.
-musculus*).
-We compared host-pathogen associations from rodent trapping studies with a global dataset and identified missing associations for several important zoonotic infectious diseases including, Lassa fever, Toxoplasmosis and Q fever in commonly occurring rodent species.
-Finally, we assessed the spatial coverage of testing for important zoonotic pathogens and found low representativeness across their host species range.
+Endemic rodent zoonoses and novel pathogen emergence from rodent hosts are predicted to have an increasing burden in West Africa and globally (@han_rodent_2015).
+Here we have synthesised data from 126 rodent trapping studies containing information on more than 72,000 rodents across 1,611 trap sites producing an estimated 942,669 trap nights.
+Trapping occurred across 14 West African countries, with locations complementary to curated datasets commonly used in studies assessing zoonosis risk based on host distributions (e.g.
+IUCN, GBIF).
+The inclusion of our synthesised dataset could counteract some of the biases inherent to these curated datasets [@Boakes].
+We identified 25 host-pathogen pairs reported from included studies, 15 of these were not included in a consolidated host-pathogen dataset.
+Pathogen sampling within identified species occurrence locations was limited, highlighting a further component of sampling bias supporting calls for further pathogen sampling across diverse species in zoonotic hotspots [@Harvey].
 
-Rodent trapping studies provide important information on the distribution of rodent species and their assemblages, we have identified important bias in the locations in which studies to data have been performed with large regions of countries including Nigeria, Guinea and Cote d'Ivoire being understudied.
-Trapping sites are clustered in areas with relatively high human population density and anthropogenic land use which can importantly bias our understanding of the distribution of rodent species and their pathogens across West Africa.
-Further effort is required to counteract this bias by sampling currently undersampled regions.
+We found that rodent trapping studies, similar to biodiversity data, showed important spatial biases [@Beck].
+Relative trapping effort bias was increased in Benin, Guinea, Senegal and Sierra Leone driven by long-standing research collaborations investigating the invasion non-native rodent species (*M.
+musculus* and *R.
+rattus*) and the hazard of endemic zoonosis outbreaks (e.g.
+Lassa mammarenavirus).
+Much of the region remains relatively understudied, particularly Burkina Faso, Côte d’Ivoire, Ghana and Nigeria despite these countries facing many of the same challenges.
+For example, annual outbreaks of Lassa fever are reported in Nigeria and there are potentially 60,000 unrecognised cases of Lassa fever each year in Côte d’Ivoire and Ghana [@Basinski].
+Rodent sampling should be targeted towards regions with minimal data to reduce this bias and assist in understanding the risk from endemic and novel zoonoses.
 
-Global datasets of rodent host distribution (e.g., IUCN) and presence (e.g., GBIF) provide useful resources to model the hazard of zoonotic disease spillover and novel pathogen emergence.
-We show here that non-detection data from rodent trapping studies can provide additional information to describe the heterogeneity of rodent assemblages across diverse habitats in West Africa.
-Combining presence data from GBIF and detection/non-detection data from rodent trapping studies can substantially increase coverage across a rodent's range, allowing improved estimates of rodent host prevalence for zoonotic infectious disease studies.
+Rodent trapping studies provide data on both presence and absence which are not available from curated datasets.
+Absence data can improve models of species distributions, however, missing data on trapping effort will confound the allocations of non-detections as true absences [@Vaclavik].
+Models of host species occurrence and abundance, improved by incorporating species absence, are important to assess the effect of land use and climate change on endemic zoonosis spillover to human populations and direct limited public health resources towards regions at greatest risk.
 
-Host-pathogen association datasets are becoming increasing comprehensive alongside increased access to molecular discovery techniques for potential pathogens.
-We show that currently available data from rodent trapping studies can enrich these resources, particularly with prevalence data and geolocation to better understand the potential hosts and reservoirs of known zoonotic pathogens.
-Most of the host-pathogen associations from CLOVER not identified from rodent trapping studies occurred in species with ranges that go beyond our study area, suggesting that these pathogens do not exist in West Africa.
-Rodent trapping data can support analyses using these global datasets to predict novel pathogen emergence and to quantify the risk of known zoonotic disease spillover based on host distributions.
+Currently available consolidated datasets on host-pathogen associations (e.g.
+CLOVER, EID2 and GMPD2) do not include spatial or temporal components [@Gibb_data_proliferation].
+The current synthesis of rodent trapping studies has shown that pathogens have only been sparsely tested for within a hosts range.
+Current zoonosis risk models dependent on these sources of data are unable to incorporate spatial heterogeneity in pathogen prevalence across a hosts range.
+Additional uncertainty in current models of zoonotic disease risk arises from host-pathogen associations that have not been reported in these consolidated datasets, for example *Hylomyscus pamfi* and Lassa fever and *R.
+rattus* and Q fever (*Coxiella burnetii*).
+Detection of zoonotic pathogens in multiple co-occurring host species supports the adoption of multi-species to better understand the potential range of endemic zoonoses [@Wilkinson].
 
-We have identified important limitations in the sampling of rodent zoonoses for known diseases across West Africa.
-For example, studies of *Lassa mammarenavirus* have taken place across 3% of *M.
-natalensis*'s range in West Africa.
-Lassa fever is known to be endemic in eight West African countries, although it is potentially more widely spread.
-Further rodent trapping studies are required to investigate this pathogen across a wider geographic range.
-Other important zoonotic diseases including Leptospirosis and Toxoplasmosis have been assessed over very limited regions of their hosts distribution.
+Few studies stratified presence-absence or pathogen detection by time at the level of rodent species, limiting observation of the changes in host and pathogen dynamics.
+This limitation prevents calculation of incidence of infection and prevalence of infectious rodents which potentially varies by both time and space [@fichet_calvet_2016].
+Understanding of temporal changes in viral burden and shedding for endemic zoonoses is required to accurately predict current and future risk of pathogen spillover.
 
 Due to data sparsity, we were unable to account for temporal change over the six decades of rodent trapping studies.
-Land use change and population density have changed dramatically over this period in West Africa.
-We attempted to mitigate this by using the median year of trapping to understand the spatial biases in trapping activity, however, it is possible that land use and population density at trapping sites varied importantly between when the study was conducted and the situation in 2005.
-Despite this limitation, the finding that trapping is biased towards high density, human dominated landscapes is unlikely to importantly change.
+Land use change and population density have changed dramatically over this period in West Africa [@hermann].
+We attempted to mitigate against this by using the median year of trapping to understand the spatial and land use biases in trapping activity, however, it is possible that land use and population density at trapping sites varied importantly between when the rodent trapping study was conducted and the conditions in 2005.
+Despite this limitation, the finding that trapping is biased towards high density, human dominated landscapes is unlikely to significantly change.
+We have shown that synthesis of rodent trapping studies to supplement curated rodent distributions can counteract some of the inherent biases in these data and that they can add further contextual data to host-pathogen association data.
+Together this supports their inclusion in efforts to model endemic zoonotic risk and novel pathogen emergence.
+However, utility can be improved by adopting reporting standards and practices consistent with Open Science [@Foster].
 
-We propose that rodent trapping studies can provide important data to supplement global datasets to quantify the hazard of zoonotic disease spillover and novel pathogen emergence.
-They can reduce some of the bias in global datasets however, important limitations remain.
-Further effort is required to survey rodent populations in non-human dominated landscapes and to assay pathogens across more of a host's range.
+Future rodent trapping studies should be targeted towards regions that are currently under-studied.
+Further information on rodent presence and abundance across West Africa will assist in modelling changing endemic zoonosis risk and the potential for novel pathogen emergence.
+Sharing of disaggregated data alongside research publications should be supported with adoption of data standards to support further data synthesis.
+Inclusion of exact locations of trapping sites, trapping effort and the dates at which trapping occurred would support more detailed understanding of the spatio-temporal dynamics of host populations and the risk of endemic zoonosis spillover events.
+Despite these challenges we propose that rodent trapping studies can provide an important source of data to supplement consolidated datasets on rodent distributions to quantify the risk of endemic zoonosis spillover events and the hazard of novel pathogen emergence.
 
@@
 
+```{r supplementary-table-3-5}
+
+read_rds(here("tables", "supplementary_table_3_5.rds"))
+
+```
+
 ## Supplementary Figure 1
@@
 
-![Supplementary Figure 3.
+![Supplementary Figure 3A.
 Relative trapping effort bias across West Africa from the subset of included studies reporting trapping effort adjusted for proportion urban land cover and proportion tree cover.
 Predictions are limited to areas around trap sites (coloured areas), uncertainty in the estimate of bias is represented by colour transparency/pallor.
-Purple regions represent areas with higher than expected trapping effort, yellow regions represent areas lower than expected trapping effort](`r here("figures", "Supplementary_figure_3.png")`)
+Purple regions represent areas with higher than expected trapping effort, yellow regions represent areas lower than expected trapping effort](`r here("figures", "Figure_2_sensitivity.png")`)
+
+![](`r here("figures", "Figure_2_ss.png")`)
 
 ## Supplementary Figure 4
+![](`r here("figures", "Supp_4a_updated.png")`)
 
 ![Supplementary Figure 4.
-L) Identified host-pathogen associations at pathogen family level through detection of acute infection (i.e.
+A) Identified host-pathogen associations at pathogen family level through detection of acute infection (i.e.
 PCR, culture).
@@
 Associations with a black border are present in the CLOVER dataset.
-R) Identified host-pathogen associations at pathogen family level through serological assays (i.e.
+B) Identified host-pathogen associations at pathogen family level through serological assays (i.e.
 ELISA).
 Percentages and colour relate to the proportion of all assays that were positive.
-Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Supplementary_figure_4_updated.png")`)
+Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Supp_4b_updated.png")`)

### 2022-05-24 main.R

@@ -50,3 +50,6 @@ source(here("scripts", "3_3_host_pathogen_associations.R"))
 # What is the spatial extent of pathogen testing within a host’s range?
 
 source(here("scripts", "3_4_host_pathogen_distributions.R"))
+
+# Supplementary material
+source(here("scripts", "4_0_supplementary.R"))

### 2022-05-24 scripts/3_2_rodent_host_distributions.R

@@ -55,6 +55,8 @@ rodent_iucn <- st_read(here("data_download", "iucn_data", "data_0.shp")) %>%
   group_by(classification) %>%
   summarise(geometry = st_union(geometry))
 
+write_rds(rodent_iucn, here("data_clean", "rodent_iucn.rds"))
+
 # GBIF data
 # We limit the figures to the top 7 most commonly trapped species
 
@@ -335,6 +337,8 @@ analysis_proportion_pixel <- function(species_name, trap_data = rodent_spatial, 
 testing_coverage_pixel <- lapply(species_names, analysis_proportion_pixel)
 
 table_1_pixel <- bind_rows(testing_coverage_pixel)
+
+write_rds(table_1_pixel, here("data_clean", "table_1_pixel.rds"))
 
 Table_1_pixel <- flextable(table_1_pixel) %>%
   bg(j = 2, bg = "grey", part = "all") %>%

### 2022-05-24 scripts/3_3_host_pathogen_associations.R

@@ -105,6 +105,7 @@ host_pathogen <- pathogen %>%
   group_by(classification) %>%
   mutate(n_pathogens_tested = n())
 
+if(!file.exists(here("data_clean", "hp_associations_dictionary.rds"))) {
 # We harmonise pathogen family names and define harmonised names for pathogens tested
 pathogen_family <- c("Arenaviridae", "Borreliaceae", "Flaviviridae", "Hantaviridae",
                      "Phenuiviridae", "Phenuiviridae", "Trypanosomatidae", "Arenaviridae",
@@ -123,6 +124,16 @@ pathogen_clean <- c("Arenaviridae sp.", "Borrelia sp.", "Flavivirus sp.", "Hanta
                     "Erhlichia sp.", "Mammarenavirus sp.", "Mycoplasma sp.", "Orentia sp.", "Ricketsia sp.", "Babesia sp.",
                     "Eimeria sp.", "Plasmodium sp.", "Strongyloides sp.")
 names(pathogen_clean) <- unique(host_pathogen$pathogen_tested)
+
+list(pathogen_family, pathogen_clean) %>%
+  write_rds(here("data_clean", "hp_associations_dictionary.rds"))
+
+} else {
+
+  pathogen_family <- read_rds(here("data_clean", "hp_associations_dictionary.rds"))[[1]]
+  pathogen_clean <- read_rds(here("data_clean", "hp_associations_dictionary.rds"))[[2]]
+
+}
 
 cleaned_pathogen <- host_pathogen %>%
   mutate(pathogen_family = recode_factor(pathogen_tested, !!!pathogen_family),
@@ -166,6 +177,8 @@ clover$combined <- bind_rows(clover$bacteria, clover$virus, clover$other) %>%
   mutate(source = "CLOVER") %>%
   arrange(classification, pathogen_name) %>%
   group_by(classification)
+
+write_rds(clover, here("data_clean", "clover_cleaned.rds"))
 
 # We produce a list of pathogens associated with each species, where there is an element per species
 clover_by_host <- clover$combined %>%
@@ -262,6 +275,7 @@ clover_not_in_trapping <- clover$combined %>%
 write_rds(confirmed_pathogen_host, here("data_clean", "host_pathogen_positive_plot_data.rds"))
 write_rds(negative_pathogen_host, here("data_clean", "host_pathogen_negative_plot_data.rds"))
[…]

### 2022-08-12 msc_data/msc_data.Rmd
mac version

---
title: "Description of available datasets."
date: "`r Sys.Date()`"
author: 
  - "David Simons"
output: 
  pdf_document
always_allow_html: yes
latex_engine: xelatex
---

```{r echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      crop = knitr::hook_pdfcrop)
library(tidyverse)
library(here)
```

# Data

Data has been obtained from a systematic literature search of studies reporting rodent trapping across West Africa. Studies have been performed for various reasons and the quality of data obtained from each study differs. Where studies have been conducted to investigate zoonoses data on pathogens have also been extracted.

The data has been divided into three datasets for you. First, a detection-non-detection dataset, second, an **abundance** dataset (not true ecological abundance), third a long-format pathogen dataset. In the following sections I will explain the variables.

## Detection-non-detection

If a study identified a rodent species at any of the trap sites this has been labelled a detection. If the species was not identified at another trap-site but was detected within the study (i.e. if it was there they would have reported it) it is termed a non-detection.

```{r detection-non-detection}

det_non_det <- read_csv(here("msc_data", "detection_non_detection.csv"))
head(det_non_det, 5)
```

| Variable name | Interpretation |
| --- | ---- |
| unique_id | This is the identifier of the study in which the data came from. This can be crosschecked against the studies page [here](https://docs.google.com/spreadsheets/d/1rQYjHhk6uk1PoKZZVsgFlmuqWGicU2tTisk9ddfAwTM/edit?usp=drive_web&ouid=109198876678825407497) |
| year_trapping | Where provided this is the period in which trapping occurred, this may have important implications due to land use change, population change, invasive rodents etc. |
| month_trapping | Where provided this is the period in which trapping occurred, this may have important implications due to seasonality |
| country & iso3c | The country in which trapping occurred with harmonised ISO3C codes |
| region | Specified region in which trapping occurred |
| town_village | If given the locations of the study site |
| habitat | The described habitat type in which trapping occurred |
| classification | The descriptor of the trapped rodent at the lowest taxonomic level available |
| trap_nights | The number of trap nights that were surveyed at this location. This value gives a measure of the trapping effort. For some studies this has been imputed where not presented based on the number of trapped rodents within a habitat |
| longitude and latitude | The GPS coordinates of the trapping location at the highest geographical resolution provided |
| detection | Whether that species was detected or not-detected at that trapping location |

## Abundance

This reports the number of individuals from a single species that were trapped at the highest geographical and temporal resolution. If you are not interested in temporal changes these can be grouped by habitat type to give the total numbers of individuals trapped across all sessions within a study.

```{r abundance}

abundance <- read_csv(here("msc_data", "abundance.csv"))
head(abundance, 5)
```

Only additional data to above is the number (abundance) variable

## Pathogens

I think you are less interested in this but perhaps this may help to focus on which rodent species you are interested in looking at in more depth. This data is in long format with each host-pathogen pair tested for separated out by study, geography, habitat and time.

```{r pathogen}

pathogen <- read_csv(here("msc_data", "pathogen.csv"))
head(pathogen, 5)
```

| Variable name | Interpretation |
| --- | ---- |
| unique_id | This is the identifier of the study in which the data came from. This can be crosschecked against the studies page [here](https://docs.google.com/spreadsheets/d/1rQYjHhk6uk1PoKZZVsgFlmuqWGicU2tTisk9ddfAwTM/edit?usp=drive_web&ouid=109198876678825407497) |
| year_trapping | Where provided this is the period in which trapping occurred, this may have important implications due to land use change, population change, invasive rodents etc. |
| month_trapping | Where provided this is the period in which trapping occurred, this may have important implications due to seasonality |
| country & iso3c | The country in which trapping occurred with harmonised ISO3C codes |
| region | Specified region in which trapping occurred |
| town_village | If given the locations of the study site |
| habitat | The described habitat type in which trapping occurred |
| classification | The descriptor of the trapped rodent at the lowest taxonomic level available |
| record_id | As several tests can be performed on the same rodent this value links all the tests for that same rodent species within the study (geography, habitat and trapping session) with one row for the assay and one row for the result for each pathogen |
| pathogen_tested | The lowest taxonomic name available for the pathogen tested as reported by the study |
| assay | The type of assay used for the pathogen (i.e. PCR or Ab/Ag) |
| number | The number of individuals tested or positive based on the value in the assay variable |
| longitude and latitude | The GPS coordinates of the trapping location at the highest geographical resolution provided |

### 2022-08-12 reports/manuscript_v4.Rmd
mac version

@@
 ---
-title: "Rodent trapping studies can improve our understanding of the hazards of zoonotic pathogen spillover and novel zoonotic pathogen emergence."
+title: "Rodent trapping studies can improve our understanding of the risk of endemic zoonoses spillover and novel zoonotic pathogen emergence."
 date: "`r Sys.Date()`"
@@
         style: false
+  # pdf_document
 always_allow_html: yes
 bibliography: citations/manuscript_citations.bib
-csl: elsevier_vancouver.csl
+csl: harvard.csl
+latex_engine: xelatex
 ---
@@
 
-NEEDS UPDATING
+Rodents are important globally distributed reservoirs of known and novel zoonotic pathogens.
+Ongoing anthropogenic land use change is altering the composition of host species assemblages and modifying the risk of endemic zoonoses spillover events.
+These changes mean that an understanding of the current distribution of rodent species is vital for accurately describing disease hazard and managing risk.
+However, available species distribution and host-pathogen association datasets such as International Union for Conservation of Nature (IUCN) distribution maps, Global Biodiversity Information Facility (GBIF) species occurrence data, or the CLOVER host-pathogen dataset are biased by the inclusion of limited data sources and spatial biases in sampled locations.
+Here, we synthesise data from West Africa from 127 rodent trapping studies, conducted between 1964-2021, including over 72,000 individual rodent detections to more comprehensively characterise the range and presence of important zoonotic pathogen host species in this region.
+We identify that rodent trapping studies, although biased towards human dominated landscapes across West Africa, can usefully complement current rodent species distribution datasets and calculate the discrepancies between sampling in GBIF and included studies.
+For five regionally important zoonotic pathogens (Arenaviridae spp., Borrelia spp., *Lassa mammarenavirus*, Leptospira spp.
+and *Toxoplasma gondii*) we identify host-pathogen associations that are not reported in the consolidated CLOVER dataset.
+These host-pathogen asssociation omissions may bias studies investigating the current risk of endemic zoonoses and the effect of land use and climate change on future novel pathogen emergence.
+Finally, we find that for these five pathogen groups, the proportion of a rodent hosts range that has been sampled remains small.
+A priority of future rodent trapping studies should be to sample rodent hosts across a greater geographic range to better characterise current and future risk.
+In the interim, future studies must incorporate a measure of the current sampling biases.
+The current synthesis of contextually rich rodent trapping data enriches available information from IUCN, GBIF and CLOVER which can support a more complete understanding of the hazard of zoonotic spillover events.
 
@@
 
-There is increasing awareness of the global health and economic impacts of novel zoonotic pathogen spillover, evidenced by the ongoing SARS-CoV-2 pandemic and previous HIV/AIDs and Spanish Influenza pandemics [@Bernstein].
+There is increasing awareness of the global health and economic impacts of novel zoonotic pathogen spillover, evidenced by the ongoing SARS-CoV-2 pandemic and previous HIV/AIDs and Spanish Influenza pandemics [@bernstein_costs_nodate].
 The number of zoonotic disease spillover events and the frequency of the emergence of novel zoonotic pathogens are predicted to increase under intensifying anthropogenic pressure driven by increased human populations, urbanisation, intensification of agriculture, climate change and wildlife defaunation [@allen_global_2017; @hassell_urbanization_2017; @garcia-pena_land-use_2021; @mcmahon_ecosystem_2018].
-The impact of endemic zoonoses meanwhile remains underestimated [@Maudlin].
-Endemic zoonoses disproportionally affect those in the poorest sections of society, those living in close contact with their animals and those with limited access to healthcare [@Halliday; @Molyneux; @judson_zoonoses_2021].
+The impact of endemic zoonoses meanwhile remains underestimated [@maudlin_neglected_2009].
+Endemic zoonoses disproportionally affect those in the poorest sections of society, those living in close contact with their animals and those with limited access to healthcare [@halliday_endemic_2015; @molyneux_zoonoses_2011; @judson_zoonoses_2021].
 West Africa is a nexus for these processes and is identified as a hotspot for increased risk of large epidemics of known zoonotic diseases and novel zoonotic pathogen emergence [@allen_global_2017; @grace_mapping_2012; @han_global_2016].
@@
 Despite the importance of understanding these complex systems, current evidence on host-pathogen associations is considerably affected by taxonomic and geographical sampling biases [@gibb_ecosystem_2020; @gibb_mammal_2021].
-Curated biodiversity datasets such as the Global Biodiversity Information Facility (GBIF) and resources produced by the International Union for Conservation of Nature suffer from well described sampling biases [@Bowler; @Boakes].
-These sampling biases can gravely distort produced species distribution models [@Beck].
-Datasets on host-pathogen associations also suffer from biases introduced from literature selection criteria and taxanomic discrepancies a recently produced consolidated dataset (CLOVER) has been produced to reduce this [@Gibb_clover].
+Curated biodiversity datasets such as the Global Biodiversity Information Facility (GBIF) and resources produced by the International Union for Conservation of Nature suffer from well described sampling biases [@bowler_temporal_nodate; @boakes_distorted_2010].
+These sampling biases can gravely distort produced species distribution models [@beck_spatial_2014].
+Datasets on host-pathogen associations also suffer from biases introduced from literature selection criteria and taxanomic discrepancies a recently produced consolidated dataset (CLOVER) has been produced to reduce this [@gibb_viralemergenceclover_2021].
 Studies identifying potential geographic hotspots of zoonotic disease spillover and novel pathogen emergence are typically based on these datasources for modelling host species distributions and host-pathogen associations which can result in biased hazard estimates [@plowright_sampling_2019; @carlson_future_2021].
@@
 Geographic locations of trapping studies was extracted using GPS locations for the most precise location presented.
-GPS coordinates for locations not reported were produced using the National Geospatial-Intelligence Agency GEOnet Names Server [@GeoNames] based on place-names and maps presented in the study.
+GPS coordinates for locations not reported were produced using the National Geospatial-Intelligence Agency GEOnet Names Server [@national_geospatial-intelligence_agency_nga_nodate] based on place-names and maps presented in the study.
 All locations were converted to decimal degrees for internal consistency.
@@
 
-To address this question, we calculated TN density within each West African level-2 administrative region, administrative boundaries were obtained from GADM 4.0.4 [@Database of Global Administrative Areas].
+To address this question, we calculated TN density within each West African level-2 administrative region, administrative boundaries were obtained from GADM 4.0.4 [@database_of_global_administrative_areas_gadm_2022 of Global Administrative Areas].
 We assigned trap nights to specific regions using GPS coordinates obtained from trapping studies as described above.
-The sf package in the R statistical language was used to handle shapefiles and geographic data [@Pebesma].
+The sf package in the R statistical language was used to handle shapefiles and geographic data [@pebesma_simple_2018].
 Density was calculated by dividing the number of trap nights by the area of a level-2 administrative area.
 For studies not reporting trap nights imputation was used as previously described.
-Human population density was obtained for the closest year (2005) to the median year of trapping (2007) from Socioeconomic Data and Applications Center (SEDAC) gridded population of the world v4 at ~ 1km resolution [@SEDAC].
+Human population density was obtained for the closest year (2005) to the median year of trapping (2007) from Socioeconomic Data and Applications Center (SEDAC) gridded population of the world v4 at ~ 1km resolution [@socioeconomic_data_and_applications_center_gridded_2021].
 Median population density was then calculated for each level-2 administrative region.
-Land cover classification was obtained from the Copernicus climate change service at ~300m resolution [@COPERNICUS].
+Land cover classification was obtained from the Copernicus climate change service at ~300m resolution [@european_space_agency_climate_change_initiative_land_2022].
 The proportion of cropland, shrubland, tree cover and urban land cover within a level-2 administrative region in 2005 was calculated.
@@
 $$
+
 \begin{equation}
@@
 \text{TN density} \sim \text{Tweedie((Longitude * Latitude))}
-\end{equation}
 
-\begin{equation}
 \tag{2}
 \text{TN density} \sim \text{Tweedie((Population density) + (Longitude * Latitude))}
-\end{equation}
 
-\begin{equation}
 \tag{3}
 \text{TN density} \sim \text{Tweedie((Population density) + (Region area) + (Longitude * Latitude))}
-\end{equation}
 
-\begin{equation}
-\tag{4}
 \text{TN density} \sim \text{Tweedie((Proportion cropland) + (Proportion tree cover) + (Proportion urban) + (Longitude * Latitude))}
-\end{equation}
 
-\begin{equation}
 \tag{5}
@@
 \end{equation}
+
 $$
@@
 Studies investigating the risk of endemic zoonoses and novel pathogen emergence generally rely on consolidated datasets of host-pathogen associations.
-To examine the usefulness of rodent trapping studies as an additional source of data we compared identified host-pathogen asociations with a consolidated zoonoses dataset (CLOVER) [@Gibb_data_proliferation].
+To examine the usefulness of rodent trapping studies as an additional source of data we compared identified host-pathogen asociations with a consolidated zoonoses dataset (CLOVER) [@gibb_mammal_2021].
 CLOVER is a synthesis of four host-pathogen datasets (GMPD2, EID2, HP3 and Shaw, 2020) and was released in 2021, it contains more than 25,000 host-pathogen associations for Bacteria, Viruses, Helminth, Protozoa and Fungi.
-An archived version of the CLOVER dataset was obtained and subset for host species trapped within included studies, spatial filtering of CLOVER for host-pathogen associations limited to the study region was not possible [@Gibb_clover_data].
+An archived version of the CLOVER dataset was obtained and subset for host species trapped within included studies, spatial filtering of CLOVER for host-pathogen associations limited to the study region was not possible [@gibb_viralemergenceclover_2021].
 
@@
 
-Endemic rodent zoonoses and novel pathogen emergence from rodent hosts are predicted to have an increasing burden in West Africa and globally (@han_rodent_2015).
-Here we have synthesised data from 126 rodent trapping studies containing information on more than 72,000 rodents across 1,611 trap sites producing an estimated 942,669 trap nights.
-Trapping occurred across 14 West African countries, with locations complementary to curated datasets commonly used in studies assessing zoonosis risk based on host distributions (e.g.
-IUCN, GBIF).
-The inclusion of our synthesised dataset could counteract some of the biases inherent to these curated datasets [@Boakes].
+Endemic rodent zoonoses and novel pathogen emergence from rodent hosts are predicted to have an increasing burden in West Africa and globally (Han et al.
+2015).
+Here we have synthesised data from 126 rodent trapping studies containing information on more than 76,000 rodents across 1,611 trap sites producing an estimated 942,669 trap nights across 14 West African countries.
+Locations studied are complementary to curated datasets (e.g.
+IUCN, GBIF), incorporation of our synthesised dataset when assessing zoonosis risk based on host distributions could counteract some of the biases inherent to these curated datasets (Boakes et al., 2010).
 We identified 25 host-pathogen pairs reported from included studies, 15 of these were not included in a consolidated host-pathogen dataset.
-Pathogen sampling within identified species occurrence locations was limited, highlighting a further component of sampling bias supporting calls for further pathogen sampling across diverse species in zoonotic hotspots [@Harvey].
+However, pathogen sampling within identified species occurrence locations was limited, highlighting a further component of sampling bias supporting calls for further pathogen sampling across diverse species in zoonotic hotspots (Harvey and Holmes, 2022).
 
-We found that rodent trapping studies, similar to biodiversity data, showed important spatial biases [@Beck].
-Relative trapping effort bias was increased in Benin, Guinea, Senegal and Sierra Leone driven by long-standing research collaborations investigating the invasion non-native rodent species (*M.
-musculus* and *R.
-rattus*) and the hazard of endemic zoonosis outbreaks (e.g.
+We found that rodent trapping studies, similar to biodiversity data, showed important spatial biases (Beck et al., 2014).
+Relative trapping effort was biased towards increased effort in Benin, Guinea, Senegal and Sierra Leone driven by long-standing research collaborations investigating the invasion of non-native rodent species (M.
+musculus and R.
+rattus) and the risk of endemic zoonosis outbreaks (e.g.
 Lassa mammarenavirus).
-Much of the region remains relatively understudied, particularly Burkina Faso, Côte d’Ivoire, Ghana and Nigeria despite these countries facing many of the same challenges.
-For example, annual outbreaks of Lassa fever are reported in Nigeria and there are potentially 60,000 unrecognised cases of Lassa fever each year in Côte d’Ivoire and Ghana [@Basinski].
+Much of West Africa remains relatively understudied, particularly Burkina Faso, Côte d’Ivoire, Ghana and Nigeria, despite these countries facing many of the same challenges.
+For example, annual outbreaks of Lassa fever are reported in Nigeria and there are potentially 60,000 unrecognised cases of Lassa fever each year in Côte d’Ivoire and Ghana (Basinski et al., 2021).
 Rodent sampling should be targeted towards regions with minimal data to reduce this bias and assist in understanding the risk from endemic and novel zoonoses.
 
-Rodent trapping studies provide data on both presence and absence which are not available from curated datasets.
-Absence data can improve models of species distributions, however, missing data on trapping effort will confound the allocations of non-detections as true absences [@Vaclavik].
-Models of host species occurrence and abundance, improved by incorporating species absence, are important to assess the effect of land use and climate change on endemic zoonosis spillover to human populations and direct limited public health resources towards regions at greatest risk.
+Rodent trapping studies provide geographic and temporally contextualised data on both species detection and non-detection which are not available from curated datasets.
+Non-detection data can improve models of species distributions, unfortunately, high levels of missing data on trapping effort will continue to confound the allocations of non-detections as true absences (Václavík and Meentemeyer, 2009).
+Models of host species occurrence and abundance, improved by incorporating species absence, are important to assess the effect of land use and climate change on endemic zoonosis spillover to human populations and direct limited public health resources towards regions at greatest risk (Plowright et al., 2019).
 
 Currently available consolidated datasets on host-pathogen associations (e.g.
-CLOVER, EID2 and GMPD2) do not include spatial or temporal components [@Gibb_data_proliferation].
-The current synthesis of rodent trapping studies has shown that pathogens have only been sparsely tested for within a hosts range.
-Current zoonosis risk models dependent on these sources of data are unable to incorporate spatial heterogeneity in pathogen prevalence across a hosts range.
-Additional uncertainty in current models of zoonotic disease risk arises from host-pathogen associations that have not been reported in these consolidated datasets, for example *Hylomyscus pamfi* and Lassa fever and *R.
-rattus* and Q fever (*Coxiella burnetii*).
-Detection of zoonotic pathogens in multiple co-occurring host species supports the adoption of multi-species to better understand the potential range of endemic zoonoses [@Wilkinson].
+CLOVER, EID2 and GMPD2) do not include spatial or temporal components (Gibb, Gregory F.
+Albery, et al., 2021a).
+The current synthesis of rodent trapping studies has highlighted that pathogens have been sparsely sampled within a hosts range.
+Current zoonosis risk models dependent on these sources of data are therefore not able to incorporate spatial heterogeneity in pathogen prevalence across the host range.
+Additional uncertainty in current models of zoonotic disease risk arises from observed host-pathogen associations that have not been reported in these consolidated datasets.
+For example, Hylomyscus pamfi infected with Lassa mammarenavirus and R.
+rattus infected with Coxiella burnetiid, will not be included when solely dependent on consolidated host-pathogen datasets.
+Further, detection of zoonotic pathogens in multiple co-occurring host species supports the adoption of multi-species to better understand the potential range of endemic zoonoses (Wilkinson et al., 2019).
 
-Few studies stratified presence-absence or pathogen detection by time at the level of rodent species, limiting observation of the changes in host and pathogen dynamics.
-This limitation prevents calculation of incidence of infection and prevalence of infectious rodents which potentially varies by both time and space [@fichet_calvet_2016].
+Few studies stratified detection and non-detection of hosts or pathogen prevalence by time, therefore limiting inference of changes in host and pathogen dynamics.
+This limitation prevents calculation of incidence of infection and the abundance of infectious rodents which potentially varies by both time and space (Fichet-Calvet et al., 2016).
 Understanding of temporal changes in viral burden and shedding for endemic zoonoses is required to accurately predict current and future risk of pathogen spillover.
 
-Due to data sparsity, we were unable to account for temporal change over the six decades of rodent trapping studies.
-Land use change and population density have changed dramatically over this period in West Africa [@hermann].
-We attempted to mitigate against this by using the median year of trapping to understand the spatial and land use biases in trapping activity, however, it is possible that land use and population density at trapping sites varied importantly between when the rodent trapping study was conducted and the conditions in 2005.
-Despite this limitation, the finding that trapping is biased towards high density, human dominated landscapes is unlikely to significantly change.
+Finally, due to data sparsity, we were unable to account for temporal change over the six decades of rodent trapping studies.
+Land use change and population density have changed dramatically over this period in West Africa (Herrmann et al., 2020).
+We attempted to mitigate against this by using the median year of trapping as a reference point to understand the spatial and land use biases in trapping activity.
+It is possible that land use and population density at trapping sites varied importantly between when the rodent trapping study was conducted and the land use and human population variables in 2005.
+Despite this limitation, the finding that trapping is biased towards high density, human dominated landscapes is unlikely to substantially change (Beck et al., 2014).
+
 We have shown that synthesis of rodent trapping studies to supplement curated rodent distributions can counteract some of the inherent biases in these data and that they can add further contextual data to host-pathogen association data.
 Together this supports their inclusion in efforts to model endemic zoonotic risk and novel pathogen emergence.
-However, utility can be improved by adopting reporting standards and practices consistent with Open Science [@Foster].
+Contribution of rodent trapping studies as data sources can be improved by adopting reporting standards and practices consistent with Open Science, namely sharing of disaggregated datasets alongside publication (Foster and Deardorff, 2017).
 
@@
 Inclusion of exact locations of trapping sites, trapping effort and the dates at which trapping occurred would support more detailed understanding of the spatio-temporal dynamics of host populations and the risk of endemic zoonosis spillover events.
-Despite these challenges we propose that rodent trapping studies can provide an important source of data to supplement consolidated datasets on rodent distributions to quantify the risk of endemic zoonosis spillover events and the hazard of novel pathogen emergence.
+Despite these challenges we propose that rodent trapping studies can provide an important source of data to supplement curated datasets on rodent distributions to quantify the risk of endemic zoonosis spillover events and the hazard of novel pathogen emergence.
 

### 2022-08-12 scripts/data_for_msc.R

@@ -1,0 +1,17 @@
+detection_non_detection <- rodent_data %>%
+  select(unique_id, year_trapping, month_trapping, country, iso3c, region, town_village, habitat, classification, number, trap_nights, longitude, latitude) %>%
+  mutate(detection = case_when(number > 0 ~ "Detected",
+                              TRUE ~ "Not detected")) %>%
+  select(- number)
+
+abundance <- rodent_data %>%
+  select(unique_id, year_trapping, month_trapping, country, iso3c, region, town_village, habitat, classification, number, trap_nights, longitude, latitude)
+
+pathogen <- long_pathogen %>%
+  select(unique_id, year_trapping, month_trapping = month, country, iso3c, region, town_village, classification, record_id, pathogen_tested, assay, number, geometry) %>%
+  mutate(longitude = st_coordinates(geometry)[, 1],
+         latitude = st_coordinates(geometry)[, 2])
+
+write_csv(detection_non_detection, here("detection_non_detection.csv"))
+write_csv(abundance, here("abundance.csv"))
+write_csv(pathogen, here("pathogen.csv"))

### 2022-08-22 reports/manuscript_v5.Rmd
updated manuscript to v5

--- 2022-08-12 reports/manuscript_v4.Rmd
@@
 ---
-title: "Rodent trapping studies can improve our understanding of the risk of endemic zoonoses spillover and novel zoonotic pathogen emergence."
+title: "Rodent trapping studies as an overlooked information source for understanding endemic and novel zoonotic spillover."
 date: "`r Sys.Date()`"
@@
 Rodents are important globally distributed reservoirs of known and novel zoonotic pathogens.
-Ongoing anthropogenic land use change is altering the composition of host species assemblages and modifying the risk of endemic zoonoses spillover events.
+Ongoing anthropogenic land use change is altering the composition of host species assemblages and modifying the risk of zoonoses spillover events.
 These changes mean that an understanding of the current distribution of rodent species is vital for accurately describing disease hazard and managing risk.
-However, available species distribution and host-pathogen association datasets such as International Union for Conservation of Nature (IUCN) distribution maps, Global Biodiversity Information Facility (GBIF) species occurrence data, or the CLOVER host-pathogen dataset are biased by the inclusion of limited data sources and spatial biases in sampled locations.
-Here, we synthesise data from West Africa from 127 rodent trapping studies, conducted between 1964-2021, including over 72,000 individual rodent detections to more comprehensively characterise the range and presence of important zoonotic pathogen host species in this region.
-We identify that rodent trapping studies, although biased towards human dominated landscapes across West Africa, can usefully complement current rodent species distribution datasets and calculate the discrepancies between sampling in GBIF and included studies.
+However, available species distribution and host-pathogen association datasets (e.g.
+IUCN, GBIF, CLOVER) are often taxonomically and spatially biased.
+Here, we synthesise data from West Africa from 127 rodent trapping studies, published between 1964-2022, as an additional source of information to characterise the range and presence of important zoonotic pathogen host species in this region.
+We identify that these rodent trapping studies, although biased towards human dominated landscapes across West Africa, can usefully complement current rodent species distribution datasets and we calculate the discrepancies between these datasets.
 For five regionally important zoonotic pathogens (Arenaviridae spp., Borrelia spp., *Lassa mammarenavirus*, Leptospira spp.
-and *Toxoplasma gondii*) we identify host-pathogen associations that are not reported in the consolidated CLOVER dataset.
-These host-pathogen asssociation omissions may bias studies investigating the current risk of endemic zoonoses and the effect of land use and climate change on future novel pathogen emergence.
-Finally, we find that for these five pathogen groups, the proportion of a rodent hosts range that has been sampled remains small.
+and *Toxoplasma gondii*), we identify host-pathogen associations that have not been previously reported in host-association datasets.
+These omissions have the potential for biasing estimates of current risk and drivers of zoonoses.
+Finally, for these five pathogen groups, we find that the proportion of a rodent hosts range that has been sampled remains small.
 A priority of future rodent trapping studies should be to sample rodent hosts across a greater geographic range to better characterise current and future risk.
-In the interim, future studies must incorporate a measure of the current sampling biases.
+In the interim, future studies of spatial pathogen risk informed by rodent distributions must incorporate a measure of the current sampling biases.
 The current synthesis of contextually rich rodent trapping data enriches available information from IUCN, GBIF and CLOVER which can support a more complete understanding of the hazard of zoonotic spillover events.
@@
 
-There is increasing awareness of the global health and economic impacts of novel zoonotic pathogen spillover, evidenced by the ongoing SARS-CoV-2 pandemic and previous HIV/AIDs and Spanish Influenza pandemics [@bernstein_costs_nodate].
-The number of zoonotic disease spillover events and the frequency of the emergence of novel zoonotic pathogens are predicted to increase under intensifying anthropogenic pressure driven by increased human populations, urbanisation, intensification of agriculture, climate change and wildlife defaunation [@allen_global_2017; @hassell_urbanization_2017; @garcia-pena_land-use_2021; @mcmahon_ecosystem_2018].
+There is increasing awareness of the global health and economic impacts of novel zoonotic pathogen spillover, evidenced by the ongoing SARS-CoV-2 pandemic and previous HIV/AIDs and Spanish Influenza pandemics [@bernstein_costs_2022].
+The number of zoonotic disease spillover events and the frequency of the emergence of novel zoonotic pathogens from rodents are predicted to increase under intensifying anthropogenic pressure driven by increased human populations, urbanisation, intensification of agriculture, climate change and wildlife defaunation [@allen_global_2017; @hassell_urbanization_2017; @garcia-pena_land-use_2021; @mcmahon_ecosystem_2018].
 The impact of endemic zoonoses meanwhile remains underestimated [@maudlin_neglected_2009].
 Endemic zoonoses disproportionally affect those in the poorest sections of society, those living in close contact with their animals and those with limited access to healthcare [@halliday_endemic_2015; @molyneux_zoonoses_2011; @judson_zoonoses_2021].
-West Africa is a nexus for these processes and is identified as a hotspot for increased risk of large epidemics of known zoonotic diseases and novel zoonotic pathogen emergence [@allen_global_2017; @grace_mapping_2012; @han_global_2016].
-Two taxonomic groups - rodents (Rodentia) and bats (Chiroptera) - contribute the greatest number of predicted novel zoonotic pathogens and known zoonoses [@han_rodent_2015; @gibb_mammal_2021].
-Of 2,220 extant rodent species, 244 (10.7%) are described as reservoirs of 85 zoonotic pathogens [@han_rodent_2015].
-Specifically, West Africa has been identified as a region at increased hazard for rodent-borne zoonotic disease spillover events under different projected future land-use change scenarios [@garcia-pena_land-use_2021].
-Currently within West Africa, rodents are involved in the transmission of multiple endemic zoonoses with large burdens on human health, these pathogens include, Lassa fever, Schistosomiasis, Leptospirosis and Toxoplasmosis [@meerburg_rodent-borne_2009; @galeh_global_2020].
 
-Rodent species form diverse assemblages, which provide important and beneficial ecosystem services including pest regulation and seed dispersal [@fischer_ecosystem_2018].
-The role of rodent species in zoonotic disease spillover or novel zoonotic pathogens emergence are examples of ecosystem disservices.
+Rodents along with bats contribute the greatest number of predicted novel zoonotic pathogens and known zoonoses [@han_rodent_2015; @gibb_mammal_2021].
+Of 2,220 extant rodent species, 244 (10.7%) are described as reservoirs of 85 zoonotic pathogens [@han_rodent_2015], although many species provide important and beneficial ecosystem services including pest regulation and seed dispersal [@fischer_ecosystem_2018].
 Rodents typically demonstrate “fast” life histories [@dobson_fast_2007] with traits such as early maturation and short gestation times (<4 days) further associated with being zoonotic reservoirs [@han_rodent_2015; @albery_fast-lived_2021].
@@
 Despite the importance of understanding these complex systems, current evidence on host-pathogen associations is considerably affected by taxonomic and geographical sampling biases [@gibb_ecosystem_2020; @gibb_mammal_2021].
-Curated biodiversity datasets such as the Global Biodiversity Information Facility (GBIF) and resources produced by the International Union for Conservation of Nature suffer from well described sampling biases [@bowler_temporal_nodate; @boakes_distorted_2010].
-These sampling biases can gravely distort produced species distribution models [@beck_spatial_2014].
-Datasets on host-pathogen associations also suffer from biases introduced from literature selection criteria and taxanomic discrepancies a recently produced consolidated dataset (CLOVER) has been produced to reduce this [@gibb_viralemergenceclover_2021].
-Studies identifying potential geographic hotspots of zoonotic disease spillover and novel pathogen emergence are typically based on these datasources for modelling host species distributions and host-pathogen associations which can result in biased hazard estimates [@plowright_sampling_2019; @carlson_future_2021].
+Curated biodiversity datasets such as the Global Biodiversity Information Facility (GBIF) and resources produced by the International Union for Conservation of Nature suffer from well described spatial and temporal sampling biases [@boakes_distorted_2010; @bowler_temporal_2022].
+These sampling biases can importantly distort produced species distribution models [@beck_spatial_2014].
+Datasets on host-pathogen associations also can suffer from biases introduced from literature selection criteria and taxanomic discrepancies.
+These biases are important because identification of potential geographic hotspots of zoonotic disease spillover and novel pathogen emergence are often based on these types of host species distributions and host-pathogen associations [@plowright_sampling_2019; @carlson_future_2021].
 For example, systematically increased sampling, over-representation of certain habitats and clustering around areas of high human population could lead to an apparent association between locations and hazard that is driven by these factors rather than underlying host-pathogen associations [@wille_how_2021; @redding_evaluating_2017; @gibb_mammal_2021].
-While other regions remain systematically undersampled (e.g., areas of sparse human populations, endemic zoonoses), potentially resulting in a reduced hazard of zoonotic spillover events or novel zoonotic pathogen emergence being attributed [@beck_spatial_2014].
 Predictions of zoonotic disease spillover and novel zoonotic pathogen emergence must account for these biases to understand the future hazard of zoonotic diseases [@carlson_future_2021].
 
-Rodent trapping studies provide contextually rich information on when, where and under what conditions rodents were trapped, potentially enriching consolidated datasets and providing additional data to minimise inherent biases [@bovendorp_optimising_2017].
-Rodent trapping studies describe rodent population assemblages, their geographic distribution and host-pathogen associations beyond what is currently available in consolidated datasets.
+West Africa has been identified as a region at increased risk for rodent-borne zoonotic disease spillover events, the probability of these events are predicted to increase under different projected future land-use change scenarios [@grace_mapping_2012; @garcia-pena_land-use_2021].
+Currently within West Africa, rodents are involved in the transmission of multiple endemic zoonoses with large burdens on human health, these pathogens include, Lassa fever, Schistosomiasis, Leptospirosis and Toxoplasmosis [@meerburg_rodent-borne_2009; @galeh_global_2020].
+Understanding of the distribution of these zoonoses are limited by biases in consolidated datasets.
+Rodent trapping studies provide contextually rich information on when, where and under what conditions rodents were trapped, potentially enriching consolidated datasets [@bovendorp_optimising_2017].
 Studies have been conducted in West Africa to investigate the distribution of rodent species, their species assemblages, the prevalence of endemic zoonoses within rodent hosts (e.g., Lassa fever, Schistosomiasis) and to identify emerging and novel zoonotic pathogens [@fichet-calvet_diversity_2009; @catalano_multihost_2020; @usaid_one_2021].
@@
 
-Here, we synthesise rodent trapping studies conducted across West Africa from a search of literature between 1964-2022.
+Here, we synthesise rodent trapping studies conducted across West Africa published between 1964-2022.
 First, we use this dataset to investigate the geographic sampling biases in relation to human population density and land use classification.
@@
 
-For further analysis, we included studies if they met all of the following inclusion criteria; i) reported findings from trapping studies where the target was a small mammal, ii) described the type of trap used or the length of trapping activity or the location of the trapping activity, iii) included trapping activity from at least one West African country, iv) recorded the genus or species of trapped individuals, and v) were published in a peer-reviewed journal or as a pre-print on a digital platform or as a report by a credible organisation.
+We included studies for further analysis if they met all of the following inclusion criteria; i) reported findings from trapping studies where the target was a small mammal, ii) described the type of trap used or the length of trapping activity or the location of the trapping activity, iii) included trapping activity from at least one West African country, iv) recorded the genus or species of trapped individuals, and v) were published in a peer-reviewed journal or as a pre-print on a digital platform or as a report by a credible organisation.
 We excluded studies if they met any of the following exclusion criteria: i) reported data that were duplicated from a previously included study, ii) no full text available, iii) not available in English.
@@
 We identified the aims of included studies, for example, whether it was conducted as a survey of small mammal species or specifically to assess the risk of zoonotic disease spillover.
-To understand potential limitations in the representativeness of rodent trapping we extracted data on study methodology, such as, the number of trap nights, the type of traps used and whether the study attempted to estimate abundance.
+we extracted data on study methodology, such as, the number of trap nights, the type of traps used and whether the study attempted to estimate abundance.
 For studies not reporting number of trap nights we used imputation based on the number of trapped individuals, stratified by the habitat type from which they were obtained.
@@
 
-The method of taxonomic identification applied within a study was recorded, reported species identification was assumed to be accurate.
-The number of individuals of these species or genera was extracted with species and genus names mapped to GBIF taxonomy [@gbif_the_global_biodiversity_information_facility_global_2021] for internal consistency.
+We also recorded how species were identified within a study and species identification was assumed to be accurate.
+The number of individuals of these species or genera was extracted with taxanomic names mapped to GBIF taxonomy [@gbif_the_global_biodiversity_information_facility_global_2021].
 We expanded species detection and non-detection records by explicitly specifying non-detection at a trap site if a species was recorded as detected at other trapping locations within the same study.
 
-Geographic locations of trapping studies was extracted using GPS locations for the most precise location presented.
-GPS coordinates for locations not reported were produced using the National Geospatial-Intelligence Agency GEOnet Names Server [@national_geospatial-intelligence_agency_nga_nodate] based on place-names and maps presented in the study.
-All locations were converted to decimal degrees for internal consistency.
+Geographic locations of trapping studies were extracted using GPS locations for the most precise location presented.
+Missing locations were found using the National Geospatial-Intelligence Agency GEOnet Names Server [@national_geospatial-intelligence_agency_nga_nodate] based on place-names and maps presented in the study.
+All locations were converted to decimal degrees.
 The year of rodent trapping was extracted alongside the length of the trapping activity to understand seasonal representativeness of trapping activity.
@@
 
-We summarised the number of studies, the year in which trapping occurred and the country in which they were conducted.
-We identified 4,692 relevant citations, with 127 rodent trapping studies included (Supplementary Table 2.).
+Out of 4,692 relevant citations, we identified 127 rodent trapping studies (Supplementary Table 2.).
 The earliest trapping studies were conducted in 1964, with a trend of increasing numbers of studies being performed annually since 2000.
@@
 
-Included studies explicitly reported on 601,184 trap nights, a further 341,445 trap nights were imputed from studies with no recording of trapping effort based on trap success.
-A total of 942,629 trap nights were estimated (Figure 1B.).
+Included studies explicitly reported on 601,184 trap nights, a further 341,445 trap nights were imputed from studies with no recording of trapping effort based on trap success, leading to an estimate of 942,629 trap nights (Figure 1B.).
 A minority of studies trapped at a single study site (30, 24%), with 46 (36%) trapping at between two and five sites, the remaining 51 studies (40%) trapped at between six and 93 study sites.
@@
 With *Rattus rattus*, *Mus musculus*, *Mastomys erythroleucus*, *Mastomys natalensis* and *Arvicanthis niloticus* being assayed for at least 18 microorganisms.
-Further descriptive information from the included studies including geolocated trapping of species, their detection and non-detection alongside microorganism data is available as supplementary material [@simons_rodents_2022].
 
@@
 
-To investigate the risk to human health from endemic zoonoses and novel pathogen emergence using synthesised data from rodent trapping data understanding of sampling biases is vital.
-Sampling of rodents has not been performed randomly across space in West Africa (Figure 1.) and at varying intensity, we aimed quantify this bias as the relationship between trapping effort (measured as trap night density (TN density)), human population density and land use type.
-We define this as relative trapping effort bias.
-
-To address this question, we calculated TN density within each West African level-2 administrative region, administrative boundaries were obtained from GADM 4.0.4 [@database_of_global_administrative_areas_gadm_2022 of Global Administrative Areas].
-We assigned trap nights to specific regions using GPS coordinates obtained from trapping studies as described above.
-The sf package in the R statistical language was used to handle shapefiles and geographic data [@pebesma_simple_2018].
-Density was calculated by dividing the number of trap nights by the area of a level-2 administrative area.
+To investigate the extent of spatial bias in the rodent trapping data, we calculated trap-night (TN) density within each West African level-2 administrative region.
+The `sf` package in the R statistical language was used to manipulate geographic data, administrative boundaries were obtained from GADM 4.0.4 [@pebesma_simple_2018; @r_core_team_r_2020; @database_of_global_administrative_areas_gadm_2022].
+Density ($\text{TN}_\text{density}$) was calculated by dividing the number of trap nights by the area of a level-2 administrative area ($\text{R}_\text{area}$).
 For studies not reporting trap nights imputation was used as previously described.
-Human population density was obtained for the closest year (2005) to the median year of trapping (2007) from Socioeconomic Data and Applications Center (SEDAC) gridded population of the world v4 at ~ 1km resolution [@socioeconomic_data_and_applications_center_gridded_2021].
+Human population density was obtained for the closest year (2005) to the median year of trapping (2007) from Socioeconomic Data and Applications Center (SEDAC) gridded population of the world v4 at ~ 1km resolution ($\text{P}_\text{density}$) [@socioeconomic_data_and_applications_center_gridded_2021].
 Median population density was then calculated for each level-2 administrative region.
 Land cover classification was obtained from the Copernicus climate change service at ~300m resolution [@european_space_agency_climate_change_initiative_land_2022].
-The proportion of cropland, shrubland, tree cover and urban land cover within a level-2 administrative region in 2005 was calculated.
+The proportion of cropland, shrubland, tree cover (${\psi}_\text{tree}$) and urban land cover (${\psi}_\text{urban}$) within a level-2 administrative region in 2005 was calculated.
 
-We investigated the role of these covariates on the spatial bias of TN density by comparing to a spatial model (equation 1), as follows:
+We investigated the association between relative trapping effort, measured as TN density, and the proportion of urban, cropland, tree cover and human population density using Generalised Additive Models (GAM) incorporating a spatial interaction term (longitude and latitude) (@pedersen_hierarchical_2019).
+The models were constructed in the `mgcv` package (@wood_generalized_2017).
+Selection of the most parsimonious model was based on Deviance explained and the Akaike Information Criterion for each model (Equations 1-5 below).
+Relative trapping effort was then predicted across West Africa using these covariates.
+We performed two sensitivity analyses, first, by removing sites with imputed trapping effort.
+Second rather than associating trap locations to level-2 administrative areas we assigned them to ~ 1km pixels.
 
 $$
-
-\begin{equation}
-\tag{1}
-\text{TN density} \sim \text{Tweedie((Longitude * Latitude))}
-
-\tag{2}
-\text{TN density} \sim \text{Tweedie((Population density) + (Longitude * Latitude))}
-
-\tag{3}
-\text{TN density} \sim \text{Tweedie((Population density) + (Region area) + (Longitude * Latitude))}
-
-\text{TN density} \sim \text{Tweedie((Proportion cropland) + (Proportion tree cover) + (Proportion urban) + (Longitude * Latitude))}
-
-\tag{5}
-\text{TN density} \sim \text{Tweedie((Population density) + (Region area) + (Proportion urban) + (Longitude * Latitude))}
-\end{equation}
+\text{TN}_\text{density} \sim \text{Tweedie}(X * Y)
+$$
+$$
+\text{TN}_\text{density} \sim \text{Tweedie}(\text{P}_\text{density} + (X * Y))
+$$
 
 $$
+\text{TN}_\text{density} \sim \text{Tweedie}(\text{P}_\text{density} + \text{R}_\text{area} + (X * Y))
+$$
 
-We formulated these models as a Generalised Additive Models (GAM) incorporating a spatial interaction term (longitude by latitude) [@pedersen_hierarchical_2019].
-These models investigated the association between TN density, proportion of urban, cropland, tree cover and human population density.
-The models were constructed in the *mgcv* package using the R programming language [@wood_generalized_2017; @r_core_team_r_2020].
-Selection of the most parsimonious model was based on Deviance explained and the Akaike information criterion for each model.
-We performed sensitivity analysis by removing sites with imputed trapping effort and performing a pixel-based analysis.
+$$
+\text{TN}_\text{density} \sim \text{Tweedie}(\text{P}_\text{density} + {\psi}_\text{tree} + {\psi}_\text{urban} + (X * Y))
+$$
+
+$$
+\text{TN}_\text{density} \sim \text{Tweedie}(\text{Pop}_\text{density} + \text{R}_\text{area} + {\psi}_\text{urban} + (X * Y))
+$$
 
@@
 
-Curated datasets (i.e., IUCN and GBIF) are commonly used in assessments of biodiversity in response to land use and climate change in addition to being used to explore zoonotic pathogen host distributions.
-To understand any biases introduced from these sources into these analyses we aimed to assess the concordance of curated rodent host distributions from IUCN and GBIF with observed rodent presence and absence from rodent trapping studies.
+We assessed the concordance of curated rodent host distributions from IUCN and GBIF with observed rodent presence and absence from rodent trapping studies for seven species with the most trap locations (*M.
+natalensis*, *R.
+rattus*, *M.
+erythroleucus*, *M.
+musculus*, *A.
+niloticus*, *Praomys daltoni* and *Cricetomys gambianus*).
 We obtained rodent species distribution maps as shapefiles from the IUCN red list and translated these to a ~20km resolution raster [@iucn_iucn_2021].
@@
 
-We assessed the taxonomic coverage of the trapped taxa and for the seven species with the most trap locations (*M.
-natalensis*, *R.
-rattus*, *M.
-erythroleucus*, *M.
-musculus*, *A.
-niloticus*, *Praomys daltoni* and *Cricetomys gambianus*) we mapped detection and non-detection, associating each location with a ~20km resolution raster.
-For these seven species we calculated the area of the IUCN expected range.
-We calculated the percentage of this range covered by presence detections in GBIF, this process was repeated for presence detections from rodent trapping studies.
-For both GBIF data and trapping studies we calculated the area of detections outside of the IUCN expected range.
-For rodent trapping studies only, we were able to calculate the area of non-detections within the IUCN expected area.
-Finally, we calculated the combined area of detection when combining GBIF and rodent trapping study data.
+For each of the seven species, we first calculated the area of the IUCN expected range, and then the percentage of this range covered by presence detections in GBIF, and from presence detections in the rodent trapping data.
+We then calculated the area of both types of detections outside of the IUCN expected range.
+For rodent trapping data, we calculated the area of non-detections within the IUCN expected area.
+Finally, we calculated the combined area of detection from both GBIF and rodent trapping data.
 
@@
 
-Studies investigating the risk of endemic zoonoses and novel pathogen emergence generally rely on consolidated datasets of host-pathogen associations.
-To examine the usefulness of rodent trapping studies as an additional source of data we compared identified host-pathogen asociations with a consolidated zoonoses dataset (CLOVER) [@gibb_mammal_2021].
+To examine the usefulness of rodent trapping studies as an additional source of data we compared identified host-pathogen associations from trapping studies investigating zoonoses with a consolidated zoonoses dataset (CLOVER) [@gibb_mammal_2021l @gibb_viralemergenceclover_2021].
 CLOVER is a synthesis of four host-pathogen datasets (GMPD2, EID2, HP3 and Shaw, 2020) and was released in 2021, it contains more than 25,000 host-pathogen associations for Bacteria, Viruses, Helminth, Protozoa and Fungi.
-An archived version of the CLOVER dataset was obtained and subset for host species trapped within included studies, spatial filtering of CLOVER for host-pathogen associations limited to the study region was not possible [@gibb_viralemergenceclover_2021].
+We compared the host-pathogen networks across the two datasets, where the CLOVER data was subset for host species present in the rodent trapping datta.
 
-We produced host-pathogen pairs from rodent trapping studies investigating zoonoses.
-For each host and pathogen, with both identified to species level, we compared our host-pathogen pair network with one produced using CLOVER, reporting concordance and discordance.
-We describe these networks and report differences in network composition.
-For host-pathogen pairs with assay results consistent with acute or prior infection we calculate the proportion positive and identify those absent from CLOVER.
+For host-pathogen pairs with assay results consistent with acute or prior infection, we calculated the proportion positive and identify those absent from CLOVER.
 We expand the analysis to host-pathogen pairs with pathogens identified to genus level in Supplementary Figure 4.
 
-### What is the spatial extent of pathogen testing within a hosts range?
+### What is the spatial extent of pathogen testing within host ranges?
 
-Sampling of zoonotic pathogens may also be biased within their hosts ranges, which may hinder inference across the entire range of the species.
-To quantify this, we focused on the five pathogen groups and species sampled at the most locations (Arenaviridae, Borreliaceae, Leptospiraceae, *Lassa mammarenavirus* and *Toxoplasma gondii*).
-We geolocated individuals of host species tested for these pathogens and calculated the area sampled.
-We adopt a similar approach to above by mapping sampled hosts to ~20km pixels and calculating the proportion of their IUCN range in which sampling has occurred.
-We also present the area covered of all areas in which the host has been detected in this study to produce a measure of relative completeness of sampling within the included rodent trapping studies.
+We use the sampled area of three pathogen groups and two pathogens (Arenaviridae, Borreliaceae, Leptospiraceae, *Lassa mammarenavirus* and *Toxoplasma gondii*) to quantify the bias of sampling within their hosts ranges.
+For each pathogen, we first describe the number of host species assated, for the five most commonly tested species we associate the locations of sampled indivuals to ~20km pixels and calculate the proportion of the IUCN range of the host in which sampling has occurred.
+We compare this figure to the total area in which the host has been detected to produce a measure of relative completeness of sampling within the included rodent trapping studies.
 
-Data and code implemented in the R statistical language has been made available alongside this manuscript to reproduce these analyses [ref_zenodo].
+All data and code to reproduce these analyses are available in an archived Zenodo repository [ref_zenodo].
 
@@
 
-Trapping sites were biased towards human modified landscapes (i.e., cropland, grassland and mosaic habitats), with under-representation of forest and bare habitats.
-However, urban habitats were comparatively undersampled (Supplementary Figure 2.).
-
-The most parsimonious GAM model (adjusted R2 = 0.3, Deviance explained = 48.7%) included smooth terms for geographic coordinates, human population density, proportion of urban landscape, and the area of the region.
-All smooth terms demonstrate significant non-linear associations (*p* < 0.002) with TN density (Supplementary Table 3.5).
-The spatial smooth term adjusted for population density, proportion of urban landscape and region area is shown in Figure 2.
-The final model identified relative trapping effort bias, with a bias towards increased trapping effort in North West Senegal, North and East Sierra Leone, Central Guinea and coastal regions of Nigeria, Benin and Ghana; in contrast South East Nigeria, Northern Nigeria and Burkina Faso had an observed bias towards a reduced trapping effort (Figure 2).
-Model summaries informing model selection are shown in Supplementary Tables 3.1-3.5.
-In sensitivity analysis, excluding sites with imputed trap nights reduced contributing data; however, Mauritania, Northern Senegal and Sierra Leone remained as regions trapped at higher rates, with Nigeria being trapped at lower than expected rates (Supplementary Figure 3a.).
-In pixel based sensitivity analysis spatial coverage was reduced with similar patterns of bias observed (Supplementary Figure 3b.).
+The most parsimonious GAM model (adjusted R2 = 0.3, Deviance explained = 48.7%) reported significant non-linear associations between relative trapping effort bias and human population densities (Effective Degrees of Freedom (EDF) = 7.13, *p* < 0.001), proportion of urban landscape (EDF = 1.92, *p* < 0.002) and region area (EDF = 3.63, *p* < 0.001), alongside significant spatial associations (EDF = 27.3, *p* < 0.001) (Supplementary table 3.1).
+Greatest trapping effort bias peaked at population densities between 5,000-7,500 individuals/km^2^, urban landscape >10% and region areas < 1,000km^2^.
+Increased trapping effort was found in North West Senegal, North and East Sierra Leone, Central Guinea and coastal regions of Nigeria, Benin and Ghana; in contrast South East Nigeria, Northern Nigeria and Burkina Faso had an observed bias towards a reduced trapping effort (Figure 2).
+In sensitivity analysis, excluding sites with imputed trap nights, Mauritania, Northern Senegal and Sierra Leone remained as regions trapped at higher rates, with Nigeria being trapped at lower than expected rates (Supplementary Figure 3a.).
+In pixel based sensitivity analysis spatial coverage was reduced with similar patterns of bias observed to the primary analysis (Supplementary Figure 3b.).
 
@@
 Uncertainty in the point estimate of relative trapping effort bias is represented by colour transparency/pallor (e.g., Eastern Liberia and Central Mali).
-Yellow regions represent areas with a bias towards increased trapping effort (e.g., North West Senegal), purple regions represent areas with a bias towards reduced trapping effort (e.g.,Northern Nigeria).](`r here("figures", "Figure_2_updated.png")`)
+Yellow regions represent areas with a bias towards increased trapping effort (e.g., North West Senegal), purple regions represent areas with a bias towards reduced trapping effort (e.g., Northern Nigeria).](`r here("figures", "Figure_2_updated.png")`)
 
@@
 
-We found that rodent trapping studies could provide a useful supplement to curated data describing rodent host distributions.
-For the seven rodent species of interest (*M.
+We found that for six of the seven rodent (*M.
 natalensis*, *R.
@@
 musculus*, *A.
-niloticus*, *P.
-daltoni* and *C.
-gambianus*), except *C.
-gambianus*, trapping studies provided more distinct locations of detection and non-detection than were available from GBIF.
+niloticus* and *P.
+daltoni*), trapping studies provided more distinct locations of detection and non-detection than were available from GBIF.
 For the endemic rodent species (*M.
@@
 rattus* and *M.
-musculus* were detected across much greater ranges than were expected from IUCN distributions (Figure 3A.
-and 3B.).
+musculus* were detected across much greater ranges than were expected from IUCN distributions.
+Comparisons for *M.
+natalensis*, *R.
+rattus* and *M.
+musculus* are shown in Figure 3A, the remaining species are shown in Supplementary Figure 4.
 
-![Figure 3A.
+![Figure 3.
 Each row corresponds to a single rodent species.
@@
 M.
-musculus has no IUCN West African range.](`r here("figures", "Figure_3_test_a.png")`)
+musculus has no IUCN West African range.](`r here("figures", "Figure_3_updated.png")`)
 
-![Figure 3B.
-Each row corresponds to a single rodent species.
-L) Presence recorded in GBIF overlaid on IUCN species range (red-shaded area).
-R) Detection (purple) and non-detection (orange) from rodent trapping studies overlaid on IUCN species ranges.](`r here("figures", "Figure_3_test_b.png")`)
-
-To further understand the sampling coverage of curated datasets and included rodent trapping studies we identified the proportion of a species IUCN range in which detections and non-detections occurred for the seven species most detected in the included studies (Table 1.).
-Apart from *M.
-musculus* these species had IUCN ranges greater than 1 million km2, with the Mastomys species having the largest geographic ranges.
-Sampling locations of these species within GBIF covered between 0.09-0.26% of expected ranges with detections occurring outside IUCN ranges for all species, most noticeably for *A.
+Comparison of proportion of a species IUCN range in which detections and non-detections occurred showed that sampling locations of these seven species within GBIF covered between 0.09-0.26% of expected ranges (Table 1.).
+Detections occurred outside IUCN ranges for all species, most noticeably for *A.
 niloticus* and *R.
 rattus*.
-Sampling locations within rodent trapping studies covered 0.03-0.24% of expected ranges with detections occurring outside IUCN ranges most noticeably for *A.
+Sampling locations within rodent trapping data covered 0.03-0.24% of expected ranges with detections occurring outside IUCN ranges most noticeably for *A.
 niloticus* and *R.
 rattus*.
-Combining GBIF and rodent trapping data increased the sampled area by a mean of 65% (SD 36.5%) , suggesting limited overlap between the locations sampled.
-Non-detection of species occurred across species ranges (mean = 0.11%, SD = 0.03%), suggestive of heterogeneity of presence across IUCN ranges.
+Combining GBIF and rodent trapping data increased the sampled area by a mean of 1.65 times, suggesting limited overlap between the locations sampled.
+Non-detection of species occurred across species ranges (mean = 0.11%, SD = 0.03%), suggestive of spatial heterogeneity of presence within IUCN ranges.
 
@@
 We identified negative associations (non-detection through specific assays) for 45 host-pathogen pairs among 35 rodent and 7 pathogen species.
-CLOVER contained 10 (40%) of our identified host-pathogen associations, the remaining 15 (60%) were not found to be present in CLOVER, additionally CLOVER recorded positive associations for 4 (9%) of the negative associations produced from the rodent trapping studies.
+CLOVER contained 10 (40%) of our identified host-pathogen associations, the remaining 15 (60%) were not found to be present in CLOVER, additionally CLOVER recorded positive associations for 4 (9%) of the negative associations produced from the rodent trapping data.
 
@@
 erythroleucus*.
-Assays were conducted on individuals trapped within less than a third of the locations in which the species were trapped (Table 2.).
+These species were assayed across between 10-20% of their trapped area, equating to ~0.02% of their IUCN range (Table 2.).
 
 Infection with species of Borreliaceae was assessed in 42 species, with evidence of infection in 17 rodent species.
-Among species with more than 500 samples the highest rates of infection were among *A.
+The greatest rates of infection were among *A.
 niloticus* (16%), *Mastomys huberti* (11%) and *M.
 erythroleucus* (9%).
-The coverage of sampling within the trapping locations was slightly greater than that for Arenaviridae but remains a small proportion (<1%) of the host species IUCN range.
-Infection with species of Leptospiraceae and *Toxoplasma gondii* was assessed in 8 species, with evidence of infection in 5 and 6 rodent species respectively.
-Testing for these pathogens was more limited with few individuals tested and reduced coverage of host ranges.
+Testing was more widespread than for Arenviruses with coverage between 15-34% of their trapped area, however, this remains a small area in relation to their IUCN ranges (<0.05%).
+Leptospiraceae and *Toxoplasma gondii* was assessed in 8 species, with evidence of infection in 5 and 6 rodent species respectively.
+The spatial coverage of testing for these pathogens was more limited within IUCN host species ranges (~0.01%).
 
@@
 table_2 <- read_rds(here("tables", "Table_2_updated_pixel.rds"))%>%
-  set_caption(caption = "Table 2: Comparison of pathogen sampling ranges for the 5 most widely sampled pathogens and the 5 most sampled potential rodent host species.
-* no IUCN range in West African") %>%
+  set_caption(caption = "Table 2: Comparison of pathogen sampling ranges for the 5 most widely sampled pathogens and the 5 most sampled rodent host species. * no IUCN range in West African") %>%
   autofit()
@@
 
-Endemic rodent zoonoses and novel pathogen emergence from rodent hosts are predicted to have an increasing burden in West Africa and globally (Han et al.
-2015).
-Here we have synthesised data from 126 rodent trapping studies containing information on more than 76,000 rodents across 1,611 trap sites producing an estimated 942,669 trap nights across 14 West African countries.
-Locations studied are complementary to curated datasets (e.g.
-IUCN, GBIF), incorporation of our synthesised dataset when assessing zoonosis risk based on host distributions could counteract some of the biases inherent to these curated datasets (Boakes et al., 2010).
+Endemic rodent zoonoses and novel pathogen emergence from rodent hosts are predicted to have an increasing burden in West Africa and globally (@han_rodent_2015).
+Here we have synthesised data from 126 rodent trapping studies containing information on more than 72,000 rodents across 1,611 trap sites producing an estimated 942,669 trap nights across 14 West African countries.
+Locations studie are complementary to curated datasets (e.g.
+IUCN, GBIF), incorporation of our synthesised dataset when assessing zoonosis risk based on host distributions could counteract some of the biases inherent to these curated datasets [@boakes_distorted_2010].
 We identified 25 host-pathogen pairs reported from included studies, 15 of these were not included in a consolidated host-pathogen dataset.
-However, pathogen sampling within identified species occurrence locations was limited, highlighting a further component of sampling bias supporting calls for further pathogen sampling across diverse species in zoonotic hotspots (Harvey and Holmes, 2022).
+Generally, the number of different species tested for a pathogen and the spatial extent of these sampling locations were limited.
+These findings highlight a further component of sampling bias, supporting calls for further pathogen sampling across diverse species in zoonotic hotspots [@harvey_diversity_2022].
 
-We found that rodent trapping studies, similar to biodiversity data, showed important spatial biases (Beck et al., 2014).
-Relative trapping effort was biased towards increased effort in Benin, Guinea, Senegal and Sierra Leone driven by long-standing research collaborations investigating the invasion of non-native rodent species (M.
-musculus and R.
-rattus) and the risk of endemic zoonosis outbreaks (e.g.
-Lassa mammarenavirus).
-Much of West Africa remains relatively understudied, particularly Burkina Faso, Côte d’Ivoire, Ghana and Nigeria, despite these countries facing many of the same challenges.
-For example, annual outbreaks of Lassa fever are reported in Nigeria and there are potentially 60,000 unrecognised cases of Lassa fever each year in Côte d’Ivoire and Ghana (Basinski et al., 2021).
-Rodent sampling should be targeted towards regions with minimal data to reduce this bias and assist in understanding the risk from endemic and novel zoonoses.
+We found that rodent trapping data, similar to biodiversity data, showed important spatial biases [@beck_spatial_2014].
+Relative trapping effort bias was greater in Benin, Guinea, Senegal and Sierra Leone driven by long-standing research collaborations investigating the invasion non-native rodent species (*M.
+musculus* and *R.
+rattus*) and the hazard of endemic zoonosis outbreaks (e.g., *Lassa mammarenavirus*).
+Much of the region remains relatively undersampled, particularly Burkina Faso, Côte d’Ivoire, Ghana and Nigeria, despite these countries facing many of the same challenges.
+For example, annual outbreaks of Lassa fever are reported in Nigeria and there are potentially 60,000 unrecognised cases of Lassa fever every year in Côte d’Ivoire and Ghana [@basinski_bridging_2021].
+Rodent sampling should be targeted towards currently under sampled regions to reduce this bias and improve our understanding of the risk from endemic and novel zoonoses.
 
 Rodent trapping studies provide geographic and temporally contextualised data on both species detection and non-detection which are not available from curated datasets.
-Non-detection data can improve models of species distributions, unfortunately, high levels of missing data on trapping effort will continue to confound the allocations of non-detections as true absences (Václavík and Meentemeyer, 2009).
-Models of host species occurrence and abundance, improved by incorporating species absence, are important to assess the effect of land use and climate change on endemic zoonosis spillover to human populations and direct limited public health resources towards regions at greatest risk (Plowright et al., 2019).
+Non-detection data can improve models of species distributions, unfortunately, high levels of missing data on trapping effort will continue to confound the allocations of non-detections as true absences [@vaclavik_invasive_2009].
+Models of host species occurrence and abundance, improved by incorporating species absence, are important to assess the effect of land use and climate change on endemic zoonosis spillover to human populations and direct limited public health resources towards regions at greatest risk [@zeimes_modelling_2012; @judson_translating_2018].
 
 Currently available consolidated datasets on host-pathogen associations (e.g.
-CLOVER, EID2 and GMPD2) do not include spatial or temporal components (Gibb, Gregory F.
-Albery, et al., 2021a).
+CLOVER, EID2 and GMPD2) do not include spatial or temporal components [@gibb_data_2021].
 The current synthesis of rodent trapping studies has highlighted that pathogens have been sparsely sampled within a hosts range.
 Current zoonosis risk models dependent on these sources of data are therefore not able to incorporate spatial heterogeneity in pathogen prevalence across the host range.
-Additional uncertainty in current models of zoonotic disease risk arises from observed host-pathogen associations that have not been reported in these consolidated datasets.
-For example, Hylomyscus pamfi infected with Lassa mammarenavirus and R.
-rattus infected with Coxiella burnetiid, will not be included when solely dependent on consolidated host-pathogen datasets.
-Further, detection of zoonotic pathogens in multiple co-occurring host species supports the adoption of multi-species to better understand the potential range of endemic zoonoses (Wilkinson et al., 2019).
+Additional uncertainty in current models of zoonotic disease risk arises from host-pathogen associations that have not been reported in these consolidated datasets.
+For example *Hylomyscus pamfi* infected with *Lassa mammarenavirus* and *R.
+rattus* infected with *Coxiella burnetii*, will not be included when solely dependent consolidated host-pathogen datasets.
+Further, detection of zoonotic pathogens in multiple co-occurring host species supports the adoption of multi-species to better understand the potential range of endemic zoonoses [@wilkinson_comparison_2019].
 
 Few studies stratified detection and non-detection of hosts or pathogen prevalence by time, therefore limiting inference of changes in host and pathogen dynamics.
-This limitation prevents calculation of incidence of infection and the abundance of infectious rodents which potentially varies by both time and space (Fichet-Calvet et al., 2016).
+This limitation prevents calculation of incidence of infection and the abundance of infectious rodents which potentially varies by both time and space [@fichet-calvet_spatial_2016].
 Understanding of temporal changes in viral burden and shedding for endemic zoonoses is required to accurately predict current and future risk of pathogen spillover.
@@
 Finally, due to data sparsity, we were unable to account for temporal change over the six decades of rodent trapping studies.
-Land use change and population density have changed dramatically over this period in West Africa (Herrmann et al., 2020).
-We attempted to mitigate against this by using the median year of trapping as a reference point to understand the spatial and land use biases in trapping activity.
-It is possible that land use and population density at trapping sites varied importantly between when the rodent trapping study was conducted and the land use and human population variables in 2005.
-Despite this limitation, the finding that trapping is biased towards high density, human dominated landscapes is unlikely to substantially change (Beck et al., 2014).
+Land use change and population density have changed dramatically over this period in West Africa [@herrmann_accelerating_2020].
+We attempted to mitigate against this by using the median year of trapping to understand the spatial and land use biases in trapping activity.
+It is possible that land use and population density at trapping sites varied importantly between when rodent trapping was conducted and the conditions in 2005.
+Despite this limitation, the finding that trapping is biased towards high density, human dominated landscapes is unlikely to substantially change.
 
@@
 Together this supports their inclusion in efforts to model endemic zoonotic risk and novel pathogen emergence.
-Contribution of rodent trapping studies as data sources can be improved by adopting reporting standards and practices consistent with Open Science, namely sharing of disaggregated datasets alongside publication (Foster and Deardorff, 2017).
+Contribution of rodent trapping studies as data sources can be improved by adopting reporting standards and practices consistent with Open Science, namely sharing of disaggregated datasets alongside publication [@foster_open_2017].
 

### 2022-08-23 reports/manuscript_v5.Rmd
push to write cover letter

@@
 
-To examine the usefulness of rodent trapping studies as an additional source of data we compared identified host-pathogen associations from trapping studies investigating zoonoses with a consolidated zoonoses dataset (CLOVER) [@gibb_mammal_2021l @gibb_viralemergenceclover_2021].
+To examine the usefulness of rodent trapping studies as an additional source of data we compared identified host-pathogen associations from trapping studies investigating zoonoses with a consolidated zoonoses dataset (CLOVER) [@gibb_mammal_2021; @gibb_viralemergenceclover_2021].
 CLOVER is a synthesis of four host-pathogen datasets (GMPD2, EID2, HP3 and Shaw, 2020) and was released in 2021, it contains more than 25,000 host-pathogen associations for Bacteria, Viruses, Helminth, Protozoa and Fungi.

### 2022-08-24 reports/manuscript_v5.Rmd
updated figures and tables

@@
 The earliest trapping studies were conducted in 1964, with a trend of increasing numbers of studies being performed annually since 2000.
-The median year of first trapping activity was 2007, with the median length of trapping activity being 1 year (IQR 0-2 years) (Supplementary Figure 1.).
-Studies were conducted in 14 West African countries, with no studies reported from The Gambia or Togo, at 1,611 trap sites (Figure 1A.).
+The median year of first trapping activity was 2007, with the median length of trapping activity being 1 year (IQR 0-2 years) (Supplementary Fig 1.).
+Studies were conducted in 14 West African countries, with no studies reported from The Gambia or Togo, at 1,611 trap sites (Fig 1A.).
 
-Included studies explicitly reported on 601,184 trap nights, a further 341,445 trap nights were imputed from studies with no recording of trapping effort based on trap success, leading to an estimate of 942,629 trap nights (Figure 1B.).
+Included studies explicitly reported on 601,184 trap nights, a further 341,445 trap nights were imputed from studies with no recording of trapping effort based on trap success, leading to an estimate of 942,629 trap nights (Fig 1B.).
 A minority of studies trapped at a single study site (30, 24%), with 46 (36%) trapping at between two and five sites, the remaining 51 studies (40%) trapped at between six and 93 study sites.
@@
 
-![Figure 1: A) The location of trapping sites in West Africa.
+![Fig 1: A) The location of trapping sites in West Africa.
 No sites were recorded from Togo or The Gambia.
@@
 For example, Senegal, Mali and Sierra Leone have generally good coverage compared to Guinea and Burkina Faso.
-B) Histogram of trap nights performed at each study site, a median of 248 trap nights (IQR 116-500) was performed at each site](`r here("figures", "Figure_1_updated.png")`)
+B) Histogram of trap nights performed at each study site, a median of 248 trap nights (IQR 116-500) was performed at each site](`r here("figures", "Figure_1_updated.pdf")`)
 
@@
 For host-pathogen pairs with assay results consistent with acute or prior infection, we calculated the proportion positive and identify those absent from CLOVER.
-We expand the analysis to host-pathogen pairs with pathogens identified to genus level in Supplementary Figure 4.
+We expand the analysis to host-pathogen pairs with pathogens identified to genus level in Supplementary Fig 4.
 
@@
 The regions with highest TN density included the capitals and large cities of Niger (Niamey), Nigeria (Ibadan), Ghana (Accra), Senegal (Dakar), Ghana (Accra) and Benin (Cotonou).
-Outside of these cities, regions in, Northern Senegal, Southern Guinea, Edo and Ogun States in Nigeria and Eastern Sierra Leone had the greatest TN density (Figure 1A.).
+Outside of these cities, regions in, Northern Senegal, Southern Guinea, Edo and Ogun States in Nigeria and Eastern Sierra Leone had the greatest TN density (Fig 1A.).
 
@@
 Greatest trapping effort bias peaked at population densities between 5,000-7,500 individuals/km^2^, urban landscape >10% and region areas < 1,000km^2^.
-Increased trapping effort was found in North West Senegal, North and East Sierra Leone, Central Guinea and coastal regions of Nigeria, Benin and Ghana; in contrast South East Nigeria, Northern Nigeria and Burkina Faso had an observed bias towards a reduced trapping effort (Figure 2).
-In sensitivity analysis, excluding sites with imputed trap nights, Mauritania, Northern Senegal and Sierra Leone remained as regions trapped at higher rates, with Nigeria being trapped at lower than expected rates (Supplementary Figure 3a.).
-In pixel based sensitivity analysis spatial coverage was reduced with similar patterns of bias observed to the primary analysis (Supplementary Figure 3b.).
+Increased trapping effort was found in North West Senegal, North and East Sierra Leone, Central Guinea and coastal regions of Nigeria, Benin and Ghana; in contrast South East Nigeria, Northern Nigeria and Burkina Faso had an observed bias towards a reduced trapping effort (Fig 2).
+In sensitivity analysis, excluding sites with imputed trap nights, Mauritania, Northern Senegal and Sierra Leone remained as regions trapped at higher rates, with Nigeria being trapped at lower than expected rates (Supplementary Fig 3a.).
+In pixel based sensitivity analysis spatial coverage was reduced with similar patterns of bias observed to the primary analysis (Supplementary Fig 3b.).
 
-![Figure 2.
+![Fig 2.
 Relative trapping effort bias across West Africa from included studies adjusted for human population density, proportion urban land cover and area of the administrative region.
-Uncertainty in the point estimate of relative trapping effort bias is represented by colour transparency/pallor (e.g., Eastern Liberia and Central Mali).
-Yellow regions represent areas with a bias towards increased trapping effort (e.g., North West Senegal), purple regions represent areas with a bias towards reduced trapping effort (e.g., Northern Nigeria).](`r here("figures", "Figure_2_updated.png")`)
+Yellow regions represent areas with a bias towards increased trapping effort (e.g., North West Senegal), purple regions represent areas with a bias towards reduced trapping effort (e.g., Northern Nigeria).](`r here("figures", "Figure_2_updated.pdf")`)
 
@@
 rattus* and *M.
-musculus* are shown in Figure 3A, the remaining species are shown in Supplementary Figure 4.
+musculus* are shown in Fig 3, the remaining species are shown in Supplementary Fig 4.
 
-![Figure 3.
+![Fig 3.
 Each row corresponds to a single rodent species.
@@
 M.
-musculus has no IUCN West African range.](`r here("figures", "Figure_3_updated.png")`)
+musculus has no IUCN West African range.](`r here("figures", "Figure_3_updated.pdf")`)
 
@@
 We found potentially important differences between the host-pathogen networks produced from included rodent trapping studies and the consolidated CLOVER dataset.
-When limited to taxonomic classification of both pathogen and host to species level we identified 25 host-pathogen pairs among 14 rodent and 6 pathogen species (Figure 4.).
+When limited to taxonomic classification of both pathogen and host to species level we identified 25 host-pathogen pairs among 14 rodent and 6 pathogen species (Fig 4.).
 We identified negative associations (non-detection through specific assays) for 45 host-pathogen pairs among 35 rodent and 7 pathogen species.
@@
 
-![](`r here("figures", "Figure_4a_updated.png")`)
-
-![Figure 4.
+![Fig 4.
 A) Identified species level host-pathogen associations through detection of acute infection (i.e.
 PCR, culture).
-Percentages and colour relate to the proportion of all assays that were positive.
-Associations with a black border are present in the CLOVER dataset.
+Percentages and colour relate to the proportion of all assays that were positive, the number of individuals tested for the pathogen is labelled N.
+Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Figure_4a_updated.pdf")`)
+
+![Fig 4.
 B) Identified species level host-pathogen associations through serological assays (i.e.
 ELISA).
-Percentages and colour relate to the proportion of all assays that were positive.
-Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Figure_4b_updated.png")`)
+Percentages and colour relate to the proportion of all assays that were positive, the number of individuals tested for the pathogen is labelled N.
+Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Figure_4b_updated.pdf")`)
 
@@
 For pathogens not identified to species level (i.e.
-family or higher taxa only), we identified 148 host-pathogen pairs among 32 rodent species and 25 pathogen families (Supplementary Figure 4.), with CLOVER containing 66 (45%) of these associations.
+family or higher taxa only), we identified 148 host-pathogen pairs among 32 rodent species and 25 pathogen families (Supplementary Fig 4.), with CLOVER containing 66 (45%) of these associations.
 
@@
 
-read_rds(here("tables", "supplementary_table_1.rds")) %>%
+read_rds(here("tables", "Supplementary_Table_1.rds")) %>%
   set_caption(caption = "Supplementary Table 1: Data extraction tool for studies meeting inclusion criteria")
@@
 
-read_rds(here("tables", "supplementary_table_2.rds"))
+read_rds(here("tables", "Supplementary_Table_2.rds"))
 
@@
 
-## Supplementary Figure 1
+## Supplementary Fig 1
 
-![Supplementary Figure 1.
+![Supplementary Fig 1.
 Green points represent the start date of rodent trapping studies, blue points representing the final trapping activity.
 Red points indicate the publication of studies.
-Increasing numbers of studies have been published since 2000 with more studies being conducted over repeated visits.](`r here("figures", "Supplementary_figure_1.png")`)
+Increasing numbers of studies have been published since 2000 with more studies being conducted over repeated visits.](`r here("figures", "Supplementary_Figure_1.pdf")`)
 
-## Supplementary Figure 2
+## Supplementary Fig 2
 
-![Supplementary Figure 2.
-Comparison between proportion of land use between regions trapped and all land use in West Africa.
-Cropland, Shrubland and Grassland were trapped at than representative rates while Bare habitats, Forested regions and Urban landscapes were trapped at lower than representative rates.](`r here("figures", "Supplementary_figure_2.png")`)
+![Supplementary Fig 2.](`r here("figures", "Supplementary_Figure_2.pdf")`)
 
-## Supplementary Figure 3
+## Supplementary Fig 3
 
-![Supplementary Figure 3A.
+![Supplementary Fig 3A.
 Relative trapping effort bias across West Africa from the subset of included studies reporting trapping effort adjusted for proportion urban land cover and proportion tree cover.
@@
 
-## Supplementary Figure 4
-![](`r here("figures", "Supp_4a_updated.png")`)
+## Supplementary Fig 4
+![](`r here("figures", "Supplementary_Figure_4.pdf")`)
 
-![Supplementary Figure 4.
+![Supplementary Fig 4.
 A) Identified host-pathogen associations at pathogen family level through detection of acute infection (i.e.

### 2022-08-24 scripts/3_2_rodent_host_distributions.R

@@ -180,31 +180,50 @@ plot_fig_3 <- function(species_name, iucn = rodent_iucn, gbif = rodent_gbif, rod
               legend = legend))
 }
 
-produced_plots <- list()
+# Three species will be shown in the main manuscript
+fig_3_species <- c("mastomys natalensis", "rattus rattus", "mus musculus")
+
+fig_3_plots <- list()
 legend <- list()
 
-for(i in 1:length(species_names)) {
+for(i in 1:length(fig_3_species)) {
 
   output <- plot_fig_3(species_names[[i]])
 
-  produced_plots[[i]] <- output[["combined_plot"]]
+  fig_3_plots[[i]] <- output[["combined_plot"]]
 
   legend <- output[["legend"]]
 
 }
 
-names(produced_plots) <- species_names
+fig_3_plots_grid <- plot_grid(fig_3_plots[[1]], fig_3_plots[[2]], fig_3_plots[[3]], legend,
+                              ncol = 1, labels = c("Mastomys natalensis", "Rattus rattus", "Mus musculus", ""),
+                              hjust = c(-0.1, -0.2, -0.165),
+                              label_size = 12, rel_heights = c(1, 1, 1, 0.4))
 
-test_plots_a <- plot_grid(produced_plots[[1]], produced_plots[[2]], produced_plots[[3]], produced_plots[[4]],
-                          legend, ncol = 1, labels = c("Mastomys natalensis", "Rattus rattus", "Mastomys erythroleucus", "Mus musculus", ""),
-                          hjust = c(-0.2, -0.32, -0.17, -0.29), label_size = 12, rel_heights = c(1, 1, 1, 1, 0.6))
-test_plots_b <- plot_grid(produced_plots[[5]], produced_plots[[6]], produced_plots[[7]],
-                          legend, ncol = 1, labels = c("Arvicanthis niloticus", "Praomys daltoni", "Cricetomys gambianus", ""),
-                          hjust = c(-0.2, -0.25, -0.165), label_size = 12, rel_heights = c(1, 1, 1, 0.6))
+save_plot(plot = fig_3_plots_grid, filename = here("figures", "Figure_3_updated.pdf"), base_height = 12, base_width = 8)
 
-save_plot(plot = test_plots_a, filename = here("figures", "Figure_3_a_updated.png"), base_height = 12, base_width = 8)
[…]

### 2022-08-24 scripts/3_3_host_pathogen_associations.R

@@ -306,37 +306,37 @@ plot_4_df$species <- fct_relevel(plot_4_df$species, levels(sum_species))
 plot_4_acute <- plot_4_df %>%
   filter(tested > 1) %>%
   filter(class == "Acute infection") %>%
-  mutate(lab = paste0(round(value, 1), "% \n N = ", tested)) %>%
+  mutate(lab = paste0(round(value, 1), "%, N = ", tested)) %>%
   ggplot() +
-  geom_tile(aes(x = pathogen_name, y = species, fill = value, colour = source, width = 0.9, height = 0.9), lwd = 1) +
+  geom_tile(aes(x = pathogen_name, y = species, fill = value, colour = source, width = 0.9, height = 0.9), lwd = 1.2) +
   geom_label(aes(x = pathogen_name, y = species, label = lab), size = 3.7) +
-  scale_fill_viridis_c(option = "viridis", direction = -1, begin = 0.3, end = 1) +
+  scale_fill_viridis_c(option = "viridis", direction = -1, begin = 0.3, end = 1, limits = c(0, 50)) +
   scale_colour_manual(na.translate = FALSE, values = "black") +
   labs(title = "Acute infection",
        fill = "Infection (%)",
        x = element_blank(),
        y = element_blank(),
        colour = element_blank()) +
-  theme_minimal()
+  theme_bw()
 
 plot_4_serology <- plot_4_df %>%
   filter(tested > 1) %>%
   filter(class == "Serology") %>%
-  mutate(lab = paste0(round(value, 1), "% N = ", tested)) %>%
+  mutate(lab = paste0(round(value, 1), "%, N = ", tested)) %>%
   ggplot() +
-  geom_tile(aes(x = pathogen_name, y = species, fill = value, colour = source, width = 0.9, height = 0.9), lwd = 1) +
+  geom_tile(aes(x = pathogen_name, y = species, fill = value, colour = source, width = 0.9, height = 0.9), lwd = 1.2) +
   geom_label(aes(x = pathogen_name, y = species, label = lab)) +
-  scale_fill_viridis_c(option = "viridis", direction = -1, begin = 0.3, end = 1) +
+  scale_fill_viridis_c(option = "viridis", direction = -1, begin = 0.3, end = 1, limits = c(0, 20)) +
   scale_colour_manual(na.translate = FALSE, values = "black") +
   labs(title = "Serology",
        fill = "Infection (%)",
        x = element_blank(),
        y = element_blank(),
        colour = element_blank()) +
[…]

### 2022-08-25 reports/manuscript_v5.Rmd
version for archiving

@@
 bibliography: citations/manuscript_citations.bib
-csl: harvard.csl
+csl: plos.csl
 latex_engine: xelatex
@@
 For example, Senegal, Mali and Sierra Leone have generally good coverage compared to Guinea and Burkina Faso.
-B) Histogram of trap nights performed at each study site, a median of 248 trap nights (IQR 116-500) was performed at each site](`r here("figures", "Figure_1_updated.pdf")`)
+B) Histogram of trap nights performed at each study site, a median of 248 trap nights (IQR 116-500) was performed at each site](`r here("figures", "Figure_1_updated.png")`)
 
@@
 To investigate the extent of spatial bias in the rodent trapping data, we calculated trap-night (TN) density within each West African level-2 administrative region.
-The `sf` package in the R statistical language was used to manipulate geographic data, administrative boundaries were obtained from GADM 4.0.4 [@pebesma_simple_2018; @r_core_team_r_2020; @database_of_global_administrative_areas_gadm_2022].
+The `sf` package in the R statistical language (R version 4.1.2) was used to manipulate geographic data, administrative boundaries were obtained from GADM 4.0.4 [@pebesma_simple_2018; @r_core_team_r_2020; @database_of_global_administrative_areas_gadm_2022].
 Density ($\text{TN}_\text{density}$) was calculated by dividing the number of trap nights by the area of a level-2 administrative area ($\text{R}_\text{area}$).
@@
 $$
-\text{TN}_\text{density} \sim \text{Tweedie}(\text{Pop}_\text{density} + \text{R}_\text{area} + {\psi}_\text{urban} + (X * Y))
+\text{TN}_\text{density} \sim \text{Tweedie}(\text{P}_\text{density} + \text{R}_\text{area} + {\psi}_\text{urban} + (X * Y))
 $$
@@
 
-We found a non-random spatial distribution of rodent trapping locations across the study region, suggestive of underlying bias in the sampling or rodents across West Africa.
+We found non-random spatial clustering of rodent trapping locations across the study region, suggestive of underlying bias in the sampling or rodents across West Africa.
 Trap sites were situated in 256 of 1,450 (17.6%) level-2 administrative regions in 14 West African nations.
@@
 Relative trapping effort bias across West Africa from included studies adjusted for human population density, proportion urban land cover and area of the administrative region.
-Yellow regions represent areas with a bias towards increased trapping effort (e.g., North West Senegal), purple regions represent areas with a bias towards reduced trapping effort (e.g., Northern Nigeria).](`r here("figures", "Figure_2_updated.pdf")`)
+Yellow regions represent areas with a bias towards increased trapping effort (e.g., North West Senegal), purple regions represent areas with a bias towards reduced trapping effort (e.g., Northern Nigeria).](`r here("figures", "Figure_2_updated.png")`)
 
@@
 M.
-musculus has no IUCN West African range.](`r here("figures", "Figure_3_updated.pdf")`)
+musculus has no IUCN West African range.](`r here("figures", "Figure_3_updated.png")`)
 
@@
 
-![Fig 4.
-A) Identified species level host-pathogen associations through detection of acute infection (i.e.
+![Fig 4) Identified species level host-pathogen associations through detection of acute infection (i.e.
 PCR, culture).
 Percentages and colour relate to the proportion of all assays that were positive, the number of individuals tested for the pathogen is labelled N.
-Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Figure_4a_updated.pdf")`)
+Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Figure_4_updated.png")`)
 
-![Fig 4.
-B) Identified species level host-pathogen associations through serological assays (i.e.
+![Fig 5) Identified species level host-pathogen associations through serological assays (i.e.
 ELISA).
 Percentages and colour relate to the proportion of all assays that were positive, the number of individuals tested for the pathogen is labelled N.
-Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Figure_4b_updated.pdf")`)
+Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Figure_5_updated.png")`)
 
@@
 Red points indicate the publication of studies.
-Increasing numbers of studies have been published since 2000 with more studies being conducted over repeated visits.](`r here("figures", "Supplementary_Figure_1.pdf")`)
+Increasing numbers of studies have been published since 2000 with more studies being conducted over repeated visits.](`r here("figures", "Supplementary_Figure_1.png")`)
 
@@
 
-![Supplementary Fig 2.](`r here("figures", "Supplementary_Figure_2.pdf")`)
+![Supplementary Fig 2.](`r here("figures", "Supplementary_Figure_2.png")`)
 
@@
 Predictions are limited to areas around trap sites (coloured areas), uncertainty in the estimate of bias is represented by colour transparency/pallor.
-Purple regions represent areas with higher than expected trapping effort, yellow regions represent areas lower than expected trapping effort](`r here("figures", "Figure_2_sensitivity.png")`)
-
-![](`r here("figures", "Figure_2_ss.png")`)
+Purple regions represent areas with higher than expected trapping effort, yellow regions represent areas lower than expected trapping effort](`r here("figures", "Supplementary_Figure_3.png")`)
 
 ## Supplementary Fig 4
-![](`r here("figures", "Supplementary_Figure_4.pdf")`)
+![](`r here("figures", "Supplementary_Figure_4a.png")`)
 
@@
 Percentages and colour relate to the proportion of all assays that were positive.
-Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Supp_4b_updated.png")`)
+Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Supplementary_Figure_4b.png")`)

### 2022-08-25 scripts/3_1_spatial_bias_analysis.R

@@ -397,6 +397,8 @@ fig_2_updated <- model_1_raster +
 
 save_plot(plot = as.grob(fig_2_updated$ggObj),
           filename = here("figures", "Figure_2_updated.pdf"), base_height = 10, base_width = 12)
+save_plot(plot = as.grob(fig_2_updated$ggObj),
+          filename = here("figures", "Figure_2_updated.png"), base_height = 10, base_width = 12)
 
 
 # Supplementary figure model sensitivity ----------------------------------
@@ -426,3 +428,5 @@ supplementary_fig_2_updated <- model_1_s_raster +
 
 save_plot(plot = as.grob(supplementary_fig_2_updated$ggObj),
           filename = here("figures", "Supplementary_Figure_2.pdf"), base_height = 10, base_width = 12)
+save_plot(plot = as.grob(supplementary_fig_2_updated$ggObj),
+          filename = here("figures", "Supplementary_Figure_2.png"), base_height = 10, base_width = 12)

### 2022-08-25 scripts/3_1b_spatial_bias_analysis.R

@@ -163,7 +163,8 @@ rect_border <- tibble(x = c(-17.64, -17.64, 15.995642, 15.995642),
                       y = c(4.2, 27.683098, 4.2, 27.683098)) %>%
   st_as_sf(coords = c("x", "y"), crs = crs(included_countries)) %>%
   summarise() %>%
-  st_cast("POLYGON")
+  st_cast("POLYGON") %>%
+  st_buffer(dist = 10)
 
 country_border <- included_countries %>%
   summarise()
@@ -203,3 +204,5 @@ supplementary_fig_3_updated <- model_1_p_raster +
 
 save_plot(plot = as.grob(supplementary_fig_3_updated$ggObj),
           filename = here("figures", "Supplementary_Figure_3.pdf"), dpi = 320, base_height = 10, base_width = 12)
+save_plot(plot = as.grob(supplementary_fig_3_updated$ggObj),
+          filename = here("figures", "Supplementary_Figure_3.png"), dpi = 320, base_height = 10, base_width = 12)

### 2022-08-25 scripts/3_2_rodent_host_distributions.R

@@ -188,7 +188,7 @@ legend <- list()
 
 for(i in 1:length(fig_3_species)) {
 
-  output <- plot_fig_3(species_names[[i]])
+  output <- plot_fig_3(fig_3_species[[i]])
 
   fig_3_plots[[i]] <- output[["combined_plot"]]
 
@@ -202,6 +202,7 @@ fig_3_plots_grid <- plot_grid(fig_3_plots[[1]], fig_3_plots[[2]], fig_3_plots[[3
                               label_size = 12, rel_heights = c(1, 1, 1, 0.4))
 
 save_plot(plot = fig_3_plots_grid, filename = here("figures", "Figure_3_updated.pdf"), base_height = 12, base_width = 8)
+save_plot(plot = fig_3_plots_grid, filename = here("figures", "Figure_3_updated.png"), base_height = 12, base_width = 8)
 
 # Remaining species will go as Supplementary
 s_fig_4_species <- species_names[!species_names %in% fig_3_species]
@@ -224,6 +225,7 @@ s_fig_4_plots_grid <- plot_grid(s_fig_4_plots[[1]], s_fig_4_plots[[2]], s_fig_4_
                           hjust = c(-0.1, -0.14, -0.165, -0.1), label_size = 12, rel_heights = c(1, 1, 1, 1, 0.4))
 
 save_plot(plot = s_fig_4_plots_grid, filename = here("figures", "Supplementary_Figure_4.pdf"), base_height = 12, base_width = 8)
+save_plot(plot = s_fig_4_plots_grid, filename = here("figures", "Supplementary_Figure_4.png"), base_height = 12, base_width = 8)
 
 # Proportion of range trapped ---------------------------------------------
 # Reference raster from pop_density

### 2022-09-18 reports/manuscript_v5.Rmd
Update for move

@@
 We extracted data from eligible studies using a standardised tool that was piloted on 5 studies (Supplementary Table 1.).
-Data was abstracted into a Google Sheets document, which was archived on completion of data extraction [ref to zenodo on finalised].
+Data was abstracted into a Google Sheets document, which was archived on completion of data extraction [@david_simons_2022_7022637].
 We identified the aims of included studies, for example, whether it was conducted as a survey of small mammal species or specifically to assess the risk of zoonotic disease spillover.
@@
 
-All data and code to reproduce these analyses are available in an archived Zenodo repository [ref_zenodo].
+All data and code to reproduce these analyses are available in an archived Zenodo repository [@david_simons_2022_7022637].
 

### 2022-10-31 reports/manuscript_v5.Rmd
update following peer review

@@
 title: "Rodent trapping studies as an overlooked information source for understanding endemic and novel zoonotic spillover."
+subtitle: "Rodent trapping to understand zoonotic spillover risk"
 date: "`r Sys.Date()`"
@@
 These omissions have the potential for biasing estimates of current risk and drivers of zoonoses.
-Finally, for these five pathogen groups, we find that the proportion of a rodent hosts range that has been sampled remains small.
-A priority of future rodent trapping studies should be to sample rodent hosts across a greater geographic range to better characterise current and future risk.
+Finally, for these five pathogen groups, we find that the proportion of a rodent hosts range that has been sampled remains small with geographic clustering.
+A priority of future rodent trapping studies should be to sample rodent hosts across a greater geographic range to better characterise current and future risk of zoonotic spillover events.
 In the interim, future studies of spatial pathogen risk informed by rodent distributions must incorporate a measure of the current sampling biases.
@@
 
+# Author Summary
+
+Emerging and endemic zoonotic diseases are projected to have increasing health impacts, particularly under changing climate and land-use scenarios.
+Rodents carry a disproportionate number of zoonotic pathogens and are abundant across West Africa, raising concerns about increasing outbreaks of endemic disease and the emergence of pandemics.
+Prior modelling studies rely on large, consolidated data sources which do not incorporate high resolution spatial and temporal data from rodent trapping studies.
+Here, we synthesise these studies to quantify the bias in the sampling of rodent hosts and their pathogens across West Africa.
+We find that rodent trapping studies are complementary to these datasets and can provide additional, high-resolution data on the distribution of hosts and their pathogens.
+Further, rodent trapping studies have identified additional potential host-pathogen associations than are recorded in consolidated host-pathogen association datasets.
+This can help to understand the risk of zoonotic diseases based on host distributions.
+Finally, we quantify the current extent of known rodent presence and pathogen sampling within a species range, highlighting that current knowledge is limited across much of the region.
+We hope that this will support work to study rodent hosts and their pathogens in currently under sampled regions to better understand the risk of emerging and endemic zoonoses in West Africa.
+
 # Introduction
 
-There is increasing awareness of the global health and economic impacts of novel zoonotic pathogen spillover, evidenced by the ongoing SARS-CoV-2 pandemic and previous HIV/AIDs and Spanish Influenza pandemics [@bernstein_costs_2022].
+There is increasing awareness of the global health and economic impacts of novel zoonotic pathogen spillover, driven by the ongoing SARS-CoV-2 pandemic and previous HIV/AIDs and Spanish Influenza pandemics [@bernstein_costs_2022].
 The number of zoonotic disease spillover events and the frequency of the emergence of novel zoonotic pathogens from rodents are predicted to increase under intensifying anthropogenic pressure driven by increased human populations, urbanisation, intensification of agriculture, climate change and wildlife defaunation [@allen_global_2017; @hassell_urbanization_2017; @garcia-pena_land-use_2021; @mcmahon_ecosystem_2018].
@@
 
-Rodents along with bats contribute the greatest number of predicted novel zoonotic pathogens and known zoonoses [@han_rodent_2015; @gibb_mammal_2021].
+Rodents along with bats contribute the greatest number of predicted novel zoonotic pathogens and known endemic zoonoses [@han_rodent_2015; @gibb_mammal_2021].
 Of 2,220 extant rodent species, 244 (10.7%) are described as reservoirs of 85 zoonotic pathogens [@han_rodent_2015], although many species provide important and beneficial ecosystem services including pest regulation and seed dispersal [@fischer_ecosystem_2018].
@@
 These sampling biases can importantly distort produced species distribution models [@beck_spatial_2014].
-Datasets on host-pathogen associations also can suffer from biases introduced from literature selection criteria and taxanomic discrepancies.
-These biases are important because identification of potential geographic hotspots of zoonotic disease spillover and novel pathogen emergence are often based on these types of host species distributions and host-pathogen associations [@plowright_sampling_2019; @carlson_future_2021].
-For example, systematically increased sampling, over-representation of certain habitats and clustering around areas of high human population could lead to an apparent association between locations and hazard that is driven by these factors rather than underlying host-pathogen associations [@wille_how_2021; @redding_evaluating_2017; @gibb_mammal_2021].
+Datasets on host-pathogen associations also can suffer from biases introduced from literature selection criteria and taxonomic discrepancies.
+These biases are important because identification of potential geographic hotspots of zoonotic disease spillover and novel pathogen emergence are often produced from these types of host species distributions and host-pathogen associations [@plowright_sampling_2019; @carlson_future_2021].For example, systematically increased sampling, over-representation of certain habitats and clustering around areas of high human population could lead to an apparent association between locations and hazard that is driven by these factors rather than underlying host-pathogen associations [@wille_how_2021; @redding_evaluating_2017; @gibb_mammal_2021].
 Predictions of zoonotic disease spillover and novel zoonotic pathogen emergence must account for these biases to understand the future hazard of zoonotic diseases [@carlson_future_2021].
@@
 We also recorded how species were identified within a study and species identification was assumed to be accurate.
-The number of individuals of these species or genera was extracted with taxanomic names mapped to GBIF taxonomy [@gbif_the_global_biodiversity_information_facility_global_2021].
+The number of individuals of these species or genera was extracted with taxanomic names mapped to GBIF taxonomy[@gbif_the_global_biodiversity_information_facility_global_2021].
 We expanded species detection and non-detection records by explicitly specifying non-detection at a trap site if a species was recorded as detected at other trapping locations within the same study.
@@
 ![Fig 1: A) The location of trapping sites in West Africa.
+A) The location of trapping sites in West Africa.
 No sites were recorded from Togo or The Gambia.
@@
 For example, Senegal, Mali and Sierra Leone have generally good coverage compared to Guinea and Burkina Faso.
-B) Histogram of trap nights performed at each study site, a median of 248 trap nights (IQR 116-500) was performed at each site](`r here("figures", "Figure_1_updated.png")`)
+B) Histogram of trap nights performed at each study site, a median of 248 trap nights (IQR 116-500) was performed at each site.](`r here("figures", "Figure_1_updated.png")`)
 
@@
 The `sf` package in the R statistical language (R version 4.1.2) was used to manipulate geographic data, administrative boundaries were obtained from GADM 4.0.4 [@pebesma_simple_2018; @r_core_team_r_2020; @database_of_global_administrative_areas_gadm_2022].
-Density ($\text{TN}_\text{density}$) was calculated by dividing the number of trap nights by the area of a level-2 administrative area ($\text{R}_\text{area}$).
+Trap-night density ($\text{TN}_\text{density}$) was calculated by dividing the number of trap nights by the area of a level-2 administrative area ($\text{R}_\text{area}$).
 For studies not reporting trap nights imputation was used as previously described.
@@
 Relative trapping effort was then predicted across West Africa using these covariates.
-We performed two sensitivity analyses, first, by removing sites with imputed trapping effort.
-Second rather than associating trap locations to level-2 administrative areas we assigned them to ~ 1km pixels.
+We performed two sensitivity analyses, first, by removing sites with imputed trapping effort, second, by associating trap locations to ~1km pixels rather than level-2 administrative areas.
 
@@
 CLOVER is a synthesis of four host-pathogen datasets (GMPD2, EID2, HP3 and Shaw, 2020) and was released in 2021, it contains more than 25,000 host-pathogen associations for Bacteria, Viruses, Helminth, Protozoa and Fungi.
-We compared the host-pathogen networks across the two datasets, where the CLOVER data was subset for host species present in the rodent trapping datta.
+We compared the host-pathogen networks across the two datasets, where the CLOVER data was subset for host species present in the rodent trapping data.
 
@@
 We use the sampled area of three pathogen groups and two pathogens (Arenaviridae, Borreliaceae, Leptospiraceae, *Lassa mammarenavirus* and *Toxoplasma gondii*) to quantify the bias of sampling within their hosts ranges.
-For each pathogen, we first describe the number of host species assated, for the five most commonly tested species we associate the locations of sampled indivuals to ~20km pixels and calculate the proportion of the IUCN range of the host in which sampling has occurred.
+For each pathogen, we first describe the number of host species assayed, for the five most commonly tested species we associate the locations of sampled individuals to ~20km pixels and calculate the proportion of the IUCN range of the host in which sampling has occurred.
 We compare this figure to the total area in which the host has been detected to produce a measure of relative completeness of sampling within the included rodent trapping studies.
 
-All data and code to reproduce these analyses are available in an archived Zenodo repository [@david_simons_2022_7022637].
+Data and code to reproduce all analyses are available in an archived Zenodo repository [@david_simons_2022_7022637].
 
@@
 
-We found non-random spatial clustering of rodent trapping locations across the study region, suggestive of underlying bias in the sampling or rodents across West Africa.
+We found non-random, spatial clustering of rodent trapping locations across the study region, suggestive of underlying bias in the sampling or rodents across West Africa.
 Trap sites were situated in 256 of 1,450 (17.6%) level-2 administrative regions in 14 West African nations.
-The regions with highest TN density included the capitals and large cities of Niger (Niamey), Nigeria (Ibadan), Ghana (Accra), Senegal (Dakar), Ghana (Accra) and Benin (Cotonou).
+The regions with the highest TN density included the capitals and large cities of Niger (Niamey), Nigeria (Ibadan), Ghana (Accra), Senegal (Dakar), Ghana (Accra) and Benin (Cotonou).
 Outside of these cities, regions in, Northern Senegal, Southern Guinea, Edo and Ogun States in Nigeria and Eastern Sierra Leone had the greatest TN density (Fig 1A.).
@@
 The most parsimonious GAM model (adjusted R2 = 0.3, Deviance explained = 48.7%) reported significant non-linear associations between relative trapping effort bias and human population densities (Effective Degrees of Freedom (EDF) = 7.13, *p* < 0.001), proportion of urban landscape (EDF = 1.92, *p* < 0.002) and region area (EDF = 3.63, *p* < 0.001), alongside significant spatial associations (EDF = 27.3, *p* < 0.001) (Supplementary table 3.1).
-Greatest trapping effort bias peaked at population densities between 5,000-7,500 individuals/km^2^, urban landscape >10% and region areas < 1,000km^2^.
+Greatest trapping effort bias peaked at population densities between 5,000-7,500 individuals/km^2^, proportion of urban landscape >10% and region areas < 1,000km2.
 Increased trapping effort was found in North West Senegal, North and East Sierra Leone, Central Guinea and coastal regions of Nigeria, Benin and Ghana; in contrast South East Nigeria, Northern Nigeria and Burkina Faso had an observed bias towards a reduced trapping effort (Fig 2).
 In sensitivity analysis, excluding sites with imputed trap nights, Mauritania, Northern Senegal and Sierra Leone remained as regions trapped at higher rates, with Nigeria being trapped at lower than expected rates (Supplementary Fig 3a.).
-In pixel based sensitivity analysis spatial coverage was reduced with similar patterns of bias observed to the primary analysis (Supplementary Fig 3b.).
+In pixel-based sensitivity analysis spatial coverage was reduced with similar patterns of bias observed to the primary analysis (Supplementary Fig 3b.).
 
 ![Fig 2.
-Relative trapping effort bias across West Africa from included studies adjusted for human population density, proportion urban land cover and area of the administrative region.
-Yellow regions represent areas with a bias towards increased trapping effort (e.g., North West Senegal), purple regions represent areas with a bias towards reduced trapping effort (e.g., Northern Nigeria).](`r here("figures", "Figure_2_updated.png")`)
+Relative trapping effort bias across West Africa.
+Modelled relative trapping effort bias adjusted for human population density, proportion urban land cover and area of the administrative region.
+Brown regions represent areas with a bias towards increased trapping effort (e.g., North West Senegal), Green regions represent areas with a bias towards reduced trapping effort (e.g., Northern Nigeria).](`r here("figures", "Figure_2_updated.png")`)
 
@@
 R) Detection and non-detection from rodent trapping studies overlaid on IUCN species ranges.
-M.
-musculus has no IUCN West African range.](`r here("figures", "Figure_3_updated.png")`)
+*M.
+musculus* has no IUCN West African range.](`r here("figures", "Figure_3_updated.png")`)
 
-Comparison of proportion of a species IUCN range in which detections and non-detections occurred showed that sampling locations of these seven species within GBIF covered between 0.09-0.26% of expected ranges (Table 1.).
-Detections occurred outside IUCN ranges for all species, most noticeably for *A.
+Comparison of the proportion of a species IUCN range in which detections and non-detections occurred showed that sampling locations of these seven species within GBIF covered between 0.09-0.26% of expected ranges (Table 1.), compared to 0.03-0.24% for rodent trapping data.
+Detections occurred outside IUCN ranges for all species in both the GBIF and rodent trapping data, most noticeably for *A.
 niloticus* and *R.
-rattus*.
-Sampling locations within rodent trapping data covered 0.03-0.24% of expected ranges with detections occurring outside IUCN ranges most noticeably for *A.
-niloticus* and *R.
-rattus*.
-Combining GBIF and rodent trapping data increased the sampled area by a mean of 1.65 times, suggesting limited overlap between the locations sampled.
-Non-detection of species occurred across species ranges (mean = 0.11%, SD = 0.03%), suggestive of spatial heterogeneity of presence within IUCN ranges.
+rattus*.Combining GBIF and rodent trapping data increased the sampled area by a mean of 1.6 times compared to the GBIF area alone, demonstrating limited overlap between the locations providing information to either dataset.
+Non-detection of a species occurred across species ranges (mean = 0.11%, SD = 0.03%), suggestive of spatial heterogeneity of presence within IUCN ranges.
 
@@
 We found potentially important differences between the host-pathogen networks produced from included rodent trapping studies and the consolidated CLOVER dataset.
-When limited to taxonomic classification of both pathogen and host to species level we identified 25 host-pathogen pairs among 14 rodent and 6 pathogen species (Fig 4.).
+When limited to taxonomic classification of both pathogen and host to species level we identified 25 host-pathogen pairs among 14 rodent and 6 pathogen species (Fig 4.
+and Fig 5.).
 We identified negative associations (non-detection through specific assays) for 45 host-pathogen pairs among 35 rodent and 7 pathogen species.
@@
 
-![Fig 4) Identified species level host-pathogen associations through detection of acute infection (i.e.
+![Fig 4.
+Host-Pathogen associations detected through acute infection.
+A) Identified species level host-pathogen associations through detection of acute infection (i.e.
 PCR, culture).
@@
 
-![Fig 5) Identified species level host-pathogen associations through serological assays (i.e.
+![Fig 5.
+Host-Pathogen associations detected through evidence of prior infection.
+B) Identified species level host-pathogen associations through serological assays (i.e.
 ELISA).
@@
 Here we have synthesised data from 126 rodent trapping studies containing information on more than 72,000 rodents across 1,611 trap sites producing an estimated 942,669 trap nights across 14 West African countries.
-Locations studie are complementary to curated datasets (e.g.
+Locations studied are complementary to curated datasets (e.g.
 IUCN, GBIF), incorporation of our synthesised dataset when assessing zoonosis risk based on host distributions could counteract some of the biases inherent to these curated datasets [@boakes_distorted_2010].
@@
 rattus*) and the hazard of endemic zoonosis outbreaks (e.g., *Lassa mammarenavirus*).
-Much of the region remains relatively undersampled, particularly Burkina Faso, Côte d’Ivoire, Ghana and Nigeria, despite these countries facing many of the same challenges.
+Much of the region remains relatively under sampled, particularly Burkina Faso, Côte d’Ivoire, Ghana and Nigeria, despite these countries facing many of the same challenges.
 For example, annual outbreaks of Lassa fever are reported in Nigeria and there are potentially 60,000 unrecognised cases of Lassa fever every year in Côte d’Ivoire and Ghana [@basinski_bridging_2021].
@@
 Future rodent trapping studies should be targeted towards regions that are currently under-studied.
-Further information on rodent presence and abundance across West Africa will assist in modelling changing endemic zoonosis risk and the potential for novel pathogen emergence.
-Sharing of disaggregated data alongside research publications should be supported with adoption of data standards to support further data synthesis.
-Inclusion of exact locations of trapping sites, trapping effort and the dates at which trapping occurred would support more detailed understanding of the spatio-temporal dynamics of host populations and the risk of endemic zoonosis spillover events.
+Further information on rodent presence and abundance across West Africa will aid the modelling of changing endemic zoonosis risk and the potential for novel pathogen emergence.
+Sharing of disaggregated data alongside research publications should be promoted with adoption of data standards to support ongoing data synthesis.
+Specifically, inclusion of exact locations of trapping sites, trapping effort and the dates at which trapping occurred would support more detailed inference of the spatio-temporal dynamics of host populations and the risk of endemic zoonosis spillover events.
 Despite these challenges we propose that rodent trapping studies can provide an important source of data to supplement curated datasets on rodent distributions to quantify the risk of endemic zoonosis spillover events and the hazard of novel pathogen emergence.

### 2022-10-31 scripts/3_2_rodent_host_distributions.R

@@ -24,7 +24,8 @@ included_countries <- level_0 %>%
 
 contiguous_boundary <- included_countries %>%
   filter(!GID_0 == "CPV") %>%
-  summarise()
+  summarise() %>%
+  st_buffer(dist = 0.001)
 
 write_rds(contiguous_boundary, here("data_clean", "WA_continental_boundary.rds"))
 
@@ -117,11 +118,11 @@ plot_fig_3 <- function(species_name, iucn = rodent_iucn, gbif = rodent_gbif, rod
 
   rodents <- rodents %>%
     filter(classification == species_name) %>%
-    select(classification, number, geometry) %>%
+    select(classification, number, town_village, geometry) %>%
     mutate(pres_abs = case_when(number == 0 ~ factor("Non-detection", levels = c("Detection", "Non-detection")),
                                 TRUE ~ factor("Detection", levels = c("Detection", "Non-detection")))) %>%
     st_intersection(., contiguous_boundary) %>%
-    distinct(classification, pres_abs, geometry)
+    distinct(classification, pres_abs, town_village, geometry)
 
   rodent_presence <- rodents %>%
     filter(pres_abs == "Detection")

### 2022-10-31 scripts/4_0_supplementary.R

@@ -223,4 +223,55 @@ save_plot(plot_grid(plot_supp_4_acute, labels = "A"), filename = here("figures",
 save_plot(plot_grid(plot_supp_4_serology, labels = "B"), filename = here("figures", "Supplementary_Figure_4b.pdf"), base_width = 12, base_height = 10)
 save_plot(plot_grid(plot_supp_4_serology, labels = "B"), filename = here("figures", "Supplementary_Figure_4b.png"), base_width = 12, base_height = 10)
 
+# Supplementary 5 WA map
+all_countries <- c("BEN", "BFA", "CIV", "CMR", "CPV", "DZA", "ESH", "GHA",
+                   "GIN", "GMB", "GNB", "LBR", "MAR", "MLI", "MRT", "NER",
+                   "NGA", "SEN", "SLE", "TCD", "TGO")
+continental_countries <- c("BEN", "BFA", "CIV", "ESH", "GHA",
+                           "GIN", "GMB", "GNB", "LBR", "MLI", "MRT",
+                           "NER", "NGA", "SEN", "SLE", "TGO")
+no_data_countries <- c("GMB", "TGO")
 
+level_0 <- read_rds(here("data_download", "admin_spatial", "level_0_admin.rds"))
+
+level_1 <- read_rds(here("data_download", "admin_spatial", "level_1_admin.rds"))
+
+level_2 <- read_rds(here("data_download", "admin_spatial", "level_2_admin.rds"))
+
+capital_cities <- c("Nouakchott", "Dakar", "Conakry", "Western Urban", "Bamako", "Greater Monrovia", "Abidjan", "Kadiogo", "Accra", "Porto-Novo", "Cotonou", "Niamey", "AbujaMun", "IbadanNorth")
+
+countries_capitals <- level_0 %>%
+  filter(GID_0 %in% continental_countries) %>%
+  bind_rows(level_2 %>%
+              filter(GID_0 %in% continental_countries) %>%
+              filter(NAME_2 %in% capital_cities) %>%
+              mutate(`Capital city` = case_when(NAME_2 == "Kadiogo" ~ "Ougadougou",
+                                                NAME_2 == "Greater Monrovia" ~ "Monrovia",
+                                                NAME_2 == "AbujaMun" ~ "Abuja",
+                                                NAME_2 == "IbadanNorth" ~ "Ibadan",
+                                                NAME_2 == "Western Urban" ~ "Freetown",
+                                                TRUE ~ NAME_2)) %>%
[…]

### 2022-10-31 scripts/manuscript_library.R

@@ -1,7 +1,10 @@
+devtools::install_github("yutannihilation/ggsflabel")
+
 if (!require("pacman")) install.packages("pacman")
 pkgs =
   c("here",
     "tidyverse",
+    "ggsflabel",
     "gt",
     "gtsummary",
     "countrycode",
@@ -12,3 +15,4 @@ pkgs =
     "lubridate"
   )
 pacman::p_load(pkgs, character.only = T)
+

### 2022-12-08 reports/archived/manuscript_v3.Rmd
archiving old scripts, manuscript versions and figures

--- 2022-03-09 reports/manuscript_v3.Rmd
@@
 
-**To complete after written**
+Rodents are important important globally distributed reservoirs of known and novel zoonotic pathogens.
+Ongoing anthropogenic land use change is altering the composition of host species assemblages and increasing the risk of zoonotic disease spillover.
+These rapid changes mean that an understanding of the current distribution of rodent species is vital for accurately describing disease hazard and managing risk.
+However, available species distribution and host-pathogen association datasets such as International Union for Conservation of Nature (IUCN) distribution maps, Global Biodiversity Information Facility (GBIF) species occurrence data, or the CLOVER host-pathogen dataset are limited by inclusion of limited data sources **further limitations** and are often biased.
+Here, we synthesise data from West Africa from 128 rodent trapping studies, from 1964-2021, including over 72,000 individual rodent detections in order to more fully categorise the range and presence of important rodent reservoir species in this region.
+We identify that, although rodent trapping studies are biased towards human dominated landscapes across West Africa, we find that they can usefully complement current rodent species distribution datasets.
+For four regionally important zoonotic pathogens (*Borrelia spp.*, *Lassa mammarenavirus*, *Leptospira spp.* and *Toxoplasma gondii*) we identify host-pathogen associations that have not been reported in the consolidated CLOVER dataset.
+These omissions may bias studies investigating the effect of land use and climate change on future disease hazard.
+Finally, we find that for these four diseases the proportion of a rodent hosts range that has been sampled remains small.
+A priority of future rodent trapping studies should be to sample rodent hosts of these diseases across a greater geographic range to investigate the hazard of zoonotic spillover across West Africa.
+In the interim, future studies must incorporate a measure of the current sampling biases.
+The current synthesis of contextually rich rodent trapping data enriches available information from IUCN, GBIF and CLOVER which can support the production of more complete understanding of the hazard of zoonotic spillover events.
 

### 2022-12-08 reports/archived/manuscript_v4.Rmd
archiving old scripts, manuscript versions and figures

--- 2022-08-12 reports/manuscript_v4.Rmd
@@
 
-Endemic rodent zoonoses and novel pathogen emergence from rodent hosts are predicted to have an increasing burden in West Africa and globally (Han et al.
-2015).
-Here we have synthesised data from 126 rodent trapping studies containing information on more than 76,000 rodents across 1,611 trap sites producing an estimated 942,669 trap nights across 14 West African countries.
-Locations studied are complementary to curated datasets (e.g.
-IUCN, GBIF), incorporation of our synthesised dataset when assessing zoonosis risk based on host distributions could counteract some of the biases inherent to these curated datasets (Boakes et al., 2010).
+Endemic rodent zoonoses and novel pathogen emergence from rodent hosts are predicted to have an increasing burden in West Africa and globally (@han_rodent_2015).
+Here we have synthesised data from 126 rodent trapping studies containing information on more than 72,000 rodents across 1,611 trap sites producing an estimated 942,669 trap nights.
+Trapping occurred across 14 West African countries, with locations complementary to curated datasets commonly used in studies assessing zoonosis risk based on host distributions (e.g.
+IUCN, GBIF).
+The inclusion of our synthesised dataset could counteract some of the biases inherent to these curated datasets [@boakes_distorted_2010].
 We identified 25 host-pathogen pairs reported from included studies, 15 of these were not included in a consolidated host-pathogen dataset.
-However, pathogen sampling within identified species occurrence locations was limited, highlighting a further component of sampling bias supporting calls for further pathogen sampling across diverse species in zoonotic hotspots (Harvey and Holmes, 2022).
+Pathogen sampling within identified species occurrence locations was limited, highlighting a further component of sampling bias supporting calls for further pathogen sampling across diverse species in zoonotic hotspots [@harvey_diversity_2022].
 
-We found that rodent trapping studies, similar to biodiversity data, showed important spatial biases (Beck et al., 2014).
-Relative trapping effort was biased towards increased effort in Benin, Guinea, Senegal and Sierra Leone driven by long-standing research collaborations investigating the invasion of non-native rodent species (M.
-musculus and R.
-rattus) and the risk of endemic zoonosis outbreaks (e.g.
+We found that rodent trapping studies, similar to biodiversity data, showed important spatial biases [@beck_spatial_2014].
+Relative trapping effort bias was increased in Benin, Guinea, Senegal and Sierra Leone driven by long-standing research collaborations investigating the invasion non-native rodent species (*M.
+musculus* and *R.
+rattus*) and the hazard of endemic zoonosis outbreaks (e.g.
 Lassa mammarenavirus).
-Much of West Africa remains relatively understudied, particularly Burkina Faso, Côte d’Ivoire, Ghana and Nigeria, despite these countries facing many of the same challenges.
-For example, annual outbreaks of Lassa fever are reported in Nigeria and there are potentially 60,000 unrecognised cases of Lassa fever each year in Côte d’Ivoire and Ghana (Basinski et al., 2021).
+Much of the region remains relatively understudied, particularly Burkina Faso, Côte d’Ivoire, Ghana and Nigeria despite these countries facing many of the same challenges.
+For example, annual outbreaks of Lassa fever are reported in Nigeria and there are potentially 60,000 unrecognised cases of Lassa fever each year in Côte d’Ivoire and Ghana [@basinski_bridging_2021].
 Rodent sampling should be targeted towards regions with minimal data to reduce this bias and assist in understanding the risk from endemic and novel zoonoses.
 
-Rodent trapping studies provide geographic and temporally contextualised data on both species detection and non-detection which are not available from curated datasets.
-Non-detection data can improve models of species distributions, unfortunately, high levels of missing data on trapping effort will continue to confound the allocations of non-detections as true absences (Václavík and Meentemeyer, 2009).
-Models of host species occurrence and abundance, improved by incorporating species absence, are important to assess the effect of land use and climate change on endemic zoonosis spillover to human populations and direct limited public health resources towards regions at greatest risk (Plowright et al., 2019).
+Rodent trapping studies provide data on both presence and absence which are not available from curated datasets.
+Absence data can improve models of species distributions, however, missing data on trapping effort will confound the allocations of non-detections as true absences [@vaclavik_invasive_2009].
+Models of host species occurrence and abundance, improved by incorporating species absence, are important to assess the effect of land use and climate change on endemic zoonosis spillover to human populations and direct limited public health resources towards regions at greatest risk.
 
 Currently available consolidated datasets on host-pathogen associations (e.g.
-CLOVER, EID2 and GMPD2) do not include spatial or temporal components (Gibb, Gregory F.
-Albery, et al., 2021a).
-The current synthesis of rodent trapping studies has highlighted that pathogens have been sparsely sampled within a hosts range.
-Current zoonosis risk models dependent on these sources of data are therefore not able to incorporate spatial heterogeneity in pathogen prevalence across the host range.
-Additional uncertainty in current models of zoonotic disease risk arises from observed host-pathogen associations that have not been reported in these consolidated datasets.
-For example, Hylomyscus pamfi infected with Lassa mammarenavirus and R.
-rattus infected with Coxiella burnetiid, will not be included when solely dependent on consolidated host-pathogen datasets.
-Further, detection of zoonotic pathogens in multiple co-occurring host species supports the adoption of multi-species to better understand the potential range of endemic zoonoses (Wilkinson et al., 2019).
+CLOVER, EID2 and GMPD2) do not include spatial or temporal components [@gibb_data_2021].
+The current synthesis of rodent trapping studies has shown that pathogens have only been sparsely tested for within a hosts range.
+Current zoonosis risk models dependent on these sources of data are unable to incorporate spatial heterogeneity in pathogen prevalence across a hosts range.
+Additional uncertainty in current models of zoonotic disease risk arises from host-pathogen associations that have not been reported in these consolidated datasets, for example *Hylomyscus pamfi* and Lassa fever and *R.
+rattus* and Q fever (*Coxiella burnetii*).
+Detection of zoonotic pathogens in multiple co-occurring host species supports the adoption of multi-species to better understand the potential range of endemic zoonoses [@wilkinson_comparison_2019].
 
-Few studies stratified detection and non-detection of hosts or pathogen prevalence by time, therefore limiting inference of changes in host and pathogen dynamics.
-This limitation prevents calculation of incidence of infection and the abundance of infectious rodents which potentially varies by both time and space (Fichet-Calvet et al., 2016).
+Few studies stratified presence-absence or pathogen detection by time at the level of rodent species, limiting observation of the changes in host and pathogen dynamics.
+This limitation prevents calculation of incidence of infection and prevalence of infectious rodents which potentially varies by both time and space [@fichet-calvet_spatial_2016].
 Understanding of temporal changes in viral burden and shedding for endemic zoonoses is required to accurately predict current and future risk of pathogen spillover.
 
-Finally, due to data sparsity, we were unable to account for temporal change over the six decades of rodent trapping studies.
-Land use change and population density have changed dramatically over this period in West Africa (Herrmann et al., 2020).
-We attempted to mitigate against this by using the median year of trapping as a reference point to understand the spatial and land use biases in trapping activity.
-It is possible that land use and population density at trapping sites varied importantly between when the rodent trapping study was conducted and the land use and human population variables in 2005.
-Despite this limitation, the finding that trapping is biased towards high density, human dominated landscapes is unlikely to substantially change (Beck et al., 2014).
+Due to data sparsity, we were unable to account for temporal change over the six decades of rodent trapping studies.
+Land use change and population density have changed dramatically over this period in West Africa [@herrmann_accelerating_2020].
+We attempted to mitigate against this by using the median year of trapping to understand the spatial and land use biases in trapping activity, however, it is possible that land use and population density at trapping sites varied importantly between when the rodent trapping study was conducted and the conditions in 2005.
+Despite this limitation, the finding that trapping is biased towards high density, human dominated landscapes is unlikely to substantially change.
 
@@
 Together this supports their inclusion in efforts to model endemic zoonotic risk and novel pathogen emergence.
-Contribution of rodent trapping studies as data sources can be improved by adopting reporting standards and practices consistent with Open Science, namely sharing of disaggregated datasets alongside publication (Foster and Deardorff, 2017).
+However, utility can be improved by adopting reporting standards and practices consistent with Open Science [@foster_open_2017].
 
@@
 Inclusion of exact locations of trapping sites, trapping effort and the dates at which trapping occurred would support more detailed understanding of the spatio-temporal dynamics of host populations and the risk of endemic zoonosis spillover events.
-Despite these challenges we propose that rodent trapping studies can provide an important source of data to supplement curated datasets on rodent distributions to quantify the risk of endemic zoonosis spillover events and the hazard of novel pathogen emergence.
+Despite these challenges we propose that rodent trapping studies can provide an important source of data to supplement consolidated datasets on rodent distributions to quantify the risk of endemic zoonosis spillover events and the hazard of novel pathogen emergence.
 

### 2022-12-08 reports/archived/test.Rmd
archiving old scripts, manuscript versions and figures

--- 2022-05-24 reports/manuscript_v4.Rmd
@@
----
-title: "Rodent trapping studies can improve our understanding of the hazards of zoonotic pathogen spillover and novel zoonotic pathogen emergence."
-date: "`r Sys.Date()`"
-author: 
-  - "David Simons"
-  - "Lauren A. Attfield"
-  - "Kate E. Jones"
-  - "Deborah Watson-Jones"
-  - "Richard Kock"
-output: 
-  officedown::rdocx_document:
-    tables:
-      caption:
-       style: false
-    plots:
-      caption:
-        style: false
-always_allow_html: yes
-bibliography: citations/manuscript_citations.bib
-csl: elsevier_vancouver.csl
----
-
 ```{r echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
@@
 
-# Abstract
-
-NEEDS UPDATING
-
-# Introduction
-
-There is increasing awareness of the global health and economic impacts of novel zoonotic pathogen spillover, evidenced by the ongoing SARS-CoV-2 pandemic and previous HIV/AIDs and Spanish Influenza pandemics [@Bernstein].
-The number of zoonotic disease spillover events and the frequency of the emergence of novel zoonotic pathogens are predicted to increase under intensifying anthropogenic pressure driven by increased human populations, urbanisation, intensification of agriculture, climate change and wildlife defaunation [@allen_global_2017; @hassell_urbanization_2017; @garcia-pena_land-use_2021; @mcmahon_ecosystem_2018].
-The impact of endemic zoonoses meanwhile remains underestimated [@Maudlin].
-Endemic zoonoses disproportionally affect those in the poorest sections of society, those living in close contact with their animals and those with limited access to healthcare [@Halliday; @Molyneux; @judson_zoonoses_2021].
-West Africa is a nexus for these processes and is identified as a hotspot for increased risk of large epidemics of known zoonotic diseases and novel zoonotic pathogen emergence [@allen_global_2017; @grace_mapping_2012; @han_global_2016].
-Two taxonomic groups - rodents (Rodentia) and bats (Chiroptera) - contribute the greatest number of predicted novel zoonotic pathogens and known zoonoses [@han_rodent_2015; @gibb_mammal_2021].
-Of 2,220 extant rodent species, 244 (10.7%) are described as reservoirs of 85 zoonotic pathogens [@han_rodent_2015].
-Specifically, West Africa has been identified as a region at increased hazard for rodent-borne zoonotic disease spillover events under different projected future land-use change scenarios [@garcia-pena_land-use_2021].
-Currently within West Africa, rodents are involved in the transmission of multiple endemic zoonoses with large burdens on human health, these pathogens include, Lassa fever, Schistosomiasis, Leptospirosis and Toxoplasmosis [@meerburg_rodent-borne_2009; @galeh_global_2020].
-
-Rodent species form diverse assemblages, which provide important and beneficial ecosystem services including pest regulation and seed dispersal [@fischer_ecosystem_2018].
-The role of rodent species in zoonotic disease spillover or novel zoonotic pathogens emergence are examples of ecosystem disservices.
-Rodents typically demonstrate “fast” life histories [@dobson_fast_2007] with traits such as early maturation and short gestation times (<4 days) further associated with being zoonotic reservoirs [@han_rodent_2015; @albery_fast-lived_2021].
-Rodent species with “fast” life histories thrive in human dominated landscapes, displacing species less likely to be reservoirs of zoonotic pathogens [@gibb_zoonotic_2020].
-The widespread occurrence of reservoir species and their proximity to human activity make the description of rodent species assemblages and host-pathogen associations vitally important to understanding the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence [@han_global_2016].
-
-Despite the importance of understanding these complex systems, current evidence on host-pathogen associations is considerably affected by taxonomic and geographical sampling biases [@gibb_ecosystem_2020; @gibb_mammal_2021].
-Curated biodiversity datasets such as the Global Biodiversity Information Facility (GBIF) and resources produced by the International Union for Conservation of Nature suffer from well described sampling biases [@Bowler; @Boakes].
-These sampling biases can gravely distort produced species distribution models [@Beck].
-Datasets on host-pathogen associations also suffer from biases introduced from literature selection criteria and taxanomic discrepancies a recently produced consolidated dataset (CLOVER) has been produced to reduce this [@Gibb_clover].
-Studies identifying potential geographic hotspots of zoonotic disease spillover and novel pathogen emergence are typically based on these datasources for modelling host species distributions and host-pathogen associations which can result in biased hazard estimates [@plowright_sampling_2019; @carlson_future_2021].
-For example, systematically increased sampling, over-representation of certain habitats and clustering around areas of high human population could lead to an apparent association between locations and hazard that is driven by these factors rather than underlying host-pathogen associations [@wille_how_2021; @redding_evaluating_2017; @gibb_mammal_2021].
-While other regions remain systematically undersampled (e.g., areas of sparse human populations, endemic zoonoses), potentially resulting in a reduced hazard of zoonotic spillover events or novel zoonotic pathogen emergence being attributed [@beck_spatial_2014].
-Predictions of zoonotic disease spillover and novel zoonotic pathogen emergence must account for these biases to understand the future hazard of zoonotic diseases [@carlson_future_2021].
-
-Rodent trapping studies provide contextually rich information on when, where and under what conditions rodents were trapped, potentially enriching consolidated datasets and providing additional data to minimise inherent biases [@bovendorp_optimising_2017].
-Rodent trapping studies describe rodent population assemblages, their geographic distribution and host-pathogen associations beyond what is currently available in consolidated datasets.
-Studies have been conducted in West Africa to investigate the distribution of rodent species, their species assemblages, the prevalence of endemic zoonoses within rodent hosts (e.g., Lassa fever, Schistosomiasis) and to identify emerging and novel zoonotic pathogens [@fichet-calvet_diversity_2009; @catalano_multihost_2020; @usaid_one_2021].
-However, individual level data from these studies have not previously been synthesised for inclusion in assessments of zoonotic disease spillover and novel zoonotic pathogen emergence.
-
-Here, we synthesise rodent trapping studies conducted across West Africa from a search of literature between 1964-2022.
-First, we use this dataset to investigate the geographic sampling biases in relation to human population density and land use classification.
-Second, we compare this to curated host datasets (IUCN and GBIF) to understand differences in reported host geographic distributions.
-Third, we compare identified host-pathogen associations with a consolidated dataset (CLOVER) to explore discrepancies in rodent host-pathogen associations and report the proportion of positive assays for pathogens of interest.
-Finally, within our dataset we investigate the spatial extent of current host-pathogen sampling to identify areas of sparse sampling of pathogens within their host ranges.
-We expect that rodent trapping studies provide an important additional source of high-resolution data that can be used to enrich available consolidated datasets to better understand the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence across West Africa.
-
-# Methods
-
-## Data sources
-
-### Host and pathogen trapping data
-
-To identify relevant literature, we conducted a search in Ovid MEDLINE, Web of Science (Core collection and Zoological Record), JSTOR, BioOne, African Journals Online, Global Health and the pre-print servers, BioRxiv and EcoEvoRxiv for the following terms as exploded keywords: (1) Rodent OR Rodent trap* AND (2) West Africa, no date limits were set.
-We also searched other resources including the UN Official Documents System, Open Grey, AGRIS FAO and Google Scholar using combinations of the above terms.
-Searches were run on 2022-05-01, and returned studies conducted between 1964-2021.
-
-For further analysis, we included studies if they met all of the following inclusion criteria; i) reported findings from trapping studies where the target was a small mammal, ii) described the type of trap used or the length of trapping activity or the location of the trapping activity, iii) included trapping activity from at least one West African country, iv) recorded the genus or species of trapped individuals, and v) were published in a peer-reviewed journal or as a pre-print on a digital platform or as a report by a credible organisation.
-We excluded studies if they met any of the following exclusion criteria: i) reported data that were duplicated from a previously included study, ii) no full text available, iii) not available in English.
-One author screened titles, abstracts and full texts against the inclusion and exclusion criteria.
-At each stage; title screening, abstract screening and full text review, a random subset (10%) was reviewed by a second author.
-
-We extracted data from eligible studies using a standardised tool that was piloted on 5 studies (Supplementary Table 1.).
-Data was abstracted into a Google Sheets document, which was archived on completion of data extraction [ref to zenodo on finalised].
-We identified the aims of included studies, for example, whether it was conducted as a survey of small mammal species or specifically to assess the risk of zoonotic disease spillover.
-To understand potential limitations in the representativeness of rodent trapping we extracted data on study methodology, such as, the number of trap nights, the type of traps used and whether the study attempted to estimate abundance.
-For studies not reporting number of trap nights we used imputation based on the number of trapped individuals, stratified by the habitat type from which they were obtained.
-This was performed by multiplying the total number of trapped individuals within that study site by the median trap success for study sites with the same reported habitat type.
-Stratification was used as trap success varied importantly between traps placed in or around buildings (13%, IQR 6-24%) compared with other habitats (3%, IQR 1-9%)
-
-The method of taxonomic identification applied within a study was recorded, reported species identification was assumed to be accurate.
-The number of individuals of these species or genera was extracted with species and genus names mapped to GBIF taxonomy [@gbif_the_global_biodiversity_information_facility_global_2021] for internal consistency.
-We expanded species detection and non-detection records by explicitly specifying non-detection at a trap site if a species was recorded as detected at other trapping locations within the same study.
-
-Geographic locations of trapping studies was extracted using GPS locations for the most precise location presented.
-GPS coordinates for locations not reported were produced using the National Geospatial-Intelligence Agency GEOnet Names Server [@GeoNames] based on place-names and maps presented in the study.
-All locations were converted to decimal degrees for internal consistency.
-The year of rodent trapping was extracted alongside the length of the trapping activity to understand seasonal representativeness of trapping activity.
-The habitats of trapping sites were mapped to the IUCN Habitat Classification Scheme (Version 3.1).
-For studies reporting multiple habitat types for a single trap, trap-line or trapping grid, a higher order classification of habitat type was recorded.
-
-For relevant studies we extracted data on all microorganisms and zoonotic pathogens tested and the method used (e.g., molecular or serological diagnosis).
-Where assays were able to identify the microorganism to species level this was recorded, non-specific assays higher order attribution was used (e.g.
-to family level).
-We recorded the species of rodent host tested, the number of individuals tested and the number of positive and negative results.
-For studies reporting summary results all testing data were extracted, this may introduce double counting of individual rodents, for example, if a single rodent was tested using both molecular and serological assays.
-Where studies reported indeterminate results, these were also recorded.
-
-We summarised the number of studies, the year in which trapping occurred and the country in which they were conducted.
-We identified 4,692 relevant citations, with 127 rodent trapping studies included (Supplementary Table 2.).
-The earliest trapping studies were conducted in 1964, with a trend of increasing numbers of studies being performed annually since 2000.
-The median year of first trapping activity was 2007, with the median length of trapping activity being 1 year (IQR 0-2 years) (Supplementary Figure 1.).
-Studies were conducted in 14 West African countries, with no studies reported from The Gambia or Togo, at 1,611 trap sites (Figure 1A.).
-
-Included studies explicitly reported on 601,184 trap nights, a further 341,445 trap nights were imputed from studies with no recording of trapping effort based on trap success.
-A total of 942,629 trap nights were estimated (Figure 1B.).
-A minority of studies trapped at a single study site (30, 24%), with 46 (36%) trapping at between two and five sites, the remaining 51 studies (40%) trapped at between six and 93 study sites.
-
-In total 76,275 small mammals were trapped with 65,628 (90%) identified to species level and 7,439 (10%)  identified to genus, with the remaining classified to higher taxonomic level.
-The majority of the 132 identified species were Rodentia (102, 78%), of which Muridae (73, 72%) were the most common family.
-Soricomorpha were the second most identified order of small mammals (28, 21%).
-57 studies tested for 32 microorganisms, defined to species or genus level that are known or potential pathogens.
-Most studies tested for a single microorganism (48, 84%).
-The most frequently assayed microorganisms were *Lassa mammarenavirus* or Arenaviridae (21, 37%), *Borrelia sp.* (9, 16%), *Bartonella sp.* (4, 7%) and *Toxoplasma gondii* (4, 7%).
-Most studies used Polymerase Chain Reaction (PCR) to detect microorganisms (37, 65%), with fewer studies using serology-based tests (11, 19%) or histological or direct visualisation assays (11, 21%).
-From 32,920 individual rodent samples we produced 351 host-pathogen pairs.
-With *Rattus rattus*, *Mus musculus*, *Mastomys erythroleucus*, *Mastomys natalensis* and *Arvicanthis niloticus* being assayed for at least 18 microorganisms.
-Further descriptive information from the included studies including geolocated trapping of species, their detection and non-detection alongside microorganism data is available as supplementary material [@simons_rodents_2022].
-
-![Figure 1: A) The location of trapping sites in West Africa.
-No sites were recorded from Togo or The Gambia.
-Heterogeneity is observed in the coverage of each country by trap night (colour) and location of sites.
-For example, Senegal, Mali and Sierra Leone have generally good coverage compared to Guinea and Burkina Faso.
-B) Histogram of trap nights performed at each study site, a median of 248 trap nights (IQR 116-500) was performed at each site](`r here("figures", "Figure_1_updated.png")`)
-
-## Analysis
-
-### What is the extent of spatial bias in the rodent trapping data?
-
-To investigate the risk to human health from endemic zoonoses and novel pathogen emergence using synthesised data from rodent trapping data understanding of sampling biases is vital.
-Sampling of rodents has not been performed randomly across space in West Africa (Figure 1.) and at varying intensity, we aimed quantify this bias as the relationship between trapping effort (measured as trap night density (TN density)), human population density and land use type.
-We define this as relative trapping effort bias.
-
-To address this question, we calculated TN density within each West African level-2 administrative region, administrative boundaries were obtained from GADM 4.0.4 [@Database of Global Administrative Areas].
-We assigned trap nights to specific regions using GPS coordinates obtained from trapping studies as described above.
-The sf package in the R statistical language was used to handle shapefiles and geographic data [@Pebesma].
-Density was calculated by dividing the number of trap nights by the area of a level-2 administrative area.
-For studies not reporting trap nights imputation was used as previously described.
-Human population density was obtained for the closest year (2005) to the median year of trapping (2007) from Socioeconomic Data and Applications Center (SEDAC) gridded population of the world v4 at ~ 1km resolution [@SEDAC].
-Median population density was then calculated for each level-2 administrative region.
-Land cover classification was obtained from the Copernicus climate change service at ~300m resolution [@COPERNICUS].
-The proportion of cropland, shrubland, tree cover and urban land cover within a level-2 administrative region in 2005 was calculated.
-
-We investigated the role of these covariates on the spatial bias of TN density by comparing to a spatial model (equation 1), as follows:
-
-$$
-\begin{equation}
-\tag{1}
-\text{TN density} \sim \text{Tweedie((Longitude * Latitude))}
-\end{equation}
-
-\begin{equation}
-\tag{2}
-\text{TN density} \sim \text{Tweedie((Population density) + (Longitude * Latitude))}
-\end{equation}
-
-\begin{equation}
-\tag{3}
-\text{TN density} \sim \text{Tweedie((Population density) + (Region area) + (Longitude * Latitude))}
-\end{equation}
-
-\begin{equation}
-\tag{4}
-\text{TN density} \sim \text{Tweedie((Proportion cropland) + (Proportion tree cover) + (Proportion urban) + (Longitude * Latitude))}
-\end{equation}
-
-\begin{equation}
-\tag{5}
-\text{TN density} \sim \text{Tweedie((Population density) + (Region area) + (Proportion urban) + (Longitude * Latitude))}
-\end{equation}
-$$
-
-We formulated these models as a Generalised Additive Models (GAM) incorporating a spatial interaction term (longitude by latitude) [@pedersen_hierarchical_2019].
-These models investigated the association between TN density, proportion of urban, cropland, tree cover and human population density.
-The models were constructed in the *mgcv* package using the R programming language [@wood_generalized_2017; @r_core_team_r_2020].
-Selection of the most parsimonious model was based on Deviance explained and the Akaike information criterion for each model.
-We performed sensitivity analysis by removing sites with imputed trapping effort and performing a pixel-based analysis.
-
-### What is the difference in rodent host distributions between curated datasets and rodent trapping studies?
-
-Curated datasets (i.e., IUCN and GBIF) are commonly used in assessments of biodiversity in response to land use and climate change in addition to being used to explore zoonotic pathogen host distributions.
-To understand any biases introduced from these sources into these analyses we aimed to assess the concordance of curated rodent host distributions from IUCN and GBIF with observed rodent presence and absence from rodent trapping studies.
-We obtained rodent species distribution maps as shapefiles from the IUCN red list and translated these to a ~20km resolution raster [@iucn_iucn_2021].
-Distributions were cropped to the study region for globally distributed rodent species.
-We obtained rodent presence locations from GBIF as point data limited to the study region [@gbif_the_global_biodiversity_information_facility_occurrence_2021].
-Presence locations were associated to cells of raster with a ~20km resolution produced for the study region.
-
-We assessed the taxonomic coverage of the trapped taxa and for the seven species with the most trap locations (*M.
-natalensis*, *R.
-rattus*, *M.
-erythroleucus*, *M.
-musculus*, *A.
-niloticus*, *Praomys daltoni* and *Cricetomys gambianus*) we mapped detection and non-detection, associating each location with a ~20km resolution raster.
-For these seven species we calculated the area of the IUCN expected range.
-We calculated the percentage of this range covered by presence detections in GBIF, this process was repeated for presence detections from rodent trapping studies.
-For both GBIF data and trapping studies we calculated the area of detections outside of the IUCN expected range.
-For rodent trapping studies only, we were able to calculate the area of non-detections within the IUCN expected area.
-Finally, we calculated the combined area of detection when combining GBIF and rodent trapping study data.
-
-### Are rodent trapping derived host-pathogen associations present in a consolidated zoonoses dataset?
-
-Studies investigating the risk of endemic zoonoses and novel pathogen emergence generally rely on consolidated datasets of host-pathogen associations.
-To examine the usefulness of rodent trapping studies as an additional source of data we compared identified host-pathogen asociations with a consolidated zoonoses dataset (CLOVER) [@Gibb_data_proliferation].
-CLOVER is a synthesis of four host-pathogen datasets (GMPD2, EID2, HP3 and Shaw, 2020) and was released in 2021, it contains more than 25,000 host-pathogen associations for Bacteria, Viruses, Helminth, Protozoa and Fungi.
-An archived version of the CLOVER dataset was obtained and subset for host species trapped within included studies, spatial filtering of CLOVER for host-pathogen associations limited to the study region was not possible [@Gibb_clover_data].
-
-We produced host-pathogen pairs from rodent trapping studies investigating zoonoses.
-For each host and pathogen, with both identified to species level, we compared our host-pathogen pair network with one produced using CLOVER, reporting concordance and discordance.
-We describe these networks and report differences in network composition.
-For host-pathogen pairs with assay results consistent with acute or prior infection we calculate the proportion positive and identify those absent from CLOVER.
-We expand the analysis to host-pathogen pairs with pathogens identified to genus level in Supplementary Figure 4.
-
-### What is the spatial extent of pathogen testing within a hosts range?
-
-Sampling of zoonotic pathogens may also be biased within their hosts ranges, which may hinder inference across the entire range of the species.
-To quantify this, we focused on the five pathogen groups and species sampled at the most locations (Arenaviridae, Borreliaceae, Leptospiraceae, *Lassa mammarenavirus* and *Toxoplasma gondii*).
-We geolocated individuals of host species tested for these pathogens and calculated the area sampled.
-We adopt a similar approach to above by mapping sampled hosts to ~20km pixels and calculating the proportion of their IUCN range in which sampling has occurred.
-We also present the area covered of all areas in which the host has been detected in this study to produce a measure of relative completeness of sampling within the included rodent trapping studies.
-
-Data and code implemented in the R statistical language has been made available alongside this manuscript to reproduce these analyses [ref_zenodo].
-
-# Results
-
-## What is the extent of spatial bias in the rodent trapping data?
-
-We found a non-random spatial distribution of rodent trapping locations across the study region, suggestive of underlying bias in the sampling or rodents across West Africa.
-Trap sites were situated in 256 of 1,450 (17.6%) level-2 administrative regions in 14 West African nations.
-The regions with highest TN density included the capitals and large cities of Niger (Niamey), Nigeria (Ibadan), Ghana (Accra), Senegal (Dakar), Ghana (Accra) and Benin (Cotonou).
-Outside of these cities, regions in, Northern Senegal, Southern Guinea, Edo and Ogun States in Nigeria and Eastern Sierra Leone had the greatest TN density (Figure 1A.).
-
-Trapping sites were biased towards human modified landscapes (i.e., cropland, grassland and mosaic habitats), with under-representation of forest and bare habitats.
-However, urban habitats were comparatively undersampled (Supplementary Figure 2.).
-
-The most parsimonious GAM model (adjusted R2 = 0.3, Deviance explained = 48.7%) included smooth terms for geographic coordinates, human population density, proportion of urban landscape, and the area of the region.
-All smooth terms demonstrate significant non-linear associations (*p* < 0.002) with TN density (Supplementary Table 3.5).
-The spatial smooth term adjusted for population density, proportion of urban landscape and region area is shown in Figure 2.
-The final model identified relative trapping effort bias, with a bias towards increased trapping effort in North West Senegal, North and East Sierra Leone, Central Guinea and coastal regions of Nigeria, Benin and Ghana; in contrast South East Nigeria, Northern Nigeria and Burkina Faso had an observed bias towards a reduced trapping effort (Figure 2).
-Model summaries informing model selection are shown in Supplementary Tables 3.1-3.5.
-In sensitivity analysis, excluding sites with imputed trap nights reduced contributing data; however, Mauritania, Northern Senegal and Sierra Leone remained as regions trapped at higher rates, with Nigeria being trapped at lower than expected rates (Supplementary Figure 3a.).
-In pixel based sensitivity analysis spatial coverage was reduced with similar patterns of bias observed (Supplementary Figure 3b.).
-
-![Figure 2.
-Relative trapping effort bias across West Africa from included studies adjusted for human population density, proportion urban land cover and area of the administrative region.
-Uncertainty in the point estimate of relative trapping effort bias is represented by colour transparency/pallor (e.g., Eastern Liberia and Central Mali).
-Yellow regions represent areas with a bias towards increased trapping effort (e.g., North West Senegal), purple regions represent areas with a bias towards reduced trapping effort (e.g.,Northern Nigeria).](`r here("figures", "Figure_2_updated.png")`)
-
-## What is the difference in rodent host distributions between curated datasets and rodent trapping studies?
-
-We found that rodent trapping studies could provide a useful supplement to curated data describing rodent host distributions.
-For the seven rodent species of interest (*M.
-natalensis*, *R.
-rattus*, *M.
-erythroleucus*, *M.
-musculus*, *A.
-niloticus*, *P.
-daltoni* and *C.
-gambianus*), except *C.
-gambianus*, trapping studies provided more distinct locations of detection and non-detection than were available from GBIF.
-For the endemic rodent species (*M.
-natalensis*, *M.
-erythroleucus*, *A.
-niloticus*, *P.
-daltoni* and *C.
-gambianus*) IUCN ranges had good concordance to both trapping studies and GBIF, however, individuals of *A.
-niloticus* and *P.
-daltoni* were detected outside of IUCN ranges.
-In contrast, the non-native species *R.
-rattus* and *M.
-musculus* were detected across much greater ranges than were expected from IUCN distributions (Figure 3A.
-and 3B.).
-
-![Figure 3A.
-Each row corresponds to a single rodent species.
-L) Presence recorded in GBIF overlaid on IUCN species range (red-shaded area).
-R) Detection and non-detection from rodent trapping studies overlaid on IUCN species ranges.
-M.
-musculus has no IUCN West African range.](`r here("figures", "Figure_3_test_a.png")`)
-
-![Figure 3B.
-Each row corresponds to a single rodent species.
-L) Presence recorded in GBIF overlaid on IUCN species range (red-shaded area).
-R) Detection (purple) and non-detection (orange) from rodent trapping studies overlaid on IUCN species ranges.](`r here("figures", "Figure_3_test_b.png")`)
-
-To further understand the sampling coverage of curated datasets and included rodent trapping studies we identified the proportion of a species IUCN range in which detections and non-detections occurred for the seven species most detected in the included studies (Table 1.).
-Apart from *M.
-musculus* these species had IUCN ranges greater than 1 million km2, with the Mastomys species having the largest geographic ranges.
-Sampling locations of these species within GBIF covered between 0.09-0.26% of expected ranges with detections occurring outside IUCN ranges for all species, most noticeably for *A.
-niloticus* and *R.
-rattus*.
-Sampling locations within rodent trapping studies covered 0.03-0.24% of expected ranges with detections occurring outside IUCN ranges most noticeably for *A.
-niloticus* and *R.
-rattus*.
-Combining GBIF and rodent trapping data increased the sampled area by a mean of 65% (SD 36.5%) , suggesting limited overlap between the locations sampled.
-Non-detection of species occurred across species ranges (mean = 0.11%, SD = 0.03%), suggestive of heterogeneity of presence across IUCN ranges.
-
 ```{r table-1}
 
-table_1 <- read_rds(here("tables", "Table_1_updated_pixel.rds")) %>%
-  set_caption(caption = "Table 1: Comparison of IUCN, GBIF and rodent trapping ranges for the 7 species trapped at the most sites.") %>%
-  autofit()
-
-table_1
+read_rds(here("tables", "Table_1_updated.rds")) %>%
+  set_caption(caption = "Table 1: Regional")
 
@@
 
-## Are rodent trapping derived host-pathogen associations present in a consolidated zoonoses dataset?
+```{r table-1-pixel}
 
-We found potentially important differences between the host-pathogen networks produced from included rodent trapping studies and the consolidated CLOVER dataset.
-When limited to taxonomic classification of both pathogen and host to species level we identified 25 host-pathogen pairs among 14 rodent and 6 pathogen species (Figure 4.).
-We identified negative associations (non-detection through specific assays) for 45 host-pathogen pairs among 35 rodent and 7 pathogen species.
-CLOVER contained 10 (40%) of our identified host-pathogen associations, the remaining 15 (60%) were not found to be present in CLOVER, additionally CLOVER recorded positive associations for 4 (9%) of the negative associations produced from the rodent trapping studies.
-
-![](`r here("figures", "Figure_4a_updated.png")`)
-
-![Figure 4.
-A) Identified species level host-pathogen associations through detection of acute infection (i.e.
-PCR, culture).
-Percentages and colour relate to the proportion of all assays that were positive.
-Associations with a black border are present in the CLOVER dataset.
-B) Identified species level host-pathogen associations through serological assays (i.e.
-ELISA).
-Percentages and colour relate to the proportion of all assays that were positive.
-Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Figure_4b_updated.png")`)
-
-CLOVER included an additional 492 host-pathogen associations we do not observe in rodent trapping studies.
-The majority of these 392 (80%) pairs are from species with global distributions (*M.
-musculus*, *R.
-rattus* and *R.
-norvegicus*), or from those with wide ranging distributions in sub-Saharan Africa (38, 8%) (i.e., *A.
-niloticus*, *M.
-natalensis* and *Atelerix albiventris*).
-
-For pathogens not identified to species level (i.e.
-family or higher taxa only), we identified 148 host-pathogen pairs among 32 rodent species and 25 pathogen families (Supplementary Figure 4.), with CLOVER containing 66 (45%) of these associations.
-
-Rodent trapping studies identified additional rodent host species for six pathogens; *Lassa mammarenavirus* (5), *Toxoplasma gondii* (4), Usutu virus (2), *Coxiella burnetii* (2), *Escherichia coli* and *Klebsiella pneumoniae* (both 1), that were not present in this consolidated host-pathogen association dataset.
-
-## What is the spatial extent of pathogen testing within a host’s range?
-
-The five most widely sampled pathogen species/families in included studies were Arenaviridae, Borreliaceae, *Lassa mammarenavirus*, Leptospiraceae and *Toxoplasma gondii* (Table 2.).
-Assays to identify Arenaviridae infection were performed in 44 rodent species with evidence of viral infection in 15 species.
-*Lassa mammarenavirus* was specifically tested for in 43 species with 10 showing evidence of viral infection.
-The most commonly infected species for both Arenaviridae, generally, and *Lassa mammarenavirus* specifically, were *M.
-natalensis* and *M.
-erythroleucus*.
-Assays were conducted on individuals trapped within less than a third of the locations in which the species were trapped (Table 2.).
-
-Infection with species of Borreliaceae was assessed in 42 species, with evidence of infection in 17 rodent species.
-Among species with more than 500 samples the highest rates of infection were among *A.
-niloticus* (16%), *Mastomys huberti* (11%) and *M.
-erythroleucus* (9%).
-The coverage of sampling within the trapping locations was slightly greater than that for Arenaviridae but remains a small proportion (<1%) of the host species IUCN range.
-Infection with species of Leptospiraceae and *Toxoplasma gondii* was assessed in 8 species, with evidence of infection in 5 and 6 rodent species respectively.
-Testing for these pathogens was more limited with few individuals tested and reduced coverage of host ranges.
-
-```{r table-2}
-
-table_2 <- read_rds(here("tables", "Table_2_updated_pixel.rds"))%>%
-  set_caption(caption = "Table 2: Comparison of pathogen sampling ranges for the 5 most widely sampled pathogens and the 5 most sampled potential rodent host species.
-* no IUCN range in West African") %>%
-  autofit()
-
-table_2
+read_rds(here("tables", "Table_1_updated_pixel.rds")) %>%
+  set_caption(caption = "Table 1: Pixel")
 
 ```
-
-# Discussion
-
-Endemic rodent zoonoses and novel pathogen emergence from rodent hosts are predicted to have an increasing burden in West Africa and globally (@han_rodent_2015).
-Here we have synthesised data from 126 rodent trapping studies containing information on more than 72,000 rodents across 1,611 trap sites producing an estimated 942,669 trap nights.
-Trapping occurred across 14 West African countries, with locations complementary to curated datasets commonly used in studies assessing zoonosis risk based on host distributions (e.g.
-IUCN, GBIF).
-The inclusion of our synthesised dataset could counteract some of the biases inherent to these curated datasets [@Boakes].
-We identified 25 host-pathogen pairs reported from included studies, 15 of these were not included in a consolidated host-pathogen dataset.
-Pathogen sampling within identified species occurrence locations was limited, highlighting a further component of sampling bias supporting calls for further pathogen sampling across diverse species in zoonotic hotspots [@Harvey].
-
-We found that rodent trapping studies, similar to biodiversity data, showed important spatial biases [@Beck].
-Relative trapping effort bias was increased in Benin, Guinea, Senegal and Sierra Leone driven by long-standing research collaborations investigating the invasion non-native rodent species (*M.
-musculus* and *R.
-rattus*) and the hazard of endemic zoonosis outbreaks (e.g.
-Lassa mammarenavirus).
-Much of the region remains relatively understudied, particularly Burkina Faso, Côte d’Ivoire, Ghana and Nigeria despite these countries facing many of the same challenges.
-For example, annual outbreaks of Lassa fever are reported in Nigeria and there are potentially 60,000 unrecognised cases of Lassa fever each year in Côte d’Ivoire and Ghana [@Basinski].
-Rodent sampling should be targeted towards regions with minimal data to reduce this bias and assist in understanding the risk from endemic and novel zoonoses.
-
-Rodent trapping studies provide data on both presence and absence which are not available from curated datasets.
-Absence data can improve models of species distributions, however, missing data on trapping effort will confound the allocations of non-detections as true absences [@Vaclavik].
-Models of host species occurrence and abundance, improved by incorporating species absence, are important to assess the effect of land use and climate change on endemic zoonosis spillover to human populations and direct limited public health resources towards regions at greatest risk.
-
-Currently available consolidated datasets on host-pathogen associations (e.g.
-CLOVER, EID2 and GMPD2) do not include spatial or temporal components [@Gibb_data_proliferation].
-The current synthesis of rodent trapping studies has shown that pathogens have only been sparsely tested for within a hosts range.
-Current zoonosis risk models dependent on these sources of data are unable to incorporate spatial heterogeneity in pathogen prevalence across a hosts range.
-Additional uncertainty in current models of zoonotic disease risk arises from host-pathogen associations that have not been reported in these consolidated datasets, for example *Hylomyscus pamfi* and Lassa fever and *R.
-rattus* and Q fever (*Coxiella burnetii*).
-Detection of zoonotic pathogens in multiple co-occurring host species supports the adoption of multi-species to better understand the potential range of endemic zoonoses [@Wilkinson].
-
-Few studies stratified presence-absence or pathogen detection by time at the level of rodent species, limiting observation of the changes in host and pathogen dynamics.
-This limitation prevents calculation of incidence of infection and prevalence of infectious rodents which potentially varies by both time and space [@fichet_calvet_2016].
-Understanding of temporal changes in viral burden and shedding for endemic zoonoses is required to accurately predict current and future risk of pathogen spillover.
-
-Due to data sparsity, we were unable to account for temporal change over the six decades of rodent trapping studies.
-Land use change and population density have changed dramatically over this period in West Africa [@hermann].
-We attempted to mitigate against this by using the median year of trapping to understand the spatial and land use biases in trapping activity, however, it is possible that land use and population density at trapping sites varied importantly between when the rodent trapping study was conducted and the conditions in 2005.
-Despite this limitation, the finding that trapping is biased towards high density, human dominated landscapes is unlikely to significantly change.
-We have shown that synthesis of rodent trapping studies to supplement curated rodent distributions can counteract some of the inherent biases in these data and that they can add further contextual data to host-pathogen association data.
-Together this supports their inclusion in efforts to model endemic zoonotic risk and novel pathogen emergence.
-However, utility can be improved by adopting reporting standards and practices consistent with Open Science [@Foster].
-
-Future rodent trapping studies should be targeted towards regions that are currently under-studied.
-Further information on rodent presence and abundance across West Africa will assist in modelling changing endemic zoonosis risk and the potential for novel pathogen emergence.
-Sharing of disaggregated data alongside research publications should be supported with adoption of data standards to support further data synthesis.
-Inclusion of exact locations of trapping sites, trapping effort and the dates at which trapping occurred would support more detailed understanding of the spatio-temporal dynamics of host populations and the risk of endemic zoonosis spillover events.
-Despite these challenges we propose that rodent trapping studies can provide an important source of data to supplement consolidated datasets on rodent distributions to quantify the risk of endemic zoonosis spillover events and the hazard of novel pathogen emergence.
-
-# References
-
-<div id="refs"></div>
-
-# Supplementary
-<!---BLOCK_LANDSCAPE_START--->
-## Supplementary Table 1
-
-```{r supplementary-table-1}
-
-read_rds(here("tables", "supplementary_table_1.rds")) %>%
-  set_caption(caption = "Supplementary Table 1: Data extraction tool for studies meeting inclusion criteria")
-
-```
-
-## Supplementary Table 2
-
-```{r supplementary-table-2}
-
-read_rds(here("tables", "supplementary_table_2.rds"))
-
-```
-<!---BLOCK_LANDSCAPE_STOP--->
-## Supplementary Table 3
-
-```{r supplementary-table-3-1}
-
-read_rds(here("tables", "supplementary_table_3_1.rds"))
-
-```
-
-```{r supplementary-table-3-2}
-
-read_rds(here("tables", "supplementary_table_3_2.rds"))
-
-```
-
-```{r supplementary-table-3-3}
-
-read_rds(here("tables", "supplementary_table_3_3.rds"))
-
-```
-
-```{r supplementary-table-3-4}
-
-read_rds(here("tables", "supplementary_table_3_4.rds"))
-
-```
-
-```{r supplementary-table-3-5}
-
-read_rds(here("tables", "supplementary_table_3_5.rds"))
-
-```
-
-## Supplementary Figure 1
-
-![Supplementary Figure 1.
-Green points represent the start date of rodent trapping studies, blue points representing the final trapping activity.
-Red points indicate the publication of studies.
-Increasing numbers of studies have been published since 2000 with more studies being conducted over repeated visits.](`r here("figures", "Supplementary_figure_1.png")`)
-
-## Supplementary Figure 2
-
-![Supplementary Figure 2.
-Comparison between proportion of land use between regions trapped and all land use in West Africa.
-Cropland, Shrubland and Grassland were trapped at than representative rates while Bare habitats, Forested regions and Urban landscapes were trapped at lower than representative rates.](`r here("figures", "Supplementary_figure_2.png")`)
-
-## Supplementary Figure 3
-
-![Supplementary Figure 3A.
-Relative trapping effort bias across West Africa from the subset of included studies reporting trapping effort adjusted for proportion urban land cover and proportion tree cover.
-Predictions are limited to areas around trap sites (coloured areas), uncertainty in the estimate of bias is represented by colour transparency/pallor.
-Purple regions represent areas with higher than expected trapping effort, yellow regions represent areas lower than expected trapping effort](`r here("figures", "Figure_2_sensitivity.png")`)
-
-![](`r here("figures", "Figure_2_ss.png")`)
-
-## Supplementary Figure 4
-![](`r here("figures", "Supp_4a_updated.png")`)
-
-![Supplementary Figure 4.
-A) Identified host-pathogen associations at pathogen family level through detection of acute infection (i.e.
-PCR, culture).
-Percentages and colour relate to the proportion of all assays that were positive.
-Associations with a black border are present in the CLOVER dataset.
-B) Identified host-pathogen associations at pathogen family level through serological assays (i.e.
-ELISA).
-Percentages and colour relate to the proportion of all assays that were positive.
-Associations with a black border are present in the CLOVER dataset.](`r here("figures", "Supp_4b_updated.png")`)
